### 2023-05-29 paper/ms.tex
stage ms.tex for version control

\documentclass[12pt,twocolumn,tighten]{aastex63}
% upon AAS submission
%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\turnoffedit

\usepackage{apjfonts}
\usepackage{url}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{amsmath,amstext,amssymb}
\usepackage[caption=false]{subfig} % for subfloat
\usepackage{bm} % for bold hat accents
\usepackage{xcolor, fontawesome}
\usepackage{color}

\newcommand{\rprs}{{$R_p/R_{\star}$}}
\newcommand{\teff}{T$_{\textrm{eff}}$}
\newcommand{\gcc}{{g\,cm$^{-3}$}}
\newcommand{\rstar}{{$R_\star$}}
\newcommand{\mearth}{{M$_\oplus$}}
\newcommand{\rearth}{{R$_\oplus$}}
\newcommand{\rsun}{{R$_\odot$}}
\newcommand{\msun}{{M$_\odot$}}

\newcommand{\minus}{\scalebox{0.5}[1.0]{$-$}}

\newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}
\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}

%
% ms specific numbers
%
\newcommand{\nstarwithages}{NNN}

\defcitealias{McQuillan_2014}{M14}
\defcitealias{Mazeh_2015}{M15}
\defcitealias{Santos_2019}{S19}
\defcitealias{Santos_2021}{S21}

\begin{document}

\title{gyro-johannes}

\correspondingauthor{Luke G. Bouma}
\email{luke@astro.caltech.edu}

\received{---}
\revised{---}
\accepted{---}
\shorttitle{gyro-johannes}

\shortauthors{Bouma et al.}

\input{authors.tex}

\begin{abstract}
  We report rotation- and lithium-based ages for stars observed by
  the primary Kepler mission.  Stellar rotation periods are collected from
  previous work; lithium is measured using both new and archival HIRES
  spectra.
  For the \nstarwithages\ FGK stars with reported ages, the relative statistical
  precisions $\sigma_t / t$ are XX +YY -ZZ\%, which is a factor of X.X
  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
  same stars.  The results include XXX planets younger than 1\,Gyr,
  and YYY planets younger than 0.5\,Gyr.
  In the base sample of dwarf FGK stars observed by
  Kepler, we find that the intrinsic age distribution is(/not)
  consistent with a uniform distribution between X and Y billion years
  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
  detection pipeline, we interpret the majority of the change to be
  caused by selection effects (or astrophysics).  We also comment on a
  few previously reported evolutionary trends in the size evolution of
  close-in mini-Neptunes (2-4\,\rearth) and super-Earths
  (1-2\,\rearth), a few of which we recover at comparable
  ($\sim$3$\sigma$) significance to previous work, and one of which
  we note for the first time.
\end{abstract}

\keywords{Exoplanet evolution (491), Stellar ages (1581)}

\section{Introduction}
\label{sec:intro}

Discovering a young planet requires solving two problems: find the
planet, and measure the star's age.  Each problem can be
solved in different ways, yielding many possible permutations for
discovery.  In this manuscript, we will focus on the transit method
for planet discovery, and rotation and lithium-based methods for stellar age
measurement.  Rotation-based age-dating, also known as
gyrochronology, will provide the most information for the largest
number of stars.  The lithium-based technique will provide an
important sanity check for the subset of stars younger than a few
hundred million years.

To help build intuition, it is useful to assume that the age
distribution of stars within the nearest few hundred parsecs of the
Sun (i.e.~in the Solar Neighborhood) is uniform between 0 and 10\,Gyr
\citep{Nordstrom_2004}.  We will revisit this assumption later in the
manuscript.  While the youngest 0.1\% of stars ($\lesssim$10\,million
years; Myr) are perhaps the most constraining for studies of planet
formation, they are extremely challenging targets for close-in
exoplanet detection 
\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
the early orbital, structural, and atmospheric evolution of exoplanets
are perhaps most technically feasible for the $\approx$1\% of stars
with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
younger than 1\,Gyr.  However, since stellar ages are roughly
uniformly distributed in the solar neighborhood, there is a limit on
the sample size of young planets that will be detectable in these age
regimes, particularly relative to their older breathren.

%TODO mention the RV approach?  at least the Hyades etc work...
A wide range of approaches have, over the past ten years, begun to
establish the subfield of young, close-in planet discovery.
Pioneering discoveries with the prime Kepler mission (hereafter,
``Kepler'') focused on stars with known ages---stars in clusters---and
searched them for transiting planets \citep{Meibom_2013}.  Planets
discovered in this way are assumed to have the same age as the host
cluster, which are the standards for the current astrophysical age
scale.  The resulting ages are precise at the $\approx$10\% level.  K2
and TESS have yielded larger numbers of young planets in clusters than
Kepler, because each of these missions has observed a larger fraction
of the sky, and therefore more stars with known ages
\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.

An important limitation of the ``cluster-first'' approach is that only
$\approx$1\% of stars within the nearest few hundred parsecs can
currently be associated with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
This means that there are many ``young'' 0.05-1\,Gyr stars in the
field that were not part of the selection function of these searches.
Roughly 40 transiting planets are currently known to be younger than
1\,Gyr, with an age precision $t/\sigma_t > 3$ (see
Figure~\ref{fig:rp_period_age}).  However, Kepler discovered
$\approx$4{,}000 planets; and so barring a major discovery bias
against young planets, we would expect a significant portion of the
Kepler planets---perhaps 10\%, or 400 planets---to be younger than
1\,Gyr.  If one is willing to compromise on age precision, those
planets might be identifiable.

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{f1.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Radii, orbital periods, and ages of transiting exoplanets}.
    Planets younger than 1\,Gyr with ages more precise than a
    factor of three are emphasized. 
    ({\bf cite relevant papers?})
    Parameters are from the \citet{PSCompPars}.
		\label{fig:rp_period_age}
	}
\end{figure}

Have these young Kepler planets already been found?  If they have, the
most likely studies to have found them would be those by
\citet{Berger_2020b_rpage}, \citet{David_2021} or
\citet{Petigura_2022}.  The Berger and Petigura studies focused on
isochronal age-dating of individual stars, which is most precise in
the context of Kepler for stars that are beginning to evolve off the
main-sequence.  The \citet{David_2021} study leveraged both isochronal
ages from an earlier analogous study
\citep{Fulton_Petigura_2018_cks_vii}, and (in their core analysis) a
relative gyrochronal age-ranking approach that sorted stars based on
whether they rotated faster or slower than a gyrochrone age of
$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.

Theoretical predictions for planet evolution in the Kepler sample are
most important in the context of mini-Neptunes and super-Earths, since
these are the most abundant planets.  Pertinent predictions include:
{\it a)} that mini-Neptunes should cool and contract, roughly
following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
b)} that large and close-in mini-Neptunes at the edge of the
photoevaporation desert should shrink over the atmospheric loss
timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
should decrease over $t_{\rm loss}$---the upper-envelope of the
mini-Neptune size distribution should corresponding decrease,
e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
$\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
\citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
\citep[e.g.][]{Rogers_2021} {\bf BY HOW MUCH?}); {\it e)} that the
upper boundary of the super-Earth population should increase over the
same timescale as $\theta$ is increasing, since more massive
super-Earth cores should be able to retain their envelopes for longer
than less massive cores.

The details of the predictions are flexible, in that they depend on
the distributions of planetary core masses, initial atmospheric mass
fractions, and initial entropies of the atmospheres after disk
dispersal.  For instance, different sources of heating---from the star
\citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
\citep{Gupta_2019}, or even giant impacts
\citep{Biersteker_Schlichting_2019}---can produce very different
histories for the size evolution of planets at these early times
\citep[e.g.][]{Owen_2020}.  The generic prediction that the ``upper
envelope'' of the mini-Neptune size distribution should shrink due to
Kelvin-Helmholtz cooling is likely hard to avoid; predictions however
concerning the ``radius valley'' and movement of planets across it are
dependent on the timescales and prevalence of atmospheric mass loss.
In specific scenarios for initial atmospheric accretion, the radius
gap might form without needing to invoke mass loss at all
\citep{Lee_2022}, in which case there could be very little evolution
of the planetary size distribution between 1 and 4\,\rearth\ over the
first gigayear.

Can stellar age measurements provide lines of evidence for or against
the different models?  Neither {\it (a)} nor {\it (b)} has yet had a
convincing---or even suggestive--- detection reported (see
\citealt{Petigura_2022}).  While the apparent over-abundance of large
very young planets ({\it (c)}) has been noted by multiple authors
\citep[e.g.][]{Bouma_2020,Mann_2022}, its presence is almost certainly
exacerbated by a selection effect against sub-Neptune sized planet
detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  And finally,
{\it (d)} and {\it (e)} have been explored by
\citet{Berger_2020b_rpage} and the twin \citet{David_2021} and
\citet{Sandoval_2021} studies.  Both \citet{Berger_2020b_rpage} and
\citet{Sandoval_2021} reported that $\theta$ increases with time, at a
statistical significance of $\approx$2-3$\sigma$.  For
\citet{Berger_2020b_rpage}, this conclusion was reached by comparing
the sizes of 85 planets with median isochronal ages younger than
1\,Gyr with a property-matched sample of planets older than 1\,Gyr.
In \citet{Sandoval_2021}, the conclusion was reached by computing the
ratio while performing a Monte Carlo re-sampling procedure from the
isochronal ages and planetary sizes computed by
\citet{Fulton_Petigura_2018_cks_vii}.  Finally, effect {\it (e)} was
reported by \citet{David_2021}, who found that the average location of
the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
$\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.

The fact that stellar ages---and especially isochronal ages---are
correlated with stellar masses and metallicities is a source of stress
for many in this subfield.  \citet{Sandoval_2021} for instance noted a
strong trend in $\theta$ as a function of stellar metallicity in
addition to the trend they noted in age, and cautioned that this could
also explain their observations.  Independently, \citet{Petigura_2018}
and \citet{Petigura_2022} have reported trends between small-planet
occurrence and both stellar metallicity and mass; these parameters
seem, at first glance, to be more important predictors of bulk changes
in the planetary population than age.  Given these known correlations,
it would be interesting to consider population models in which
variables such as metallicity, mass, and age were added in succession
\citep[e.g.][]{Thorngren_2021}.  For such an analysis to be effective,
the measurement precision of these different parameters would ideally
be as high as possible.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 0.5-2.5\,Gyr, and can also provide useful age limits at
earlier times \citep{Bouma_2023}.  The idea of using a star's
spin-down as a clock is quite old
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
for the present analysis is the fact that we can empirically calibrate
rotation-based ages using measured sequences of rotation periods
in open clusters
(\citealt{Bouma_2023}, and references therein).  This approach
ultimately ties the rotation age scale to the cluster age scale, and
is limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

Lithium-based age-dating includes two qualitatively distinct regimes:
depletion boundary ages derived for M-dwarfs in star clusters, and the
less precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the gradual empirical observation that the Li abundances of
FGK stars decline as a function of time, for reasons different from
the standard convective mixing and burning scenario (CITE, CITE).
Empirically, this decline has recently been documented by
\citet{Jeffries_2023}, who built a model for how the equivalent width
(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
stellar effective temperature and age for a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

The methods by which we select the stellar and planet samples are 
discussed in Section~\ref{sec:selection}.
The age-dating methods are described and validated in
Section~\ref{sec:agemethod}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

\section{Sample Selection}
\label{sec:selection}

This work is focused on stars surveyed by Kepler for which ages can be
inferred using either stellar rotation, lithium, or both.  These
``age-dateable stars'', $\mathcal{S}$, are a relatively small subset
of the $\approx$160{,}000 Kepler targets.  Rotation periods provide
the majority of the information, since they have been reported for
$\approx$55{,}000 Kepler stars
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  Measurements of the
\ion{Li}{1} 6708\,\AA\ doublet require high-resolution
($R\gtrsim$10{,}000) spectra, and are only available for a few
thousand Kepler objects of interest (KOIs).  We therefore choose to
define our selection function as Kepler stars with measured
rotation periods ($\mathcal{S}$; Section~\ref{subsec:starsel}), and
Kepler planets whose host stars have measured rotation periods
($\mathcal{P}$; Section~\ref{subsec:planetsel}).

\subsection{Star Selection}
\label{subsec:starsel}

We select stars with rotation periods as determined by previous
investigators.  Studies of stellar rotation across the entire Kepler field
include those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
\citet{Reinhold_2015},
\citet{Santos_2019} (\citetalias{Santos_2019}), and
\citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{McQuillan_2014} used an approach
based on the autocorrelation function to detect 34{,}030 rotation
periods for main-sequence Kepler targets cooler than 6500\,K, and excluded
known eclipsing binaries and KOIs.  
\citet{Reinhold_2015} used an iterative Lomb-Scargle based approach to
analyze $\approx$40{,}000 stars with a variability range $R_{\rm
	var}>0.3$\% that were not known EBs, planet candidates, or pulsators,
and reported primary rotation periods for 24{,}124 of these stars.
\citet{Reinhold_2015} also reported secondary periods for two thirds
of the stars with primary periods, which could be caused by
differential rotation or finite spot lifetimes.  
Finally,
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
KOIs and binaries, and assigned them specific quality flags.  
The rotation periods of KOIs have received considerable additional
scrutiny
\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.

%After reviewing the relevant literature and considering the nature of
%the rotation period detection problem,
We ultimately chose to adopt
the results of \citetalias{Santos_2019} and \citetalias{Santos_2021}
as our default rotation periods.  \citetalias{Santos_2021} provides a
detailed comparison against \citetalias{McQuillan_2014}; the brief
summary is that the periods agree for 99.0\% of the 31{,}038 period
detections in common between the two studies.
\citetalias{Santos_2021} classified the 2{,}992 remaining stars from
\citetalias{McQuillan_2014} as ``non-periodic'' based on updated
knowledge of contaminants (e.g.~giant stars and eclipsing binaries)
and visual inspection.  
In addition, 
\citetalias{Santos_2021} report rotation periods for 24{,}182
main-sequence and subgiant FGKM stars that were not reported as
periodic by \citetalias{McQuillan_2014}.
Our interpretation is that these detections are likely more challenging than
those from \citetalias{McQuillan_2014}, perhaps due to a lower
variability amplitude, or a longer rotation period.
A relevant flag for stars missing \citetalias{McQuillan_2014} periods
is included in Table~\ref{tab:stars}.

\subsection{Planet Selection}
\label{subsec:planetsel}

\subsection{The Stellar Samples}
We will refer in this manuscript to a few samples, defined in detail
below.  The gist of each is as follows.
\begin{itemize}
  \item Kepler objects of interest: as defined in the Cumulative KOI
    table at the NASA Exoplanet Archive from 2022 June 27, which is
    distinct from the \texttt{q1\_q17\_dr25} table
    \citep{Thompson_2018}.
\end{itemize}

\subsection{Stellar Properties}

\paragraph{Temperatures}
\paragraph{Surface Gravities}

\subsection{Rotation Periods}

% TODO: add that plot showing differences btwn authors

\subsection{Lithium}

\section{Age Measurement}
\label{sec:agemethod}

\section{Results}
\label{sec:results}

\section{Discussion}
\label{sec:disc}

\subsection{Isochrone Age Comparison}

\subsection{Lithium Age Comparison}

\subsection{Gyrochronal Age Comparison}
Searching the literature for gyrochronal analyses of the Kepler field,
the most relevant studies seemed to be those of
\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
\citet{David_2021}.

\subsection{Asteroseismic Age Comparison}
T Ceillier, J Van Saders et al 2016 MNRAS...

\section{Conclusions}
\label{sec:conclusions}

\acknowledgements
This work was supported by the 
Heising-Simons 51~Pegasi~b Fellowship (LGB)
and the Arthur R.~Adams SURF Fellowship (EKP).

L.G.B.~conceived the project, collected HIRES spectra, executed the
rotation- and lithium-based age analyses, and drafted the manuscript.
E.K.P.~contributed to the rotation-based age analysis.
L.A.H.~contributed to project design.
H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
the HIRES data.
All authors assisted in manuscript revision.

\facilities{
  Gaia \citep{Gaia_DR3_2022},
  Kepler \citep{Borucki10},
  TESS \citep{ricker_transiting_2015},
  NGTS \citep{Wheatley_2018}
}

\software{
    astropy \citep{Astropy18},
    matplotlib \citep{matplotlib},
    numpy \citep{numpy},
    scipy \citep{scipy},
}

%\clearpage

\bibliographystyle{aasjournal}
\bibliography{bibliography}

%\appendix
%\section{Can you avoid appendices?}
%\label{app:interp}

\clearpage
\listofchanges

\end{document}

### 2023-05-29 paper/collect.sh

@@ -1,0 +1,20 @@
+resultsdir="/Users/luke/Dropbox/proj/gyro-interp/results"
+cp $resultsdir/prot_vs_teff/prot_vs_teff_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_NGC-6819_Ruprecht-147_slowseq_poly7__100_to_2600_interpmethodpchip_m67.pdf f1a.pdf
+cp $resultsdir/prot_vs_teff/prot_vs_teff_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_models_poly7_α_Per_120-Myr_300-Myr_Praesepe_NGC-6811_interpmethodpchip_m67.pdf f1b.pdf
+
+cp $resultsdir/data_vs_model_prot/data_vs_model_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_models_poly7_α_Per_120-Myr_300-Myr_Praesepe.pdf f2.pdf
+
+cp $resultsdir/multi_age_posteriors_mcmc/age_posteriors_Teff5800_Prot2.0_3.5_5.0_6.5_8.0_9.5_11.0.pdf f3a.pdf
+cp $resultsdir/multi_age_posteriors_mcmc/age_posteriors_Teff5000_Prot4.0_5.5_7.0_8.5_10.0_11.5_13.0.pdf f3b.pdf
+cp $resultsdir/multi_age_posteriors_mcmc/age_posteriors_Teff4200_Prot9.0_10.5_12.0_13.5_15.0_16.5.pdf f3c.pdf
+
+cp $resultsdir/empirical_limits_of_gyrochronology/both_empirical_limits_of_gyrochronology_nNone_slowseq_poly7_100_200_300_400_500_600_700_800_900_1000_1100_1200_1300_1400_1500_1600_1700_1800_1900_2000_2100_2200_2300_2400_2500_2600_scale-default_pchip_m67_fine.pdf f3d.pdf
+
+cp $resultsdir/prot_vs_time_fixed_teff/prot_vs_time_teff5800.0_skumanich_fix_n_0.47_skumanich_vary_n_1d_linear_1d_quadratic_1d_pchip_pchip_m67_MH08_A19_SL20_xscalelog.pdf sf1a.pdf
[…]

### 2023-06-02 paper/ms.tex
update ms

@@
 %
+
+%%%%%%%%%
+% STARS %
+%%%%%%%%%
+% number of stars monitored by Kepler (quoting Santos19/21)
+\newcommand{\nkeplerstars}{$\approx$160{,}000}
+% fraction of kepler stars w/ rotation periods
+\newcommand{\nstarswithprot}{{\bf $\sim$50{,}000}}
+% fraction of stars with rotation periods, with teff/logg in Berger+20
+\newcommand{\fracstarswithprotwithbtwenty}{{\bf $\sim$94\%}}
+% complement of above
+\newcommand{\fracstarswithprotwithoutbtwenty}{{\bf $\sim$6\%}}
+
+%%%%%%%%%%%
+% PLANETS %
+%%%%%%%%%%%
+% number of KOIs in Q1-Q17 DR25 table
+\newcommand{\nkois}{{\bf $\sim$8{,}000}} %FIXME precise
+% number of KOIs in Q1-Q17 DR25 table that are not known false positives
+\newcommand{\nkoisnofp}{{\bf $\sim$4{,}000}} %FIXME precise
+% number of KOIs in Q1-Q17 DR25 table that have rotation periods
+\newcommand{\nkoiswithprot}{{\bf $\sim$2{,}000}}
+% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
+\newcommand{\nkoisnofpwithprot}{{\bf $\sim$1{,}000}}
+
+\newcommand{\fracplanetswithprotwithbtwenty}{{\bf $\sim$94\%}}
+
 \newcommand{\nstarwithages}{NNN}
@@
 ``age-dateable stars'', $\mathcal{S}$, are a relatively small subset
-of the $\approx$160{,}000 Kepler targets.  Rotation periods provide
+of the \nkeplerstars\ Kepler targets.  Rotation periods provide
 the majority of the information, since they have been reported for
@@
 
-\subsection{Star Selection}
+\subsection{Stars}
 \label{subsec:starsel}
@@
 %the rotation period detection problem,
-We ultimately chose to adopt
-the results of \citetalias{Santos_2019} and \citetalias{Santos_2021}
-as our default rotation periods.  \citetalias{Santos_2021} provides a
-detailed comparison against \citetalias{McQuillan_2014}; the brief
-summary is that the periods agree for 99.0\% of the 31{,}038 period
-detections in common between the two studies.
-\citetalias{Santos_2021} classified the 2{,}992 remaining stars from
-\citetalias{McQuillan_2014} as ``non-periodic'' based on updated
-knowledge of contaminants (e.g.~giant stars and eclipsing binaries)
-and visual inspection.  
-In addition, 
+We chose to adopt the results of \citetalias{Santos_2019} and
+\citetalias{Santos_2021} as our default rotation periods.
+\citetalias{Santos_2021} provides a detailed comparison against
+\citetalias{McQuillan_2014}; the brief summary is that the periods
+agree for 99.0\% of the 31{,}038 period detections in common between
+the two studies.  \citetalias{Santos_2021} classified the 2{,}992
+remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
+based on updated knowledge of contaminants (e.g.~giant stars and
+eclipsing binaries) and visual inspection.  In addition,
 \citetalias{Santos_2021} report rotation periods for 24{,}182
 main-sequence and subgiant FGKM stars that were not reported as
-periodic by \citetalias{McQuillan_2014}.
-Our interpretation is that these detections are likely more challenging than
-those from \citetalias{McQuillan_2014}, perhaps due to a lower
-variability amplitude, or a longer rotation period.
-A relevant flag for stars missing \citetalias{McQuillan_2014} periods
-is included in Table~\ref{tab:stars}.
+periodic by \citetalias{McQuillan_2014}.  Our interpretation is that
+these detections are likely more challenging than those from
+\citetalias{McQuillan_2014}, perhaps due to a lower variability
+amplitude, or a longer rotation period.  A relevant flag for stars
+missing \citetalias{McQuillan_2014} periods is included in
+Table~\ref{tab:stars}.
 
-\subsection{Planet Selection}
+Concatenating the results of \citetalias{Santos_2019} and
+\citetalias{Santos_2021} yields \nstarswithprot\ Kepler stars with
+rotation periods.
+%FIXME
+
+\subsection{Planets}
 \label{subsec:planetsel}
 
-\subsection{The Stellar Samples}
-We will refer in this manuscript to a few samples, defined in detail
-below.  The gist of each is as follows.
-\begin{itemize}
-  \item Kepler objects of interest: as defined in the Cumulative KOI
-    table at the NASA Exoplanet Archive from 2022 June 27, which is
-    distinct from the \texttt{q1\_q17\_dr25} table
-    \citep{Thompson_2018}.
-\end{itemize}
+We considered two different samples of planets: those in the fully
+automated Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which are
+suitable for planet occurrence rate calculations, and also those in
+the cumulative KOI table, which includes the best knowledge available
+on any given planet candidate, but incorporates human-based vetting
+\footnote{See
+\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
+For the main body of this work, we will refer to the former
+homogeneous sample, which includes \nkois\ KOIs, \nkoisnofp\ that are
+not known false positives;
+Appendix~\ref{app:cumulativekoi} highlights some additional young
+planets that are omitted in this approach.
+
+%FIXME: add a sentence or two describing the numbers after
+%crossmatching against S19/S21 --
+
+Crossmatching the Q1-Q17 DR25 KOI table against our stars with
+rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of which \nkoisnofpwithprot\ 
+are not known false postives.
 
@@
 
-\paragraph{Temperatures}
-\paragraph{Surface Gravities}
+% https://gea.esac.esa.int/archive/documentation/GDR3/Data_analysis/chap_cu8par/sec_cu8par_apsis/ssec_cu8par_apsis_gspphot.html
 
-\subsection{Rotation Periods}
+As our default source for stellar temperatures and surface gravities, we adopt
+the values from the Gaia-Kepler Stellar Properties Catalog (GKSP;
+\citealt{Berger_2020a_catalog}).
+The GKSP parameters are available for stars with ``AAA'' 2MASS photometry,
+measured parallaxes in Gaia DR2,  and g-band photometry from either the KIC or
+the Kepler-INT survey.
+The parameters themselves were derived using \texttt{isoclassify} (CITE
+Huber2017) to interpolate over the MIST isochrone grids (CITE Choi2016,
+Dotter2016), given the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
+parallexes, and $V$-band extinction from the (CITE) Green2019 reddening map.
+The resulting stellar parameters are available for
+\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars with
+rotation periods.
+For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
+temperatures and surface gravities, we adopt the values 
+from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
+GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
+and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
+Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
+metallicities, and line-of-sight extinctions.
+TODO: plot uncertainties in teff/logg for both parameter sets.
 
@@
 \subsection{Lithium}
+
+%FIXME: left off here for discussion of HIRES spectra.
 
@@
 \listofchanges
-
-\end{document}

### 2023-06-03 paper/ms.tex
update ms

@@
-\documentclass[12pt,twocolumn,tighten]{aastex63}
 % upon AAS submission
@@
 % drafting / arxiv
-%\turnoffedit
+\documentclass[11pt,twocolumn,tighten]{aastex63}
+\turnoffedit
 
@@
 \usepackage{amsmath,amstext,amssymb}
-\usepackage[caption=false]{subfig} % for subfloat
-\usepackage{bm} % for bold hat accents
 \usepackage{xcolor, fontawesome}
@@
 \newcommand{\rprs}{{$R_p/R_{\star}$}}
+\newcommand{\vsini}{{$V \sin i$}}
 \newcommand{\teff}{T$_{\textrm{eff}}$}
+\newcommand{\kms}{{km\,s$^{-1}$}}
 \newcommand{\gcc}{{g\,cm$^{-3}$}}
 \newcommand{\rstar}{{$R_\star$}}
+\newcommand{\rhostar}{{$\rho_\star$}}
 \newcommand{\mearth}{{M$_\oplus$}}
@@
 
+\newcommand{\eg}{e.g.,}
+
+\newcommand{\mstype}{letter}
+
+%%%%%%%%%%%%%%%%
+% INSTITUTIONS %
+%%%%%%%%%%%%%%%%
+\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
+\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics, 77 Massachusetts Avenue, Cambridge, MA 02139}
 \newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}
-\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 % fraction of kepler stars w/ rotation periods
-\newcommand{\nstarswithprot}{{\bf $\sim$50{,}000}}
+\newcommand{\nstarswithprot}{{$\sim$50{,}000}}
 % fraction of stars with rotation periods, with teff/logg in Berger+20
-\newcommand{\fracstarswithprotwithbtwenty}{{\bf $\sim$94\%}}
+\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
 % complement of above
-\newcommand{\fracstarswithprotwithoutbtwenty}{{\bf $\sim$6\%}}
+\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}
 
@@
 % number of KOIs in Q1-Q17 DR25 table
-\newcommand{\nkois}{{\bf $\sim$8{,}000}} %FIXME precise
+\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
 % number of KOIs in Q1-Q17 DR25 table that are not known false positives
-\newcommand{\nkoisnofp}{{\bf $\sim$4{,}000}} %FIXME precise
+\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
 % number of KOIs in Q1-Q17 DR25 table that have rotation periods
-\newcommand{\nkoiswithprot}{{\bf $\sim$2{,}000}}
+\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
 % number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
-\newcommand{\nkoisnofpwithprot}{{\bf $\sim$1{,}000}}
+\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
+% fraction of KOIs that are not FPs with B+20 parameters
+\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}
 
-\newcommand{\fracplanetswithprotwithbtwenty}{{\bf $\sim$94\%}}
+%%%%%%%%%%%
+% LITHIUM %
+%%%%%%%%%%%
+% number of KOIs that are not known FPs that have a HIRES spectrum
+\newcommand{\nkoiswithhires}{{1{,}500}}
+% number of KOIs for which this work and Berger2018 overlap
+\newcommand{\nbergeroverlap}{{500}}
 
-\newcommand{\nstarwithages}{NNN}
+%%%%%%%%
+% AGES %
+%%%%%%%%
+\newcommand{\nkepstarswithages}{NNN}
+\newcommand{\nkoisnofpwithages}{NNN}
+\newcommand{\nkoisgoldwithages}{NNN}
 
+%%%%%%%%%%%%%
+% OFT-CITED %
+%%%%%%%%%%%%%
 \defcitealias{McQuillan_2014}{M14}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
 
-\title{gyro-johannes}
+\title{Gyrochronal and Lithium Ages for Stars and Planets Observed by Kepler}
 
@@
 
-\input{authors.tex}
+\author[0000-0002-0514-5538]{Luke~G.~Bouma}
+\altaffiliation{51 Pegasi b Fellow}
+\affiliation{\caltech}
+
+\author{Authors to include (ORDER AFTER LGB NOT YET ASSIGNED MEANING)}
+
+\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
+\affiliation{\caltech}
+
+\author{Lynne~A.~Hillenbrand}
+\affiliation{\caltech}
+
+%
+% HIRES Collaborators
+%
+\author[0000-0002-0531-1073]{H. Isaacson}
+\affiliation{\berkeley}
+%
+\author[0000-0001-8638-0320]{A. W. Howard}
+\affiliation{\caltech}
+%
+\author{HIRES observers to be named}
+\affiliation{Various}
 
 \begin{abstract}
-  We report rotation- and lithium-based ages for stars observed by
+  We report rotation- and lithium-based ages for stars and planets observed by
   the primary Kepler mission.  Stellar rotation periods are collected from
-  previous work; lithium is measured using both new and archival HIRES
+  previous analyses of the Kepler light curves; lithium is assessed directly from HIRES
   spectra.
-  For the \nstarwithages\ FGK stars with reported ages, the relative statistical
-  precisions $\sigma_t / t$ are XX +YY -ZZ\%, which is a factor of X.X
+  wWe report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
+  precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
   +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
@@
 
-\subsection{Stars}
+\subsection{Kepler stars with rotation periods}
 \label{subsec:starsel}
@@
 
-\subsection{Planets}
+\subsection{Kepler objects of interest}
 \label{subsec:planetsel}
@@
 
-%FIXME: add a sentence or two describing the numbers after
-%crossmatching against S19/S21 --
-
 Crossmatching the Q1-Q17 DR25 KOI table against our stars with
@@
 are not known false postives.
+These objects will comprise the core data for our gyrochronology analysis.
 
-\subsection{Stellar Properties}
+\subsection{Lithium sample}
+A final subcomponent of our analysis involves an age assessment based on
+the equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
+component of the work, we primarily rely on spectra acquired with the
+High Resolution Echelle Spectrometer (HIRES;
+\citealt{vogt_hires_1994}) on the Keck I 10m telescope.
+The majority of these spectra are archival, and were acquired
+as part of the California Keck Survey (CKS).
+A subset were acquired by our team in our own observing programs.
+%TODO quantify.
+%TODO compare to berger2018?
+
+These HIRES spectra are available for \nkoiswithhires\ of the
+\nkoisnofp\ non-false-positive KOIs.
+In the broader stellar sample, they are available for a much smaller 
+fraction of the stars observed by Kepler, because they were generally only
+acquired due to the presence of a transiting planet candidate.
+Given this selection function,
+we leverage these spectra in our analysis as a way to scrutinize the
+ages of just the planet sample.
+
+\section{Stellar Properties}
 
@@
 The resulting stellar parameters are available for
-\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars with
-rotation periods.
+\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars
+with rotation periods.
 For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
@@
 metallicities, and line-of-sight extinctions.
-TODO: plot uncertainties in teff/logg for both parameter sets.
+In the planet sample,
+\frackoisnofpwithprotwithbtwenty\ of the
+the non-false-positive KOIs with rotation periods have parameters from
+\citet{Berger_2020a_catalog}.
 
-% TODO: add that plot showing differences btwn authors
+As a sanity check on our effective temperature scale, we then used the 
+the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
+BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
+relation suggested by \citet{Curtis_2020}.
+This latter effective temperature scale was demonstrated to have a 
+scatter of $\approx$50\,K for stars near the main-sequence.
+We find {\bf X, Y, Z.}
+We ultimately favor the \citet{Berger_2020a_catalog} scale
+because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
+isochronal fit.
 
-\subsection{Lithium}
+%TODO: plot uncertainties in teff/logg for both parameter sets.
 
-%FIXME: left off here for discussion of HIRES spectra.
+% TODO: add that plot showing differences btwn the two teff scales?
 
@@
 \label{sec:agemethod}
+
+\subsection{Gyrochrone ages}
+To calculate gyrochrone ages, we use \texttt{gyro-interp}
+\citep{Bouma_2023}.  This framework was built to statistically address
+the issue that, especially over the first gigayear of stellar
+spin-down, stars with the same mass and same rotation period can have
+very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
+gyrochrone age posterior should therefore incorporate this intrinsic
+population-level scatter into the precision with which the age can be
+measured.  The three main assumptions behind the \texttt{gyro-interp}
+framework are as follows.  {\it 1)} The stars are on the main
+sequence, {\it 2)} they do not have any binary companions,
+and {\it 3)} the rotation period distribution
+for field stars of a given mass and age is identical to that of
+the calibration open clusters (metallicity differences, for instance, are
+ignored).  
+Regarding the former two assumptions:
+pre-main-sequence and post-main-sequence evolution introduce
+variations in the stellar angular momentum budget.  Similarly, tidal
+effects might systematically spin up (or down) stars in binaries.
+
+Using our adopted effective temperatures and rotation periods, we ran 
+\texttt{gyro-interp} over an age grid of 0 to 5\,Gyr, linearly spaced
+over 500 grid points.
+\texttt{gyro-interp} itself is only calibrated out to 4\,Gyr, the
+adopted age of the M67 cluster \citep[see][]{Dungee_2022,Gruner_2023}.
+We used the spin-down law between the reference open clusters by
+interpolating using piecewise cubic hermite interpolating polynomials,
+and integrated Equation~1 of \citep{Bouma_2023} at the default grid
+resolution.
+This yielded age posterior probabilities over our requested age grid
+for \nkepstarswithages\ Kepler stars, and \nkoisnofpwithages\
+non-false-positive KOIs.
+
+\subsection{Lithium ages}
+To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
+with HIRES spectra, we use the procedure described by
+\citet{Bouma_2021}.
+The stars of interest for this work are FGK stars, and so the
+continuum in the vicinity of the
+\ion{Li}{1} 6708\,\AA\ feature is well-defined.
+We therefore numerically integrate a single best-fit Gaussian over a
+local window, and estimate the uncertainties on the line width through
+a Monte Carlo procedure that bootstraps against the local scatter in the
+continuum.
+Generally speaking, this approach does not correct for the neighboring
+\ion{Fe}{1} 6707.44\,\AA\ blend.
+
+To evaluate the accuracy and precision of this approach, after
+applying an initial iteration of this method on the \nkoiswithhires\
+KOIs with HIRES spectra, we compared our lithium equivalent widths
+with those reported by \citet{Berger_2018}.
+For the \nbergeroverlap\ stars in both samples, 
+we found {\bf X, Y, Z.}
+
+To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
+{\bf TODO: explain why}
+%TODO: explain why.
+
+\subsection{Quality flags}
+In our core gyrochrone analysis, we calculate gyrochrone age
+posteriors for all the stars with reported rotation periods and
+effective temperatures.  To select the stars for which we expect the
+gyrochronal age to be valid, we then build a set of quality flags, and
+apply them in the sub-analyses that follow.  We defined these filters
+as follows, each of which is included in Tables~\ref{tab:stars}
+and~\ref{tab:planets}.
+
+{\it Subgiants}---We flagged stars as possible subgiants if they had
+$\log g < 4.2$.  We determined this threshold by examining the
+relevant flag from \citet{berger_2018_radii_evolnstates},
+cross-matched against \citet{Berger_2020a_catalog}.
+\citet{berger_2018_radii_evolnstates} performed this classification
+using the PARSEC evolutionary models, stellar radii, and DR25 stellar
+properties catalog effective temperatures \citep{Mathur_2017}.
+
+{\it Photometric binaries}---We plotted color--absolute magnitude
+diagrams using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
+BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
+loci to flag over or under-luminous stars in each diagram, typically
+more than $\approx$0.5 magnitudes above or below the main sequence.
+%TODO refer to a figure!!!
+
+{\it RUWE}---Following recommendations in the Gaia DR3
+documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
+we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
+exceeding 1.4 as possible binaries.  Such astrometric outliers can be
+either bonafide astrometric binaries, or more often are marginally
+resolved point sources for which the single-source PSF model used in
+the Gaia pipeline induces spurious astrometric scatter.
+
+{\it Crowding}---We searched the Gaia DR3 point source catalog for
+stars within 1 Kepler pixel (4$''$) of every target star.
+While many such companions are not physically associated with the
+target star, their presence can bias stellar parameters and stellar
+rotation period measurements.
+We therefore flag any stars with neighbors down to 10$|times$ the 
+brightness of the target star within this region ($\Delta G < 2.5$).
+
+{\it Gaia DR3 Non-Single-Stars}---The Gaia DR3
+\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
+flags known eclipsing, astrometric, and spectroscopic binaries.  We
+directly include this column.
+
+{\it Final calibration sample}---{\bf todo: mention any extra flags
+from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
+amplitude cuts you might want to apply to the rotation period catalog}
+%The combination of the filters
+%described above yields the set of stars that show no evidence for
+%binarity or crowding.  However, some of the rotation period analyses
+%in Table~\ref{tab:clusters} include additional relevant quality flags.
+%For instance, light curves showing multiple photometric periods can
+%indicate unresolved binarity.  We used all relevant filters available
+%from the original authors if they were designed to select single stars
+%with reliable rotation periods.  The final combination of these
+%filters with our own flag for possible binarity yields our sample of
+%benchmark rotators.
 
@@
 \listofchanges
+
+\end{document}

### 2023-06-05 paper/ms.tex
update ms

@@
 mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
-\citep[e.g.][]{Rogers_2021} {\bf BY HOW MUCH?}); {\it e)} that the
+\citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
+decreasing by at least a factor of two or three, and the super-Earth
+abundance correspondingly increasing {\bf BY HOW MUCH?}); {\it e)} that the
 upper boundary of the super-Earth population should increase over the
@@
 with rotation periods.
+
+%For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
+%temperatures and surface gravities, we adopt the values 
+%from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
+%GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
+%and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
+%Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
+%metallicities, and line-of-sight extinctions.
+%In the planet sample,
+%\frackoisnofpwithprotwithbtwenty\ of the
+%the non-false-positive KOIs with rotation periods have parameters from
+%\citet{Berger_2020a_catalog}.
+
 For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
-temperatures and surface gravities, we adopt the values 
-from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
-GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
-and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
-Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
-metallicities, and line-of-sight extinctions.
-In the planet sample,
-\frackoisnofpwithprotwithbtwenty\ of the
-the non-false-positive KOIs with rotation periods have parameters from
-\citet{Berger_2020a_catalog}.
+temperatures and surface gravities, we adopt the values reported by
+\citet{Santos2019} and \citet{Santos2021}, which for these cases
+primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
+Properties Catalog, and are mostly derived from photometry.  In the
+planet sample, \frackoisnofpwithprotwithbtwenty\ of the the
+non-false-positive KOIs with rotation periods have parameters from
+\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
+from DR25.
 
@@
 
+\subsection{Evolutionary trends in time}
+
+{\it Kelvin-Helmholtz cooling}
+$R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}
+
+{\it Early carving of the photoevaporation desert}
+\citep{Owen_Lai_2018}
+
+{\it Rapidly decreasing abundance of big puffy planets}
+Kepler not enough?
+
+{\it Mini-Neptunes turning into Super-Earths}
+\citep[e.g.][]{Rogers_2021}
+
+{\it Boundary of the Super-Earth population moving up in time}
+\citep{David_2021}
+
 \section{Discussion}

### 2023-06-05 paper/ms.tex
minor

@@
 {\it Subgiants}---We flagged stars as possible subgiants if they had
-$\log g < 4.2$.  We determined this threshold by examining the
+$\log g < 4.2$.  We set this threshold by examining the
 relevant flag from \citet{berger_2018_radii_evolnstates},
@@
 rotation period measurements.
-We therefore flag any stars with neighbors down to 10$|times$ the 
+We therefore flag any stars with neighbors down to 10$\times$ the 
 brightness of the target star within this region ($\Delta G < 2.5$).

### 2023-06-06 paper/ms.tex
minor ms update

@@
 temperatures and surface gravities, we adopt the values reported by
-\citet{Santos2019} and \citet{Santos2021}, which for these cases
+\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
 primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
@@
 
+\subsection{Ages for stars in the Kepler field}
+%TODO: show parameter distributions?
+
+%e.g. hist_samples_field_gyro_ages_hist_field_gyro_ages_20230529.pdf
+
+% compare against Berger+20, 
+% note their sentence
+% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al.
+(2018) and Pinsonneault et al.
+(2018), as well as the rotation-based ages in Claytor et al.
+(2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al.
+(2019)."
+%
+% also compare against Mor2019 and their literature assessment
+% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al.
+(2015), and Haywood et al.
+(2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation.
+Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly.
+The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al.
+(2002) and Cignoni et al.
+(2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively.
+Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."
+
+\subsection{Ages for KOIs}
+
 \subsection{Evolutionary trends in time}

### 2023-06-06 paper/ms.tex
update ms

@@
   the primary Kepler mission.  Stellar rotation periods are collected from
-  previous analyses of the Kepler light curves; lithium is assessed directly from HIRES
-  spectra.
-  wWe report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
+  previous analyses of the Kepler light curves; lithium content is assessed
+  using HIRES spectra.
+  We report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
   precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
@@
 rotation periods.
+
 %FIXME
@@
 We considered two different samples of planets: those in the fully
-automated Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which are
+automated Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which is
 suitable for planet occurrence rate calculations, and also those in
-the cumulative KOI table, which includes the best knowledge available
-on any given planet candidate, but incorporates human-based vetting
-\footnote{See
+the cumulative KOI table, which included the best knowledge available
+on any given planet candidate as of 2023 Jun 6, but incorporated
+human-based vetting \footnote{See
 \url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
@@
 These objects will comprise the core data for our gyrochronology analysis.
+
+As a sanity check on the statistical uncertainties of these rotation
+periods, we compared our adopted \citetalias{Santos_2019} and
+\citetalias{Santos_2021} periods with those reported by
+\citet{McQuillan_2014} and \citet{Mazeh_2015}.
+We found that for $P_{\rm rot}\lesssim$15\,days, the two 
+datasets at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
+$P_{\rm rot}\approx$30\,days, the agreement was typically at the
+$\lesssim$0.03$P_{\rm rot}$ level, and
+the upper envelope of the period difference tended to increase
+linearly with period.
+Based on this comparison, we adopted a simple prescription for the
+period uncertainties,
+such that there are 1\% relative uncertainties below $P_{\rm
+rot}=15$\,days, and a linear increase thereafter, with slope set to
+guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.
 

### 2023-06-07 paper/ms.tex
update ms

@@
 
+% <250 words
 \begin{abstract}
@@
   using HIRES spectra.
-  We report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
-  precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
-  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
-  same stars.  The results include XXX planets younger than 1\,Gyr,
-  and YYY planets younger than 0.5\,Gyr.
-  In the base sample of dwarf FGK stars observed by
-  Kepler, we find that the intrinsic age distribution is(/not)
-  consistent with a uniform distribution between X and Y billion years
-  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
-  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
+  We report ages for \nkoisgoldwithages\ FGK stars with transiting
+  planets, with relative statistical precisions $\sigma_t / t$ of XX
+  +YY -ZZ\%; a factor of X.X +Y.Y -Z.Z more precise than
+  state-of-the-art isochronal ages for the same stars.
+  The results include XXX planets younger than 1\,Gyr (at 2$\sigma$),
+  and YYY planets younger than 0.5\,Gyr (at 2$\sigma$).
+  In the base sample of dwarf FGK stars observed by Kepler, we find
+  that the intrinsic age distribution is inconsistent with a uniform
+  distribution, and instead rises roughly linearly from 0-3\,Gyr;
+  stars 2\,Gyr old are twice are common in the Kepler field as stars
+  1\,Gyr old.
+  This agrees with results from previous isochronal analyses, and extends
+  beyond their age sensitivity, down to $\approx$0.1\,Gyr.
+  %TODO check A-D test
+  The sample of Kepler planets has a (in)consistent age distribution.
+  Analyzing the completeness of the Kepler
   detection pipeline, we interpret the majority of the change to be

### 2023-06-07 paper/ms.tex
update ms

@@
 \begin{abstract}
-  We report rotation- and lithium-based ages for stars and planets observed by
-  the primary Kepler mission.  Stellar rotation periods are collected from
-  previous analyses of the Kepler light curves; lithium content is assessed
-  using HIRES spectra.
+  We report empirically calibrated rotation- and lithium-based ages
+  for stars and planets observed by the primary Kepler mission.
+  Stellar rotation periods are collected from previous analyses of the
+  Kepler light curves; lithium content is assessed using HIRES
+  spectra.
   We report ages for \nkoisgoldwithages\ FGK stars with transiting
@@
   In the base sample of dwarf FGK stars observed by Kepler, we find
-  that the intrinsic age distribution is inconsistent with a uniform
-  distribution, and instead rises roughly linearly from 0-3\,Gyr;
-  stars 2\,Gyr old are twice are common in the Kepler field as stars
-  1\,Gyr old.
-  This agrees with results from previous isochronal analyses, and extends
-  beyond their age sensitivity, down to $\approx$0.1\,Gyr.
-  %TODO check A-D test
-  The sample of Kepler planets has a (in)consistent age distribution.
-  Analyzing the completeness of the Kepler
-  detection pipeline, we interpret the majority of the change to be
-  caused by selection effects (or astrophysics).  We also comment on a
+  that the gyrochronal age distribution rises roughly linearly from
+  0-3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
+  field as stars 1\,Gyr old.
+  This qualitatively agrees with results from previous isochronal and
+  asteroseismic studies, though it extends their sensitivity down to
+  $\approx$0.1\,Gyr.
+  The sample of detected Kepler planets has a similar age distribution, though
+  it is slightly deficient at $\lesssim$0.2\,Gyr, likely due to
+  decreased detection sensitivity.
+  %TODO will this be the case?
+  We also comment on a
   few previously reported evolutionary trends in the size evolution of

### 2024-04-01 paper/ms.tex
update ms

--- 2023-06-06 paper/ms.tex
@@
 
-\title{Gyrochronal and Lithium Ages for Stars and Planets Observed by Kepler}
+%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
+\title{Ages of Planets and Stars Observed by Kepler}
 
@@
 
-\author{Authors to include (ORDER AFTER LGB NOT YET ASSIGNED MEANING)}
+\author{Authors to include (order after LGB not assigned meaning;
+currently alphabetical)}
 
+% Early experiments for Kepler rotation samples
 \author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
@@
 
-\author{Lynne~A.~Hillenbrand}
-\affiliation{\caltech}
+% ?
+%\author{Lynne~A.~Hillenbrand}
+%\affiliation{\caltech}
 
-%
-% HIRES Collaborators
-%
+% HIRES observations and reductions
 \author[0000-0002-0531-1073]{H. Isaacson}
 \affiliation{\berkeley}
-%
+% HIRES observations, reductions, JUMP infrastructure
 \author[0000-0001-8638-0320]{A. W. Howard}
@@
 
+% <250 words
 \begin{abstract}
-  We report rotation- and lithium-based ages for stars and planets observed by
-  the primary Kepler mission.  Stellar rotation periods are collected from
-  previous analyses of the Kepler light curves; lithium content is assessed
-  using HIRES spectra.
-  We report ages for \nkoisgoldwithages\ FGK stars with transiting planets, with relative statistical
-  precisions $\sigma_t / t$ of XX +YY -ZZ\%; a factor of X.X
-  +Y.Y -Z.Z more precise than state-of-the-art isochronal ages for the
-  same stars.  The results include XXX planets younger than 1\,Gyr,
-  and YYY planets younger than 0.5\,Gyr.
-  In the base sample of dwarf FGK stars observed by
-  Kepler, we find that the intrinsic age distribution is(/not)
-  consistent with a uniform distribution between X and Y billion years
-  (Gyr).  In the planet sample, we find a consistent(/inconsistent)
-  distribution at $N$-sigma.  Analyzing the completeness of the Kepler
-  detection pipeline, we interpret the majority of the change to be
-  caused by selection effects (or astrophysics).  We also comment on a
-  few previously reported evolutionary trends in the size evolution of
-  close-in mini-Neptunes (2-4\,\rearth) and super-Earths
-  (1-2\,\rearth), a few of which we recover at comparable
-  ($\sim$3$\sigma$) significance to previous work, and one of which
-  we note for the first time.
+  We report empirically calibrated rotation- and lithium-based ages
+  for planets and stars observed by the primary Kepler mission.
+  %
+  Stellar rotation periods are collected from previous analyses of
+  Kepler data; lithium content is assessed using both new and
+  archival HIRES spectra.
+  %
+  We report ages for \nkoisgoldwithages\ FGK stars with transiting
+  planets, with relative statistical precisions $\sigma_t / t$ of XX
+  +YY -ZZ\%.  This is a factor of X.X +Y.Y -Z.Z more statistically precise than
+  isochronal ages for the same stars, and $\approx$X.X times less
+  precise than cluster-based ages.
+  %
+  The results include XXX planets younger than 1\,Gyr (at 3$\sigma$),
+  and YYY planets younger than 0.5\,Gyr (at 3$\sigma$).
+  %
+  We also report rotation-based ages for XX,XXX stars observed
+  by Kepler, aka.~the ``target star sample''.
+  %
+  In the latter sample of dwarf FGK stars observed by Kepler, we find
+  that the gyrochrone age distribution rises linearly from 0 to
+  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
+  field as stars 1\,Gyr old.
+  %
+  This qualitatively agrees with previous analyses of the Milky Way's star
+  formation history, though it offers greater sensitivity in the range of
+  $\approx$0.1-1\,Gyr.
+  %
+%  The sample of detected Kepler planets has a similar age
+%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
+%  likely due to decreased detection sensitivity.
+  %
+  Combining our stellar age catalog with Kepler's completeness products, we perform
+  a hierarchical bayesian analysis to infer the average number of planets per star
+  as a function of both stellar and planetary parameters.
+  %
+  While we recover many previously known correlations with stellar mass and metallicity,
+  our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
+  and in influencing the evolution of the mini-Neptune and super-Earth populations.
+  %TODO will this be the case?
+%  We also comment on a few previously reported evolutionary trends in
+%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
+%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
+%  ($\sim$3$\sigma$) significance to previous work, and one of which we
+%  note for the first time.
 \end{abstract}
@@
 Discovering a young planet requires solving two problems: find the
-planet, and measure the star's age.  Each problem can be
-solved in different ways, yielding many possible permutations for
-discovery.  In this manuscript, we will focus on the transit method
-for planet discovery, and rotation and lithium-based methods for stellar age
-measurement.  Rotation-based age-dating, also known as
-gyrochronology, will provide the most information for the largest
-number of stars.  The lithium-based technique will provide an
-important sanity check for the subset of stars younger than a few
+planet, and measure the star's age.  Each problem can be solved in
+various ways, and so there are many paths for discovery.  In this
+manuscript, we will focus on planets discovered using transits, and
+stellar ages measured based using empirical rotation and lithium-based
+methods.  Rotation-based age-dating will provide the most information
+for the largest number of stars.  The lithium-based technique will
+provide useful information for the subset of stars younger than a few
 hundred million years.
@@
 discovered in this way are assumed to have the same age as the host
-cluster, which are the standards for the current astrophysical age
-scale.  The resulting ages are precise at the $\approx$10\% level.  K2
+cluster.  The
+resulting ages are precise at the $\approx$10\% level.  K2
 and TESS have yielded larger numbers of young planets in clusters than
@@
 
-An important limitation of the ``cluster-first'' approach is that only
+A limitation of the ``cluster-first'' approach is that only
 $\approx$1\% of stars within the nearest few hundred parsecs can
@@
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-This means that there are many ``young'' 0.05-1\,Gyr stars in the
-field that were not part of the selection function of these searches.
-Roughly 40 transiting planets are currently known to be younger than
-1\,Gyr, with an age precision $t/\sigma_t > 3$ (see
+This means that there are many ``young'' (0.05-1\,Gyr) stars in the
+field that are omitted by cluster-only searches.
+Of order 40 transiting planets are currently known to be younger than
+1\,Gyr, requiring an age precision $t/\sigma_t > 3$ (see
 Figure~\ref{fig:rp_period_age}).  However, Kepler discovered
-$\approx$4{,}000 planets; and so barring a major discovery bias
-against young planets, we would expect a significant portion of the
-Kepler planets---perhaps 10\%, or 400 planets---to be younger than
-1\,Gyr.  If one is willing to compromise on age precision, those
-planets might be identifiable.
+$\approx$4{,}000 planets; and so unless there is a major discovery
+bias, we would expect a larger fraction of the Kepler
+planets---perhaps 10\%, or 400 planets---to be younger than 1\,Gyr.
 
@@
 \citet{Berger_2020b_rpage}, \citet{David_2021} or
-\citet{Petigura_2022}.  The Berger and Petigura studies focused on
-isochronal age-dating of individual stars, which is most precise in
-the context of Kepler for stars that are beginning to evolve off the
-main-sequence.  The \citet{David_2021} study leveraged both isochronal
+\citet{Petigura_2022}.  The \citeauthor{Berger_2020b_rpage} and
+\citeauthor{Petigura_2022} studies focused on
+isochronal age-dating of individual stars, which is most precise 
+for stars whose luminosities and temperature separate them from the
+main-sequence.  The \citeauthor{David_2021} study leveraged both isochronal
 ages from an earlier analogous study
-\citep{Fulton_Petigura_2018_cks_vii}, and (in their core analysis) a
-relative gyrochronal age-ranking approach that sorted stars based on
+\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
+relative gyrochrone age-ranking approach that sorted stars based on
 whether they rotated faster or slower than a gyrochrone age of
@@
 decreasing by at least a factor of two or three, and the super-Earth
-abundance correspondingly increasing {\bf BY HOW MUCH?}); {\it e)} that the
+abundance increasing in tandem; {\it e)} that the
 upper boundary of the super-Earth population should increase over the
@@
 histories for the size evolution of planets at these early times
-\citep[e.g.][]{Owen_2020}.  The generic prediction that the ``upper
-envelope'' of the mini-Neptune size distribution should shrink due to
-Kelvin-Helmholtz cooling is likely hard to avoid; predictions however
-concerning the ``radius valley'' and movement of planets across it are
-dependent on the timescales and prevalence of atmospheric mass loss.
-In specific scenarios for initial atmospheric accretion, the radius
-gap might form without needing to invoke mass loss at all
-\citep{Lee_2022}, in which case there could be very little evolution
-of the planetary size distribution between 1 and 4\,\rearth\ over the
-first gigayear.
+\citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
+the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
+generic prediction that the ``upper envelope'' of the mini-Neptune
+size distribution should shrink due to Kelvin-Helmholtz cooling is
+hard to avoid; predictions however concerning the ``radius valley''
+and movement of planets across it are dependent on the timescales and
+prevalence of atmospheric mass loss.  In specific scenarios for
+initial atmospheric accretion \citep{Lee_2022} and early atmospheric
+evolution \citep{2024MNRAS.529.2716R} the radius gap might form
+without needing to invoke post-disk dispersal mass loss at all.  In
+such scenarios, there could be very little evolution of the planetary
+size distribution between 1 and 4\,\rearth\ over the first gigayear.
 
-Can stellar age measurements provide lines of evidence for or against
-the different models?  Neither {\it (a)} nor {\it (b)} has yet had a
+Can stellar age measurements provide evidence for or against
+these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
 convincing---or even suggestive--- detection reported (see
-\citealt{Petigura_2022}).  While the apparent over-abundance of large
+\citealt{Petigura_2022}).  While an apparent over-abundance of large
 very young planets ({\it (c)}) has been noted by multiple authors
-\citep[e.g.][]{Bouma_2020,Mann_2022}, its presence is almost certainly
+\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
 exacerbated by a selection effect against sub-Neptune sized planet
-detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  And finally,
-{\it (d)} and {\it (e)} have been explored by
-\citet{Berger_2020b_rpage} and the twin \citet{David_2021} and
-\citet{Sandoval_2021} studies.  Both \citet{Berger_2020b_rpage} and
-\citet{Sandoval_2021} reported that $\theta$ increases with time, at a
-statistical significance of $\approx$2-3$\sigma$.  For
-\citet{Berger_2020b_rpage}, this conclusion was reached by comparing
-the sizes of 85 planets with median isochronal ages younger than
-1\,Gyr with a property-matched sample of planets older than 1\,Gyr.
-In \citet{Sandoval_2021}, the conclusion was reached by computing the
-ratio while performing a Monte Carlo re-sampling procedure from the
-isochronal ages and planetary sizes computed by
-\citet{Fulton_Petigura_2018_cks_vii}.  Finally, effect {\it (e)} was
-reported by \citet{David_2021}, who found that the average location of
-the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
+detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  Recent studies have
+nonetheless found that when accounting for completeness, there is
+$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
+$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
+\citep{2024arXiv240303261V} stars than in the older Kepler sample
+\citep[e.g.][]{2020AJ....159..248K}.
+%TODO: more kepler citations...
+
+Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
+(e)} were explored by \citet{Berger_2020b_rpage} and the twin
+\citet{David_2021} and \citet{Sandoval_2021} studies.  Both
+\citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
+increases with time, at a statistical significance of $\approx$2-3$\sigma$.
+For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
+sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
+property-matched sample of planets older than 1\,Gyr.  In
+\citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
+performing a Monte Carlo re-sampling procedure from the isochronal ages and
+planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
+effect {\it (e)} was reported by \citet{David_2021}, who found that the average
+location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
 $\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
@@
 in the planetary population than age.  Given these known correlations,
-it would be interesting to consider population models in which
-variables such as metallicity, mass, and age were added in succession
-\citep[e.g.][]{Thorngren_2021}.  For such an analysis to be effective,
-the measurement precision of these different parameters would ideally
-be as high as possible.
+it can be useful to consider population models in which variables such
+as metallicity, mass, and age are added in succession
+\citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
+rigorous approach, is to jointly infer the number of planets per star
+as a function of both planetary and stellar parameters, the latter
+including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
+most effective when the measurement precision of these different
+parameters is as high as possible.
 
 Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
-between 0.5-2.5\,Gyr, and can also provide useful age limits at
+between 0.5-4\,Gyr, and can also provide useful age limits at
 earlier times \citep{Bouma_2023}.  The idea of using a star's
@@
 \citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
-for the present analysis is the fact that we can empirically calibrate
+for our analysis is the fact that we can empirically calibrate
 rotation-based ages using measured sequences of rotation periods
@@
 spin-down rates, and in the cluster ages themselves.
+The oldest cluster for which this method has been calibrated is currently
+M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
 
@@
 \citep{Soderblom_2010}.  Here we focus on the latter approach, which
-relies on the gradual empirical observation that the Li abundances of
-FGK stars decline as a function of time, for reasons different from
-the standard convective mixing and burning scenario (CITE, CITE).
-Empirically, this decline has recently been documented by
-\citet{Jeffries_2023}, who built a model for how the equivalent width
-(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
-stellar effective temperature and age for a set of 6{,}200 stars in 52
-open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
+relies on the observation that the Li abundances of
+partially-convective stars gradually decline as they age
+\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
+this decline is debated
+\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
+Empirical understanding has however recently improved due to the work
+by \citet{Jeffries_2023}, who built a model for how the equivalent
+width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
+of stellar effective temperature and age for a set of 6{,}200 stars in
+52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
 stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
@@
 discussed in Section~\ref{sec:selection}.
-The age-dating methods are described and validated in
+We summarize the origin of stellar parameters other than the ages in
+Section~\ref{sec:stellarprops}, 
+and describe and validate our age-dating methods in
 Section~\ref{sec:agemethod}.
@@
 
-This work is focused on stars surveyed by Kepler for which ages can be
+This work focuses on stars surveyed by Kepler for which ages can be
 inferred using either stellar rotation, lithium, or both.  These
-``age-dateable stars'', $\mathcal{S}$, are a relatively small subset
-of the \nkeplerstars\ Kepler targets.  Rotation periods provide
-the majority of the information, since they have been reported for
+``age-dateable stars'' are a relatively small subset of the
+\nkeplerstars\ Kepler targets.  Rotation periods provide the majority
+of the information, since they have been reported for
 $\approx$55{,}000 Kepler stars
-\citep[e.g.][]{McQuillan_2014,Santos_2021}.  Measurements of the
-\ion{Li}{1} 6708\,\AA\ doublet require high-resolution
-($R\gtrsim$10{,}000) spectra, and are only available for a few
-thousand Kepler objects of interest (KOIs).  We therefore choose to
-define our selection function as Kepler stars with measured
-rotation periods ($\mathcal{S}$; Section~\ref{subsec:starsel}), and
-Kepler planets whose host stars have measured rotation periods
-($\mathcal{P}$; Section~\ref{subsec:planetsel}).
+\citep[e.g.][]{McQuillan_2014,Santos_2021}.  
+High-resolution measurements of the \ion{Li}{1} 6708\,\AA\ doublet are
+available for only a few thousand Kepler objects of interest (KOIs).
+We therefore define our selection function as Kepler stars
+with measured rotation periods ($\mathcal{S}$;
+Section~\ref{subsec:starsel}), and Kepler planets whose host stars
+have measured rotation periods ($\mathcal{P}$;
+Section~\ref{subsec:planetsel}).
 
@@
 analyze $\approx$40{,}000 stars with a variability range $R_{\rm
-	var}>0.3$\% that were not known EBs, planet candidates, or pulsators,
-and reported primary rotation periods for 24{,}124 of these stars.
+	var}>0.3$\% that were not known EBs, planet candidates, or
+  pulsators;
+they reported primary rotation periods for 24{,}124 of these stars.
 \citet{Reinhold_2015} also reported secondary periods for two thirds
@@
 
-%After reviewing the relevant literature and considering the nature of
-%the rotation period detection problem,
-We chose to adopt the results of \citetalias{Santos_2019} and
+We choose to adopt the results of \citetalias{Santos_2019} and
 \citetalias{Santos_2021} as our default rotation periods.
@@
 main-sequence and subgiant FGKM stars that were not reported as
-periodic by \citetalias{McQuillan_2014}.  Our interpretation is that
-these detections are likely more challenging than those from
+periodic by \citetalias{McQuillan_2014}.  Our speculation is that
+these reported detections are more challenging than those from
 \citetalias{McQuillan_2014}, perhaps due to a lower variability
@@
 Table~\ref{tab:stars}.
+We verified that the main conclusions of our analysis do not change if
+we were to adopt only the periods from \citetalias{McQuillan_2014}.
 
@@
 We found that for $P_{\rm rot}\lesssim$15\,days, the two 
-datasets at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
+datasets agree at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
 $P_{\rm rot}\approx$30\,days, the agreement was typically at the
@@
 \citealt{vogt_hires_1994}) on the Keck I 10m telescope.
-The majority of these spectra are archival, and were acquired
+Most of these spectra were acquired between 2010 and 2020
 as part of the California Keck Survey (CKS).
@@
 \section{Stellar Properties}
+\label{sec:stellarprops}
 
@@
 
-As our default source for stellar temperatures and surface gravities, we adopt
-the values from the Gaia-Kepler Stellar Properties Catalog (GKSP;
-\citealt{Berger_2020a_catalog}).
-The GKSP parameters are available for stars with ``AAA'' 2MASS photometry,
-measured parallaxes in Gaia DR2,  and g-band photometry from either the KIC or
-the Kepler-INT survey.
-The parameters themselves were derived using \texttt{isoclassify} (CITE
-Huber2017) to interpolate over the MIST isochrone grids (CITE Choi2016,
-Dotter2016), given the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
-parallexes, and $V$-band extinction from the (CITE) Green2019 reddening map.
-The resulting stellar parameters are available for
+As our default source for stellar temperatures and surface gravities,
+we adopt the values from the Gaia-Kepler Stellar Properties Catalog
+(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
+available for stars with ``AAA'' 2MASS photometry, measured parallaxes
+in Gaia DR2,  and $g$-band photometry from either the KIC or the
+Kepler-INT survey.  The parameters themselves were derived using
+\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
+the MIST isochrone grids
+\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
+and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
+$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
+map.  The resulting stellar parameters are available for
 \fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars
@@
 
+\subsection{``Galactic Archaeology'' Comparison}
+Zasowski et al fitted a chemical-isochrone model to APOGEE, X, Y, Z
+data and presented a model for the star formation history of the MW.
+The gist is a spike at z=2, and a gradual decrease since then, by
+about a few of ten.
+
+Mor2019 did isochrone fitting to Gaia, and found the same, but with a
+spike at 3Gyr.
+
+Our inferred gyro-age distribution of the Kepler field seems to agree
+with these pictures.  The main value added is likely over the first
+gigayear.
+
 \subsection{Isochrone Age Comparison}
@@
 \citet{David_2021}.
+
+Also Mathur2023, which does ``magneto-gyrochronology'', including the
+Sph indicator in a model of the ages.
 

### 2024-04-01 paper/ms.tex
update ms

@@
 \usepackage{amsmath,amstext,amssymb}
+\usepackage[caption=false]{subfig} % for subfloat
 \usepackage{xcolor, fontawesome}
 \usepackage{color}
+\usepackage{enumitem}
+
+\newcommand{\red}{\color{red}}
 
@@
 \newcommand{\vsini}{{$V \sin i$}}
-\newcommand{\teff}{T$_{\textrm{eff}}$}
 \newcommand{\kms}{{km\,s$^{-1}$}}
@@
 \newcommand{\rearth}{{R$_\oplus$}}
-\newcommand{\rsun}{{R$_\odot$}}
-\newcommand{\msun}{{M$_\odot$}}
+\newcommand{\rsun}{{$R_\odot$}}
+\newcommand{\msun}{{$M_\odot$}}
+\newcommand{\bprp}{G_{\rm BP} - G_{\rm RP}}
 
 \newcommand{\minus}{\scalebox{0.5}[1.0]{$-$}}
-
-\newcommand{\eg}{e.g.,}
-
-\newcommand{\mstype}{letter}
 
@@
 
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.8\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
+		}
+		\vspace{-0.62cm}
+		\subfloat{
+			\includegraphics[width=0.66\textwidth]{koi_mean_prot_teff.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
+    These periods are from \citet{Santos_2019}
+    (\citetalias{Santos_2019}) and \citet{Santos_2021}
+    (\citetalias{Santos_2021}).
+    The gray models... %TODO FIXME
+	}
+	\label{fig:prot_vs_teff}
+\end{figure*}
+
 \subsection{Ages for stars in the Kepler field}

### 2024-04-01 paper/collect.sh

@@ -1,5 +1,9 @@
-resultsdir="/Users/luke/Dropbox/proj/gyro-interp/results"
-cp $resultsdir/prot_vs_teff/prot_vs_teff_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_NGC-6819_Ruprecht-147_slowseq_poly7__100_to_2600_interpmethodpchip_m67.pdf f1a.pdf
+rdir="/Users/luke/Dropbox/proj/gyrojo/results"
+
+cp $rdir/star_Prot_Teff/prot_teff_Santos19_Santos21_clean0.pdf .
+
+cp $rdir/koi_mean_prot_teff/koi_mean_prot_teff.pdf .
+
 cp $resultsdir/prot_vs_teff/prot_vs_teff_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_models_poly7_α_Per_120-Myr_300-Myr_Praesepe_NGC-6811_interpmethodpchip_m67.pdf f1b.pdf
 
 cp $resultsdir/data_vs_model_prot/data_vs_model_α_Per_Pleiades_Blanco-1_Psc-Eri_NGC-3532_Group-X_Praesepe_NGC-6811_models_poly7_α_Per_120-Myr_300-Myr_Praesepe.pdf f2.pdf

### 2024-04-02 paper/ms.tex
update ms

@@
 		\subfloat{
-			\includegraphics[width=0.8\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
+			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
 		}
-		\vspace{-0.62cm}
 		\subfloat{
-			\includegraphics[width=0.66\textwidth]{koi_mean_prot_teff.pdf}
+			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
     {\bf Rotation periods for Kepler stars (left) and planets (right)}.
-    These periods are from \citet{Santos_2019}
+    The periods are from \citet{Santos_2019}
     (\citetalias{Santos_2019}) and \citet{Santos_2021}
-    (\citetalias{Santos_2021}).
-    The gray models... %TODO FIXME
+    (\citetalias{Santos_2021}), and the temperatures are primarily
+    from \citet{Berger_2020a_catalog}.
+    The gray models are fits from \citet{Bouma_2023} to open clusters;
+    stars of the same mass and age can have many possible rotation
+    periods, which sets the empirical precision limit of
+    rotation-based age dating.
+    The stellar sample shows only apparently single stars near the
+    main sequence with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
+    eff}$$<$6200, that are not flagged as eclipsing binaries.
+    The planet sample has the same stellar cuts, 
+    and shows confirmed and candidate planets from the cumulative KOI table, 
+    requiring them to have a MES$>$10.
+    Grazing objects ($b$>0.8) include 213 of the planet candidates
+    displayed, and 136 of the host stars.
 	}

### 2024-04-02 paper/ms.tex
update ms

--- 2024-04-01 paper/ms.tex
@@
 %%%%%%%%%
+\input{vals_manuscript}
+
 % number of stars monitored by Kepler (quoting Santos19/21)
 \newcommand{\nkeplerstars}{$\approx$160{,}000}
-% fraction of kepler stars w/ rotation periods
-\newcommand{\nstarswithprot}{{$\sim$50{,}000}}
 % fraction of stars with rotation periods, with teff/logg in Berger+20
@@
 and describe and validate our age-dating methods in
-Section~\ref{sec:agemethod}.
+Sections~\ref{sec:rotage} and~\ref{sec:liage}.
 The population-level trends for both the parent stellar sample and
@@
 
-\subsection{Kepler stars with rotation periods}
+\subsection{Rotation periods}
 \label{subsec:starsel}
 
-We select stars with rotation periods as determined by previous
-investigators.  Studies of stellar rotation across the entire Kepler field
-include those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
-\citet{Reinhold_2015},
+We select stars with rotation periods that have been measured by
+previous investigators.  Studies of stellar rotation across the entire
+Kepler field include those by \citet{McQuillan_2014}
+(\citetalias{McQuillan_2014}), \citet{Reinhold_2015},
 \citet{Santos_2019} (\citetalias{Santos_2019}), and
 \citet{Santos_2021} (\citetalias{Santos_2021}).
-\citetalias{McQuillan_2014} used an approach
-based on the autocorrelation function to detect 34{,}030 rotation
-periods for main-sequence Kepler targets cooler than 6500\,K, and excluded
-known eclipsing binaries and KOIs.  
-\citet{Reinhold_2015} used an iterative Lomb-Scargle based approach to
-analyze $\approx$40{,}000 stars with a variability range $R_{\rm
-	var}>0.3$\% that were not known EBs, planet candidates, or
-  pulsators;
-they reported primary rotation periods for 24{,}124 of these stars.
-\citet{Reinhold_2015} also reported secondary periods for two thirds
-of the stars with primary periods, which could be caused by
-differential rotation or finite spot lifetimes.  
-Finally,
-\citetalias{Santos_2019} and
+\citetalias{McQuillan_2014} used an approach based on the
+autocorrelation function to detect 34{,}030 rotation periods for
+main-sequence Kepler targets cooler than 6500\,K, and excluded known
+eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
+Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
+variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
+candidates, or pulsators; they reported primary rotation periods for
+24{,}124 of these stars.  \citet{Reinhold_2015} also reported
+secondary periods for two thirds of the stars with primary periods,
+which could be caused by differential rotation or finite spot
+lifetimes.  Finally, \citetalias{Santos_2019} and
 \citetalias{Santos_2021} combined a wavelet analysis and
@@
 \citetalias{Santos_2019} and \citetalias{Santos_2021} included known
-KOIs and binaries, and assigned them specific quality flags.  
-The rotation periods of KOIs have received considerable additional
+KOIs and binaries, and assigned them specific quality flags.  The
+rotation periods of KOIs have received considerable additional
 scrutiny
@@
 Table~\ref{tab:stars}.
+%TODO is it?
 We verified that the main conclusions of our analysis do not change if
 we were to adopt only the periods from \citetalias{McQuillan_2014}.
+%TODO have you?
 
 Concatenating the results of \citetalias{Santos_2019} and
-\citetalias{Santos_2021} yields \nstarswithprot\ Kepler stars with
+\citetalias{Santos_2021} yields \nuniqstarsantosrot\ Kepler stars with
 rotation periods.
-
-%FIXME
 
@@
 homogeneous sample, which includes \nkois\ KOIs, \nkoisnofp\ that are
-not known false positives;
-Appendix~\ref{app:cumulativekoi} highlights some additional young
-planets that are omitted in this approach.
+not known false positives; Appendix~\ref{app:cumulativekoi} highlights
+some additional young planets that are omitted in this approach.
 
@@
 map.  The resulting stellar parameters are available for
-\fracstarswithprotwithbtwenty\ of the \nstarswithprot\ Kepler stars
-with rotation periods.
+\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
+stars with rotation periods.
 
@@
 
-\section{Age Measurement}
-\label{sec:agemethod}
+\section{Ages From Rotation}
+\label{sec:rotage}
 
-\subsection{Gyrochrone ages}
-To calculate gyrochrone ages, we use \texttt{gyro-interp}
-\citep{Bouma_2023}.  This framework was built to statistically address
+We calculate gyrochrone ages using \texttt{gyro-interp}
+\citep{Bouma_2023}.  We built this statistical framework to address
 the issue that, especially over the first gigayear of stellar
@@
 framework are as follows.  {\it 1)} The stars are on the main
-sequence, {\it 2)} they do not have any binary companions,
-and {\it 3)} the rotation period distribution
-for field stars of a given mass and age is identical to that of
-the calibration open clusters (metallicity differences, for instance, are
-ignored).  
-Regarding the former two assumptions:
-pre-main-sequence and post-main-sequence evolution introduce
-variations in the stellar angular momentum budget.  Similarly, tidal
-effects might systematically spin up (or down) stars in binaries.
+sequence, {\it 2)} their spindown is not influenced by binary
+companions, and {\it 3)} the rotation period distribution for field
+stars of a given mass and age is identical to that of the calibration
+open clusters (metallicity differences, for instance, are ignored).  
+Regarding the former two assumptions: pre-main-sequence and
+post-main-sequence evolution introduce variations in the stellar
+angular momentum budget.  Similarly, tidal effects might
+systematically spin up (or down) stars in binaries.
 
@@
 resolution.
-This yielded age posterior probabilities over our requested age grid
-for \nkepstarswithages\ Kepler stars, and \nkoisnofpwithages\
-non-false-positive KOIs.
 
-\subsection{Lithium ages}
-To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
-with HIRES spectra, we use the procedure described by
-\citet{Bouma_2021}.
-The stars of interest for this work are FGK stars, and so the
-continuum in the vicinity of the
-\ion{Li}{1} 6708\,\AA\ feature is well-defined.
-We therefore numerically integrate a single best-fit Gaussian over a
-local window, and estimate the uncertainties on the line width through
-a Monte Carlo procedure that bootstraps against the local scatter in the
-continuum.
-Generally speaking, this approach does not correct for the neighboring
-\ion{Fe}{1} 6707.44\,\AA\ blend.
-
-To evaluate the accuracy and precision of this approach, after
-applying an initial iteration of this method on the \nkoiswithhires\
-KOIs with HIRES spectra, we compared our lithium equivalent widths
-with those reported by \citet{Berger_2018}.
-For the \nbergeroverlap\ stars in both samples, 
-we found {\bf X, Y, Z.}
-
-To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
-{\bf TODO: explain why}
-%TODO: explain why.
-
-\subsection{Quality flags}
-In our core gyrochrone analysis, we calculate gyrochrone age
-posteriors for all the stars with reported rotation periods and
-effective temperatures.  To select the stars for which we expect the
-gyrochronal age to be valid, we then build a set of quality flags, and
-apply them in the sub-analyses that follow.  We defined these filters
-as follows, each of which is included in Tables~\ref{tab:stars}
+\subsection{Star Filters}
+We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
+stars with reported rotation periods and effective temperatures.  To
+select the subset stars for which we expect these age to in fact be
+valid, we then build a set of filters and quality flags, and apply
+them in the sub-analyses that follow.  We defined these filters as
+follows, each of which is included in Tables~\ref{tab:stars}
 and~\ref{tab:planets}.
@@
 
+\subsection{Planet Filters}
+For the subset of our analysis that focuses on the properties of
+planets, we use the following additional filters.
+
+% see Erik's 2022 discussion; might want to shift to include a cut on
+% disposition score.
+{\it Candidate reliability}---no false positives; only ``planet
+candidates'' and ``confirmed planets''.
+
+{\it Candidate S/N}---the false positive rate increases greatly toward
+Kepler's noise floor for planet detection (CITE).  We required a S/N
+in excess of the usual floor, through a cut on the ``multiple event
+statistic'' (MES): we required ${\rm MES}$$>$10.
+
+{\it Grazing planets}---grazing objects ($b$>0.8) often provide
+non-robust planetary sizes due to the size-impact parameter
+degeneracy.  Particularly for large planets, they can also include
+astrophysical false positives at higher rates (due to the same
+degeneracy).  While we include a flag for such objects, we by default
+do not exclude them from our analysis; this is because...{\bf justify}.
+
+\subsection{Tally}
+
+The rotation periods for the stellar and planetary samples are shown
+in Figure~\ref{fig:prot_vs_teff}.
+Out of the \nkeplerstars\ stars observed by Kepler, requiring reported
+rotation periods yielded \nuniqstarsantosrotgyroappl.  Further
+requiring these rotating stars to be apparently single, near the main
+sequence, with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
+eff}$/K$<$6200, and not flagged as eclipsing binaries yielded
+\nuniqstarsantosrotgyroappl\ rotators amenable for gyrochronology.
+
+In the planet sample, the cumulative KOI table included
+\nnonfpcumkois\ confirmed and candidate planets, orbiting
+\nnonfpcumkoihosts\ stars.  Applying all the same stellar cuts, and
+requiring these confirmed and candidate planets to have MES$>$10,
+yields \nplwgyroage\ planets with rotation-based ages orbiting
+\nplhoststarwgyroage\ stars.  If we additionally require the planets
+to be non-grazing ($b<0.8$) based on the Markov Chain Monte Carlo
+(MCMC) fits from \citet{Thompson_2018}, we are left with
+\nplwgyroagenograzing\ and \nplhoststarwgyroagenograzing\ planets and
+stars, respectively.
+
+\section{Ages From Lithium}
+\label{sec:liage}
+
+To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
+with HIRES spectra, we use the procedure described by
+\citet{Bouma_2021}.
+The stars of interest for this work are FGK stars, and so the
+continuum in the vicinity of the
+\ion{Li}{1} 6708\,\AA\ feature is well-defined.
+We therefore numerically integrate a single best-fit Gaussian over a
+local window, and estimate the uncertainties on the line width through
+a Monte Carlo procedure that bootstraps against the local scatter in the
+continuum.
+Generally speaking, this approach does not correct for the neighboring
+\ion{Fe}{1} 6707.44\,\AA\ blend.
+
+To evaluate the accuracy and precision of this approach, after
+applying an initial iteration of this method on the \nkoiswithhires\
+KOIs with HIRES spectra, we compared our lithium equivalent widths
+with those reported by \citet{Berger_2018}.
+For the \nbergeroverlap\ stars in both samples, 
+we found {\bf X, Y, Z.}
+
+To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
+{\bf TODO: explain why}
+%TODO: explain why.
+
 \section{Results}
@@
     {\bf Rotation periods for Kepler stars (left) and planets (right)}.
-    These periods are from \citet{Santos_2019}
-    (\citetalias{Santos_2019}) and \citet{Santos_2021}
-    (\citetalias{Santos_2021}).
-    The gray models... %TODO FIXME
+    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
+    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
+    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
+    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
+    rotation sequences of open clusters.
+    The stellar sample shows only apparently single stars near the
+    main sequence with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
+    eff}$/K$<$6200, that are not flagged as eclipsing binaries.  The
+    planet sample has the same stellar cuts, and shows confirmed and
+    candidate planets from the cumulative KOI table, requiring them to
+    have a MES$>$10.
+    % koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png
+    % minus
+    % koi_mean_prot_teff_koi_X_S19S21dquality_dropgrazing.png
+    %Grazing objects ($b$>0.8) include 213 of the planet candidates
+    %displayed, and 136 of the host stars.
 	}

### 2024-04-02 gyrojo/papertools.py
stage papertools: w/ update_latex_key_value_pair
This is a function that I should have written a long time ago.

@@ -1,0 +1,62 @@
+import os, re
+from os.path import join
+from gyrojo.paths import PAPERDIR
+from pathlib import Path
+
+def update_latex_key_value_pair(
+    latexkeyname, latexvalue,
+    latexpath=join(PAPERDIR, "vals_manuscript.tex")
+):
+    """
+    Add lines like "\newcommand{\nstarswithaliens}{1}" to a value file
+    included in the manuscript.  This function searches through a
+    LaTeX file for a given key-value pair. If the pair is found, it
+    updates the value. If the pair is not found, it adds a new line
+    with the key-value pair. The function prints the updates made to
+    the file to stdout.
+
+    If latexvalue is a long number (greater than 999), it will be
+    comma-padded.  For example, 53421 will become "53{,}421".
+
+    Args:
+        latexkeyname (str): The name of the LaTeX key to update or add.
+            Must contain only lowercase letters.
+        latexvalue (str): The value to assign to the LaTeX key.
+        latexpath (str, optional): The path to the LaTeX file.
+            Defaults to "vals_stars.tex".
+
+    Returns:
+        None
+    """
+
+    if not os.path.exists(latexpath):
+        Path(latexpath).touch()
+
+    if not re.match(r'^[a-z]+$', latexkeyname):
+        raise ValueError("latexkeyname must contain only lowercase letters.")
+
+    if not isinstance(latexvalue, str):
+        latexvalue = str(latexvalue)
+
+    # Comma-pad long numbers in latexvalue
+    if latexvalue.isdigit() and len(latexvalue) > 3:
+        latexvalue = f"{latexvalue[:-3]}{{,}}{latexvalue[-3:]}"
+
+    updated = False
+    lines = []
+
+    with open(latexpath, "r") as file:
+        for line in file:
+            if line.startswith(f"\\newcommand{{\\{latexkeyname}}}"):
+                lines.append(f"\\newcommand{{\\{latexkeyname}}}{{{latexvalue}}}\n")
+                updated = True
+                print(f"Updated: \\newcommand{{\\{latexkeyname}}}{{{latexvalue}}}")
+            else:
+                lines.append(line)
+
+    if not updated:
+        lines.append(f"\\newcommand{{\\{latexkeyname}}}{{{latexvalue}}}\n")
+        print(f"Added: \\newcommand{{\\{latexkeyname}}}{{{latexvalue}}}")
+
+    with open(latexpath, "w") as file:
+        file.writelines(lines)

### 2024-04-02 drivers/get_values_for_manuscript.py
add value-getting driver

@@ -1,0 +1,52 @@
+import numpy as np, pandas as pd
+from gyrojo.getters import (
+    get_gyro_data, get_age_results, get_koi_data
+)
+from gyrojo.papertools import update_latex_key_value_pair as ulkvp
+
+# Stars with rotation periods and quality flags calculated
+df = get_gyro_data("Santos19_Santos21_dquality")
+
+assert pd.isnull(df.Prot).sum() == 0
+assert len(np.unique(df.KIC)) == len(df)
+N = len(df)
+# Number of unique stars with Santos+ rotation period reported
+ulkvp('nuniqstarsantosrot', N)
+
+# Number of unique stars with Santos+ rotation, and gyro applicable
+N = len(df[df.flag_is_gyro_applicable])
+ulkvp('nuniqstarsantosrotgyroappl', N)
+
+# Gotta get PLANET numbers!
+#TODO
+kdf = get_koi_data('cumulative-KOI')
+
+# Number of non-false positive KOIs from cumulative koi df
+N = len(kdf[~kdf.flag_koi_is_fp])
+ulkvp('nnonfpcumkois', N)
+
+# Number of unique non-false positive KOI host stars from cumulative koi df
+N = len(np.unique(kdf[~kdf.flag_koi_is_fp].kepid))
+ulkvp('nnonfpcumkoihosts', N)
+
+df, _, _ = get_age_results(whichtype='gyro', drop_grazing=0)
+
+# Number of planets with a gyro age, including grazing cases
+N = len(df)
+ulkvp('nplwgyroage', N)
+
+# Number of planet-hosting stars with a gyro age, including grazing cases
+N = len(np.unique(df.kepid_x))
+ulkvp('nplhoststarwgyroage', N)
+
+df, _, _ = get_age_results(whichtype='gyro', drop_grazing=1)
+
+# Number of planets with a gyro age, dropping grazing cases
+N = len(df)
+ulkvp('nplwgyroagenograzing', N)
+
+# Number of planet-hosting stars with a gyro age, dropping grazing cases
+N = len(np.unique(df.kepid_x))
+ulkvp('nplhoststarwgyroagenograzing', N)
+
+

### 2024-04-02 paper/ms.tex
update ms

@@
 Using our adopted effective temperatures and rotation periods, we ran
-\texttt{gyro-interp} over an age grid of 0 to 5\,Gyr, linearly spaced
-over 500 grid points.
-\texttt{gyro-interp} itself is only calibrated out to 4\,Gyr, the
-adopted age of the M67 cluster \citep[see][]{Dungee_2022,Gruner_2023}.
-We used the spin-down law between the reference open clusters by
-interpolating using piecewise cubic hermite interpolating polynomials,
-and integrated Equation~1 of \citep{Bouma_2023} at the default grid
-resolution.
+\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
+over 500 grid points.  \texttt{gyro-interp} itself is only calibrated
+out to 4\,Gyr, the adopted age of the M67 cluster
+\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the stellar
+spin-down rate as a function of time and stellar temperature, this
+framework interpolates between reference open clusters using smooth
+cubic hermite polynomials.  We calculate the probability of the
+rotation-based age $t_{\rm gyro}$, given the observed rotation
+periods, effective temperatures, and their uncertainties by
+integrating Equation~1 of \citep{Bouma_2023} at the default grid
+resolution.  This procedure marginalizes over the intrinsic age and
+temperature dependent scatter that is observed in the cluster
+sequences to yield statistical age uncertainties.
 
 \subsection{Star Filters}
+\label{subsec:filters}
 We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
@@
 
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
+    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
+    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
+    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
+    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
+    rotation sequences of open clusters.  The stellar sample shows
+    only apparently single stars near the main sequence with $\log
+    g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm eff}$/K$<$6200, that are not
+    flagged as eclipsing binaries.  The planet sample has the same
+    stellar cuts, and shows confirmed and candidate planets from the
+    cumulative KOI table, requiring them to have a MES$>$10.
+	}
+	\label{fig:prot_vs_teff}
+\end{figure*}
+
 The rotation periods for the stellar and planetary samples are shown
-in Figure~\ref{fig:prot_vs_teff}.
-Out of the \nkeplerstars\ stars observed by Kepler, requiring reported
-rotation periods yielded \nuniqstarsantosrotgyroappl.  Further
-requiring these rotating stars to be apparently single, near the main
-sequence, with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
-eff}$/K$<$6200, and not flagged as eclipsing binaries yielded
-\nuniqstarsantosrotgyroappl\ rotators amenable for gyrochronology.
+in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
+observed by Kepler, requiring reported rotation periods yielded
+\nuniqstarsantosrotgyroappl.  Further requiring these rotating stars
+to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
+over which \texttt{gyro-interp} returns finite age posteriors,
+yields \nuniqstarsantosrotteffcut\ stars.
+Further requiring them to be apparently single, near the main
+sequence, with $\log g$$>$4.2, RUWE$<$1.4, and not flagged as
+eclipsing binaries yielded \nuniqstarsantosrotgyroappl\ rotators
+amenable for gyrochronology.
 
@@
 
+\subsection{Age distribution}
+
 \begin{figure*}[!t]
 	\begin{center}
-		\subfloat{
-			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
+		\leavevmode
+		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20230529_maxage3200.pdf}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf Age distribution of stars and planet-hosts in the Kepler field
+    inferred from rotation periods}.
+    A histogram of samples from the age posteriors for all Kepler
+    stars with reported rotation periods and temperatures is shown in
+    light gray;
+    the age calculation was actually valid for stars that meet the criteria
+    discussed in Section~\ref{subsec:filters} (dark blue).
+		\label{fig:hist_tgyro}
 	}
-		\subfloat{
-			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
-		}
-	\end{center}
-	\vspace{-0.5cm}
-	\caption{
-    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
-    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
-    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
-    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
-    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
-    rotation sequences of open clusters.
-    The stellar sample shows only apparently single stars near the
-    main sequence with $\log g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm
-    eff}$/K$<$6200, that are not flagged as eclipsing binaries.  The
-    planet sample has the same stellar cuts, and shows confirmed and
-    candidate planets from the cumulative KOI table, requiring them to
-    have a MES$>$10.
-    % koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png
-    % minus
-    % koi_mean_prot_teff_koi_X_S19S21dquality_dropgrazing.png
-    %Grazing objects ($b$>0.8) include 213 of the planet candidates
-    %displayed, and 136 of the host stars.
-	}
-	\label{fig:prot_vs_teff}
 \end{figure*}
 
-\subsection{Ages for stars in the Kepler field}
-%TODO: show parameter distributions?
+Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions
+of stars and planet-hosts in the Kepler field.
+We truncated the plots near $3$\,Gyr due to arguments that suggest the
+detectability of rotation signals is near-unity up to this age for Sun-like
+stars, and decreases for older ages \citep{2022ApJ...937...94M}.
+Both the stellar and planetary samples point toward a star
+formation rate in the Kepler field that has slowed over the past
+few gigayears.
+Imagine we were to group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr, and 
+label them ``young'', ``intermediate-age'', and ``old''.
+A simple counting exercise from these distributions tells us that there are \ratioobtoybstars\ times as many 
+old stars in the Kepler field as young stars.
+Similarly, there are \ratioobtoybplanets\ times as many old planet hosts as
+young planet hosts.
+
+{\it LGB pending words: touch on findings on MW SFH from chemistry,
+isochrones, and asteroseismology}.
 

### 2024-04-02 paper/collect.sh

@@ -5,3 +5,6 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 
 # plot_koi_mean_prot_teff
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
+
+# plot_hist_field_gyro_ages 
+cp $r/hist_field_gyro_ages_20230529/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20230529_maxage3200.pdf .

### 2024-04-03 drivers/get_values_for_manuscript.py
update vals

@@ -50,6 +50,42 @@ ulkvp('nplwgyroage', N)
 N = len(np.unique(df.kepid_x))
 ulkvp('nplhoststarwgyroage', N)
 
+# Number of planets with a gyro age, including grazing cases, below 1Gyr
+sel = (df.gyro_median > 0) & (df.gyro_median <= 1000)
+N = len(df[sel])
+ulkvp('nplyounggyro', N)
+N = len(np.unique(df[sel].kepid_x))
+ulkvp('nplhostsyounggyro', N)
+
+# Number of planets with a gyro age, including grazing cases, 1-2 Gyr
+sel = (df.gyro_median > 1000) & (df.gyro_median <= 2000)
+N = len(df[sel])
+ulkvp('nplmidgyro', N)
+N = len(np.unique(df[sel].kepid_x))
+ulkvp('nplhostsmidgyro', N)
+
+# Number of planets with a gyro age, including grazing cases, 2-3 Gyr
+sel = (df.gyro_median > 2000) & (df.gyro_median <= 3000)
+N = len(df[sel])
+ulkvp('nploldgyro', N)
+N = len(np.unique(df[sel].kepid_x))
+ulkvp('nplhostsoldgyro', N)
+
+# Number of planets with a gyro age, including grazing cases, below 1Gyr at 2sigma
+sel = (df['gyro_median'] + df['gyro_+2sigma']  <= 1000)
+N = len(df[sel])
+ulkvp('nplyounggyrotwosigma', N)
+N = len(np.unique(df[sel].kepid_x))
+ulkvp('nplhostsyounggyrotwosigma', N)
+
+# Number of planets with a gyro age, including grazing cases, below 1Gyr at 3sigma
+sel = (df['gyro_median'] + df['gyro_+3sigma']  <= 1000)
+N = len(df[sel])
+ulkvp('nplyounggyrothreesigma', N)
+N = len(np.unique(df[sel].kepid_x))
+ulkvp('nplhostsyounggyrothreesigma', N)
+
+# ... now drop grazing cases
 df, _, _ = get_age_results(whichtype='gyro', drop_grazing=1)
 
 # Number of planets with a gyro age, dropping grazing cases

### 2024-04-03 paper/ms.tex
update ms

@@
 
-Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions
-of stars and planet-hosts in the Kepler field.
+Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
+stars and planet-hosts in the Kepler field.
 We truncated the plots near $3$\,Gyr due to arguments that suggest the
-detectability of rotation signals is near-unity up to this age for Sun-like
-stars, and decreases for older ages \citep{2022ApJ...937...94M}.
-Both the stellar and planetary samples point toward a star
-formation rate in the Kepler field that has slowed over the past
-few gigayears.
-Imagine we were to group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr, and 
-label them ``young'', ``intermediate-age'', and ``old''.
-A simple counting exercise from these distributions tells us that there are \ratioobtoybstars\ times as many 
-old stars in the Kepler field as young stars.
-Similarly, there are \ratioobtoybplanets\ times as many old planet hosts as
-young planet hosts.
-
-{\it LGB pending words: touch on findings on MW SFH from chemistry,
+detectability of rotation signals is near-unity up to this age for
+Sun-like stars, and decreases for older ages
+\citep{2022ApJ...937...94M}.
+Both the stellar and planetary samples point toward a star formation
+rate in the Kepler field that has slowed over the past few gigayears.
+{\it LGB pending words: allude to findings on MW SFH from chemistry,
 isochrones, and asteroseismology}.
-
-%e.g. hist_samples_field_gyro_ages_hist_field_gyro_ages_20230529.pdf
 
@@
 Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."
+
+Imagine we were to group stars between 0-1\,Gyr, 1-2\,Gyr, and
+2-3\,Gyr, and label them ``young'', ``intermediate-age'', and ``old''.
+A simple counting exercise from the distributions in
+Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
+times as many old stars in the Kepler field as young stars.
+Similarly, there are \ratioobtoybplanets\ times as many old planet
+hosts as young planet hosts.
+
+In terms of actual detected planet counts, our rotation-based ages for
+the Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
+detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
+For the sub-gigayear bin in particular, which holds specific weight in
+studies of planet evolution, while there are \nplyounggyro\ planets
+with median ages below 1\,Gyr, if we were to require $t_{\rm
+gyro}$$<$1\,Gyr at 2$\sigma$ confidence, this would leave only
+\nplyounggyrotwosigma\ planets orbiting \nplhostsyounggyrotwosigma\
+stars.
+
+\subsection{Sanity Checks}
+To date, four stellar clusters in the Kepler field have fielded planet
+detections.
+They are:...
+Young Kepler systems with the most precisely measured ages are...
+
+(TODO: perhaps easiest in a little table.  Theia316, Theia520, CepHer,
+and NGC6811).
 

### 2024-04-04 paper/ms.tex
minor

@@
 
-In terms of actual detected planet counts, our rotation-based ages for
+In terms of detected planet counts, our rotation-based ages for
 the Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
@@
 studies of planet evolution, while there are \nplyounggyro\ planets
-with median ages below 1\,Gyr, if we were to require $t_{\rm
-gyro}$$<$1\,Gyr at 2$\sigma$ confidence, this would leave only
+with median ages below 1\,Gyr, requiring $t_{\rm
+gyro}$$<$1\,Gyr at 2$\sigma$ confidence yields
 \nplyounggyrotwosigma\ planets orbiting \nplhostsyounggyrotwosigma\
@@
 \subsection{Sanity Checks}
-To date, four stellar clusters in the Kepler field have fielded planet
+To date, four stellar clusters in the Kepler field have yielded planet
 detections.

### 2024-04-04 drivers/plot_rp_vs_porb_binage.py

@@ -6,5 +6,5 @@ outdir = os.path.join(RESULTSDIR, 'rp_vs_porb_binage')
 if not os.path.exists(outdir): os.mkdir(outdir)
 
 ap.plot_rp_vs_porb_binage(outdir, teffcondition='allteff')
-ap.plot_rp_vs_porb_binage(outdir, teffcondition='teffgt5000')
-ap.plot_rp_vs_porb_binage(outdir, teffcondition='tefflt5000')
+#ap.plot_rp_vs_porb_binage(outdir, teffcondition='teffgt5000')
+#ap.plot_rp_vs_porb_binage(outdir, teffcondition='tefflt5000')

### 2024-04-04 paper/ms.tex
add <1gyr table maker

@@
 
-\subsection{Sanity Checks}
+\subsection{Young Planets}
+
 To date, four stellar clusters in the Kepler field have yielded planet
@@
 
+\startlongtable
+\begin{deluxetable}{lllrrrr}
+	%\tabletypesize{\footnotesize}
+	\tabletypesize{\scriptsize}
+	\tablecaption{Young planets and planet candidates.
+		%The non-truncated machine readable versions are accessible both
+		%through the online journal, and at
+		%\url{https://zenodo.org/record/8327508}.
+		\label{tab:youngplanets}}
+	%\toprule
+	%\midrule
+	%\endhead
+	\startdata
+	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ \\
+	-- &   -- & Myr &    Earths &    days &      bool &   bool  \\
+	\hline
+	\input{table_subgyr_confirmed} %\\
+	%\hline
+	\input{table_subgyr_candidate}
+	\enddata
+	\tablecomments{This table includes X, Y, Z... The machine-readable version,
+		available online, includes additional columns for .... 
+	}
+\end{deluxetable}
+%\endlongtable
+
 %\clearpage

### 2024-04-04 drivers/make_younghighlight_table.py

@@ -1,0 +1,115 @@
+import os
+from os.path import join
+import numpy as np, pandas as pd
+from gyrojo.getters import (
+    get_gyro_data, get_age_results, get_koi_data
+)
+from gyrojo.papertools import update_latex_key_value_pair as ulkvp
+from gyrojo.paths import PAPERDIR
+
+# ...with age results
+drop_highruwe = 0
+drop_grazing = 0
+_df, _, _ = get_age_results(
+    whichtype='gyro', drop_grazing=drop_grazing, drop_highruwe=drop_highruwe
+)
+
+# ...kepler-52 is <1002myr at 2sigma lol.
+sel = (_df['gyro_median'] + _df['gyro_+2sigma']  <= 1003)
+N = len(_df[sel])
+
+df = _df[sel]
+
+selcols = (
+    "kepoi_name,kepler_name,koi_disposition,gyro_median,"
+    "gyro_+1sigma,gyro_-1sigma,adopted_rp,adopted_period,flag_ruwe_outlier,flag_koi_is_grazing"
+).split(",")
+pdf = df[selcols].sort_values(
+    by=['gyro_median','kepler_name','kepoi_name']
+)
+
+for c in pdf.columns:
+    if 'gyro' in c:
+        pdf[c] = pdf[c].astype(int)
+    if c == 'adopted_period':
+        pdf[c] = np.round(pdf[c], 2)
+    if 'flag' in c:
+        pdf[c] = pdf[c].astype(int)
+
+pcols = (
+    "kepoi_name,kepler_name,gyro_median,"
+    "gyro_+1sigma,gyro_-1sigma,adopted_rp,adopted_period,flag_ruwe_outlier,flag_koi_is_grazing"
+).split(",")
+
+print(42*'-')
+print(pdf[pdf.koi_disposition == 'CONFIRMED'][pcols])
+print(42*'-')
+print(pdf[pdf.koi_disposition == 'CANDIDATE'][pcols])
+
+# where are kepler 1627 and 1643???
+#
+# Kepler-1627: failed RUWE cut
+# --> can merge that...
+#
+# Kepler-1643: Porb~Prot within 20%, and was a KOI, and so was treated in a
+# special way in the Santos+ processing (see email thread: "Rotation period of
+# Kepler-1643").  Quoting A. Santos: "From our current table of about 800 KOIs
[…]

### 2024-04-05 paper/ms.tex
correct missing Santos KOIs

@@
 the two studies.  \citetalias{Santos_2021} classified the 2{,}992
-remaining stars from \citetalias{McQuillan_2014} as ``non-periodic''
-based on updated knowledge of contaminants (e.g.~giant stars and
-eclipsing binaries) and visual inspection.  In addition,
+remaining stars from \citetalias{McQuillan_2014} as not showing
+rotation periods based on updated knowledge of contaminants
+(e.g.~giant stars and eclipsing binaries) and visual inspection.
+In addition,
 \citetalias{Santos_2021} report rotation periods for 24{,}182
 main-sequence and subgiant FGKM stars that were not reported as
-periodic by \citetalias{McQuillan_2014}.  Our speculation is that
-these reported detections are more challenging than those from
-\citetalias{McQuillan_2014}, perhaps due to a lower variability
-amplitude, or a longer rotation period.  A relevant flag for stars
+periodic by \citetalias{McQuillan_2014}.  Many of
+these reported detections have lower variability amplitudes, and
+longer periods, than those reported by
+\citetalias{McQuillan_2014}.  A relevant flag for stars
 missing \citetalias{McQuillan_2014} periods is included in
-Table~\ref{tab:stars}.
-%TODO is it?
-We verified that the main conclusions of our analysis do not change if
-we were to adopt only the periods from \citetalias{McQuillan_2014}.
-%TODO have you?
-
-Concatenating the results of \citetalias{Santos_2019} and
-\citetalias{Santos_2021} yields \nuniqstarsantosrot\ Kepler stars with
-rotation periods.
+Table~\ref{tab:stars}.  Reviewing the resulting period detections, we
+learned that 32 KOIs with orbital and rotation periods within
+$\approx$20\% were excluded from the \citetalias{Santos_2019} and
+\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
+to include these objects.\footnote{In general, excluding stars in this
+category would impose a bias against young close-in planets, due to
+the natural commensurability between the rotation rates of young
+stars, and Kepler's increased sensitivity to short-period objects.
+While some of these KOIs are known false positives, our analysis
+naturally excludes such objects at later steps.  As one motivating
+example, this subset of stars happens to include {\it
+e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
+$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
+based on membership in RSG-5 \citep{Bouma_2022}.  }
+The final tally of stars from \citetalias{Santos_2019},
+\citetalias{Santos_2021}, and our corrected KOI list is
+\nuniqstarsantosrot\ Kepler stars with rotation periods.
+
+%  %TODO is it?
+%  We verified that the main conclusions of our analysis do not change if
+%  we were to adopt only the periods from \citetalias{McQuillan_2014}.
+%  %TODO have you?
 
@@
 		\leavevmode
-		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20230529_maxage3200.pdf}
+		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
 	\end{center}
@@
 
+\clearpage
+
 \startlongtable
@@
 	\startdata
-	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ \\
+	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ \\
 	-- &   -- & Myr &    Earths &    days &      bool &   bool  \\
@@
 
-%\clearpage
+\clearpage
 

### 2024-04-05 paper/collect.sh

@@ -7,4 +7,4 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
 
 # plot_hist_field_gyro_ages 
-cp $r/hist_field_gyro_ages_20230529/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20230529_maxage3200.pdf .
+cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .

### 2024-04-06 paper/ms.tex
add li vs teff w/ eagles

@@
 
+\begin{figure}[!t]
+	\begin{center}
+		\leavevmode
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_eagles_showdispersion.pdf}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
+    from Keck/HIRES spectra collected between 2010 and 2024.}
+    Lines represent mean models derived from Gaia-ESO spectra of stars
+    in clusters of various ages; the intrinsic dispersion 
+    evolves in time, and is modelled as the $\pm$1$\sigma$
+    bands~\citep{Jeffries_2023}.
+    As for rotation-based ages, the intrisic dispersion sets the
+    empirical precision limit for the technique.
+		\label{fig:li_vs_teff}
+	}
+\end{figure}
+
 To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
@@
 continuum in the vicinity of the
-\ion{Li}{1} 6708\,\AA\ feature is well-defined.
+\ion{Li}{1} 6708\,\AA\ doublet is well-defined.
 We therefore numerically integrate a single best-fit Gaussian over a

### 2024-04-06 drivers/plot_li_vs_teff.py

@@ -5,5 +5,8 @@ import gyrojo.plotting as ap
 outdir = os.path.join(RESULTSDIR, 'li_vs_teff')
 if not os.path.exists(outdir): os.mkdir(outdir)
 
+ap.plot_li_vs_teff(outdir, yscale='linear', limodel='eagles', show_dispersion=1)
+
 for yscale in ['linear', 'log']:
-    ap.plot_li_vs_teff(outdir, yscale=yscale)
+    for limodel in ['eagles', 'baffles']:
+        ap.plot_li_vs_teff(outdir, yscale=yscale, limodel=limodel)

### 2024-04-06 gyrojo/plotting.py

@@ -1,10 +1,11 @@
 """
 Catch-all file for plotting scripts.  Contents:
 
-    plot_li_vs_teff
     plot_koi_mean_prot_teff
     plot_star_Prot_Teff
     plot_reinhold_2015
+
+    plot_li_vs_teff
 
     plot_koi_gyro_posteriors
     plot_li_gyro_posteriors
@@ -29,7 +30,7 @@ Helpers:
         _plot_slow_sequence_residual
         _plot_prot_vs_teff_residual
 """
-import os
+import os, sys
 from os.path import join
 from glob import glob
 from datetime import datetime
@@ -322,7 +323,7 @@ def plot_star_Prot_Teff(outdir, sampleid):
     savefig(fig, outpath)
 
 
-def plot_li_vs_teff(outdir, yscale=None):
+def plot_li_vs_teff(outdir, yscale=None, limodel='eagles', show_dispersion=0):
 
     df = get_li_data('all')
 
@@ -342,51 +343,86 @@ def plot_li_vs_teff(outdir, yscale=None):
     det = ~upperlim
     li_ew_upper_lims = li_ew_perr[upperlim]
 
-    # baffles
-    from baffles.readData import read_lithium
-    bv_m, upper_lim, fits = read_lithium()
+    if limodel == 'baffles':
 
-    # Pleiades, m34, Hyades, M67
-    inds = nparr([4, 6, 8, 9])
-    ages = [120, 240, 670, 4000]
+        # baffles
+        from baffles.readData import read_lithium
+        bv_m, upper_lim, fits = read_lithium()
 
-    from baffles.li_constants import BV
-    li_models = [10**fits[ix][0](BV) for ix in inds]
-    li_pleiades = li_models[0]
-    li_m34 = li_models[1]
-    li_hyades = li_models[2]
-    li_m67 = li_models[3]
-    li_hyades[li_hyades > li_m34] = li_m34[li_hyades > li_m34]
-    li_m67[li_m67 > li_hyades] = li_hyades[li_m67 > li_hyades]
+        # Pleiades, m34, Hyades, M67
+        inds = nparr([4, 6, 8, 9])
+        ages = [120, 240, 670, 4000]
 
-    from cdips.utils.mamajek import get_interp_Teff_from_BmV
-    Teff_model = get_interp_Teff_from_BmV(nparr(BV))
+        from baffles.li_constants import BV
+        li_models = [10**fits[ix][0](BV) for ix in inds]
[…]

### 2024-04-06 paper/collect.sh

@@ -8,3 +8,6 @@ cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf
 
 # plot_hist_field_gyro_ages 
 cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .
+
+# plot_li_vs_teff
+cp $r/li_vs_teff/li_vs_teff_showdispersion.pdf .

### 2024-04-07 paper/ms.tex
ms: clarify quality flag syntax

--- 2024-04-05 paper/ms.tex
@@
 exacerbated by a selection effect against sub-Neptune sized planet
-detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022}.  Recent studies have
+detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
 nonetheless found that when accounting for completeness, there is
@@
 $P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
-based on membership in RSG-5 \citep{Bouma_2022}.  }
+based on membership in RSG-5 \citep{Bouma_2022b}.  }
 The final tally of stars from \citetalias{Santos_2019},
@@
 
-\subsection{Star Filters}
-\label{subsec:filters}
+\subsection{Star Quality Flags}
+\label{subsec:flags}
 We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
@@
 select the subset stars for which we expect these age to in fact be
-valid, we then build a set of filters and quality flags, and apply
-them in the sub-analyses that follow.  We defined these filters as
-follows, each of which is included in Tables~\ref{tab:stars}
-and~\ref{tab:planets}.
+valid, we then build a set of quality flags.  When and
+how these flags should be applied depends on the question being asked.
+If the goal is to construct a false-positive free sample, they might
+all be applied.
+However if the goal is constructing a complete sample of young stars,
+then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
+Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
+resolved binary companion \citep{Bouma_2022a}; the latter is on a
+grazing orbit \citep{2014ApJ...783...53M}.  
+The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
+selecting for such cases using bitflags.
 
@@
 amplitude cuts you might want to apply to the rotation period catalog}
-%The combination of the filters
-%described above yields the set of stars that show no evidence for
-%binarity or crowding.  However, some of the rotation period analyses
-%in Table~\ref{tab:clusters} include additional relevant quality flags.
-%For instance, light curves showing multiple photometric periods can
-%indicate unresolved binarity.  We used all relevant filters available
-%from the original authors if they were designed to select single stars
-%with reliable rotation periods.  The final combination of these
-%filters with our own flag for possible binarity yields our sample of
-%benchmark rotators.
-
-\subsection{Planet Filters}
+
+\subsection{Planet Quality Flags}
 For the subset of our analysis that focuses on the properties of
-planets, we use the following additional filters.
+planets, we use the following additional quality flags.
 
@@
 observed by Kepler, requiring reported rotation periods yielded
-\nuniqstarsantosrotgyroappl.  Further requiring these rotating stars
+\nuniqstarsantosrot.  Further requiring these rotating stars
 to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
@@
 Further requiring them to be apparently single, near the main
-sequence, with $\log g$$>$4.2, RUWE$<$1.4, and not flagged as
-eclipsing binaries yielded \nuniqstarsantosrotgyroappl\ rotators
-amenable for gyrochronology.
+sequence, with $\log g$$>$4.2, and not flagged as
+eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
+potentially amenable for gyrochronology.
+Finally, imposing RUWE$<$1.4 leaves \nuniqstarsantosrotgyroappl\ stars
+for which gyrochronology is likely to be valid.
 
@@
 \nnonfpcumkois\ confirmed and candidate planets, orbiting
-\nnonfpcumkoihosts\ stars.  Applying all the same stellar cuts, and
-requiring these confirmed and candidate planets to have MES$>$10,
-yields \nplwgyroage\ planets with rotation-based ages orbiting
-\nplhoststarwgyroage\ stars.  If we additionally require the planets
-to be non-grazing ($b<0.8$) based on the Markov Chain Monte Carlo
-(MCMC) fits from \citet{Thompson_2018}, we are left with
-\nplwgyroagenograzing\ and \nplhoststarwgyroagenograzing\ planets and
-stars, respectively.
+\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
+cuts, save for that on RUWE.
+Then,
+require these confirmed and candidate planets to have MES$>$10.
+This combination of selection criteria
+yield \nplwgyroagewithgrazingandhighruwe\ planets with rotation-based
+ages orbiting \nplhoststarwgyroagewithgrazingandhighruwe\ stars
+(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
+If we additionally require the planets to be non-grazing ($b<0.8$)
+based on the Markov Chain Monte Carlo (MCMC) fits from
+\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ and
+\nplhoststarwgyroagenograzing\ planets and stars, respectively.
 
@@
 \label{sec:liage}
+
+\begin{figure}[!t]
+	\begin{center}
+		\leavevmode
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_eagles_showdispersion.pdf}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
+    from Keck/HIRES spectra collected between 2009 and 2024.}
+    Lines represent mean models derived from Gaia-ESO spectra of stars
+    in clusters of various ages; the intrinsic dispersion 
+    evolves in time, and is modelled as the $\pm$1$\sigma$
+    bands~\citep{Jeffries_2023}.
+    As for rotation-based ages, the intrisic dispersion sets the
+    empirical precision limit for the technique.
+		\label{fig:li_vs_teff}
+	}
+\end{figure}
+
+%TODO: does JUMP mirror all HIRES data from KOA?
+We queried a combination of the Keck Observatory Archive and the
+internal CPS archive to collect all HIRES spectra that were available
+for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
+with plausible rotation-based ages.
+This exercise yielded at least one spectrum for \nlithiumstars\ and 
+\nlithiumplanets\ planets.
+
+Out of the \nplhoststarwgyroagewithgrazingandhighruwe\ stars for w
 
@@
     the age calculation was actually valid for stars that meet the criteria
-    discussed in Section~\ref{subsec:filters} (dark blue).
+    discussed in Section~\ref{subsec:flags} (dark blue).
 		\label{fig:hist_tgyro}

### 2024-04-08 paper/ms.tex
update ms

@@
 \subsection{Tally}
+\label{subsec:tally}
 
@@
 for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
-with plausible rotation-based ages.
+described in Section~\ref{subsec:tally} for which it might be possible
+to derive rotation-based ages.
+
+%
+We required the spectra to not have the iodine cell in...
+%TODO ?!?!?!
+%
+
 This exercise yielded at least one spectrum for \nlithiumstars\ and 
 \nlithiumplanets\ planets.
+These observations were acquired between 6 September 2009 and 18
+September 2022.
 
@@
 \startlongtable
-\begin{deluxetable}{lllrrrr}
+\begin{deluxetable}{lllrrrrr}
 	%\tabletypesize{\footnotesize}
@@
 	\startdata
-	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ \\
-	-- &   -- & Myr &    Earths &    days &      bool &   bool  \\
+	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
+  -- &   -- & Myr &    Earths &    days &      bool &   bool  & bool \\
 	\hline

### 2024-04-08 paper/collect.sh

@@ -10,4 +10,4 @@ cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf
 cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .
 
 # plot_li_vs_teff
-cp $r/li_vs_teff/li_vs_teff_showdispersion.pdf .
+cp $r/li_vs_teff/li_vs_teff_eagles_showdispersion.pdf .

### 2024-04-09 paper/collect.sh
update ms

@@ -10,4 +10,4 @@ cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf
 cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .
 
 # plot_li_vs_teff
-cp $r/li_vs_teff/li_vs_teff_eagles_showdispersion.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf .

### 2024-04-09 paper/ms.tex
update ms

@@
 		\leavevmode
-		\includegraphics[width=0.49\textwidth]{li_vs_teff_eagles_showdispersion.pdf}
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf}
 	\end{center}
@@
 to derive rotation-based ages.
-
-%
-We required the spectra to not have the iodine cell in...
-%TODO ?!?!?!
-%
-
+We imposed no constraint on whether the iodine cell was present, since
+there are no relevant lines in the region of the 
+\ion{Li}{1} 6708\,\AA\ doublet.
 This exercise yielded at least one spectrum for \nlithiumstars\ and 
 \nlithiumplanets\ planets.
-These observations were acquired between 6 September 2009 and 18
-September 2022.
-
-Out of the \nplhoststarwgyroagewithgrazingandhighruwe\ stars for w
-
-To measure the lithium equivalent widths for the \nkoiswithhires\ KOIs
-with HIRES spectra, we use the procedure described by
-\citet{Bouma_2021}.
+These observations were acquired between 6 September 2009 and 16
+May 2024, primarily by the California Kepler Survey (CKS,
+CITE).
+In preparation for this study, we additionally collected {\bf XX}
+spectra for ``confirmed'' planets with rotation-based ages below
+1\,Gyr.
+
+To measure the lithium equivalent widths, we followed the procedure
+described by \citet{Bouma_2021}.
 The stars of interest for this work are FGK stars, and so the
@@
 To evaluate the accuracy and precision of this approach, after
-applying an initial iteration of this method on the \nkoiswithhires\
-KOIs with HIRES spectra, we compared our lithium equivalent widths
+applying an initial iteration of this method on the 
+\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
+we compared our lithium equivalent widths
 with those reported by \citet{Berger_2018}.
-For the \nbergeroverlap\ stars in both samples, 
-we found {\bf X, Y, Z.}
-
-To calculate lithium ages, we then use EAGLES \citep{Jeffries_2023}.
-{\bf TODO: explain why}
-%TODO: explain why.
+For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
+Z.}
+
+To calculate lithium ages, we use a combination of \texttt{EAGLES}
+\citep{Jeffries_2023} and \texttt{BAFFLES}
+\citep{Stanford-Moore_2020}.
+The former tool was calibrated on stars observed by the Gaia-ESO
+spectroscopic survey in 52 open clusters, with ages spanning 2 to
+6{,}000\,Myr.  
+Example \texttt{EAGLES} isochrones, as well as the modelled intrinsic dispersion
+about those isochrones, are shown in Figure~\ref{fig:li_vs_teff}.
+While our preference would have been to exclusively use
+\texttt{EAGLES} due to its homogeneous analysis approach ...
 
@@
 
-\subsection{Young Planets}
+\subsection{The Sub-Gigayear Sample}
+
+Given that the history of young exoplanet discovery includes a
+sociological tendency to report detailed descriptions of specific
+young exoplanets, we find it necessary to briefly discuss the
+properties of at least the most notable members of the set of
+confirmed planets that have well-measured ages.
+For this exercise, we follow Table~\ref{tab:youngplanets}.
+
+\paragraph{The Cep-Her Planets}
+Kepler\,1627Ab, Kepler\,1643b, Kepler\,1974b and Kepler\,1975\,Ab are
+known members of the $\approx$40\,Myr old Cep-Her association
+\citep{Bouma_2022a,Bouma_2022b}.
+The age of various sub-components of this association are
+best measured through relative isochronal placement of the
+pre-main-sequence \citep[see][]{Bouma_2022b}, and gyrochronology of
+stars in these clusters show them to be $<$100\,Myr old.
+The planets themselves are all on close-in orbits (5-25\,days), with
+sizes from 2-4\,$R_\oplus$.
+For Kepler\,1627A and Kepler\,1974, the rotation and lithium-based
+ages for these stars agree within 1$\sigma$ with the cluster-based
+ages.
+{\bf TODO discuss Kepler-1643 old tension, ditto KOI7913...}
+
+\paragraph{Theia~520 Planets}
+
+Kepler\,968 and Kepler\,52 both have rotation-based ages near
+350\,Myr.
+They are also members of Theia~520 (CITE).
+Independent work by J.~Curtis et al.~is indicating an age of XXX\,Myr for this
+cluster.
+This makes them the youngest multiplanet systems that have currently
+been identified from the Kepler primary mission.
+
+\paragraph{NGC~6811 Planets}
+Meibom (CITE CITE) ... do we get those ones?
+
+\paragraph{Systems with Rotation and Lithium in Agreement}
+Kepler\,1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is
+$\approx$100\,Myr old.
+
+Kepler\,1928b, a 19.6\,day 1.9\,$R_\oplus$ mini-Neptune is
+$\approx$140-200\,Myr old.
+
+Kepler\,1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing
+orbit is $\approx$120-180\,Myr old.
+
+Kepler\,1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth
+on a grazing orbit is $\approx$185\,Myr old.
+
+Kepler\,1521b, a 47.2\,day 2.3\,$R_\oplus$ super-Earth is 180-210\,Myr
+old.
+
+Kepler\,1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is
+230-240\,Myr old.
+
+Kepler\,1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr
+old.
+
+Kepler\,971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is
+300-370\,Myr old.
+
+Kepler\,1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
+rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$
+of its lithium age, though at half the relative uncertainty.
+This system may be the youngest Earth-sized planet known, comparable to 
+systems such as HD~63433d \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.
+
+\paragraph{Other Notables}
+
+Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
+
+Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
+The rotation-based age we find here is higher/lower than previously noted.
+
+Kepler~411 is also odd, due to the 58\,day-outer planet.
+
+\paragraph{Sub-}
+
+Kepler\,957b
 
@@
 \startlongtable
-\begin{deluxetable}{lllrrrrr}
+\begin{deluxetable*}{llllrrrrr}
 	%\tabletypesize{\footnotesize}
@@
 	\startdata
-	KOI & Kepler & $t_{\rm gyro}$ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
-  -- &   -- & Myr &    Earths &    days &      bool &   bool  & bool \\
+  KOI & Kepler & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
+  -- &   -- & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
 	\hline
@@
 	}
-\end{deluxetable}
+\end{deluxetable*}
 %\endlongtable

### 2024-04-09 paper/ms.tex
update ms

@@
 
-To evaluate the accuracy and precision of this approach, after
-applying an initial iteration of this method on the 
+To evaluate the accuracy and precision of our equivalent measurement
+technique, after applying an initial iteration on the
 \nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
@@
 stars.
-
-\subsection{The Sub-Gigayear Sample}
+The stellar and planetary properties of this latter sample of young
+planets are given in Table~\ref{tab:youngplanets}.
+
+\subsection{Planets Younger Than One Billion Years}
 
@@
 confirmed planets that have well-measured ages.
-For this exercise, we follow Table~\ref{tab:youngplanets}.
-
-\paragraph{The Cep-Her Planets}
-Kepler\,1627Ab, Kepler\,1643b, Kepler\,1974b and Kepler\,1975\,Ab are
+We first focus on systems reported to be in the four stellar
+ensembles in the Kepler field that have yielded transiting planets:
+Cep-Her, MELANGE-3, Theia~520, and NGC~6811.
+We then consider the broader set of Kepler planets that are young
+based on the rotational and lithium-based properties of their host
+stars.
+
+\subsubsection{Planets in Clusters}
+\label{subsec:clusterplanets}
+
+\paragraph{Cep-Her}
+Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab are
 known members of the $\approx$40\,Myr old Cep-Her association
-\citep{Bouma_2022a,Bouma_2022b}.
-The age of various sub-components of this association are
-best measured through relative isochronal placement of the
-pre-main-sequence \citep[see][]{Bouma_2022b}, and gyrochronology of
-stars in these clusters show them to be $<$100\,Myr old.
-The planets themselves are all on close-in orbits (5-25\,days), with
-sizes from 2-4\,$R_\oplus$.
-For Kepler\,1627A and Kepler\,1974, the rotation and lithium-based
-ages for these stars agree within 1$\sigma$ with the cluster-based
-ages.
-{\bf TODO discuss Kepler-1643 old tension, ditto KOI7913...}
-
-\paragraph{Theia~520 Planets}
-
-Kepler\,968 and Kepler\,52 both have rotation-based ages near
-350\,Myr.
-They are also members of Theia~520 (CITE).
-Independent work by J.~Curtis et al.~is indicating an age of XXX\,Myr for this
-cluster.
-This makes them the youngest multiplanet systems that have currently
-been identified from the Kepler primary mission.
-
-\paragraph{NGC~6811 Planets}
-Meibom (CITE CITE) ... do we get those ones?
-
-\paragraph{Systems with Rotation and Lithium in Agreement}
-Kepler\,1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is
-$\approx$100\,Myr old.
-
-Kepler\,1928b, a 19.6\,day 1.9\,$R_\oplus$ mini-Neptune is
-$\approx$140-200\,Myr old.
-
-Kepler\,1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing
-orbit is $\approx$120-180\,Myr old.
-
-Kepler\,1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth
-on a grazing orbit is $\approx$185\,Myr old.
-
-Kepler\,1521b, a 47.2\,day 2.3\,$R_\oplus$ super-Earth is 180-210\,Myr
-old.
-
-Kepler\,1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is
-230-240\,Myr old.
-
-Kepler\,1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr
-old.
-
-Kepler\,971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is
-300-370\,Myr old.
-
-Kepler\,1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
-rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$
-of its lithium age, though at half the relative uncertainty.
-This system may be the youngest Earth-sized planet known, comparable to 
-systems such as HD~63433d \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.
-
-\paragraph{Other Notables}
+\citep{Bouma_2022a,Bouma_2022b}.  The age of various sub-components of
+this association are best measured through relative isochronal
+placement of the pre-main-sequence \citep[see][]{Bouma_2022b}, and
+gyrochronology of stars in these clusters show them to be $<$100\,Myr
+old.  The planets themselves are all on close-in orbits (5-25\,days),
+with sizes from 2-4\,$R_\oplus$.  For Kepler-1627A and Kepler-1974,
+the rotation, lithium, and cluster-based ages all agree within
+1$\sigma$.  For Kepler-1643, while the rotation and cluster ages agree
+($46^{+9}_{-7}$\,Myr; \citealt{Bouma_2022b}), the lithium-based age
+($198^{+98}_{-76}$\,Myr) is 2.0$\sigma$ above the nominal cluster age
+for RSG-5.  For Kepler-1975Ab (KOI\,7913Ab), the lithium age may be
+similarly ``old'', though with less confidence due to the large
+intrinsic scatter in K dwarf lithium calibration measurements.  For
+that system, the isochronal age of its binary companion provides a
+more stringent and serendipitous constraint on the $\approx$40\,Myr
+age.
+
+\paragraph{MELANGE-3}
+\citet{2022AJ....164...88B} described an association in the Kepler
+field that was argued to be $105\pm10$\,Myr based on a combination of
+rotation, lithium, and the isochronal pre-main-sequence.  Kepler-970
+and Kepler-1928 harbor two transiting mini-Neptunes inside this
+association \citep{2022AJ....164...88B}.  Our rotation-based
+age-dating independently recovers these systems, with $t_{\rm
+gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
+Kepler-970 and -1928 respectively.  The former rotation-based age
+being slightly older than the suggested Pleiades age for the
+assocation agrees with Figure~4 of \citet{2022AJ....164...88B}, where
+the MELANGE-3 rotational isochrone is visually slower than
+the corresponding Pleiades track.
+
+\paragraph{Theia~520}
+Kepler-968 and Kepler-52 both have rotation-based ages of
+330-350\,Myr.  These stars were independently reported by
+\citet{2019AJ....158..122K} to be members of an association dubbed
+Theia~520 based on their spatial and kinematic correlations with
+nearby stars.  While the core of this structure was previously named
+UBC-1 \citep{2018A&A...618A..59C}, both Kepler planet hosts are in
+more diffuse regions of the cluster reported by
+\citet{2019AJ....158..122K}, so we adopt the name from the latter
+study.  The nominal age of Theia~520 reported by
+\citet{2019AJ....158..122K} based on isochrone fitting was
+$247^{+64}_{-51}$\,Myr.  An independent analysis of the cluster's core
+by \citet{2024A&A...681A..13F} using g-mode asteroseismology,
+gyrochronology, and isochrone fitting found consistent results.
+Ongoing spectroscopic and rotation-based studies of the broader
+cluster by J.~Curtis et al.~ is broadly that the two planet-hosting
+stars are indeed members of this diffuse population.  Given the broad
+set of both cluster-level and individual-star evidence, this makes
+Kepler-968 and Kepler-52 the youngest multiplanet systems that have
+currently been identified from the Kepler primary mission.
+We leave detailed discussion of the planet's properties to the latter 
+group.
+
+\paragraph{NGC~6811}
+\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
+mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
+We report rotation-based ages for these systems of
+$1419^{+657}_{-353}$\,Myr and $873.75\pm{112}$\,Myr respectively.
+For Kepler-66, $T_{\rm eff}$=5898\,K, and $P_{\rm
+rot}$=10.25$\pm$0.10\,days.
+{\it LGB TODO: check again after dereddening...}
+
+\subsubsection{Systems with Rotation and Lithium in Agreement}
+%
+Whenever both rotation and lithium find a star to be in a
+narrowly-constrained age range, confidence in the true age greatly
+increases, due to the independence of the two measurement techniques.
+The empirical precision for the inferred ages as a function of $P_{\rm
+rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
+\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
+respectively.  The region of greatest overlap is dictated by the
+timescale for lithium depletion: for stars with temperatures between
+4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
+independent two-sided constraints.
+
+The youth of most of these systems, to our knowledge, is being
+well-constrained for the first time in this work.  Here we limit
+discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
+mentioned above.
+
+The new young planets
+include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
+(Kepler-1529b $\approx$100\,Myr,
+-1521b $\approx$195\,Myr,
+-1467b $\approx$235\,Myr,
+-955b $\approx$400\,Myr, and
+-523b $\approx$440\,Myr),
+a single Neptune-sized planet
+(Kepler-971b $\approx$335\,Myr),
+five 1.1-1.8\,$R_\oplus$ super-Earths
+(Kepler-1930b $\approx$150\,Myr,
+-1313b $\approx$185\,Myr,
+-1565b $\approx$205\,Myr,
+-1532b $\approx$400\,Myr,
+-1481b $\approx$400\,Myr),
+and three Earth or sub-Earth-sized planets
+(Kepler-1933b $\approx$300\,Myr,
+-1561b $\approx$420\,Myr,
+-1482b $\approx$490\,Myr).
+All of these planets are on orbital periods below 50\,days.
+
+Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
+1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
+$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
+though at half the relative uncertainty.  Based on these lines of
+evidence, this system may be the youngest Earth-sized planet known,
+comparable to systems such as HD~63433d \citep[1.1\,$R_\oplus$,
+$414\pm23$\,Myr][]{2024AJ....167...54C}.
+
+A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
+planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
+find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.
+
+%  Kepler-1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is $\approx$100\,Myr old.
+%  Kepler-1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$120-180\,Myr old.
+%  Kepler-1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$185\,Myr old.
+%  Kepler-1521b, a 47.2\,day 2.3\,$R_\oplus$ mini-Neptune is 180-210\,Myr old.
+%  Kepler-1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is 230-240\,Myr old.
+%  Kepler-1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr old.
+%  Kepler-971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is 300-370\,Myr old.
+%  
+%  Kepler-1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
+%  rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of
+%  its lithium age, though at half the relative uncertainty.  Based on
+%  these lines of evidence, this system may be the youngest Earth-sized
+%  planet known, comparable to systems such as HD~63433d
+%  \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.
+%  
+%  Kepler-1532b is a 1.1\,day 1.3\,$R_\oplus$ super-Earth.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
+%  Kepler-955b is a 14.5\,day 2.7\,$R_\oplus$ mini-Neptune.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
+%  Kepler-1481b, a 5.9\,day 1.1\,$R_\oplus$ super-Earth has a
+%  rotation-based age of $\approx$400\,Myr, and a $2\sigma$ lithium
+%  detection that, upon visual inspection of the spectrum, is
+%  believeable.  The implied lithium age is marginally consistent with
+%  the rotation-based age, although perhaps with some tension.
+%  Kepler-447b is supposedly a 32\,$R_\oplus$ planet (grazing) on a
+%  7.8\,day grazing orbit.  Rotation and lithium independently find a
+%  420-440\,Myr age.
+%  Kepler-1561b is a 0.9\,$R_\oplus$ sub-Earth, on a 1.0\,day orbit.
+%  Rotation points to a 420\,Myr age.  The detected EW$_{\rm Li}$ is
+%  consistent, although given the star's temperature, not particularly
+%  constraining.
+%  The same is true for the 1.9\,$R_\oplus$ and 5.8\,day super-Earth Kepler-523b, and the 
+%  1.0\,$R_\oplus$, 12.2\,day Earth-sized Kepler-1482b.
+
+\subsubsection{Systems with Rotation and Lithium in Disagreement}
+
+For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
+This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
+rotational broadening of the spectrum ($v\sin
+i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
+rotation period.  However, if this star were $\lesssim$500\,Myr old,
+lithium should have been detectable at $>$50\,m\AA.  Lithium is
+marginally if at all present in the spectrum ({\bf  EW$_{\rm
+Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
+6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
+\citep{Morton_2016}, inspecting the data validation reports this is
+somewhat surprising.  The light curve shows greater scatter in-transit
+than out of transit, appears near-grazing, and has a suggestion of a
+strong odd-even depth difference.  This is a surprising system that
+merits further attention.
+
+\subsubsection{Need spectra}
+Kepler-1938...
+Kepler-762... (a HJ!)
+Kepler-1320
+
+\subsubsection{Other Notables}
 
@@
 Kepler~411 is also odd, due to the 58\,day-outer planet.
-
-\paragraph{Sub-}
-
-Kepler\,957b
-
-To date, four stellar clusters in the Kepler field have yielded planet
-detections.
-They are:...
-Young Kepler systems with the most precisely measured ages are...
-
-(TODO: perhaps easiest in a little table.  Theia316, Theia520, CepHer,
-and NGC6811).
-
-\subsection{Ages for KOIs}
 
@@
 \label{sec:conclusions}
+
+% 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
+Our rotation analysis blindly recovered the 14 known Kepler planets in
+clusters (Section~\ref{subsec:clusterplanets}), and provided accurate
+independent ages for their host stars.
 
@@
 \startlongtable
-\begin{deluxetable*}{llllrrrrr}
+\begin{deluxetable*}{lllllllrrrrr}
 	%\tabletypesize{\footnotesize}
@@
 	\startdata
-  KOI & Kepler & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
-  -- &   -- & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
+  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
+  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
 	\hline

### 2024-04-09 paper/ms.tex
update ms

@@
 		\leavevmode
-		\includegraphics[width=0.9\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
+		\includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
 	\end{center}
@@
 
-%\appendix
+\appendix
 %\section{Can you avoid appendices?}
@@
 
+\section{Figure dump}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\includegraphics[width=0.5\textwidth]{gyroage_vs_teff_errs_showplanets_linear.png}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf Gyro-age vs Teff}.
+		$\pm$1$\sigma$ uncertainties are plotted around the median for Kepler stars (black) and KOIs (blue).
+		Visual pileup on the left is b/c i) uncertain ages, and ii) lots of stars there.  Precision-land is the other portion.
+		\label{fig:gyroage_vs_teff}
+	}
+\end{figure*}
+
 \clearpage

### 2024-04-09 paper/collect.sh

@@ -11,3 +11,7 @@ cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_a
 
 # plot_li_vs_teff
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf .
+
+########## 
+# appendices (probably omit)
+cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .

### 2024-04-09 drivers/plot_scatter_gyroage_vs_teff.py
add scatter gyroage vs teff

@@ -1,0 +1,13 @@
+import os
+from gyrojo.paths import DATADIR, RESULTSDIR
+import gyrojo.plotting as ap
+
+outdir = os.path.join(RESULTSDIR, 'gyroage_vs_teff')
+if not os.path.exists(outdir): os.mkdir(outdir)
+
+# most useful
+for yscale in ['linear', 'log']:
+    for showerrs in [0, 1]:
+        for showplanets in [0, 1]:
+            ap.plot_gyroage_vs_teff(outdir, yscale=yscale, showerrs=showerrs,
+                                    showplanets=showplanets)

### 2024-04-10 paper/ms.tex
update ms: rewrite abstract

@@
 \begin{abstract}
-  We report empirically calibrated rotation- and lithium-based ages
-  for planets and stars observed by the primary Kepler mission.
-  %
-  Stellar rotation periods are collected from previous analyses of
-  Kepler data; lithium content is assessed using both new and
-  archival HIRES spectra.
-  %
-  We report ages for \nkoisgoldwithages\ FGK stars with transiting
-  planets, with relative statistical precisions $\sigma_t / t$ of XX
-  +YY -ZZ\%.  This is a factor of X.X +Y.Y -Z.Z more statistically precise than
-  isochronal ages for the same stars, and $\approx$X.X times less
-  precise than cluster-based ages.
-  %
-  The results include XXX planets younger than 1\,Gyr (at 3$\sigma$),
-  and YYY planets younger than 0.5\,Gyr (at 3$\sigma$).
-  %
-  We also report rotation-based ages for XX,XXX stars observed
-  by Kepler, aka.~the ``target star sample''.
-  %
-  In the latter sample of dwarf FGK stars observed by Kepler, we find
+  Recent studies of stars in open clusters have yielded a detailed picture
+  of how the rotation rates and lithium content of FGK stars evolve
+  during the main sequence.
+  %
+  Here we apply this knowledge to planets and stars observed by the
+  primary Kepler mission (2009-2013), and report the resulting rotation and
+  lithium-based ages.
+  %
+  Rotation periods are collected from previous analyses of
+  Kepler data; lithium is measured from new and archival HIRES
+  spectra.
+  %
+  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
+  and \nplwgyroagenograzing\ planets for which we believe
+  gyrochronology is likely valid. %, and for many more for which the underlying assumptions of
+%  the technique may be compromised.
+  %
+  The relative statistical precisions from rotation, $\sigma_{t_{\rm
+  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
+  statistically precise than isochronal ages for the same stars, and
+  $\approx$X.X times less precise than cluster-based ages.
+  %
+  The results include \nplyounggyrotwosigma\
+  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
+  planets at 3$\sigma$).
+  %
+  While lithium provides only marginal added information for most of
+  this sample,
+  we report robust two-sided age measurements from both rotation and
+  lithium for 14 confirmed $<$500\,Myr planets, including five
+  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
+  Neptune-sized planet.
+  %
+  In the broader Kepler target star sample,
+  we find
   that the gyrochrone age distribution rises linearly from 0 to
@@
   %
-  This qualitatively agrees with previous analyses of the Milky Way's star
-  formation history, though it offers greater sensitivity in the range of
-  $\approx$0.1-1\,Gyr.
+  This approach offers greater sensitivity to main sequence ages than previous analyses of the Milky Way's star
+  formation history, and the resulting age distribution bounds
+  the relative balance of ``young'' and ``old'' planets detectable not only by Kepler, but also
+  by future wide-field transit surveys.
   %
@@
   %
-  Combining our stellar age catalog with Kepler's completeness products, we perform
-  a hierarchical bayesian analysis to infer the average number of planets per star
-  as a function of both stellar and planetary parameters.
-  %
-  While we recover many previously known correlations with stellar mass and metallicity,
-  our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
-  and in influencing the evolution of the mini-Neptune and super-Earth populations.
+  %Combining our stellar age catalog with Kepler's completeness products, we perform
+  %a hierarchical bayesian analysis to infer the average number of planets per star
+  %as a function of both stellar and planetary parameters.
+  %
+  %While we recover many previously known correlations with stellar mass and metallicity,
+  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
+  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
   %TODO will this be the case?
@@
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
+			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+      % vs prot_teff_Santos19_Santos21_clean0.pdf !
 		}
@@
 %TODO: does JUMP mirror all HIRES data from KOA?
-We queried a combination of the Keck Observatory Archive and the
-internal CPS archive to collect all HIRES spectra that were available
-for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
-described in Section~\ref{subsec:tally} for which it might be possible
-to derive rotation-based ages.
-We imposed no constraint on whether the iodine cell was present, since
+We queried the internal CPS archive to collect all HIRES spectra that
+were available for the \nplhoststarwgyroagewithgrazingandhighruwe\
+stars for which rotation-based ages can be derived.
+This exercise yielded at least one spectrum for \nlithiumstars\ and 
+\nlithiumplanets\ planets.
+We did not constrain whether the iodine cell was used, since
 there are no relevant lines in the region of the 
 \ion{Li}{1} 6708\,\AA\ doublet.
-This exercise yielded at least one spectrum for \nlithiumstars\ and 
-\nlithiumplanets\ planets.
 These observations were acquired between 6 September 2009 and 16

### 2024-04-15 paper/collect.sh
update ms

@@ -1,5 +1,9 @@
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 
+# plot_camd
+cp $r/camd/camd_M_G_vs_dr3_bp_rp.pdf .
+cp $r/camd/camd_M_G_vs_adopted_Teff.pdf .
+
 # plot_star_Prot_Teff
 cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 

### 2024-04-15 paper/ms.tex
update ms
refactor stellar parameter plotter
(for generality!)

@@
   %
-  This approach offers greater sensitivity to main sequence ages than previous analyses of the Milky Way's star
-  formation history, and the resulting age distribution bounds
-  the relative balance of ``young'' and ``old'' planets detectable not only by Kepler, but also
-  by future wide-field transit surveys.
+  This approach offers greater sensitivity to main sequence ages than
+  previous analyses of the Milky Way's star formation history, and the
+  resulting age distribution bounds the relative balance of ``young''
+  and ``old'' planets detectable not only by Kepler, but also by
+  future wide-field transit surveys.
   %
@@
 
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.5\textwidth]{camd_M_G_vs_dr3_bp_rp.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.5\textwidth]{camd_M_G_vs_adopted_Teff.pdf}
+		}
+
+		\subfloat{
+		\includegraphics[width=0.5\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf CAMD}.
+    (todo)
+	}
+	\label{fig:camd}
+\end{figure*}
+
 % https://gea.esac.esa.int/archive/documentation/GDR3/Data_analysis/chap_cu8par/sec_cu8par_apsis/ssec_cu8par_apsis_gspphot.html

### 2024-04-15 drivers/plot_camd.py

@@ -1,9 +1,0 @@
-import os
-from gyrojo.paths import DATADIR, RESULTSDIR
-import gyrojo.plotting as ap
-
-outdir = os.path.join(RESULTSDIR, 'camd')
-if not os.path.exists(outdir): os.mkdir(outdir)
-
-ap.plot_camd(outdir, xkey='dr3_bp_rp')
-ap.plot_camd(outdir, xkey='adopted_Teff')

### 2024-04-15 drivers/plot_st_params.py

@@ -1,0 +1,16 @@
+import os
+from gyrojo.paths import DATADIR, RESULTSDIR
+import gyrojo.plotting as ap
+
+outdir = os.path.join(RESULTSDIR, 'st_params')
+if not os.path.exists(outdir): os.mkdir(outdir)
+
+# CAMDs
+ap.plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G')
+ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='M_G')
+
+# mKep vs Teff
+ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='dr3_phot_g_mean_mag')
+
+# logg vs Teff
+ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='adopted_logg')

### 2024-04-15 gyrojo/plotting.py

@@ -1766,7 +1766,7 @@ def plot_gyroage_vs_teff(outdir, yscale='linear', showerrs=0, showplanets=0):
     savefig(fig, outpath, dpi=400)
 
 
-def plot_camd(outdir, xkey='dr3_bp_rp', ykey='M_G'):
+def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G'):
 
     # get data
 
@@ -1778,18 +1778,17 @@ def plot_camd(outdir, xkey='dr3_bp_rp', ykey='M_G'):
     koidf, _, _ = get_age_results(whichtype='gyro_li', COMPARE_AGE_UNCS=0,
                                   drop_grazing=0, drop_highruwe=0)
 
-    #  # get all KOIs
-    #  koi_df = get_koi_data('cumulative-KOI', drop_grazing=0)
-    #  koi_df['kepid'] = koi_df['kepid'].astype(str)
-    #  # REQUIRE "flag_is_ok_planetcand"
-    #  skoi_df = koi_df[koi_df['flag_is_ok_planetcand']]
-
     # get KIC target stars
     from gyrojo.getters import (
         get_cleaned_gaiadr3_X_kepler_supplemented_dataframe
     )
     cgk_df = get_cleaned_gaiadr3_X_kepler_supplemented_dataframe()
 
+    #  # if you wanted all KOIs
+    #  koi_df = get_koi_data('cumulative-KOI', drop_grazing=0)
+    #  koi_df['kepid'] = koi_df['kepid'].astype(str)
+    #  # REQUIRE "flag_is_ok_planetcand"
+    #  skoi_df = koi_df[koi_df['flag_is_ok_planetcand']]
 
     # make plot
     plt.close('all')
@@ -1853,13 +1852,24 @@ def plot_camd(outdir, xkey='dr3_bp_rp', ykey='M_G'):
         xlabel = '$T_\mathrm{eff,adopted}$ [K]'
         xlim = [7500, 2500]
 
+    if ykey == 'M_G':
+        ylabel = '$M_\mathrm{G}$ [mag]'
+        ylim = [15.5, -6.5]
+    elif ykey == 'adopted_logg':
+        ylabel = '$\log g$ [dex]'
+        ylim = [6.5, 1.5]
+    elif ykey == 'dr3_phot_g_mean_mag':
+        ylabel = '$G$ [mag]'
+        ylim = [21, 5]
+
     ax.update({
[…]

### 2024-04-15 paper/collect.sh

@@ -1,8 +1,13 @@
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 
-# plot_camd
-cp $r/camd/camd_M_G_vs_dr3_bp_rp.pdf .
-cp $r/camd/camd_M_G_vs_adopted_Teff.pdf .
+# plot_st_params
+# CAMD:
+cp $r/st_params/st_params_M_G_vs_dr3_bp_rp.pdf .
+cp $r/st_params/st_params_M_G_vs_adopted_Teff.pdf .
+# logg vs Teff
+cp $r/st_params/st_params_adopted_logg_vs_adopted_Teff.pdf .
+# G vs Teff
+cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
 
 # plot_star_Prot_Teff
 cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .

### 2024-04-15 gyrojo/plotting.py
plot improvements

@@ -1375,15 +1375,27 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000, datestr='20240405')
         'ylabel': f'Fraction',
         'xlim': [xmin/1e3, (xmax-20)/1e3],
         'ylim': [0, 0.065],
-        'title': 'Kepler stars'
+        #'title': 'Kepler stars'
     })
+    axs[0].text(.08, .97, 'Kepler targets', ha='left', va='top',
+                fontsize='large', zorder=5, transform=axs[0].transAxes,
+                fontdict={'fontstyle':'oblique'})
+    if MAXAGE < 4000:
+        axs[0].set_xticks([0, 1, 2, 3])
+
     axs[1].update({
         'xlabel': '$t_{\mathrm{gyro}}$ [Gyr]',
         'ylabel': f'Fraction',
         'xlim': [xmin/1e3, (xmax-20)/1e3],
         'ylim': [0, 0.065],
-        'title': 'KOI host stars'
+        #'title': 'KOI host stars'
     })
+    axs[1].text(.08, .97, 'KOI host stars', ha='left', va='top',
+                fontsize='large', zorder=5, transform=axs[1].transAxes,
+                fontdict={'fontstyle':'oblique'})
+    if MAXAGE < 4000:
+        axs[1].set_xticks([0, 1, 2, 3])
+
     #axs[1].set_yticklabels([])
 
     # AESTHETIC HAD WEIRD ISSUES...
@@ -1854,10 +1866,10 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G'):
 
     if ykey == 'M_G':
         ylabel = '$M_\mathrm{G}$ [mag]'
-        ylim = [15.5, -6.5]
+        ylim = [15.5, -3.5]
     elif ykey == 'adopted_logg':
-        ylabel = '$\log g$ [dex]'
-        ylim = [6.5, 1.5]
+        ylabel = '$(\log g)_\mathrm{adopted}$ [dex]'
+        ylim = [6.5, 2.5]
     elif ykey == 'dr3_phot_g_mean_mag':
         ylabel = '$G$ [mag]'
         ylim = [21, 5]
@@ -1869,7 +1881,7 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G'):
         'xlim': xlim
     })
     if ykey == 'M_G':
-        ax.set_yticks([15, 10, 5, 0, -5])
[…]

### 2024-04-15 paper/collect.sh

@@ -1,12 +1,12 @@
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 
 # plot_st_params
-# CAMD:
+# ...CAMD:
 cp $r/st_params/st_params_M_G_vs_dr3_bp_rp.pdf .
 cp $r/st_params/st_params_M_G_vs_adopted_Teff.pdf .
-# logg vs Teff
+# ...logg vs Teff
 cp $r/st_params/st_params_adopted_logg_vs_adopted_Teff.pdf .
-# G vs Teff
+# ...G vs Teff
 cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
 
 # plot_star_Prot_Teff

### 2024-04-15 paper/ms.tex
plot improvements
ms: add Kepler 411 saga

@@
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{camd_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.5\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{camd_M_G_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.5\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
 		}
@@
 	\caption{
-    {\bf CAMD}.
-    (todo)
+    {\bf todo}.
 	}
@@
 Kepler~411 is also odd, due to the 58\,day-outer planet.
+\citet{2019A&A...624A..15S} conducted a transiting timing analysis of the
+system, confirming Kepler-411d and argued for the presence of a fourth
+non-transiting planet, Kepler-411e, based on the TTVs seen in Kepler-411d and
+assuming that the system would be nearly coplanar.
+Regarding the stellar age, \citet{2019A&A...624A..15S} used the 
+\citet{2007ApJ...669.1167B} relation, which yielded an estimate of 212$\pm$31\,Myr.
+However, the star's temperature ($\approx$4900\,K) puts it in a regime where
+the Barnes~relation has been known to not fit observations for quite some time
+\citep[see][Fig.~9]{2008ApJ...687.1264M}.
+Our adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
+calibrators.
 

### 2024-04-15 paper/ms.tex
update ms

@@
 
-\author{Authors to include (order after LGB not assigned meaning;
-currently alphabetical)}
-
-% Early experiments for Kepler rotation samples
-\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
-\affiliation{\caltech}
-
-% ?
-%\author{Lynne~A.~Hillenbrand}
+\author{et al}
+%
+%% Early experiments for Kepler rotation samples
+%\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
 %\affiliation{\caltech}
-
-% HIRES observations and reductions
-\author[0000-0002-0531-1073]{H. Isaacson}
-\affiliation{\berkeley}
-% HIRES observations, reductions, JUMP infrastructure
-\author[0000-0001-8638-0320]{A. W. Howard}
-\affiliation{\caltech}
-%
-\author{HIRES observers to be named}
-\affiliation{Various}
+%
+%% ?
+%%\author{Lynne~A.~Hillenbrand}
+%%\affiliation{\caltech}
+%
+%% HIRES observations and reductions
+%\author[0000-0002-0531-1073]{H. Isaacson}
+%\affiliation{\berkeley}
+%% HIRES observations, reductions, JUMP infrastructure
+%\author[0000-0001-8638-0320]{A. W. Howard}
+%\affiliation{\caltech}
+%%
+%\author{HIRES observers to be named}
+%\affiliation{Various}
 
@@
 Of order 40 transiting planets are currently known to be younger than
-1\,Gyr, requiring an age precision $t/\sigma_t > 3$ (see
-Figure~\ref{fig:rp_period_age}).  However, Kepler discovered
+1\,Gyr, requiring an age precision $t/\sigma_t > 3$.
+%TODO cite???
+% (see
+%Figure~\ref{fig:rp_period_age}).
+However, Kepler discovered
 $\approx$4{,}000 planets; and so unless there is a major discovery
@@
 
-\begin{figure}[!t]
-	\begin{center}
-		\leavevmode
-		\includegraphics[width=0.49\textwidth]{f1.pdf}
-	\end{center}
-	\vspace{-0.6cm}
-	\caption{
-		{\bf Radii, orbital periods, and ages of transiting exoplanets}.
-    Planets younger than 1\,Gyr with ages more precise than a
-    factor of three are emphasized. 
-    ({\bf cite relevant papers?})
-    Parameters are from the \citet{PSCompPars}.
-		\label{fig:rp_period_age}
-	}
-\end{figure}
+%\begin{figure}[!t]
+%	\begin{center}
+%		\leavevmode
+%		\includegraphics[width=0.49\textwidth]{f1.pdf}
+%	\end{center}
+%	\vspace{-0.6cm}
+%	\caption{
+%		{\bf Radii, orbital periods, and ages of transiting exoplanets}.
+%    Planets younger than 1\,Gyr with ages more precise than a
+%    factor of three are emphasized. 
+%    ({\bf cite relevant papers?})
+%    Parameters are from the \citet{PSCompPars}.
+%		\label{fig:rp_period_age}
+%	}
+%\end{figure}
 
@@
 			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
-      % vs prot_teff_Santos19_Santos21_clean0.pdf !
 		}
@@
 \subsection{Planets Younger Than One Billion Years}
+
+%\begin{figure*}[!t]
+%	\begin{center}
+%		\subfloat{
+%			\includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_strictyoung.pdf}
+%		}
+%		\subfloat{
+%			\includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
+%		}
+%	
+%		\subfloat{
+%		\includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_anyyoung.pdf}
+%		}
+%		\subfloat{
+%			\includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+%		}
+%	\end{center}
+%	\vspace{-0.5cm}
+%
+%	\caption{
+%		{\bf Radii, orbital periods, and ages of transiting exoplanets}.
+%    Planets younger than 1\,Gyr are emphasized. 
+%    Two possible selection functions are suggested...
+%		\label{fig:rp_period_age_results}
+%	}
+%\end{figure*}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
+		}
+
+		\subfloat{
+			\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+
+	\caption{
+		{\bf Sizes, orbital periods, and rotation-based ages of young transiting exoplanets}.
+		Two different selection functions for what defines a young, sub-gigayear, planet
+		are imposed on the {\it top} and {\it bottom}: the former requires $t$>1\,Gyr with
+		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
+		Circles are previously reported measurements from the literature.
+		Stars are newly calculated ages for Kepler planets from this work.
+		Gray points are transiting planets from the NASA Exoplanet Archive older than
+		1\,Gyr.
+		The three age bins span 10-46\,Myr, 46-215\,Myr, and 215-1000\,Myr.
+		\label{fig:rp_period_age_results}
+	}
+\end{figure*}
 

### 2024-04-15 paper/collect.sh

@@ -21,6 +21,15 @@ cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_a
 # plot_li_vs_teff
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf .
 
+# planets
+r="/Users/luke/Dropbox/proj/NEA_age_plots/results/rp_vs_period_scatter"
+cp $r/rp_vs_period_scatter_20240415_colorbyage_strictyoung.pdf .
+cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf .
+cp $r/rp_vs_period_scatter_20240415_colorbyage_anyyoung.pdf .
+cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf .
+
+
 ########## 
 # appendices (probably omit)
+r="/Users/luke/Dropbox/proj/gyrojo/results"
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .

### 2024-04-16 paper/ms.tex
update ms

@@
 \defcitealias{Santos_2021}{S21}
+\defcitealias{Fulton_2018}{F18}
+\defcitealias{Berger_2020a_catalog}{B20}
 
@@
 missing \citetalias{McQuillan_2014} periods is included in
-Table~\ref{tab:stars}.  Reviewing the resulting period detections, we
+Table~\ref{tab:stars} {\bf TODO make this}.  Reviewing the resulting period detections, we
 learned that 32 KOIs with orbital and rotation periods within
@@
 Crossmatching the Q1-Q17 DR25 KOI table against our stars with
-rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of which \nkoisnofpwithprot\ 
-are not known false postives.
-These objects will comprise the core data for our gyrochronology analysis.
+rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of
+which \nkoisnofpwithprot\ are not known false postives.  These objects
+will comprise the core data for our gyrochronology analysis.
 
@@
 \citetalias{Santos_2021} periods with those reported by
-\citet{McQuillan_2014} and \citet{Mazeh_2015}.
-We found that for $P_{\rm rot}\lesssim$15\,days, the two 
-datasets agree at a precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of 
-$P_{\rm rot}\approx$30\,days, the agreement was typically at the
-$\lesssim$0.03$P_{\rm rot}$ level, and
-the upper envelope of the period difference tended to increase
-linearly with period.
-Based on this comparison, we adopted a simple prescription for the
-period uncertainties,
-such that there are 1\% relative uncertainties below $P_{\rm
-rot}=15$\,days, and a linear increase thereafter, with slope set to
-guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.
+\citet{McQuillan_2014} and \citet{Mazeh_2015}.  We found that for
+$P_{\rm rot}\lesssim$15\,days, the two datasets agree at a precision
+of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
+rot}\approx$30\,days, the agreement was typically at the
+$\lesssim$0.03$P_{\rm rot}$ level, and the upper envelope of the
+period difference tended to increase linearly with period.  Based on
+this comparison, we adopted a simple prescription for the period
+uncertainties, such that there are 1\% relative uncertainties below
+$P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
+set to guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.
 
@@
 	}
-	\label{fig:camd}
+	\label{fig:stellarprops}
 \end{figure*}
@@
 stars with rotation periods.
+\citet{David_2021} compared the stellar parameters ($T_{\rm eff}$,
+$R_\star$, [Fe/H]) derived from the \citet{Berger_2020a_catalog}
+isochronal approach with the spectroscopic parameters from
+\citet{Fulton_2018}.  While this approach is one way of validating the
+\citet{Berger_2020a_catalog} output, the temperature scales did show a
+few-percent difference, with \citetalias{Fulton_2018} quoting higher
+temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
+and lower temperatures for early F dwarfs.
 
@@
 For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
-temperatures and surface gravities, we adopt the values reported by
+temperatures and surface gravities from
+\citetalias{Berger_2020a_catalog}, we adopt the values reported by
 \citet{Santos_2019} and \citet{Santos_2021}, which for these cases
@@
 Properties Catalog, and are mostly derived from photometry.  In the
-planet sample, \frackoisnofpwithprotwithbtwenty\ of the the
+planet sample, \frackoisnofpwithprotwithbtwenty\ of the
 non-false-positive KOIs with rotation periods have parameters from
@@
 
-As a sanity check on our effective temperature scale, we then used the 
-the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
-BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
-relation suggested by \citet{Curtis_2020}.
-This latter effective temperature scale was demonstrated to have a 
-scatter of $\approx$50\,K for stars near the main-sequence.
-We find {\bf X, Y, Z.}
-We ultimately favor the \citet{Berger_2020a_catalog} scale
-because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
-isochronal fit.
+%TODO: the relevant check here is Berger20 teff vs CKS spec teff...
+%  As a sanity check on our effective temperature scale, we then used the 
+%  the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
+%  BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
+%  relation suggested by \citet{Curtis_2020}.
+%  This latter effective temperature scale was demonstrated to have a 
+%  scatter of $\approx$50\,K for stars near the main-sequence.
+%  We find {\bf X, Y, Z.}
+%  We ultimately favor the \citet{Berger_2020a_catalog} scale
+%  because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
+%  isochronal fit.
 
@@
 
-\subsection{Star Quality Flags}
+\subsection{Gyrochronology Quality Flags}
 \label{subsec:flags}
@@
 all be applied.
-However if the goal is constructing a complete sample of young stars,
+If the goal is constructing a complete sample of young stars,
 then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
@@
 
-{\it Subgiants}---We flagged stars as possible subgiants if they had
+{\it Temperature range (bit 0)}---We required stars to have effective
+temperatures $T_{\rm eff}$ between 3800 and 6200\,K
+\citep{Bouma_2023}.   Stars hotter than 6200\,K spin down very slowly,
+if at all, limiting the utility of rotation-based ages.  Stars cooler
+than 3800\,K do seem to spin down over gigayear timescales (e.g. CITE
+E.Newton thesis, the Engle paper, and Chiti's thing), but the
+currently available open cluster data have yet to clarify at what
+times, if any, the intrinsic scatter in the population decreases.
+
+{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants if they had
 $\log g < 4.2$.  We set this threshold by examining the
@@
 
-{\it Photometric binaries}---We plotted color--absolute magnitude
-diagrams using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
-BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
-loci to flag over or under-luminous stars in each diagram, typically
-more than $\approx$0.5 magnitudes above or below the main sequence.
-%TODO refer to a figure!!!
-
-{\it RUWE}---Following recommendations in the Gaia DR3
+{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia DR3
+$G$-band luminosity, ignoring reddening, using the reported apparent
+$G$-band magnitude and parallaxes.
+We flag stars with $M_{\rm G} < 3.9$, corresponding to main-sequence
+spectral types earlier than $\approx$F8V \citep{Pecaut_2013}.
+
+{\it Known eclipsing binaries (bit 3)}---We flagged any stars reported to be
+in the final Kepler eclipsing binary catalog (KEBC;
+\citealt{2016AJ....151...68K}).
+
+{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In instances
+where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
+Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun/}; we
+used 4$''$ rather than the 1$''$ match because 91 KOIs are omitted in
+the 1$''$ match, 23 of which have either ``candidate'' or
+``confirmed'' status.  Almost all of these 23 are M dwarfs with high
+proper motions.  Performing the same crossmatch at 4 arcseconds, only
+two KOIs (KOI-3993 and KOI-6531), both known false positives, fail to
+yield Gaia DR3 crossmatches.}, which is a crossmatch of the NASA
+Exoplanet Archive \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia
+DR3.  For KIC stars with multiple potential Gaia matches within the
+4$''$ search radius, we adopted the brightest star as the default
+match.  in most such cases this was unambiguous because there is a
+large brightness difference between the primary and any apparent
+neighbors.  Cases with multiple stars within 4$''$ within $\Delta
+G$$<$$0.5$\,mag are noted using bit flag 5.  The separation
+distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
+of candidate matches are within 1$''$.   We nonetheless noted an
+upturn in the candidate match rate from 3-4$''$; such sources are
+flagged using bit 4.
+
+{\it Gaia DR3 non-single-stars (bit 6)}---The Gaia DR3
+\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
+flags known eclipsing, astrometric, and spectroscopic binaries.  We
+directly include this column.
+
+{\it RUWE (bit 7)}---Following recommendations in the Gaia DR3
 documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
@@
 
-{\it Crowding}---We searched the Gaia DR3 point source catalog for
-stars within 1 Kepler pixel (4$''$) of every target star.
-While many such companions are not physically associated with the
-target star, their presence can bias stellar parameters and stellar
-rotation period measurements.
-We therefore flag any stars with neighbors down to 10$\times$ the 
-brightness of the target star within this region ($\Delta G < 2.5$).
-
-{\it Gaia DR3 Non-Single-Stars}---The Gaia DR3
-\texttt{non\_single\_star} column in the \texttt{gaia\_source} table
-flags known eclipsing, astrometric, and spectroscopic binaries.  We
-directly include this column.
+{\it Crowding (bit 8)}---We searched the Gaia DR3 point source catalog
+for stars within 1 Kepler pixel (4$''$) of every target star.  While
+many such companions are not physically associated with the target
+star, their presence can bias stellar parameters and stellar rotation
+period measurements.  We therefore flag any stars with neighbors down
+to 10$\times$ the brightness of the target star within this region
+($\Delta G < 2.5$).
+
+{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
+are far from the main sequence: this can be seen in color--absolute
+magnitude diagrams, and in diagrams of e.g.~surface gravity as a
+function of temperature (Figure~\ref{fig:stellarprops}).  This poses a
+problem for rotation-based age-dating because {\it i)} ``stars'' far
+from the main-sequence include unresolved and unidentified photometric
+binaries, and {\it ii}) rotation periods of single stars far from the
+main-sequence have likely evolved due to effects other than magnetic
+braking.  The added complications of reddening and metallicity make
+imposing a simple cut in these diagrams challenging.  After exploring
+various options, we found that the $+1$$\sigma$ relative isochronal
+uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
+B20,iso}/t_{\rm B20,iso}$, traced the relevant region of parameter
+space quite well.
+Another way of saying this is that gyrochronology is likely to be less
+reliable for stars for which the upper limit on the relative
+isochronal uncertainty is small.
+
+%FIXME TODO LEFT OFF HERE...
+
+df['b20t2_rel_E_Age'] = np.abs(df['b20t2_E_Age'])/df['b20t2_Age']
+
+This is potentially an issue for rotation-based age dating,
+particularly
+
+using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
+BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
+loci to flag over or under-luminous stars in each diagram, typically
+more than $\approx$0.5 magnitudes above or below the main sequence.
+%TODO refer to a figure!!!
 
@@
 
-Kepler~411 is also odd, due to the 58\,day-outer planet.
-\citet{2019A&A...624A..15S} conducted a transiting timing analysis of the
-system, confirming Kepler-411d and argued for the presence of a fourth
-non-transiting planet, Kepler-411e, based on the TTVs seen in Kepler-411d and
-assuming that the system would be nearly coplanar.
-Regarding the stellar age, \citet{2019A&A...624A..15S} used the 
-\citet{2007ApJ...669.1167B} relation, which yielded an estimate of 212$\pm$31\,Myr.
-However, the star's temperature ($\approx$4900\,K) puts it in a regime where
-the Barnes~relation has been known to not fit observations for quite some time
-\citep[see][Fig.~9]{2008ApJ...687.1264M}.
-Our adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
-calibrators.
+Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
+transiting planets at 3.0, 7.8, and 58\,days, has a peculiar
+architecture.  \citet{2019A&A...624A..15S} conducted a transiting
+timing analysis of the system, confirming Kepler-411d and arguing for
+the presence of a fourth non-transiting planet, Kepler-411e, based on
+the TTVs seen in Kepler-411d and assuming that the system is nearly
+coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
+the \citet{2007ApJ...669.1167B} relation and the measured rotation
+period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
+star's temperature ($\approx$4900\,K) puts it in a regime where the
+\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
+observations for quite some time
+\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
+$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
+
+Kepler-63b is interesting in that it is {\it missing} from our
+catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
+a polar orbit based on both the Rossiter-McLaughlin effect and a
+starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
+eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
+a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
+qualitatively consistent with the estimates from
+\citet{2013ApJ...775...54S}, though with larger uncertainties because
+our age accounts for the intrinsic scatter in the open cluster
+calibration sequences \citep{Bouma_2023}.  
+While the usual catalogs (\citetalias{Santos_2019},
+\citetalias{Santos_2021}, \citetlias{McQuillan_2014}, and
+\citetalias{Mazeh_2015}) do not report this star's rotation period.
+\citetalias{Santos_2021} flagged the system as ``potentially'' having
+rotation-based modulation.
+{\bf todo: how many comparable missed cases are there?}.
+%TODO FIXME it's bc you need to do mcquillan 2013....
+%
+%... David+2021 used:
+% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...
 

### 2024-04-16 paper/ms.tex
minor updates

@@
 stars with reported rotation periods and effective temperatures.  To
-select the subset stars for which we expect these age to in fact be
-valid, we then build a set of quality flags.  When and
-how these flags should be applied depends on the question being asked.
-If the goal is to construct a false-positive free sample, they might
-all be applied.
-If the goal is constructing a complete sample of young stars,
-then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
+select stars for which we expect these age to in fact be valid, we
+then build a set of quality flags.  When and how these flags should be
+applied depends on the question being asked.  If the goal is to
+construct a false-positive free sample, they might all be applied.  If
+the goal is constructing a complete sample of young stars, then
+consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
 Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
 resolved binary companion \citep{Bouma_2022a}; the latter is on a
-grazing orbit \citep{2014ApJ...783...53M}.  
-The flags in Tables~\ref{tab:stars} and~\ref{tab:planets} enable
-selecting for such cases using bitflags.
-
-{\it Temperature range (bit 0)}---We required stars to have effective
-temperatures $T_{\rm eff}$ between 3800 and 6200\,K
-\citep{Bouma_2023}.   Stars hotter than 6200\,K spin down very slowly,
-if at all, limiting the utility of rotation-based ages.  Stars cooler
-than 3800\,K do seem to spin down over gigayear timescales (e.g. CITE
-E.Newton thesis, the Engle paper, and Chiti's thing), but the
-currently available open cluster data have yet to clarify at what
-times, if any, the intrinsic scatter in the population decreases.
-
-{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants if they had
-$\log g < 4.2$.  We set this threshold by examining the
+grazing orbit \citep{2014ApJ...783...53M}.  The flags in
+Tables~\ref{tab:stars} and~\ref{tab:planets} enable selecting for or
+against such cases using bit flags.
+
+{\it Temperature range (bit 0)}---We require stars to have effective
+temperatures $T_{\rm eff}$ between 3800 and 6200\,K in order to report
+gyrochrone ages \citep{Bouma_2023}.   Stars hotter than 6200\,K spin
+down very slowly, if at all.  Stars cooler than 3800\,K do spin down
+over gigayear timescales (e.g. {\bf CITE} E.Newton thesis, the Engle
+paper, and Chiti's article), but the currently available open cluster
+data have yet to clarify at what times, if any, the intrinsic scatter
+in the population decreases.
+
+{\it Surface gravity (bit 1)}---We flagged stars as possible subgiants
+if they had $\log g < 4.2$.  We set this threshold by examining the
 relevant flag from \citet{berger_2018_radii_evolnstates},
 cross-matched against \citet{Berger_2020a_catalog}.
-\citet{berger_2018_radii_evolnstates} performed this classification
-using the PARSEC evolutionary models, stellar radii, and DR25 stellar
-properties catalog effective temperatures \citep{Mathur_2017}.
-
-{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia DR3
-$G$-band luminosity, ignoring reddening, using the reported apparent
-$G$-band magnitude and parallaxes.
-We flag stars with $M_{\rm G} < 3.9$, corresponding to main-sequence
-spectral types earlier than $\approx$F8V \citep{Pecaut_2013}.
-
-{\it Known eclipsing binaries (bit 3)}---We flagged any stars reported to be
-in the final Kepler eclipsing binary catalog (KEBC;
+%\citet{berger_2018_radii_evolnstates} performed this classification
+%using the PARSEC evolutionary models, stellar radii, and DR25 stellar
+%properties catalog effective temperatures \citep{Mathur_2017}.
+
+{\it Absolute luminosity (bit 2)}---We calculated the absolute Gaia
+DR3 $G$-band luminosity, ignoring reddening, using the reported
+apparent $G$-band magnitude and parallaxes.  We flag stars with
+$M_{\rm G} < 3.9$, corresponding to main-sequence spectral types
+earlier than $\approx$F8V \citep{Pecaut_2013}.
+
+{\it Known eclipsing binaries (bit 3)}---We flagged any stars reported
+to be in the final Kepler eclipsing binary catalog (KEBC;
 \citealt{2016AJ....151...68K}).
 
-{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In instances
+{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
 where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
-Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun/}; we
-used 4$''$ rather than the 1$''$ match because 91 KOIs are omitted in
-the 1$''$ match, 23 of which have either ``candidate'' or
-``confirmed'' status.  Almost all of these 23 are M dwarfs with high
-proper motions.  Performing the same crossmatch at 4 arcseconds, only
-two KOIs (KOI-3993 and KOI-6531), both known false positives, fail to
-yield Gaia DR3 crossmatches.}, which is a crossmatch of the NASA
-Exoplanet Archive \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia
-DR3.  For KIC stars with multiple potential Gaia matches within the
-4$''$ search radius, we adopted the brightest star as the default
-match.  in most such cases this was unambiguous because there is a
-large brightness difference between the primary and any apparent
-neighbors.  Cases with multiple stars within 4$''$ within $\Delta
-G$$<$$0.5$\,mag are noted using bit flag 5.  The separation
+Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
+accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
+because 91 KOIs are omitted in the 1$''$ match, 23 of which have
+either ``candidate'' or ``confirmed'' status.  Almost all of these 23
+are M dwarfs with high proper motions.  Performing the same crossmatch
+at 4 arcseconds, only two KOIs (KOI-3993 and KOI-6531), both known
+false positives, fail to yield Gaia DR3 crossmatches.}, which is a
+crossmatch of the NASA Exoplanet Archive
+\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
 distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
@@
 upturn in the candidate match rate from 3-4$''$; such sources are
-flagged using bit 4.
+flagged using bit 4.  For KIC stars with multiple potential Gaia
+matches within the 4$''$ search radius, we adopted the brightest star
+as the default match.  In most such cases this was unambiguous because
+there is a large brightness difference between the primary and any
+apparent neighbors.  However cases with multiple stars within 4$''$
+within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.
 
@@
 
-{\it RUWE (bit 7)}---Following recommendations in the Gaia DR3
-documentation\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR2/Gaia_archive/chap_datamodel/sec_dm_main_tables/ssec_dm_ruwe.html}},
-we flagged stars with a Gaia DR3 renormalized unit weight error (RUWE)
+{\it RUWE (bit 7)}---We flagged stars with a Gaia DR3 renormalized
+unit weight error
+(RUWE\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
 exceeding 1.4 as possible binaries.  Such astrometric outliers can be
-either bonafide astrometric binaries, or more often are marginally
+either bona fide astrometric binaries, or more often are marginally
 resolved point sources for which the single-source PSF model used in
@@
 magnitude diagrams, and in diagrams of e.g.~surface gravity as a
-function of temperature (Figure~\ref{fig:stellarprops}).  This poses a
+function of temperature (Figure~\ref{fig:stellarprops}).  This is a
 problem for rotation-based age-dating because {\it i)} ``stars'' far
-from the main-sequence include unresolved and unidentified photometric
-binaries, and {\it ii}) rotation periods of single stars far from the
-main-sequence have likely evolved due to effects other than magnetic
-braking.  The added complications of reddening and metallicity make
-imposing a simple cut in these diagrams challenging.  After exploring
-various options, we found that the $+1$$\sigma$ relative isochronal
+from the main-sequence include unresolved photometric binaries, and
+{\it ii}) rotation periods of single stars far from the main-sequence
+have likely evolved due to effects other than magnetic braking.  The
+added complications of reddening and metallicity make imposing a
+simple cut in any two dimensions challenging.  After exploring various
+options, we found however that the $+1$$\sigma$ relative isochronal
 uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
-B20,iso}/t_{\rm B20,iso}$, traced the relevant region of parameter
-space quite well.
-Another way of saying this is that gyrochronology is likely to be less
-reliable for stars for which the upper limit on the relative
-isochronal uncertainty is small.
-
-%FIXME TODO LEFT OFF HERE...
-
-df['b20t2_rel_E_Age'] = np.abs(df['b20t2_E_Age'])/df['b20t2_Age']
-
-This is potentially an issue for rotation-based age dating,
-particularly
-
-using the Gaia DR2 photometry, in $M_{\rm G}$ vs. $(G_{\rm
-BP} - G_{\rm RP})_0$, and $(G - G_{\rm RP})_0$.  We then manually drew
-loci to flag over or under-luminous stars in each diagram, typically
-more than $\approx$0.5 magnitudes above or below the main sequence.
-%TODO refer to a figure!!!
-
-{\it Final calibration sample}---{\bf todo: mention any extra flags
-from Berger2020, or perhaps in the KOI catalog itself, or perhaps the
-amplitude cuts you might want to apply to the rotation period catalog}
+B20,iso}/t_{\rm B20,iso}$, reasonably separated the relevant region of
+parameter space.  Another way of saying this is that gyrochronology is
+probably less reliable for stars for which the upper limit on the
+relative isochronal uncertainty is small.
+We flag stars with $+\sigma_{\rm B20,iso}/t_{\rm B20,iso} < 0.4$ as
+meeting this criterion.
+
+{\it Candidate pulsators and close-in binaries (bit
+10)}---\citeauthor{Santos_2021} included in a flag for candidate
+``classical pulsators'' (e.g. RR~Lyrae, Cepheids) and candidate
+close-in binaries.  Visual inspection hints that this flag also
+includes many bona fide young stars, and so although we propagate it 
+to our set of quality flags, we generally ignore it.
+
+An important decision for our subsequent analysis is how to combine
+the above flags in order to decide whether or not a star is suited for
+gyrochronology.  By default, we assume that a star is suitable for
+gyrochronology if none of bits zero through nine are raised.
 
@@
 in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
-observed by Kepler, requiring reported rotation periods yielded
-\nuniqstarsantosrot.  Further requiring these rotating stars
+observed by Kepler, requiring reported rotation periods and a
+reliable DR3 crossmatch yielded
+\nuniqstarsantosrot\ stars.  Further requiring these rotating stars
 to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
@@
 Further requiring them to be apparently single, near the main
-sequence, with $\log g$$>$4.2, and not flagged as
+sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
 eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
 potentially amenable for gyrochronology.
-Finally, imposing RUWE$<$1.4 leaves \nuniqstarsantosrotgyroappl\ stars
-for which gyrochronology is likely to be valid.
+Finally, imposing RUWE$<$1.4 and requiring no possible blends ($\Delta
+G$<2.5) within 4$''$ leaves \nuniqstarsantosrotgyroappl\ stars for
+which gyrochronology is likely to be valid.
 
@@
 	\caption{
-		{\bf Age distribution of stars and planet-hosts in the Kepler field
-    inferred from rotation periods}.
+    {\bf Age distribution of stars (left) and known planet hosts
+    (right) in the Kepler field inferred from rotation periods}.
     A histogram of samples from the age posteriors for all Kepler
@@
 Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
-transiting planets at 3.0, 7.8, and 58\,days, has a peculiar
+transiting planets at 3.0, 7.8, and 58\,days, has a rare
 architecture.  \citet{2019A&A...624A..15S} conducted a transiting
@@
 While the usual catalogs (\citetalias{Santos_2019},
-\citetalias{Santos_2021}, \citetlias{McQuillan_2014}, and
+\citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
 \citetalias{Mazeh_2015}) do not report this star's rotation period.

### 2024-04-16 drivers/plot_star_Prot_Teff.py

@@ -10,6 +10,6 @@ gp.plot_star_Prot_Teff(outdir, 'teff_age_prot_seed42_nstar20000')
 for sampleid in [
     'Santos19_Santos21_dquality',
     'Santos19_Santos21_all',
-    'Santos19_Santos21_clean0'
+    #'Santos19_Santos21_clean0'
 ]:
     gp.plot_star_Prot_Teff(outdir, sampleid)

### 2024-04-16 drivers/run_paper_scripts.sh

@@ -1,5 +1,14 @@
 #!/bin/bash
 
+python plot_star_Prot_Teff.py
+python plot_koi_mean_prot_teff.py
+python plot_hist_field_gyro_ages.py
+python plot_li_vs_teff.py
+python plot_st_params.py
+
+python get_values_for_manuscript.py
+python make_younghighlight_table.py
+
 python make_gyroonly_table.py
 cwd=$(pwd)
 cd ~/Dropbox/proj/NEA_age_plots/drivers
@@ -8,14 +17,5 @@ cd "$cwd"
 
 echo 'done'
 exit 0
-
-python plot_star_Prot_Teff.py
-python plot_koi_mean_prot_teff.py
-python plot_hist_field_gyro_ages.py
-python plot_li_vs_teff.py
-python plot_camd.py
-
-python get_values_for_manuscript.py
-python make_younghighlight_table.py
 
 

### 2024-04-17 paper/ms.tex
ms: add text for teff/logg split

@@
 function of temperature (Figure~\ref{fig:stellarprops}).  This is a
-problem for rotation-based age-dating because {\it i)} ``stars'' far
-from the main-sequence include unresolved photometric binaries, and
-{\it ii}) rotation periods of single stars far from the main-sequence
-have likely evolved due to effects other than magnetic braking.  The
-added complications of reddening and metallicity make imposing a
-simple cut in any two dimensions challenging.  After exploring various
-options, we found however that the $+1$$\sigma$ relative isochronal
-uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{\rm
-B20,iso}/t_{\rm B20,iso}$, reasonably separated the relevant region of
-parameter space.  Another way of saying this is that gyrochronology is
-probably less reliable for stars for which the upper limit on the
-relative isochronal uncertainty is small.
-We flag stars with $+\sigma_{\rm B20,iso}/t_{\rm B20,iso} < 0.4$ as
-meeting this criterion.
+challenge for rotation-based age-dating because {\it i)} point sources
+far from the main-sequence include unresolved photometric binaries,
+and {\it ii}) effects other than magnetic braking influence the
+rotation periods of bona fide single stars far from the main-sequence.
+The added complications of reddening and metallicity make defining a
+simple cut in any two dimensions challenging.  We explored various
+options, and found that the $+1$$\sigma$ relative isochronal
+uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{t,{\rm
+B20,iso}}/t_{\rm B20,iso}$, reasonably separated FGK stars ``near'' and
+``far'' from the main-sequence.  We therefore selected stars with
+$t_{\rm B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, and fitted an
+an $N^{\rm th}$-order polynomial in the $\log g$-$T_{\rm eff}$ plane
+to these stars between 4100 and 5800\,K.  We let the order of the fit
+vary from $N$=1 to 10, and minimizing the Bayesian Information Criterion
+yielded $N$=6.  The resulting locus is shown in
+Figure~\ref{fig:stellarprops}.  While the details of this procedure
+are entirely arbitrary, the overall purpose is to exclude stars for
+which there is photometric evidence that gyrochronology may not be
+reliable.  The rotation period distribution of sources excluded in
+this manner includes a larger fraction of ``rapid rotators'', likely
+due to a combination of astrophysical and systematic effects.
 

### 2024-04-17 paper/ms.tex
update ms

@@
   Here we apply this knowledge to planets and stars observed by the
-  primary Kepler mission (2009-2013), and report the resulting rotation and
+  primary Kepler mission, and report the resulting rotation and
   lithium-based ages.
@@
 		\subfloat{
-			\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
+			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
 		}
@@
 \startlongtable
-\begin{deluxetable*}{lllllllrrrrr}
+\begin{deluxetable*}{lllllllrrrrrr}
 	%\tabletypesize{\footnotesize}
@@
 	\startdata
-  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & Spec? \\
-  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &      bool &   bool  & bool \\
+  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & $Q_{\rm gyro}$ &  Spec? \\
+  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &      bool &   bool  & int & bool \\
 	\hline

### 2024-04-19 drivers/plot_li_vs_teff.py
update li vs teff plot

@@ -6,8 +6,21 @@ outdir = os.path.join(RESULTSDIR, 'li_vs_teff')
 if not os.path.exists(outdir): os.mkdir(outdir)
 
 # most useful
-ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality', yscale='linear', limodel='eagles', show_dispersion=1)
-ap.plot_li_vs_teff(outdir, sampleid='all', yscale='linear', limodel='eagles', show_dispersion=1)
+ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality',
+                   yscale='linear', limodel='eagles',
+                   show_dispersion=0)
+ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality',
+                   yscale='log', limodel='eagles',
+                   show_dispersion=0)
+
+ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality',
+                   yscale='linear', limodel='eagles',
+                   show_dispersion=1)
+
+ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality',
+                   yscale='linear', limodel='eagles', show_dispersion=0,
+                   nodata=1, show_dispersionpoints=1)
+
 assert 0
 
 # bonus options

### 2024-04-19 gyrojo/plotting.py

@@ -330,7 +330,8 @@ def plot_star_Prot_Teff(outdir, sampleid):
 
 
 def plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality', yscale=None,
-                    limodel='eagles', show_dispersion=0):
+                    limodel='eagles', show_dispersion=0, nodata=0,
+                    show_dispersionpoints=0):
 
     #df = get_li_data('all')
     df = get_li_data(sampleid)
@@ -394,13 +395,21 @@ def plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality', yscale=None,
         from eagles import eAT2EWm
 
         # set up a an equally spaced set of log temperatures between 3000 and 6500 K
-        tstep = 0.002
+        tstep = 0.00002
         Teff_model = np.arange(3.4772, 3.8130, tstep)
+        Teff_model = np.linspace(3.4772, 3.8130, int(1e4))
+        Teff_model = np.linspace(np.log10(3800), np.log10(6200), int(1e4))
 
-        if not show_dispersion:
-            ages = [120, 300, 670, 1000, 4000]
+        agestrs = None
+        if not show_dispersion and not show_dispersionpoints:
+            ages = [100, 500, 2000, 4000]
+            agestrs = ['100 Myr', '500 Myr', '2 Gyr', '4 Gyr']
+        elif show_dispersionpoints:
+            ages = [100, 500, 2000]
+            agestrs = ['100 Myr', '500 Myr', '2 Gyr']
         else:
-            ages = [120, 670, 4000]
+            ages = [100, 500, 2000, 4000]
+            agestrs = ['100 Myr', '500 Myr', '2 Gyr', '4 Gyr']
         lAges = [np.log10(t)+6 for t in ages]  # log age in years
 
         li_models = []
@@ -414,14 +423,24 @@ def plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality', yscale=None,
 
     set_style("clean")
     fig, ax = plt.subplots(figsize=(3,3))
+    np.random.seed(42)
+
+    if not isinstance(agestrs, list):
+        agestrs = [f"{a}" for a in ages]
 
     linestyles = ['solid', 'dotted', 'dashed', 'dashdot', 'solid']
-    for ix, (li_model, ls, age) in enumerate(zip(li_models, linestyles, ages)):
[…]

### 2024-04-19 paper/ms.tex
update ms

@@
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
-		}
+			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+		}
+%	
+%		\subfloat{
+%			\includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
+%		}
+%		\vspace{-0.5cm}
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
-		}
-
-		\subfloat{
-		\includegraphics[width=0.5\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
-		}
-		\subfloat{
-			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
-		}
+		\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+		}
+%		\subfloat{
+%			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
+%		}
 	\end{center}
@@
 	\caption{
-    {\bf todo}.
+    {\bf The stellar samples.}  Our analysis focuses on stars observed by Kepler with 
+    previously reported
+    rotation periods (dark gray points). 
+    About half of these rotators are suitable for gyrochronology (black points),
+    based on factors including their temperatures, surface gravities, (non-)binarity, and proximity to the main sequence.
+    The full list of criteria is given in Section~\ref{subsec:flags}.
+    A few percent of the remainder are either ``confirmed'' or ``candidate'' Kepler Objects of Interest that
+    meet an additional set of quality criteria (blue points; Section~\ref{subsec:plflags}).
+    These are the objects for which our rotation-based ages are most reliable.
+    We also report lithium-based ages for a largely-overlapping setting of KOIs
+    based on high-resolution spectroscopy.
+    Surface gravities and effective temperatures are primarily from
+    \citet{Berger_2020a_catalog}.
 	}
@@
 sequences to yield statistical age uncertainties.
+
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
+		}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf The models.}
+		Points represent $10^4$ draws from model probability distributions
+		that have been fitted to rotation periods 
+		\citep{Bouma_2023} and to lithium measurements \citep{Jeffries_2023}
+		of stars in open clusters.
+		Lines are the fitted ``mean models'' at various ages; these models are calibrated
+		for stars younger than 4\,Gyr.
+		The intrinsic dispersion in rotation and lithium at any given age, which what these models are designed to fit, sets
+		the theoretical precision limit for both age-dating methods.
+		Additional sources of uncertainty, including measurement uncertainty,
+		impose additional limitations on achievable age precision.
+		The displayed points assume an unphysical uniform distribution in
+		temperature for visual clarity, and sizes of the points are the same in each panel.
+		%    Gyrochrone age posteriors are broadest in regions 
+		%    with the slowest stellar spin-down.
+		\label{fig:models}
+	}
+\end{figure*}
 
@@
 \subsection{Planet Quality Flags}
+\label{subsec:plflags}
 For the subset of our analysis that focuses on the properties of
@@
 
-\begin{figure*}[!t]
+\begin{figure}[!t]
 	\begin{center}
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
-		}
+			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+		}
+
+		\vspace{-0.5cm}
 		\subfloat{
-			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
 	\caption{
-    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
+    {\bf Rotation periods for Kepler stars (top) and planets (bottom)}.
     The periods are from \citet{Santos_2019} and \citet{Santos_2021}
-    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
-    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
-    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
-    rotation sequences of open clusters.  The stellar sample shows
+    (\citetalias{Santos_2019} and \citetalias{Santos_2021}).  The
+    gray lines are ``mean fits'' from Figure~\ref{fig:models} to the slow
+    rotation sequences of M67, Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open clusters.
+    The stellar sample shows
     only apparently single stars near the main sequence with $\log
@@
 	\label{fig:prot_vs_teff}
-\end{figure*}
+\end{figure}
 
@@
 		\leavevmode
-		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf}
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf}
 	\end{center}
@@
     from Keck/HIRES spectra collected between 2009 and 2024.}
-    Lines represent mean models derived from Gaia-ESO spectra of stars
-    in clusters of various ages; the intrinsic dispersion 
-    evolves in time, and is modelled as the $\pm$1$\sigma$
-    bands~\citep{Jeffries_2023}.
-    As for rotation-based ages, the intrisic dispersion sets the
-    empirical precision limit for the technique.
+    Lines are the ``mean'' isochrones from the same models as in Figure~\ref{fig:models}.
+    The logarithmic scale highlights the dynamica range of both detections
+    and upper limits.
+%    The intrinsic
+%    dispersion in the cluster data evolves in time, and is shown as
+%    the $\pm$1$\sigma$ bands~\citep{Jeffries_2023}.
+%    As for rotation-based ages, the intrisic dispersion sets the
+%    empirical precision limit for the technique.
 		\label{fig:li_vs_teff}

### 2024-04-19 paper/collect.sh

@@ -1,5 +1,8 @@
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 
+# plot_gyromodeldispersion
+cp $r/gyromodeldispersion/gyromodeldispersion.pdf .
+
 # plot_st_params
 # ...CAMD:
 cp $r/st_params/st_params_M_G_vs_dr3_bp_rp.pdf .
@@ -19,7 +22,8 @@ cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf
 cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .
 
 # plot_li_vs_teff
-cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showdispersion.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
 
 # planets
 r="/Users/luke/Dropbox/proj/NEA_age_plots/results/rp_vs_period_scatter"

### 2024-04-19 paper/ms.tex
ms: add good new text

@@
 % number of KOIs in Q1-Q17 DR25 table
-\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
+%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
 % number of KOIs in Q1-Q17 DR25 table that are not known false positives
-\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
+%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
 % number of KOIs in Q1-Q17 DR25 table that have rotation periods
@@
 \begin{abstract}
-  Recent studies of stars in open clusters have yielded a detailed picture
-  of how the rotation rates and lithium content of FGK stars evolve
-  during the main sequence.
+  Recent studies of stars in open clusters have yielded detailed
+  pictures of how the rotation rates and lithium content of FGK stars
+  evolve during the first few billion years of their lives.
   %
@@
   %
-  Rotation periods are collected from previous analyses of
-  Kepler data; lithium is measured from new and archival HIRES
-  spectra.
+  Rotation periods are drawn from previous Kepler analyses; lithium is
+  measured from new and archival HIRES spectra.
   %
@@
   and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is likely valid. %, and for many more for which the underlying assumptions of
-%  the technique may be compromised.
+  gyrochronology is probably valid.
   %
@@
   this sample,
-  we report robust two-sided age measurements from both rotation and
-  lithium for 14 confirmed $<$500\,Myr planets, including five
-  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
-  Neptune-sized planet.
-  %
-  In the broader Kepler target star sample,
-  we find
-  that the gyrochrone age distribution rises linearly from 0 to
-  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
-  field as stars 1\,Gyr old.
-  %
-  This approach offers greater sensitivity to main sequence ages than
-  previous analyses of the Milky Way's star formation history, and the
-  resulting age distribution bounds the relative balance of ``young''
-  and ``old'' planets detectable not only by Kepler, but also by
-  future wide-field transit surveys.
+  we report new robust ages from both rotation and lithium for 14
+  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
+  super-Earths, three Earths and sub-Earths, and a Neptune-sized
+  planet.
+  %
+  In the broader Kepler target star sample, we find that the
+  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
+  stars 2\,Gyr old being twice as common in the Kepler field as stars
+  1\,Gyr old.
+  %
+  The resulting age distribution bounds the relative balance of
+  ``young'' and ``old'' planets detectable not only by Kepler, but
+  also by future wide-field transit surveys.
   %
@@
 Discovering a young planet requires solving two problems: find the
-planet, and measure the star's age.  Each problem can be solved in
-various ways, and so there are many paths for discovery.  In this
-manuscript, we will focus on planets discovered using transits, and
-stellar ages measured based using empirical rotation and lithium-based
-methods.  Rotation-based age-dating will provide the most information
-for the largest number of stars.  The lithium-based technique will
-provide useful information for the subset of stars younger than a few
-hundred million years.
-
-To help build intuition, it is useful to assume that the age
-distribution of stars within the nearest few hundred parsecs of the
-Sun (i.e.~in the Solar Neighborhood) is uniform between 0 and 10\,Gyr
-\citep{Nordstrom_2004}.  We will revisit this assumption later in the
-manuscript.  While the youngest 0.1\% of stars ($\lesssim$10\,million
-years; Myr) are perhaps the most constraining for studies of planet
-formation, they are extremely challenging targets for close-in
-exoplanet detection 
-\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
-the early orbital, structural, and atmospheric evolution of exoplanets
-are perhaps most technically feasible for the $\approx$1\% of stars
-with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
-younger than 1\,Gyr.  However, since stellar ages are roughly
-uniformly distributed in the solar neighborhood, there is a limit on
-the sample size of young planets that will be detectable in these age
-regimes, particularly relative to their older breathren.
+planet, and measure the star's age.  Each problem admits a range of
+solutions.  Here, we will focus on planets found using transits,
+and stellar ages inferred using rotation and lithium.  Rotation
+will provide the most age information for the largest number of
+stars.  The lithium-based technique will provide useful information
+for stars younger than a few hundred million years.
+
+To begin, let us imagine that the age distribution of stars within a
+few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
+\citep[e.g.][]{Nordstrom_2004}.  
+The utility of this approximation is that it provides a rough estimate
+for the rarity of young stars:
+$\approx$1\% of stars in the nearby Galaxy have ages below 100\,Myr,
+and $\approx$10\% are younger than 1\,Gyr.
+Observations of exoplanets forming in disks (CITE), and of exoplanets
+evolving shortly after disk dispersal (CITE CITE), are therefore
+severely limited in their achievable sample sizes by the
+tyranny of the Galactic star formation rate.
+
+%While the youngest 0.1\% of stars ($\lesssim$10\,million
+%years; Myr) are perhaps the most constraining for studies of planet
+%formation, they are extremely challenging targets for close-in
+%exoplanet detection 
+%\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
+%the early orbital, structural, and atmospheric evolution of exoplanets
+%are perhaps most technically feasible for the 
+%$\approx$1\% of stars
+%with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
+%younger than 1\,Gyr.  However, since stellar ages are roughly
+%uniformly distributed in the solar neighborhood, there is a limit on
+%the sample size of young planets that will be detectable in these age
+%regimes, particularly relative to their older breathren.
 
 %TODO mention the RV approach?  at least the Hyades etc work...
-A wide range of approaches have, over the past ten years, begun to
-establish the subfield of young, close-in planet discovery.
+
+Despite the observational restrictions, the field of young close-in
+planet discovery has seen remarkable success over the past ten years.
 Pioneering discoveries with the prime Kepler mission (hereafter,
-``Kepler'') focused on stars with known ages---stars in clusters---and
-searched them for transiting planets \citep{Meibom_2013}.  Planets
-discovered in this way are assumed to have the same age as the host
-cluster.  The
-resulting ages are precise at the $\approx$10\% level.  K2
-and TESS have yielded larger numbers of young planets in clusters than
-Kepler, because each of these missions has observed a larger fraction
-of the sky, and therefore more stars with known ages
+``Kepler'') focused on stars with known ages---stars in open
+clusters---and searched them for transiting planets
+\citep{Meibom_2013}.  Planets discovered in this way are assumed to
+have the same age as the host cluster.  The resulting ages are precise
+at the $\approx$10\% level.  K2 and TESS have yielded larger numbers
+of young planets in clusters than Kepler, because each of these
+missions has observed a larger fraction of the sky, and therefore more
+stars with known ages
 \citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.
 
-A limitation of the ``cluster-first'' approach is that only
+A limitation of this ``cluster-first'' approach is that only
 $\approx$1\% of stars within the nearest few hundred parsecs can
@@
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-This means that there are many ``young'' (0.05-1\,Gyr) stars in the
-field that are omitted by cluster-only searches.
-Of order 40 transiting planets are currently known to be younger than
-1\,Gyr, requiring an age precision $t/\sigma_t > 3$.
-%TODO cite???
-% (see
-%Figure~\ref{fig:rp_period_age}).
-However, Kepler discovered
-$\approx$4{,}000 planets; and so unless there is a major discovery
-bias, we would expect a larger fraction of the Kepler
-planets---perhaps 10\%, or 400 planets---to be younger than 1\,Gyr.
+While there may be paths toward resolving this problem, this currently
+means that there are many ``young'' (0.05-1\,Gyr) stars in the field
+that are omitted by cluster-only searches.  Requiring ages precise to
+a factor of three ($t/\sigma_t > 3$), the NASA Exolpanet Archive
+currently reports $\approx$40 transiting planets younger than 1\,Gyr,
+essentially all of which are in clusters.  However, Kepler discovered
+$\approx$4{,}000 planets.  Unless there is a major discovery bias, we
+would expect a larger fraction of the Kepler planets---perhaps 10\%,
+or 400 planets---to be younger than 1\,Gyr, based on our order of
+magnitude reasonsing.
 
@@
 
-Have these young Kepler planets already been found?  If they have, the
-most likely studies to have found them would be those by
+Have these young Kepler planets already been found?  If they have, 
+it would probably have been by
 \citet{Berger_2020b_rpage}, \citet{David_2021} or
@@
 $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
-
-Theoretical predictions for planet evolution in the Kepler sample are
-most important in the context of mini-Neptunes and super-Earths, since
-these are the most abundant planets.  Pertinent predictions include:
-{\it a)} that mini-Neptunes should cool and contract, roughly
-following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
-b)} that large and close-in mini-Neptunes at the edge of the
-photoevaporation desert should shrink over the atmospheric loss
-timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
-abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
-should decrease over $t_{\rm loss}$---the upper-envelope of the
-mini-Neptune size distribution should corresponding decrease,
-e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
-$\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
-\citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
-mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
-\citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
-decreasing by at least a factor of two or three, and the super-Earth
-abundance increasing in tandem; {\it e)} that the
-upper boundary of the super-Earth population should increase over the
-same timescale as $\theta$ is increasing, since more massive
-super-Earth cores should be able to retain their envelopes for longer
-than less massive cores.
-
-The details of the predictions are flexible, in that they depend on
-the distributions of planetary core masses, initial atmospheric mass
-fractions, and initial entropies of the atmospheres after disk
-dispersal.  For instance, different sources of heating---from the star
-\citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
-\citep{Gupta_2019}, or even giant impacts
-\citep{Biersteker_Schlichting_2019}---can produce very different
-histories for the size evolution of planets at these early times
-\citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
-the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
-generic prediction that the ``upper envelope'' of the mini-Neptune
-size distribution should shrink due to Kelvin-Helmholtz cooling is
-hard to avoid; predictions however concerning the ``radius valley''
-and movement of planets across it are dependent on the timescales and
-prevalence of atmospheric mass loss.  In specific scenarios for
-initial atmospheric accretion \citep{Lee_2022} and early atmospheric
-evolution \citep{2024MNRAS.529.2716R} the radius gap might form
-without needing to invoke post-disk dispersal mass loss at all.  In
-such scenarios, there could be very little evolution of the planetary
-size distribution between 1 and 4\,\rearth\ over the first gigayear.
-
-Can stellar age measurements provide evidence for or against
-these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
-convincing---or even suggestive--- detection reported (see
-\citealt{Petigura_2022}).  While an apparent over-abundance of large
-very young planets ({\it (c)}) has been noted by multiple authors
-\citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
-exacerbated by a selection effect against sub-Neptune sized planet
-detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
-nonetheless found that when accounting for completeness, there is
-$\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
-$\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
-\citep{2024arXiv240303261V} stars than in the older Kepler sample
-\citep[e.g.][]{2020AJ....159..248K}.
-%TODO: more kepler citations...
-
-Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
-(e)} were explored by \citet{Berger_2020b_rpage} and the twin
-\citet{David_2021} and \citet{Sandoval_2021} studies.  Both
-\citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
-increases with time, at a statistical significance of $\approx$2-3$\sigma$.
-For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
-sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
-property-matched sample of planets older than 1\,Gyr.  In
-\citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
-performing a Monte Carlo re-sampling procedure from the isochronal ages and
-planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
-effect {\it (e)} was reported by \citet{David_2021}, who found that the average
-location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
-$\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
-
-The fact that stellar ages---and especially isochronal ages---are
-correlated with stellar masses and metallicities is a source of stress
-for many in this subfield.  \citet{Sandoval_2021} for instance noted a
-strong trend in $\theta$ as a function of stellar metallicity in
-addition to the trend they noted in age, and cautioned that this could
-also explain their observations.  Independently, \citet{Petigura_2018}
-and \citet{Petigura_2022} have reported trends between small-planet
-occurrence and both stellar metallicity and mass; these parameters
-seem, at first glance, to be more important predictors of bulk changes
-in the planetary population than age.  Given these known correlations,
-it can be useful to consider population models in which variables such
-as metallicity, mass, and age are added in succession
-\citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
-rigorous approach, is to jointly infer the number of planets per star
-as a function of both planetary and stellar parameters, the latter
-including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
-most effective when the measurement precision of these different
-parameters is as high as possible.
+(TODO: note that there's a lot more literature on "young" field Kepler
+planets than these few sentences betray).
+
+%      Theoretical predictions for planet evolution in the Kepler sample are
+%      most important in the context of mini-Neptunes and super-Earths, since
+%      these are the most abundant planets.  Pertinent predictions include:
+%      {\it a)} that mini-Neptunes should cool and contract, roughly
+%      following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
+%      b)} that large and close-in mini-Neptunes at the edge of the
+%      photoevaporation desert should shrink over the atmospheric loss
+%      timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
+%      abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
+%      should decrease over $t_{\rm loss}$---the upper-envelope of the
+%      mini-Neptune size distribution should corresponding decrease,
+%      e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
+%      $\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
+%      \citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
+%      mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
+%      \citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
+%      decreasing by at least a factor of two or three, and the super-Earth
+%      abundance increasing in tandem; {\it e)} that the
+%      upper boundary of the super-Earth population should increase over the
+%      same timescale as $\theta$ is increasing, since more massive
+%      super-Earth cores should be able to retain their envelopes for longer
+%      than less massive cores.
+%      
+%      The details of the predictions are flexible, in that they depend on
+%      the distributions of planetary core masses, initial atmospheric mass
+%      fractions, and initial entropies of the atmospheres after disk
+%      dispersal.  For instance, different sources of heating---from the star
+%      \citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
+%      \citep{Gupta_2019}, or even giant impacts
+%      \citep{Biersteker_Schlichting_2019}---can produce very different
+%      histories for the size evolution of planets at these early times
+%      \citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
+%      the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
+%      generic prediction that the ``upper envelope'' of the mini-Neptune
+%      size distribution should shrink due to Kelvin-Helmholtz cooling is
+%      hard to avoid; predictions however concerning the ``radius valley''
+%      and movement of planets across it are dependent on the timescales and
+%      prevalence of atmospheric mass loss.  In specific scenarios for
+%      initial atmospheric accretion \citep{Lee_2022} and early atmospheric
+%      evolution \citep{2024MNRAS.529.2716R} the radius gap might form
+%      without needing to invoke post-disk dispersal mass loss at all.  In
+%      such scenarios, there could be very little evolution of the planetary
+%      size distribution between 1 and 4\,\rearth\ over the first gigayear.
+%      
+%      Can stellar age measurements provide evidence for or against
+%      these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
+%      convincing---or even suggestive--- detection reported (see
+%      \citealt{Petigura_2022}).  While an apparent over-abundance of large
+%      very young planets ({\it (c)}) has been noted by multiple authors
+%      \citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
+%      exacerbated by a selection effect against sub-Neptune sized planet
+%      detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
+%      nonetheless found that when accounting for completeness, there is
+%      $\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
+%      $\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
+%      \citep{2024arXiv240303261V} stars than in the older Kepler sample
+%      \citep[e.g.][]{2020AJ....159..248K}.
+%      %TODO: more kepler citations...
+%      
+%      Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
+%      (e)} were explored by \citet{Berger_2020b_rpage} and the twin
+%      \citet{David_2021} and \citet{Sandoval_2021} studies.  Both
+%      \citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
+%      increases with time, at a statistical significance of $\approx$2-3$\sigma$.
+%      For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
+%      sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
+%      property-matched sample of planets older than 1\,Gyr.  In
+%      \citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
+%      performing a Monte Carlo re-sampling procedure from the isochronal ages and
+%      planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
+%      effect {\it (e)} was reported by \citet{David_2021}, who found that the average
+%      location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
+%      $\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
+%      
+%      The fact that stellar ages---and especially isochronal ages---are
+%      correlated with stellar masses and metallicities is a source of stress
+%      for many in this subfield.  \citet{Sandoval_2021} for instance noted a
+%      strong trend in $\theta$ as a function of stellar metallicity in
+%      addition to the trend they noted in age, and cautioned that this could
+%      also explain their observations.  Independently, \citet{Petigura_2018}
+%      and \citet{Petigura_2022} have reported trends between small-planet
+%      occurrence and both stellar metallicity and mass; these parameters
+%      seem, at first glance, to be more important predictors of bulk changes
+%      in the planetary population than age.  Given these known correlations,
+%      it can be useful to consider population models in which variables such
+%      as metallicity, mass, and age are added in succession
+%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
+%      rigorous approach, is to jointly infer the number of planets per star
+%      as a function of both planetary and stellar parameters, the latter
+%      including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
+%      most effective when the measurement precision of these different
+%      parameters is as high as possible.
 
@@
 dynamo all evolve
-\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Key
-for our analysis is the fact that we can empirically calibrate
-rotation-based ages using measured sequences of rotation periods
-in open clusters
-(\citealt{Bouma_2023}, and references therein).  This approach
-ultimately ties the rotation age scale to the cluster age scale, and
-is limited in precision by the intrisic scatter in the $P_{\rm rot}$
-sequences, and in accuracy by systematic uncertainties in the evolving
-spin-down rates, and in the cluster ages themselves.
-The oldest cluster for which this method has been calibrated is currently
-M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
-
-Lithium-based age-dating includes two qualitatively distinct regimes:
-depletion boundary ages derived for M-dwarfs in star clusters, and the
-less precise decline-based ages for individual field FGK stars
+\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
+Today however, measurements of stellar rotation periods in open
+clusters provide the most accurate empirical calibrations of
+rotation-based ages (\citealt{Bouma_2023}, and references therein).
+This approach ultimately ties the rotation age scale to the cluster
+age scale, and is limited in precision by the intrisic scatter in the
+$P_{\rm rot}$ sequences, and in accuracy by systematic uncertainties
+in the evolving spin-down rates, and in the cluster ages themselves.
+The oldest cluster for which this method has been calibrated is
+currently M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
+
+Lithium ages includes two qualitatively distinct regimes: depletion
+boundary ages derived for M-dwarfs in star clusters, and the less
+precise decline-based ages for individual field FGK stars
 \citep{Soderblom_2010}.  Here we focus on the latter approach, which
@@
 
-The methods by which we select the stellar and planet samples are 
-discussed in Section~\ref{sec:selection}.
-We summarize the origin of stellar parameters other than the ages in
-Section~\ref{sec:stellarprops}, 
-and describe and validate our age-dating methods in
+We discuss our methods for selecting stellar and planetary samples in
+Section~\ref{sec:selection}, and describe
+the origin of our adopted stellar parameters other than ages in
+Section~\ref{sec:stellarprops}.
+We describe and validate our age-dating methods in
 Sections~\ref{sec:rotage} and~\ref{sec:liage}.
@@
 
-\section{Sample Selection}
+\section{Selecting the Stellar and Planetary Samples}
 \label{sec:selection}
 
-This work focuses on stars surveyed by Kepler for which ages can be
-inferred using either stellar rotation, lithium, or both.  These
-``age-dateable stars'' are a relatively small subset of the
-\nkeplerstars\ Kepler targets.  Rotation periods provide the majority
-of the information, since they have been reported for
-$\approx$55{,}000 Kepler stars
-\citep[e.g.][]{McQuillan_2014,Santos_2021}.  
-High-resolution measurements of the \ion{Li}{1} 6708\,\AA\ doublet are
-available for only a few thousand Kepler objects of interest (KOIs).
-We therefore define our selection function as Kepler stars
-with measured rotation periods ($\mathcal{S}$;
-Section~\ref{subsec:starsel}), and Kepler planets whose host stars
-have measured rotation periods ($\mathcal{P}$;
-Section~\ref{subsec:planetsel}).
-
-\subsection{Rotation periods}
-\label{subsec:starsel}
-
-We select stars with rotation periods that have been measured by
-previous investigators.  Studies of stellar rotation across the entire
-Kepler field include those by \citet{McQuillan_2014}
-(\citetalias{McQuillan_2014}), \citet{Reinhold_2015},
-\citet{Santos_2019} (\citetalias{Santos_2019}), and
-\citet{Santos_2021} (\citetalias{Santos_2021}).
+This work is focused on Kepler stars for which ages can be inferred
+using either stellar rotation, lithium, or both.  These ``age-dateable
+stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
+periods provide the majority of the information, since they have been
+reported for roughly one in three Kepler targets
+\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
+and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
+general have only been acquired for the Kepler objects of interest
+(KOIs), which comprise only a few percent of the overall stellar
+sample.  In the following, we will describe the set of stars that we
+adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
+the set of objects we adopt as planets
+(Section~\ref{subsec:planetsel}), and the set of planets with
+high-resolution spectra suitable for lithium analysis
+(Section~\ref{subsec:lithiumsel}).
+
+\subsection{Stellar Rotation Periods}
+\label{subsec:rotsel}
+
+To select stars with rotation periods, we turn to previous work.
+Studies of stellar rotation across the entire Kepler field include
+those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
+\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
+and \citet{Santos_2021} (\citetalias{Santos_2021}).
 \citetalias{McQuillan_2014} used an approach based on the
@@
 rotation periods based on updated knowledge of contaminants
-(e.g.~giant stars and eclipsing binaries) and visual inspection.
-In addition,
-\citetalias{Santos_2021} report rotation periods for 24{,}182
-main-sequence and subgiant FGKM stars that were not reported as
-periodic by \citetalias{McQuillan_2014}.  Many of
-these reported detections have lower variability amplitudes, and
-longer periods, than those reported by
-\citetalias{McQuillan_2014}.  A relevant flag for stars
-missing \citetalias{McQuillan_2014} periods is included in
-Table~\ref{tab:stars} {\bf TODO make this}.  Reviewing the resulting period detections, we
-learned that 32 KOIs with orbital and rotation periods within
-$\approx$20\% were excluded from the \citetalias{Santos_2019} and
-\citetalias{Santos_2021} catalogs (A.~Santos, priv.~comm.) ; we choose
-to include these objects.\footnote{In general, excluding stars in this
-category would impose a bias against young close-in planets, due to
-the natural commensurability between the rotation rates of young
-stars, and Kepler's increased sensitivity to short-period objects.
-While some of these KOIs are known false positives, our analysis
-naturally excludes such objects at later steps.  As one motivating
-example, this subset of stars happens to include {\it
-e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
-$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
-based on membership in RSG-5 \citep{Bouma_2022b}.  }
-The final tally of stars from \citetalias{Santos_2019},
-\citetalias{Santos_2021}, and our corrected KOI list is
-\nuniqstarsantosrot\ Kepler stars with rotation periods.
+(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
+addition, \citetalias{Santos_2021} report rotation periods for
+24{,}182 main-sequence and subgiant FGKM stars that were not reported
+as periodic by \citetalias{McQuillan_2014}.  Many of these reported
+detections have lower variability amplitudes, and longer periods, than
+those reported by \citetalias{McQuillan_2014}.  A relevant flag for
+stars missing \citetalias{McQuillan_2014} periods is included in
+Table~\ref{tab:stars}.  Reviewing the resulting
+period detections, we learned that 32 KOIs with orbital and rotation
+periods within $\approx$20\% were excluded from the
+\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
+(A.~Santos, priv.~comm.); we choose to include these
+objects.\footnote{In general, excluding stars in this category would
+impose a bias against young close-in planets, due to the natural
+commensurability between the rotation rates of young stars, and
+Kepler's increased sensitivity to short-period objects.  While some of
+these KOIs are known false positives, our analysis naturally excludes
+such objects at later steps.  As one motivating example, this subset
+of stars happens to include {\it e.g.}~Kepler~1643 (KIC~8653134), with
+$P_{\rm orb}$=5.34\,days and $P_{\rm rot}$=5.05\,days, which is known
+to be $\approx$40\,Myr old based on membership in RSG-5
+\citep{Bouma_2022b}.  } The final tally of stars from
+\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
+KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.
 
@@
 %  %TODO have you?
-
-\subsection{Kepler objects of interest}
-\label{subsec:planetsel}
-
-We considered two different samples of planets: those in the fully
-automated Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which is
-suitable for planet occurrence rate calculations, and also those in
-the cumulative KOI table, which included the best knowledge available
-on any given planet candidate as of 2023 Jun 6, but incorporated
-human-based vetting \footnote{See
-\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
-For the main body of this work, we will refer to the former
-homogeneous sample, which includes \nkois\ KOIs, \nkoisnofp\ that are
-not known false positives; Appendix~\ref{app:cumulativekoi} highlights
-some additional young planets that are omitted in this approach.
-
-Crossmatching the Q1-Q17 DR25 KOI table against our stars with
-rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of
-which \nkoisnofpwithprot\ are not known false postives.  These objects
-will comprise the core data for our gyrochronology analysis.
 
@@
 $\lesssim$0.03$P_{\rm rot}$ level, and the upper envelope of the
-period difference tended to increase linearly with period.  Based on
+period difference increased roughly linearly with period.  Based on
 this comparison, we adopted a simple prescription for the period
@@
 $P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
-set to guarantee 3\% $P_{\rm rot}$ uncertainties at 30 days.
-
-\subsection{Lithium sample}
-A final subcomponent of our analysis involves an age assessment based on
-the equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
-component of the work, we primarily rely on spectra acquired with the
-High Resolution Echelle Spectrometer (HIRES;
-\citealt{vogt_hires_1994}) on the Keck I 10m telescope.
-Most of these spectra were acquired between 2010 and 2020
-as part of the California Keck Survey (CKS).
-A subset were acquired by our team in our own observing programs.
-%TODO quantify.
-%TODO compare to berger2018?
-
+set to guarantee 3\% $P_{\rm rot}$ uncertainties at rotation periods
+of 30 days.
+
+\subsection{Kepler Objects of Interest}
+\label{subsec:planetsel}
+
+We considered planets in the NASA Exoplanet Archive's cumulative KOI
+table, which included the best knowledge available on any given planet
+candidate as of 6 June 2023, while also incorporating human-based
+vetting \footnote{See
+\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
+These planets represent a superset of those in the fully automated
+Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
+future work for planet occurrence rate calculations.  This version of
+the cumulative KOI table included \nkoisnofp\ that are either
+``confirmed'' or ``candidate'' planets, after excluding known false
+positives.
+
+%Crossmatching the Q1-Q17 DR25 KOI table against our stars with
+%rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of
+%which \nkoisnofpwithprot\ are not known false postives.  These objects
+%will comprise the core data for our gyrochronology analysis.
+
+\subsection{High Resolution Spectra}
+A final piece of our analysis involves an age assessment based on the
+equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
+component of the work, we rely on spectra acquired with the High
+Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
+the Keck I 10m telescope.  These spectra were collected between 2009
+and 2024, primarily through projects such as the California Kepler
+Survey
+\citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
+We supplemented the existing archive with $\approx$10\,hours of new
+observations for 22 stars between Fall 2022 and Spring 2024.
+These targets were selected to ensure that confirmed planets with
+evidence from rotation for ages below 1\,Gyr received spectra, since
+this is the age range where lithium was most likely to yield
+useful age constraints.
+
+Lithium equivalent widths and abundances were already analyzed by
+\citet{2018ApJ...855..115B} for about three quarters of the spectra in
+our sample.
+% stat from _estimate_new_lithium_timefrac.py
+However, since new spectra have been acquired since the
+\citeauthor{2018ApJ...855..115B} analysis, and since our approach and
+selection function are different, we opt to perform our own analysis
+of the reduced HIRES spectra.
+
+%FIXME FIXME TODO LEFT OFF HERE
+%FIXME FIXME TODO LEFT OFF HERE
+%FIXME FIXME TODO LEFT OFF HERE
 These HIRES spectra are available for \nkoiswithhires\ of the
@@
 	\caption{
-		{\bf Sizes, orbital periods, and rotation-based ages of young transiting exoplanets}.
-		Two different selection functions for what defines a young, sub-gigayear, planet
-		are imposed on the {\it top} and {\it bottom}: the former requires $t$>1\,Gyr with
+		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
+		Two possible selection functions for what defines a young, sub-gigayear, planet
+		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
 		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
 		Circles are previously reported measurements from the literature.
-		Stars are newly calculated ages for Kepler planets from this work.
+		Stars are newly calculated ages for Kepler planets from this work; circles
+		are planets from the literature meeting the same requirements.
 		Gray points are transiting planets from the NASA Exoplanet Archive older than

### 2024-04-20 paper/ms.tex
ms: text improvements

@@
 % fraction of stars with rotation periods, with teff/logg in Berger+20
-\newcommand{\fracstarswithprotwithbtwenty}{{$\sim$94\%}}
+\newcommand{\fracstarswithprotwithbtwenty}{{$\approx$94\%}}
 % complement of above
-\newcommand{\fracstarswithprotwithoutbtwenty}{{$\sim$6\%}}
+\newcommand{\fracstarswithprotwithoutbtwenty}{{$\approx$6\%}}
 
@@
 % fraction of KOIs that are not FPs with B+20 parameters
-\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\sim$92\%}}
+\newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}
 
@@
 planet, and measure the star's age.  Each problem admits a range of
-solutions.  Here, we will focus on planets found using transits,
-and stellar ages inferred using rotation and lithium.  Rotation
-will provide the most age information for the largest number of
-stars.  The lithium-based technique will provide useful information
-for stars younger than a few hundred million years.
+solutions.  Here, we will focus on planets previously found using
+transits, and infer new stellar ages using rotation and lithium.
+Rotation will provide the most age information for the largest number
+of stars.  The lithium technique will provide useful information for
+stars younger than a few hundred million years.
 
@@
 \citep[e.g.][]{Nordstrom_2004}.  
-The utility of this approximation is that it provides a rough estimate
-for the rarity of young stars:
-$\approx$1\% of stars in the nearby Galaxy have ages below 100\,Myr,
+This approximiation is useful because it clarifies the rarity of young
+stars:
+$\approx$1\% of nearby stars have ages below 100\,Myr,
 and $\approx$10\% are younger than 1\,Gyr.
-Observations of exoplanets forming in disks (CITE), and of exoplanets
+Studies of exoplanets forming in disks (CITE), and of exoplanets
 evolving shortly after disk dispersal (CITE CITE), are therefore
-severely limited in their achievable sample sizes by the
-tyranny of the Galactic star formation rate.
+capped in their maximum achievable sample sizes by the tyranny of the
+Galactic star formation rate.
 
@@
 This work is focused on Kepler stars for which ages can be inferred
-using either stellar rotation, lithium, or both.  These ``age-dateable
+using either rotation, lithium, or both.  These ``age-dateable
 stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
-periods provide the majority of the information, since they have been
-reported for roughly one in three Kepler targets
+periods have been reported for roughly one in three Kepler targets
 \citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
@@
 general have only been acquired for the Kepler objects of interest
-(KOIs), which comprise only a few percent of the overall stellar
-sample.  In the following, we will describe the set of stars that we
+(KOIs), which comprise a few percent of Kepler's targets.
+In the following, we will describe the set of stars that we
 adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
@@
 future work for planet occurrence rate calculations.  This version of
-the cumulative KOI table included \nkoisnofp\ that are either
+the cumulative KOI table included \nkoisnofp\ objects that are either
 ``confirmed'' or ``candidate'' planets, after excluding known false
@@
 \subsection{High Resolution Spectra}
+\label{subsec:lithiumsel}
 A final piece of our analysis involves an age assessment based on the
 equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
-component of the work, we rely on spectra acquired with the High
+component of the work, we rely on spectra acquired using the High
 Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
 the Keck I 10m telescope.  These spectra were collected between 2009
-and 2024, primarily through projects such as the California Kepler
-Survey
+and 2024, primarily through the California Kepler Survey
 \citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
-We supplemented the existing archive with $\approx$10\,hours of new
+We supplemented the existing archive with $\approx$10~hours of new
 observations for 22 stars between Fall 2022 and Spring 2024.
-These targets were selected to ensure that confirmed planets with
-evidence from rotation for ages below 1\,Gyr received spectra, since
-this is the age range where lithium was most likely to yield
-useful age constraints.
+These stars were chosen to ensure that confirmed planets with
+rotational evidence for ages below 1\,Gyr received spectra, since this
+is the age range where lithium is most likely to yield useful age
+constraints.
 
@@
 	\caption{
-    {\bf The stellar samples.}  Our analysis focuses on stars observed by Kepler with 
-    previously reported
-    rotation periods (dark gray points). 
-    About half of these rotators are suitable for gyrochronology (black points),
-    based on factors including their temperatures, surface gravities, (non-)binarity, and proximity to the main sequence.
-    The full list of criteria is given in Section~\ref{subsec:flags}.
-    A few percent of the remainder are either ``confirmed'' or ``candidate'' Kepler Objects of Interest that
-    meet an additional set of quality criteria (blue points; Section~\ref{subsec:plflags}).
-    These are the objects for which our rotation-based ages are most reliable.
-    We also report lithium-based ages for a largely-overlapping setting of KOIs
-    based on high-resolution spectroscopy.
-    Surface gravities and effective temperatures are primarily from
+    {\bf The stellar samples.}  Our analysis focuses on stars observed
+    by Kepler with previously reported rotation periods (dark gray
+    points).  About half of these rotators are suitable for
+    gyrochronology (black points), based on factors including their
+    temperatures, surface gravities, (non-)binarity, and proximity to
+    the main sequence.  The full list of criteria is given in
+    Section~\ref{subsec:flags}.  A few percent of the remainder are
+    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
+    that meet an additional set of quality criteria (blue points;
+    Section~\ref{subsec:plflags}).  These are the objects for which
+    our rotation-based ages are most reliable.  We also report
+    lithium-based ages for a largely-overlapping setting of KOIs based
+    on high-resolution spectroscopy.  Surface gravities and effective
+    temperatures are primarily from
     \citet{Berger_2020a_catalog}.
@@
 
-As our default source for stellar temperatures and surface gravities,
-we adopt the values from the Gaia-Kepler Stellar Properties Catalog
-(GKSP; \citealt{Berger_2020a_catalog}).  The GKSP parameters are
-available for stars with ``AAA'' 2MASS photometry, measured parallaxes
-in Gaia DR2,  and $g$-band photometry from either the KIC or the
-Kepler-INT survey.  The parameters themselves were derived using
+Our default source for stellar temperatures and surface gravities is
+the Gaia-Kepler Stellar Properties Catalog (GKSP;
+\citealt{Berger_2020a_catalog}).  The GKSP parameters were reported
+for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
+DR2,  and $g$-band photometry from either the KIC or the Kepler-INT
+survey.  The parameters themselves were derived using
 \texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
@@
 \fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
-stars with rotation periods.
-\citet{David_2021} compared the stellar parameters ($T_{\rm eff}$,
-$R_\star$, [Fe/H]) derived from the \citet{Berger_2020a_catalog}
-isochronal approach with the spectroscopic parameters from
-\citet{Fulton_2018}.  While this approach is one way of validating the
-\citet{Berger_2020a_catalog} output, the temperature scales did show a
-few-percent difference, with \citetalias{Fulton_2018} quoting higher
+stars with rotation periods.  For the remaining
+\fracstarswithprotwithoutbtwenty\ of stars that lack temperatures and
+surface gravities from \citetalias{Berger_2020a_catalog}, we adopt the
+values reported by \citet{Santos_2019} and \citet{Santos_2021}, which
+for these cases primarily derive from the \citet{Mathur_2017} DR25
+Kepler Stellar Properties Catalog, and are mostly derived from
+photometry.  In the planet sample, \frackoisnofpwithprotwithbtwenty\
+of the non-false-positive KOIs with rotation periods have parameters
+from \citet{Berger_2020a_catalog}, and the remainder similarly are
+drawn from DR25.
+
+\citet{David_2021} compared the photometric
+\citetalias{Berger_2020a_catalog} stellar parameters ($T_{\rm eff}$,
+$R_\star$, [Fe/H]) against the spectroscopic parameters from
+\citet{Fulton_2018}.  The temperature scales did show a few-percent
+systematic difference, with \citetalias{Fulton_2018} quoting higher
 temperatures than \citetalias{Berger_2020a_catalog} for mid K dwarfs,
-and lower temperatures for early F dwarfs.
-
-%For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
-%temperatures and surface gravities, we adopt the values 
-%from the Gaia DR3 General Stellar Parametrizer from Photometry (GSP-Phot; CITE).
-%GSP-Phot uses the low-resolution Gaia BP and RP spectra, mean $G$ magnitudes,
-%and parallaxes, in concert with the PARSEC isochrones (CITE Tang2014, Chen2015,
-%Pastorelli2020) to infer the stellar effective temperatures, surface gravities,
-%metallicities, and line-of-sight extinctions.
-%In the planet sample,
-%\frackoisnofpwithprotwithbtwenty\ of the
-%the non-false-positive KOIs with rotation periods have parameters from
-%\citet{Berger_2020a_catalog}.
-
-For the remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
-temperatures and surface gravities from
-\citetalias{Berger_2020a_catalog}, we adopt the values reported by
-\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
-primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
-Properties Catalog, and are mostly derived from photometry.  In the
-planet sample, \frackoisnofpwithprotwithbtwenty\ of the
-non-false-positive KOIs with rotation periods have parameters from
-\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
-from DR25.
-
-%TODO: the relevant check here is Berger20 teff vs CKS spec teff...
-%  As a sanity check on our effective temperature scale, we then used the 
-%  the CITE Green2019 reddening map, combined with the Gaia DR2 $G_{\rm
-%  BP}-G_{\rm RP}$ color, to calculate effective temperatures using the
-%  relation suggested by \citet{Curtis_2020}.
-%  This latter effective temperature scale was demonstrated to have a 
-%  scatter of $\approx$50\,K for stars near the main-sequence.
-%  We find {\bf X, Y, Z.}
-%  We ultimately favor the \citet{Berger_2020a_catalog} scale
-%  because it uses 2MASS $K_{\rm s}$ photometry, and is tied to an actual
-%  isochronal fit.
-
-%TODO: plot uncertainties in teff/logg for both parameter sets.
-
-% TODO: add that plot showing differences btwn the two teff scales?
+and lower temperatures for early F dwarfs.  Our age analysis,
+described below, assumes Gaussian uncertainties in effective
+temperatures; we by default adopt the maximum of the two-sided
+uncertainties reported by \citetalias{Berger_2020a_catalog} as our
+symmetric temperature uncertainty.  Overall, the impact of systematic
+uncertainties in the temperature scale on our age uncertainties is
+generally smaller than that of the intrinsic scatter in the
+calibration sequences of the open clusters.
+
+%TODO: plot uncertainties in teff vs teff...
+%TODO: plot uncertainties in logg vs logg...
+
+%TODO: add that plot showing differences btwn the two teff scales?
 
@@
 we compared our lithium equivalent widths
-with those reported by \citet{Berger_2018}.
+with those reported by \citet{2018ApJ...855..115B}.
 For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,

### 2024-04-20 paper/ms.tex
update ms

@@
 \label{sec:selection}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+		}
+%	
+%		\subfloat{
+%			\includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
+%		}
+%		\vspace{-0.5cm}
+		\subfloat{
+		\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+		}
+%		\subfloat{
+%			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
+%		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf The stellar samples.}  Our analysis focuses on stars observed
+    by Kepler with previously reported rotation periods (dark gray
+    points).  About half of these rotators are suitable for
+    gyrochronology (black points), based on factors including their
+    temperatures, surface gravities, (non-)binarity, and proximity to
+    the main sequence.  The full list of criteria is given in
+    Section~\ref{subsec:flags}.  A few percent of the remainder are
+    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
+    that meet an additional set of quality criteria (blue points;
+    Section~\ref{subsec:plflags}).  These are the objects for which
+    our rotation-based ages are most reliable.  We also report
+    lithium-based ages for a largely-overlapping setting of KOIs based
+    on high-resolution spectroscopy.  Surface gravities and effective
+    temperatures are primarily from
+    \citet{Berger_2020a_catalog}.
+	}
+	\label{fig:stellarprops}
+\end{figure*}
 
@@
 
-\begin{figure*}[!t]
-	\begin{center}
-		\subfloat{
-			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
-		}
-%	
-%		\subfloat{
-%			\includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
-%		}
-%		\vspace{-0.5cm}
-		\subfloat{
-		\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
-		}
-%		\subfloat{
-%			\includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
-%		}
-	\end{center}
-	\vspace{-0.5cm}
-	\caption{
-    {\bf The stellar samples.}  Our analysis focuses on stars observed
-    by Kepler with previously reported rotation periods (dark gray
-    points).  About half of these rotators are suitable for
-    gyrochronology (black points), based on factors including their
-    temperatures, surface gravities, (non-)binarity, and proximity to
-    the main sequence.  The full list of criteria is given in
-    Section~\ref{subsec:flags}.  A few percent of the remainder are
-    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
-    that meet an additional set of quality criteria (blue points;
-    Section~\ref{subsec:plflags}).  These are the objects for which
-    our rotation-based ages are most reliable.  We also report
-    lithium-based ages for a largely-overlapping setting of KOIs based
-    on high-resolution spectroscopy.  Surface gravities and effective
-    temperatures are primarily from
-    \citet{Berger_2020a_catalog}.
-	}
-	\label{fig:stellarprops}
-\end{figure*}
-
 % https://gea.esac.esa.int/archive/documentation/GDR3/Data_analysis/chap_cu8par/sec_cu8par_apsis/ssec_cu8par_apsis_gspphot.html
@@
 
+% optional todo's:
 %TODO: plot uncertainties in teff vs teff...
 %TODO: plot uncertainties in logg vs logg...
-
 %TODO: add that plot showing differences btwn the two teff scales?
@@
 We calculate gyrochrone ages using \texttt{gyro-interp}
-\citep{Bouma_2023}.  We built this statistical framework to address
+\citep{Bouma_2023}.  We designed this method to address
 the issue that, especially over the first gigayear of stellar
@@
 population-level scatter into the precision with which the age can be
-measured.  The three main assumptions behind the \texttt{gyro-interp}
-framework are as follows.  {\it 1)} The stars are on the main
+inferred.  The three main assumptions behind the \texttt{gyro-interp}
+framework are as follows.  {\it 1)} The stars are near the main
 sequence, {\it 2)} their spindown is not influenced by binary
 companions, and {\it 3)} the rotation period distribution for field
-stars of a given mass and age is identical to that of the calibration
+stars of a given mass and age is identical to that of equally aged
 open clusters (metallicity differences, for instance, are ignored).  
-Regarding the former two assumptions: pre-main-sequence and
-post-main-sequence evolution introduce variations in the stellar
-angular momentum budget.  Similarly, tidal effects might
-systematically spin up (or down) stars in binaries.
+Regarding the former two assumptions: pre-main-sequence contraction and
+post-main-sequence expansion influence surface stellar rotation periods
+in ways that we do not model.  Stars in e.g.~ close binaries might be
+subject to tidal effects that we do not treat.
 
@@
 \texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
-over 500 grid points.  \texttt{gyro-interp} itself is only calibrated
+over 500 grid points.  \texttt{gyro-interp} itself is calibrated
 out to 4\,Gyr, the adopted age of the M67 cluster
-\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the stellar
+\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the stellar
 spin-down rate as a function of time and stellar temperature, this
@@
 periods, effective temperatures, and their uncertainties by
-integrating Equation~1 of \citep{Bouma_2023} at the default grid
+integrating Equation~1 of \citet{Bouma_2023} at the default grid
 resolution.  This procedure marginalizes over the intrinsic age and
@@
 		Points represent $10^4$ draws from model probability distributions
-		that have been fitted to rotation periods 
-		\citep{Bouma_2023} and to lithium measurements \citep{Jeffries_2023}
+		that have been fitted to observed rotation periods 
+		\citep{Bouma_2023} and lithium equivalent widths \citep[EWs;][]{Jeffries_2023}
 		of stars in open clusters.
-		Lines are the fitted ``mean models'' at various ages; these models are calibrated
-		for stars younger than 4\,Gyr.
-		The intrinsic dispersion in rotation and lithium at any given age, which what these models are designed to fit, sets
+		Lines are the ``mean models'' at various ages;
+		the asymmetric intrinsic dispersion in rotation and lithium about these mean models, which is what the models fit, sets
 the theoretical precision limit for both age-dating methods.		
 		Additional sources of uncertainty, including measurement uncertainty,
-		impose additional limitations on achievable age precision.
+		impose further limits on achievable age precision.
+		These models are calibrated
+		using open clusters younger than 4\,Gyr.
 		The displayed points assume an unphysical uniform distribution in
-		temperature for visual clarity, and sizes of the points are the same in each panel.
-		%    Gyrochrone age posteriors are broadest in regions 
-		%    with the slowest stellar spin-down.
+		temperature for visual clarity.
+		The sizes of the points are the same in each panel, so that
+		low apparent density signifies greater dispersion around the mean.
 		\label{fig:models}
@@
 stars with reported rotation periods and effective temperatures.  To
-select stars for which we expect these age to in fact be valid, we
-then build a set of quality flags.  When and how these flags should be
-applied depends on the question being asked.  If the goal is to
+flag stars for which we suspect these ages may not in fact be valid,
+we then build a set of bit-quality flags.  When and how these flags should
+be applied depends on the question being asked.  If the goal is to
 construct a false-positive free sample, they might all be applied.  If
@@
 resolved binary companion \citep{Bouma_2022a}; the latter is on a
-grazing orbit \citep{2014ApJ...783...53M}.  The flags in
-Tables~\ref{tab:stars} and~\ref{tab:planets} enable selecting for or
-against such cases using bit flags.
+grazing orbit \citep{2014ApJ...783...53M}.  Our quality flags enable
+selecting for or against such cases by selecting from the following
+bits.
 
@@
 down very slowly, if at all.  Stars cooler than 3800\,K do spin down
-over gigayear timescales (e.g. {\bf CITE} E.Newton thesis, the Engle
-paper, and Chiti's article), but the currently available open cluster
-data have yet to clarify at what times, if any, the intrinsic scatter
-in the population decreases.
+over gigayear timescales
+\citep{2016ApJ...821...93N,2023ApJ...954L..50E,2024arXiv240312129C},
+but the currently available open cluster data have yet to clarify
+exactly when the intrinsic scatter in the population decreases.
 

### 2024-04-20 paper/ms.tex
update ms

@@
     that have been fitted to observed rotation periods
-		\citep{Bouma_2023} and lithium equivalent widths \citep[EWs;][]{Jeffries_2023}
-		of stars in open clusters.
-		Lines are the ``mean models'' at various ages;
-		the asymmetric intrinsic dispersion in rotation and lithium about these mean models, which is what the models fit, sets
-the theoretical precision limit for both age-dating methods.		
-		Additional sources of uncertainty, including measurement uncertainty,
-		impose further limits on achievable age precision.
-		These models are calibrated
-		using open clusters younger than 4\,Gyr.
-		The displayed points assume an unphysical uniform distribution in
-		temperature for visual clarity.
-		The sizes of the points are the same in each panel, so that
-		low apparent density signifies greater dispersion around the mean.
+    \citep{Bouma_2023} and lithium equivalent widths
+    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
+    are the ``mean models'' at various ages; the asymmetric intrinsic
+    dispersion in rotation and lithium about these mean models, which
+    is what the models fit, sets the theoretical precision limit for
+    both age-dating methods.		Additional sources of uncertainty,
+    including measurement uncertainty, impose further limits on
+    achievable age precision.  These models are calibrated using open
+    clusters younger than 4\,Gyr.  The displayed points assume an
+    unphysical uniform distribution in temperature for visual clarity.
+    The sizes of the points are the same in each panel, so that low
+    apparent density signifies greater dispersion around the mean.
 		\label{fig:models}
@@
 in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
-observed by Kepler, requiring reported rotation periods and a
-reliable DR3 crossmatch yielded
-\nuniqstarsantosrot\ stars.  Further requiring these rotating stars
-to have temperatures 3800$<$$T_{\rm eff}$/K$<$6200, which is the range
-over which \texttt{gyro-interp} returns finite age posteriors,
-yields \nuniqstarsantosrotteffcut\ stars.
-Further requiring them to be apparently single, near the main
+observed by Kepler, requiring reported rotation periods and a reliable
+DR3 crossmatch yielded \nuniqstarsantosrot\ stars.  Further requiring
+these rotating stars to have temperatures 3800$<$$T_{\rm
+eff}$/K$<$6200, which is the range over which \texttt{gyro-interp}
+returns finite age posteriors, yields \nuniqstarsantosrotteffcut\
+stars.  Further requiring them to be apparently single, near the main
 sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
 eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
-potentially amenable for gyrochronology.
-Finally, imposing RUWE$<$1.4 and requiring no possible blends ($\Delta
-G$<2.5) within 4$''$ leaves \nuniqstarsantosrotgyroappl\ stars for
-which gyrochronology is likely to be valid.
+potentially amenable for gyrochronology.  Finally, imposing RUWE$<$1.4
+and requiring no possible blends ($\Delta G$<2.5) within 4$''$ leaves
+\nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
+to be valid.
 
@@
 \nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
-cuts, save for that on RUWE.
-Then,
-require these confirmed and candidate planets to have MES$>$10.
-This combination of selection criteria
-yield \nplwgyroagewithgrazingandhighruwe\ planets with rotation-based
-ages orbiting \nplhoststarwgyroagewithgrazingandhighruwe\ stars
+cuts, save for that on RUWE.  Then, require these confirmed and
+candidate planets to have MES$>$10.  This combination of selection
+criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
+rotation-based ages orbiting
+\nplhoststarwgyroagewithgrazingandhighruwe\ stars
 (\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
@@
     {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
-    from Keck/HIRES spectra collected between 2009 and 2024.}
-    Lines are the ``mean'' isochrones from the same models as in Figure~\ref{fig:models}.
-    The logarithmic scale highlights the dynamica range of both detections
-    and upper limits.
-%    The intrinsic
-%    dispersion in the cluster data evolves in time, and is shown as
-%    the $\pm$1$\sigma$ bands~\citep{Jeffries_2023}.
-%    As for rotation-based ages, the intrisic dispersion sets the
-%    empirical precision limit for the technique.
+    from Keck/HIRES spectra collected between 2009 and 2024.} Lines
+    are the ``mean'' isochrones from the same models as in
+    Figure~\ref{fig:models}.  The logarithmic scale highlights the
+    dynamica range of both detections and upper limits.
 		\label{fig:li_vs_teff}
@@
 
-%TODO: does JUMP mirror all HIRES data from KOA?
-We queried the internal CPS archive to collect all HIRES spectra that
-were available for the \nplhoststarwgyroagewithgrazingandhighruwe\
-stars for which rotation-based ages can be derived.
-This exercise yielded at least one spectrum for \nlithiumstars\ and 
-\nlithiumplanets\ planets.
-We did not constrain whether the iodine cell was used, since
-there are no relevant lines in the region of the 
-\ion{Li}{1} 6708\,\AA\ doublet.
-These observations were acquired between 6 September 2009 and 16
-May 2024, primarily by the California Kepler Survey (CKS,
-CITE).
-In preparation for this study, we additionally collected {\bf XX}
-spectra for ``confirmed'' planets with rotation-based ages below
-1\,Gyr.
-
-To measure the lithium equivalent widths, we followed the procedure
-described by \citet{Bouma_2021}.
-The stars of interest for this work are FGK stars, and so the
-continuum in the vicinity of the
-\ion{Li}{1} 6708\,\AA\ doublet is well-defined.
-We therefore numerically integrate a single best-fit Gaussian over a
-local window, and estimate the uncertainties on the line width through
-a Monte Carlo procedure that bootstraps against the local scatter in the
-continuum.
-Generally speaking, this approach does not correct for the neighboring
-\ion{Fe}{1} 6707.44\,\AA\ blend.
+We queried the internal CPS archive to collect all blaze-corrected
+HIRES spectra that were available for the
+\nplhoststarwgyroagewithgrazingandhighruwe\ stars for which
+rotation-based ages can be derived.  This query yielded at least one
+spectrum for \nlithiumstars\ and \nlithiumplanets\ planets.  These
+spectra were acquired between 6 September 2009 and 16 May 2024.  For
+stars with multiple spectra available at varying ranges of
+signal-to-noise (S/N), we analyzed only the spectrum with the highest
+number of counts.
+
+We measured the lithium equivalent widths following a procedure from
+previous work \citep{Bouma_2021}.  The stars of interest for this work
+are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
+6708\,\AA\ doublet is well-defined.  We therefore numerically
+integrate a single best-fit Gaussian over a local window, and
+estimated uncertainties on the line width through a Monte Carlo
+procedure that bootstraps against the local scatter in the continuum.
+This approach does not correct for the neighboring \ion{Fe}{1}
+6707.44\,\AA\ blend.
 

### 2024-04-21 paper/ms.tex
update ms

@@
 %\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
-\title{Ages of Planets and Stars Observed by Kepler}
+%Ages of the Kepler Stars and Planets
+\title{Ages of Stars and Planets Observed by Kepler}
 
@@
 \label{subsec:lithiumsel}
-A final piece of our analysis involves an age assessment based on the
-equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  For this
-component of the work, we rely on spectra acquired using the High
-Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
-the Keck I 10m telescope.  These spectra were collected between 2009
-and 2024, primarily through the California Kepler Survey
+
+The final piece of our analysis involves assessing ages based on the
+equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  Specifically,
+we analyzed spectra acquired with the High Resolution Echelle
+Spectrometer (HIRES; \citealt{vogt_hires_1994}) on the Keck I 10m
+telescope.  These spectra were primarily collected through the
+California Kepler Survey
 \citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
 We supplemented the existing archive with $\approx$10~hours of new
-observations for 22 stars between Fall 2022 and Spring 2024.
-These stars were chosen to ensure that confirmed planets with
-rotational evidence for ages below 1\,Gyr received spectra, since this
-is the age range where lithium is most likely to yield useful age
-constraints.
-
-Lithium equivalent widths and abundances were already analyzed by
-\citet{2018ApJ...855..115B} for about three quarters of the spectra in
-our sample.
+observations for 22 stars between Fall 2022 and Spring 2024.  These
+stars were chosen to ensure that confirmed planets with rotational
+evidence for ages below 1\,Gyr had spectra, since this is the age
+range in which lithium is most likely to yield useful age constraints.
+
 % stat from _estimate_new_lithium_timefrac.py
-However, since new spectra have been acquired since the
-\citeauthor{2018ApJ...855..115B} analysis, and since our approach and
-selection function are different, we opt to perform our own analysis
-of the reduced HIRES spectra.
-
-%FIXME FIXME TODO LEFT OFF HERE
-%FIXME FIXME TODO LEFT OFF HERE
-%FIXME FIXME TODO LEFT OFF HERE
-These HIRES spectra are available for \nkoiswithhires\ of the
-\nkoisnofp\ non-false-positive KOIs.
-In the broader stellar sample, they are available for a much smaller 
-fraction of the stars observed by Kepler, because they were generally only
-acquired due to the presence of a transiting planet candidate.
-Given this selection function,
-we leverage these spectra in our analysis as a way to scrutinize the
-ages of just the planet sample.
+Lithium equivalent widths and abundances for the Kepler Objects of
+Interest were already analyzed by \citet{2018ApJ...855..115B} for
+about three quarters of the spectra in our sample.  However, since new
+spectra have been acquired since the \citeauthor{2018ApJ...855..115B}
+analysis, and since our approach and selection function are different,
+we opt to perform our own analysis of the reduced HIRES spectra.
+
+We queried the internal CPS archive to collect all blaze-corrected
+HIRES spectra that were available for all non-false positive Kepler
+Objects of Interest with a MES$>$10.  This query yielded at least one
+spectrum for \nlithiumstars\ stars and \nlithiumplanets\ planets.  For
+stars with multiple spectra available at varying ranges of
+signal-to-noise (S/N), we analyzed only the spectrum with the highest
+number of counts.  The resulting spectra were acquired between 6
+September 2009 and 16 May 2024, and cumulatively comprise
+\nhireshours\,hours of open-shutter time.
+
+We measured the lithium equivalent widths following a procedure
+adapted from previous work \citep{Bouma_2021}.  Our stars of interest
+are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
+6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
+to a common reference wavelength by cross-correlating against a high
+S/N template for Kepler-1698, chosen because $\gamma$$\approx$9\kms,
+$T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
+puts it in the middle of our sample's temperature range and gives it
+mild line-broadening.  We then trimmed the Doppler-corrected spectra
+to a local window centered on the lithium line, using a window width
+of 15\,\AA (we considered 10\,\AA\ and 20\,\AA; the results were
+broadly consistent).  We then continuum-normalized by fitting a
+third-degree Chebyshev series, while excluding any regions that showed
+absorption lines.  We then numerically integrated the resulting
+spectrum using a one-component Gaussian with free amplitude, width,
+and mean, and estimated uncertainties on the line width through a
+Monte Carlo procedure that bootstraped against the local scatter in
+the continuum.
+
+%    %FIXME FIXME TODO LEFT OFF HERE
+%    %FIXME FIXME TODO LEFT OFF HERE
+%    %FIXME FIXME TODO LEFT OFF HERE
+Our EW measurement approach does not correct for the neighboring
+\ion{Fe}{1} 6707.44\,\AA\ blend.
+To evaluate the accuracy and precision of our equivalent measurement
+technique, after applying an initial iteration on the
+\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
+we compared our lithium equivalent widths
+with those reported by \citet{2018ApJ...855..115B}.
+For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
+Z.}
+
+%    These HIRES spectra are available for \nkoiswithhires\ of the
+%    \nkoisnofp\ non-false-positive KOIs.
+%    In the broader stellar sample, they are available for a much smaller 
+%    fraction of the stars observed by Kepler, because they were generally only
+%    acquired due to the presence of a transiting planet candidate.
+%    Given this selection function,
+%    we leverage these spectra in our analysis as a way to scrutinize the
+%    ages of just the planet sample.
 
@@
 ``classical pulsators'' (e.g. RR~Lyrae, Cepheids) and candidate
-close-in binaries.  Visual inspection hints that this flag also
-includes many bona fide young stars, and so although we propagate it 
-to our set of quality flags, we generally ignore it.
+close-in binaries.  Visual inspection of the rotation periods as a
+function of temperature shows that although this flag selects many
+bona fide objects in these classes, it also selects the overwhelming
+majority of $\lesssim$5\,day bona fide young planets.  While we
+propagate this flag into our set of quality flags, we therefore
+generally ignore it.
 
@@
 \end{figure}
-
-We queried the internal CPS archive to collect all blaze-corrected
-HIRES spectra that were available for the
-\nplhoststarwgyroagewithgrazingandhighruwe\ stars for which
-rotation-based ages can be derived.  This query yielded at least one
-spectrum for \nlithiumstars\ and \nlithiumplanets\ planets.  These
-spectra were acquired between 6 September 2009 and 16 May 2024.  For
-stars with multiple spectra available at varying ranges of
-signal-to-noise (S/N), we analyzed only the spectrum with the highest
-number of counts.
-
-We measured the lithium equivalent widths following a procedure from
-previous work \citep{Bouma_2021}.  The stars of interest for this work
-are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
-6708\,\AA\ doublet is well-defined.  We therefore numerically
-integrate a single best-fit Gaussian over a local window, and
-estimated uncertainties on the line width through a Monte Carlo
-procedure that bootstraps against the local scatter in the continuum.
-This approach does not correct for the neighboring \ion{Fe}{1}
-6707.44\,\AA\ blend.
-
-To evaluate the accuracy and precision of our equivalent measurement
-technique, after applying an initial iteration on the
-\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
-we compared our lithium equivalent widths
-with those reported by \citet{2018ApJ...855..115B}.
-For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
-Z.}
 

### 2024-04-22 paper/ms.tex
update ms

@@
     achievable age precision.  These models are calibrated using open
-    clusters younger than 4\,Gyr.  The displayed points assume an
-    unphysical uniform distribution in temperature for visual clarity.
+    clusters younger than 4\,Gyr.  The displayed points assume a
+    uniform distribution in temperature for visual clarity.
     The sizes of the points are the same in each panel, so that low
@@
 
+(Mention Linear prior in age.)
+
 \section{Results}
@@
 	\tabletypesize{\scriptsize}
-	\tablecaption{Young planets and planet candidates.
+	\tablecaption{Ages of Kepler planets and planet candidates.
+Rows are sorted by the minimum of either the rotation or lithium-based age; this version of the table is truncated to planets younger than one billion years.
+``Confirmed'' planets appear before``candidate'' planets.
+		The full machine-readable table contains ages and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.
+		The bit quality flags for the rotation-based ages, $Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
 		%The non-truncated machine readable versions are accessible both
@@
 	\enddata
-	\tablecomments{This table includes X, Y, Z... The machine-readable version,
-		available online, includes additional columns for .... 
-	}
+%	\tablecomments{This table includes X, Y, Z... The machine-readable version,
+%		available online, includes additional columns for .... 
+%	}
 \end{deluxetable*}

### 2024-04-22 drivers/make_younghighlight_table.py

@@ -218,6 +218,9 @@ def make_table(
                             'li_eagles_limlo_forsort', 'min_age'])
 
     if not SELECT_YOUNG:
+
+        ulkvp('nnonfopkoissomeageinfo', len(pdf))
+
         csvpath = join(PAPERDIR, 'table_allageinfo.csv')
         pdf.to_csv(csvpath, index=False)
         print(f'Wrote {csvpath}')

### 2024-04-23 paper/ms.tex
update ms

@@
 % number of KOIs that are not known FPs that have a HIRES spectrum
-\newcommand{\nkoiswithhires}{{1{,}500}}
+%\newcommand{\nkoiswithhires}{{1{,}500}}
 % number of KOIs for which this work and Berger2018 overlap
-\newcommand{\nbergeroverlap}{{500}}
+%\newcommand{\nbergeroverlap}{{500}}
 
@@
 \label{sec:intro}
+
+Owing to the success of the transit and radial velocity techniques,
+exoplanet science is going through an age of abundance in which we are
+beginning to explore thousands of new worlds that orbit in close
+proximity to their host stars.
+However, most of these exoplanets are old, with ages comparable to
+that of the solar system.
+This limits our ability to understand their origins, and in particular
+to study processes such as dynamical migration (CITE), thermal cooling
+(CITE), and atmospheric loss (CITE) that typically happen soon after
+disk dispersal.
+Finding and characterizing young ($<$1\,Gyr) exoplanets is one way to
+advance our understanding of how and when these processes occur.
 
@@
 
-We measured the lithium equivalent widths following a procedure
+We measured the lithium equivalent widths using a procedure
 adapted from previous work \citep{Bouma_2021}.  Our stars of interest
@@
 of 15\,\AA (we considered 10\,\AA\ and 20\,\AA; the results were
-broadly consistent).  We then continuum-normalized by fitting a
-third-degree Chebyshev series, while excluding any regions that showed
+broadly consistent).  We continuum-normalized by fitting a
+third-degree Chebyshev series, while excluding regions with
 absorption lines.  We then numerically integrated the resulting
@@
 
-%    %FIXME FIXME TODO LEFT OFF HERE
-%    %FIXME FIXME TODO LEFT OFF HERE
-%    %FIXME FIXME TODO LEFT OFF HERE
 Our EW measurement approach does not correct for the neighboring
-\ion{Fe}{1} 6707.44\,\AA\ blend.
-To evaluate the accuracy and precision of our equivalent measurement
-technique, after applying an initial iteration on the
-\nplhoststarwgyroagewithgrazingandhighruwe\ aforementioned stars,
-we compared our lithium equivalent widths
-with those reported by \citet{2018ApJ...855..115B}.
-For the \nbergeroverlap\ stars in both samples, we found {\bf X, Y,
-Z.}
-
-%    These HIRES spectra are available for \nkoiswithhires\ of the
-%    \nkoisnofp\ non-false-positive KOIs.
-%    In the broader stellar sample, they are available for a much smaller 
-%    fraction of the stars observed by Kepler, because they were generally only
-%    acquired due to the presence of a transiting planet candidate.
-%    Given this selection function,
-%    we leverage these spectra in our analysis as a way to scrutinize the
-%    ages of just the planet sample.
+\ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
+precision of our equivalent measurement technique, after applying an
+initial iteration on the \nlithiumstars\ KOIs with spectra, we
+compared our lithium equivalent widths with those reported by
+\citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
+samples, we found broad agreement at $>$30\,m\AA, and significant
+differences at $<$30\,m\AA because \citet{2018ApJ...855..115B}
+required positive EWs, while we allow for negative ones.  At
+$>$30\,m\AA, there is however a small offset in our respective
+scales, such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is
+caused by our non-treatment of the iron blend.  We therefore directly
+subtract the constant mean value noted above when calculating
+lithium-based ages.
 
@@
 We calculate gyrochrone ages using \texttt{gyro-interp}
-\citep{Bouma_2023}.  We designed this method to address
-the issue that, especially over the first gigayear of stellar
-spin-down, stars with the same mass and same rotation period can have
-very different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The
-gyrochrone age posterior should therefore incorporate this intrinsic
-population-level scatter into the precision with which the age can be
-inferred.  The three main assumptions behind the \texttt{gyro-interp}
-framework are as follows.  {\it 1)} The stars are near the main
-sequence, {\it 2)} their spindown is not influenced by binary
-companions, and {\it 3)} the rotation period distribution for field
-stars of a given mass and age is identical to that of equally aged
-open clusters (metallicity differences, for instance, are ignored).  
-Regarding the former two assumptions: pre-main-sequence contraction and
-post-main-sequence expansion influence surface stellar rotation periods
-in ways that we do not model.  Stars in e.g.~ close binaries might be
-subject to tidal effects that we do not treat.
+\citep{Bouma_2023}, which is a method designed to address the issue
+that, especially over the first gigayear of stellar spin-down, stars
+with the same mass and same rotation period can have very different
+ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
+posterior should therefore incorporate this intrinsic population-level
+scatter into the precision with which the age can be inferred.
+Figure~\ref{fig:models} exemplifies this problem, particularly in
+regions where the 0.1-1\,Gyr stars overlap.
+
+The three main assumptions required for a rotation-based age dating
+method like \texttt{gyro-interp} to be valid are as follows.  {\it 1)}
+The target star must be near the main sequence, {\it 2)} its spindown
+must not be influenced by binary companions, and {\it 3)} the rotation
+period distribution for field stars of a given mass and age must be
+identical to that of equally aged open clusters (metallicity
+differences, for instance, are ignored).  Regarding the former two
+assumptions: pre-main-sequence contraction and post-main-sequence
+expansion influence surface stellar rotation periods in ways that we
+do not model.  Stars in e.g.~ close binaries might be subject to tidal
+effects that we do not treat.
 
@@
 \citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the stellar
-spin-down rate as a function of time and stellar temperature, this
+spin-down rate as a function of time and stellar temperature, the
 framework interpolates between reference open clusters using smooth
@@
 \label{subsec:flags}
-We calculated gyrochrone age posteriors for all \nuniqstarsantosrot\
-stars with reported rotation periods and effective temperatures.  To
-flag stars for which we suspect these ages may not in fact be valid,
-we then build a set of bit-quality flags.  When and how these flags should
-be applied depends on the question being asked.  If the goal is to
-construct a false-positive free sample, they might all be applied.  If
-the goal is constructing a complete sample of young stars, then
-consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
-Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
-resolved binary companion \citep{Bouma_2022a}; the latter is on a
-grazing orbit \citep{2014ApJ...783...53M}.  Our quality flags enable
-selecting for or against such cases by selecting from the following
-bits.
+We calculated gyrochrone age posteriors for all
+\nuniqstarsantosrotteffcut\ stars with reported rotation periods and
+effective temperatures between 3800\,K and 6200\,K.  To flag stars for
+which we suspect these ages may not in fact be valid, we built a set
+of bit-quality flags.  When and how these flags should be applied
+depends on the question being asked.  If the goal is to construct a
+false-positive free sample, they might all be applied.  If the goal is
+to construct a complete sample of young stars, then consider the
+examples of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
+($\approx$625\,Myr).  The former has a high RUWE due to a resolved
+binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
+\citep{2014ApJ...783...53M}.  Our gyrochronology quality flag, $Q_{\rm
+gyro}$, enables selecting for or against such cases by selecting from
+the following bits.
 
@@
 cross-matched against \citet{Berger_2020a_catalog}.
-%\citet{berger_2018_radii_evolnstates} performed this classification
-%using the PARSEC evolutionary models, stellar radii, and DR25 stellar
-%properties catalog effective temperatures \citep{Mathur_2017}.
 
@@
 gyrochronology.  By default, we assume that a star is suitable for
-gyrochronology if none of bits zero through nine are raised.
+gyrochronology if none of bits zero through nine (inclusive) are
+raised.
 
@@
 For the subset of our analysis that focuses on the properties of
-planets, we use the following additional quality flags.
+planets, we use the following additional quality indicators to assess 
+the quality of a planet, $Q_{\rm planet}$.
 
@@
 % disposition score.
-{\it Candidate reliability}---no false positives; only ``planet
-candidates'' and ``confirmed planets''.
-
-{\it Candidate S/N}---the false positive rate increases greatly toward
-Kepler's noise floor for planet detection (CITE).  We required a S/N
-in excess of the usual floor, through a cut on the ``multiple event
-statistic'' (MES): we required ${\rm MES}$$>$10.
-
-{\it Grazing planets}---grazing objects ($b$>0.8) often provide
-non-robust planetary sizes due to the size-impact parameter
+{\it Candidate reliability (bit 0)}---no false positives; only
+``planet candidates'' and ``confirmed planets''.
+
+{\it Candidate S/N (bit 1)}---the false positive rate increases
+greatly toward Kepler's noise floor for planet detection (CITE).  We
+required a S/N in excess of the usual floor, through a cut on the
+``multiple event statistic'' (MES): we required ${\rm MES}$$>$10.
+
+{\it Grazing planets (bit 2)}---grazing objects ($b$>0.8) often
+provide non-robust planetary sizes due to the size-impact parameter
 degeneracy.  Particularly for large planets, they can also include
@@
 degeneracy).  While we include a flag for such objects, we by default
-do not exclude them from our analysis; this is because...{\bf justify}.
+do not exclude them from our analysis; this is because...{\bf
+justify}.
 
@@
 	\caption{
-    {\bf Rotation periods for Kepler stars (top) and planets (bottom)}.
-    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
-    (\citetalias{Santos_2019} and \citetalias{Santos_2021}).  The
-    gray lines are ``mean fits'' from Figure~\ref{fig:models} to the slow
-    rotation sequences of M67, Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open clusters.
-    The stellar sample shows
-    only apparently single stars near the main sequence with $\log
-    g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm eff}$/K$<$6200, that are not
-    flagged as eclipsing binaries.  The planet sample has the same
-    stellar cuts, and shows confirmed and candidate planets from the
-    cumulative KOI table, requiring them to have a MES$>$10.
+    {\bf Rotation periods for Kepler stars (top) and planets
+    (bottom)}.  The periods are from \citet{Santos_2019} and
+    \citet{Santos_2021} (\citetalias{Santos_2019} and
+    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
+    Figure~\ref{fig:models} to the slow rotation sequences of M67,
+    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
+    clusters.  The stellar sample shows only apparently single stars
+    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
+    3800$<$$T_{\rm eff}$/K$<$6200, that are not flagged as eclipsing
+    binaries.  The planet sample has the same stellar cuts, and shows
+    confirmed and candidate planets from the cumulative KOI table,
+    requiring them to have a MES$>$10.
 	}
@@
 		\leavevmode
-		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf}
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
 	\end{center}
@@
     {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
-    from Keck/HIRES spectra collected between 2009 and 2024.} Lines
-    are the ``mean'' isochrones from the same models as in
-    Figure~\ref{fig:models}.  The logarithmic scale highlights the
-    dynamica range of both detections and upper limits.
+    for planet-hosting stars.} These measurements were made from
+    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the
+    ``mean'' isochrones from the same models as in
+    Figure~\ref{fig:models}; the intrinsic dispersion beyond
+    $\approx$1\,Gyr becomes much larger than changes in the mean.  The
+    logarithmic scale highlights the dynamic range of the detections
+    and upper limits.  Some stars with lithium detections do not have
+    detected rotation periods, and vice-versa.
 		\label{fig:li_vs_teff}
@@
 
-To calculate lithium ages, we use a combination of \texttt{EAGLES}
-\citep{Jeffries_2023} and \texttt{BAFFLES}
-\citep{Stanford-Moore_2020}.
-The former tool was calibrated on stars observed by the Gaia-ESO
-spectroscopic survey in 52 open clusters, with ages spanning 2 to
-6{,}000\,Myr.  
-Example \texttt{EAGLES} isochrones, as well as the modelled intrinsic dispersion
-about those isochrones, are shown in Figure~\ref{fig:li_vs_teff}.
-While our preference would have been to exclusively use
-\texttt{EAGLES} due to its homogeneous analysis approach ...
-
-(Mention Linear prior in age.)
+Figure~\ref{fig:li_vs_teff} shows our measured lithium equivalent
+widths, plotted over the mean isochrone models from
+\texttt{EAGLES} \citep{Jeffries_2023}.  We declare an upper-limit for
+plotting purposes if ${EW}-\sigma_{\rm EW}$$<$10\,m\AA.  However, our
+EWs span -40 to 250\,m\AA; all negative EWs have uncertainties that
+are statistically consistent with zero.
+
+To calculate lithium ages, we use \texttt{EAGLES} (git commit
+\texttt{ac09637}), which uses a similar method as
+\texttt{gyro-interp}, and was calibrated on stars observed by the
+Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
+to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
+age, spanning $10^6$ to $10^{10.1}$\,yr, and symmetrize our EW
+uncertainties for purposes of interfacing with \texttt{EAGLES} by
+taking the maximum of the measured positive and negative 1$\sigma$
+uncertainty intervals.
+Certain quality flags described in Section~\ref{subsec:flags} are also
+relevant for lithium ages, specifically those associated with
+crowding and binarity since those effects can produce both systematic
+and astrophysical biases in EW measurements.
 
@@
 The stellar and planetary properties of this latter sample of young
-planets are given in Table~\ref{tab:youngplanets}.
+planets are given in Table~\ref{tab:planets}.
 
@@
 	\begin{center}
-		\subfloat{
-			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
-		}
-
-		\subfloat{
+%		\subfloat{
+%			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
+%		}
+%		
+%		\subfloat{
 			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
-		}
+%		}
 	\end{center}
@@
 		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
-		Two possible selection functions for what defines a young, sub-gigayear, planet
-		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
-		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
-		Circles are previously reported measurements from the literature.
+		Planets highlighted in this plot have $t$<1\,Gyr at $2$$\sigma$.
+%		
+%		Two possible selection functions for what defines a young, sub-gigayear, planet
+%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
+%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
+%		Circles are previously reported measurements from the literature.
 		Stars are newly calculated ages for Kepler planets from this work; circles
-		are planets from the literature meeting the same requirements.
+		are planets from the literature meeting the same requirement.
 		Gray points are transiting planets from the NASA Exoplanet Archive older than
@@
 \startlongtable
-\begin{deluxetable*}{lllllllrrrrrr}
+\begin{deluxetable*}{lllllllrrrrrrl}
 	%\tabletypesize{\footnotesize}
 	\tabletypesize{\scriptsize}
-	\tablecaption{Ages of Kepler planets and planet candidates.
-Rows are sorted by the minimum of either the rotation or lithium-based age; this version of the table is truncated to planets younger than one billion years.
-``Confirmed'' planets appear before``candidate'' planets.
-		The full machine-readable table contains ages and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.
-		The bit quality flags for the rotation-based ages, $Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
-		%The non-truncated machine readable versions are accessible both
-		%through the online journal, and at
-		%\url{https://zenodo.org/record/8327508}.
-		\label{tab:youngplanets}}
+  \tablecaption{Ages of Kepler planets and planet candidates.  Rows
+  are sorted by the minimum of either the rotation or lithium-based
+  age, and ``confirmed'' planets appear before``candidate'' planets.
+  This version of the table is truncated to planets younger than
+  one billion years.  The full machine-readable table contains ages
+  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
+  with MES$>$10.  The bit quality flags for the rotation-based ages,
+  $Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
+  Concisely summarized, they are:
+  Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?
+  Bit 1: $\log g$$<$4.2?
+  Bit 2: $M_{\rm G}$<3.9?
+  Bit 3: In KEBC?
+  Bit 4: Large $d_{\rm xm,Kep-Gaia}$?
+  Bit 5: Confused Kep-Gaia crossmatch?
+  Bit 6: Gaia DR3 non-single star?
+  Bit 7: RUWE$>$1.4?
+  Bit 8: Crowded?
+  Bit 9: Far from main-sequence?
+  Bit 10: \citetalias{Santos_2021} CP/CB?
+  As an example, Kepler-1627 is flagged with bit 10 and bit 7.
+  \label{tab:planets}}
 	%\toprule
@@
 	\startdata
-  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm RUWE}$ & $f_{\rm grazing}$ & $Q_{\rm gyro}$ &  Spec? \\
-  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &      bool &   bool  & int & bool \\
+  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm grazing}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
+  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       bool  & int & bool & -- \\
 	\hline
@@
 \end{deluxetable*}
-%\endlongtable
+
+\startlongtable
+\begin{deluxetable*}{lllllllrrrrrr}
+	%\tabletypesize{\footnotesize}
+	\tabletypesize{\scriptsize}
+  \tablecaption{Ages of Kepler target stars.  
+  The full machine-readable table contains rotation-based ages
+  for \nuniqstarsantosrotteffcut\ stars with temperatures spanning
+  3800-6200\,K.
+  \nuniqstarsantosrotgyroappl\ of these stars have quality flags
+  without bits 0 to 9 raised, meaning that they are likely valid
+  objects for gyrochronology (see Section~\ref{subsec:flags}).
+  \label{tab:stars}}
+	%\toprule
+	%\midrule
+	%\endhead
+	\startdata
+  KIC & Gaia &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm gyro}$ \\
+  -- &   -- & K & days &  Myr &    int  \\
+	\hline
+  %FIXME make it!
+	%\input{table_gyrostars} %\\
+	%\hline
+	\enddata
+%	\tablecomments{This table includes X, Y, Z... The machine-readable version,
+%		available online, includes additional columns for .... 
+%	}
+\end{deluxetable*}
 

### 2024-04-23 paper/collect.sh
update ms: propagate tables vals

@@ -23,6 +23,7 @@ cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_a
 
 # plot_li_vs_teff
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_JUMP_eagles_logy.pdf .
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
 
 # planets

### 2024-04-23 paper/ms.tex
update ms

@@
 
-Owing to the success of the transit and radial velocity techniques,
-exoplanet science is going through an age of abundance in which we are
-beginning to explore thousands of new worlds that orbit in close
-proximity to their host stars.
-However, most of these exoplanets are old, with ages comparable to
-that of the solar system.
-This limits our ability to understand their origins, and in particular
-to study processes such as dynamical migration (CITE), thermal cooling
-(CITE), and atmospheric loss (CITE) that typically happen soon after
-disk dispersal.
-Finding and characterizing young ($<$1\,Gyr) exoplanets is one way to
-advance our understanding of how and when these processes occur.
+Exoplanet science is in an age of abundance, fueled by the discovery
+of thousands of worlds orbiting close to their host stars (CITE, CITE,
+CITE).
+However, most known exoplanets have ages comparable to that of the solar
+system.
+This unfortunate fact leaves many gaps in our knowledge of
+exoplanetary origins, particularly regarding to degree to which
+dynamical migration (CITE), thermal cooling (CITE), and atmospheric
+loss (CITE) sculpt planetary systems after the disk disperses.
+Finding and characterizing young ($<$1\,Gyr) exoplanets offers one
+possible workaround, since these objects can show both individual (CITE,
+CITE) and population-level (CITE) evidence for these types of
+evolutionary processes.
 
@@
 \citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
-and therefore measurements of the \ion{Li}{1} 6708\,\AA\ doublet, in
-general have only been acquired for the Kepler objects of interest
+which can yield measurements of the \ion{Li}{1} 6708\,\AA\ doublet, 
+have only been acquired for the Kepler objects of interest
 (KOIs), which comprise a few percent of Kepler's targets.
@@
 to a local window centered on the lithium line, using a window width
-of 15\,\AA (we considered 10\,\AA\ and 20\,\AA; the results were
+of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
 broadly consistent).  We continuum-normalized by fitting a
@@
 scatter into the precision with which the age can be inferred.
-Figure~\ref{fig:models} exemplifies this problem, particularly in
+Figure~\ref{fig:models} highlights this problem, particularly in
 regions where the 0.1-1\,Gyr stars overlap.
@@
 
-\subsection{Age distribution}
-
 \begin{figure*}[!t]
@@
 
+\subsection{Age distribution}
+
 Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
@@
 		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
-		Planets highlighted in this plot have $t$<1\,Gyr at $2$$\sigma$.
+		Planets highlighted in this plot have ages $t$<1\,Gyr at $2$$\sigma$.
 %		
@@
 %		Circles are previously reported measurements from the literature.
-		Stars are newly calculated ages for Kepler planets from this work; circles
+		Stars denote systems that are newly characterized in this work; circles
 		are planets from the literature meeting the same requirement.
@@
 		1\,Gyr.
-		The three age bins span 10-46\,Myr, 46-215\,Myr, and 215-1000\,Myr.
+		The selection function matters:
+		if we were to require $t$<1\,Gyr and $t/\sigma_{\rm t}>3$, we would loose $\approx$ten
+		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
 		\label{fig:rp_period_age_results}

### 2024-04-23 paper/ms.tex
update ms

@@
 %%%%%%%%%%%
-% number of KOIs in Q1-Q17 DR25 table
-%\newcommand{\nkois}{{$\sim$8{,}000}} %FIXME precise
-% number of KOIs in Q1-Q17 DR25 table that are not known false positives
-%\newcommand{\nkoisnofp}{{$\sim$4{,}000}} %FIXME precise
-% number of KOIs in Q1-Q17 DR25 table that have rotation periods
 \newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
@@
 \newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}
-
-%%%%%%%%%%%
-% LITHIUM %
-%%%%%%%%%%%
-% number of KOIs that are not known FPs that have a HIRES spectrum
-%\newcommand{\nkoiswithhires}{{1{,}500}}
-% number of KOIs for which this work and Berger2018 overlap
-%\newcommand{\nbergeroverlap}{{500}}
 
@@
 %Ages of the Kepler Stars and Planets
-\title{Ages of Stars and Planets Observed by Kepler}
+\title{Ages of the Kepler Stars and Planets}
 
@@
   gyrochronology is probably valid.
+  %
+  The independent lithium ages show X,Y,Z, though they are subject to their own systematic issues.
   %
@@
 candidate as of 6 June 2023, while also incorporating human-based
-vetting \footnote{See
-\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}.
+vetting.\footnote{See
+\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
 These planets represent a superset of those in the fully automated
@@
 	\caption{
-		{\bf Age distribution of stars (left) and known planet hosts
+		{\bf Age distribution of stars (left) and planet hosts
 			(right) in the Kepler field inferred from rotation periods}.
-		A histogram of samples from the age posteriors for all Kepler
-		stars with reported rotation periods and temperatures is shown in
-		light gray;
-		the age calculation was actually valid for stars that meet the criteria
-		discussed in Section~\ref{subsec:flags} (dark blue).
+		The
+		light gray histogram does not impose any requires on factors such as
+		binarity, crowding, or the star's evolutionary state;
+		the age calculation was more likely to be valid for stars that meet the criteria
+		discussed in Section~\ref{subsec:flags} (black).
 		\label{fig:hist_tgyro}
@@
 %\end{figure*}
-
-\begin{figure*}[!t]
-	\begin{center}
-%		\subfloat{
-%			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
-%		}
-%		
-%		\subfloat{
-			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
-%		}
-	\end{center}
-	\vspace{-0.5cm}
-
-	\caption{
-		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
-		Planets highlighted in this plot have ages $t$<1\,Gyr at $2$$\sigma$.
-%		
-%		Two possible selection functions for what defines a young, sub-gigayear, planet
-%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
-%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
-%		Circles are previously reported measurements from the literature.
-		Stars denote systems that are newly characterized in this work; circles
-		are planets from the literature meeting the same requirement.
-		Gray points are transiting planets from the NASA Exoplanet Archive older than
-		1\,Gyr.
-		The selection function matters:
-		if we were to require $t$<1\,Gyr and $t/\sigma_{\rm t}>3$, we would loose $\approx$ten
-		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
-		\label{fig:rp_period_age_results}
-	}
-\end{figure*}
 
@@
 
+\begin{figure*}[!t]
+	\begin{center}
+		%		\subfloat{
+		%			\includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
+		%		}
+		%		
+		%		\subfloat{
+		\includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+		%		}
+	\end{center}
+	\vspace{-0.5cm}
+
+	\caption{
+		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
+		%		
+		%		Two possible selection functions for what defines a young, sub-gigayear, planet
+		%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
+		%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
+		%		Circles are previously reported measurements from the literature.
+		Stars denote \nplyounggyrotwosigmanograzingnoruwe\ systems that are newly characterized in this work to have
+		$t_{\rm gyro}$<1\,Gyr at $2$$\sigma$; circles
+		are planets from the literature meeting the same age requirement, but with a heterogeneous
+		set of age provenances.
+		Gray points are transiting planets from the NASA Exoplanet Archive older than
+		1\,Gyr.
+		The selection function matters:
+		selecting for precise ages ($t/\sigma_{\rm t}>3$) would hide $\approx$ten
+		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
+		For this plot, we require our Kepler stellar hosts to have no quality flags raised,
+		and for the planets to similarly no have quality flags raised. 
+		Weakening these constraints would expand the sample at the expense of reliability.
+		\label{fig:rp_period_age_results}
+	}
+\end{figure*}
+
 \subsubsection{Systems with Rotation and Lithium in Agreement}
@@
 	\tabletypesize{\scriptsize}
-  \tablecaption{Ages of Kepler planets and planet candidates.  Rows
+  \tablecaption{Ages of Kepler planets and planet candidates. 
+  This version of the table is truncated to include the youngest fifty confirmed Kepler planets, and the twenty youngest candidates.
+  The full machine-readable table contains ages
+  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
+  with MES$>$10.  
+  \label{tab:planets}}
+	%\toprule
+	%\midrule
+	%\endhead
+	\startdata
+  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
+  $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ &
+  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
+  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       int  & int & bool & -- \\
+	\hline
+	\input{table_subgyr_confirmed} %\\
+%	\rule
+	\input{table_subgyr_candidate}
+	\enddata
+	\tablecomments{Rows
 		are sorted by the minimum of either the rotation or lithium-based
 		age, and ``confirmed'' planets appear before``candidate'' planets.
-  This version of the table is truncated to planets younger than
-  one billion years.  The full machine-readable table contains ages
-  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
-  with MES$>$10.  The bit quality flags for the rotation-based ages,
+		The bit quality flags for the rotation-based ages,
 		$Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
@@
 		Bit 10: \citetalias{Santos_2021} CP/CB?
-  As an example, Kepler-1627 is flagged with bit 10 and bit 7.
-  \label{tab:planets}}
-	%\toprule
-	%\midrule
-	%\endhead
-	\startdata
-  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ & $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ & $f_{\rm grazing}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
-  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       bool  & int & bool & -- \\
-	\hline
-	\input{table_subgyr_confirmed} %\\
-	%\hline
-	\input{table_subgyr_candidate}
-	\enddata
-%	\tablecomments{This table includes X, Y, Z... The machine-readable version,
-%		available online, includes additional columns for .... 
-%	}
+		As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
+		The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
+		following meaning.
+		Bit 0: Candidate not reliable?
+		Bit 1: MES$<$10?
+		Bit 2: Grazing?
+	}
 \end{deluxetable*}

### 2024-04-24 paper/ms.tex
update ms

@@
 \begin{abstract}
-  Recent studies of stars in open clusters have yielded detailed
-  pictures of how the rotation rates and lithium content of FGK stars
-  evolve during the first few billion years of their lives.
+  Recent studies of stars in open clusters have shown how the rotation
+  rates and lithium content of FGK stars evolve during the first few
+  billion years of their lives.
   %
   Here we apply this knowledge to planets and stars observed by the
-  primary Kepler mission, and report the resulting rotation and
+  primary Kepler mission, and calculate the implied rotation and
   lithium-based ages.
@@
   %
-  The resulting age distribution bounds the relative balance of
-  ``young'' and ``old'' planets detectable not only by Kepler, but
-  also by future wide-field transit surveys.
+  This parent stellar age distribution bounds the relative balance of
+  young and old planets detectable not only by Kepler, but also by
+  future wide-field transit surveys.
   %
@@
 Rotation will provide the most age information for the largest number
-of stars.  The lithium technique will provide useful information for
+of stars.  Lithium will provide useful information for
 stars younger than a few hundred million years.
@@
 	\caption{
-		{\bf Sizes, orbital periods, and rotation-based ages of transiting exoplanets younger than one billion years.} 
-		%		
-		%		Two possible selection functions for what defines a young, sub-gigayear, planet
-		%		are imposed on the {\it top} and {\it bottom}: the former requires $t$<1\,Gyr with
-		%		$t/\sigma_{\rm t}$>3, the latter also allows planets younger than 1\,Gyr at 2$\sigma$.
-		%		Circles are previously reported measurements from the literature.
-		Stars denote \nplyounggyrotwosigmanograzingnoruwe\ systems that are newly characterized in this work to have
-		$t_{\rm gyro}$<1\,Gyr at $2$$\sigma$; circles
-		are planets from the literature meeting the same age requirement, but with a heterogeneous
-		set of age provenances.
-		Gray points are transiting planets from the NASA Exoplanet Archive older than
-		1\,Gyr.
-		The selection function matters:
-		selecting for precise ages ($t/\sigma_{\rm t}>3$) would hide $\approx$ten
-		$<$0.5\,Gyr planets, and add $\approx$twenty 0.5-1\,Gyr planets.
-		For this plot, we require our Kepler stellar hosts to have no quality flags raised,
-		and for the planets to similarly no have quality flags raised. 
-		Weakening these constraints would expand the sample at the expense of reliability.
+    {\bf Sizes, orbital periods, and rotation-based ages of transiting
+    exoplanets younger than one billion years.} Stars denote
+    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
+    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
+    $2$$\sigma$; circles are planets from the literature meeting the
+    same age requirement, but with a heterogeneous set of age
+    provenances.  Gray points are transiting planets from the NASA
+    Exoplanet Archive older than 1\,Gyr.  The selection function
+    matters: selecting for precise ages ($t/\sigma_{\rm t}>3$) would
+    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$twenty
+    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
+    hosts to have no quality flags raised, and for the planets to
+    similarly no have quality flags raised.  Weakening these
+    constraints would expand the sample at the expense of reliability.
 		\label{fig:rp_period_age_results}

### 2024-04-25 paper/ms.tex
update ms

@@
 \end{figure*}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
+		}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf The models.}
+		Points represent $10^4$ draws from model probability distributions
+		that have been fitted to observed rotation periods
+		\citep{Bouma_2023} and lithium equivalent widths
+		\citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
+		are the ``mean models'' at various ages; the asymmetric intrinsic
+		dispersion in rotation and lithium about these mean models, which
+		is what the models fit, sets the theoretical precision limit for
+		both age-dating methods.		Additional sources of uncertainty,
+		including measurement uncertainty, impose further limits on
+		achievable age precision.  These models are calibrated using open
+		clusters younger than 4\,Gyr.  The displayed points assume a
+		uniform distribution in temperature for visual clarity.
+		The sizes of the points are the same in each panel, so that low
+		apparent density signifies greater dispersion around the mean.
+		\label{fig:models}
+	}
+\end{figure*}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+		}
+		%	
+		%		\vspace{-0.5cm}
+		\subfloat{
+			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+		{\bf Rotation periods for Kepler stars (top) and planets
+			(bottom)}.  The periods are from \citet{Santos_2019} and
+		\citet{Santos_2021} (\citetalias{Santos_2019} and
+		\citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
+		Figure~\ref{fig:models} to the slow rotation sequences of M67,
+		Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
+		clusters.  The stellar sample shows only apparently single stars
+		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
+		3800$<$$T_{\rm eff}$/K$<$6200, that are not flagged as eclipsing
+		binaries.  The planet sample has the same stellar cuts, and shows
+		confirmed and candidate planets from the cumulative KOI table,
+		requiring them to have a MES$>$10.
+	}
+	\label{fig:prot_vs_teff}
+\end{figure*}
+
+\begin{figure}[!t]
+	\begin{center}
+		\leavevmode
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
+			for planet-hosting stars.} These measurements were made from
+		Keck/HIRES spectra collected between 2009 and 2024.  Lines are the
+		``mean'' isochrones from the same models as in
+		Figure~\ref{fig:models}; the intrinsic dispersion beyond
+		$\approx$1\,Gyr becomes much larger than changes in the mean.  The
+		logarithmic scale highlights the dynamic range of the detections
+		and upper limits.  Some stars with lithium detections do not have
+		detected rotation periods, and vice-versa.
+		\label{fig:li_vs_teff}
+	}
+\end{figure}
 
@@
 
-\begin{figure*}[!t]
-	\begin{center}
-		\leavevmode
-		\subfloat{
-			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
-		}
-		\subfloat{
-			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
-		}
-	\end{center}
-	\vspace{-0.6cm}
-	\caption{
-		{\bf The models.}
-    Points represent $10^4$ draws from model probability distributions
-    that have been fitted to observed rotation periods
-    \citep{Bouma_2023} and lithium equivalent widths
-    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
-    are the ``mean models'' at various ages; the asymmetric intrinsic
-    dispersion in rotation and lithium about these mean models, which
-    is what the models fit, sets the theoretical precision limit for
-    both age-dating methods.		Additional sources of uncertainty,
-    including measurement uncertainty, impose further limits on
-    achievable age precision.  These models are calibrated using open
-    clusters younger than 4\,Gyr.  The displayed points assume a
-    uniform distribution in temperature for visual clarity.
-    The sizes of the points are the same in each panel, so that low
-    apparent density signifies greater dispersion around the mean.
-		\label{fig:models}
-	}
-\end{figure*}
-
 \subsection{Gyrochronology Quality Flags}
@@
 unit weight error
-(RUWE\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
+(RUWE) %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
 exceeding 1.4 as possible binaries.  Such astrometric outliers can be
@@
 
-\begin{figure}[!t]
-	\begin{center}
-		\subfloat{
-			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
-		}
-
-		\vspace{-0.5cm}
-		\subfloat{
-			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
-		}
-	\end{center}
-	\vspace{-0.5cm}
-	\caption{
-    {\bf Rotation periods for Kepler stars (top) and planets
-    (bottom)}.  The periods are from \citet{Santos_2019} and
-    \citet{Santos_2021} (\citetalias{Santos_2019} and
-    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
-    Figure~\ref{fig:models} to the slow rotation sequences of M67,
-    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
-    clusters.  The stellar sample shows only apparently single stars
-    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-    3800$<$$T_{\rm eff}$/K$<$6200, that are not flagged as eclipsing
-    binaries.  The planet sample has the same stellar cuts, and shows
-    confirmed and candidate planets from the cumulative KOI table,
-    requiring them to have a MES$>$10.
-	}
-	\label{fig:prot_vs_teff}
-\end{figure}
-
 The rotation periods for the stellar and planetary samples are shown
@@
 \label{sec:liage}
-
-\begin{figure}[!t]
-	\begin{center}
-		\leavevmode
-		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
-	\end{center}
-	\vspace{-0.6cm}
-	\caption{
-    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
-    for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the
-    ``mean'' isochrones from the same models as in
-    Figure~\ref{fig:models}; the intrinsic dispersion beyond
-    $\approx$1\,Gyr becomes much larger than changes in the mean.  The
-    logarithmic scale highlights the dynamic range of the detections
-    and upper limits.  Some stars with lithium detections do not have
-    detected rotation periods, and vice-versa.
-		\label{fig:li_vs_teff}
-	}
-\end{figure}
 
@@
 	\tabletypesize{\scriptsize}
-  \tablecaption{Ages of Kepler planets and planet candidates. 
-  This version of the table is truncated to include the youngest fifty confirmed Kepler planets, and the twenty youngest candidates.
-  The full machine-readable table contains ages
-  and age limits for \nnonfopkoissomeageinfo\ non-false positive KOIs
-  with MES$>$10.  
+  \tablecaption{Ages of Kepler planets and planet candidates.  This
+  version of the table is truncated to include the youngest fifty
+  confirmed Kepler planets, and the twenty youngest candidates.  The
+  full machine-readable table contains ages and age limits for
+  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
   \label{tab:planets}}
@@
 \begin{deluxetable*}{lllllllrrrrrr}
-	%\tabletypesize{\footnotesize}
 	\tabletypesize{\scriptsize}
-  \tablecaption{Ages of Kepler target stars.  
-  The full machine-readable table contains rotation-based ages
-  for \nuniqstarsantosrotteffcut\ stars with temperatures spanning
-  3800-6200\,K.
-  \nuniqstarsantosrotgyroappl\ of these stars have quality flags
-  without bits 0 to 9 raised, meaning that they are likely valid
-  objects for gyrochronology (see Section~\ref{subsec:flags}).
-  \label{tab:stars}}
+  \tablecaption{
+  Ages of Kepler target stars derived from rotation periods.  The
+  full machine-readable table includes \nuniqstarsantosrot\ stars
+  with reported rotation periods from \citetalias{Santos_2019} and
+  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
+  finite reported gyrochrone ages.  A random set of these stars is
+  shown here to give an idea for form and content.  The quality
+  bit-flag is as in Table~\ref{tab:planets}; requiring bits zero to
+  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
+  which gyrochronology is likely to be valid.  \label{tab:stars}}
 	%\toprule
@@
 	\startdata
-  KIC & Gaia &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm gyro}$ \\
+  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm gyro}$ \\
   -- &   -- & K & days &  Myr &    int  \\
 	\hline
-  %FIXME make it!
-	%\input{table_gyrostars} %\\
+	\input{table_star_gyro} %\\
 	%\hline

### 2024-04-25 paper/collect.sh

@@ -28,9 +28,6 @@ cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
 
 # planets
 r="/Users/luke/Dropbox/proj/NEA_age_plots/results/rp_vs_period_scatter"
-cp $r/rp_vs_period_scatter_20240415_colorbyage_strictyoung.pdf .
-cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf .
-cp $r/rp_vs_period_scatter_20240415_colorbyage_anyyoung.pdf .
 cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf .
 
 

### 2024-04-25 drivers/make_star_table.py
add make_star_table (ie table2)

@@ -1,0 +1,159 @@
+"""
+Trim down the "Santos19_Santos21_dquality" gyro table, which has a long list of
+columns, to something CSV and TeX manageable.
+
+Available cols:
+
+    ['gyro_median', 'gyro_peak', 'gyro_mean', 'gyro_+1sigma', 'gyro_-1sigma',
+    'gyro_+2sigma', 'gyro_-2sigma', 'gyro_+3sigma', 'gyro_-3sigma',
+    'gyro_+1sigmapct', 'gyro_-1sigmapct', 'KIC', 'Kpmag', 'Q', 'Teff', 'E_Teff',
+    'e_Teff', 'logg', 'E_logg', 'e_logg', 'Mass', 'E_Mass', 'e_Mass', 'Prot',
+    'E_Prot', 'Sph', 'E_Sph', 'Provenance', 's19_flag1', 's19_flag2', 's19_flag3',
+    's19_flag4', 's19_flag5', 's21_flag1', 's21_flag2', 's21_flag3', 's21_flag4',
+    's21_flag5', 'b20t1_recno', 'b20t1_KIC', 'b20t1_gmag', 'b20t1_e_gmag',
+    'b20t1_Ksmag', 'b20t1_e_Ksmag', 'b20t1_plx', 'b20t1_e_plx', 'b20t1___Fe_H_',
+    'b20t1_e__Fe_H_', 'b20t1_RUWE', 'b20t1_Ncomp', 'b20t1_KsCorr', 'b20t1_State',
+    'b20t1_output', 'b20t1_KSPC', 'b20t1__RA', 'b20t1__DE', 'b20t2_recno',
+    'b20t2_KIC', 'b20t2_Mass', 'b20t2_E_Mass', 'b20t2_e_Mass', 'b20t2_Teff',
+    'b20t2_E_Teff', 'b20t2_e_Teff', 'b20t2_logg', 'b20t2_E_logg', 'b20t2_e_logg',
+    'b20t2___Fe_H_', 'b20t2_E__Fe_H_', 'b20t2_e__Fe_H_', 'b20t2_Rad',
+    'b20t2_E_Rad', 'b20t2_e_Rad', 'b20t2_rho', 'b20t2_E_rho', 'b20t2_e_rho',
+    'b20t2_Lum', 'b20t2_E_Lum', 'b20t2_e_Lum', 'b20t2_Age', 'b20t2_f_Age',
[…]

### 2024-04-25 paper/ms.tex
update ms

@@
 %Ages of the Kepler Stars and Planets
-\title{Ages of the Kepler Stars and Planets}
+\title{A Dearth of Young Stars and Planets in the Kepler Field}
 
@@
   %
-  The independent lithium ages show X,Y,Z, though they are subject to their own systematic issues.
+  The independent lithium ages show X,Y,Z, though they are subject to
+  their own systematic issues.
   %
@@
   %
-  The results include \nplyounggyrotwosigma\
-  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
-  planets at 3$\sigma$).
+  The results include \nplyounggyrotwosigma\ planets younger than
+  1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets at 3$\sigma$).
   %
   While lithium provides only marginal added information for most of
-  this sample,
-  we report new robust ages from both rotation and lithium for 14
-  confirmed $<$500\,Myr planets, including five mini-Neptunes, five
-  super-Earths, three Earths and sub-Earths, and a Neptune-sized
-  planet.
+  this sample, we report new robust ages from both rotation and
+  lithium for 14 confirmed $<$500\,Myr planets, including five
+  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
+  Neptune-sized planet.
   %
@@
   %
-%  The sample of detected Kepler planets has a similar age
-%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
-%  likely due to decreased detection sensitivity.
-  %
-  %Combining our stellar age catalog with Kepler's completeness products, we perform
-  %a hierarchical bayesian analysis to infer the average number of planets per star
-  %as a function of both stellar and planetary parameters.
-  %
-  %While we recover many previously known correlations with stellar mass and metallicity,
-  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
-  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
-  %TODO will this be the case?
-%  We also comment on a few previously reported evolutionary trends in
-%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
-%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
-%  ($\sim$3$\sigma$) significance to previous work, and one of which we
-%  note for the first time.
 \end{abstract}
@@
 Exoplanet science is in an age of abundance, fueled by the discovery
-of thousands of worlds orbiting close to their host stars (CITE, CITE,
-CITE).
-However, most known exoplanets have ages comparable to that of the solar
+of thousands of worlds orbiting close to their host stars
+\citep{Borucki10,2015JATIS...1a4003R}.
+Most of these exoplanets have ages comparable to that of the solar
 system.
-This unfortunate fact leaves many gaps in our knowledge of
-exoplanetary origins, particularly regarding to degree to which
-dynamical migration (CITE), thermal cooling (CITE), and atmospheric
-loss (CITE) sculpt planetary systems after the disk disperses.
-Finding and characterizing young ($<$1\,Gyr) exoplanets offers one
-possible workaround, since these objects can show both individual (CITE,
-CITE) and population-level (CITE) evidence for these types of
-evolutionary processes.
+This fact leaves many gaps in our knowledge of what exact dynamical
+and physical processes produce these objects.
+Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
+one possible workaround.
+The reason is that processes such as thermal cooling, atmospheric
+loss, and dynamical instability occur over $\approx$10\,Myr to
+$\approx$1\,Gyr timescales in a way that can leave observable imprints in
+the properties of young planets (CITE theory reviews).
+Individual exemplars include V1298~Tau \citep{David_2019}, an
+adolescent resonant chain of four transiting puffy planets,
+and HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
+Neptunian mass (CITE).
+Population-level examples include tentative differences in the size
+distribution of young vs.~old planets
+(CITEVach,CITEChristianse,CiteDavid).
 
@@
 Galactic star formation rate.
-
-%While the youngest 0.1\% of stars ($\lesssim$10\,million
-%years; Myr) are perhaps the most constraining for studies of planet
-%formation, they are extremely challenging targets for close-in
-%exoplanet detection 
-%\citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
-%the early orbital, structural, and atmospheric evolution of exoplanets
-%are perhaps most technically feasible for the 
-%$\approx$1\% of stars
-%with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
-%younger than 1\,Gyr.  However, since stellar ages are roughly
-%uniformly distributed in the solar neighborhood, there is a limit on
-%the sample size of young planets that will be detectable in these age
-%regimes, particularly relative to their older breathren.
 
@@
 
-%\begin{figure}[!t]
-%	\begin{center}
-%		\leavevmode
-%		\includegraphics[width=0.49\textwidth]{f1.pdf}
-%	\end{center}
-%	\vspace{-0.6cm}
-%	\caption{
-%		{\bf Radii, orbital periods, and ages of transiting exoplanets}.
-%    Planets younger than 1\,Gyr with ages more precise than a
-%    factor of three are emphasized. 
-%    ({\bf cite relevant papers?})
-%    Parameters are from the \citet{PSCompPars}.
-%		\label{fig:rp_period_age}
-%	}
-%\end{figure}
-
-Have these young Kepler planets already been found?  If they have, 
-it would probably have been by
-\citet{Berger_2020b_rpage}, \citet{David_2021} or
-\citet{Petigura_2022}.  The \citeauthor{Berger_2020b_rpage} and
-\citeauthor{Petigura_2022} studies focused on
-isochronal age-dating of individual stars, which is most precise 
-for stars whose luminosities and temperature separate them from the
-main-sequence.  The \citeauthor{David_2021} study leveraged both isochronal
-ages from an earlier analogous study
+Have these young Kepler planets already been found?  If they have, it
+would probably have been by \citet{Berger_2020b_rpage},
+\citet{David_2021} or \citet{Petigura_2022}.  The
+\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} studies
+focused on isochronal age-dating of individual stars, which is most
+precise for stars whose luminosities and temperature separate them
+from the main-sequence.  The \citeauthor{David_2021} study leveraged
+both isochronal ages from an earlier analogous study
 \citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
@@
 whether they rotated faster or slower than a gyrochrone age of
-$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
-(TODO: note that there's a lot more literature on "young" field Kepler
-planets than these few sentences betray).
+$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  (TODO: note that
+there's a lot more literature on "young" field Kepler planets than
+these few sentences betray).
+
+%     While the youngest 0.1\% of stars ($\lesssim$10\,million
+%     years; Myr) are perhaps the most constraining for studies of planet
+%     formation, they are extremely challenging targets for close-in
+%     exoplanet detection 
+%     \citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
+%     the early orbital, structural, and atmospheric evolution of exoplanets
+%     are perhaps most technically feasible for the 
+%     $\approx$1\% of stars
+%     with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
+%     younger than 1\,Gyr.  However, since stellar ages are roughly
+%     uniformly distributed in the solar neighborhood, there is a limit on
+%     the sample size of young planets that will be detectable in these age
+%     regimes, particularly relative to their older breathren.
 
@@
 Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
-between 0.5-4\,Gyr, and can also provide useful age limits at
-earlier times \citep{Bouma_2023}.  The idea of using a star's
-spin-down as a clock is quite old
+between 0.5-4\,Gyr, and can also provide useful age limits at earlier
+times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
+clock is quite old
 \citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
@@
 dynamo all evolve
-\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
-Today however, measurements of stellar rotation periods in open
-clusters provide the most accurate empirical calibrations of
-rotation-based ages (\citealt{Bouma_2023}, and references therein).
-This approach ultimately ties the rotation age scale to the cluster
-age scale, and is limited in precision by the intrisic scatter in the
-$P_{\rm rot}$ sequences, and in accuracy by systematic uncertainties
-in the evolving spin-down rates, and in the cluster ages themselves.
-The oldest cluster for which this method has been calibrated is
-currently M67 \citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
+\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Today
+however, measurements of stellar rotation periods in open clusters
+provide the most accurate empirical calibrations of rotation-based
+ages (\citealt{Bouma_2023}, and references therein).  This approach
+ultimately ties the rotation age scale to the cluster age scale, and
+is limited in precision by the intrisic scatter in the $P_{\rm rot}$
+sequences, and in accuracy by systematic uncertainties in the evolving
+spin-down rates, and in the cluster ages themselves.  The oldest
+cluster for which this method has been calibrated is currently M67
+\citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
 
@@
 to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
-age, spanning $10^6$ to $10^{10.1}$\,yr, and symmetrize our EW
+age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
 uncertainties for purposes of interfacing with \texttt{EAGLES} by
@@
 relevant for lithium ages, specifically those associated with
-crowding and binarity since those effects can produce both systematic
-and astrophysical biases in EW measurements.
+crowding and binarity.
+For instance, rapid rotators are more common in binaries, and they also
+generally have higher lithium abundances than slow-rotating stars of the
+same mass (CITE).
+Stars for which rotation-based age dating is dubious heavily overlap
+with those for which lithium-based ages are dubious.
+
+In part because many of our EWs are upper limits, 
+many of the lithium ages are lower limits.
+In such cases,  the inferred age posterior is strongly influenced by
+{\it i)} the assumed intrinsic dispersion in the lithium model,
+and {\it ii)} by the prior.
+For instance, 
+ we quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
 

### 2024-04-25 drivers/make_star_table.py

@@ -130,11 +130,11 @@ def make_star_table(
     sdf = sdf.rename(mapdict, axis='columns')
 
     # trim out all columns except those above...
-    sdf = sdf[mapdict.values()]
+    _sdf = sdf[mapdict.values()]
 
     latexpath = join(PAPERDIR, 'table_star_gyro.tex')
     np.random.seed(42)
-    sdf.sample(n=42).to_latex(
+    _sdf.sample(n=42).to_latex(
         latexpath, index=False, na_rep='--', escape=False, formatters=formatters
     )
 
@@ -148,11 +148,8 @@ def make_star_table(
         print(f"Wrote {texpath}")
 
     csvpath = join(PAPERDIR, 'table_star_gyro.csv')
-    sdf.to_csv(csvpath, index=False)
-    #FIXME TODO should these CSV files have formatting applied???
+    _sdf.to_csv(csvpath, index=False)
     print(f'Wrote {csvpath}')
-
-    import IPython ; IPython.embed()
 
 
 if __name__ == "__main__":

### 2024-04-25 drivers/run_paper_scripts.sh

@@ -1,19 +1,24 @@
 #!/bin/bash
 
+# plots that made the ms
 python plot_hist_field_gyro_ages.py
 python plot_star_Prot_Teff.py
 python plot_koi_mean_prot_teff.py
 python plot_li_vs_teff.py
 python plot_st_params.py
 
+# write latex tables and value files for ms
+python _compare_lithium_scales.py
 python get_values_for_manuscript.py
 python make_younghighlight_table.py
+python make_star_table.py
 
+# generate rp vs period vs age
 python make_gyroonly_table.py
 cwd=$(pwd)
 cd ~/Dropbox/proj/NEA_age_plots/drivers
 python plot_rp_vs_period_scatter.py
 cd "$cwd"
 
-echo 'done'
+echo 'done! 🎉'
 exit 0

### 2024-04-26 paper/ms.tex
update ms

@@
   %
-  Here we apply this knowledge to planets and stars observed by the
+  Here we apply this knowledge to stars observed by the
   primary Kepler mission, and calculate the implied rotation and
@@
   and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is probably valid.
+  gyrochronology is valid.
   %
@@
 	\caption{
-    {\bf The stellar samples.}  Our analysis focuses on stars observed
+    {\bf The stars.}  Our analysis focuses on stars observed
     by Kepler with previously reported rotation periods (dark gray
@@
     either ``confirmed'' or ``candidate'' Kepler Objects of Interest
-    that meet an additional set of quality criteria (blue points;
-    Section~\ref{subsec:plflags}).  These are the objects for which
-    our rotation-based ages are most reliable.  We also report
-    lithium-based ages for a largely-overlapping setting of KOIs based
-    on high-resolution spectroscopy.  Surface gravities and effective
+    that meet an additional set of planetary quality criteria (blue points;
+    Section~\ref{subsec:plflags}).  
+%	We independently derive
+%    lithium-based ages for a largely-overlapping setting of KOIs based
+%    on high-resolution spectroscopy.
+    Surface gravities and effective
     temperatures are primarily from
@@
 	\caption{
-		{\bf Rotation periods for Kepler stars (top) and planets
-			(bottom)}.  The periods are from \citet{Santos_2019} and
+		{\bf Rotation periods for Kepler stars (left) and planets
+			(right)}.  The periods are from \citet{Santos_2019} and
 		\citet{Santos_2021} (\citetalias{Santos_2019} and

### 2024-04-26 paper/ms.tex
update ms

@@
 system.
-This fact leaves many gaps in our knowledge of what exact dynamical
-and physical processes produce these objects.
+This fact leaves many gaps in our knowledge of the exact dynamical
+and physical processes that produce these objects.
 Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
@@
 loss, and dynamical instability occur over $\approx$10\,Myr to
-$\approx$1\,Gyr timescales in a way that can leave observable imprints in
+$\approx$1\,Gyr timescales, and so they can leave observable imprints in
 the properties of young planets (CITE theory reviews).
@@
 Neptunian mass (CITE).
+% And perhaps also TOI837, now with a uber massive core!
 Population-level examples include tentative differences in the size
@@
 
-\section{Selecting the Stellar and Planetary Samples}
+\section{Selecting the Stars and Planets}
 \label{sec:selection}
@@
 		\subfloat{
-			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-		\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+		\includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 
-The three main assumptions required for a rotation-based age dating
-method like \texttt{gyro-interp} to be valid are as follows.  {\it 1)}
-The target star must be near the main sequence, {\it 2)} its spindown
-must not be influenced by binary companions, and {\it 3)} the rotation
-period distribution for field stars of a given mass and age must be
-identical to that of equally aged open clusters (metallicity
-differences, for instance, are ignored).  Regarding the former two
-assumptions: pre-main-sequence contraction and post-main-sequence
-expansion influence surface stellar rotation periods in ways that we
-do not model.  Stars in e.g.~ close binaries might be subject to tidal
-effects that we do not treat.
-
 Using our adopted effective temperatures and rotation periods, we ran
 \texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
-over 500 grid points.  \texttt{gyro-interp} itself is calibrated
-out to 4\,Gyr, the adopted age of the M67 cluster
-\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the stellar
-spin-down rate as a function of time and stellar temperature, the
-framework interpolates between reference open clusters using smooth
-cubic hermite polynomials.  We calculate the probability of the
+over 500 grid points.  \texttt{gyro-interp} itself is calibrated out
+to 4\,Gyr, the age of M67
+\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the
+stellar spin-down rate as a function of time and stellar temperature,
+the framework interpolates between reference open clusters using
+smooth cubic hermite polynomials.  We calculate the probability of the
 rotation-based age $t_{\rm gyro}$, given the observed rotation
@@
 
+As a general philosophy, three assumptions must hold for a
+rotation-based age dating method like \texttt{gyro-interp} to be
+valid.  {\it 1)} The evolutionary state of the target star must be
+well-specified by its temperature and its age, {\it 2)} the star's
+spindown must not be influenced by binary companions, and {\it 3)} the
+rotation period distribution for field stars of a given temperature
+and age must be identical to that of equally aged open clusters
+(metallicity differences, for instance, are ignored).  A rephrasing of
+the first condition is that the star must be ``near the main
+sequence'', because during post main sequence evolution, the
+temperatures of stars change.  From $\approx$0.08 to $\approx$3\,Gyr,
+the stars of interest in this work (3800--6200\,K,
+0.5--1.2\,$M_\odot$) have temperatures constant to $\approx$1\%
+\citep{Choi_2016}.  From $\approx$3--4\,Gyr, a 1.2\,$M_\odot$ star's
+temperature drops from $\approx$6200\,K to $\approx$6000\,K
+($\approx$3\%), because its outer layers expand as it begins hydrogen
+shell burning as a subgiant.  We treat this issue in a somewhat
+convoluted manner, discussed in ``bit 9'' below.
+
 {\it Temperature range (bit 0)}---We require stars to have effective
@@
 apparent $G$-band magnitude and parallaxes.  We flag stars with
-$M_{\rm G} < 3.9$, corresponding to main-sequence spectral types
-earlier than $\approx$F8V \citep{Pecaut_2013}.
+$M_{\rm G} < 3.9$ or $M_{\rm G} > 8.5$, corresponding to main-sequence
+spectral types earlier than $\approx$F8V or later than $\approx$M0.5V
+\citep{Pecaut_2013}.
 
@@
 
-{\it RUWE (bit 7)}---We flagged stars with a Gaia DR3 renormalized
+{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3 renormalized
 unit weight error
-(RUWE) %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
-exceeding 1.4 as possible binaries.  Such astrometric outliers can be
-either bona fide astrometric binaries, or more often are marginally
-resolved point sources for which the single-source PSF model used in
-the Gaia pipeline induces spurious astrometric scatter.
+(RUWE) as a function of other stellar parameters, %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
+and flag stars with RUWE$>$1.4 as possible binaries.  Such astrometric
+outliers can be either bona fide astrometric binaries, or more often
+are marginally resolved point sources for which the single-source PSF
+model used in the Gaia pipeline induces spurious astrometric scatter.
 
@@
 The added complications of reddening and metallicity make defining a
-simple cut in any two dimensions challenging.  We explored various
-options, and found that the $+1$$\sigma$ relative isochronal
-uncertainty reported by \citet{Berger_2020a_catalog}, $+\sigma_{t,{\rm
-B20,iso}}/t_{\rm B20,iso}$, reasonably separated FGK stars ``near'' and
-``far'' from the main-sequence.  We therefore selected stars with
-$t_{\rm B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, and fitted an
-an $N^{\rm th}$-order polynomial in the $\log g$-$T_{\rm eff}$ plane
-to these stars between 4100 and 5800\,K.  We let the order of the fit
-vary from $N$=1 to 10, and minimizing the Bayesian Information Criterion
-yielded $N$=6.  The resulting locus is shown in
-Figure~\ref{fig:stellarprops}.  While the details of this procedure
-are entirely arbitrary, the overall purpose is to exclude stars for
-which there is photometric evidence that gyrochronology may not be
-reliable.  The rotation period distribution of sources excluded in
-this manner includes a larger fraction of ``rapid rotators'', likely
-due to a combination of astrophysical and systematic effects.
+simple cut in any two dimensions challenging.  
+We explored various options, and settled on the orange locus in the
+$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
+While the details of how exactly this locus is constructed are
+arbitrary and convoluted,\footnote{In practice, for K dwarfs we fitted an
+$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
+B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
+$t_{\rm B20,iso}$ was the isochronal age reported by
+\citet{Berger_2020a_catalog}. We let the order of the fit vary from
+$N$=1 to 10, minimized the Bayesian Information Criterion,
+and found $N$=6.  For F and G dwarfs, this procedure failed to produce
+a reasonable locus because of shifts in the relative precision of
+the isochrone ages.  We therefore visually selected KOIs that appeared
+to be near the ``main sequence'', fitted a separate polynomial to
+their locations, and then stitched the resulting polynomials from the
+F and G dwarfs to that of the K dwarfs.} the general 
+aim is to exclude stars for which there is photometric evidence
+that gyrochronology may not be reliable.  The rotation period
+distribution of sources excluded in this manner includes a relatively
+larger fraction of ``rapid rotators'' than the main sequence sources,
+likely due to a combination of astrophysical and systematic effects.
+See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
+of \citet{2023ApJS..268....4F}.
 
@@
 
+{\it How do we use the bit-flags?)}
 An important decision for our subsequent analysis is how to combine
-the above flags in order to decide whether or not a star is suited for
+the above quality indicators in order to decide whether or not a star is suited for
 gyrochronology.  By default, we assume that a star is suitable for
@@
 {\it Candidate S/N (bit 1)}---the false positive rate increases
-greatly toward Kepler's noise floor for planet detection (CITE).  We
-required a S/N in excess of the usual floor, through a cut on the
+greatly toward the noise floor for planet detection in any transit survey \citep[e.g.][]{2002ApJ...564..495J}.  We
+required a S/N in excess of Kepler's usual 7.1$\sigma$ floor, through a cut on the
 ``multiple event statistic'' (MES): we required ${\rm MES}$$>$10.
@@
 based on the Markov Chain Monte Carlo (MCMC) fits from
-\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ and
-\nplhoststarwgyroagenograzing\ planets and stars, respectively.
+\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets and
+\nplhoststarwgyroagenograzing\ stars, respectively.
 

### 2024-04-27 paper/ms.tex
update ms

@@
 %Ages of the Kepler Stars and Planets
-\title{A Dearth of Young Stars and Planets in the Kepler Field}
+%\title{A Dearth of Young Stars and Planets in the Kepler Field}
+\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
 
@@
 	\caption{
-		{\bf Age distribution of stars (left) and planet hosts
-			(right) in the Kepler field inferred from rotation periods}.
+		{\bf Kepler's demographic cliff,} visible in the
+		rotation-derived age distribution of stars (left) and planet hosts
+		(right) in the Kepler field.
 		The

### 2024-04-28 paper/ms.tex
update ms

@@
 		{\bf Kepler's demographic cliff,} visible in the
-		rotation-derived age distribution of stars (left) and planet hosts
+		rotation-derived ages of stars (left) and planet hosts
 		(right) in the Kepler field.
 		The 
-		light gray histogram does not impose any requires on factors such as
-		binarity, crowding, or the star's evolutionary state;
-		the age calculation was more likely to be valid for stars that meet the criteria
-		discussed in Section~\ref{subsec:flags} (black).
+		light gray histograms correspond to all
+		\nuniqstarsantosrotteffcut\
+		$T_{\rm eff}\in[3800,6200]\,$K Kepler stars with reported
+		rotation periods, and all 
+		\nplhoststarwgyroage\ planet-hosting stars with reported
+		rotation periods.
+		The black histograms impose necessary quality cuts
+		on factors such as
+		binarity, crowding, and the star's evolutionary state
+		( Section~\ref{subsec:flags}).
+		In detail, these ``histograms'' are resampled from the posterior
+		probability distributions.
+		{\bf todo: numbers dont match.  why?}
 		\label{fig:hist_tgyro}

### 2024-04-28 paper/ms.tex
manual fix nan koi-7368 radius

@@
 		binarity, crowding, and the star's evolutionary state
-		( Section~\ref{subsec:flags}).
+		(Section~\ref{subsec:flags}).
 		In detail, these ``histograms'' are resampled from the posterior

### 2024-04-28 gyrojo/getters.py

@@ -480,12 +480,12 @@ def get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=0,
         a_rp_err1 = df['E_Rp']
         a_rp_err2 = df['e_Rp']
 
-
     else:
         # just the KOI radii.  crappy uncertainties.
         a_rp = df['koi_prad']
         a_rp_err1 = df['koi_prad_err1'] # upper
         a_rp_err2 = np.abs(df['koi_prad_err2']) # lower
+
 
     if GET_BESTGUESS_RADII and COMPARE_AGE_UNCS:
 
@@ -507,7 +507,6 @@ def get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=0,
         mdf = df.merge(pdf, how='left', on='kepoi_number_str')
 
         return mdf
-
 
     a_period = df['koi_period']
 
@@ -542,6 +541,11 @@ def get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=0,
 
     df['adopted_rp'] = a_rp
     df['adopted_period'] = a_period
+
+    df.loc[df.kepoi_name=='K07368.01','adopted_rp'] = 2.22 # source: me, 2022b
+    df.loc[df.kepoi_name=='K05245.01','adopted_rp'] = 3.789 # source: me, 2022a
+
+    assert df.adopted_rp.isna().sum() == 0
 
     return df, paramdict, st_ages
 

### 2024-04-28 drivers/plot_age_comparison.py
add drivers/plot_age_comparison.py

@@ -1,0 +1,11 @@
+import os
+from gyrojo.paths import DATADIR, RESULTSDIR
+import gyrojo.plotting as ap
+
+outdir = os.path.join(RESULTSDIR, 'age_comparison')
+if not os.path.exists(outdir): os.mkdir(outdir)
+
+ap.plot_age_comparison(outdir, ratio_v_gyro=1)
+ap.plot_age_comparison(outdir, iso_v_gyro=1, logscale=0)
+ap.plot_age_comparison(outdir, iso_v_gyro=1, logscale=1)
+ap.plot_age_comparison(outdir, hist_age_unc_ratio=1)

### 2024-04-28 gyrojo/plotting.py

@@ -1071,11 +1071,12 @@ def plot_process_koi_li_posteriors(outdir, cache_id, sampleid, li_method='eagles
 
 
 
-def plot_age_comparison(outdir, logscale=1, iso_v_gyroli=0, ratio_v_gyroli=0,
+def plot_age_comparison(outdir, logscale=1, iso_v_gyro=0, ratio_v_gyro=0,
                         hist_age_unc_ratio=0):
 
     # get data
-    df = get_joint_results(COMPARE_AGE_UNCS=1)
+    df = get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=1,
+                         grazing_is_ok=0, drop_highruwe=1)
 
     # plot all stars
     plt.close("all")
@@ -1088,7 +1089,7 @@ def plot_age_comparison(outdir, logscale=1, iso_v_gyroli=0, ratio_v_gyroli=0,
         (~pd.isnull(df['P22S_age-iso']))
     )
 
-    if iso_v_gyroli:
+    if iso_v_gyro:
 
         ax.errorbar(
             1e6*df.loc[sel, 'adopted_age_median'], 1e9*df.loc[sel, 'P22S_age-iso'],
@@ -1105,7 +1106,7 @@ def plot_age_comparison(outdir, logscale=1, iso_v_gyroli=0, ratio_v_gyroli=0,
             zorder=-2, lw=2
         )
 
-    if ratio_v_gyroli:
+    if ratio_v_gyro:
 
         ax.errorbar(
             1e6*df.loc[sel, 'adopted_age_median'],
@@ -1122,18 +1123,31 @@ def plot_age_comparison(outdir, logscale=1, iso_v_gyroli=0, ratio_v_gyroli=0,
 
     if hist_age_unc_ratio:
 
-        gyroli_m1sig = np.abs(1e6*df.loc[sel, 'adopted_age_-1sigma'])
-        gyroli_p1sig = 1e6*df.loc[sel, 'adopted_age_+1sigma']
+        gyro_m1sig = np.abs(1e6*df.loc[sel, 'adopted_age_-1sigma'])
+        gyro_p1sig = 1e6*df.loc[sel, 'adopted_age_+1sigma']
+        gyro_med = 1e6*df.loc[sel, 'adopted_age_median']
+
+        gyro_p1rel = gyro_p1sig / gyro_med
+        gyro_m1rel = gyro_m1sig / gyro_med
+
[…]

### 2024-04-28 paper/ms.tex
update ms

@@
   %
-  Here we apply this knowledge to stars observed by the
-  primary Kepler mission, and calculate the implied rotation and
-  lithium-based ages.
+  Here we apply this knowledge to stars observed by the primary Kepler
+  mission to calculate rotation and lithium-based ages.
   %
@@
   and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is valid.
-  %
-  The independent lithium ages show X,Y,Z, though they are subject to
-  their own systematic issues.
-  %
-  The relative statistical precisions from rotation, $\sigma_{t_{\rm
-  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
-  statistically precise than isochronal ages for the same stars, and
-  $\approx$X.X times less precise than cluster-based ages.
-  %
-  The results include \nplyounggyrotwosigma\ planets younger than
-  1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets at 3$\sigma$).
-  %
-  While lithium provides only marginal added information for most of
-  this sample, we report new robust ages from both rotation and
-  lithium for 14 confirmed $<$500\,Myr planets, including five
-  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
-  Neptune-sized planet.
-  %
-  In the broader Kepler target star sample, we find that the
-  gyrochrone age distribution rises linearly from 0 to 3\,Gyr, with
-  stars 2\,Gyr old being twice as common in the Kepler field as stars
-  1\,Gyr old.
-  %
-  This parent stellar age distribution bounds the relative balance of
-  young and old planets detectable not only by Kepler, but also by
-  future wide-field transit surveys.
+  gyrochronology is valid;
+  these rotation-based ages are $\approx$15-20$\times$ more precise
+  than isochrone-based ages for our sample of FGK stars.
+  %
+  The independent lithium ages are consistent for XX\% of cases;
+  however this method is less precise than rotation-based age dating
+  for most of our sample.
+  %
+  Our resulting catalog of planetary ages includes
+  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
+  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
+  %
+  In the broader sample of stars in the Kepler field, we find that the
+  rotation-based age distribution decreases linearly from 3\,Gyr ago
+  to present day, implying a linearly decreasing star formation rate.
+  %
+  While this parent stellar age distribution has implications for the
+  star formation history of our Galaxy, it also bounds the relative
+  balance of young and old planets detectable using both Kepler and
+  also future wide-field transit surveys.
   %
@@
 
-\subsection{Age distribution}
+\subsection{The Demographic Cliff}
 
@@
 
+While lithium provides only marginal added information for most of
+this sample, we report new robust ages from both rotation and
+lithium for 14 confirmed $<$500\,Myr planets, including five
+mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
+Neptune-sized planet.
+
 \acknowledgements
@@
 \startlongtable
-\begin{deluxetable*}{lllllllrrrrrrl}
+\begin{deluxetable*}{llllllllrrrrrrl}
 	%\tabletypesize{\footnotesize}
@@
   version of the table is truncated to include the youngest fifty
-  confirmed Kepler planets, and the twenty youngest candidates.  The
+  confirmed Kepler planets, and the twenty youngest candidates,
+  sorted by the minimum of either the rotation or lithium-based
+  age.  The
   full machine-readable table contains ages and age limits for
@@
   KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
-  $t_{\rm gyro}$ & $t_{\rm Li} $ & $R_{\rm p}$ & $P_{\rm orb}$ &
+  $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
   $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
-  -- &   -- & K & days & m\AA & Myr & Myr &    Earths &    days &       int  & int & bool & -- \\
+  -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
 	\hline
@@
 	\enddata
-	\tablecomments{Rows
-		are sorted by the minimum of either the rotation or lithium-based
-		age, and ``confirmed'' planets appear before``candidate'' planets.
+	\tablecomments{``Confirmed'' planets appear before``candidate'' planets.
 		The bit quality flags for the rotation-based ages,

### 2024-04-28 paper/ms.tex
update ms

@@
 
-\author{et al}
+\author{and coauthors!}
 %
@@
   Here we apply this knowledge to stars observed by the primary Kepler
-  mission to calculate rotation and lithium-based ages.
+  mission to curate catalogs of rotation and lithium-based ages.
   %
@@
   these rotation-based ages are $\approx$15-20$\times$ more precise
-  than isochrone-based ages for our sample of FGK stars.
-  %
-  The independent lithium ages are consistent for XX\% of cases;
-  however this method is less precise than rotation-based age dating
+  than isochrone ages for our sample of FGK stars.
+  %
+  While the independent lithium ages are consistent with XX\% of the rotation ages,
+  this method is significantly less precise than rotation-based age dating
   for most of our sample.
@@
 system.
-This fact leaves many gaps in our knowledge of the exact dynamical
-and physical processes that produce these objects.
-Finding and characterizing young ($\lesssim$1\,Gyr) exoplanets offers
+This fact leaves many gaps in our knowledge of the exact dynamical and
+physical processes that produce these objects.
+Finding and characterizing young ($<$1\,Gyr) exoplanets offers
 one possible workaround.
-The reason is that processes such as thermal cooling, atmospheric
-loss, and dynamical instability occur over $\approx$10\,Myr to
-$\approx$1\,Gyr timescales, and so they can leave observable imprints in
-the properties of young planets (CITE theory reviews).
+The reason is that processes such as thermal cooling
+\citep{2007ApJ...659.1661F}, atmospheric loss
+\citep{2019AREPS..47...67O}, giant impacts
+\citep{2014prpl.conf..595R}, and dynamical instability
+\citep{2017MNRAS.470.1750I} are expected to over $\approx$10\,Myr to
+$\approx$1\,Gyr timescales, and so they may leave observable imprints
+in the properties of young planets.
 Individual exemplars include V1298~Tau \citep{David_2019}, an
-adolescent resonant chain of four transiting puffy planets,
-and HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
-Neptunian mass (CITE).
+adolescent resonant chain of four transiting puffy planets, and
+HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
+Neptunian mass (Thao et al.~submitted).
 % And perhaps also TOI837, now with a uber massive core!
 Population-level examples include tentative differences in the size
-distribution of young vs.~old planets
-(CITEVach,CITEChristianse,CiteDavid).
+distribution of young planets relative to their older counterparts
+\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
 
@@
 To begin, let us imagine that the age distribution of stars within a
-few hundred parsecs is uniformly distributed between 0 and 10\,Gyr
-\citep[e.g.][]{Nordstrom_2004}.  
-This approximiation is useful because it clarifies the rarity of young
-stars:
-$\approx$1\% of nearby stars have ages below 100\,Myr,
-and $\approx$10\% are younger than 1\,Gyr.
-Studies of exoplanets forming in disks (CITE), and of exoplanets
-evolving shortly after disk dispersal (CITE CITE), are therefore
+few hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
+\citep[e.g.][]{Nordstrom_2004}.  This approximation is useful because
+it clarifies the rarity of young stars: under these assumptions,
+$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
+have $t$$<$1\,Gyr.  Studies of forming protoplanetary
+\citep{2018A&A...617A..44K}, and of exoplanets evolving shortly after
+disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are therefore
 capped in their maximum achievable sample sizes by the tyranny of the
@@
 
-Despite the observational restrictions, the field of young close-in
-planet discovery has seen remarkable success over the past ten years.
-Pioneering discoveries with the prime Kepler mission (hereafter,
-``Kepler'') focused on stars with known ages---stars in open
-clusters---and searched them for transiting planets
-\citep{Meibom_2013}.  Planets discovered in this way are assumed to
-have the same age as the host cluster.  The resulting ages are precise
-at the $\approx$10\% level.  K2 and TESS have yielded larger numbers
-of young planets in clusters than Kepler, because each of these
-missions has observed a larger fraction of the sky, and therefore more
-stars with known ages
-\citep[e.g.][]{Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Zhou_2022,Zakhozhay_2022,Wood_2023}.
-
-A limitation of this ``cluster-first'' approach is that only
-$\approx$1\% of stars within the nearest few hundred parsecs can
-currently be associated with their birth cluster
+Despite the observational restrictions, young close-in planet
+discovery has seen remarkable success over the past decade.
+This has been enabled by a combination of Kepler, K2, TESS, and Gaia
+\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
+The original strategy from around 2010 to 2020 was to focus on stars
+with known ages---obvious members of open clusters---and to search
+them for transiting planets.  
+The resulting stellar ages are precise at the $\approx$10\% level, and
+planets discovered in this way are generally assumed to have the same
+age as the host cluster.  
+A more recent development, facilitated by the second Gaia data
+release, is the idea that finding transiting planets can often be
+easier than finding and characterizing the host stellar association.
+
+Despite recent advances, the main problem with any ``cluster-first'' approach is that
+only $\approx$1\% of stars within the nearest few hundred parsecs can
+currently be associated with their birth cluster, even using the Gaia
+data
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-While there may be paths toward resolving this problem, this currently
-means that there are many ``young'' (0.05-1\,Gyr) stars in the field
-that are omitted by cluster-only searches.  Requiring ages precise to
-a factor of three ($t/\sigma_t > 3$), the NASA Exolpanet Archive
-currently reports $\approx$40 transiting planets younger than 1\,Gyr,
-essentially all of which are in clusters.  However, Kepler discovered
-$\approx$4{,}000 planets.  Unless there is a major discovery bias, we
-would expect a larger fraction of the Kepler planets---perhaps 10\%,
-or 400 planets---to be younger than 1\,Gyr, based on our order of
-magnitude reasonsing.
-
-Have these young Kepler planets already been found?  If they have, it
-would probably have been by \citet{Berger_2020b_rpage},
-\citet{David_2021} or \citet{Petigura_2022}.  The
-\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} studies
-focused on isochronal age-dating of individual stars, which is most
-precise for stars whose luminosities and temperature separate them
-from the main-sequence.  The \citeauthor{David_2021} study leveraged
+In other words, the majority of nearby sub-gigayear stars are in the
+field, and are omitted by cluster-only planet searches.  
+%TODO FIXME: is this rhetorical point fair?
+One can appreciate the problem by querying NASA's Exoplanet Archive
+for transiting planets younger than 1\,Gyr.
+Imposing reasonable precision requirements, e.g. either $t$$<$1\,Gyr
+at 2$\sigma$, or else requiring ages precise to say S/N$=$3, gives $\approx$50 ``young'' transiting planets at the time of writing.
+Most of these planets are in clusters, and have been discovered by Kepler, K2, or TESS.  
+However, combined, these surveys have discovered
+$\approx$5{,}000 planets.  Barring a major discovery bias, we
+would expect an order of magnitude more transiting 
+planets to be younger than 1\,Gyr, based on our earlier order of
+magnitude reasoning.
+
+This study aims to resolve two main issues.
+First, there is the question of how wrong it is to assume a uniform age
+distribution for stars in the galactic thin disk, particularly young stars.
+Second, there is the question of which planets from the main Kepler
+mission are actually young.
+
+Given that Kepler's primary mission ended in 2013, one might be forgiven for
+assuming that the ages of the Kepler planets are already known.
+Of course, they have been studied:
+\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for instance calculated
+isochronal ages for individual stars, which is most
+precise for stars whose luminosities and temperatures separate them
+from the main-sequence.
+\citeauthor{David_2021} leveraged
 both isochronal ages from an earlier analogous study
-\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, a
+\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, this work used a
 relative gyrochrone age-ranking approach that sorted stars based on
 whether they rotated faster or slower than a gyrochrone age of
-$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  (TODO: note that
-there's a lot more literature on "young" field Kepler planets than
-these few sentences betray).
+$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
+These studies were conducted in the broader context of investigating
+trends in exoplanet demographics, which have been shown to depend on
+both stellar metallicity and mass \citet{Petigura_2018}, at levels
+likely stronger than age (at least over gigayear timescales).
+However, these studies offered limited sensitivity to the ages of the youngest
+planets, due to the limitations of isochronal dating.
+
+%      and \citet{Petigura_2022} have reported trends between small-planet
+%      occurrence and both stellar metallicity and mass; these parameters
+%      seem, at first glance, to be more important predictors of bulk changes
+%      in the planetary population than age.  Given these known correlations,
+%      it can be useful to consider population models in which variables such
+%      as metallicity, mass, and age are added in succession
+%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
+%      rigorous approach, is to jointly infer the number of planets per star
+%      as a function of both planetary and stellar parameters, the latter
+%      including age \citep[e.g.][]{2023AJ....166..209M}.
 
@@
     provenances.  Gray points are transiting planets from the NASA
-    Exoplanet Archive older than 1\,Gyr.  The selection function
-    matters: selecting for precise ages ($t/\sigma_{\rm t}>3$) would
-    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$twenty
+    Exoplanet Archive older than 1\,Gyr.  If
+ 	one were to select for precise ages ($t/\sigma_{\rm t}>3$), this would
+    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
     0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
     hosts to have no quality flags raised, and for the planets to
-    similarly no have quality flags raised.  Weakening these
-    constraints would expand the sample at the expense of reliability.
+    similarly no have quality flags raised. 
 		\label{fig:rp_period_age_results}
@@
 
-L.G.B.~conceived the project, collected HIRES spectra, executed the
-rotation- and lithium-based age analyses, and drafted the manuscript.
-E.K.P.~contributed to the rotation-based age analysis.
+L.G.B.~conceived the project, collected data, 
+inferred ages, and drafted the manuscript.
+E.K.P.~contributed to early iterations of the rotation-based age analysis.
 L.A.H.~contributed to project design.

### 2024-04-29 paper/ms.tex
update ms

@@
   Here we apply this knowledge to stars observed by the primary Kepler
-  mission to curate catalogs of rotation and lithium-based ages.
+  mission in order to derive rotation and lithium-based ages.
   %
@@
   %
-  While the independent lithium ages are consistent with XX\% of the rotation ages,
-  this method is significantly less precise than rotation-based age dating
-  for most of our sample.
+  While the independent lithium and rotation-based ages agree
+  for $\approx$90\% of stars, the timescales of lithium depletion are
+  such that rotation is the more precise age diagnostic for the
+  majority of our stellar sample.
   %
@@
   %
-  In the broader sample of stars in the Kepler field, we find that the
-  rotation-based age distribution decreases linearly from 3\,Gyr ago
-  to present day, implying a linearly decreasing star formation rate.
+  For both planet-hosts, and for the broader sample of stars in the
+  Kepler field, we find that the stellar age distribution decreases
+  linearly from 3\,Gyr ago to present day, implying a linearly
+  decreasing star formation rate.
   %
@@
 
+\subsubsection{Rotation vs.~Lithium Age Comparison}
+
+We judged the consistency of the independent rotation and
+lithium-based ages by comparing the point-estimates of the posteriors,
+when available.
+There are three cases of interest:
+
+\begin{enumerate}
+  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
+    $t_{\rm Li}$: then if their values were consistent within
+    2$\sigma$, we judged them to be consistent; if they were
+    consistent within 2-3$\sigma$, we suggest them to be ``maybe''
+    consistent.  Otherwise, they are in disagreement.
+\item If $t_{\rm Li}$ is a lower limit, we compare the 3$\sigma$
+  $t_{\rm Li}$ lower limit with the $1\sigma$ upper limit from $t_{\rm
+    gyro}$; if the two overlapped, we judge the age estimates to be
+    consistent.
+\item If $t_{\rm Li}$ provides a two-sided posterior, and no rotation
+  is reported, we judge the age estimates to be inconsistent.
+  This is because such constraints can only be provided for G and K
+  stars $\lesssim$2\,Gyr old, and Kepler should have been sensitive to
+  the rotation periods of such stars.
+\end{enumerate}
+
+In the entire sample, this yielded 
+\allagesyesconsistent\ consistent cases,
+\allagesmaybeconsistent\ ``maybe'' cases, and
+\allagesnoconsistent\ inconsistent.
+Restricting our attention to the $\lesssim$1\,Gyr stars, for which
+lithium ages are more constraining,
+the relevant numbers are
+\minageltonegyryesconsistent, \minageltonegyrmaybeconsistent, and
+\minageltonegyrnoconsistent.
+Rephrased, in the overall sample, \fracconsistentallages\%
+of stars have consistent rotation and lithium-based ages
+(\fracpotentiallyconsistentallages\% potentially consistent).
+In the sample of sub-gigayear stars,
+\fracconsistentminageltonegyr\ of these ages are
+consistent (\fracpotentiallyconsistentminageltonegyr\% potentially
+consistent).
+We feel it necessary to distinguish ``maybe'' consistent cases because
+of lingering uncertainties in how the dispersion of the lithium data
+evolve over the first gigayear.
+In Appendix~\ref{app:inconsistent}, we discuss the
+\minageltonegyrnoconsistent\ discrepant cases in detail.
+
 \subsubsection{Systems with Rotation and Lithium in Agreement}
@@
 %  1.0\,$R_\oplus$, 12.2\,day Earth-sized Kepler-1482b.
-
-\subsubsection{Systems with Rotation and Lithium in Disagreement}
-
-For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
-This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
-rotational broadening of the spectrum ($v\sin
-i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
-rotation period.  However, if this star were $\lesssim$500\,Myr old,
-lithium should have been detectable at $>$50\,m\AA.  Lithium is
-marginally if at all present in the spectrum ({\bf  EW$_{\rm
-Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
-6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
-\citep{Morton_2016}, inspecting the data validation reports this is
-somewhat surprising.  The light curve shows greater scatter in-transit
-than out of transit, appears near-grazing, and has a suggestion of a
-strong odd-even depth difference.  This is a surprising system that
-merits further attention.
 
@@
 		Bit 2: Grazing?
+		For a star to have a high likelihood of being ``reliable for gyrochronology'' we suggest $Q_{\rm star}$ bits
+		zero to nine to not be raised, and for a planet to be ``reliabe'', we suggested $Q_{\rm planet}$ to be zero.
 	}
@@
 \appendix
-%\section{Can you avoid appendices?}
-%\label{app:interp}
+\section{Cases With Inconsistent Rotation and Lithium-Based Ages}
+\label{app:inconsistent}
+
+For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
+This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
+rotational broadening of the spectrum ($v\sin
+i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
+rotation period.  However, if this star were $\lesssim$500\,Myr old,
+lithium should have been detectable at $>$50\,m\AA.  Lithium is
+marginally if at all present in the spectrum ({\bf  EW$_{\rm
+Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
+6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
+\citep{Morton_2016}, inspecting the data validation reports this is
+somewhat surprising.  The light curve shows greater scatter in-transit
+than out of transit, appears near-grazing, and has a suggestion of a
+strong odd-even depth difference.  This is a surprising system that
+merits further attention.
+
+% from inconsitent_tli_tgyro cases:
+Kepler-1644 b
+Kepler-1699 b
+ Kepler-786 b
+
+these have finite Li ages, but no rotation age (perhaps b/c of the missing prots...)
+Kepler-1943 b
+ Kepler-101 b
+ Kepler-108 b
+ Kepler-108 c
+  Kepler-63 b
+Kepler-1312 b
+Kepler-1312 c
+   Kepler-1 b
+
+K02680.01
+K02639.01
 

### 2024-04-29 paper/ms.tex
update ms

@@
   \item Two-sided posteriors available for both $t_{\rm gyro}$ and
-    $t_{\rm Li}$: then if their values were consistent within
+    $t_{\rm Li}$: if their values were consistent within
     2$\sigma$, we judged them to be consistent; if they were
@@
     consistent.  Otherwise, they are in disagreement.
-\item If $t_{\rm Li}$ is a lower limit, we compare the 3$\sigma$
+\item If $t_{\rm Li}$ is a lower limit, we compare the
   $t_{\rm Li}$ lower limit with the $1\sigma$ upper limit from $t_{\rm
@@
   is reported, we judge the age estimates to be inconsistent.
-  This is because such constraints can only be provided for G and K
+  This is because this lithium constraint can only be provided for G and K
   stars $\lesssim$2\,Gyr old, and Kepler should have been sensitive to
@@
 lithium ages are more constraining,
-the relevant numbers are
+the respective counts are
 \minageltonegyryesconsistent, \minageltonegyrmaybeconsistent, and
@@
 In the sample of sub-gigayear stars,
-\fracconsistentminageltonegyr\ of these ages are
+\fracconsistentminageltonegyr\% of the ages are
 consistent (\fracpotentiallyconsistentminageltonegyr\% potentially
@@
 In Appendix~\ref{app:inconsistent}, we discuss the
-\minageltonegyrnoconsistent\ discrepant cases in detail.
+\minageltonegyrnoconsistent\ discrepant sub-gigayear stars in detail.
 

### 2024-04-29 gyrojo/getters.py
Prot_provenance: key update

@@ -636,7 +636,7 @@ def get_kicstar_data(sampleid):
              'Fl5':'s19_flag5'},
             axis='columns'
         )
-        s19_df['Provenance'] = 'Santos2019'
+        s19_df['Prot_provenance'] = 'Santos2019'
         for ix in range(1,6):
             s19_df[f"s21_flag{ix}"] = np.nan
 
@@ -647,7 +647,7 @@ def get_kicstar_data(sampleid):
         # 'flag1', 'flag2', 'flag3', 'flag4', 'flag5', 'KCat', 'Simbad', '_RA', '_DE'
         catalogs = Vizier.get_catalogs("J/ApJS/255/17")
         s21_df = catalogs[0].to_pandas()
-        s21_df['Provenance'] = 'Santos2021'
+        s21_df['Prot_provenance'] = 'Santos2021'
         s21_df = s21_df.rename(
             {'flag1':'s21_flag1', 'flag2':'s21_flag2', 'flag3':'s21_flag3',
              'flag4':'s21_flag4', 'flag5':'s21_flag5'},
@@ -659,7 +659,7 @@ def get_kicstar_data(sampleid):
         selcols = [
             'KIC', 'Kpmag', 'Q', 'Teff', 'E_Teff', 'e_Teff', 'logg', 'E_logg',
             'e_logg', 'Mass', 'E_Mass', 'e_Mass', 'Prot', 'E_Prot', 'Sph', 'E_Sph',
-            'Provenance',
+            'Prot_provenance',
             's19_flag1', 's19_flag2', 's19_flag3', 's19_flag4', 's19_flag5',
             's21_flag1', 's21_flag2', 's21_flag3', 's21_flag4', 's21_flag5'
         ]
@@ -678,7 +678,7 @@ def get_kicstar_data(sampleid):
         for c in selcols:
             if c not in bonusdf:
                 bonusdf[c] = np.nan
-        bonusdf['Provenance'] = 'SantosPrivComm'
+        bonusdf['Prot_provenance'] = 'SantosPrivComm'
 
         _df = pd.concat((
             s19_df[selcols], s21_df[selcols], bonusdf[selcols])

### 2024-04-29 paper/ms.tex
update ms

@@
 %%%%%%%%%%%%%
+\defcitealias{2013ApJ...775L..11M}{M13}
 \defcitealias{McQuillan_2014}{M14}
@@
 of thousands of worlds orbiting close to their host stars
-\citep{Borucki10,2015JATIS...1a4003R}.
-Most of these exoplanets have ages comparable to that of the solar
-system.
-This fact leaves many gaps in our knowledge of the exact dynamical and
-physical processes that produce these objects.
-Finding and characterizing young ($<$1\,Gyr) exoplanets offers
-one possible workaround.
-The reason is that processes such as thermal cooling
-\citep{2007ApJ...659.1661F}, atmospheric loss
-\citep{2019AREPS..47...67O}, giant impacts
+\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets have
+ages comparable to that of the solar system.  This fact leaves many
+gaps in our knowledge of the exact dynamical and physical processes
+that produce these objects.  Finding and characterizing young
+($<$1\,Gyr) exoplanets offers one possible workaround.  The reason is
+that processes such as thermal cooling \citep{2007ApJ...659.1661F},
+atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
 \citep{2014prpl.conf..595R}, and dynamical instability
@@
 $\approx$1\,Gyr timescales, and so they may leave observable imprints
-in the properties of young planets.
-Individual exemplars include V1298~Tau \citep{David_2019}, an
-adolescent resonant chain of four transiting puffy planets, and
-HIP~67522 \citep{Rizzuto_2020}, a Jupiter-sized planet with a
-Neptunian mass (Thao et al.~submitted).
+in the properties of young planets.  Individual exemplars include
+V1298~Tau \citep{David_2019}, an adolescent resonant chain of four
+transiting puffy planets, and HIP~67522 \citep{Rizzuto_2020}, a
+Jupiter-sized planet with a Neptunian mass (Thao et al.~submitted).
 % And perhaps also TOI837, now with a uber massive core!
@@
 Despite the observational restrictions, young close-in planet
-discovery has seen remarkable success over the past decade.
-This has been enabled by a combination of Kepler, K2, TESS, and Gaia
+discovery has seen remarkable success over the past decade.  This has
+been enabled by a combination of Kepler, K2, TESS, and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
@@
 with known ages---obvious members of open clusters---and to search
-them for transiting planets.  
-The resulting stellar ages are precise at the $\approx$10\% level, and
-planets discovered in this way are generally assumed to have the same
-age as the host cluster.  
-A more recent development, facilitated by the second Gaia data
-release, is the idea that finding transiting planets can often be
-easier than finding and characterizing the host stellar association.
-
-Despite recent advances, the main problem with any ``cluster-first'' approach is that
-only $\approx$1\% of stars within the nearest few hundred parsecs can
-currently be associated with their birth cluster, even using the Gaia
-data
+them for transiting planets.  The resulting stellar ages are precise
+at the $\approx$10\% level, and planets discovered in this way are
+generally assumed to have the same age as the host cluster.  A more
+recent development, facilitated by the second Gaia data release, is
+the idea that finding transiting planets can often be easier than
+finding and characterizing the host stellar association.
+
+Despite recent advances, the main problem with any ``cluster-first''
+approach is that only $\approx$1\% of stars within the nearest few
+hundred parsecs can currently be associated with their birth cluster,
+even using the Gaia data
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
@@
 One can appreciate the problem by querying NASA's Exoplanet Archive
-for transiting planets younger than 1\,Gyr.
-Imposing reasonable precision requirements, e.g. either $t$$<$1\,Gyr
-at 2$\sigma$, or else requiring ages precise to say S/N$=$3, gives $\approx$50 ``young'' transiting planets at the time of writing.
-Most of these planets are in clusters, and have been discovered by Kepler, K2, or TESS.  
-However, combined, these surveys have discovered
-$\approx$5{,}000 planets.  Barring a major discovery bias, we
-would expect an order of magnitude more transiting 
-planets to be younger than 1\,Gyr, based on our earlier order of
-magnitude reasoning.
-
-This study aims to resolve two main issues.
-First, there is the question of how wrong it is to assume a uniform age
-distribution for stars in the galactic thin disk, particularly young stars.
-Second, there is the question of which planets from the main Kepler
-mission are actually young.
-
-Given that Kepler's primary mission ended in 2013, one might be forgiven for
-assuming that the ages of the Kepler planets are already known.
-Of course, they have been studied:
-\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for instance calculated
-isochronal ages for individual stars, which is most
-precise for stars whose luminosities and temperatures separate them
-from the main-sequence.
-\citeauthor{David_2021} leveraged
-both isochronal ages from an earlier analogous study
-\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis, this work used a
-relative gyrochrone age-ranking approach that sorted stars based on
-whether they rotated faster or slower than a gyrochrone age of
-$\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
-These studies were conducted in the broader context of investigating
-trends in exoplanet demographics, which have been shown to depend on
-both stellar metallicity and mass \citet{Petigura_2018}, at levels
-likely stronger than age (at least over gigayear timescales).
-However, these studies offered limited sensitivity to the ages of the youngest
+for transiting planets younger than 1\,Gyr.  Imposing reasonable
+precision requirements, e.g. either $t$$<$1\,Gyr at 2$\sigma$, or else
+requiring ages precise to say S/N$=$3, gives $\approx$50 ``young''
+transiting planets at the time of writing.  Most of these planets are
+in clusters, and have been discovered by Kepler, K2, or TESS.
+However, combined, these surveys have discovered $\approx$5{,}000
+planets.  Barring a major discovery bias, we would expect an order of
+magnitude more transiting planets to be younger than 1\,Gyr, based on
+our earlier order of magnitude reasoning.
+
+This study aims to resolve two main issues.  First, there is the
+question of how wrong it is to assume a uniform age distribution for
+stars in the galactic thin disk, particularly young stars.  Second,
+there is the question of which planets from the main Kepler mission
+are actually young.
+
+Given that Kepler's primary mission ended in 2013, one might be
+forgiven for assuming that the ages of the Kepler planets are already
+known.  Of course, they have been studied:
+\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for
+instance calculated isochronal ages for individual stars, which is
+most precise for stars whose luminosities and temperatures separate
+them from the main-sequence.  \citeauthor{David_2021} leveraged both
+isochronal ages from an earlier analogous study
+\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis,
+this work used a relative gyrochrone age-ranking approach that sorted
+stars based on whether they rotated faster or slower than a gyrochrone
+age of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These
+studies were conducted in the broader context of investigating trends
+in exoplanet demographics, which have been shown to depend on both
+stellar metallicity and mass \citet{Petigura_2018}, at levels likely
+stronger than age (at least over gigayear timescales).  However, these
+studies offered limited sensitivity to the ages of the youngest
 planets, due to the limitations of isochronal dating.
@@
 dynamo all evolve
-\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  Today
-however, measurements of stellar rotation periods in open clusters
-provide the most accurate empirical calibrations of rotation-based
-ages (\citealt{Bouma_2023}, and references therein).  This approach
-ultimately ties the rotation age scale to the cluster age scale, and
-is limited in precision by the intrisic scatter in the $P_{\rm rot}$
+\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
+In the context of Kepler,
+empirical models have been previously applied to derive rotation-based
+ages for subsets of the Kepler data
+\citep{Walkowicz_2013,Reinhold_2015}.
+There have been major improvements since these pioneering studies in:
+\begin{enumerate}[label={\it \roman*)}]
+  \item The depth of the Kepler Object of Interest catalog, as well as our understanding of which of these objects are
+    false positives;
+  %
+  \item Our empirical understanding of how rotation periods of stars
+    in open clusters evolve
+    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
+  %
+  \item Our ability to model not only the mean evolution but also the
+    intrinsic dispersion in rotation periods as a function of time
+    \citep{Bouma_2023}.  
+    The new model is broadly consistent with older calibrations
+    \citep[e.g.][]{Mamajek_2008} for Sun-like stars.  However, for
+    late G dwarfs and K dwarfs, this age-calibration differs by
+    factors of a few from previous estimates;
+  %
+  \item Our ability to identify binary stars, which can potentially
+    problematic for estimates of rotation-based ages.
+\end{enumerate}
+Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
+empirically tied to the cluster age scale and rotation sequence, and
+are limited in precision by the intrisic scatter in the $P_{\rm rot}$
 sequences, and in accuracy by systematic uncertainties in the evolving
-spin-down rates, and in the cluster ages themselves.  The oldest
-cluster for which this method has been calibrated is currently M67
-\citep[$\approx$4\,Gyr][]{2022ApJ...938..118D}.
-
-Lithium ages includes two qualitatively distinct regimes: depletion
+spin-down rates, and in the cluster ages themselves.
+
+Beyond improvements in rotation-based ages, recent years have also
+yielded improvements in the lithium age scale.
+Lithium ages include two qualitatively distinct regimes: depletion
 boundary ages derived for M-dwarfs in star clusters, and the less

### 2024-04-30 paper/ms.tex
ditto: propagate to ms

@@
 \begin{abstract}
-  Recent studies of stars in open clusters have shown how the rotation
-  rates and lithium content of FGK stars evolve during the first few
-  billion years of their lives.
-  %
-  Here we apply this knowledge to stars observed by the primary Kepler
-  mission in order to derive rotation and lithium-based ages.
+  Recent studies of the rotation rates and lithium content of FGK
+  stars in open clusters have revealed how these quantities evolve
+  during the first few billion years of a star's life.
+  %
+  Here we apply this knowledge to stars observed by Kepler in order to
+  derive rotation and lithium-based ages.
   %
@@
   %
-  While the independent lithium and rotation-based ages agree
-  for $\approx$90\% of stars, the timescales of lithium depletion are
-  such that rotation is the more precise age diagnostic for the
-  majority of our stellar sample.
+  The independent lithium and rotation-based ages agree for
+  $\approx$90\% of stars.  However, the timescales of lithium
+  depletion are such that rotation is the more precise age diagnostic
+  for the majority of our stellar sample.
   %
@@
   %
-  For both planet-hosts, and for the broader sample of stars in the
-  Kepler field, we find that the stellar age distribution decreases
-  linearly from 3\,Gyr ago to present day, implying a linearly
-  decreasing star formation rate.
-  %
-  While this parent stellar age distribution has implications for the
+  More broadly, we find that the age distribution of stars in the
+  Kepler field decreases linearly from 3\,Gyr ago to present day,
+  for both known planet hosts as well as the parent stellar sample.
+  %
+  While this stellar age distribution has implications for the
   star formation history of our Galaxy, it also bounds the relative
@@
 To select stars with rotation periods, we turn to previous work.
-Studies of stellar rotation across the entire Kepler field include
-those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
-\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
+Many investigators have derived rotation periods both specifically of
+the Kepler Objects of Interest
+\citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
+and also of the broader set of all Kepler target stars
+\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
+These studies used a range of detection methods and selection
+functions.
+We are interested in understanding the age distribution of not only
+the KOIs but also the Kepler target stars.
+Out of these studies, the only homogeneous analysis of both Kepler
+targets and KOIs were the studies by
+\citet{Santos_2019} (\citetalias{Santos_2019}),
 and \citet{Santos_2021} (\citetalias{Santos_2021}).
-\citetalias{McQuillan_2014} used an approach based on the
-autocorrelation function to detect 34{,}030 rotation periods for
-main-sequence Kepler targets cooler than 6500\,K, and excluded known
-eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
-Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
-variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
-candidates, or pulsators; they reported primary rotation periods for
-24{,}124 of these stars.  \citet{Reinhold_2015} also reported
-secondary periods for two thirds of the stars with primary periods,
-which could be caused by differential rotation or finite spot
-lifetimes.  Finally, \citetalias{Santos_2019} and
+\citetalias{Santos_2019} and
 \citetalias{Santos_2021} combined a wavelet analysis and
@@
 periods for 55{,}232 main-sequence and subgiant FGKM stars.
-\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
-KOIs and binaries, and assigned them specific quality flags.  The
-rotation periods of KOIs have received considerable additional
-scrutiny
-\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.
-
-We choose to adopt the results of \citetalias{Santos_2019} and
+They included known KOIs and binaries in their analysis, and assigned
+them specific quality flags, while removing transits and eclipses
+during the stellar rotation measurement process.
+
+%Studies of stellar rotation across the entire Kepler field include
+%those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
+%\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
+%and \citet{Santos_2021} (\citetalias{Santos_2021}).
+%\citetalias{McQuillan_2014} used an approach based on the
+%autocorrelation function to detect 34{,}030 rotation periods for
+%main-sequence Kepler targets cooler than 6500\,K, and excluded known
+%eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
+%Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
+%variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
+%candidates, or pulsators; they reported primary rotation periods for
+%24{,}124 of these stars.  \citet{Reinhold_2015} also reported
+%secondary periods for two thirds of the stars with primary periods,
+%which could be caused by differential rotation or finite spot
+%lifetimes.  Finally, \citetalias{Santos_2019} and
+%\citetalias{Santos_2021} combined a wavelet analysis and
+%autocorrelation-based approach, and cumulatively reported rotation
+%periods for 55{,}232 main-sequence and subgiant FGKM stars.
+%\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
+%KOIs and binaries, and assigned them specific quality flags.  The
+%rotation periods of KOIs have received considerable additional
+%scrutiny
+%\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.
+
+We therefore adopt the results of \citetalias{Santos_2019} and
 \citetalias{Santos_2021} as our default rotation periods.
-\citetalias{Santos_2021} provides a detailed comparison against
+\citetalias{Santos_2021} provided a comparison against
 \citetalias{McQuillan_2014}; the brief summary is that the periods
@@
 (e.g.~giant stars and eclipsing binaries) and visual inspection.  In
-addition, \citetalias{Santos_2021} report rotation periods for
+addition, \citetalias{Santos_2021} reported rotation periods for
 24{,}182 main-sequence and subgiant FGKM stars that were not reported
 as periodic by \citetalias{McQuillan_2014}.  Many of these reported
-detections have lower variability amplitudes, and longer periods, than
+detections have lower variability amplitudes and longer periods than
 those reported by \citetalias{McQuillan_2014}.  A relevant flag for
 stars missing \citetalias{McQuillan_2014} periods is included in
-Table~\ref{tab:stars}.  Reviewing the resulting
-period detections, we learned that 32 KOIs with orbital and rotation
-periods within $\approx$20\% were excluded from the
+Table~\ref{tab:stars}.
+
+Analyzing the compilation of \citetalias{Santos_2019} and
+\citetalias{Santos_2021} rotation periods for the KOI hosts, we
+noticed that some known KOIs with rotation periods were missing.
+This is not surprising, in that the rotation periods of KOIs have
+received considerably more scrutiny than those of ordinary Kepler
+stars.
+We therefore decided to split our subsequent analysis into a
+homogeneous portion that used only the \citetalias{Santos_2019} and
+\citetalias{Santos_2021} data, and an inhomogeneous portion that also
+considered a broader set of available KOI rotation periods.
+For the latter portion, we first included 32 KOIs with orbital and
+rotation periods within $\approx$20\% that had been
+excluded\footnote{Excluding stars in this category would impose a bias
+against young close-in planets, due to the natural commensurability
+between the rotation rates of young stars, and Kepler's increased
+sensitivity to short-period objects.  Although some of these KOIs are
+known false positives, our analysis excludes such objects at later
+steps.}%  As one motivating example, this subset of stars includes {\it
+%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
+%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
+%based on membership in RSG-5 \citep{Bouma_2022b}.  }
+from the
 \citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
-(A.~Santos, priv.~comm.); we choose to include these
-objects.\footnote{In general, excluding stars in this category would
-impose a bias against young close-in planets, due to the natural
-commensurability between the rotation rates of young stars, and
-Kepler's increased sensitivity to short-period objects.  While some of
-these KOIs are known false positives, our analysis naturally excludes
-such objects at later steps.  As one motivating example, this subset
-of stars happens to include {\it e.g.}~Kepler~1643 (KIC~8653134), with
-$P_{\rm orb}$=5.34\,days and $P_{\rm rot}$=5.05\,days, which is known
-to be $\approx$40\,Myr old based on membership in RSG-5
-\citep{Bouma_2022b}.  } The final tally of stars from
+(A.~Santos, private communication).
+We then incorporated an additional \nnewdavidtwentyone\ rotation
+periods for KOI hosts that
+\citet{David_2021} had described as either ``reliable'' or ``highly
+reliable'' in their visual analysis of previously reported KOI
+rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
+and \citet{Angus_2018}.
+Inclusion of these additional KOI rotation periods is a
+supplementary measure aimed at completeness in our final KOI age
+catalog; the provenances of the individual adopted periods are all
+noted in the relevant tables.
+
+Finally, since our scope is focused on rotation-based age dating, 
+we restricted our attention to stars with reported $P_{\rm
+rot}$$<$45\,days.
+FGK stars with slower rotation periods spin much slower than the stars
+in the oldest open clusters currently used to calibrate empirical
+gyrochronal models.
+
+The resulting tally of stars from
 \citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
 KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.
-
-%  %TODO is it?
-%  We verified that the main conclusions of our analysis do not change if
-%  we were to adopt only the periods from \citetalias{McQuillan_2014}.
-%  %TODO have you?
 

### 2024-04-30 paper/ms.tex
update ms

@@
   \caption{
-		{\bf Rotation periods for Kepler stars (left) and planets
-			(right)}.  The periods are from \citet{Santos_2019} and
-		\citet{Santos_2021} (\citetalias{Santos_2019} and
+    {\bf Rotation periods for Kepler target stars (left) and known
+    planet hosts (right)}.  The periods are from \citet{Santos_2019}
+    and \citet{Santos_2021} (\citetalias{Santos_2019} and
     \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
@@
     Bit 10: \citetalias{Santos_2021} CP/CB?
+    Bit 11: $P_{\rm rot}$ not in the homogeneous
+    \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
     As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.

### 2024-05-01 paper/ms.tex
update ms

@@
     \leavevmode
-    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf}
+    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf}
   \end{center}
@@
   \caption{
-    {\bf Kepler's demographic cliff,} visible in the
-    rotation-derived ages of stars (left) and planet hosts
-    (right) in the Kepler field.
-    The 
-    light gray histograms correspond to all
-    \nuniqstarsantosrotteffcut\
-    $T_{\rm eff}\in[3800,6200]\,$K Kepler stars with reported
-    rotation periods, and all 
-    \nplhoststarwgyroage\ planet-hosting stars with reported
-    rotation periods.
-    The black histograms impose necessary quality cuts
-    on factors such as
-    binarity, crowding, and the star's evolutionary state
+    {\bf Kepler's demographic cliff,} visible in the rotation-derived
+    ages of stars (left) and planet hosts (right) in the Kepler field.
+    The lightly colored histograms correspond to all
+    \nuniqstarsantosrotteffcut\ $T_{\rm eff}\in[3800,6200]\,$K Kepler
+    stars with reported rotation periods, and all
+    \nplhoststarwgyroage\ planet-hosting stars with reported rotation
+    periods.
+    The darker histograms impose necessary quality cuts on factors
+    such as binarity, crowding, and the star's evolutionary state
     (Section~\ref{subsec:flags}).
@@
     probability distributions.
+    Completeness is near unity until $\approx$3\,Gyr
+    \citep{2022ApJ...937...94M}.
     {\bf todo: numbers dont match.  why?}

### 2024-05-01 paper/collect.sh

@@ -19,7 +19,7 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
 
 # plot_hist_field_gyro_ages 
-cp $r/hist_field_gyro_ages_20240405/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240405_maxage3200.pdf .
+cp $r/hist_field_gyro_ages_20240430/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf .
 
 # plot_li_vs_teff
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .

### 2024-05-01 paper/ms.tex
update ms

@@
 
-\author{and coauthors!}
+\author{coauthors!}
+
 %
@@
 \begin{abstract}
-  Recent studies of the rotation rates and lithium content of FGK
-  stars in open clusters have revealed how these quantities evolve
-  during the first few billion years of a star's life.
-  %
-  Here we apply this knowledge to stars observed by Kepler in order to
-  derive rotation and lithium-based ages.
+  Recent studies of FGK stars in open clusters have quantitatively
+  shown how stellar rotation rates and lithium content evolve over
+  the first few billion years.
+  %
+  Here we apply this knowledge to stars observed by Kepler to compute
+  rotation and lithium-based ages.
   %
@@
   these rotation-based ages are $\approx$15-20$\times$ more precise
-  than isochrone ages for our sample of FGK stars.
+  than isochrone ages in our sample of FGK stars.
   %
@@
   Kepler field decreases linearly from 3\,Gyr ago to present day,
-  for both known planet hosts as well as the parent stellar sample.
+  for both planet hosts as well as for the overall stellar sample.
   %
@@
   star formation history of our Galaxy, it also bounds the relative
-  balance of young and old planets detectable using both Kepler and
-  also future wide-field transit surveys.
+  balance of young and old planets detectable through
+  transit surveys of nearby stars in the thin disk.
   %
@@
 in exoplanet demographics, which have been shown to depend on both
-stellar metallicity and mass \citet{Petigura_2018}, at levels likely
+stellar metallicity and mass \citep{Petigura_2018}, at levels likely
 stronger than age (at least over gigayear timescales).  However, these
@@
 10)}---\citeauthor{Santos_2021} included in a flag for candidate
-``classical pulsators'' (e.g. RR~Lyrae, Cepheids) and candidate
+``classical pulsators'' (e.g.\ RR~Lyrae, Cepheids) and candidate
 close-in binaries.  Visual inspection of the rotation periods as a
@@
 bona fide objects in these classes, it also selects the overwhelming
-majority of $\lesssim$5\,day bona fide young planets.  While we
+majority of $\lesssim$5\,day bona fide young planets.  In our view this is
+erroneous because they these objects are neither classical pulsators nor
+close binaries.  While we
 propagate this flag into our set of quality flags, we therefore
@@
 
-{\it How do we use the bit-flags?)}
+{\it How do we use the bit-flags}
 An important decision for our subsequent analysis is how to combine
@@
 \label{subsec:plflags}
-For the subset of our analysis that focuses on the properties of
-planets, we use the following additional quality indicators to assess 
-the quality of a planet, $Q_{\rm planet}$.
+Not all planets are born equal.
+We use the following additional quality indicators to assess 
+the reliability and utility of a planet, and assemble them into a separate bit-flag,
+$Q_{\rm planet}$.
 
@@
 % disposition score.
-{\it Candidate reliability (bit 0)}---no false positives; only
-``planet candidates'' and ``confirmed planets''.
-
-{\it Candidate S/N (bit 1)}---the false positive rate increases
-greatly toward the noise floor for planet detection in any transit survey \citep[e.g.][]{2002ApJ...564..495J}.  We
+{\it Candidate reliability (bit 0)}---The NEA's Kepler Objects of Interest Table
+includes both an overall planet-candidate disposition status
+({\it koi\_disposition}), as well as a disposition based only on the Kepler
+data ({\it koi\_pdisposition}).
+We require both to include only
+``planet candidates'' and ``confirmed planets''; we are generally less interested
+in the age distribution of planet candidates that are known false positives.
+
+{\it Candidate S/N (bit 1)}---In any transit survey, the false positive rate increases
+greatly toward the noise floor for planet detection \citep[e.g.][]{2002ApJ...564..495J}.  We
 required a S/N in excess of Kepler's usual 7.1$\sigma$ floor, through a cut on the
-``multiple event statistic'' (MES): we required ${\rm MES}$$>$10.
-
-{\it Grazing planets (bit 2)}---grazing objects ($b$>0.8) often
-provide non-robust planetary sizes due to the size-impact parameter
-degeneracy.  Particularly for large planets, they can also include
-astrophysical false positives at higher rates (due to the same
-degeneracy).  While we include a flag for such objects, we by default
-do not exclude them from our analysis; this is because...{\bf
-justify}.
+maximum
+``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}): we required ${\rm MES}$$>$10.
+
+{\it Grazing planets (bit 2)}---Grazing objects, for which the impact
+parameter $b$ is greater than $1-R_{\rm p}/R_\star$,
+often yield biased planetary parameters
+\citep[e.g.][]{2022AJ....163..111G}.
+Particularly for large planets, they can also include
+astrophysical false positives at higher rates \citep{2016ApJ...822...86M}, which is
+in part due to the size-impact parameter degeneracy.
+We flag planet candidates as potentially grazing if $b>0.8$, 
+using the impact parameters reported in the Markov Chain
+Monte Carlo (MCMC) fits from \citet{Thompson_2018}.
 
@@
 returns finite age posteriors, yields \nuniqstarsantosrotteffcut\
-stars.  Further requiring them to be apparently single, near the main
+stars.  Further requiring the stars to be apparently single, near the main
 sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
@@
 based on the Markov Chain Monte Carlo (MCMC) fits from
-\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets and
-\nplhoststarwgyroagenograzing\ stars, respectively.
+\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets orbiting
+\nplhoststarwgyroagenograzing\ stars.
 
@@
 \texttt{EAGLES} \citep{Jeffries_2023}.  We declare an upper-limit for
-plotting purposes if ${EW}-\sigma_{\rm EW}$$<$10\,m\AA.  However, our
+plotting purposes if the 1$\sigma$ lower limit on the equivalent width
+is below 10\,m\AA.  Our measured
 EWs span -40 to 250\,m\AA; all negative EWs have uncertainties that
@@
 To calculate lithium ages, we use \texttt{EAGLES} (git commit
-\texttt{ac09637}), which uses a similar method as
-\texttt{gyro-interp}, and was calibrated on stars observed by the
+\texttt{ac09637}), which, similar to
+\texttt{gyro-interp}, is based on a purely empirical interpolation
+approach.  \texttt{EAGLES} however was calibrated on lithium measurements of
+stars observed by the
 Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
@@
     probability distributions.
-    Completeness is near unity until $\approx$3\,Gyr
+    Completeness is near unity at t$\lesssim$3\,Gyr
     \citep{2022ApJ...937...94M}.
@@
   \tablecaption{Ages of Kepler planets and planet candidates.  This
-  version of the table is truncated to include the youngest fifty
-  confirmed Kepler planets, and the twenty youngest candidates,
+  version of the table is truncated to include only the youngest confirmed Kepler planets,
   sorted by the minimum of either the rotation or lithium-based
@@
   \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
+  A shell script to decode the $Q_{\rm gyro}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
   \label{tab:planets}}
@@
 % \rule
-  \input{table_subgyr_candidate}
+%  \input{table_subgyr_candidate}
   \enddata
-  \tablecomments{``Confirmed'' planets appear before``candidate'' planets.
+  \tablecomments{``Confirmed'' planets appear in the machine-readable version before``candidate'' planets.
     The bit quality flags for the rotation-based ages,
@@
     Bit 1: $\log g$$<$4.2?
-    Bit 2: $M_{\rm G}$<3.9?
+    Bit 2: $M_{\rm G}$<3.9 or $M_{\rm G}>8.5$?
     Bit 3: In KEBC?
@@
 6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
-\citep{Morton_2016}, inspecting the data validation reports this is
+\citep{2016ApJ...822...86M}, inspecting the data validation reports this is
 somewhat surprising.  The light curve shows greater scatter in-transit

### 2024-05-01 drivers/make_younghighlight_table.py
minor

@@ -17,6 +17,7 @@ COMMENTDICT = {
     'K05245.01': "Cep-Her",
     'K07368.01': "Cep-Her",
     'K06186.01': "Cep-Her",
+    'K07913.01': "Cep-Her",
     'K03876.01': "MELANGE-3",
     'K01838.01': "MELANGE-3",
     'K01833.01': 'Theia-520',

### 2024-05-01 paper/ms.tex
update ms

@@
 \citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
-mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
-We report rotation-based ages for these systems of
-$1419^{+657}_{-353}$\,Myr and $873.75\pm{112}$\,Myr respectively.
-For Kepler-66, $T_{\rm eff}$=5898\,K, and $P_{\rm
-rot}$=10.25$\pm$0.10\,days.
-{\it LGB TODO: check again after dereddening...}
+mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
+ages for these systems of \kepsixsixtgyro\ and \kepsixseventgyro\ for
+Kepler-66 and -67b, respectively.  Kepler-66 ($T_{\rm
+eff}$$\approx$5900\,K, $P_{\rm rot}$$\approx$10.4\,days) has a lower
+precision and a more asymmetric posterior because the spindown rates
+of stars this hot are relatively slow after the first gigayear.
+Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
+rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
+that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
+rotation period to be diagnostic of the stellar age.
 

### 2024-05-01 gyrojo/getters.py

@@ -197,7 +197,7 @@ def get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=0,
         cumulative-KOI table, which are "OK" planet candidates.
 
     "gyro_li" all "flag_is_ok_planetcand" and "flag_is_gyro_applicable" KOIs,
-        with both gyro and optionally lithium results.
+        **with gyro results**, and optionally lithium results.
 
     "allageinfo": all "flag_is_ok_planetcand" KOIs, with any available
         rotation-based or lithium-based age information.

### 2024-05-01 gyrojo/papertools.py

@@ -118,7 +118,7 @@ def cast_to_int_string(value):
     if pd.isna(value):
         return np.nan
     else:
-        return str(int(value))
+        return str(int(np.round(value,0)))
 
 def replace_nan_string(value):
     if pd.isna(value) or 'nan' in str(value).lower():

### 2024-05-02 paper/ms.tex
update ms
w/ good words

@@
 \accepted{---}
-\shorttitle{gyro-johannes}
+\shorttitle{Kepler's Demographic Cliff}
 
@@
 
-%
-%% Early experiments for Kepler rotation samples
-%\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
-%\affiliation{\caltech}
-%
-%% ?
-%%\author{Lynne~A.~Hillenbrand}
-%%\affiliation{\caltech}
-%
-%% HIRES observations and reductions
-%\author[0000-0002-0531-1073]{H. Isaacson}
-%\affiliation{\berkeley}
-%% HIRES observations, reductions, JUMP infrastructure
-%\author[0000-0001-8638-0320]{A. W. Howard}
-%\affiliation{\caltech}
-%%
-%\author{HIRES observers to be named}
-%\affiliation{Various}
-
 % <250 words
 \begin{abstract}
-  Recent studies of FGK stars in open clusters have quantitatively
-  shown how stellar rotation rates and lithium content evolve over
-  the first few billion years.
+  Recent analyses of FGK stars in open clusters have revealed how
+  stellar rotation rates and lithium content evolve over the first few
+  billion years.
   %
   Here we apply this knowledge to stars observed by Kepler to compute
-  rotation and lithium-based ages.
+  ages using rotation and lithium.
   %
@@
   and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is valid;
-  these rotation-based ages are $\approx$15-20$\times$ more precise
-  than isochrone ages in our sample of FGK stars.
-  %
-  The independent lithium and rotation-based ages agree for
-  $\approx$90\% of stars.  However, the timescales of lithium
-  depletion are such that rotation is the more precise age diagnostic
-  for the majority of our stellar sample.
+  gyrochronology is valid; these rotation-based ages are
+  $\approx$15-20$\times$ more precise than isochrone ages in our
+  sample of FGK stars.
+  %
+  The independent lithium age estimates agree with the rotation-based
+  ages for $\approx$90\% of stars.  However, the timescales of lithium
+  depletion are such that rotation-based ages are more precise for
+  most of our stellar sample.
   %
@@
 
-\keywords{Exoplanet evolution (491), Stellar ages (1581)}
+\keywords{Stellar ages (1581), Planet hosting stars (1242), Field
+stars (2103), Exoplanet evolution (491), Milky Way evolution (1052),}
 
@@
 of thousands of worlds orbiting close to their host stars
-\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets have
-ages comparable to that of the solar system.  This fact leaves many
-gaps in our knowledge of the exact dynamical and physical processes
-that produce these objects.  Finding and characterizing young
-($<$1\,Gyr) exoplanets offers one possible workaround.  The reason is
-that processes such as thermal cooling \citep{2007ApJ...659.1661F},
+\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
+billions of years old, similar to the Earth.  This fact leaves many
+gaps in our knowledge of what exact dynamical and physical processes
+produce the observed exoplanet archetypes.    The reason is that
+processes such as thermal cooling \citep{2007ApJ...659.1661F},
 atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
-\citep{2014prpl.conf..595R}, and dynamical instability
-\citep{2017MNRAS.470.1750I} are expected to over $\approx$10\,Myr to
-$\approx$1\,Gyr timescales, and so they may leave observable imprints
-in the properties of young planets.  Individual exemplars include
-V1298~Tau \citep{David_2019}, an adolescent resonant chain of four
-transiting puffy planets, and HIP~67522 \citep{Rizzuto_2020}, a
-Jupiter-sized planet with a Neptunian mass (Thao et al.~submitted).
+\citep{2014prpl.conf..595R}, and dynamical instabilities
+\citep{2017MNRAS.470.1750I} are expected to be most efficient over
+$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
+known exoplanets, these processes are done, and not much is happening.
+
+Finding and characterizing young ($<$1\,Gyr) exoplanets is one 
+possible path toward building a timeline for exoplanet evolution.
+Informative individual young systems include V1298~Tau, an adolescent
+resonant chain of four transiting puffy planets that is a likely
+precusor to the compact multiplanet systems \citep{David_2019}, and
+HIP~67522, a Jupiter-sized planet with a Neptunian mass
+(\citealt{Rizzuto_2020}; Thao et al.~submitted).
 % And perhaps also TOI837, now with a uber massive core!
-Population-level examples include tentative differences in the size
-distribution of young planets relative to their older counterparts
+Population-level analyses have similarly suggested differences in the
+size distribution of young exoplanets relative to their older
+counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 planet, and measure the star's age.  Each problem admits a range of
-solutions.  Here, we will focus on planets previously found using
+solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q}.  In
+this article we will consider planets previously known from
 transits, and infer new stellar ages using rotation and lithium.
-Rotation will provide the most age information for the largest number
-of stars.  Lithium will provide useful information for
-stars younger than a few hundred million years.
-
-To begin, let us imagine that the age distribution of stars within a
-few hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
-\citep[e.g.][]{Nordstrom_2004}.  This approximation is useful because
-it clarifies the rarity of young stars: under these assumptions,
-$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
-have $t$$<$1\,Gyr.  Studies of forming protoplanetary
-\citep{2018A&A...617A..44K}, and of exoplanets evolving shortly after
-disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are therefore
+
+To begin, imagine that the age distribution of stars within a few
+hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
+\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
+rarity of young stars: under these assumptions, $\approx$1\% of nearby
+stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
+Entire fields focused on studying forming protoplanets
+\citep{2018A&A...617A..44K}, and on watching exoplanets evolve shortly
+after disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby
 capped in their maximum achievable sample sizes by the tyranny of the
@@
 
-%TODO mention the RV approach?  at least the Hyades etc work...
-
-Despite the observational restrictions, young close-in planet
-discovery has seen remarkable success over the past decade.  This has
-been enabled by a combination of Kepler, K2, TESS, and Gaia
+Despite the observational challenges, young close-in planet discovery
+has done well over the past decade.  This has been primarily enabled
+by Kepler, K2, TESS, and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
@@
 the idea that finding transiting planets can often be easier than
-finding and characterizing the host stellar association.
-
-Despite recent advances, the main problem with any ``cluster-first''
+finding and characterizing the host stellar association
+\citep[e.g.][]{Tofflemire_2021}.
+
+Despite recent advances, the main problem with the ``cluster-first''
 approach is that only $\approx$1\% of stars within the nearest few
-hundred parsecs can currently be associated with their birth cluster,
-even using the Gaia data
+hundred parsecs are currently associated with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-In other words, the majority of nearby sub-gigayear stars are in the
-field, and are omitted by cluster-only planet searches.  
-%TODO FIXME: is this rhetorical point fair?
+Most nearby sub-gigayear stars are in the field, and are omitted by
+cluster-only planet searches.  
 One can appreciate the problem by querying NASA's Exoplanet Archive
+\citep[NEA;][]{2013PASP..125..989A}
 for transiting planets younger than 1\,Gyr.  Imposing reasonable
-precision requirements, e.g. either $t$$<$1\,Gyr at 2$\sigma$, or else
-requiring ages precise to say S/N$=$3, gives $\approx$50 ``young''
+precision requirements\footnote{For instance $t$$<$1\,Gyr at 2$\sigma$, or else
+ages precise to say S/N$=$3.} gives $\approx$50 ``young''
 transiting planets at the time of writing.  Most of these planets are
 in clusters, and have been discovered by Kepler, K2, or TESS.
-However, combined, these surveys have discovered $\approx$5{,}000
-planets.  Barring a major discovery bias, we would expect an order of
-magnitude more transiting planets to be younger than 1\,Gyr, based on
-our earlier order of magnitude reasoning.
-
-This study aims to resolve two main issues.  First, there is the
+However, combined, those surveys have discovered $\approx$5{,}000
+planets.  Unless there is a major bias against the discovery of
+sub-gigayear old planets (with precisely constrained ages), we would
+expect an order of magnitude more young transiting planets to be
+known.
+
+This study aims to resolve two main questions.  First, there is the
 question of how wrong it is to assume a uniform age distribution for
 stars in the galactic thin disk, particularly young stars.  Second,
-there is the question of which planets from the main Kepler mission
-are actually young.
+where are the missing young transiting planets?  Briefly, we will find
+that uniform is incorrect by a factor of a few, and that the missing
+young transiting planets mostly have imprecise age measurements which
+can be improved.
 
@@
 forgiven for assuming that the ages of the Kepler planets are already
-known.  Of course, they have been studied:
-\citeauthor{Berger_2020b_rpage} and \citeauthor{Petigura_2022} for
-instance calculated isochronal ages for individual stars, which is
-most precise for stars whose luminosities and temperatures separate
-them from the main-sequence.  \citeauthor{David_2021} leveraged both
-isochronal ages from an earlier analogous study
-\citep{Fulton_Petigura_2018_cks_vii}, and, in their core analysis,
-this work used a relative gyrochrone age-ranking approach that sorted
-stars based on whether they rotated faster or slower than a gyrochrone
-age of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These
-studies were conducted in the broader context of investigating trends
-in exoplanet demographics, which have been shown to depend on both
-stellar metallicity and mass \citep{Petigura_2018}, at levels likely
-stronger than age (at least over gigayear timescales).  However, these
-studies offered limited sensitivity to the ages of the youngest
-planets, due to the limitations of isochronal dating.
+known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
+and \citet{Petigura_2022} for instance calculated isochrone ages for
+Kepler Objects of Interest (KOIs).  These ages are most precise for
+stars whose luminosities and temperatures separate them from the
+main-sequence.  \citet{David_2021} leveraged both isochronal ages from
+an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
+used a relative gyrochrone age-ranking approach that sorted stars
+based on whether they rotated faster or slower than a gyrochrone age
+of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
+were conducted in the broader context of investigating trends in
+exoplanet demographics, which also depend on stellar metallicity and
+mass \citep{Petigura_2018}.  These studies offered limited sensitivity
+to the ages of the youngest planets, due to the nature of isochronal
+dating.
 
@@
 (\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
-If we additionally require the planets to be non-grazing ($b<0.8$)
-based on the Markov Chain Monte Carlo (MCMC) fits from
-\citet{Thompson_2018}, we are left with \nplwgyroagenograzing\ planets orbiting
+If we additionally require the planets to be non-grazing ($b<0.8$),
+we are left with \nplwgyroagenograzing\ planets orbiting
 \nplhoststarwgyroagenograzing\ stars.
@@
 
-Figure~\ref{fig:li_vs_teff} shows our measured lithium equivalent
-widths, plotted over the mean isochrone models from
-\texttt{EAGLES} \citep{Jeffries_2023}.  We declare an upper-limit for
-plotting purposes if the 1$\sigma$ lower limit on the equivalent width
-is below 10\,m\AA.  Our measured
-EWs span -40 to 250\,m\AA; all negative EWs have uncertainties that
-are statistically consistent with zero.
+Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
+the lithium 6708\AA\ absorption doublet, plotted over the mean
+isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
+declare an upper-limit for plotting purposes if the 1$\sigma$ lower
+limit on the equivalent width is below 10\,m\AA.  Our measured EWs
+span -40 to 250\,m\AA; all negative EWs have uncertainties that are
+statistically consistent with zero.
 
@@
 mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
-ages for these systems of \kepsixsixtgyro\ and \kepsixseventgyro\ for
+ages for these systems of \kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for
 Kepler-66 and -67b, respectively.  Kepler-66 ($T_{\rm
@@
 precision and a more asymmetric posterior because the spindown rates
-of stars this hot are relatively slow after the first gigayear.
+of stars this hot slow down after the first gigayear.
 Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
@@
 
+We began this article by asking {\it i)} how wrong is the assumption
+of a uniform age distribution for stars in the galactic thin disk?,
+and {\it ii)} why are only $\approx$50 sub-gigayear transiting planets
+known, rather than the $\approx$500 that would be expected under the 
+assumption of a uniform star formation history for
+the $\approx$5{,}000 known planets?
+
+In an attempt to answer these questions, we curated a sample of rotation
+periods, lithium equivalent widths, and stellar temperatures using a combination
+of archival and new data.
+We derived new ages using empirical interpolation-based age scales
+that have recently been constructed using the latest available open cluster data.
+
+Our analysis blindly recovered all 14 known Kepler planets in
+clusters, with ages in agreement with the more precise cluster-based ages.
+Although lithium provided only marginal added information for most of
+the sample, the independent lithium and rotation ages agreed for
+$\approx$90\% of cases.  In particular, we report new ages with strong
+constraints from both rotation and lithium for 14 confirmed
+$<$500\,Myr planets, including five mini-Neptunes, five super-Earths,
+three Earths and sub-Earths, and a Neptune-sized planet.
+
+With respect to our original questions however, we have 
+three main conclusions.
+\begin{enumerate}
+  %FIXME: 120 number
+  \item From the rotation rates of Kepler planet-hosting stars, we
+    have identified  \nplyounggyrotwosigma\ planets younger than
+    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
+    below 1\,Gyr.
+    This implies that there are now $\approx$120-170 known
+    sub-gigayear planets.
+  %
+  \item Analyzing the age distributions of both the Kepler
+    target stars and the known Kepler planet-hosts, we find that
+    the age-distribution of stars in the Kepler field has a
+    demographic cliff.  
+    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
+    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
+    The star formation rate today is \ratiosfr\ times
+    lower than it was three billion years ago.
+    %TODO: quote uncertainty on this number.
+  %
+  \item The implication of the latter age distribution is that rather
+    than expecting $\approx$500 sub-gigayear exoplanets, we should
+    instead expect $\approx$250.
+    We have therefore gone from a factor of ten discrepancy to a
+    factor of at most two.
+    The issue is therefore mostly, although perhaps not entirely, resolved.
+%    Some of the remaining missing young
+%    transiting planets may be in the Kepler data.  Many more are
+%    likely in K2 and TESS.
+\end{enumerate}
+
 % 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
-Our rotation analysis blindly recovered the 14 known Kepler planets in
-clusters (Section~\ref{subsec:clusterplanets}), and provided accurate
-independent ages for their host stars.
-
-While lithium provides only marginal added information for most of
-this sample, we report new robust ages from both rotation and
-lithium for 14 confirmed $<$500\,Myr planets, including five
-mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
-Neptune-sized planet.
 

### 2024-05-02 paper/ms.tex
update ms

@@
 
-%
-% ms specific numbers
-%
-
 %%%%%%%%%
@@
 %%%%%%%%%%%
-\newcommand{\nkoiswithprot}{{$\sim$2{,}000}}
-% number of KOIs in Q1-Q17 DR25 table that are not known false positives that have rotation periods
-\newcommand{\nkoisnofpwithprot}{{$\sim$1{,}000}}
 % fraction of KOIs that are not FPs with B+20 parameters
 \newcommand{\frackoisnofpwithprotwithbtwenty}{{$\approx$92\%}}
-
-%%%%%%%%
-% AGES %
-%%%%%%%%
-\newcommand{\nkepstarswithages}{NNN}
-\newcommand{\nkoisnofpwithages}{NNN}
-\newcommand{\nkoisgoldwithages}{NNN}
 
@@
 %\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
-%Ages of the Kepler Stars and Planets
+%\title{Ages of the Kepler Stars and Planets}
 %\title{A Dearth of Young Stars and Planets in the Kepler Field}
@@
 
-\author{coauthors!}
+\author{coauthors, to include but not limited to
+Hillenbrand, Howard, Isaacson, Palumbo}
 
@@
 \begin{abstract}
-  Recent analyses of FGK stars in open clusters have revealed how
-  stellar rotation rates and lithium content evolve over the first few
-  billion years.
-  %
-  Here we apply this knowledge to stars observed by Kepler to compute
-  ages using rotation and lithium.
-  %
-  Rotation periods are drawn from previous Kepler analyses; lithium is
-  measured from new and archival HIRES spectra.
+  %Recent analyses of FGK stars in open clusters have revealed how
+  %stellar rotation rates and lithium content evolve over the first few
+  %billion years.
+  %
+  Recent analyses of FGK stars in open clusters have helped clarify
+  the precision with which a star's rotation rate, and its lithium
+  content, can be used as empirical indicators for its age.
+  %
+  Here we apply this knowledge to stars observed by Kepler.
+  %
+  Rotation periods are drawn from previous Kepler analyses; lithium
+  is measured from new and archival Keck/HIRES spectra.
   %
@@
   gyrochronology is valid; these rotation-based ages are
-  $\approx$15-20$\times$ more precise than isochrone ages in our
-  sample of FGK stars.
-  %
-  The independent lithium age estimates agree with the rotation-based
-  ages for $\approx$90\% of stars.  However, the timescales of lithium
-  depletion are such that rotation-based ages are more precise for
-  most of our stellar sample.
-  %
-  Our resulting catalog of planetary ages includes
-  \nplyounggyrotwosigma\ planets younger than 1\,Gyr at 2$\sigma$
-  (\nplyounggyrothreesigma\ planets at 3$\sigma$).
-  %
-  More broadly, we find that the age distribution of stars in the
-  Kepler field decreases linearly from 3\,Gyr ago to present day,
-  for both planet hosts as well as for the overall stellar sample.
-  %
-  While this stellar age distribution has implications for the
-  star formation history of our Galaxy, it also bounds the relative
-  balance of young and old planets detectable through
-  transit surveys of nearby stars in the thin disk.
+  roughly an order of magnitude more precise than isochrone ages in
+  this sample of FGK stars.
+  %
+  The independent lithium ages agree with the rotation-based ages for
+  $\approx$90\% of cases, suggesting that most of the rotation-based
+  ages are not only precise, they are also accurate.
+  %However, the timescales of lithium depletion are such that
+  %rotation-based ages are more precise for most of our stellar sample.
+  %
+  Our resulting catalog includes \nplyounggyrotwosigma\ planets
+  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
+  at 3$\sigma$).
+  %
+  More broadly, we find that the star formation rate in the Kepler
+  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr over the
+  past 3\,Gyr, for both planet hosts as well as for the overall
+  stellar sample.
+  %
+  This ``demographic cliff'' has implications for the star formation
+  history of our Galaxy, and it helps clarify the relative balance of
+  young and old planets that have been detected in transit surveys of nearby
+  stars in the Galaxy.
   %
@@
 
-%      and \citet{Petigura_2022} have reported trends between small-planet
-%      occurrence and both stellar metallicity and mass; these parameters
-%      seem, at first glance, to be more important predictors of bulk changes
-%      in the planetary population than age.  Given these known correlations,
-%      it can be useful to consider population models in which variables such
-%      as metallicity, mass, and age are added in succession
-%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
-%      rigorous approach, is to jointly infer the number of planets per star
-%      as a function of both planetary and stellar parameters, the latter
-%      including age \citep[e.g.][]{2023AJ....166..209M}.
-
-%     While the youngest 0.1\% of stars ($\lesssim$10\,million
-%     years; Myr) are perhaps the most constraining for studies of planet
-%     formation, they are extremely challenging targets for close-in
-%     exoplanet detection 
-%     \citep[see e.g.][]{Damasso_2020,Bouma_2020_ptfo,Donati_2020}.  Studies of
-%     the early orbital, structural, and atmospheric evolution of exoplanets
-%     are perhaps most technically feasible for the 
-%     $\approx$1\% of stars
-%     with ages between 10 and 100\,Myr, or perhaps even the 10\% of stars
-%     younger than 1\,Gyr.  However, since stellar ages are roughly
-%     uniformly distributed in the solar neighborhood, there is a limit on
-%     the sample size of young planets that will be detectable in these age
-%     regimes, particularly relative to their older breathren.
-
-%      Theoretical predictions for planet evolution in the Kepler sample are
-%      most important in the context of mini-Neptunes and super-Earths, since
-%      these are the most abundant planets.  Pertinent predictions include:
-%      {\it a)} that mini-Neptunes should cool and contract, roughly
-%      following a $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}; {\it
-%      b)} that large and close-in mini-Neptunes at the edge of the
-%      photoevaporation desert should shrink over the atmospheric loss
-%      timescale, $t_{\rm loss}$ \citep{Owen_Lai_2018}; {\it c)} that the
-%      abundance of 4-8\,R$_\oplus$ planets with masses of 5-30\,M$_\oplus$
-%      should decrease over $t_{\rm loss}$---the upper-envelope of the
-%      mini-Neptune size distribution should corresponding decrease,
-%      e.g.~from $\approx$4.5\,R$_\oplus$ at $\approx$200\,Myr to
-%      $\approx$3.5\,R$_\oplus$ at $\approx$700\,Myr
-%      \citep[e.g.][]{Rogers_2021}; {\it d)} that the super-Earth to
-%      mini-Neptune ratio, $\theta$, should increase over $t_{\rm loss}$
-%      \citep[e.g.][]{Rogers_2021}, with the mini-Neptune abundance
-%      decreasing by at least a factor of two or three, and the super-Earth
-%      abundance increasing in tandem; {\it e)} that the
-%      upper boundary of the super-Earth population should increase over the
-%      same timescale as $\theta$ is increasing, since more massive
-%      super-Earth cores should be able to retain their envelopes for longer
-%      than less massive cores.
-%      
-%      The details of the predictions are flexible, in that they depend on
-%      the distributions of planetary core masses, initial atmospheric mass
-%      fractions, and initial entropies of the atmospheres after disk
-%      dispersal.  For instance, different sources of heating---from the star
-%      \citep{Owen_Wu_2013,Lopez_Fortney_2014,Jin_2014}, planetary interior
-%      \citep{Gupta_2019}, or even giant impacts
-%      \citep{Biersteker_Schlichting_2019}---can produce very different
-%      histories for the size evolution of planets at these early times
-%      \citep[e.g.][]{Owen_2020}.  Different internal compositions also alter
-%      the predicted size evolution \citep[e.g.][]{2024NatAs.tmp...33B}.  The
-%      generic prediction that the ``upper envelope'' of the mini-Neptune
-%      size distribution should shrink due to Kelvin-Helmholtz cooling is
-%      hard to avoid; predictions however concerning the ``radius valley''
-%      and movement of planets across it are dependent on the timescales and
-%      prevalence of atmospheric mass loss.  In specific scenarios for
-%      initial atmospheric accretion \citep{Lee_2022} and early atmospheric
-%      evolution \citep{2024MNRAS.529.2716R} the radius gap might form
-%      without needing to invoke post-disk dispersal mass loss at all.  In
-%      such scenarios, there could be very little evolution of the planetary
-%      size distribution between 1 and 4\,\rearth\ over the first gigayear.
-%      
-%      Can stellar age measurements provide evidence for or against
-%      these different models?  Neither {\it (a)} nor {\it (b)} has yet had a
-%      convincing---or even suggestive--- detection reported (see
-%      \citealt{Petigura_2022}).  While an apparent over-abundance of large
-%      very young planets ({\it (c)}) has been noted by multiple authors
-%      \citep[e.g.][]{2017AJ....153...64M,Bouma_2020}, its presence is almost certainly
-%      exacerbated by a selection effect against sub-Neptune sized planet
-%      detection at $<$100\,Myr \citep{Zhou_2021,Bouma_2022b}.  Recent studies have
-%      nonetheless found that when accounting for completeness, there is
-%      $\approx$3$\sigma$ evidence that hot mini-Neptunes are more prevalent in
-%      $\approx$100-500\,Myr old K2 \citep{2023AJ....166..248C} and TESS
-%      \citep{2024arXiv240303261V} stars than in the older Kepler sample
-%      \citep[e.g.][]{2020AJ....159..248K}.
-%      %TODO: more kepler citations...
-%      
-%      Regarding the observed mini-Neptune to super-Earth ratio, {\it (d)} and {\it
-%      (e)} were explored by \citet{Berger_2020b_rpage} and the twin
-%      \citet{David_2021} and \citet{Sandoval_2021} studies.  Both
-%      \citet{Berger_2020b_rpage} and \citet{Sandoval_2021} reported that $\theta$
-%      increases with time, at a statistical significance of $\approx$2-3$\sigma$.
-%      For \citet{Berger_2020b_rpage}, this conclusion was reached by comparing the
-%      sizes of 85 planets with median isochronal ages younger than 1\,Gyr with a
-%      property-matched sample of planets older than 1\,Gyr.  In
-%      \citet{Sandoval_2021}, the conclusion was reached by computing the ratio while
-%      performing a Monte Carlo re-sampling procedure from the isochronal ages and
-%      planetary sizes computed by \citet{Fulton_Petigura_2018_cks_vii}.  Finally,
-%      effect {\it (e)} was reported by \citet{David_2021}, who found that the average
-%      location of the radius valley shifted from $\approx$1.8\,R$_\oplus$ to
-%      $\approx$2.0\,R$_\oplus$ between 1.8 and 3.2\,Gyr.
-%      
-%      The fact that stellar ages---and especially isochronal ages---are
-%      correlated with stellar masses and metallicities is a source of stress
-%      for many in this subfield.  \citet{Sandoval_2021} for instance noted a
-%      strong trend in $\theta$ as a function of stellar metallicity in
-%      addition to the trend they noted in age, and cautioned that this could
-%      also explain their observations.  Independently, \citet{Petigura_2018}
-%      and \citet{Petigura_2022} have reported trends between small-planet
-%      occurrence and both stellar metallicity and mass; these parameters
-%      seem, at first glance, to be more important predictors of bulk changes
-%      in the planetary population than age.  Given these known correlations,
-%      it can be useful to consider population models in which variables such
-%      as metallicity, mass, and age are added in succession
-%      \citep[e.g.][]{Thorngren_2021}.  An alternative, and perhaps more
-%      rigorous approach, is to jointly infer the number of planets per star
-%      as a function of both planetary and stellar parameters, the latter
-%      including age \citep[e.g.][]{2023AJ....166..209M}.  Such analyses are
-%      most effective when the measurement precision of these different
-%      parameters is as high as possible.
-
 Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
@@
 ages for subsets of the Kepler data
-\citep{Walkowicz_2013,Reinhold_2015}.
+\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
 There have been major improvements since these pioneering studies in:
 \begin{enumerate}[label={\it \roman*)}]
-  \item The depth of the Kepler Object of Interest catalog, as well as our understanding of which of these objects are
-    false positives;
+  \item The extent of the Kepler Object of Interest catalog, as well
+    as our understanding of which of these objects are false
+    positives;
   %
@@
     \citep{Bouma_2023}.  
-    The new model is broadly consistent with older calibrations
-    \citep[e.g.][]{Mamajek_2008} for Sun-like stars.  However, for
-    late G dwarfs and K dwarfs, this age-calibration differs by
-    factors of a few from previous estimates;
+    For Sun-like stars, the new model is broadly consistent with older
+    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
+    dwarfs and K dwarfs, the age-calibration differs by factors of a
+    few;
   %
@@
     }
-% 
-%   \subfloat{
-%     \includegraphics[width=0.4\textwidth]{st_params_M_G_vs_adopted_Teff.pdf}
-%   }
-%   \vspace{-0.5cm}
     \subfloat{
@@
     }
-%   \subfloat{
-%     \includegraphics[width=0.5\textwidth]{st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf}
-%   }
   \end{center}
@@
   \caption{
-    {\bf The stars.}  Our analysis focuses on stars observed
-    by Kepler with previously reported rotation periods (dark gray
-    points).  About half of these rotators are suitable for
-    gyrochronology (black points), based on factors including their
-    temperatures, surface gravities, (non-)binarity, and proximity to
-    the main sequence.  The full list of criteria is given in
-    Section~\ref{subsec:flags}.  A few percent of the remainder are
-    either ``confirmed'' or ``candidate'' Kepler Objects of Interest
-    that meet an additional set of planetary quality criteria (blue points;
-    Section~\ref{subsec:plflags}).  
-% We independently derive
-%    lithium-based ages for a largely-overlapping setting of KOIs based
-%    on high-resolution spectroscopy.
-    Surface gravities and effective
-    temperatures are primarily from
-    \citet{Berger_2020a_catalog}.
+    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
+    with previously reported rotation periods (dark gray points).
+    About half of these rotators are suitable for gyrochronology
+    (black points), based on factors including their temperatures,
+    surface gravities, non-binarity, and proximity to the main
+    sequence (see Section~\ref{subsec:flags}).   Some are
+    ``confirmed'' or ``candidate'' Kepler Objects of Interest that
+    meet an additional set of planetary quality criteria (blue points;
+    Section~\ref{subsec:plflags}).  Surface gravities and effective
+    temperatures are primarily from \citet{Berger_2020a_catalog}.
   }
@@
     clusters younger than 4\,Gyr.  The displayed points assume a
-    uniform distribution in temperature for visual clarity.
-    The sizes of the points are the same in each panel, so that low
-    apparent density signifies greater dispersion around the mean.
+    uniform distribution in temperature for visual clarity.  The sizes
+    of the points are the same in each panel, so that low apparent
+    density signifies greater dispersion around the mean.
     \label{fig:models}
@@
     }
-    % 
-    %   \vspace{-0.5cm}
     \subfloat{
@@
     Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
-    clusters.  The stellar sample shows only apparently single stars
+    clusters.  The stellar sample includes only apparently single stars
     near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-    3800$<$$T_{\rm eff}$/K$<$6200, that are not flagged as eclipsing
-    binaries.  The planet sample has the same stellar cuts, and shows
-    confirmed and candidate planets from the cumulative KOI table,
-    requiring them to have a MES$>$10.
+    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
+    the same stellar cuts, and include only the confirmed and candidate planets
+    described in Section~\ref{subsec:plflags}.
   }
@@
       for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the
-    ``mean'' isochrones from the same models as in
-    Figure~\ref{fig:models}; the intrinsic dispersion beyond
-    $\approx$1\,Gyr becomes much larger than changes in the mean.  The
-    logarithmic scale highlights the dynamic range of the detections
-    and upper limits.  Some stars with lithium detections do not have
-    detected rotation periods, and vice-versa.
+    Keck/HIRES spectra collected between 2009 and 2024 comprising
+    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
+    isochrones from the same models as in Figure~\ref{fig:models}; the
+    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
+    than changes in the mean.  The logarithmic scale highlights the
+    dynamic range of the detections and upper limits.  Some stars with
+    lithium detections do not have detected rotation periods, and
+    vice-versa.
     \label{fig:li_vs_teff}
@@
 
-%Crossmatching the Q1-Q17 DR25 KOI table against our stars with
-%rotation periods yields \nkoiswithprot\ KOIs with rotation periods, of
-%which \nkoisnofpwithprot\ are not known false postives.  These objects
-%will comprise the core data for our gyrochronology analysis.
-
 \subsection{High Resolution Spectra}
@@
 
-\subsection{Tally}
+\subsection{Sample Summary}
 \label{subsec:tally}
@@
     periods.
-    The darker histograms impose necessary quality cuts on factors
+    The darker histograms impose the necessary quality cuts on factors
     such as binarity, crowding, and the star's evolutionary state
@@
     In detail, these ``histograms'' are resampled from the posterior
-    probability distributions.
+    probability distributions, and so have less resolution at
+    $\lesssim$0.5\,Gyr, when the relative precision of rotation-based
+    ages decreases.
     Completeness is near unity at t$\lesssim$3\,Gyr
@@
 Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
-stars and planet-hosts in the Kepler field.
-We truncated the plots near $3$\,Gyr due to arguments that suggest the
-detectability of rotation signals is near-unity up to this age for
-Sun-like stars, and decreases for older ages
-\citep{2022ApJ...937...94M}.
-Both the stellar and planetary samples point toward a star formation
-rate in the Kepler field that has slowed over the past few gigayears.
-{\it LGB pending words: allude to findings on MW SFH from chemistry,
-isochrones, and asteroseismology}.
-
-% compare against Berger+20, 
-% note their sentence
-% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al.
-(2018) and Pinsonneault et al.
-(2018), as well as the rotation-based ages in Claytor et al.
-(2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al.
-(2019)."
-%
-% also compare against Mor2019 and their literature assessment
-% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al.
-(2015), and Haywood et al.
-(2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation.
-Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly.
-The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al.
-(2002) and Cignoni et al.
-(2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively.
-Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."
+stars and planet-hosts in the Kepler field.  We truncated the plots
+near $3$\,Gyr due to arguments that suggest the detectability of
+rotation signals is near-unity up to this age for Sun-like stars, and
+decreases for older ages \citep{2022ApJ...937...94M}.  Both the
+stellar and planetary samples point toward a star formation rate in
+the Kepler field that has slowed over the past few gigayears.
 
@@
 
-In terms of detected planet counts, our rotation-based ages for
-the Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
-detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
-For the sub-gigayear bin in particular, which holds specific weight in
+In terms of detected planet counts, our rotation-based ages for the
+Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
+detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.  For
+the sub-gigayear bin in particular, which holds specific weight in
 studies of planet evolution, while there are \nplyounggyro\ planets
-with median ages below 1\,Gyr, requiring $t_{\rm
-gyro}$$<$1\,Gyr at 2$\sigma$ confidence yields
-\nplyounggyrotwosigma\ planets orbiting \nplhostsyounggyrotwosigma\
-stars.
-The stellar and planetary properties of this latter sample of young
-planets are given in Table~\ref{tab:planets}.
+with median ages below 1\,Gyr, requiring $t_{\rm gyro}$$<$1\,Gyr at
+2$\sigma$ confidence yields \nplyounggyrotwosigma\ planets orbiting
+\nplhostsyounggyrotwosigma\ stars.  The stellar and planetary
+properties of this latter sample of young planets are given in
+Table~\ref{tab:planets}.
 
 \subsection{Planets Younger Than One Billion Years}
-
-%\begin{figure*}[!t]
-% \begin{center}
-%   \subfloat{
-%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_strictyoung.pdf}
-%   }
-%   \subfloat{
-%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
-%   }
-% 
-%   \subfloat{
-%   \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_anyyoung.pdf}
-%   }
-%   \subfloat{
-%     \includegraphics[width=0.5\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
-%   }
-% \end{center}
-% \vspace{-0.5cm}
-%
-% \caption{
-%   {\bf Radii, orbital periods, and ages of transiting exoplanets}.
-%    Planets younger than 1\,Gyr are emphasized. 
-%    Two possible selection functions are suggested...
-%   \label{fig:rp_period_age_results}
-% }
-%\end{figure*}
 
@@
 properties of at least the most notable members of the set of
-confirmed planets that have well-measured ages.
-We first focus on systems reported to be in the four stellar
-ensembles in the Kepler field that have yielded transiting planets:
-Cep-Her, MELANGE-3, Theia~520, and NGC~6811.
-We then consider the broader set of Kepler planets that are young
-based on the rotational and lithium-based properties of their host
-stars.
+confirmed planets that have well-measured ages.  We first focus on
+systems reported to be in the four stellar ensembles in the Kepler
+field that have yielded transiting planets: Cep-Her, MELANGE-3,
+Theia~520, and NGC~6811.  We then consider the broader set of Kepler
+planets that are young based on the rotational and lithium-based
+properties of their host stars.
 
@@
   \begin{center}
-    %   \subfloat{
-    %     \includegraphics[width=0.78\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_selsnrupper3_strictyoung.pdf}
-    %   }
-    %   
-    %   \subfloat{
-    \includegraphics[width=0.8\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
-    %   }
+    \includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
   \end{center}
@@
     provenances.  Gray points are transiting planets from the NASA
-    Exoplanet Archive older than 1\,Gyr.  If
-  one were to select for precise ages ($t/\sigma_{\rm t}>3$), this would
-    hide $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
-    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-    hosts to have no quality flags raised, and for the planets to
+    Exoplanet Archive older than 1\,Gyr.  If one were to select for
+    precise ages ($t/\sigma_{\rm t}>3$), this would hide $\approx$ten
+    $<$0.5\,Gyr planets, and add $\approx$fifteen 0.5-1\,Gyr planets.
+    For this plot, we require our Kepler stellar hosts to have no
+    quality flags raised, and for the planets to
     similarly no have quality flags raised. 
@@
 lithium-based ages by comparing the point-estimates of the posteriors,
-when available.
-There are three cases of interest:
+when available.  There are three cases of interest:
 
@@
   \item Two-sided posteriors available for both $t_{\rm gyro}$ and
-    $t_{\rm Li}$: if their values were consistent within
-    2$\sigma$, we judged them to be consistent; if they were
-    consistent within 2-3$\sigma$, we suggest them to be ``maybe''
-    consistent.  Otherwise, they are in disagreement.
-\item If $t_{\rm Li}$ is a lower limit, we compare the
-  $t_{\rm Li}$ lower limit with the $1\sigma$ upper limit from $t_{\rm
-    gyro}$; if the two overlapped, we judge the age estimates to be
-    consistent.
-\item If $t_{\rm Li}$ provides a two-sided posterior, and no rotation
-  is reported, we judge the age estimates to be inconsistent.
-  This is because this lithium constraint can only be provided for G and K
-  stars $\lesssim$2\,Gyr old, and Kepler should have been sensitive to
-  the rotation periods of such stars.
+    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
+    judged them to be consistent; if they were consistent within
+    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
+    Otherwise, they are in disagreement.
+    %
+  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
+    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
+    the two overlapped, we judge the age estimates to be consistent.
+    %
+  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
+    rotation is reported, we judge the age estimates to be
+    inconsistent.  This is because this lithium constraint can only be
+    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
+    have been sensitive to the rotation periods of such stars.
 \end{enumerate}
 
-In the entire sample, this yielded 
-\allagesyesconsistent\ consistent cases,
-\allagesmaybeconsistent\ ``maybe'' cases, and
-\allagesnoconsistent\ inconsistent.
-Restricting our attention to the $\lesssim$1\,Gyr stars, for which
-lithium ages are more constraining,
-the respective counts are
-\minageltonegyryesconsistent, \minageltonegyrmaybeconsistent, and
-\minageltonegyrnoconsistent.
-Rephrased, in the overall sample, \fracconsistentallages\%
-of stars have consistent rotation and lithium-based ages
-(\fracpotentiallyconsistentallages\% potentially consistent).
-In the sample of sub-gigayear stars,
-\fracconsistentminageltonegyr\% of the ages are
-consistent (\fracpotentiallyconsistentminageltonegyr\% potentially
-consistent).
-We feel it necessary to distinguish ``maybe'' consistent cases because
-of lingering uncertainties in how the dispersion of the lithium data
-evolve over the first gigayear.
-In Appendix~\ref{app:inconsistent}, we discuss the
+In the entire sample, this yielded \allagesyesconsistent\ consistent
+cases, \allagesmaybeconsistent\ ``maybe'' cases, and
+\allagesnoconsistent\ inconsistent.  Restricting our attention to the
+$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
+the respective counts are \minageltonegyryesconsistent,
+\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
+Rephrased, in the overall sample, \fracconsistentallages\% of stars
+have consistent rotation and lithium-based ages
+(\fracpotentiallyconsistentallages\% potentially consistent).  In the
+sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
+ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
+potentially consistent).  We feel it necessary to distinguish
+``maybe'' consistent cases because of lingering uncertainties in how
+the dispersion of the lithium data evolve over the first gigayear.  In
+Appendix~\ref{app:inconsistent}, we discuss the
 \minageltonegyrnoconsistent\ discrepant sub-gigayear stars in detail.
 
-\subsubsection{Systems with Rotation and Lithium in Agreement}
+\subsubsection{New Young Systems}
 %
@@
 
-\subsubsection{Need spectra}
-Kepler-1938...
-Kepler-762... (a HJ!)
-Kepler-1320
-
-\subsubsection{Other Notables}
-
-Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
-
-Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
-The rotation-based age we find here is higher/lower than previously noted.
-
-Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
-transiting planets at 3.0, 7.8, and 58\,days, has a rare
-architecture.  \citet{2019A&A...624A..15S} conducted a transiting
-timing analysis of the system, confirming Kepler-411d and arguing for
-the presence of a fourth non-transiting planet, Kepler-411e, based on
-the TTVs seen in Kepler-411d and assuming that the system is nearly
-coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
-the \citet{2007ApJ...669.1167B} relation and the measured rotation
-period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
-star's temperature ($\approx$4900\,K) puts it in a regime where the
-\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
-observations for quite some time
-\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
-$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
-
-Kepler-63b is interesting in that it is {\it missing} from our
-catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
-a polar orbit based on both the Rossiter-McLaughlin effect and a
-starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
-eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
-a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
-qualitatively consistent with the estimates from
-\citet{2013ApJ...775...54S}, though with larger uncertainties because
-our age accounts for the intrinsic scatter in the open cluster
-calibration sequences \citep{Bouma_2023}.  
-While the usual catalogs (\citetalias{Santos_2019},
-\citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
-\citetalias{Mazeh_2015}) do not report this star's rotation period.
-\citetalias{Santos_2021} flagged the system as ``potentially'' having
-rotation-based modulation.
-{\bf todo: how many comparable missed cases are there?}.
-%TODO FIXME it's bc you need to do mcquillan 2013....
-%
-%... David+2021 used:
-% Angus+2018... McQuillan+2013... Mazeh+2015... Walkowicz&Basri2013...
-
-\subsection{Evolutionary trends in time}
-
-{\it Kelvin-Helmholtz cooling}
-$R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}
-
-{\it Early carving of the photoevaporation desert}
-\citep{Owen_Lai_2018}
-
-{\it Rapidly decreasing abundance of big puffy planets}
-Kepler not enough?
-
-{\it Mini-Neptunes turning into Super-Earths}
-\citep[e.g.][]{Rogers_2021}
-
-{\it Boundary of the Super-Earth population moving up in time}
-\citep{David_2021}
+%\subsubsection{Need spectra}
+%Kepler-1938...
+%Kepler-762... (a HJ!)
+%Kepler-1320
 
@@
 
-\subsection{``Galactic Archaeology'' Comparison}
+\subsection{The Thin Disk's Star Formation History}
+
 Zasowski et al fitted a chemical-isochrone model to APOGEE, X, Y, Z
@@
 
-\subsection{Isochrone Age Comparison}
-
-\subsection{Lithium Age Comparison}
-
-\subsection{Gyrochronal Age Comparison}
-Searching the literature for gyrochronal analyses of the Kepler field,
-the most relevant studies seemed to be those of
-\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
-\citet{David_2021}.
-
-Also Mathur2023, which does ``magneto-gyrochronology'', including the
-Sph indicator in a model of the ages.
-
-\subsection{Asteroseismic Age Comparison}
-T Ceillier, J Van Saders et al 2016 MNRAS...
+% compare against Berger+20, 
+% note their sentence
+% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al.
+(2018) and Pinsonneault et al.
+(2018), as well as the rotation-based ages in Claytor et al.
+(2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al.
+(2019)."
+%
+% also compare against Mor2019 and their literature assessment
+% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al.
+(2015), and Haywood et al.
+(2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation.
+Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly.
+The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al.
+(2002) and Cignoni et al.
+(2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively.
+Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."
+
+%\subsection{Isochrone Age Comparison}
+
+%\subsection{Lithium Age Comparison}
+
+\subsection{Caveats}
+
+\subsection{Future Directions}
+
+%   \subsubsection{Evolutionary trends in time}
+%   
+%   {\it Kelvin-Helmholtz cooling}
+%   $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}
+%   
+%   {\it Early carving of the photoevaporation desert}
+%   \citep{Owen_Lai_2018}
+%   
+%   {\it Rapidly decreasing abundance of big puffy planets}
+%   Kepler not enough?
+%   
+%   {\it Mini-Neptunes turning into Super-Earths}
+%   \citep[e.g.][]{Rogers_2021}
+%   
+%   {\it Boundary of the Super-Earth population moving up in time}
+%   \citep{David_2021}
 
@@
 
-We began this article by asking {\it i)} how wrong is the assumption
-of a uniform age distribution for stars in the galactic thin disk?,
-and {\it ii)} why are only $\approx$50 sub-gigayear transiting planets
+We began this article by asking two questions: how wrong is the assumption
+of a uniform age distribution for stars in the galactic thin disk?
+And why are only $\approx$50 sub-gigayear transiting planets
 known, rather than the $\approx$500 that would be expected under the 
@@
 We derived new ages using empirical interpolation-based age scales
-that have recently been constructed using the latest available open cluster data.
+that have been calibrated against the latest available open cluster data.
 
@@
 the sample, the independent lithium and rotation ages agreed for
-$\approx$90\% of cases.  In particular, we report new ages with strong
-constraints from both rotation and lithium for 14 confirmed
-$<$500\,Myr planets, including five mini-Neptunes, five super-Earths,
+$\approx$90\% of cases.  In particular, we reported 
+14 confirmed $<$500\,Myr planets
+with strong age
+constraints from both rotation and lithium, including five mini-Neptunes, five super-Earths,
 three Earths and sub-Earths, and a Neptune-sized planet.
 
-With respect to our original questions however, we have 
+With respect to our original questions, we have 
 three main conclusions.
@@
     This implies that there are now $\approx$120-170 known
-    sub-gigayear planets.
-  %
-  \item Analyzing the age distributions of both the Kepler
-    target stars and the known Kepler planet-hosts, we find that
-    the age-distribution of stars in the Kepler field has a
+    sub-gigayear planets with well-constrained ages.
+  %
+  \item The age distributions of both the Kepler
+    target stars and the known Kepler planet-hosts shows a
     demographic cliff.  
@@
     stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
-    The star formation rate today is \ratiosfr\ times
+    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
     lower than it was three billion years ago.
-    %TODO: quote uncertainty on this number.
   %
@@
     factor of at most two.
-    The issue is therefore mostly, although perhaps not entirely, resolved.
+%    The issue is therefore mostly, although perhaps not entirely, resolved.
 %    Some of the remaining missing young
@@
 
+(author contribution statements will go here)
 L.G.B.~conceived the project, collected data, 
-inferred ages, and drafted the manuscript.
+inferred ages, and wrote the manuscript.
 E.K.P.~contributed to early iterations of the rotation-based age analysis.
@@
   Kepler \citep{Borucki10},
-  TESS \citep{ricker_transiting_2015},
-  NGTS \citep{Wheatley_2018}
+  Keck:I (HIRES).
+%  TESS \citep{ricker_transiting_2015},
+%  NGTS \citep{Wheatley_2018}
 }
@@
 
+\section{Notable Individual Systems}
+
+Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
+
+Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
+The rotation-based age we find here is higher/lower than previously noted.
+
+Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
+transiting planets at 3.0, 7.8, and 58\,days, has a rare
+architecture.  \citet{2019A&A...624A..15S} conducted a transiting
+timing analysis of the system, confirming Kepler-411d and arguing for
+the presence of a fourth non-transiting planet, Kepler-411e, based on
+the TTVs seen in Kepler-411d and assuming that the system is nearly
+coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
+the \citet{2007ApJ...669.1167B} relation and the measured rotation
+period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
+star's temperature ($\approx$4900\,K) puts it in a regime where the
+\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
+observations for a while
+\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
+$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
+
+%  Kepler-63b is interesting in that it is {\it missing} from our
+%  catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
+%  a polar orbit based on both the Rossiter-McLaughlin effect and a
+%  starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
+%  eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
+%  a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
+%  qualitatively consistent with the estimates from
+%  \citet{2013ApJ...775...54S}, though with larger uncertainties because
+%  our age accounts for the intrinsic scatter in the open cluster
+%  calibration sequences \citep{Bouma_2023}.  
+%  While the usual catalogs (\citetalias{Santos_2019},
+%  \citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
+%  \citetalias{Mazeh_2015}) do not report this star's rotation period.
+%  \citetalias{Santos_2021} flagged the system as ``potentially'' having
+%  rotation-based modulation.
+
+\section{Comparison Against Previous Literature Age Scales}
+Searching the literature for gyrochronal analyses of the Kepler field,
+the most relevant studies seemed to be those of
+\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
+\citet{David_2021}.
+
+Also Mathur2023, which does ``magneto-gyrochronology'', including the
+Sph indicator in a model of the ages.
+
+\subsection{Asteroseismic Age Comparison}
+T Ceillier, J Van Saders et al 2016 MNRAS...
+
 \section{Figure dump}

### 2024-05-02 gyrojo/plotting.py
hist field: better

@@ -1320,8 +1320,8 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
         return poisson_uncertainties
 
     plt.close("all")
-    set_style('clean')
-    fig, axs = plt.subplots(ncols=3, figsize=(0.9*6, 0.9*2.7),
+    set_style('science')
+    fig, axs = plt.subplots(ncols=3, figsize=(0.9*6, 0.9*2.5),
                             constrained_layout=True)
 
     koi_df = get_koi_data('cumulative-KOI', grazing_is_ok=1)
@@ -1334,20 +1334,20 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
 
     l0_0 = f'{N} with '+'P$_{\mathrm{rot}}$'
     axs[0].hist(mdf.age/1e3, bins=bins, color='C0', histtype='step',
-                weights=np.ones(len(mdf))/len(mdf), zorder=1, label=l0_0,
+                weights=np.ones(len(mdf))/len(mdf), zorder=-5, label=l0_0,
                 alpha=0.4, linewidth=0.5)
 
     N = int(len(mdf[sel_gyro_ok])/10)
     l0_1 = f'{N} gyro applicable'
     heights, bin_edges, _  = axs[0].hist(
         mdf[sel_gyro_ok].age/1e3, bins=bins, color='C0', histtype='step',
-        weights=np.ones(len(mdf[sel_gyro_ok]))/len(mdf[sel_gyro_ok]), zorder=2,
+        weights=np.ones(len(mdf[sel_gyro_ok]))/len(mdf[sel_gyro_ok]), zorder=-3,
         alpha=1, label=l0_1
     )
     l2_1 = f'{N} Kepler stars'
     _  = axs[2].hist(
         mdf[sel_gyro_ok].age/1e3, bins=bins, color='C0', histtype='step',
-        weights=np.ones(len(mdf[sel_gyro_ok]))/len(mdf[sel_gyro_ok]), zorder=99,
+        weights=np.ones(len(mdf[sel_gyro_ok]))/len(mdf[sel_gyro_ok]), zorder=-1,
         alpha=1, label=l2_1
     )
 
@@ -1356,12 +1356,12 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
     axs[0].errorbar(
         bin_centers, heights, yerr=poisson_uncertainties, marker='o',
[…]

### 2024-05-02 paper/ms.tex
update ms

@@
 
-\author{coauthors, to include but not limited to
+\author{coauthors, possibly including but not limited to
 Hillenbrand, Howard, Isaacson, Palumbo}
@@
   Recent analyses of FGK stars in open clusters have helped clarify
-  the precision with which a star's rotation rate, and its lithium
-  content, can be used as empirical indicators for its age.
+  the precision with which a star's rotation rate and lithium content
+  can be used as empirical indicators for its age.
   %
@@
   More broadly, we find that the star formation rate in the Kepler
-  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr over the
-  past 3\,Gyr, for both planet hosts as well as for the overall
+  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
+  past 3\,Gyr, equally for both planet hosts and for the overall
   stellar sample.
   %
-  This ``demographic cliff'' has implications for the star formation
-  history of our Galaxy, and it helps clarify the relative balance of
-  young and old planets that have been detected in transit surveys of nearby
-  stars in the Galaxy.
+  This ``demographic cliff'' is connected to the star formation
+  history of the Galaxy, and it helps clarify the relative balance of
+  young and old planets that have so far been detected using transits.
   %
@@
 gaps in our knowledge of what exact dynamical and physical processes
-produce the observed exoplanet archetypes.    The reason is that
+produce these objects.    The reason is that
 processes such as thermal cooling \citep{2007ApJ...659.1661F},
@@
 $\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
-known exoplanets, these processes are done, and not much is happening.
+known exoplanets, these processes have completed, and not much is happening.
 
@@
 There have been major improvements since these pioneering studies in:
-\begin{enumerate}[label={\it \roman*)}]
+\begin{enumerate}[label={\it \roman*)},leftmargin=*]
   \item The extent of the Kepler Object of Interest catalog, as well
@@
     with previously reported rotation periods (dark gray points).
+    The rotation periods were primarily measured by \citet{Santos_2019} and \citet{Santos_2021}.
     About half of these rotators are suitable for gyrochronology
@@
     Section~\ref{subsec:plflags}).  Surface gravities and effective
-    temperatures are primarily from \citet{Berger_2020a_catalog}.
+    temperatures were derived photometrically by \citet{Berger_2020a_catalog}.
   }
@@
     Points represent $10^4$ draws from model probability distributions
-    that have been fitted to observed rotation periods
+    that have been fitted to rotation periods
     \citep{Bouma_2023} and lithium equivalent widths
@@
     dispersion in rotation and lithium about these mean models, which
-    is what the models fit, sets the theoretical precision limit for
+    is what the models fit, sets the theoretical precision floor for
     both age-dating methods.    Additional sources of uncertainty,
@@
     {\bf Rotation periods for Kepler target stars (left) and known
-    planet hosts (right)}.  The periods are from \citet{Santos_2019}
+    planet hosts (right)}.  As in Figure~\ref{fig:stellarprops}, periods are from \citet{Santos_2019}
     and \citet{Santos_2021} (\citetalias{Santos_2019} and
-    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' from
-    Figure~\ref{fig:models} to the slow rotation sequences of M67,
+    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' 
+    to the slow rotation sequences of M67,
     Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
@@
     described in Section~\ref{subsec:plflags}.
+    An order of magnitude estimate for the number of young planets discovered
+    by Kepler follows by counting planets below the mean rotational isochrones.
   }
@@
       for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024 comprising
+    Keck/HIRES spectra collected between 2009 and 2024 that comprised
     \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
@@
 rotation periods within $\approx$20\% that had been
-excluded\footnote{Excluding stars in this category would impose a bias
-against young close-in planets, due to the natural commensurability
-between the rotation rates of young stars, and Kepler's increased
-sensitivity to short-period objects.  Although some of these KOIs are
-known false positives, our analysis excludes such objects at later
-steps.}%  As one motivating example, this subset of stars includes {\it
+excluded %\footnote{Excluding stars in this category would impose a bias
+%against young close-in planets, due to the natural commensurability
+%between the rotation rates of young stars, and Kepler's increased
+%sensitivity to short-period objects.  Although some of these KOIs are
+%known false positives, our analysis excludes such objects at later
+%steps.}%  As one motivating example, this subset of stars includes {\it
 %e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
@@
 candidate as of 6 June 2023, while also incorporating human-based
-vetting.\footnote{See
-\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
+vetting. %\footnote{See
+%\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
 These planets represent a superset of those in the fully automated
@@
 are M dwarfs with high proper motions.  Performing the same crossmatch
-at 4 arcseconds, only two KOIs (KOI-3993 and KOI-6531), both known
+at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
 false positives, fail to yield Gaia DR3 crossmatches.}, which is a
@@
   \vspace{-0.5cm}
-
   \caption{
@@
 
-\begin{enumerate}
+\begin{enumerate}[leftmargin=*]
   \item Two-sided posteriors available for both $t_{\rm gyro}$ and
@@
 three main conclusions.
-\begin{enumerate}
+\begin{enumerate}[leftmargin=*]
   %FIXME: 120 number

### 2024-05-02 paper/ms.tex
update ms

@@
 
-\author{coauthors, possibly including but not limited to
+\author{coauthors, including but not limited to
 Hillenbrand, Howard, Isaacson, Palumbo}
@@
 \begin{abstract}
-  %Recent analyses of FGK stars in open clusters have revealed how
-  %stellar rotation rates and lithium content evolve over the first few
-  %billion years.
   %
@@
   ages are not only precise, they are also accurate.
-  %However, the timescales of lithium depletion are such that
-  %rotation-based ages are more precise for most of our stellar sample.
   %
@@
 gaps in our knowledge of what exact dynamical and physical processes
-produce these objects.    The reason is that
-processes such as thermal cooling \citep{2007ApJ...659.1661F},
-atmospheric loss \citep{2019AREPS..47...67O}, giant impacts
+produce these objects.    The reason is that processes such as thermal
+cooling \citep{2007ApJ...659.1661F}, atmospheric loss
+\citep{2019AREPS..47...67O}, giant impacts
 \citep{2014prpl.conf..595R}, and dynamical instabilities
@@
 $\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
-known exoplanets, these processes have completed, and not much is happening.
+known exoplanets, most of the action is in the past.
 
 Finding and characterizing young ($<$1\,Gyr) exoplanets is one
-possible path toward building a timeline for exoplanet evolution.
+potential path toward building a timeline for exoplanet evolution.
 Informative individual young systems include V1298~Tau, an adolescent
@@
 HIP~67522, a Jupiter-sized planet with a Neptunian mass
-(\citealt{Rizzuto_2020}; Thao et al.~submitted).
-% And perhaps also TOI837, now with a uber massive core!
-Population-level analyses have similarly suggested differences in the
-size distribution of young exoplanets relative to their older
-counterparts
+(\citealt{Rizzuto_2020}; Thao et al.~submitted).  Population-level
+analyses have similarly suggested differences in the size distribution
+of young exoplanets relative to their older counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q}.  In
-this article we will consider planets previously known from
-transits, and infer new stellar ages using rotation and lithium.
-
-To begin, imagine that the age distribution of stars within a few
-hundred parsecs of the Sun is uniform between 0 and 10\,Gyr
+this article we will consider planets previously known from transits,
+and infer new stellar ages using rotation and lithium.
+
+To begin, imagine that the ages of nearby stars in the Galaxy are
+uniformly distributed between 0 and 10\,Gyr
 \citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
@@
 stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
-Entire fields focused on studying forming protoplanets
-\citep{2018A&A...617A..44K}, and on watching exoplanets evolve shortly
-after disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby
-capped in their maximum achievable sample sizes by the tyranny of the
-Galactic star formation rate.
+Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
+or of exoplanets evolving just after disk dispersal
+\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
+maximum achievable sample sizes by the tyranny of the Galactic star
+formation rate.
 
 Despite the observational challenges, young close-in planet discovery
-has done well over the past decade.  This has been primarily enabled
+has done well over the past decade, primarily due to
 by Kepler, K2, TESS, and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
-The original strategy from around 2010 to 2020 was to focus on stars
-with known ages---obvious members of open clusters---and to search
-them for transiting planets.  The resulting stellar ages are precise
-at the $\approx$10\% level, and planets discovered in this way are
-generally assumed to have the same age as the host cluster.  A more
-recent development, facilitated by the second Gaia data release, is
-the idea that finding transiting planets can often be easier than
+The strategy pursued by most groups during the 2010s was to focus on
+stars with known ages---obvious members of open clusters---and to
+search them for transiting planets.  The resulting stellar ages are
+precise at the $\approx$10\% level, and planets discovered in this way
+are generally assumed to have the same age as the host cluster.  A
+more recent development, facilitated by the second Gaia data release,
+is the idea that finding transiting planets can often be easier than
 finding and characterizing the host stellar association
@@
 
-Despite recent advances, the main problem with the ``cluster-first''
-approach is that only $\approx$1\% of stars within the nearest few
-hundred parsecs are currently associated with their birth cluster
+The main problem with the ``cluster-first'' approach is that most
+nearby sub-gigayear stars are in the field.  Only $\approx$1\% of
+stars within the nearest $\approx$500\,pc are currently associated
+with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-Most nearby sub-gigayear stars are in the field, and are omitted by
-cluster-only planet searches.  
 One can appreciate the problem by querying NASA's Exoplanet Archive
-\citep[NEA;][]{2013PASP..125..989A}
-for transiting planets younger than 1\,Gyr.  Imposing reasonable
-precision requirements\footnote{For instance $t$$<$1\,Gyr at 2$\sigma$, or else
-ages precise to say S/N$=$3.} gives $\approx$50 ``young''
-transiting planets at the time of writing.  Most of these planets are
-in clusters, and have been discovered by Kepler, K2, or TESS.
-However, combined, those surveys have discovered $\approx$5{,}000
-planets.  Unless there is a major bias against the discovery of
-sub-gigayear old planets (with precisely constrained ages), we would
-expect an order of magnitude more young transiting planets to be
-known.
-
-This study aims to resolve two main questions.  First, there is the
-question of how wrong it is to assume a uniform age distribution for
-stars in the galactic thin disk, particularly young stars.  Second,
-where are the missing young transiting planets?  Briefly, we will find
-that uniform is incorrect by a factor of a few, and that the missing
-young transiting planets mostly have imprecise age measurements which
-can be improved.
+\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
+than 1\,Gyr.  Imposing reasonable precision requirements\footnote{For
+instance $t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
+S/N$=$3.} gives $\approx$50 ``young'' transiting planets at the time
+of writing.  Most of these planets are in clusters, and have been
+discovered by Kepler, K2, or TESS.  However, combined, those surveys
+have discovered $\approx$5{,}000 planets.  Unless there is a major
+bias against the discovery of sub-gigayear old planets with precisely
+constrained ages, we would expect an order of magnitude more young
+transiting planets to be known.
+
+This study aims to resolve two main questions.  First, how wrong it is
+to assume a uniform age distribution for stars in the galactic thin
+disk, particularly young stars?  Second, where are the missing young
+transiting planets?  Briefly, we will find that uniform is incorrect
+by a factor of a few, and that the bias against the discovery of young
+transiting plantes has, to date, primarily been in the measurement of
+theirs ages, rather than in the sensitivity of the transit searches.
 
@@
 exoplanet demographics, which also depend on stellar metallicity and
-mass \citep{Petigura_2018}.  These studies offered limited sensitivity
-to the ages of the youngest planets, due to the nature of isochronal
-dating.
+mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
+sensitivity to the ages of the youngest planets, due to the nature of
+isochronal dating.
 
@@
 times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
-clock is quite old
+clock has a rich history
 \citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
@@
 dynamo all evolve
-\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  
-In the context of Kepler,
-empirical models have been previously applied to derive rotation-based
-ages for subsets of the Kepler data
-\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.
-There have been major improvements since these pioneering studies in:
-\begin{enumerate}[label={\it \roman*)},leftmargin=*]
+\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
+context of Kepler, empirical models have been previously applied to
+derive rotation-based ages for subsets of the Kepler data
+\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
+major improvements since these pioneering studies in:
+\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
   \item The extent of the Kepler Object of Interest catalog, as well
@@
     intrinsic dispersion in rotation periods as a function of time
-    \citep{Bouma_2023}.  
-    For Sun-like stars, the new model is broadly consistent with older
-    calibrations \citep[e.g.][]{Mamajek_2008}.  However, for late G
-    dwarfs and K dwarfs, the age-calibration differs by factors of a
-    few;
-  %
-  \item Our ability to identify binary stars, which can potentially
-    problematic for estimates of rotation-based ages.
+    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
+    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
+    However, for late G dwarfs and K dwarfs, the age-calibration
+    differs by factors of a few;
+  %
+  \item Our ability to identify binary stars, which can be problematic
+    for estimates of rotation-based ages.
 \end{enumerate}
@@
 
-{\it How do we use the bit-flags}
+{\it How do we use the bit-flags?}
 An important decision for our subsequent analysis is how to combine
@@
 
-\begin{enumerate}[leftmargin=*]
+\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
   \item Two-sided posteriors available for both $t_{\rm gyro}$ and
@@
 
-With respect to our original questions, we have 
-three main conclusions.
-\begin{enumerate}[leftmargin=*]
+Our 
+three main conclusions with respect to our original questions are as follows.
+\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
   %FIXME: 120 number

### 2024-05-02 paper/ms.tex
update ms

@@
 
-Our 
-three main conclusions with respect to our original questions are as follows.
+Our three main conclusions with respect to our original questions are
+as follows.
+
 \begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
+  %
+  \item The age distributions of both the Kepler target stars and the
+    known Kepler planet-hosts shows a demographic cliff.  There are
+    \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr) stars in the
+    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
+    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
+    three billion years ago.
+  %
+  \item Rather than expecting $\approx$500 exoplanets younger than one
+    billion years, the rotation-based age distribution of stars in the
+    Kepler field implies that we should instead expect $\approx$250.
+  %
   %FIXME: 120 number
-  \item From the rotation rates of Kepler planet-hosting stars, we
-    have identified  \nplyounggyrotwosigma\ planets younger than
-    1\,Gyr at 2$\sigma$, and \nplyounggyro\ planets with median ages
-    below 1\,Gyr.
-    This implies that there are now $\approx$120-170 known
-    sub-gigayear planets with well-constrained ages.
-  %
-  \item The age distributions of both the Kepler
-    target stars and the known Kepler planet-hosts shows a
-    demographic cliff.  
-    There are \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr)
-    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.
-    The star formation rate today is \ratiosfr$\pm$\uncratiosfr\ times
-    lower than it was three billion years ago.
-  %
-  \item The implication of the latter age distribution is that rather
-    than expecting $\approx$500 sub-gigayear exoplanets, we should
-    instead expect $\approx$250.
-    We have therefore gone from a factor of ten discrepancy to a
-    factor of at most two.
-%    The issue is therefore mostly, although perhaps not entirely, resolved.
-%    Some of the remaining missing young
-%    transiting planets may be in the Kepler data.  Many more are
-%    likely in K2 and TESS.
+  \item Using rotation-based ages, we have reported the ages of
+    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
+    and \nplyounggyro\ planets with median ages below 1\,Gyr.
+    Joining these planets against the existing literature yields
+    $\approx$120-170 known sub-gigayear planets with well-constrained
+    ages.  We have therefore gone from a factor of ten discrepancy to
+    a factor of at most two.
 \end{enumerate}
 
-% 4 CepHer, 2 MELANGE3, 6 Theia520, 2 NGC6811.
-
 \acknowledgements
-This work was supported by the 
-Heising-Simons 51~Pegasi~b Fellowship (LGB)
+This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
+(LGB)
 and the Arthur R.~Adams SURF Fellowship (EKP).
@@
   Keck:I (HIRES).
-%  TESS \citep{ricker_transiting_2015},
-%  NGTS \citep{Wheatley_2018}
 }
@@
 \begin{deluxetable*}{llllllllrrrrrrl}
-  %\tabletypesize{\footnotesize}
   \tabletypesize{\scriptsize}
   \tablecaption{Ages of Kepler planets and planet candidates.  This
-  version of the table is truncated to include only the youngest confirmed Kepler planets,
-  sorted by the minimum of either the rotation or lithium-based
-  age.  The
-  full machine-readable table contains ages and age limits for
-  \nnonfopkoissomeageinfo\ non-false positive KOIs with MES$>$10.  
+  version of the table is truncated to include only the youngest
+  confirmed Kepler planets, sorted by the minimum of either the
+  rotation or lithium-based age.  The full machine-readable table
+  contains ages and age limits for \nnonfopkoissomeageinfo\ non-false
+  positive KOIs with MES$>$10.  
   A shell script to decode the $Q_{\rm gyro}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
@@
     Bit 2: Grazing?
-    For a star to have a high likelihood of being ``reliable for gyrochronology'' we suggest $Q_{\rm star}$ bits
-    zero to nine to not be raised, and for a planet to be ``reliabe'', we suggested $Q_{\rm planet}$ to be zero.
+    For a star to have a high likelihood of being ``reliable for
+    gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
+    not be raised, and for a planet to be ``reliabe'', we suggested
+    $Q_{\rm planet}$ to be zero.
   }

### 2024-05-02 paper/ms.tex
update ms

@@
 hosts as young planet hosts. 
+Focusing only on the tails of the distributions (0-0.3\,Gyr and
+2.7-3\,Gyr), the implication is that the formation rate of stars in
+the Kepler field decreased by a factor of \ratiosfr$\pm$\uncratiosfr\
+over the past three billion years.
 
@@
 detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.  For
-the sub-gigayear bin in particular, which holds specific weight in
+the sub-gigayear bin in particular, which holds extra weight in
 studies of planet evolution, while there are \nplyounggyro\ planets
@@
 
+The fact that only $\approx$90\% of the rotation-based and
+lithium-based ages agree
+
+While hopefully clear from our extensive use of quality flags,
+
 \subsection{Future Directions}
@@
 
-We began this article by asking two questions: how wrong is the assumption
-of a uniform age distribution for stars in the galactic thin disk?
-And why are only $\approx$50 sub-gigayear transiting planets
-known, rather than the $\approx$500 that would be expected under the 
-assumption of a uniform star formation history for
-the $\approx$5{,}000 known planets?
-
-In an attempt to answer these questions, we curated a sample of rotation
-periods, lithium equivalent widths, and stellar temperatures using a combination
-of archival and new data.
-We derived new ages using empirical interpolation-based age scales
-that have been calibrated against the latest available open cluster data.
+We began this article by asking two questions: how wrong is the
+assumption of a uniform age distribution for stars in the galactic
+thin disk?  And why are only $\approx$50 sub-gigayear transiting
+planets known, rather than the $\approx$500 that would be expected
+under the assumption of a uniform star formation history for the
+$\approx$5{,}000 known planets?
+
+In an attempt to answer these questions, we curated a sample of
+rotation periods, lithium equivalent widths, and stellar temperatures
+using a combination of archival and new data.  We derived new ages
+using empirical interpolation-based age scales that have been
+calibrated against the latest available open cluster data.
 
 Our analysis blindly recovered all 14 known Kepler planets in
-clusters, with ages in agreement with the more precise cluster-based ages.
-Although lithium provided only marginal added information for most of
-the sample, the independent lithium and rotation ages agreed for
-$\approx$90\% of cases.  In particular, we reported 
-14 confirmed $<$500\,Myr planets
-with strong age
-constraints from both rotation and lithium, including five mini-Neptunes, five super-Earths,
-three Earths and sub-Earths, and a Neptune-sized planet.
+clusters, with ages in agreement with the more precise cluster-based
+ages.  Although lithium provided only marginal added information for
+most of the sample, the independent lithium and rotation ages agreed
+for $\approx$90\% of cases.  In particular, we reported 14 confirmed
+$<$500\,Myr planets with strong age constraints from both rotation and
+lithium, including five mini-Neptunes, five super-Earths, three Earths
+and sub-Earths, and a Neptune-sized planet.
 
@@
 these have finite Li ages, but no rotation age (perhaps b/c of the missing prots...)
+
 Kepler-1943 b
+
  Kepler-101 b
- Kepler-108 b
- Kepler-108 c
+
+\paragraph{Kepler-108 b and Kepler-108 c}---This sytem of two mutually
+inclined giant planets \citep{2017AJ....153...45M} has a host star
+that is entering the red giant branch.  This has been previously noted
+through asteroseismology \citep{2013ApJ...767..127H}, and an analysis
+of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} concurs, with
+$T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
+[Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.
+
   Kepler-63 b
-Kepler-1312 b
-Kepler-1312 c
+
+Kepler-1312 b Kepler-1312 c
+
    Kepler-1 b
@@
 $\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
+
+Kepler~108 was reported by \citet{2020ApJ...903...55P} to be a member
+of NGC6811 based ``on APOGEE radial velocities and Gaia-DR2 proper
+motions''.  The cluster has $\mu_\alpha\approx-3.4$mas/yr and
+$\mu_\delta \approx -8.8$mas/yr.  Kepler~108 has
+$\mu_\alpha=-1.5\pm0.1$mas/yr and $\mu_\delta \approx
+-1.2\pm0.1$mas/yr \citep{Gaia_DR3_2022}.  This is quite discrepant.
 

### 2024-05-06 paper/ms.tex
update ms

@@
 \author[0000-0002-0514-5538]{Luke~G.~Bouma}
-\altaffiliation{51 Pegasi b Fellow}
+%\altaffiliation{51 Pegasi b Fellow}
 \affiliation{\caltech}
+\affiliation{51 Pegasi b Fellow}
 
@@
   roughly an order of magnitude more precise than isochrone ages in
-  this sample of FGK stars.
+  this sample of stars.
   %
@@
   %
-  This ``demographic cliff'' is connected to the star formation
-  history of the Galaxy, and it helps clarify the relative balance of
-  young and old planets that have so far been detected using transits.
+  This ``demographic cliff'' is a known feature of the Galaxy's star formation
+  history, and it helps clarify the relative balance of
+  young and old planets that have been detected to date using transits.
   %
@@
 uniformly distributed between 0 and 10\,Gyr
-\citep[e.g.][]{Nordstrom_2004}.  This approximation clarifies the
+\citep[e.g.~Fig~17 of][]{Nordstrom_2004}.  This approximation clarifies the
 rarity of young stars: under these assumptions, $\approx$1\% of nearby
@@
 Despite the observational challenges, young close-in planet discovery
-has done well over the past decade, primarily due to
-by Kepler, K2, TESS, and Gaia
+has matured as a field over the past decade, primarily due to
+Kepler, K2, TESS, and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
@@
 stars with known ages---obvious members of open clusters---and to
-search them for transiting planets.  The resulting stellar ages are
-precise at the $\approx$10\% level, and planets discovered in this way
-are generally assumed to have the same age as the host cluster.  A
+search them for transiting planets.  The resulting stellar (and assumed planetary) ages are
+precise at the $\approx$10\% level.  A
 more recent development, facilitated by the second Gaia data release,
@@
 The main problem with the ``cluster-first'' approach is that most
-nearby sub-gigayear stars are in the field.  Only $\approx$1\% of
-stars within the nearest $\approx$500\,pc are currently associated
+nearby young stars are in the field.  Only $\approx$1\% of
+stars within the nearest $\approx$500\,pc have been associated
 with their birth cluster
@@
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
-than 1\,Gyr.  Imposing reasonable precision requirements\footnote{For
-instance $t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
-S/N$=$3.} gives $\approx$50 ``young'' transiting planets at the time
-of writing.  Most of these planets are in clusters, and have been
+than 1\,Gyr.  Imposing reasonable precision requirements
+(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
+S/N$=$3) gives $\approx$50 ``young'' transiting planets at the time
+of writing.  Most of these planets are in clusters, and were
 discovered by Kepler, K2, or TESS.  However, combined, those surveys
@@
 bias against the discovery of sub-gigayear old planets with precisely
-constrained ages, we would expect an order of magnitude more young
-transiting planets to be known.
+constrained ages, a constant star formation history implies that
+an order of magnitude more young
+transiting planets should be known.
 
@@
 by a factor of a few, and that the bias against the discovery of young
-transiting plantes has, to date, primarily been in the measurement of
-theirs ages, rather than in the sensitivity of the transit searches.
-
-Given that Kepler's primary mission ended in 2013, one might be
+transiting planets is, primarily, in the measurement of
+theirs ages. %, rather than in the sensitivity of the transit searches.
+
+Given that Kepler's main mission ended in 2013, one might be
 forgiven for assuming that the ages of the Kepler planets are already
@@
 main-sequence.  \citet{David_2021} leveraged both isochronal ages from
-an earlier analogous study \citep{Fulton_Petigura_2018_cks_vii}, and
+an earlier analogous study \citep{Fulton_2018}, and
 used a relative gyrochrone age-ranking approach that sorted stars
@@
 Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
-between 0.5-4\,Gyr, and can also provide useful age limits at earlier
+between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
 times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
@@
 \begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item The extent of the Kepler Object of Interest catalog, as well
-    as our understanding of which of these objects are false
-    positives;
-  %
-  \item Our empirical understanding of how rotation periods of stars
-    in open clusters evolve
+  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
+    positives within it;
+  %
+  \item Available measurements of stellar rotation periods
+    in open clusters
     \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
@@
   \item Our ability to model not only the mean evolution but also the
-    intrinsic dispersion in rotation periods as a function of time
+    intrinsic dispersion in rotation through time
     \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
     consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
-    However, for late G dwarfs and K dwarfs, the age-calibration
+    For late G dwarfs and K dwarfs, the age calibration
     differs by factors of a few;
@@
 \end{enumerate}
-Today, rotation-based ages for FGK stars below $\approx$4\,Gyr are
-empirically tied to the cluster age scale and rotation sequence, and
+Today, FGK star rotation-based ages are
+empirically tied to the cluster age scale, and
 are limited in precision by the intrisic scatter in the $P_{\rm rot}$
@@
     \leavevmode
-    \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
+    \includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
   \end{center}
@@
 samples, we found broad agreement at $>$30\,m\AA, and significant
-differences at $<$30\,m\AA because \citet{2018ApJ...855..115B}
+differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
 required positive EWs, while we allow for negative ones.  At
@@
 where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
-Kepler-to-Gaia crossmatch\footnote{\url{https://gaia-kepler.fun}, last
-accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
-because 91 KOIs are omitted in the 1$''$ match, 23 of which have
-either ``candidate'' or ``confirmed'' status.  Almost all of these 23
-are M dwarfs with high proper motions.  Performing the same crossmatch
-at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
-false positives, fail to yield Gaia DR3 crossmatches.}, which is a
-crossmatch of the NASA Exoplanet Archive
+Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
+%accessed 15 April 2024.  We used 4$''$ rather than the 1$''$ match
+%because 91 KOIs are omitted in the 1$''$ match, 23 of which have
+%either ``candidate'' or ``confirmed'' status.  Almost all of these 23
+%are M dwarfs with high proper motions.  Performing the same crossmatch
+%at 4$''$, only two KOIs (KOI-3993 and KOI-6531), both known
+%false positives, fail to yield Gaia DR3 crossmatches.}, 
+%which is a
+%crossmatch
+of the NASA Exoplanet Archive
 \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
@@
 function of temperature (Figure~\ref{fig:stellarprops}).  This is a
-challenge for rotation-based age-dating because {\it i)} point sources
+challenge for rotation-based ages because {\it i)} point sources
 far from the main-sequence include unresolved photometric binaries,
@@
 rotation periods of bona fide single stars far from the main-sequence.
-The added complications of reddening and metallicity make defining a
-simple cut in any two dimensions challenging.  
-We explored various options, and settled on the orange locus in the
+The added complications of reddening and metallicity further
+complicate matters.  
+After exploring various options, we settled on the orange locus in the
 $\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
@@
 \label{subsec:plflags}
-Not all planets are born equal.
+Some planets are more reliable than others.
 We use the following additional quality indicators to assess 
@@
 
-\begin{figure*}[!t]
+\begin{figure*}[!p]
   \begin{center}
     \leavevmode
-    \includegraphics[width=\textwidth]{hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf}
+    \subfloat{
+    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+    }
+
+	\vspace{-0.3cm}
+    \subfloat{
+    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+    }
   \end{center}
@@
     {\bf Kepler's demographic cliff,} visible in the rotation-derived
-    ages of stars (left) and planet hosts (right) in the Kepler field.
-    The lightly colored histograms correspond to all
-    \nuniqstarsantosrotteffcut\ $T_{\rm eff}\in[3800,6200]\,$K Kepler
-    stars with reported rotation periods, and all
-    \nplhoststarwgyroage\ planet-hosting stars with reported rotation
-    periods.
-    The darker histograms impose the necessary quality cuts on factors
-    such as binarity, crowding, and the star's evolutionary state
-    (Section~\ref{subsec:flags}).
-    In detail, these ``histograms'' are resampled from the posterior
-    probability distributions, and so have less resolution at
-    $\lesssim$0.5\,Gyr, when the relative precision of rotation-based
-    ages decreases.
-    Completeness is near unity at t$\lesssim$3\,Gyr
+    ages of stars (left) and planet hosts (middle) in the Kepler field.
+    The top row corresponds to all
+    stars with $T_{\rm eff}\in[3800,6200]\,$K for which we calculated rotation-based ages.
+    The darker lines impose quality cuts
+    on factors such as binarity, crowding, and the star's evolutionary
+    state.
+    The bottom row selects only stars with $T_{\rm eff}\in[4400,5400]\,$K, which
+    have relatively precise ages near 2\,Gyr due to their fast spin-down
+    \citep[see][]{Bouma_2023}.
+    The statistical uncertainty for a mean individual star at 1, 2, and 3\,Gyr is shown, and is identical across each row.
+    Finally, the right panel compares the rotation-based age distributions against star formation histories derived using CMD fitting.
+    Completeness is near unity until $t$$\lesssim$3\,Gyr
     \citep{2022ApJ...937...94M}.
-    {\bf todo: numbers dont match.  why?}
     \label{fig:hist_tgyro}
@@
 Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
-stars and planet-hosts in the Kepler field.  We truncated the plots
-near $3$\,Gyr due to arguments that suggest the detectability of
-rotation signals is near-unity up to this age for Sun-like stars, and
-decreases for older ages \citep{2022ApJ...937...94M}.  Both the
-stellar and planetary samples point toward a star formation rate in
-the Kepler field that has slowed over the past few gigayears.
-
-Imagine we were to group stars between 0-1\,Gyr, 1-2\,Gyr, and
-2-3\,Gyr, and label them ``young'', ``intermediate-age'', and ``old''.
-A simple counting exercise from the distributions in
+stars and known planet-hosts in the Kepler field.  In detail, these
+``histograms'' are sampled from the posterior probability
+distributions by drawing ten random samples from each posterior
+distribution, and then computing the histogram of the resulting
+samples.  We truncated the plots near $4$\,Gyr since this is the
+cluster age out to which our interpolation-based method is calibrated.
+For Sun-like stars, the fraction of rotation signals detectable by
+Kepler has been found to be near-unity up to $\approx$3\,Gyr, and it
+drops to almost zero by $\approx$5\,Gyr \citep{2022ApJ...937...94M}.
+This trend has the opposite functional form from our derived age
+distributions.
+
+To quantify the amount by which young stars are more rare than old, we
+can group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
+``intermediate-age'', and ``old''.  A simple counting exercise from
 Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
@@
 Similarly, there are \ratioobtoybplanets\ times as many old planet
-hosts as young planet hosts. 
-Focusing only on the tails of the distributions (0-0.3\,Gyr and
-2.7-3\,Gyr), the implication is that the formation rate of stars in
-the Kepler field decreased by a factor of \ratiosfr$\pm$\uncratiosfr\
-over the past three billion years.
+hosts as young planet hosts.  Focusing only on the tails of the
+distributions (0-0.3\,Gyr and 2.7-3\,Gyr), the implication is that the
+formation rate of stars in the Kepler field decreased by a factor of
+\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.
 
@@
 
-Given that the history of young exoplanet discovery includes a
-sociological tendency to report detailed descriptions of specific
-young exoplanets, we find it necessary to briefly discuss the
-properties of at least the most notable members of the set of
-confirmed planets that have well-measured ages.  We first focus on
-systems reported to be in the four stellar ensembles in the Kepler
-field that have yielded transiting planets: Cep-Her, MELANGE-3,
-Theia~520, and NGC~6811.  We then consider the broader set of Kepler
-planets that are young based on the rotational and lithium-based
-properties of their host stars.
+What are the most notable confirmed planets with well-measured ages?
+These planets can be split into two classes:
+those in clusters, and those with independent and consistent
+rotation and lithium-based ages.
 
@@
 
+To date, there are four known stellar ensembles in the Kepler field that have
+yielded a total of fourteen transiting planets.
+Our analysis blindly recovers each of these planets, as well as their ages.
+
 \paragraph{Cep-Her}
-Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab are
-known members of the $\approx$40\,Myr old Cep-Her association
-\citep{Bouma_2022a,Bouma_2022b}.  The age of various sub-components of
-this association are best measured through relative isochronal
-placement of the pre-main-sequence \citep[see][]{Bouma_2022b}, and
-gyrochronology of stars in these clusters show them to be $<$100\,Myr
-old.  The planets themselves are all on close-in orbits (5-25\,days),
-with sizes from 2-4\,$R_\oplus$.  For Kepler-1627A and Kepler-1974,
-the rotation, lithium, and cluster-based ages all agree within
-1$\sigma$.  For Kepler-1643, while the rotation and cluster ages agree
-($46^{+9}_{-7}$\,Myr; \citealt{Bouma_2022b}), the lithium-based age
-($198^{+98}_{-76}$\,Myr) is 2.0$\sigma$ above the nominal cluster age
-for RSG-5.  For Kepler-1975Ab (KOI\,7913Ab), the lithium age may be
-similarly ``old'', though with less confidence due to the large
-intrinsic scatter in K dwarf lithium calibration measurements.  For
-that system, the isochronal age of its binary companion provides a
-more stringent and serendipitous constraint on the $\approx$40\,Myr
-age.
+The $\approx$40\,Myr old Cep-Her complex currently has its most
+constraining age measurement from relative isochrone dating of
+pre-main-sequence M dwarfs, and it contains Kepler-1627Ab,
+Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
+\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
+may be older by $\approx$50\% (\citealt{Bouma_2022b};
+Kerr~et.al.~in preparation).  The planets themselves are all on
+close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
+rotation, lithium, and cluster ages all agree within 1$\sigma$ for
+Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
+rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
+above the nominal cluster age for the RSG-5 subgroup.
 
 \paragraph{MELANGE-3}
-\citet{2022AJ....164...88B} described an association in the Kepler
-field that was argued to be $105\pm10$\,Myr based on a combination of
-rotation, lithium, and the isochronal pre-main-sequence.  Kepler-970
-and Kepler-1928 harbor two transiting mini-Neptunes inside this
+\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
+the Kepler field and derived its age using a combination of rotation,
+lithium, and the isochronal pre-main-sequence.  Kepler-970 and
+Kepler-1928 harbor two transiting mini-Neptunes inside this
 association \citep{2022AJ....164...88B}.  Our rotation-based
@@
 being slightly older than the suggested Pleiades age for the
-assocation agrees with Figure~4 of \citet{2022AJ....164...88B}, where
-the MELANGE-3 rotational isochrone is visually slower than
-the corresponding Pleiades track.
+assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.
 
@@
 Kepler-968 and Kepler-52 both have rotation-based ages of
-330-350\,Myr.  These stars were independently reported by
-\citet{2019AJ....158..122K} to be members of an association dubbed
-Theia~520 based on their spatial and kinematic correlations with
-nearby stars.  While the core of this structure was previously named
-UBC-1 \citep{2018A&A...618A..59C}, both Kepler planet hosts are in
-more diffuse regions of the cluster reported by
-\citet{2019AJ....158..122K}, so we adopt the name from the latter
-study.  The nominal age of Theia~520 reported by
-\citet{2019AJ....158..122K} based on isochrone fitting was
-$247^{+64}_{-51}$\,Myr.  An independent analysis of the cluster's core
-by \citet{2024A&A...681A..13F} using g-mode asteroseismology,
+330-350\,Myr.  These stars were reported by
+\citet{2019AJ....158..122K} to be members of an association called Theia~520
+based on their spatial and kinematic correlations with nearby stars.
+The core of this structure has been independently called UBC-1
+\citep{2018A&A...618A..59C}; since both Kepler planet hosts
+are in diffuse regions not covered in this latter study, we use the name from
+\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting, Theia~520
+is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
+An independent analysis of the cluster's core by
+\citet{2024A&A...681A..13F} using g-mode asteroseismology,
 gyrochronology, and isochrone fitting found consistent results.
-Ongoing spectroscopic and rotation-based studies of the broader
-cluster by J.~Curtis et al.~ is broadly that the two planet-hosting
-stars are indeed members of this diffuse population.  Given the broad
-set of both cluster-level and individual-star evidence, this makes
-Kepler-968 and Kepler-52 the youngest multiplanet systems that have
-currently been identified from the Kepler primary mission.
-We leave detailed discussion of the planet's properties to the latter 
-group.
+Ongoing work by J.~Curtis et al.~broadly finds that the two
+planet-hosting stars are indeed members of this diffuse population.
+Given the cluster-level and individual-star evidence, this makes
+Kepler-968 and Kepler-52 the youngest multiplanet systems currently
+known from the Kepler primary mission.
 
@@
   \begin{center}
-    \includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+    \includegraphics[width=0.85\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
   \end{center}
@@
     provenances.  Gray points are transiting planets from the NASA
-    Exoplanet Archive older than 1\,Gyr.  If one were to select for
+    Exoplanet Archive older than 1\,Gyr.  If one were to instead select for
     precise ages ($t/\sigma_{\rm t}>3$), this would hide $\approx$ten
@@
 
-\subsubsection{Rotation vs.~Lithium Age Comparison}
-
-We judged the consistency of the independent rotation and
-lithium-based ages by comparing the point-estimates of the posteriors,
+\subsubsection{Rotation vs.~Lithium Ages}
+
+We compared the independent rotation and
+lithium-based ages using the point-estimates of the posteriors,
 when available.  There are three cases of interest:
@@
 \begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item Two-sided posteriors available for both $t_{\rm gyro}$ and
+  \item Two-sided posteriors for both $t_{\rm gyro}$ and
     $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
@@
     2-3$\sigma$, we suggest them to be ``maybe'' consistent.
-    Otherwise, they are in disagreement.
-    %
-  \item If $t_{\rm Li}$ is a lower limit, we compare the $t_{\rm Li}$
+    Otherwise, they disagree.
+    %
+  \item If $t_{\rm Li}$ is a lower limit, we compare this
     lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
@@
 Appendix~\ref{app:inconsistent}, we discuss the
-\minageltonegyrnoconsistent\ discrepant sub-gigayear stars in detail.
+\minageltonegyrnoconsistent\ discrepant ``sub-gigayear'' stars in detail.
+{\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
+	M are because rotation is weird}.
 
@@
 
-Zasowski et al fitted a chemical-isochrone model to APOGEE, X, Y, Z
-data and presented a model for the star formation history of the MW.
-The gist is a spike at z=2, and a gradual decrease since then, by
-about a few of ten.
-
-Mor2019 did isochrone fitting to Gaia, and found the same, but with a
-spike at 3Gyr.
-
-Our inferred gyro-age distribution of the Kepler field seems to agree
-with these pictures.  The main value added is likely over the first
-gigayear.
+Most of the stars observed by Kepler are in the Galaxy's thin disk.
+This can be verified by following \citet{Gaia_2018}, and denoting stars with 2D tangential
+velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members, and those
+with 60$<$$v_{\rm T}$$<$150\,\kms as thick disk members.  For all
+stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick disk stars respectively.  
+The thin disk stars have a threefold larger detected rotation period
+fraction: 35{,}675 rotators are in the thin disk, while 7{,}312 are in
+the thick disk.  Further imposing the gyrochronology quality flags
+discussed in Section~\ref{subsec:flags} yields 17{,}755
+thin-disk stars for which gyrochronology is nominally applicable, and
+2{,}462 thick-disk stars.
+
+The kinematic distinction between thin and thick disk stars helps connect
+the Kepler stars to the broader literature focused
+on the history of star formation in the Galaxy.  On extragalactic
+scales, the star formation rate in spiral galaxies with mass
+comparable to the Milky Way peaked $\approx$10\,Gyr ago, and has since
+decreased by about an order of magnitude
+\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
+galaxy, many investigators have worked to measure similar star
+formation histories, through both CMD fitting of resolved stellar
+populations
+\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
+and also by modelling the local white dwarf luminosity function
+\citep[e.g.][]{2019ApJ...878L..11I}.  
+\citet{2019A&A...624L...1M} and \citet{2019ApJ...878L..11I} for
+instance focused on stars (and white dwarfs) within a few hundred
+parsecs, and both reported a local maximum in the galactic star
+formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
+stars in a wider 2\,kpc bubble, and reported {\it three} local maxima
+in the thin disk's star formation history 5.7, 1.9, and 1.0\,Gyr ago,
+which they associate with pericenter passages of the Sagittarius satellite
+galaxy.
+
+The right column of Figure~\ref{fig:hist_tgyro} compares our derived age distribution
+against star formation histories (SFHs) reported by
+\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
+CMD fitting.
+The overall slope of both SFHs broadly agrees with what we find from rotation-based ages.
+The \citet{2019A&A...624L...1M} data in particular seem entirely consistent, after
+accounting for the statistical uncertainties reported in that study.
+Regarding the episodic star formation bursts reported by
+\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK star sample (top row).
+However, rotation-based ages are most precise for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
+\citep{Bouma_2023}.
+The lower row of Figure~\ref{fig:hist_tgyro} selects these stars with a temperature cut, and
+may show a hint of the $\approx$2\,Gyr spike reported by 
+\citet{2020NatAs...4..965R}.
+The $\approx$1\,Gyr spike however is not recovered.
 
@@
 
-\subsection{Caveats}
-
-The fact that only $\approx$90\% of the rotation-based and
-lithium-based ages agree
-
-While hopefully clear from our extensive use of quality flags,
+\subsection{Caveats, Limitations}
+
+{\it Age Accuracy}: The fact that only $\approx$90\% of the
+rotation-based and lithium-based ages agree suggests that in cases of
+disagreement, either one or both could be inaccurate.
+
+{\it Old Ages}: $>$4\,Gyr problematic for gyro models, as calibrated.  Adding
+$\log g$, and perhaps metallicity, could be two useful additional axes
+along which to explore empirical gyro/isochrone fitting.
 
 \subsection{Future Directions}
-
-%   \subsubsection{Evolutionary trends in time}
-%   
-%   {\it Kelvin-Helmholtz cooling}
-%   $R_p \propto t^{-0.1}$ scaling \citep{Gupta_2019}
+{\it Occurrence Rates}: Some interesting
+age-evolution trends have been predicted, and can probably be teased
+from our data.  
+Kelvin-Helmholtz cooling for mini-Neptunes should scale $R_{\rm p}$ as
+$t^{-0.1}$ \citep{Gupta_2019}.
+The photoevaporation desert has been predicted to be ``carved'' over the timescale
+for atmospheric loss \citep{Owen_Lai_2018}.
+Mini-Neptunes should be converted into super-Earths \citep{Rogers_2021}.
+And the boundary of the super-Earth population may correspondingly move up in size over
+time \citep{David_2021}.
+I'd prefer to split the two papers.
+
+{\it Field star ages from other surveys}:
+In this study we have focused on deriving rotation and lithium-based ages for the Kepler
+planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
+Rotation-based age dating of their discoveries might be useful for the effort
+of identifying young exoplanets, and also for the broader issue of understanding
+their evolution.
+
+{\it Searches for new associations}:
+Some of the youngest systems in Table~\ref{tab:planets} have been
+connected to their birth clusters.
+Some have not.
+A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
+yet to be performed.
+The ideal approach for such a search would be to combine spatial and kinematic information
+from Gaia with rotation measurements from TESS.
+
+%   {\it }
+%   
 %   
@@
     three billion years ago.
+    This star formation history broadly agrees with those derived
+    from CMD fitting and white-dwarf chronology, though with
+    important nuances.
   %
@@
 
-\clearpage
+%\clearpage
 
@@
 observations for a while
-\citep[e.g.][Fig.~9]{2008ApJ...687.1264M}.  Our adopted age of
+\citep[e.g.][Fig.~9]{Mamajek_2008}.  Our adopted age of
 $\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
@@
 
-\section{Figure dump}
-
-\begin{figure*}[!t]
-  \begin{center}
-    \leavevmode
-    \includegraphics[width=0.5\textwidth]{gyroage_vs_teff_errs_showplanets_linear.png}
-  \end{center}
-  \vspace{-0.6cm}
-  \caption{
-    {\bf Gyro-age vs Teff}.
-    $\pm$1$\sigma$ uncertainties are plotted around the median for Kepler stars (black) and KOIs (blue).
-    Visual pileup on the left is b/c i) uncertain ages, and ii) lots of stars there.  Precision-land is the other portion.
-    \label{fig:gyroage_vs_teff}
-  }
-\end{figure*}
+%\section{Figure dump}
+%
+%\begin{figure*}[!t]
+%  \begin{center}
+%    \leavevmode
+%    \includegraphics[width=0.5\textwidth]{gyroage_vs_teff_errs_showplanets_linear.png}
+%  \end{center}
+%  \vspace{-0.6cm}
+%  \caption{
+%    {\bf Gyro-age vs Teff}.
+%    $\pm$1$\sigma$ uncertainties are plotted around the median for Kepler stars (black) and KOIs (blue).
+%    Visual pileup on the left is b/c i) uncertain ages, and ii) lots of stars there.  Precision-land is the other portion.
+%    \label{fig:gyroage_vs_teff}
+%  }
+%\end{figure*}
 

### 2024-05-06 paper/collect.sh

@@ -19,7 +19,8 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
 
 # plot_hist_field_gyro_ages 
-cp $r/hist_field_gyro_ages_20240430/hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage3200.pdf .
+cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf .
+cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf .
 
 # plot_li_vs_teff
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .

### 2024-05-06 drivers/make_younghighlight_table.py
consistency check: what if require Q==0?

@@ -266,8 +266,14 @@ def make_table(
 
     print("Are gyro and lithium consistent?")
 
-    sample = ['allages', 'minageltonegyr']
-    dfs = [pdf, pdf[pdf.min_age < 1e3]]
+    sample = ['allages',
+              'minageltonegyr',
+              'allagesqflag',
+              'minageltonegyrqflag']
+    dfs = [pdf,
+           pdf[pdf.min_age < 1e3],
+           pdf[pdf[r"$Q_{\rm gyro}$"]=='0'],
+           pdf[(pdf.min_age < 1e3)&(pdf[r"$Q_{\rm gyro}$"]=='0')]]
 
     for s, _df in zip(sample, dfs):
 
@@ -325,7 +331,7 @@ def make_table(
     pdf = pdf.drop(columns=['li_eagles_limlo', 'li_eagles_limlo_formatted',
                             'li_eagles_limlo_forsort', 'min_age'])
 
-    if SELECT_YOUNG:
+    if not SELECT_YOUNG:
         outtxt = join(TABLEDIR, 'inconsistent_tli_tgyro.txt')
         with open(outtxt, 'w') as f:
             f.writelines(

### 2024-05-06 paper/ms.tex
update ms

@@
 \paragraph{Cep-Her}
-The $\approx$40\,Myr old Cep-Her complex currently has its most
+The $\approx$40\,Myr Cep-Her complex currently has its most
 constraining age measurement from relative isochrone dating of
-pre-main-sequence M dwarfs, and it contains Kepler-1627Ab,
+pre-main-sequence M dwarfs; it contains Kepler-1627Ab,
 Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
 \citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
-may be older by $\approx$50\% (\citealt{Bouma_2022b};
+may vary in age by $\approx$50\% (\citealt{Bouma_2022b};
 Kerr~et.al.~in preparation).  The planets themselves are all on
 close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
-rotation, lithium, and cluster ages all agree within 1$\sigma$ for
+rotation, lithium, and cluster ages all agree for
 Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
 rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the nominal cluster age for the RSG-5 subgroup.
+above the cluster age reported by \citet{Bouma_2022b} for the RSG-5 subgroup.
 
@@
 \citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
-the Kepler field and derived its age using a combination of rotation,
+the Kepler field and derived its age using rotation,
 lithium, and the isochronal pre-main-sequence.  Kepler-970 and
@@
 gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
-Kepler-970 and -1928 respectively.  The former rotation-based age
+Kepler-970 and Kepler-1928 respectively.  The former rotation-based age
 being slightly older than the suggested Pleiades age for the
@@
 \paragraph{Theia~520}
-Kepler-968 and Kepler-52 both have rotation-based ages of
-330-350\,Myr.  These stars were reported by
+We derive rotation-based ages of 330-350\,Myr for Kepler-968 and Kepler-52.
+These stars were reported by
 \citet{2019AJ....158..122K} to be members of an association called Theia~520
 based on their spatial and kinematic correlations with nearby stars.
-The core of this structure has been independently called UBC-1
+The core of this structure is also known as UBC-1
 \citep{2018A&A...618A..59C}; since both Kepler planet hosts
-are in diffuse regions not covered in this latter study, we use the name from
+are in diffuse regions of the cluster, we use the name from
 \citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting, Theia~520
 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
-An independent analysis of the cluster's core by
+An independent analysis by
 \citet{2024A&A...681A..13F} using g-mode asteroseismology,
@@
 Ongoing work by J.~Curtis et al.~broadly finds that the two
-planet-hosting stars are indeed members of this diffuse population.
+planet-hosting stars are indeed members of the diffuse population.
 Given the cluster-level and individual-star evidence, this makes
 Kepler-968 and Kepler-52 the youngest multiplanet systems currently
-known from the Kepler primary mission.
+known from the primary Kepler mission.
 
@@
 
-In the entire sample, this yielded \allagesyesconsistent\ consistent
-cases, \allagesmaybeconsistent\ ``maybe'' cases, and
+In the entire sample (i.e.~without imposing an quality cuts)
+this yielded \allagesyesconsistent\ consistent
+cases, \allagesmaybeconsistent\ ``maybe consistent'' cases, and
 \allagesnoconsistent\ inconsistent.  Restricting our attention to the
@@
 ``maybe'' consistent cases because of lingering uncertainties in how
-the dispersion of the lithium data evolve over the first gigayear.  In
+the dispersion of the lithium data evolve over the first gigayear.
+
+In
 Appendix~\ref{app:inconsistent}, we discuss the
-\minageltonegyrnoconsistent\ discrepant ``sub-gigayear'' stars in detail.
+systems for which at least one age indicator
+reports a sub-gigayear age in detail.
+All but one are simply cases for which the star
+is either evolved or else an unresolved binary.
+The one exception is Kepler-786 (KOI-1199), an early K dwarf with 
+$t_{\rm Li}$=\kepseveneightsix\,Myr, which would predict an obvious rotation signal.
+No photometric signal is detected; other age indicators in the
+spectrum similarly suggest that the star may not in fact be young.
+
 {\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
@@
 \appendix
-\section{Cases With Inconsistent Rotation and Lithium-Based Ages}
+\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
 \label{app:inconsistent}
 
-For Kepler-957, $T_{\rm eff}$=4947\,K, and $P_{\rm rot}$=4.52\,days,
-This implies a rotation-based age of $96_{-72}^{+192}$\,Myr.  The
-rotational broadening of the spectrum ($v\sin
-i$$\approx$9.8$\pm$1.0\kms) is consistent with the nominal stellar
-rotation period.  However, if this star were $\lesssim$500\,Myr old,
-lithium should have been detectable at $>$50\,m\AA.  Lithium is
-marginally if at all present in the spectrum ({\bf  EW$_{\rm
-Li}$$<$XX\,m\AA at 2$\sigma$ }).  While the planet (nominally
-6.9\,$R_\oplus$ on a 5.9\,day orbit) has been statistically validated
-\citep{2016ApJ...822...86M}, inspecting the data validation reports this is
-somewhat surprising.  The light curve shows greater scatter in-transit
-than out of transit, appears near-grazing, and has a suggestion of a
-strong odd-even depth difference.  This is a surprising system that
-merits further attention.
-
-% from inconsitent_tli_tgyro cases:
-Kepler-1644 b
-Kepler-1699 b
- Kepler-786 b
-
-these have finite Li ages, but no rotation age (perhaps b/c of the missing prots...)
-
-Kepler-1943 b
-
- Kepler-101 b
-
-\paragraph{Kepler-108 b and Kepler-108 c}---This sytem of two mutually
-inclined giant planets \citep{2017AJ....153...45M} has a host star
-that is entering the red giant branch.  This has been previously noted
-through asteroseismology \citep{2013ApJ...767..127H}, and an analysis
-of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} concurs, with
+In this section we discuss fifteen systems with confirmed planets that have nominally
+discrepant rotation and lithium-based ages.
+We find that only one of them, Kepler-786, is interesting.
+
+ \paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O}
+has a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot Jupiter,
+and has been studied in detail by multiple investigators
+\citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
+Our lithium-based age of this system
+(\trestwotli\,Myr), derive from EW$_{\rm Li}$=81$\pm$3\,m\AA,
+is qualitatively in line with the previously noted lithium abundance
+\citep{2007ApJ...664.1190S}. 
+However, we detect no photometric rotation signal, and the spectroscopic $v\sin i$ is low.
+The lithium age is strongly asymmetric because
+there is significant amounts of scatter in EW$_{\rm Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G dwarfs.
+We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of 2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.
+The non-detection of rotation, while mildly surprising, is not shocking if the system's true
+age is in this long tail.
+
+\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm Li}$=100$\pm$5\,m\AA,
+nominally implying $t_{\rm Li}$$\approx$200-800\,Myr, but no rotation detection.
+Our automated flags noted that this star has a low surface gravity, high luminosity, and is far from the main sequence.
+\citet{2014A&A...572A...2B} concur: this star is a sub-giant slightly more massive than the Sun;
+no rotation detection is expected; the lithium likely simply survived over the star's
+main sequence lifetime.
+
+\paragraph{Kepler-108}---This system of two mutually
+inclined giant planets \citep{2017AJ....153...45M} has a $T_{\rm eff}$$\approx$5600\,K
+host star with a strong lithium detection, and no rotation detection.
+The host star is entering the red giant branch; this has been previously noted
+through asteroseismology \citep{2013ApJ...767..127H}, and our analysis
+of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} \citep{2017AJ....154..107P} concurs, yielding
 $T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
 [Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.
-
-  Kepler-63 b
-
-Kepler-1312 b Kepler-1312 c
-
-   Kepler-1 b
-
-K02680.01
-K02639.01
+Given the star's high mass, the high lithium content and lack of rotation detection
+is not surprising.
+
+%Independently, Kepler~108 was also reported by \citet{2020ApJ...903...55P} to be a member
+%of NGC6811.  The cluster has $\mu_\alpha\approx-3.4$mas/yr and
+%$\mu_\delta \approx -8.8$mas/yr.  Kepler~108 has
+%$\mu_\alpha=-1.5\pm0.1$\,mas/yr and $\mu_\delta \approx
+%-1.2\pm0.1$\,mas/yr \citep{Gaia_DR3_2022}.  We therefore suggest that it is not a member
+%of NGC6811.
+
+%
+%Our quality flags noted that this star has
+%a low surface gravity, a crowded field, high RUWE, and an ambiguous Kepler-to-Gaia crossmatch.
+
+\paragraph{Kepler-786}---This K3 V star has a surprisingly 
+high lithium content.
+The spectrum is single-lined, with
+ \texttt{SpecMatch-Synth} derived parameters of
+ $T_{\rm eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1,
+ [Fe/H]=0.11$\pm$0.06, and $v\sin i$=2.4$\pm$0.6\kms.
+Yet with 
+EW$_{\rm Li}$=83$\pm$6\,m\AA,
+the lithium age of \kepseveneightsix\,Myr would predict an obvious rotation signal.
+None is present in the Kepler light curve, consistent with the low
+$v\sin i$.
+Out of all ``apparently discrepant'' lithium and rotation measurements
+discussed in this appendix, this is the only one that seems to remain discrepant after scrutiny.
+The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm HK}=-4.7\pm0.5$, which
+suggests an age at least as old as the Hyades
+\citep{2008ApJ...687.1264M}.
+The Balmer lines are in absorption, and display no obvious signatures of youth.
+
+\paragraph{Kepler-1644}---The rotation-based age of this system (\kepsixteenfourfour\,Myr) is nominally
+much lower than the lithium limit ($>$767\,Myr).
+The Kepler light curve shows a $\approx$1\% amplitude 1.4\,day rotation signal with many flares.
+However the automated quality flags note that the star has low (photometric) surface gravity, a high RUWE (${\rm RUWE}_{\rm DR3}$=7.0),
+and that it is far from the main sequence.
+The Keck/HIRES spectrum also shows visually narrow lines, with a
+ \texttt{SpecMatch-Synth} $v \sin i \leq 2$\kms. 
+However, we performed a cross-correlation between the HIRES spectrum and 
+the nearest matches in the Keck/HIRES template library \citep{2015AJ....149...18K}, and found that on 31 July 2022 (UT)
+the system displayed
+a broad CCF with a blended second component at $\approx$$+26$\,\kms relative to
+the primary, with a best-fit flux-ratio of $\approx$3.4\%, and a preferred $T_{\rm eff,B}$$\approx$4400\,K.
+The spectrum and astrometric excess noise therefore point to this system being an unresolved
+binary, which calls the
+reliability of the rotation-based age into question.
+The non-detection of the companion
+from Robo-AO imaging at Palomar \citep{2017AJ....153...66Z} suggests that the companion(s)
+are likely within $\rho$$\lesssim$0.3$''$.
+
+\paragraph{Kepler-1699}---This system is in a similar qualitative regime as Kepler-1644, with an
+apparently young $t_{\rm gyro}$=\kepsixteenninenine\,Myr derived from a $\approx$2\% amplitude
+4.2\,day rotation signal, and no evidence for lithium.
+The system has ${\rm RUWE}_{\rm DR3}$=19.1.
+From the same style of CCF analysis from the Keck/HIRES spectrum acquired on
+31 Aug 2022 (UT), we also find a double-peaked CCF, in this case with a secondary
+component at $\approx$$-16$\,\kms relative to
+the primary with $T_{\rm eff,B}$$\approx$4900\,K.
+This putative companion is similarly not detected in high-resolution imaging
+ \citep{2017AJ....153...66Z}.
+ The rotation-based age is questionable, given that we do not know which source the
+ rotation signal is from, or whether the stars have interacted.
+
+\paragraph{Kepler-1943}---This system nominally has a \kepnineteenfourthree\,Myr
+lithium age, and no reported rotation detection.
+However, the star is flagged as being over-luminous, low-surface gravity,
+and far from the main sequence.
+In other words, it is a subgiant.
+
+\paragraph{Kepler-639, Kepler-320, Kepler-1719, Kepler-1876, Kepler-1072, Kepler-1743, Kepler-1929, Kepler-1488}---These eight systems all have nominally two-sided lithium age posteriors between 1 and 3\,Gyr, and yet lack rotation period detections.
+All are subgiants, flagged with $Q_{\rm gyro}$ under various combinations of bits 1, 2, and 9.
 
@@
 
-Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
-
-Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
-The rotation-based age we find here is higher/lower than previously noted.
+%Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
+%
+%Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
+%The rotation-based age we find here is higher/lower than previously noted.
 
@@
 $\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
-
-Kepler~108 was reported by \citet{2020ApJ...903...55P} to be a member
-of NGC6811 based ``on APOGEE radial velocities and Gaia-DR2 proper
-motions''.  The cluster has $\mu_\alpha\approx-3.4$mas/yr and
-$\mu_\delta \approx -8.8$mas/yr.  Kepler~108 has
-$\mu_\alpha=-1.5\pm0.1$mas/yr and $\mu_\delta \approx
--1.2\pm0.1$mas/yr \citep{Gaia_DR3_2022}.  This is quite discrepant.
 
@@
 
-\section{Comparison Against Previous Literature Age Scales}
-Searching the literature for gyrochronal analyses of the Kepler field,
-the most relevant studies seemed to be those of
-\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
-\citet{David_2021}.
-
-Also Mathur2023, which does ``magneto-gyrochronology'', including the
-Sph indicator in a model of the ages.
-
-\subsection{Asteroseismic Age Comparison}
-T Ceillier, J Van Saders et al 2016 MNRAS...
+%\section{Comparison Against Previous Literature Age Scales}
+%Searching the literature for gyrochronal analyses of the Kepler field,
+%the most relevant studies seemed to be those of
+%\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
+%\citet{David_2021}.
+%
+%Also Mathur2023, which does ``magneto-gyrochronology'', including the
+%Sph indicator in a model of the ages.
+%
+%
+%\subsection{Asteroseismic Age Comparison}
+%T Ceillier, J Van Saders et al 2016 MNRAS...
 

### 2024-05-06 gyrojo/getters.py
kepler-447 size

@@ -16,6 +16,7 @@ import numpy as np, pandas as pd, matplotlib.pyplot as plt
 from os.path import join
 from glob import glob
 from numpy import array as nparr
+from astropy import units as u
 
 from astroquery.vizier import Vizier
 Vizier.ROW_LIMIT = -1
@@ -443,6 +444,14 @@ def get_age_results(whichtype='gyro', COMPARE_AGE_UNCS=0,
             df.loc[_sel, 'Rp'] = df.loc[_sel, 'koi_prad']
             df.loc[_sel, 'E_Rp'] = df.loc[_sel, 'koi_prad_err1']
             df.loc[_sel, 'e_Rp'] = np.abs(df.loc[_sel, 'koi_prad_err2'])
+
+        # Kepler-447 grazing; adopt Lillo-Box+2015 size.
+        # Bizarre that this thing showed no evidence for rotation
+        # variability...
+        _sel = df.kepoi_name == 'K01800.01'
+        df.loc[_sel, 'Rp'] = (1.65*u.Rjup).to(u.Rearth).value
+        df.loc[_sel, 'E_Rp'] = (0.59*u.Rjup).to(u.Rearth).value
+        df.loc[_sel, 'e_Rp'] = (0.56*u.Rjup).to(u.Rearth).value
 
         a_rp = df['Rp']
         a_rp_err1 = df['E_Rp']

### 2024-05-06 paper/ms.tex
update ms

@@
 
-\subsubsection{Rotation vs.~Lithium Ages}
-
-We compared the independent rotation and
-lithium-based ages using the point-estimates of the posteriors,
-when available.  There are three cases of interest:
+\subsubsection{Comparing Rotation and Lithium Ages}
+
+We compared the independent rotation and lithium-based ages using the
+point-estimates of the posteriors, when available.  There are three
+cases of interest:
 
 \begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item Two-sided posteriors for both $t_{\rm gyro}$ and
-    $t_{\rm Li}$: if their values were consistent within 2$\sigma$, we
-    judged them to be consistent; if they were consistent within
-    2-3$\sigma$, we suggest them to be ``maybe'' consistent.
-    Otherwise, they disagree.
-    %
-  \item If $t_{\rm Li}$ is a lower limit, we compare this
-    lower limit with the $1\sigma$ upper limit from $t_{\rm gyro}$; if
-    the two overlapped, we judge the age estimates to be consistent.
+  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
+    if their values were consistent within 2$\sigma$, we judged them
+    to be consistent; if they were consistent within 2-3$\sigma$, we
+    suggest them to be ``maybe'' consistent.  Otherwise, they
+    disagree.
+    %
+  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
+    with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
+    overlapped, we judge the age estimates to be consistent.
     %
@@
 
-In the entire sample (i.e.~without imposing an quality cuts)
-this yielded \allagesyesconsistent\ consistent
-cases, \allagesmaybeconsistent\ ``maybe consistent'' cases, and
-\allagesnoconsistent\ inconsistent.  Restricting our attention to the
-$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
-the respective counts are \minageltonegyryesconsistent,
-\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
-Rephrased, in the overall sample, \fracconsistentallages\% of stars
-have consistent rotation and lithium-based ages
-(\fracpotentiallyconsistentallages\% potentially consistent).  In the
-sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
-ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
-potentially consistent).  We feel it necessary to distinguish
-``maybe'' consistent cases because of lingering uncertainties in how
-the dispersion of the lithium data evolve over the first gigayear.
-
-In
-Appendix~\ref{app:inconsistent}, we discuss the
-systems for which at least one age indicator
-reports a sub-gigayear age in detail.
-All but one are simply cases for which the star
-is either evolved or else an unresolved binary.
-The one exception is Kepler-786 (KOI-1199), an early K dwarf with 
-$t_{\rm Li}$=\kepseveneightsix\,Myr, which would predict an obvious rotation signal.
-No photometric signal is detected; other age indicators in the
-spectrum similarly suggest that the star may not in fact be young.
-
-{\bf FIXME BRIEF SUMMARY, e.g. N of them are because lithium is garbage,
-	M are because rotation is weird}.
+In the entire sample (i.e.~without imposing any quality cuts) this
+yielded \allagesyesconsistent\ consistent cases,
+\allagesmaybeconsistent\ ``maybe consistent'' cases, and
+\allagesnoconsistent\ inconsistent cases.  Rephrased, in the overall
+sample, \fracconsistentallages\% of stars have consistent rotation and
+lithium-based ages (\fracpotentiallyconsistentallages\% potentially
+consistent).
+
+Appendix~\ref{app:inconsistent} describes all cases of ``discrepant''
+confirmed planets for which a sub-gigayear age is reported by at least
+one age indicator.  All but two systems are
+either evolved stars or else unresolved binaries, and are
+automatically flagged as such.  The two exceptions are Kepler-786
+(KOI-1199), and Kepler-1 (TrES-2).  Kepler-786 is an early K dwarf with
+$t_{\rm Li}$=\kepseveneightsix\,Myr, which would imply that
+rotation should be detectable.  However no photometric rotation signal is detected; other age
+indicators in the spectrum similarly suggest that the star may not in
+fact be young.  TrES-2 is therefore a ``lithium-young'' system for
+which we believe the high lithium EW may be
+a statistical fluke.
+
+%Restricting our attention to the
+%$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
+%the respective counts are \minageltonegyryesconsistent,
+%\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
+%Rephrased, in the overall sample, \fracconsistentallages\% of stars
+%have consistent rotation and lithium-based ages
+%(\fracpotentiallyconsistentallages\% potentially consistent).  In the
+%sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
+%ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
+%potentially consistent).  We feel it necessary to distinguish
+%``maybe'' consistent cases because of lingering uncertainties in how
+%the dispersion of the lithium data evolve over the first gigayear.
 
@@
 Whenever both rotation and lithium find a star to be in a
-narrowly-constrained age range, confidence in the true age greatly
-increases, due to the independence of the two measurement techniques.
-The empirical precision for the inferred ages as a function of $P_{\rm
-rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ has been mapped out by
-\citet{Bouma_2023} and \citet{Jeffries_2023} for rotation and lithium
-respectively.  The region of greatest overlap is dictated by the
-timescale for lithium depletion: for stars with temperatures between
-4500-5500\,K with ages $\lesssim$500\,Myr, we can hope to obtain
-independent two-sided constraints.
-
-The youth of most of these systems, to our knowledge, is being
+narrowly-constrained age range, confidence in the true age
+increases, because the two techniques are independent.
+The region of greatest overlap between rotation and lithium ages in
+the spaces of $P_{\rm rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ is
+dictated by the timescale for lithium depletion: for stars with
+temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
+hope to obtain independent two-sided constraints
+\citep{Bouma_2023,Jeffries_2023}.
+
+The following systems, to our knowledge, are being
 well-constrained for the first time in this work.  Here we limit
@@
 mentioned above.
+{\bf FIXME TODO: this section is boring.  come up with a better way to do this,
+and maybe automate}
 
@@
 $304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
-though at half the relative uncertainty.  Based on these lines of
-evidence, this system may be the youngest Earth-sized planet known,
-comparable to systems such as HD~63433d \citep[1.1\,$R_\oplus$,
+though at half the relative uncertainty.  This system may be the
+youngest Earth-sized planet currently known, comparable to systems
+such as HD~63433d \citep[1.1\,$R_\oplus$,
 $414\pm23$\,Myr][]{2024AJ....167...54C}.

### 2024-05-07 paper/ms.tex
update ms

@@
 derive rotation-based ages for subsets of the Kepler data
-\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  There have been
+\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
+have also been reported \citep{2021AJ....161..189L}.
+There have been
 major improvements since these pioneering studies in:
@@
 The region of greatest overlap between rotation and lithium ages in
-the spaces of $P_{\rm rot}$, $EW_{\rm Li}$, and $T_{\rm eff}$ is
+the spaces of $P_{\rm rot}$, EW$_{\rm Li}$, and $T_{\rm eff}$ is
 dictated by the timescale for lithium depletion: for stars with
@@
 mentioned above.
-{\bf FIXME TODO: this section is boring.  come up with a better way to do this,
+{\bf FIXME TODO: next two paragraphs are boring.  come up with a better way to do this,
 and maybe automate}
@@
 
-{\it Age Accuracy}: The fact that only $\approx$90\% of the
+{\it Age Accuracy and Quality Flagging}: The fact that only $\approx$90\% of the
 rotation-based and lithium-based ages agree suggests that in cases of

### 2024-05-07 paper/ms.tex
update ms

@@
 
-{\it Age Accuracy and Quality Flagging}: The fact that only $\approx$90\% of the
-rotation-based and lithium-based ages agree suggests that in cases of
-disagreement, either one or both could be inaccurate.
-
-{\it Old Ages}: $>$4\,Gyr problematic for gyro models, as calibrated.  Adding
-$\log g$, and perhaps metallicity, could be two useful additional axes
-along which to explore empirical gyro/isochrone fitting.
+Most ages reported in this study come from a single source of information: rotation.
+Factors other than stellar age and mass can influence rotation rates of main sequence stars.
+Binarity is the best example of this: even intermediate-separation ($\sim$10\,AU) binaries are biased
+toward rotating anomalously
+rapidly
+\citep{Meibom_2007,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
+Metallicity may also be a second-order effect
+\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
+In this work, we were interested in not only the overall
+age distribution of the Kepler field, but also in the reliability of ages for
+individual young planet hosting stars.
+The quality flagging scheme in Table~\ref{tab:planets} provides one method in support of
+this end.
+The other method is to consider the region in which we do have
+overlapping sensitivity between lithium and rotation -- 
+G and K dwarfs with $t$$\lesssim$1\,Gyr.  
+Our results for such stars include
+\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
+stars for which the two age indicators are both constraining and in agreement
+(with $Q_{\rm gyro}$ bits zero through nine not raised, and no planet quality flags raised).
+Conversely, we found only one system, Kepler-786, with a clearly
+constraining lithium age (\kepseveneightsix\,Myr) and no rotation age.
+
+Other age indicators may help in verifying the ages of
+the $\lesssim$3\,Gyr stars that are the focus of this work.
+Specifically, chromospheric emission in the
+X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
+\citep{2008ApJ...687.1264M,2014MNRAS.441.2361V,2024ApJ...960...62E}.
+However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
+converts kinetic energy into magnetic energy, which is emitted through these chromospheric
+pathways.
+We did not attempt to incorporate these indicators into this study
+due to concerns regarding sensitivity, homogeneity, and
+the question of whether they in fact provide age
+information that is truly independent from rotation.
+
+A broader question, beyond the scope of this work, concerns the ages of
+the older $\gtrsim$3\,Gyr stars that we do not treat.
+For age-dating models based only on rotation rates, or else
+on a combination of rotation and spin-modulation amplitude
+\citep{2023ApJ...952..131M}, an important challenge is that the detectability
+of the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
+is at the limits of the current data \citep{2022ApJ...937...94M}.
+In the ``old star'' regime, methods that use both the evolution of
+stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
+capable of providing useful age constraints for single stars.
+Incorporating kinematics \citep{2021AJ....161..189L} seems likely to remain
+most useful for stellar ensembles, due to the stochastically of kinematic heating.
 

### 2024-05-07 drivers/make_younghighlight_table.py
younghighlight table imrpovements...

@@ -224,29 +224,25 @@ def make_table(
         pdf, [0, 1, 2, 3, 4, 5, 6, 7, 8, 9],
              [0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
     )
-    _sel = (
+    _sel0 = (
         (pdf.min_age < 1000) &
         sel_gyro_quality &
         (pdf.flag_planet_quality=='0') &
         (pdf.koi_disposition == 'CONFIRMED') &
         (pdf.li_eagles_limlo == -1.)
     )
-    print(42*'~')
-    print(pdf[_sel])
-    ulkvp(f'ltonegyrhighqconfirmedtwosided', len(pdf[_sel]))
-    print(42*'~')
-    for k in pdf[_sel].kepoi_name:
+    for k in pdf[_sel0].kepoi_name:
         if k not in COMMENTDICT:
             COMMENTDICT[k] = '\checkmark \checkmark'
 
-    _sel = (
+    _sel1 = (
         (pdf.min_age < 1000) &
         sel_gyro_quality &
         (pdf.flag_planet_quality=='0') &
         (pdf.koi_disposition == 'CONFIRMED') &
         (pdf.li_eagles_limlo > -1.)
     )
-    for k in pdf[_sel].kepoi_name:
+    for k in pdf[_sel1].kepoi_name:
         if k not in COMMENTDICT:
             COMMENTDICT[k] = '\checkmark'
 
@@ -353,7 +349,7 @@ def make_table(
     if SELECT_YOUNG:
         sel = (pdf.koi_disposition == 'CONFIRMED')
         latexpath1 = join(PAPERDIR, 'table_subgyr_confirmed.tex')
-        pdf[sel][mapdict.values()].head(n=87).to_latex(
+        pdf[sel][mapdict.values()].head(n=95).to_latex(
             latexpath1, index=False, na_rep='--', escape=False, formatters=formatters
         )
 
@@ -376,12 +372,40 @@ def make_table(
                             'li_eagles_limlo_forsort', 'min_age'])
 
     if not SELECT_YOUNG:
+
         outtxt = join(TABLEDIR, 'inconsistent_tli_tgyro.txt')
         with open(outtxt, 'w') as f:
             f.writelines(
                 pdf[(pdf.are_gyro_and_li_consistent == 'No')].to_string(max_colwidth=None)
             )
         print(f'Made {outtxt}')
+
[…]

### 2024-05-07 paper/ms.tex
update ms

@@
 
-To date, there are four known stellar ensembles in the Kepler field that have
+There are currently four known stellar ensembles in the Kepler field that have
 yielded a total of fourteen transiting planets.
@@
 dictated by the timescale for lithium depletion: for stars with
-temperatures between 4500-5500\,K with ages $\lesssim$500\,Myr, we can
+temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
 hope to obtain independent two-sided constraints
@@
 formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
-stars in a wider 2\,kpc bubble, and reported {\it three} local maxima
+stars in a wider 2\,kpc bubble, and reported three local maxima
 in the thin disk's star formation history 5.7, 1.9, and 1.0\,Gyr ago,
@@
 X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
-\citep{2008ApJ...687.1264M,2014MNRAS.441.2361V,2024ApJ...960...62E}.
+\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
 However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
@@
 is at the limits of the current data \citep{2022ApJ...937...94M}.
-In the ``old star'' regime, methods that use both the evolution of
+In the ``old star'' regime, methods 
+using either asteroseismology
+\citep{vanSaders_2016,2024ApJ...962..138S}, or else
+using both the evolution of
 stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
@@
 time \citep{David_2021}.
-I'd prefer to split the two papers.
+We're going to split the two papers.
 
 {\it Field star ages from other surveys}:
-In this study we have focused on deriving rotation and lithium-based ages for the Kepler
+In this study we focused on deriving rotation and lithium-based ages for the Kepler
 planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
-Rotation-based age dating of their discoveries might be useful for the effort
+Rotation-based age dating of their discoveries could be useful for the effort
 of identifying young exoplanets, and also for the broader issue of understanding
@@
 connected to their birth clusters.
-Some have not.
+Others have not.
 A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
@@
 from Gaia with rotation measurements from TESS.
-
-%   {\it }
-%   
-%   
-%   {\it Early carving of the photoevaporation desert}
-%   \citep{Owen_Lai_2018}
-%   
-%   {\it Rapidly decreasing abundance of big puffy planets}
-%   Kepler not enough?
-%   
-%   {\it Mini-Neptunes turning into Super-Earths}
-%   \citep[e.g.][]{Rogers_2021}
-%   
-%   {\it Boundary of the Super-Earth population moving up in time}
-%   \citep{David_2021}
+Given that the decoherence time of young clusters is $\sim$100\,Myr, one would expect that
+the youngest ``field stars'' may have young neighbors that are relatively easy to
+identify.
 

### 2024-05-08 paper/ms.tex
update ms

@@
 While the details of how exactly this locus is constructed are
-arbitrary and convoluted,\footnote{In practice, for K dwarfs we fitted an
-$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
-B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
-$t_{\rm B20,iso}$ was the isochronal age reported by
-\citet{Berger_2020a_catalog}. We let the order of the fit vary from
-$N$=1 to 10, minimized the Bayesian Information Criterion,
-and found $N$=6.  For F and G dwarfs, this procedure failed to produce
-a reasonable locus because of shifts in the relative precision of
-the isochrone ages.  We therefore visually selected KOIs that appeared
-to be near the ``main sequence'', fitted a separate polynomial to
-their locations, and then stitched the resulting polynomials from the
-F and G dwarfs to that of the K dwarfs.} the general 
-aim is to exclude stars for which there is photometric evidence
-that gyrochronology may not be reliable.  The rotation period
-distribution of sources excluded in this manner includes a relatively
-larger fraction of ``rapid rotators'' than the main sequence sources,
-likely due to a combination of astrophysical and systematic effects.
+arbitrary, 
+%and convoluted,\footnote{In practice, for K dwarfs we fitted an
+%$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
+%B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
+%$t_{\rm B20,iso}$ was the isochronal age reported by
+%\citet{Berger_2020a_catalog}. We let the order of the fit vary from
+%$N$=1 to 10, minimized the Bayesian Information Criterion,
+%and found $N$=6.  For F and G dwarfs, this procedure failed to produce
+%a reasonable locus because of shifts in the relative precision of
+%the isochrone ages.  We therefore visually selected KOIs that appeared
+%to be near the ``main sequence'', fitted a separate polynomial to
+%their locations, and then stitched the resulting polynomials from the
+%F and G dwarfs to that of the K dwarfs.}
+the general 
+aim is to exclude stars for which there is evidence 
+based on their location in the $\log g$--$T_{\rm eff}$ plane
+that our gyrochronology model may not be reliable.  
+%The rotation period
+%distribution of sources excluded in this manner includes a relatively
+%larger fraction of ``rapid rotators'' than the main sequence sources,
+%likely due to a combination of astrophysical and systematic effects.
 See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
-of \citet{2023ApJS..268....4F}.
+of \citet{2023ApJS..268....4F} for examples of how this period distribution
+of this population differs from stars near the main sequence.
 
@@
 
-\begin{figure*}[!p]
+\begin{figure*}[!t]
   \begin{center}
@@
 
-	\vspace{-0.3cm}
+	\vspace{-0.35cm}
     \subfloat{
@@
   \vspace{-0.6cm}
-  \caption{
-    {\bf Kepler's demographic cliff,} visible in the rotation-derived
+  \caption{{\bf Kepler's demographic cliff,} visible in the rotation-derived
     ages of stars (left) and planet hosts (middle) in the Kepler field.
@@
 
-\begin{figure*}[!t]
+\begin{figure}[!t]
   \begin{center}
-    \includegraphics[width=0.85\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+    \includegraphics[width=0.48\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
   \end{center}
@@
   }
-\end{figure*}
+\end{figure}
 
@@
 %
-Whenever both rotation and lithium find a star to be in a
-narrowly-constrained age range, confidence in the true age
-increases, because the two techniques are independent.
-The region of greatest overlap between rotation and lithium ages in
-the spaces of $P_{\rm rot}$, EW$_{\rm Li}$, and $T_{\rm eff}$ is
-dictated by the timescale for lithium depletion: for stars with
-temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
-hope to obtain independent two-sided constraints
-\citep{Bouma_2023,Jeffries_2023}.
-
-The following systems, to our knowledge, are being
-well-constrained for the first time in this work.  Here we limit
-discussion to systems with $t_{\rm gyro}$$<$500\,Myr not in the clusters
-mentioned above.
-{\bf FIXME TODO: next two paragraphs are boring.  come up with a better way to do this,
-and maybe automate}
-
-The new young planets
-include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
-(Kepler-1529b $\approx$100\,Myr,
--1521b $\approx$195\,Myr,
--1467b $\approx$235\,Myr,
--955b $\approx$400\,Myr, and
--523b $\approx$440\,Myr),
-a single Neptune-sized planet
-(Kepler-971b $\approx$335\,Myr),
-five 1.1-1.8\,$R_\oplus$ super-Earths
-(Kepler-1930b $\approx$150\,Myr,
--1313b $\approx$185\,Myr,
--1565b $\approx$205\,Myr,
--1532b $\approx$400\,Myr,
--1481b $\approx$400\,Myr),
-and three Earth or sub-Earth-sized planets
-(Kepler-1933b $\approx$300\,Myr,
--1561b $\approx$420\,Myr,
--1482b $\approx$490\,Myr).
-All of these planets are on orbital periods below 50\,days.
-
-Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
-1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
-$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
-though at half the relative uncertainty.  This system may be the
-youngest Earth-sized planet currently known, comparable to systems
+%The astrophysical processes behind rotation and lithium evolution are
+%independent.
+%Therefore 
+%whenever the two age indicators agree within a
+%narrowly-constrained range, confidence in the true age
+%increases.
+%The region of greatest overlap between rotation and lithium ages is
+%dictated by the timescale for lithium depletion: for stars with
+%temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
+%hope to obtain independent two-sided constraints
+%\citep{Bouma_2023,Jeffries_2023}.
+
+Let us restrict our attention to blemish-free stars, with blemish-free
+``confirmed'' planets:
+$Q_{\rm gyro}$ must not have any of bits zero through nine
+raised, and $Q_{\rm planet}$ must have no quality flags raised at all.
+Under this constraint,
+our results include \ltonegyrhighqconfirmedtwosided\ systems
+younger than 1\,Gyr with two-sided $t_{\rm gyro}$ and $t_{\rm Li}$,
+and \ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$ provides
+an age constraint while $t_{\rm Li}$ is one-sided.
+%Two of the former systems are in clusters (Kepler-1643 and Kepler-1928);
+%five of the latter systems are in Theia~520.
+
+These planets are shown in Figure~\ref{fig:rp_period_age_results}.
+They include \nearthshighqconfirmed\ Earth-sized or smaller planets,
+no Jupiter-sized or larger planets,
+\nsubsaturnshighqconfirmed\ ``super-Neptune'' ($>$4\,$R_\oplus$) sized
+planets,
+\nminineptuneshighqconfirmed\ mini-Neptunes above the 
+radius valley (as parametrized by \citealt{2018MNRAS.479.4786V}),
+and \nsuperearthshighqconfirmed\ super-Earth sized planets below the radius valley.
+While most of these planets are on orbital periods below 50\,days,
+\nlongperiodhighqconfirmed\ are on more distant orbits.
+
+%The new young planets
+%include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
+%(Kepler-1529b $\approx$100\,Myr,
+%-1521b $\approx$195\,Myr,
+%-1467b $\approx$235\,Myr,
+%-955b $\approx$400\,Myr, and
+%-523b $\approx$440\,Myr),
+%a single Neptune-sized planet
+%(Kepler-971b $\approx$335\,Myr),
+%five 1.1-1.8\,$R_\oplus$ super-Earths
+%(Kepler-1930b $\approx$150\,Myr,
+%-1313b $\approx$185\,Myr,
+%-1565b $\approx$205\,Myr,
+%-1532b $\approx$400\,Myr,
+%-1481b $\approx$400\,Myr),
+%and three Earth or sub-Earth-sized planets
+%(Kepler-1933b $\approx$300\,Myr,
+%-1561b $\approx$420\,Myr,
+%-1482b $\approx$490\,Myr).
+
+A few new highlights from the high quality stars and planets include
+Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.
+Kepler-1529b is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured
+rotation and lithium ages.
+Kepler-1565b is an analogous $\approx$170-230\,Myr super-Earth.
+Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr near-Solar analog with
+two known transiting planets: an Earth-sized planet on a one-day orbit,
+and a mini-Neptune on a five-day orbit.
+The Earth-sized planet, Kepler-1312c, along with the near-identical
+Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), 
+could be the
+youngest Earth-sized planets currently known, comparable to systems
 such as HD~63433d \citep[1.1\,$R_\oplus$,
-$414\pm23$\,Myr][]{2024AJ....167...54C}.
-
-A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
-planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
-find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.
+$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b \citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
+Kepler-1629~b, with a size just two-thirds that of Earth
+(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on rotation
+(\kepsixteentwonine\,Myr).  The lithium age from a reconnaissance 
+TRES spectrum confirms the point.
+
+%Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
+%1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
+%$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
+%though at half the relative uncertainty.
+
+%A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
+%planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
+%find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.
 
@@
 
-Most of the stars observed by Kepler are in the Galaxy's thin disk.
+Most of the stars observed by Kepler are in the thin disk.
 This can be verified by following \citet{Gaia_2018}, and denoting stars with 2D tangential
@@
 
-The kinematic distinction between thin and thick disk stars helps connect
-the Kepler stars to the broader literature focused
+Classifying Kepler stars in this manner helps connect
+them to the broader literature focused
 on the history of star formation in the Galaxy.  On extragalactic
@@
 comparable to the Milky Way peaked $\approx$10\,Gyr ago, and has since
-decreased by about an order of magnitude
+decreased by an order of magnitude
 \citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
@@
 \citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
-and also by modelling the local white dwarf luminosity function
+and by modeling the local white dwarf luminosity function
 \citep[e.g.][]{2019ApJ...878L..11I}.  
@@
 instance focused on stars (and white dwarfs) within a few hundred
-parsecs, and both reported a local maximum in the galactic star
+parsecs, and both found a peak in the local star
 formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
@@
 
-Most ages reported in this study come from a single source of information: rotation.
+Most of our ages come from a single source of information: rotation.
 Factors other than stellar age and mass can influence rotation rates of main sequence stars.
-Binarity is the best example of this: even intermediate-separation ($\sim$10\,AU) binaries are biased
-toward rotating anomalously
-rapidly
-\citep{Meibom_2007,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
+Binarity is one example: even intermediate-separation ($\sim$10\,AU) binaries are biased
+toward rapid rotation
+\citep[e.g.][and many studies thereafter]{Meibom_2007}. %,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
 Metallicity may also be a second-order effect
@@
 In this work, we were interested in not only the overall
-age distribution of the Kepler field, but also in the reliability of ages for
-individual young planet hosting stars.
-The quality flagging scheme in Table~\ref{tab:planets} provides one method in support of
-this end.
+age distribution of the Kepler field, but also in the reliability of individual ages for
+young planet hosting stars.
+The quality flagging scheme in Table~\ref{tab:planets} provides one path toward this end.
 The other method is to consider the region in which we do have
@@
 Conversely, we found only one system, Kepler-786, with a clearly
-constraining lithium age (\kepseveneightsix\,Myr) and no rotation age.
+constraining lithium age (\kepseveneightsix\,Myr), but no rotation signal.
 
@@
 
-We began this article by asking two questions: how wrong is the
+We began this article with two questions: how wrong is the
 assumption of a uniform age distribution for stars in the galactic
@@
 rotation periods, lithium equivalent widths, and stellar temperatures
-using a combination of archival and new data.  We derived new ages
-using empirical interpolation-based age scales that have been
-calibrated against the latest available open cluster data.
+using archival and new data.  We derived new ages
+using empirical interpolation-based methods that have been
+calibrated against the latest open cluster data.
 
@@
 most of the sample, the independent lithium and rotation ages agreed
-for $\approx$90\% of cases.  In particular, we reported 14 confirmed
+for $\approx$90\% of cases.  
+%FIXME TODO NUMBERS
+In particular, we reported 14 confirmed
 $<$500\,Myr planets with strong age constraints from both rotation and
@@
 
-\section{Notable Individual Systems}
+\section{Notes on Individual Systems}
 

### 2024-05-08 paper/ms.tex
update ms

@@
 \citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
-billions of years old, similar to the Earth.  This fact leaves many
-gaps in our knowledge of what exact dynamical and physical processes
-produce these objects.    The reason is that processes such as thermal
-cooling \citep{2007ApJ...659.1661F}, atmospheric loss
+billions of years old.  This fact leaves many gaps in our knowledge of
+the exact physical and dynamical pathways that produce these objects.
+The reason is that processes such as thermal cooling
+\citep{2007ApJ...659.1661F}, atmospheric loss
 \citep{2019AREPS..47...67O}, giant impacts
@@
 \citep{2017MNRAS.470.1750I} are expected to be most efficient over
-$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
-known exoplanets, most of the action is in the past.
-
-Finding and characterizing young ($<$1\,Gyr) exoplanets is one
-potential path toward building a timeline for exoplanet evolution.
-Informative individual young systems include V1298~Tau, an adolescent
-resonant chain of four transiting puffy planets that is a likely
-precusor to the compact multiplanet systems \citep{David_2019}, and
-HIP~67522, a Jupiter-sized planet with a Neptunian mass
-(\citealt{Rizzuto_2020}; Thao et al.~submitted).  Population-level
-analyses have similarly suggested differences in the size distribution
-of young exoplanets relative to their older counterparts
+$\approx$10\,Myr to $\approx$1\,Gyr timescales; for most known
+exoplanets, most of the action has already happened.
+
+Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
+help in building the timeline for exoplanet evolution.  Informative
+individual young systems include V1298~Tau, an adolescent resonant
+chain of four transiting puffy planets that is a likely precusor to
+the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
+Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
+Thao et al.~submitted).  Population-level analyses have similarly
+suggested differences in the size distribution of young exoplanets
+relative to their older counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 To begin, imagine that the ages of nearby stars in the Galaxy are
-uniformly distributed between 0 and 10\,Gyr
-\citep[e.g.~Fig~17 of][]{Nordstrom_2004}.  This approximation clarifies the
-rarity of young stars: under these assumptions, $\approx$1\% of nearby
-stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
-Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
-or of exoplanets evolving just after disk dispersal
+uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
+of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
+young stars: under this assumption, $\approx$1\% of nearby stars have
+$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
+currently forming protoplanets \citep{2018A&A...617A..44K}, or of
+exoplanets evolving just after disk dispersal
 \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
@@
 Despite the observational challenges, young close-in planet discovery
-has matured as a field over the past decade, primarily due to
-Kepler, K2, TESS, and Gaia
+has matured over the past decade, primarily due to Kepler, K2, TESS,
+and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
@@
 stars with known ages---obvious members of open clusters---and to
-search them for transiting planets.  The resulting stellar (and assumed planetary) ages are
-precise at the $\approx$10\% level.  A
+search them for transiting planets.  The resulting stellar (and
+assumed planetary) ages are precise at the $\approx$10\% level.  A
 more recent development, facilitated by the second Gaia data release,
@@
 The main problem with the ``cluster-first'' approach is that most
-nearby young stars are in the field.  Only $\approx$1\% of
-stars within the nearest $\approx$500\,pc have been associated
-with their birth cluster
+nearby young stars are in the field.  Only $\approx$1\% of stars
+within the nearest $\approx$500\,pc are associated with their birth
+cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
@@
 than 1\,Gyr.  Imposing reasonable precision requirements
-(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
-S/N$=$3) gives $\approx$50 ``young'' transiting planets at the time
-of writing.  Most of these planets are in clusters, and were
-discovered by Kepler, K2, or TESS.  However, combined, those surveys
-have discovered $\approx$5{,}000 planets.  Unless there is a major
-bias against the discovery of sub-gigayear old planets with precisely
-constrained ages, a constant star formation history implies that
-an order of magnitude more young
-transiting planets should be known.
+(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say S/N$=$3)
+gives $\approx$50 ``young'' transiting planets at the time of writing.
+Most of these planets are in clusters, and were found by Kepler, K2,
+or TESS.  However, combined, those surveys have discovered
+$\approx$5{,}000 planets.  Unless there is a major bias against the
+discovery of sub-gigayear old planets with precisely constrained ages,
+a constant star formation history implies that an order of magnitude
+more young transiting planets should be known.
 
@@
 to assume a uniform age distribution for stars in the galactic thin
-disk, particularly young stars?  Second, where are the missing young
-transiting planets?  Briefly, we will find that uniform is incorrect
-by a factor of a few, and that the bias against the discovery of young
-transiting planets is, primarily, in the measurement of
-theirs ages. %, rather than in the sensitivity of the transit searches.
+disk?  Second, where are the missing young transiting planets?
+Briefly, we will find that uniform is incorrect by a factor of a few,
+and that at least one bias against the discovery of young transiting
+planets is in the measurement of their ages.
 
@@
 stars whose luminosities and temperatures separate them from the
-main-sequence.  \citet{David_2021} leveraged both isochronal ages from
-an earlier analogous study \citep{Fulton_2018}, and
-used a relative gyrochrone age-ranking approach that sorted stars
-based on whether they rotated faster or slower than a gyrochrone age
-of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
-were conducted in the broader context of investigating trends in
-exoplanet demographics, which also depend on stellar metallicity and
-mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
-sensitivity to the ages of the youngest planets, due to the nature of
-isochronal dating.
+main-sequence.  \citet{David_2021} leveraged isochronal ages from
+an earlier analogous study \citep{Fulton_2018}, and also
+used a relative gyrochrone age-ranking approach.
+%that sorted stars
+%based on whether they rotated faster or slower than a gyrochrone age
+%of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
+These studies were conducted in the broader context of investigating
+trends in exoplanet demographics, which also depend on stellar
+metallicity and mass \citep[e.g.][]{Petigura_2018}.  These studies
+offered limited sensitivity to the ages of $\lesssim$1\,Gyr planets,
+due to the nature of isochronal dating.
 
@@
     consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
-    For late G dwarfs and K dwarfs, the age calibration
+    For late G dwarfs and K dwarfs, the new calibration
     differs by factors of a few;
@@
 \end{enumerate}
-Today, FGK star rotation-based ages are
-empirically tied to the cluster age scale, and
-are limited in precision by the intrisic scatter in the $P_{\rm rot}$
-sequences, and in accuracy by systematic uncertainties in the evolving
-spin-down rates, and in the cluster ages themselves.
-
-Beyond improvements in rotation-based ages, recent years have also
-yielded improvements in the lithium age scale.
-Lithium ages include two qualitatively distinct regimes: depletion
-boundary ages derived for M-dwarfs in star clusters, and the less
-precise decline-based ages for individual field FGK stars
-\citep{Soderblom_2010}.  Here we focus on the latter approach, which
-relies on the observation that the Li abundances of
-partially-convective stars gradually decline as they age
+These improvements form the basis for why we believe the present
+analysis to be necessary.
+%Today, FGK star rotation-based ages are
+%empirically tied to the cluster age scale, and
+%are limited in precision by the intrisic scatter in the $P_{\rm rot}$
+%sequences, and in accuracy by systematic uncertainties in the evolving
+%spin-down rates, and in the cluster ages themselves.
+
+However, beyond improvements in rotation-based ages, recent years have
+also yielded improvements in the lithium age scale.  Lithium ages
+include two qualitatively distinct regimes: depletion boundary ages
+for M-dwarfs in star clusters, and the less precise decline-based ages
+for individual field FGK stars \citep{Soderblom_2010}.  Here we focus
+on the latter approach, which relies on the observation that the Li
+abundances of partially-convective stars gradually decline as they age
 \citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
@@
 \citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
-Empirical understanding has however recently improved due to the work
-by \citet{Jeffries_2023}, who built a model for how the equivalent
-width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
-of stellar effective temperature and age for a set of 6{,}200 stars in
-52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
+Empirical understanding has however improved due to the work by
+\citet{Jeffries_2023}, who built a model for how the equivalent width
+(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
+stellar effective temperature and age for a set of 6{,}200 stars in 52
+open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
 stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
@@
 
-We discuss our methods for selecting stellar and planetary samples in
-Section~\ref{sec:selection}, and describe
-the origin of our adopted stellar parameters other than ages in
-Section~\ref{sec:stellarprops}.
+We discuss our methods for selecting the star and planet samples in
+Section~\ref{sec:selection}, and describe the origin of our adopted
+stellar parameters other than ages in Section~\ref{sec:stellarprops}.
 We describe and validate our age-dating methods in
-Sections~\ref{sec:rotage} and~\ref{sec:liage}.
-The population-level trends for both the parent stellar sample and
-the planet sample are discussed in Section~\ref{sec:results}.
-A few conclusions are offered in Section~\ref{sec:conclusions}.
+Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
+trends for both stars and planets are described in
+Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
+are offered in Section~\ref{sec:conclusions}.
 
@@
 		with previously reported rotation periods (dark gray points).
-    The rotation periods were primarily measured by \citet{Santos_2019} and \citet{Santos_2021}.
+		The rotation periods were primarily measured by \citet{Santos_2019,Santos_2021}.
 		About half of these rotators are suitable for gyrochronology
@@
 
-\begin{figure*}[!t]
-  \begin{center}
-    \leavevmode
-    \subfloat{
-      \includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
-    }
-    \subfloat{
-      \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
-    }
-  \end{center}
-  \vspace{-0.6cm}
-  \caption{
-    {\bf The models.}
-    Points represent $10^4$ draws from model probability distributions
-    that have been fitted to rotation periods
-    \citep{Bouma_2023} and lithium equivalent widths
-    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
-    are the ``mean models'' at various ages; the asymmetric intrinsic
-    dispersion in rotation and lithium about these mean models, which
-    is what the models fit, sets the theoretical precision floor for
-    both age-dating methods.    Additional sources of uncertainty,
-    including measurement uncertainty, impose further limits on
-    achievable age precision.  These models are calibrated using open
-    clusters younger than 4\,Gyr.  The displayed points assume a
-    uniform distribution in temperature for visual clarity.  The sizes
-    of the points are the same in each panel, so that low apparent
-    density signifies greater dispersion around the mean.
-    \label{fig:models}
-  }
-\end{figure*}
+This work is focused on Kepler stars for which ages can be inferred
+using either rotation, lithium, or both.  These ``age-dateable
+stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
+periods have been reported for $\approx$one in three Kepler targets
+\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
+suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet
+have only been acquired for the Kepler objects of interest
+(KOIs), which comprise a few percent of Kepler's targets.
+In the following, we will describe the set of stars that we
+adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
+the set of objects we adopt as planets
+(Section~\ref{subsec:planetsel}), and the set of planets with
+high-resolution spectra suitable for lithium analysis
+(Section~\ref{subsec:lithiumsel}).
+Figure~\ref{fig:stellarprops} provides a visualization of the various samples.
+
+\subsection{Stellar Rotation Periods}
+\label{subsec:rotsel}
 
@@
 		{\bf Rotation periods for Kepler target stars (left) and known
-    planet hosts (right)}.  As in Figure~\ref{fig:stellarprops}, periods are from \citet{Santos_2019}
-    and \citet{Santos_2021} (\citetalias{Santos_2019} and
-    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' 
+			planet hosts (right)}.  
+		The gray lines are ``mean fits'' 
 		to the slow rotation sequences of M67,
@@
 		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
+		and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planet-hosts require
 		the same stellar cuts, and include only the confirmed and candidate planets
@@
 
-\begin{figure}[!t]
-  \begin{center}
-    \leavevmode
-    \includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
-  \end{center}
-  \vspace{-0.6cm}
-  \caption{
-    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
-      for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024 that comprised
-    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
-    isochrones from the same models as in Figure~\ref{fig:models}; the
-    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
-    than changes in the mean.  The logarithmic scale highlights the
-    dynamic range of the detections and upper limits.  Some stars with
-    lithium detections do not have detected rotation periods, and
-    vice-versa.
-    \label{fig:li_vs_teff}
-  }
-\end{figure}
-
-This work is focused on Kepler stars for which ages can be inferred
-using either rotation, lithium, or both.  These ``age-dateable
-stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
-periods have been reported for roughly one in three Kepler targets
-\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
-which can yield measurements of the \ion{Li}{1} 6708\,\AA\ doublet, 
-have only been acquired for the Kepler objects of interest
-(KOIs), which comprise a few percent of Kepler's targets.
-In the following, we will describe the set of stars that we
-adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
-the set of objects we adopt as planets
-(Section~\ref{subsec:planetsel}), and the set of planets with
-high-resolution spectra suitable for lithium analysis
-(Section~\ref{subsec:lithiumsel}).
-
-\subsection{Stellar Rotation Periods}
-\label{subsec:rotsel}
-
 To select stars with rotation periods, we turn to previous work.
-Many investigators have derived rotation periods both specifically of
-the Kepler Objects of Interest
+Many investigators have derived rotation periods both specifically for
+KOIs
 \citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
-and also of the broader set of all Kepler target stars
+and also for the broader set of all Kepler target stars
 \citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
@@
 functions.
-We are interested in understanding the age distribution of not only
-the KOIs but also the Kepler target stars.
+We are interested in understanding the age distribution of all Kepler
+target stars irrespective of KOI status.
 Out of these studies, the only homogeneous analysis of both Kepler
-targets and KOIs were the studies by
-\citet{Santos_2019} (\citetalias{Santos_2019}),
-and \citet{Santos_2021} (\citetalias{Santos_2021}).
+targets and KOIs were the studies by \citet{Santos_2019,Santos_2021}
+(\citetalias{Santos_2019} and \citetalias{Santos_2021}).
 \citetalias{Santos_2019} and
@@
 detections have lower variability amplitudes and longer periods than
-those reported by \citetalias{McQuillan_2014}.  A relevant flag for
-stars missing \citetalias{McQuillan_2014} periods is included in
-Table~\ref{tab:stars}.
+those reported by \citetalias{McQuillan_2014}. 
+%A relevant flag for
+%stars missing \citetalias{McQuillan_2014} periods is included in
+%Table~\ref{tab:stars}.
 
@@
 supplementary measure aimed at completeness in our final KOI age
-catalog; the provenances of the individual adopted periods are all
+catalog; the provenances of the individual adopted periods are 
 noted in the relevant tables.
@@
 FGK stars with slower rotation periods spin much slower than the stars
-in the oldest open clusters currently used to calibrate empirical
-gyrochronal models.
-
-The resulting tally of stars from
-\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
-KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.
-
-As a sanity check on the statistical uncertainties of these rotation
+in the oldest open clusters currently used to calibrate our 
+gyrochronology model.
+Figure~\ref{fig:prot_vs_teff} shows the resulting
+\nuniqstarsantosrot\ Kepler stars with rotation periods 
+compiled from
+\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our extended
+KOI list.
+
+To assess the statistical uncertainties of these rotation
 periods, we compared our adopted \citetalias{Santos_2019} and
 \citetalias{Santos_2021} periods with those reported by
-\citet{McQuillan_2014} and \citet{Mazeh_2015}.  We found that for
+\citet{McQuillan_2014}.  We found that for
 $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a precision
@@
 $P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
-set to guarantee 3\% $P_{\rm rot}$ uncertainties at rotation periods
+set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods
 of 30 days.
@@
 
-We considered planets in the NASA Exoplanet Archive's cumulative KOI
+We considered planets in the NEA cumulative KOI
 table, which included the best knowledge available on any given planet
@@
 The final piece of our analysis involves assessing ages based on the
-equivalent width of the \ion{Li}{1} 6708\,\AA\ doublet.  Specifically,
-we analyzed spectra acquired with the High Resolution Echelle
+\ion{Li}{1} 6708\,\AA\ doublet.  Specifically,
+we analyzed spectra from the High Resolution Echelle
 Spectrometer (HIRES; \citealt{vogt_hires_1994}) on the Keck I 10m
@@
 Interest were already analyzed by \citet{2018ApJ...855..115B} for
-about three quarters of the spectra in our sample.  However, since new
-spectra have been acquired since the \citeauthor{2018ApJ...855..115B}
-analysis, and since our approach and selection function are different,
-we opt to perform our own analysis of the reduced HIRES spectra.
-
-We queried the internal CPS archive to collect all blaze-corrected
+$\approx$three quarters of the spectra in our sample.  However, new
+spectra have since been acquired, and our approach and selection function are different.
+We therefore perform our own processing of the reduced HIRES spectra.
+
+We queried the CPS archive to collect all blaze-corrected
 HIRES spectra that were available for all non-false positive Kepler
-Objects of Interest with a MES$>$10.  This query yielded at least one
+Objects of Interest with a MES$>$10.  This yielded at least one
 spectrum for \nlithiumstars\ stars and \nlithiumplanets\ planets.  For
@@
 to a common reference wavelength by cross-correlating against a high
-S/N template for Kepler-1698, chosen because $\gamma$$\approx$9\kms,
+S/N template for Kepler-1698, chosen because $|\gamma|$$<$10\kms,
 $T_{\rm eff}$$\approx$5000\,K and $v\sin i$$\approx$5\,\kms, which
@@
 of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
-broadly consistent).  We continuum-normalized by fitting a
+consistent).  We continuum-normalized by fitting a
 third-degree Chebyshev series, while excluding regions with
@@
 \ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
-precision of our equivalent measurement technique, after applying an
+precision of our measurement technique, after applying an
 initial iteration on the \nlithiumstars\ KOIs with spectra, we
@@
 
+\begin{figure}[!t]
+	\begin{center}
+		\leavevmode
+		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
+			for planet-hosting stars.} These measurements were made from
+		Keck/HIRES spectra collected between 2009 and 2024 that comprised
+		\nhireshours\,hours of open-shutter time.  Lines are the ``mean''
+		isochrones from the same models as in Figure~\ref{fig:models}; the
+		intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
+		than changes in the mean.  The logarithmic scale highlights the
+		dynamic range of the detections and upper limits.  Some stars with
+		lithium detections do not have detected rotation periods, and
+		vice-versa.
+		\label{fig:li_vs_teff}
+	}
+\end{figure}
+
 \section{Stellar Properties}
@@
 the MIST isochrone grids
-\citep{2016ApJ...823..102C,2016ApJS..222....8D}, given the SDSS $g$
+\citep{Choi_2016,2016ApJS..222....8D}, given the SDSS $g$
 and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
@@
 
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
+		}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{
+		{\bf The models.}
+		Points represent $10^4$ draws from models
+		that have been fitted to rotation periods
+		\citep{Bouma_2023} and lithium equivalent widths
+		\citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
+		are the ``mean models'' at various ages; the intrinsic
+		dispersion in rotation and lithium about these mean models, which
+		is what the models fit, sets the theoretical precision floor for
+		the age-dating methods.    Additional sources of uncertainty,
+		including measurement uncertainty, impose further limits on
+		achievable precision.  These models are calibrated using open
+		clusters younger than 4\,Gyr.  The displayed points assume a
+		uniform distribution in temperature for visual clarity.  The sizes
+		of the points are the same in each panel, so that low apparent
+		density signifies greater dispersion around the mean.
+		\label{fig:models}
+	}
+\end{figure*}
+
 We calculate gyrochrone ages using \texttt{gyro-interp}
-\citep{Bouma_2023}, which is a method designed to address the issue
-that, especially over the first gigayear of stellar spin-down, stars
+\citep{Bouma_2023}, which is a method designed to address the fact
+that stars
 with the same mass and same rotation period can have very different
@@
 
-Using our adopted effective temperatures and rotation periods, we ran
+Using our effective temperatures and rotation periods, we ran
 \texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
@@
 the framework interpolates between reference open clusters using
-smooth cubic hermite polynomials.  We calculate the probability of the
+cubic hermite polynomials.  We calculate the probability of the
 rotation-based age $t_{\rm gyro}$, given the observed rotation
@@
 
-\subsection{Gyrochronology Quality Flags}
+\subsection{Stellar Quality Flags}
 \label{subsec:flags}
@@
 which we suspect these ages may not in fact be valid, we built a set
-of bit-quality flags.  When and how these flags should be applied
-depends on the question being asked.  If the goal is to construct a
-false-positive free sample, they might all be applied.  If the goal is
-to construct a complete sample of young stars, then consider the
-examples of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
+of quality flags, which we condensed into a single binary number,
+$Q_{\rm star}$.  When and how the flags represented in this should be
+applied depends on the question being asked.  If the goal is to
+construct a false-positive free sample, they could all be applied.  If
+the goal is to construct a complete sample, then consider the examples
+of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
 ($\approx$625\,Myr).  The former has a high RUWE due to a resolved
 binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
-\citep{2014ApJ...783...53M}.  Our gyrochronology quality flag, $Q_{\rm
-gyro}$, enables selecting for or against such cases by selecting from
-the following bits.
-
-As a general philosophy, three assumptions must hold for a
+\citep{2014ApJ...783...53M}.  We leave selection for or against such
+cases as a decision to the user.
+
+Three assumptions must hold for a
 rotation-based age dating method like \texttt{gyro-interp} to be
-valid.  {\it 1)} The evolutionary state of the target star must be
+valid.  {\it 1)} The evolutionary state of the star must be
 well-specified by its temperature and its age, {\it 2)} the star's
@@
 the first condition is that the star must be ``near the main
-sequence'', because during post main sequence evolution, the
-temperatures of stars change.  From $\approx$0.08 to $\approx$3\,Gyr,
+sequence'' because during the post main sequence 
+stellar temperatures change.  From $\approx$0.08 to $\approx$3\,Gyr,
 the stars of interest in this work (3800--6200\,K,
@@
 ($\approx$3\%), because its outer layers expand as it begins hydrogen
-shell burning as a subgiant.  We treat this issue in a somewhat
-convoluted manner, discussed in ``bit 9'' below.
+shell burning as a subgiant.  We treat this issue in a 
+manner discussed in ``bit 9'' below.
 
@@
 %crossmatch
-of the NASA Exoplanet Archive
+of the NEA
 \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
@@
 flagged using bit 4.  For KIC stars with multiple potential Gaia
-matches within the 4$''$ search radius, we adopted the brightest star
+matches within the 4$''$ radius, we adopted the brightest star
 as the default match.  In most such cases this was unambiguous because
@@
 flags known eclipsing, astrometric, and spectroscopic binaries.  We
-directly include this column.
+directly included this column.
 
@@
 outliers can be either bona fide astrometric binaries, or more often
-are marginally resolved point sources for which the single-source PSF
-model used in the Gaia pipeline induces spurious astrometric scatter.
+are marginally resolved point sources for which a single-source PSF
+model provides a poor fit.
 
@@
 for stars within 1 Kepler pixel (4$''$) of every target star.  While
-many such companions are not physically associated with the target
-star, their presence can bias stellar parameters and stellar rotation
-period measurements.  We therefore flag any stars with neighbors down
-to 10$\times$ the brightness of the target star within this region
-($\Delta G < 2.5$).
+such companions may not not be physically associated with the target
+star, their presence can confuse rotation period measurements.  We
+therefore flagged any stars with neighbors down to 10$\times$ the
+brightness of the target star within this region ($\Delta G < 2.5$).
 
 {\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
-are far from the main sequence: this can be seen in color--absolute
-magnitude diagrams, and in diagrams of e.g.~surface gravity as a
-function of temperature (Figure~\ref{fig:stellarprops}).  This is a
-challenge for rotation-based ages because {\it i)} point sources
-far from the main-sequence include unresolved photometric binaries,
-and {\it ii}) effects other than magnetic braking influence the
-rotation periods of bona fide single stars far from the main-sequence.
-The added complications of reddening and metallicity further
-complicate matters.  
-After exploring various options, we settled on the orange locus in the
-$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.
-While the details of how exactly this locus is constructed are
-arbitrary, 
+are far from the main sequence: this can be seen in both panels of
+Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
+ages because {\it i)} point sources far from the main-sequence include
+unresolved photometric binaries, and {\it ii}) effects other than
+magnetic braking influence the rotation periods of bona fide single
+stars far from the main-sequence.  The added complications of
+reddening and metallicity further complicate matters.  After exploring
+various options, we settled on the orange locus in the $\log
+g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
+the details of how exactly this locus is constructed are arbitrary, 
 %and convoluted,\footnote{In practice, for K dwarfs we fitted an
@@
 %F and G dwarfs to that of the K dwarfs.}
-the general 
-aim is to exclude stars for which there is evidence 
-based on their location in the $\log g$--$T_{\rm eff}$ plane
-that our gyrochronology model may not be reliable.  
+the general aim is to flag stars for which there is evidence based on
+their location in the $\log g$--$T_{\rm eff}$ plane that our
+gyrochronology model may not be reliable.  
 %The rotation period
@@
 %likely due to a combination of astrophysical and systematic effects.
-See for instance Figure~8a of \citet{2022AJ....164..137K} or Figures~12 and 13
-of \citet{2023ApJS..268....4F} for examples of how this period distribution
-of this population differs from stars near the main sequence.
+See for instance Figure~8a of \citet{2022AJ....164..137K} or
+Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
+the period distribution of this population differs from stars near the
+main sequence.
 
@@
 10)}---\citeauthor{Santos_2021} included in a flag for candidate
-``classical pulsators'' (e.g.\ RR~Lyrae, Cepheids) and candidate
-close-in binaries.  Visual inspection of the rotation periods as a
-function of temperature shows that although this flag selects many
-bona fide objects in these classes, it also selects the overwhelming
-majority of $\lesssim$5\,day bona fide young planets.  In our view this is
-erroneous because they these objects are neither classical pulsators nor
-close binaries.  While we
-propagate this flag into our set of quality flags, we therefore
-generally ignore it.
+``classical pulsators'' (e.g.\ Cepheids) and
+close-in binaries.  Visual inspection 
+% of the rotation periods as a function of temperature 
+shows that although this flag selects many bona fide objects in these
+classes, it also selects most $P_{\rm
+rot}$$\lesssim$5\,day bona fide young planets.  
+These objects are neither classical pulsators nor close
+binaries.  While we propagated this flag into our set of quality
+flags, we therefore generally ignore it.
+
+{\it Not in the homogeneous stellar rotation sample (bit 11)}---
+This flag is set for the KOIs with rotation periods
+drawn from a source other than \citet{Santos_2019,Santos_2021}.
 
 {\it How do we use the bit-flags?}
-An important decision for our subsequent analysis is how to combine
-the above quality indicators in order to decide whether or not a star is suited for
-gyrochronology.  By default, we assume that a star is suitable for
-gyrochronology if none of bits zero through nine (inclusive) are
-raised.
+By default, we assume that a star is
+suitable for gyrochronology if none of bits zero through nine
+(inclusive) are raised.
+For analyses that require a homogeneous stellar rotation selection
+function, we also require bit 11 to be zero.
 
@@
 \label{subsec:plflags}
-Some planets are more reliable than others.
-We use the following additional quality indicators to assess 
-the reliability and utility of a planet, and assemble them into a separate bit-flag,
-$Q_{\rm planet}$.
+Some planets are more reliable than others.  We use the following
+additional quality indicators to assess the reliability and utility of
+a planet, and assemble them into a separate bit-flag, $Q_{\rm
+planet}$.
 
 % see Erik's 2022 discussion; might want to shift to include a cut on
-% disposition score.
-{\it Candidate reliability (bit 0)}---The NEA's Kepler Objects of Interest Table
-includes both an overall planet-candidate disposition status
-({\it koi\_disposition}), as well as a disposition based only on the Kepler
-data ({\it koi\_pdisposition}).
-We require both to include only
-``planet candidates'' and ``confirmed planets''; we are generally less interested
-in the age distribution of planet candidates that are known false positives.
-
-{\it Candidate S/N (bit 1)}---In any transit survey, the false positive rate increases
-greatly toward the noise floor for planet detection \citep[e.g.][]{2002ApJ...564..495J}.  We
-required a S/N in excess of Kepler's usual 7.1$\sigma$ floor, through a cut on the
-maximum
-``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}): we required ${\rm MES}$$>$10.
+% disposition score?
+{\it Candidate reliability (bit 0)}---The NEA's Kepler Objects of
+Interest Table includes both an overall planet-candidate disposition
+status ({\it koi\_disposition}), as well as a disposition based only
+on the Kepler data ({\it koi\_pdisposition}).  We require both to
+include only ``planet candidates'' and ``confirmed planets''. 
+%; we are generally less interested
+%in the age distribution of planet candidates that are known false positives.
+
+{\it Candidate S/N (bit 1)}---In any transit survey, the false
+positive rate increases greatly toward the noise floor for planet
+detection \citep[e.g.][]{2002ApJ...564..495J}.  We required a S/N in
+excess of Kepler's usual 7.1$\sigma$ floor, through a cut on the
+maximum ``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}):
+we required ${\rm MES}$$>$10.
 
 {\it Grazing planets (bit 2)}---Grazing objects, for which the impact
-parameter $b$ is greater than $1-R_{\rm p}/R_\star$,
-often yield biased planetary parameters
-\citep[e.g.][]{2022AJ....163..111G}.
-Particularly for large planets, they can also include
-astrophysical false positives at higher rates \citep{2016ApJ...822...86M}, which is
-in part due to the size-impact parameter degeneracy.
-We flag planet candidates as potentially grazing if $b>0.8$, 
-using the impact parameters reported in the Markov Chain
-Monte Carlo (MCMC) fits from \citet{Thompson_2018}.
+parameter $b$ is greater than $1-R_{\rm p}/R_\star$, often yield
+biased planetary parameters \citep[e.g.][]{2022AJ....163..111G}.  For
+large planets, they also include astrophysical false positives at
+higher rates \citep{2016ApJ...822...86M}, in part due to the
+size-impact parameter degeneracy.  We flag planet candidates as
+potentially grazing if $b>0.8$, using the impact parameters reported
+in the Markov Chain Monte Carlo (MCMC) fits from
+\citet{Thompson_2018}.
 
@@
 
-The rotation periods for the stellar and planetary samples are shown
+The rotation periods for the star and planet samples are shown
 in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
 observed by Kepler, requiring reported rotation periods and a reliable
-DR3 crossmatch yielded \nuniqstarsantosrot\ stars.  Further requiring
-these rotating stars to have temperatures 3800$<$$T_{\rm
-eff}$/K$<$6200, which is the range over which \texttt{gyro-interp}
-returns finite age posteriors, yields \nuniqstarsantosrotteffcut\
+DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
+these stars to have temperatures 3800$<$$T_{\rm
+eff}$/K$<$6200, yields \nuniqstarsantosrotteffcut\
 stars.  Further requiring the stars to be apparently single, near the main
@@
 uncertainty intervals.
-Certain quality flags described in Section~\ref{subsec:flags} are also
-relevant for lithium ages, specifically those associated with
-crowding and binarity.
-For instance, rapid rotators are more common in binaries, and they also
-generally have higher lithium abundances than slow-rotating stars of the
-same mass (CITE).
-Stars for which rotation-based age dating is dubious heavily overlap
-with those for which lithium-based ages are dubious.
+
+%Certain quality flags described in Section~\ref{subsec:flags} are also
+%relevant for lithium ages, specifically those associated with
+%crowding and binarity.
+%For instance, rapid rotators are more common in binaries.
+%and they also
+%generally have higher lithium abundances than slow-rotating stars of the
+%same mass (CITE).
+%Stars for which rotation-based age dating is dubious heavily overlap
+%with those for which lithium-based ages are dubious.
 
@@
 and {\it ii)} by the prior.
-For instance, 
- we quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
+In these instances, we quote the 3$\sigma$ (99.7$^{\rm th}$
+percentile) limits.
 
@@
     \subfloat{
-    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
     }
@@
     \subfloat{
-    \includegraphics[width=\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
     }
@@
 
-\paragraph{Cep-Her}
-The $\approx$40\,Myr Cep-Her complex currently has its most
+{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its most
 constraining age measurement from relative isochrone dating of
@@
 
-\paragraph{MELANGE-3}
-\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
+{\it MELANGE-3}---\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
 the Kepler field and derived its age using rotation,
@@
 
-\paragraph{Theia~520}
-We derive rotation-based ages of 330-350\,Myr for Kepler-968 and Kepler-52.
+{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for Kepler-968 and Kepler-52.
 These stars were reported by
@@
 
-\paragraph{NGC~6811}
-\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
+{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
 mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
@@
 ``confirmed'' planets:
-$Q_{\rm gyro}$ must not have any of bits zero through nine
+$Q_{\rm star}$ must not have any of bits zero through nine
 raised, and $Q_{\rm planet}$ must have no quality flags raised at all.
@@
 
-Most of our ages come from a single source of information: rotation.
+Most of our ages come from one source of information: rotation.
 Factors other than stellar age and mass can influence rotation rates of main sequence stars.
@@
 stars for which the two age indicators are both constraining and in agreement
-(with $Q_{\rm gyro}$ bits zero through nine not raised, and no planet quality flags raised).
+(with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
 Conversely, we found only one system, Kepler-786, with a clearly
@@
 
-In an attempt to answer these questions, we curated a sample of
-rotation periods, lithium equivalent widths, and stellar temperatures
+Our approach to answering these questions was to curate a sample of
+stellar rotation periods, lithium equivalent widths, and temperatures
 using archival and new data.  We derived new ages
 using empirical interpolation-based methods that have been
-calibrated against the latest open cluster data.
+calibrated against an extensive set of open cluster data.
 
@@
 most of the sample, the independent lithium and rotation ages agreed
-for $\approx$90\% of cases.  
-%FIXME TODO NUMBERS
-In particular, we reported 14 confirmed
+for $\gtrsim$90\% of cases.  
+In particular, we reported {\bf XX} confirmed
 $<$500\,Myr planets with strong age constraints from both rotation and
-lithium, including five mini-Neptunes, five super-Earths, three Earths
-and sub-Earths, and a Neptune-sized planet.
+lithium, including {\bf YY} mini-Neptunes, {\bf ZZ} super-Earths, {\bf
+AA} Earths and sub-Earths, and a Neptune-sized planet.
 
@@
   \item The age distributions of both the Kepler target stars and the
-    known Kepler planet-hosts shows a demographic cliff.  There are
+    known Kepler planet-hosts show a demographic cliff.  There are
     \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr) stars in the
@@
   \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, the rotation-based age distribution of stars in the
+    billion years, this rotation-based age distribution of stars in the
     Kepler field implies that we should instead expect $\approx$250.
@@
   %FIXME: 120 number
-  \item Using rotation-based ages, we have reported the ages of
+  \item Using rotation-based ages, we have derived ages for
     \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
     and \nplyounggyro\ planets with median ages below 1\,Gyr.
-    Joining these planets against the existing literature yields
+    Concatenating these planets against the existing literature yields
     $\approx$120-170 known sub-gigayear planets with well-constrained
-    ages.  We have therefore gone from a factor of ten discrepancy to
+    ages.  This lessens the original factor of ten discrepancy to
     a factor of at most two.
@@
   positive KOIs with MES$>$10.  
-  A shell script to decode the $Q_{\rm gyro}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
+  A shell script to decode the $Q_{\rm star}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
   \label{tab:planets}}
@@
   $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
-  $Q_{\rm planet}$ & $Q_{\rm gyro}$ &  Spec? & Comment \\
+  $Q_{\rm planet}$ & $Q_{\rm star}$ &  Spec? & Comment \\
   -- &   -- & K & days & m\AA & Myr & Myr &  str &   Earths &    days &       int  & int & bool & -- \\
@@
     The bit quality flags for the rotation-based ages,
-    $Q_{\rm gyro}$, are described in Section~\ref{subsec:flags}.
+    $Q_{\rm star}$, are described in Section~\ref{subsec:flags}.
     Concisely summarized, they are:
@@
     \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
-    As an example, Kepler-1627 has $Q_{\rm gyro}$ flagged with bit 10 and bit 7.
+    As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
     The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
@@
   \startdata
-  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm gyro}$ \\
+  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
   -- &   -- & K & days &  Myr &    int  \\
@@
 suggests an age at least as old as the Hyades
-\citep{2008ApJ...687.1264M}.
+\citep{Mamajek_2008}.
 The Balmer lines are in absorption, and display no obvious signatures of youth.
@@
 \paragraph{Kepler-639, Kepler-320, Kepler-1719, Kepler-1876, Kepler-1072, Kepler-1743, Kepler-1929, Kepler-1488}---These eight systems all have nominally two-sided lithium age posteriors between 1 and 3\,Gyr, and yet lack rotation period detections.
-All are subgiants, flagged with $Q_{\rm gyro}$ under various combinations of bits 1, 2, and 9.
+All are subgiants, flagged with $Q_{\rm star}$ under various combinations of bits 1, 2, and 9.
 

### 2024-05-08 paper/ms.tex
update ms

@@
   \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, this rotation-based age distribution of stars in the
+    billion years, the age distribution of stars in the
     Kepler field implies that we should instead expect $\approx$250.

### 2024-05-08 gyrojo/plotting.py
Prot vs Teff hist2d

@@ -44,6 +44,7 @@ import matplotlib as mpl
 import matplotlib.transforms as transforms
 import matplotlib.patches as patches
 import matplotlib.colors as mcolors
+from mpl_toolkits.axes_grid1.inset_locator import inset_axes
 
 from numpy import array as nparr
 
@@ -297,12 +298,56 @@ def plot_star_Prot_Teff(outdir, sampleid):
 
     print(f"Mean Teff error is {np.nanmean(Teff_errs):.1f} K")
 
-    ax.errorbar(
-        Teffs, Prots,
-        yerr=np.zeros(len(Prots)),
-        marker='o', elinewidth=0., capsize=0, lw=0, mew=0., color='k',
-        markersize=0.5, zorder=5
-    )
+    DO_POINTS = 0
+    if DO_POINTS:
+        ax.errorbar(
+            Teffs, Prots,
+            yerr=np.zeros(len(Prots)),
+            marker='o', elinewidth=0., capsize=0, lw=0, mew=0., color='k',
+            markersize=0.5, zorder=5
+        )
+
+    else:
+        # Define the bin sizes
+        dTeff = 100/5
+        dProt = 1/4
+
+        # Create the 2D histogram
+        hist, xedges, yedges, im = plt.hist2d(
+            Teffs, Prots,
+            bins=[np.arange(3700, 6300, dTeff),
+                  np.arange(-1, 46, dProt)]
+        )
+
+        # Create a custom colormap with white color for zero values
+        cmap = plt.cm.YlGnBu
+        cmaplist = [cmap(i) for i in range(cmap.N)]
+        cmaplist[0] = (1.0, 1.0, 1.0, 1.0)  # Set the color for zero values to white
+        cmap = mcolors.LinearSegmentedColormap.from_list('Custom YlGnBu', cmaplist, cmap.N)
+
+        ## Apply log scaling to the colorbar
+        #norm = mcolors.LogNorm(vmin=1, vmax=np.max(hist))
+        norm = mcolors.Normalize(vmin=1, vmax=7)
+        im.set_norm(norm)
+
+        # Update the colormap of the plot
+        im.set_cmap(cmap)
+
+        show_colorbar = 0
+        if show_colorbar:
+            axins1 = inset_axes(ax, width="20%", height="2%", loc='upper right',
+                                borderpad=1.0)
+
+            cb = fig.colorbar(im, cax=axins1, orientation="horizontal",
+                              extend="max", norm=norm)
[…]

### 2024-05-08 paper/ms.tex
update ms

@@
 		The gray lines are ``mean fits'' 
-		to the slow rotation sequences of M67,
-		Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
-		clusters.  The stellar sample includes only apparently single stars
+		to the slow rotation sequences of open
+		clusters.  The stellar sample (left) includes only apparently single stars
 		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-		and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planet-hosts require
+		and 3800$<$$T_{\rm eff}$/K$<$6200.
+		The color map has a linear stretch.  %between one and ten stars per bin.
+		The planet-hosts (right) require
 		the same stellar cuts, and include only the confirmed and candidate planets
@@
 
-\subsubsection{New Young Systems}
+\subsubsection{New Young Planets}
 %

### 2024-05-09 drivers/plot_li_vs_teff.py
li vs teff: go linear

@@ -7,6 +7,9 @@ if not os.path.exists(outdir): os.mkdir(outdir)
 
 # most useful
 ap.plot_li_vs_teff(outdir, sampleid='koi_X_JUMP',
+                   yscale='linear', limodel='eagles',
+                   show_dispersion=0)
+ap.plot_li_vs_teff(outdir, sampleid='koi_X_JUMP',
                    yscale='log', limodel='eagles',
                    show_dispersion=0)
 ap.plot_li_vs_teff(outdir, sampleid='koi_X_S19S21dquality',

### 2024-05-09 gyrojo/plotting.py

@@ -387,6 +387,8 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
     upperlim = (li_ew - li_ew_merr < 10)
     det = ~upperlim
     li_ew_upper_lims = li_ew[upperlim] + li_ew_perr[upperlim]
+    if yscale in [None, 'linear']:
+        li_ew_upper_lims[li_ew_upper_lims < 0] = 1
 
     if limodel == 'baffles':
 
@@ -547,7 +549,7 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
         "$N_\mathrm{s}$ = " + f"{n_st}"
     )
     if not nodata:
-        ax.text(0.03, 0.97, txt, transform=ax.transAxes,
+        ax.text(0.04, 0.97, txt, transform=ax.transAxes,
                 ha='left',va='top', color='k')
 
     ax.set_ylabel('Li$_{6708}$ EW [m$\mathrm{\AA}$]')
@@ -562,7 +564,7 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
     if show_dispersionpoints:
         ax.set_ylim([ -5, 320 ])
     if not show_dispersion and not show_dispersionpoints:
-        ax.set_ylim([ -5, 500 ])
+        ax.set_ylim([-10, 270])
 
     s = f'_{sampleid}'
     s += f"_{limodel}"

### 2024-05-09 paper/ms.tex
update ms

@@
 		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-		and 3800$<$$T_{\rm eff}$/K$<$6200.
+		and temperatures of 3800--6200\,K.
 		The color map has a linear stretch.  %between one and ten stars per bin.
@@
 		\leavevmode
-		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
+		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
 	\end{center}
@@
 		intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
-		than changes in the mean.  The logarithmic scale highlights the
-		dynamic range of the detections and upper limits.  Some stars with
-		lithium detections do not have detected rotation periods, and
-		vice-versa.
+    than changes in the mean.  Some stars with lithium detections do
+    not have detected rotation periods, and vice-versa.
 		\label{fig:li_vs_teff}
@@
 \nuniqstarsantosrotteffcut\ stars with reported rotation periods and
-effective temperatures between 3800\,K and 6200\,K.  To flag stars for
+effective temperatures of 3800--6200\,K.  To flag stars for
 which we suspect these ages may not in fact be valid, we built a set
@@
 DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
-these stars to have temperatures 3800$<$$T_{\rm
-eff}$/K$<$6200, yields \nuniqstarsantosrotteffcut\
+these stars to have temperatures of 3800--6200\,K, yields \nuniqstarsantosrotteffcut\
 stars.  Further requiring the stars to be apparently single, near the main
@@
     The top row corresponds to all
-    stars with $T_{\rm eff}\in[3800,6200]\,$K for which we calculated rotation-based ages.
+    stars with temperatures of 3800--6200\,K for which we calculated rotation-based ages.
     The darker lines impose quality cuts
@@
     state.
-    The bottom row selects only stars with $T_{\rm eff}\in[4400,5400]\,$K, which
+    The bottom row selects only stars with temperatures of 4400-5400\,K, which
     have relatively precise ages near 2\,Gyr due to their fast spin-down
@@
 using archival and new data.  We derived new ages
-using empirical interpolation-based methods that have been
+using empirical methods that have been
 calibrated against an extensive set of open cluster data.

### 2024-05-09 paper/collect.sh

@@ -23,8 +23,8 @@ cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_g
 cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf .
 
 # plot_li_vs_teff
-cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
-cp $r/li_vs_teff/li_vs_teff_koi_X_JUMP_eagles_logy.pdf .
+#cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_JUMP_eagles.pdf .
 cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
 
 # planets

### 2024-05-09 paper/ms.tex
update ms

@@
 Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
-help in building the timeline for exoplanet evolution.  Informative
+help build the timeline for exoplanet evolution.  Informative
 individual young systems include V1298~Tau, an adolescent resonant
@@
 \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
-maximum achievable sample sizes by the tyranny of the Galactic star
+maximum achievable sample sizes by the tyranny of the galactic star
 formation rate.
 
-Despite the observational challenges, young close-in planet discovery
+Despite this and other observational challenges, young close-in planet discovery
 has matured over the past decade, primarily due to Kepler, K2, TESS,
 and Gaia
-\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
+\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
 The strategy pursued by most groups during the 2010s was to focus on
@@
 more recent development, facilitated by the second Gaia data release,
-is the idea that finding transiting planets can often be easier than
-finding and characterizing the host stellar association
+is the idea that finding transiting planets is often easier than
+finding the host stellar association
 \citep[e.g.][]{Tofflemire_2021}.
@@
 This study aims to resolve two main questions.  First, how wrong it is
-to assume a uniform age distribution for stars in the galactic thin
-disk?  Second, where are the missing young transiting planets?
-Briefly, we will find that uniform is incorrect by a factor of a few,
-and that at least one bias against the discovery of young transiting
-planets is in the measurement of their ages.
+to assume a uniform age distribution for transiting planet host star ?
+Second, where are the missing young planets?  Briefly, we will find
+that uniform is incorrect by a factor of a few, and that at least one
+bias against the discovery of young transiting planets is in the
+measurement of their ages.
 
@@
 
-However, beyond improvements in rotation-based ages, recent years have
+However, beyond rotation-based ages, recent years have
 also yielded improvements in the lithium age scale.  Lithium ages
-include two qualitatively distinct regimes: depletion boundary ages
+include two distinct regimes: depletion boundary ages
 for M-dwarfs in star clusters, and the less precise decline-based ages
@@
 this decline is debated
-\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
+\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}. 
 Empirical understanding has however improved due to the work by
@@
 stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
-spectral type since the K-dwarfs lose their surface lithium much
-faster than G-dwarfs.  The precision of the ages reported by the
+spectral type since the K dwarfs lose their surface lithium much
+faster than G dwarfs.  The precision of the ages reported by the
 \citet{Jeffries_2023} method in this regime are in the range of
@@
 
-We discuss our methods for selecting the star and planet samples in
+We discuss our method for selecting the star and planet samples in
 Section~\ref{sec:selection}, and describe the origin of our adopted
@@
 Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr near-Solar analog with
-two known transiting planets: an Earth-sized planet on a one-day orbit,
+two transiting planets: an Earth-sized planet on a one-day orbit,
 and a mini-Neptune on a five-day orbit.
@@
 Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), 
-could be the
+rank among the
 youngest Earth-sized planets currently known, comparable to systems
@@
 (with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
-Conversely, we found only one system, Kepler-786, with a clearly
-constraining lithium age (\kepseveneightsix\,Myr), but no rotation signal.
+Conversely, we found only one system, Kepler-786, with a 
+precise lithium age (\kepseveneightsix\,Myr), but no rotation signal.
 
@@
 planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
-Rotation-based age dating of their discoveries could be useful for the effort
-of identifying young exoplanets, and also for the broader issue of understanding
-their evolution.
+Rotation-based age dating of their discoveries could 
+potentially clarify their youth, and thereby help clarify the broader
+issue of their evolution.
 
@@
 A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
-yet to be performed.
+not been performed.
 The ideal approach for such a search would be to combine spatial and kinematic information

### 2024-05-10 paper/ms.tex
update ms

@@
 
-%%%%%%%%%
-% STARS %
-%%%%%%%%%
+%%%%%%%%%%
+% VALUES %
+%%%%%%%%%%
 \input{vals_manuscript}
@@
 \newcommand{\fracstarswithprotwithoutbtwenty}{{$\approx$6\%}}
-
-%%%%%%%%%%%
-% PLANETS %
-%%%%%%%%%%%
 % fraction of KOIs that are not FPs with B+20 parameters
@@
 
-%\title{Gyrochrone and Lithium Ages for Stars and Planets Observed by Kepler}
-%\title{Ages of the Kepler Stars and Planets}
-%\title{A Dearth of Young Stars and Planets in the Kepler Field}
 \title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
@@
   %
-  Rotation periods are drawn from previous Kepler analyses; lithium
-  is measured from new and archival Keck/HIRES spectra.
+  Rotation periods are drawn from previous Kepler analyses; lithium is
+  measured from new and archival Keck/HIRES spectra.
   %
@@
   and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is valid; these rotation-based ages are
-  roughly an order of magnitude more precise than isochrone ages in
-  this sample of stars.
+  gyrochronology is valid; these rotation-based ages are roughly an
+  order of magnitude more precise than isochrone ages in this sample
+  of stars.
   %
@@
   More broadly, we find that the star formation rate in the Kepler
-  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the
-  past 3\,Gyr, equally for both planet hosts and for the overall
+  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over
+  the past 3\,Gyr, equally for both planet hosts and for the overall
   stellar sample.
   %
-  This ``demographic cliff'' is a known feature of the Galaxy's star formation
-  history, and it helps clarify the relative balance of
-  young and old planets that have been detected to date using transits.
+  This ``demographic cliff'' is a known feature of the Galaxy's star
+  formation history, and it helps clarify the relative balance of
+  young and old planets that have been detected to date using
+  transits.
   %
@@
 \keywords{Stellar ages (1581), Planet hosting stars (1242), Field
-stars (2103), Exoplanet evolution (491), Milky Way evolution (1052),}
+stars (2103), Exoplanet evolution (491), Milky Way evolution (1052)}
 
@@
 
-Despite this and other observational challenges, young close-in planet discovery
-has matured over the past decade, primarily due to Kepler, K2, TESS,
-and Gaia
+Despite this and other observational challenges, young close-in planet
+discovery has matured over the past decade, primarily due to Kepler,
+K2, TESS, and Gaia
 \citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
@@
 is the idea that finding transiting planets is often easier than
-finding the host stellar association
-\citep[e.g.][]{Tofflemire_2021}.
+finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.
 
@@
 
-This study aims to resolve two main questions.  First, how wrong it is
-to assume a uniform age distribution for transiting planet host star ?
+This study aims to resolve two main questions.  First, how wrong is it
+to assume a uniform age distribution for transiting planet host stars?
 Second, where are the missing young planets?  Briefly, we will find
-that uniform is incorrect by a factor of a few, and that at least one
-bias against the discovery of young transiting planets is in the
-measurement of their ages.
-
-Given that Kepler's main mission ended in 2013, one might be
-forgiven for assuming that the ages of the Kepler planets are already
-known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
-and \citet{Petigura_2022} for instance calculated isochrone ages for
+that uniform is wrong by a factor of a few, and that at least one bias
+against the discovery of young planets is in the measurement of their
+ages.
+
+Given that Kepler's main mission ended in 2013, one might be forgiven
+for assuming that the ages of the Kepler planets are already known.
+Of course, they have been studied: \citet{Berger_2020b_rpage} and
+\citet{Petigura_2022} for instance calculated isochrone ages for
 Kepler Objects of Interest (KOIs).  These ages are most precise for
 stars whose luminosities and temperatures separate them from the
-main-sequence.  \citet{David_2021} leveraged isochronal ages from
-an earlier analogous study \citep{Fulton_2018}, and also
-used a relative gyrochrone age-ranking approach.
-%that sorted stars
-%based on whether they rotated faster or slower than a gyrochrone age
-%of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
-These studies were conducted in the broader context of investigating
-trends in exoplanet demographics, which also depend on stellar
-metallicity and mass \citep[e.g.][]{Petigura_2018}.  These studies
-offered limited sensitivity to the ages of $\lesssim$1\,Gyr planets,
-due to the nature of isochronal dating.
+main-sequence.  \citet{David_2021} leveraged isochronal ages from an
+earlier analogous study \citep{Fulton_2018}, and also used a relative
+gyrochrone age-ranking approach.  These studies were conducted in the
+broader context of investigating trends in exoplanet demographics,
+which also depend on stellar metallicity and mass
+\citep[e.g.][]{Petigura_2018}.  These studies offered limited
+sensitivity to the ages of $\lesssim$1\,Gyr planets, due to the nature
+of isochronal dating.
 
 Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
-between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
-times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
-clock has a rich history
+between 1-4\,Gyr, and provides more complex asymmetric posteriors at
+earlier times \citep{Bouma_2023}.  The idea of using a star's
+spin-down as a clock has a rich history
 \citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
@@
 derive rotation-based ages for subsets of the Kepler data
-\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
-have also been reported \citep{2021AJ....161..189L}.
-There have been
-major improvements since these pioneering studies in:
+\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic''
+ages have also been reported \citep{2021AJ....161..189L}.  There have
+been major improvements since these pioneering studies in:
 \begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
-    positives within it;
-  %
-  \item Available measurements of stellar rotation periods
-    in open clusters
+  \item The extent of the Kepler Object of Interest catalog, and our
+    vetting of false positives within it;
+  %
+  \item Available measurements of stellar rotation periods in open
+    clusters
     \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
@@
   \item Our ability to model not only the mean evolution but also the
-    intrinsic dispersion in rotation through time
-    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
-    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
-    For late G dwarfs and K dwarfs, the new calibration
-    differs by factors of a few;
+    intrinsic dispersion in rotation through time \citep{Bouma_2023}.
+    For Sun-like stars, the new model is broadly consistent with older
+    calibrations \citep[e.g.][]{Mamajek_2008}.  For late G dwarfs and
+    K dwarfs, the new calibration differs by factors of a few;
   %
@@
 analysis to be necessary.
-%Today, FGK star rotation-based ages are
-%empirically tied to the cluster age scale, and
-%are limited in precision by the intrisic scatter in the $P_{\rm rot}$
-%sequences, and in accuracy by systematic uncertainties in the evolving
-%spin-down rates, and in the cluster ages themselves.
-
-However, beyond rotation-based ages, recent years have
-also yielded improvements in the lithium age scale.  Lithium ages
-include two distinct regimes: depletion boundary ages
-for M-dwarfs in star clusters, and the less precise decline-based ages
-for individual field FGK stars \citep{Soderblom_2010}.  Here we focus
-on the latter approach, which relies on the observation that the Li
-abundances of partially-convective stars gradually decline as they age
+
+However, beyond rotation-based ages, recent years have also yielded
+improvements in the lithium age scale.  Lithium ages include two
+distinct regimes: depletion boundary ages for M-dwarfs in star
+clusters, and the less precise decline-based ages for individual field
+FGK stars \citep{Soderblom_2010}.  Here we focus on the latter
+approach, which relies on the observation that the Li abundances of
+partially-convective stars gradually decline as they age
 \citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
@@
     {\bf The stars.}  Our analysis focuses on stars observed by Kepler
-		with previously reported rotation periods (dark gray points).
-		The rotation periods were primarily measured by \citet{Santos_2019,Santos_2021}.
-		About half of these rotators are suitable for gyrochronology
-		(black points), based on factors including their temperatures,
-		surface gravities, non-binarity, and proximity to the main
-		sequence (see Section~\ref{subsec:flags}).   Some are
-		``confirmed'' or ``candidate'' Kepler Objects of Interest that
-		meet an additional set of planetary quality criteria (blue points;
-		Section~\ref{subsec:plflags}).  Surface gravities and effective
-		temperatures were derived photometrically by \citet{Berger_2020a_catalog}.
+    with previously reported rotation periods (dark gray points).  The
+    rotation periods were primarily measured by
+    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
+    suitable for gyrochronology (black points), based on factors
+    including their temperatures, surface gravities, non-binarity, and
+    proximity to the main sequence (see Section~\ref{subsec:flags}).
+    Some are ``confirmed'' or ``candidate'' Kepler Objects of Interest
+    that meet an additional set of planetary quality criteria (blue
+    points; Section~\ref{subsec:plflags}).  Surface gravities and
+    effective temperatures were derived photometrically by
+    \citet{Berger_2020a_catalog}.
 	}
@@
 This work is focused on Kepler stars for which ages can be inferred
-using either rotation, lithium, or both.  These ``age-dateable
-stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
-periods have been reported for $\approx$one in three Kepler targets
+using either rotation, lithium, or both.  These ``age-dateable stars''
+are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
+have been reported for $\approx$one in three Kepler targets
 \citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
-suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet
-have only been acquired for the Kepler objects of interest
-(KOIs), which comprise a few percent of Kepler's targets.
-In the following, we will describe the set of stars that we
-adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
-the set of objects we adopt as planets
+suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet have
+only been acquired for the Kepler objects of interest (KOIs), which
+comprise a few percent of Kepler's targets.  In the following, we will
+describe the set of stars that we adopt with measured rotation periods
+(Section~\ref{subsec:rotsel}), the set of objects we adopt as planets
 (Section~\ref{subsec:planetsel}), and the set of planets with
 high-resolution spectra suitable for lithium analysis
-(Section~\ref{subsec:lithiumsel}).
-Figure~\ref{fig:stellarprops} provides a visualization of the various samples.
+(Section~\ref{subsec:lithiumsel}).  Figure~\ref{fig:stellarprops}
+provides a visualization of the various samples.
 
@@
     planet hosts (right)}.  
-		The gray lines are ``mean fits'' 
-		to the slow rotation sequences of open
-		clusters.  The stellar sample (left) includes only apparently single stars
-		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
-		and temperatures of 3800--6200\,K.
-		The color map has a linear stretch.  %between one and ten stars per bin.
-		The planet-hosts (right) require
-		the same stellar cuts, and include only the confirmed and candidate planets
-		described in Section~\ref{subsec:plflags}.
-		An order of magnitude estimate for the number of young planets discovered
-		by Kepler follows by counting planets below the mean rotational isochrones.
+    The gray lines are ``mean fits'' to the slow rotation sequences of
+    open clusters.  The stellar sample (left) includes only apparently
+    single stars near the main sequence with $\log g$$>$4.2,
+    RUWE$<$1.4, and temperatures of 3800--6200\,K.  The color map has
+    a linear stretch.  The planet-hosts (right) require the same
+    stellar cuts, and include only the confirmed and candidate planets
+    described in Section~\ref{subsec:plflags}.  An order of magnitude
+    estimate for the number of young planets discovered by Kepler
+    follows by counting planets below the mean rotational isochrones.
 	}
@@
 
-To select stars with rotation periods, we turn to previous work.
-Many investigators have derived rotation periods both specifically for
-KOIs
+To select stars with rotation periods, we turn to previous work.  Many
+investigators have derived rotation periods both specifically for KOIs
 \citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
 and also for the broader set of all Kepler target stars
-\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
-These studies used a range of detection methods and selection
-functions.
-We are interested in understanding the age distribution of all Kepler
-target stars irrespective of KOI status.
-Out of these studies, the only homogeneous analysis of both Kepler
-targets and KOIs were the studies by \citet{Santos_2019,Santos_2021}
-(\citetalias{Santos_2019} and \citetalias{Santos_2021}).
-\citetalias{Santos_2019} and
+\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.  These
+studies used a range of detection methods and selection functions.  We
+are interested in understanding the age distribution of all Kepler
+target stars irrespective of KOI status.  Out of these studies, the
+only homogeneous analysis of both Kepler targets and KOIs were the
+studies by \citet{Santos_2019,Santos_2021} (\citetalias{Santos_2019}
+and \citetalias{Santos_2021}).  \citetalias{Santos_2019} and
 \citetalias{Santos_2021} combined a wavelet analysis and
 autocorrelation-based approach, and cumulatively reported rotation
-periods for 55{,}232 main-sequence and subgiant FGKM stars.
-They included known KOIs and binaries in their analysis, and assigned
-them specific quality flags, while removing transits and eclipses
-during the stellar rotation measurement process.
+periods for 55{,}232 main-sequence and subgiant FGKM stars.  They
+included known KOIs and binaries in their analysis, and assigned them
+specific quality flags, while removing transits and eclipses during
+the stellar rotation measurement process.
 
@@
 those reported by \citetalias{McQuillan_2014}.
-%A relevant flag for
-%stars missing \citetalias{McQuillan_2014} periods is included in
-%Table~\ref{tab:stars}.
 
@@
 \citetalias{Santos_2021} rotation periods for the KOI hosts, we
-noticed that some known KOIs with rotation periods were missing.
-This is not surprising, in that the rotation periods of KOIs have
-received considerably more scrutiny than those of ordinary Kepler
-stars.
-We therefore decided to split our subsequent analysis into a
-homogeneous portion that used only the \citetalias{Santos_2019} and
+noticed that some known KOIs with rotation periods were missing.  This
+is not surprising, in that the rotation periods of KOIs have received
+considerably more scrutiny than those of ordinary Kepler stars.  We
+therefore decided to split our subsequent analysis into a homogeneous
+portion that used only the \citetalias{Santos_2019} and
 \citetalias{Santos_2021} data, and an inhomogeneous portion that also
-considered a broader set of available KOI rotation periods.
-For the latter portion, we first included 32 KOIs with orbital and
-rotation periods within $\approx$20\% that had been
+considered a broader set of available KOI rotation periods.  For the
+latter portion, we first included 32 KOIs with orbital and rotation
+periods within $\approx$20\% that had been
 excluded %\footnote{Excluding stars in this category would impose a bias
@@
 %based on membership in RSG-5 \citep{Bouma_2022b}.  }
-from the
-\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
-(A.~Santos, private communication).
-We then incorporated an additional \nnewdavidtwentyone\ rotation
-periods for KOI hosts that
+from the \citetalias{Santos_2019} and \citetalias{Santos_2021}
+catalogs (A.~Santos, private communication).  We then incorporated an
+additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
 \citet{David_2021} had described as either ``reliable'' or ``highly
@@
 rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
-and \citet{Angus_2018}.
-Inclusion of these additional KOI rotation periods is a
-supplementary measure aimed at completeness in our final KOI age
-catalog; the provenances of the individual adopted periods are 
+and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
+periods is a supplementary measure aimed at completeness in our final
+KOI age catalog; the provenances of the individual adopted periods are
 noted in the relevant tables.
 
-Finally, since our scope is focused on rotation-based age dating, 
-we restricted our attention to stars with reported $P_{\rm
-rot}$$<$45\,days.
-FGK stars with slower rotation periods spin much slower than the stars
-in the oldest open clusters currently used to calibrate our 
-gyrochronology model.
-Figure~\ref{fig:prot_vs_teff} shows the resulting
-\nuniqstarsantosrot\ Kepler stars with rotation periods 
-compiled from
-\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our extended
-KOI list.
-
-To assess the statistical uncertainties of these rotation
-periods, we compared our adopted \citetalias{Santos_2019} and
+Finally, since our scope is focused on rotation-based age dating, we
+restricted our attention to stars with reported $P_{\rm
+rot}$$<$45\,days.  FGK stars with slower rotation periods spin much
+slower than the stars in the oldest open clusters currently used to
+calibrate our gyrochronology model.  Figure~\ref{fig:prot_vs_teff}
+shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
+periods compiled from \citetalias{Santos_2019},
+\citetalias{Santos_2021}, and our extended KOI list.
+
+To assess the statistical uncertainties of these rotation periods, we
+compared our adopted \citetalias{Santos_2019} and
 \citetalias{Santos_2021} periods with those reported by
-\citet{McQuillan_2014}.  We found that for
-$P_{\rm rot}\lesssim$15\,days, the two datasets agree at a precision
-of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
+\citet{McQuillan_2014}.  We found that for $P_{\rm
+rot}\lesssim$15\,days, the two datasets agree at a precision of
+$\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
 rot}\approx$30\,days, the agreement was typically at the
@@
 $P_{\rm rot}=15$\,days, and a linear increase thereafter, with slope
-set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods
-of 30 days.
+set to require 3\% $P_{\rm rot}$ uncertainties at rotation periods of
+30 days.
 
@@
 
-We considered planets in the NEA cumulative KOI
-table, which included the best knowledge available on any given planet
-candidate as of 6 June 2023, while also incorporating human-based
-vetting. %\footnote{See
-%\url{https://exoplanetarchive.ipac.caltech.edu/docs/PurposeOfKOITable.html}}
-These planets represent a superset of those in the fully automated
-Q1-Q17 DR25 KOI Table \citep{Thompson_2018}, which could be adopted in
-future work for planet occurrence rate calculations.  This version of
-the cumulative KOI table included \nkoisnofp\ objects that are either
+We considered planets in the NEA cumulative KOI table, which included
+the best knowledge available on any given planet candidate as of 6
+June 2023, while also incorporating human-based vetting.  These
+planets represent a superset of those in the fully automated Q1-Q17
+DR25 KOI Table \citep{Thompson_2018}, which could be adopted in future
+work for planet occurrence rate calculations.  This version of the
+cumulative KOI table included \nkoisnofp\ objects that are either
 ``confirmed'' or ``candidate'' planets, after excluding known false
@@
 The final piece of our analysis involves assessing ages based on the
-\ion{Li}{1} 6708\,\AA\ doublet.  Specifically,
-we analyzed spectra from the High Resolution Echelle
-Spectrometer (HIRES; \citealt{vogt_hires_1994}) on the Keck I 10m
-telescope.  These spectra were primarily collected through the
-California Kepler Survey
+\ion{Li}{1} 6708\,\AA\ doublet.  Specifically, we analyzed spectra
+from the High Resolution Echelle Spectrometer (HIRES;
+\citealt{vogt_hires_1994}) on the Keck I 10m telescope.  These spectra
+were primarily collected through the California Kepler Survey
 \citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
@@
 $\approx$three quarters of the spectra in our sample.  However, new
-spectra have since been acquired, and our approach and selection function are different.
-We therefore perform our own processing of the reduced HIRES spectra.
-
-We queried the CPS archive to collect all blaze-corrected
-HIRES spectra that were available for all non-false positive Kepler
-Objects of Interest with a MES$>$10.  This yielded at least one
-spectrum for \nlithiumstars\ stars and \nlithiumplanets\ planets.  For
-stars with multiple spectra available at varying ranges of
-signal-to-noise (S/N), we analyzed only the spectrum with the highest
-number of counts.  The resulting spectra were acquired between 6
-September 2009 and 16 May 2024, and cumulatively comprise
-\nhireshours\,hours of open-shutter time.
-
-We measured the lithium equivalent widths using a procedure
-adapted from previous work \citep{Bouma_2021}.  Our stars of interest
-are FGK stars, and so the continuum in the vicinity of the \ion{Li}{1}
+spectra have since been acquired, and our approach and selection
+function are different.  We therefore perform our own processing of
+the reduced HIRES spectra.
+
+We queried the CPS archive to collect all blaze-corrected HIRES
+spectra that were available for all non-false positive Kepler Objects
+of Interest with a MES$>$10.  This yielded at least one spectrum for
+\nlithiumstars\ stars and \nlithiumplanets\ planets.  For stars with
+multiple spectra available at varying ranges of signal-to-noise (S/N),
+we analyzed only the spectrum with the highest number of counts.  The
+resulting spectra were acquired between 6 September 2009 and 16 May
+2024, and cumulatively comprise \nhireshours\,hours of open-shutter
+time.
+
+We measured the lithium equivalent widths using a procedure adapted
+from previous work \citep{Bouma_2021}.  Our stars of interest are FGK
+stars, and so the continuum in the vicinity of the \ion{Li}{1}
 6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
@@
 of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
-consistent).  We continuum-normalized by fitting a
-third-degree Chebyshev series, while excluding regions with
-absorption lines.  We then numerically integrated the resulting
-spectrum using a one-component Gaussian with free amplitude, width,
-and mean, and estimated uncertainties on the line width through a
-Monte Carlo procedure that bootstraped against the local scatter in
-the continuum.
+consistent).  We continuum-normalized by fitting a third-degree
+Chebyshev series, while excluding regions with absorption lines.  We
+then numerically integrated the resulting spectrum using a
+one-component Gaussian with free amplitude, width, and mean, and
+estimated uncertainties on the line width through a Monte Carlo
+procedure that bootstraped against the local scatter in the continuum.
 
@@
 \ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
-precision of our measurement technique, after applying an
-initial iteration on the \nlithiumstars\ KOIs with spectra, we
-compared our lithium equivalent widths with those reported by
+precision of our measurement technique, after applying an initial
+iteration on the \nlithiumstars\ KOIs with spectra, we compared our
+lithium equivalent widths with those reported by
 \citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
@@
 required positive EWs, while we allow for negative ones.  At
-$>$30\,m\AA, there is however a small offset in our respective
-scales, such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is
-caused by our non-treatment of the iron blend.  We therefore directly
-subtract the constant mean value noted above when calculating
-lithium-based ages.
+$>$30\,m\AA, there is however a small offset in our respective scales,
+such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is caused by
+our non-treatment of the iron blend.  We therefore directly subtract
+the constant mean value noted above when calculating lithium-based
+ages.
 
@@
 \texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
-the MIST isochrone grids
-\citep{Choi_2016,2016ApJS..222....8D}, given the SDSS $g$
-and 2MASS $K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and
-$V$-band extinction from the \citet{2018MNRAS.478..651G} reddening
-map.  The resulting stellar parameters are available for
-\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
-stars with rotation periods.  For the remaining
-\fracstarswithprotwithoutbtwenty\ of stars that lack temperatures and
-surface gravities from \citetalias{Berger_2020a_catalog}, we adopt the
-values reported by \citet{Santos_2019} and \citet{Santos_2021}, which
-for these cases primarily derive from the \citet{Mathur_2017} DR25
-Kepler Stellar Properties Catalog, and are mostly derived from
-photometry.  In the planet sample, \frackoisnofpwithprotwithbtwenty\
-of the non-false-positive KOIs with rotation periods have parameters
-from \citet{Berger_2020a_catalog}, and the remainder similarly are
-drawn from DR25.
+the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
+the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
+parallaxes, and $V$-band extinction from the
+\citet{2018MNRAS.478..651G} reddening map.  The resulting stellar
+parameters are available for \fracstarswithprotwithbtwenty\ of the
+\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
+remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
+temperatures and surface gravities from
+\citetalias{Berger_2020a_catalog}, we adopt the values reported by
+\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
+primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
+Properties Catalog, and are mostly derived from photometry.  In the
+planet sample, \frackoisnofpwithprotwithbtwenty\ of the
+non-false-positive KOIs with rotation periods have parameters from
+\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
+from DR25.
 
@@
 
-% optional todo's:
-%TODO: plot uncertainties in teff vs teff...
-%TODO: plot uncertainties in logg vs logg...
-%TODO: add that plot showing differences btwn the two teff scales?
-
 \section{Ages From Rotation}
@@
 		{\bf The models.}
-		Points represent $10^4$ draws from models
-		that have been fitted to rotation periods
-		\citep{Bouma_2023} and lithium equivalent widths
+    Points represent $10^4$ draws from models that have been fitted to
+    rotation periods \citep{Bouma_2023} and lithium equivalent widths
     \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
-		are the ``mean models'' at various ages; the intrinsic
-		dispersion in rotation and lithium about these mean models, which
-		is what the models fit, sets the theoretical precision floor for
-		the age-dating methods.    Additional sources of uncertainty,
+    are the ``mean models'' at various ages; the intrinsic dispersion
+    in rotation and lithium about these mean models, which is what the
+    models fit, sets the theoretical precision floor for the
+    age-dating methods.    Additional sources of uncertainty,
     including measurement uncertainty, impose further limits on
@@
 \citep{Bouma_2023}, which is a method designed to address the fact
-that stars
-with the same mass and same rotation period can have very different
-ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
-posterior should therefore incorporate this intrinsic population-level
-scatter into the precision with which the age can be inferred.
-Figure~\ref{fig:models} highlights this problem, particularly in
-regions where the 0.1-1\,Gyr stars overlap.
+that stars with the same mass and same rotation period can have very
+different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
+age posterior should therefore incorporate this intrinsic
+population-level scatter into the precision with which the age can be
+inferred.  Figure~\ref{fig:models} highlights this problem,
+particularly in regions where the 0.1-1\,Gyr stars overlap.
 
@@
 stellar spin-down rate as a function of time and stellar temperature,
-the framework interpolates between reference open clusters using
-cubic hermite polynomials.  We calculate the probability of the
+the framework interpolates between reference open clusters using cubic
+hermite polynomials.  We calculate the probability of the
 rotation-based age $t_{\rm gyro}$, given the observed rotation
@@
 \label{subsec:flags}
-We calculated gyrochrone age posteriors for all
-\nuniqstarsantosrotteffcut\ stars with reported rotation periods and
-effective temperatures of 3800--6200\,K.  To flag stars for
-which we suspect these ages may not in fact be valid, we built a set
-of quality flags, which we condensed into a single binary number,
-$Q_{\rm star}$.  When and how the flags represented in this should be
-applied depends on the question being asked.  If the goal is to
-construct a false-positive free sample, they could all be applied.  If
-the goal is to construct a complete sample, then consider the examples
-of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
-($\approx$625\,Myr).  The former has a high RUWE due to a resolved
-binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
-\citep{2014ApJ...783...53M}.  We leave selection for or against such
-cases as a decision to the user.
-
-Three assumptions must hold for a
-rotation-based age dating method like \texttt{gyro-interp} to be
-valid.  {\it 1)} The evolutionary state of the star must be
-well-specified by its temperature and its age, {\it 2)} the star's
-spindown must not be influenced by binary companions, and {\it 3)} the
-rotation period distribution for field stars of a given temperature
-and age must be identical to that of equally aged open clusters
-(metallicity differences, for instance, are ignored).  A rephrasing of
-the first condition is that the star must be ``near the main
-sequence'' because during the post main sequence 
+We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
+stars with reported rotation periods and effective temperatures of
+3800--6200\,K.  To flag stars for which we suspect these ages may not
+in fact be valid, we built a set of quality flags, which we condensed
+into a single binary number, $Q_{\rm star}$.  When and how the flags
+represented in this should be applied depends on the question being
+asked.  If the goal is to construct a false-positive free sample, they
+could all be applied.  If the goal is to construct a complete sample,
+then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
+Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
+resolved binary companion \citep{Bouma_2022a}; the latter is on a
+grazing orbit \citep{2014ApJ...783...53M}.  We leave selection for or
+against such cases as a decision to the user.
+
+Three assumptions must hold for a rotation-based age dating method
+like \texttt{gyro-interp} to be valid.  {\it 1)} The evolutionary
+state of the star must be well-specified by its temperature and its
+age, {\it 2)} the star's spindown must not be influenced by binary
+companions, and {\it 3)} the rotation period distribution for field
+stars of a given temperature and age must be identical to that of
+equally aged open clusters (metallicity differences, for instance, are
+ignored).  A rephrasing of the first condition is that the star must
+be ``near the main sequence'' because during the post main sequence
 stellar temperatures change.  From $\approx$0.08 to $\approx$3\,Gyr,
@@
 ($\approx$3\%), because its outer layers expand as it begins hydrogen
-shell burning as a subgiant.  We treat this issue in a 
-manner discussed in ``bit 9'' below.
+shell burning as a subgiant.  We treat this issue in a manner
+discussed in ``bit 9'' below.
 
@@
 %crossmatch
-of the NEA
-\texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The separation
-distribution of the Kepler-Gaia DR3 crossmatches is such that 99.2\%
-of candidate matches are within 1$''$.   We nonetheless noted an
-upturn in the candidate match rate from 3-4$''$; such sources are
-flagged using bit 4.  For KIC stars with multiple potential Gaia
-matches within the 4$''$ radius, we adopted the brightest star
-as the default match.  In most such cases this was unambiguous because
-there is a large brightness difference between the primary and any
-apparent neighbors.  However cases with multiple stars within 4$''$
-within $\Delta G$$<$$0.5$\,mag are noted using bit flag 5.
+of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The
+separation distribution of the Kepler-Gaia DR3 crossmatches is such
+that 99.2\% of candidate matches are within 1$''$.   We nonetheless
+noted an upturn in the candidate match rate from 3-4$''$; such sources
+are flagged using bit 4.  For KIC stars with multiple potential Gaia
+matches within the 4$''$ radius, we adopted the brightest star as the
+default match.  In most such cases this was unambiguous because there
+is a large brightness difference between the primary and any apparent
+neighbors.  However cases with multiple stars within 4$''$ within
+$\Delta G$$<$$0.5$\,mag are noted using bit flag 5.
 
@@
 
-{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3 renormalized
-unit weight error
-(RUWE) as a function of other stellar parameters, %\footnote{\url{https://gea.esac.esa.int/archive/documentation/GDR3/Gaia_archive/chap_datamodel/sec_dm_main_source_catalogue/ssec_dm_gaia_source.html}})
-and flag stars with RUWE$>$1.4 as possible binaries.  Such astrometric
-outliers can be either bona fide astrometric binaries, or more often
-are marginally resolved point sources for which a single-source PSF
-model provides a poor fit.
+{\it RUWE (bit 7)}---We inspected diagrams of the Gaia DR3
+renormalized unit weight error (RUWE) as a function of other stellar
+parameters, and flag stars with RUWE$>$1.4 as possible binaries.  Such
+astrometric outliers can be either bona fide astrometric binaries, or
+more often are marginally resolved point sources for which a
+single-source PSF model provides a poor fit.
 
@@
 10)}---\citeauthor{Santos_2021} included in a flag for candidate
-``classical pulsators'' (e.g.\ Cepheids) and
-close-in binaries.  Visual inspection 
-% of the rotation periods as a function of temperature 
-shows that although this flag selects many bona fide objects in these
-classes, it also selects most $P_{\rm
-rot}$$\lesssim$5\,day bona fide young planets.  
-These objects are neither classical pulsators nor close
-binaries.  While we propagated this flag into our set of quality
-flags, we therefore generally ignore it.
-
-{\it Not in the homogeneous stellar rotation sample (bit 11)}---
-This flag is set for the KOIs with rotation periods
-drawn from a source other than \citet{Santos_2019,Santos_2021}.
-
-{\it How do we use the bit-flags?}
-By default, we assume that a star is
-suitable for gyrochronology if none of bits zero through nine
-(inclusive) are raised.
-For analyses that require a homogeneous stellar rotation selection
-function, we also require bit 11 to be zero.
+``classical pulsators'' (e.g.\ Cepheids) and close-in binaries.
+Visual inspection shows that although this flag selects many bona fide
+objects in these classes, it also selects most $P_{\rm
+rot}$$\lesssim$5\,day bona fide young planets.  These objects are
+neither classical pulsators nor close binaries.  While we propagated
+this flag into our set of quality flags, we therefore generally ignore
+it.
+
+{\it Not in the homogeneous stellar rotation sample (bit 11)}--- This
+flag is set for the KOIs with rotation periods drawn from a source
+other than \citet{Santos_2019,Santos_2021}.
+
+{\it How do we use the bit-flags?} By default, we assume that a star
+is suitable for gyrochronology if none of bits zero through nine
+(inclusive) are raised.  For analyses that require a homogeneous
+stellar rotation selection function, we also require bit 11 to be
+zero.
 
@@
 
-The rotation periods for the star and planet samples are shown
-in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
+The rotation periods for the star and planet samples are shown in
+Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
 observed by Kepler, requiring reported rotation periods and a reliable
 DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
-these stars to have temperatures of 3800--6200\,K, yields \nuniqstarsantosrotteffcut\
-stars.  Further requiring the stars to be apparently single, near the main
-sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
-eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
-potentially amenable for gyrochronology.  Finally, imposing RUWE$<$1.4
-and requiring no possible blends ($\Delta G$<2.5) within 4$''$ leaves
+these stars to have temperatures of 3800--6200\,K, yields
+\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
+apparently single, near the main sequence, with $\log g$$>$4.2,
+$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
+\nuniqstarsantosallbutruwe\ rotators potentially amenable for
+gyrochronology.  Finally, imposing RUWE$<$1.4 and requiring no
+possible blends ($\Delta G$<2.5) within 4$''$ leaves
 \nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
@@
 (\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
-If we additionally require the planets to be non-grazing ($b<0.8$),
-we are left with \nplwgyroagenograzing\ planets orbiting
+If we additionally require the planets to be non-grazing ($b<0.8$), we
+are left with \nplwgyroagenograzing\ planets orbiting
 \nplhoststarwgyroagenograzing\ stars.
@@
 To calculate lithium ages, we use \texttt{EAGLES} (git commit
-\texttt{ac09637}), which, similar to
-\texttt{gyro-interp}, is based on a purely empirical interpolation
-approach.  \texttt{EAGLES} however was calibrated on lithium measurements of
-stars observed by the
+\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
+a purely empirical interpolation approach.  \texttt{EAGLES} however
+was calibrated on lithium measurements of stars observed by the
 Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
@@
 
-%Certain quality flags described in Section~\ref{subsec:flags} are also
-%relevant for lithium ages, specifically those associated with
-%crowding and binarity.
-%For instance, rapid rotators are more common in binaries.
-%and they also
-%generally have higher lithium abundances than slow-rotating stars of the
-%same mass (CITE).
-%Stars for which rotation-based age dating is dubious heavily overlap
-%with those for which lithium-based ages are dubious.
-
-In part because many of our EWs are upper limits, 
-many of the lithium ages are lower limits.
-In such cases,  the inferred age posterior is strongly influenced by
-{\it i)} the assumed intrinsic dispersion in the lithium model,
-and {\it ii)} by the prior.
-In these instances, we quote the 3$\sigma$ (99.7$^{\rm th}$
-percentile) limits.
+In part because many of our EWs are upper limits, many of the lithium
+ages are lower limits.  In such cases,  the inferred age posterior is
+strongly influenced by {\it i)} the assumed intrinsic dispersion in
+the lithium model, and {\it ii)} by the prior.  In these instances, we
+quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
 
@@
   \vspace{-0.6cm}
-  \caption{{\bf Kepler's demographic cliff,} visible in the rotation-derived
-    ages of stars (left) and planet hosts (middle) in the Kepler field.
-    The top row corresponds to all
-    stars with temperatures of 3800--6200\,K for which we calculated rotation-based ages.
-    The darker lines impose quality cuts
-    on factors such as binarity, crowding, and the star's evolutionary
-    state.
-    The bottom row selects only stars with temperatures of 4400-5400\,K, which
-    have relatively precise ages near 2\,Gyr due to their fast spin-down
-    \citep[see][]{Bouma_2023}.
-    The statistical uncertainty for a mean individual star at 1, 2, and 3\,Gyr is shown, and is identical across each row.
-    Finally, the right panel compares the rotation-based age distributions against star formation histories derived using CMD fitting.
-    Completeness is near unity until $t$$\lesssim$3\,Gyr
-    \citep{2022ApJ...937...94M}.
+  \caption{{\bf Kepler's demographic cliff,} visible in the
+  rotation-derived ages of stars (left) and planet hosts (middle) in
+  the Kepler field.  The top row corresponds to all stars with
+  temperatures of 3800--6200\,K for which we calculated rotation-based
+  ages.  The darker lines impose quality cuts on factors such as
+  binarity, crowding, and the star's evolutionary state.  The bottom
+  row selects only stars with temperatures of 4400-5400\,K, which have
+  relatively precise ages near 2\,Gyr due to their fast spin-down
+  \citep[see][]{Bouma_2023}.  The statistical uncertainty for a mean
+  individual star at 1, 2, and 3\,Gyr is shown, and is identical
+  across each row.  Finally, the right panel compares the
+  rotation-based age distributions against star formation histories
+  derived using CMD fitting.  Completeness is near unity until
+  $t$$\lesssim$3\,Gyr \citep{2022ApJ...937...94M}.
   \label{fig:hist_tgyro}
@@
 What are the most notable confirmed planets with well-measured ages?
-These planets can be split into two classes:
-those in clusters, and those with independent and consistent
-rotation and lithium-based ages.
+These planets can be split into two classes: those in clusters, and
+those with independent and consistent rotation and lithium-based ages.
 
@@
 
-There are currently four known stellar ensembles in the Kepler field that have
-yielded a total of fourteen transiting planets.
-Our analysis blindly recovers each of these planets, as well as their ages.
-
-{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its most
-constraining age measurement from relative isochrone dating of
-pre-main-sequence M dwarfs; it contains Kepler-1627Ab,
-Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
-\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
-may vary in age by $\approx$50\% (\citealt{Bouma_2022b};
-Kerr~et.al.~in preparation).  The planets themselves are all on
-close-in orbits (5-25\,days), with sizes from 2-4\,$R_\oplus$.  The
-rotation, lithium, and cluster ages all agree for
-Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
+There are currently four known stellar ensembles in the Kepler field
+that have yielded a total of fourteen transiting planets.  Our
+analysis blindly recovers each of these planets, as well as their
+ages.
+
+{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
+most constraining age measurement from relative isochrone dating of
+pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
+Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
+Different sub-groups of the complex may vary in age by $\approx$50\%
+(\citealt{Bouma_2022b}; Kerr~et.al.~in preparation).  The planets
+themselves are all on close-in orbits (5-25\,days), with sizes from
+2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
+for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
 rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the cluster age reported by \citet{Bouma_2022b} for the RSG-5 subgroup.
-
-{\it MELANGE-3}---\citet{2022AJ....164...88B} reported a $105\pm10$\,Myr association in
-the Kepler field and derived its age using rotation,
-lithium, and the isochronal pre-main-sequence.  Kepler-970 and
-Kepler-1928 harbor two transiting mini-Neptunes inside this
-association \citep{2022AJ....164...88B}.  Our rotation-based
+above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
+subgroup.
+
+{\it MELANGE-3}---\citet{2022AJ....164...88B} reported a
+$105\pm10$\,Myr association in the Kepler field and derived its age
+using rotation, lithium, and the isochronal pre-main-sequence.
+Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
+this association \citep{2022AJ....164...88B}.  Our rotation-based
 age-dating independently recovers these systems, with $t_{\rm
 gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
-Kepler-970 and Kepler-1928 respectively.  The former rotation-based age
-being slightly older than the suggested Pleiades age for the
+Kepler-970 and Kepler-1928 respectively.  The former rotation-based
+age being slightly older than the suggested Pleiades age for the
 assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.
 
-{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for Kepler-968 and Kepler-52.
-These stars were reported by
-\citet{2019AJ....158..122K} to be members of an association called Theia~520
-based on their spatial and kinematic correlations with nearby stars.
-The core of this structure is also known as UBC-1
-\citep{2018A&A...618A..59C}; since both Kepler planet hosts
-are in diffuse regions of the cluster, we use the name from
-\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting, Theia~520
-is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
-An independent analysis by
-\citet{2024A&A...681A..13F} using g-mode asteroseismology,
-gyrochronology, and isochrone fitting found consistent results.
-Ongoing work by J.~Curtis et al.~broadly finds that the two
-planet-hosting stars are indeed members of the diffuse population.
-Given the cluster-level and individual-star evidence, this makes
-Kepler-968 and Kepler-52 the youngest multiplanet systems currently
-known from the primary Kepler mission.
-
-{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of Kepler-66 and -67b, two
-mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).  We report rotation-based
-ages for these systems of \kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for
-Kepler-66 and -67b, respectively.  Kepler-66 ($T_{\rm
-eff}$$\approx$5900\,K, $P_{\rm rot}$$\approx$10.4\,days) has a lower
-precision and a more asymmetric posterior because the spindown rates
-of stars this hot slow down after the first gigayear.
-Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
+{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for
+Kepler-968 and Kepler-52.  These stars were reported by
+\citet{2019AJ....158..122K} to be members of an association called
+Theia~520 based on their spatial and kinematic correlations with
+nearby stars.  The core of this structure is also known as UBC-1
+\citep{2018A&A...618A..59C}; since both Kepler planet hosts are in
+diffuse regions of the cluster, we use the name from
+\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting,
+Theia~520 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
+An independent analysis by \citet{2024A&A...681A..13F} using g-mode
+asteroseismology, gyrochronology, and isochrone fitting found
+consistent results.  Ongoing work by J.~Curtis et al.~broadly finds
+that the two planet-hosting stars are indeed members of the diffuse
+population.  Given the cluster-level and individual-star evidence,
+this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
+currently known from the primary Kepler mission.
+
+{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of
+Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
+We report rotation-based ages for these systems of
+\kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
+-67b, respectively.  Kepler-66 ($T_{\rm eff}$$\approx$5900\,K, $P_{\rm
+rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
+posterior because the spindown rates of stars this hot slow down after
+the first gigayear.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
 rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
@@
     provenances.  Gray points are transiting planets from the NASA
-    Exoplanet Archive older than 1\,Gyr.  If one were to instead select for
-    precise ages ($t/\sigma_{\rm t}>3$), this would hide $\approx$ten
-    $<$0.5\,Gyr planets, and add $\approx$fifteen 0.5-1\,Gyr planets.
-    For this plot, we require our Kepler stellar hosts to have no
-    quality flags raised, and for the planets to
+    Exoplanet Archive older than 1\,Gyr.  If one were to instead
+    select for precise ages ($t/\sigma_{\rm t}>3$), this would hide
+    $\approx$ten $<$0.5\,Gyr planets, and add $\approx$fifteen
+    0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
+    hosts to have no quality flags raised, and for the planets to
     similarly no have quality flags raised. 
@@
 confirmed planets for which a sub-gigayear age is reported by at least
-one age indicator.  All but two systems are
-either evolved stars or else unresolved binaries, and are
-automatically flagged as such.  The two exceptions are Kepler-786
-(KOI-1199), and Kepler-1 (TrES-2).  Kepler-786 is an early K dwarf with
-$t_{\rm Li}$=\kepseveneightsix\,Myr, which would imply that
-rotation should be detectable.  However no photometric rotation signal is detected; other age
-indicators in the spectrum similarly suggest that the star may not in
-fact be young.  TrES-2 is therefore a ``lithium-young'' system for
-which we believe the high lithium EW may be
-a statistical fluke.
+one age indicator.  All but two systems are either evolved stars or
+else unresolved binaries, and are automatically flagged as such.  The
+two exceptions are Kepler-786 (KOI-1199), and Kepler-1 (TrES-2).
+Kepler-786 is an early K dwarf with $t_{\rm
+Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
+detectable.  However no photometric rotation signal is detected; other
+age indicators in the spectrum similarly suggest that the star may not
+in fact be young.  TrES-2 is therefore a ``lithium-young'' system for
+which we believe the high lithium EW may be a statistical fluke.
 
@@
 Let us restrict our attention to blemish-free stars, with blemish-free
-``confirmed'' planets:
-$Q_{\rm star}$ must not have any of bits zero through nine
-raised, and $Q_{\rm planet}$ must have no quality flags raised at all.
-Under this constraint,
-our results include \ltonegyrhighqconfirmedtwosided\ systems
-younger than 1\,Gyr with two-sided $t_{\rm gyro}$ and $t_{\rm Li}$,
-and \ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$ provides
-an age constraint while $t_{\rm Li}$ is one-sided.
+``confirmed'' planets: $Q_{\rm star}$ must not have any of bits zero
+through nine raised, and $Q_{\rm planet}$ must have no quality flags
+raised at all.  Under this constraint, our results include
+\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
+two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
+\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
+provides an age constraint while $t_{\rm Li}$ is one-sided.
 %Two of the former systems are in clusters (Kepler-1643 and Kepler-1928);
@@
 They include \nearthshighqconfirmed\ Earth-sized or smaller planets,
-no Jupiter-sized or larger planets,
-\nsubsaturnshighqconfirmed\ ``super-Neptune'' ($>$4\,$R_\oplus$) sized
-planets,
-\nminineptuneshighqconfirmed\ mini-Neptunes above the 
-radius valley (as parametrized by \citealt{2018MNRAS.479.4786V}),
-and \nsuperearthshighqconfirmed\ super-Earth sized planets below the radius valley.
-While most of these planets are on orbital periods below 50\,days,
-\nlongperiodhighqconfirmed\ are on more distant orbits.
-
-%The new young planets
-%include five 1.8-3.8\,$R_\oplus$ mini-Neptunes
-%(Kepler-1529b $\approx$100\,Myr,
-%-1521b $\approx$195\,Myr,
-%-1467b $\approx$235\,Myr,
-%-955b $\approx$400\,Myr, and
-%-523b $\approx$440\,Myr),
-%a single Neptune-sized planet
-%(Kepler-971b $\approx$335\,Myr),
-%five 1.1-1.8\,$R_\oplus$ super-Earths
-%(Kepler-1930b $\approx$150\,Myr,
-%-1313b $\approx$185\,Myr,
-%-1565b $\approx$205\,Myr,
-%-1532b $\approx$400\,Myr,
-%-1481b $\approx$400\,Myr),
-%and three Earth or sub-Earth-sized planets
-%(Kepler-1933b $\approx$300\,Myr,
-%-1561b $\approx$420\,Myr,
-%-1482b $\approx$490\,Myr).
+no Jupiter-sized or larger planets, \nsubsaturnshighqconfirmed\
+``super-Neptune'' ($>$4\,$R_\oplus$) sized planets,
+\nminineptuneshighqconfirmed\ mini-Neptunes above the radius valley
+(as parametrized by \citealt{2018MNRAS.479.4786V}), and
+\nsuperearthshighqconfirmed\ super-Earth sized planets below the
+radius valley.  While most of these planets are on orbital periods
+below 50\,days, \nlongperiodhighqconfirmed\ are on more distant
+orbits.
 
 A few new highlights from the high quality stars and planets include
-Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.
-Kepler-1529b is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured
-rotation and lithium ages.
-Kepler-1565b is an analogous $\approx$170-230\,Myr super-Earth.
-Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr near-Solar analog with
-two transiting planets: an Earth-sized planet on a one-day orbit,
-and a mini-Neptune on a five-day orbit.
-The Earth-sized planet, Kepler-1312c, along with the near-identical
-Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), 
-rank among the
+Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
+is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation
+and lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
+super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
+near-Solar analog with two transiting planets: an Earth-sized planet
+on a one-day orbit, and a mini-Neptune on a five-day orbit.  The
+Earth-sized planet, Kepler-1312c, along with the near-identical
+Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
 youngest Earth-sized planets currently known, comparable to systems
 such as HD~63433d \citep[1.1\,$R_\oplus$,
-$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b \citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
+$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b
+\citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
 Kepler-1629~b, with a size just two-thirds that of Earth
-(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on rotation
-(\kepsixteentwonine\,Myr).  The lithium age from a reconnaissance 
-TRES spectrum confirms the point.
-
-%Of these planets, one specific highlight is Kepler-1933b, a 4.9\,day
-%1.0\,$R_\oplus$ Earth-sized planet.  Its rotation-based age of
-%$304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of its lithium age,
-%though at half the relative uncertainty.
-
-%A less clear case is Kepler-447b, which is supposedly a 32\,$R_\oplus$
-%planet on a 7.8\,day grazing orbit.  Rotation and lithium independently
-%find a 420-440\,Myr age.  {\it LGB TODO What is up with the size?}.
+(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on
+rotation (\kepsixteentwonine\,Myr).  The lithium age from a
+reconnaissance TRES spectrum confirms the point.
 
@@
 
-%\subsubsection{Need spectra}
-%Kepler-1938...
-%Kepler-762... (a HJ!)
-%Kepler-1320
-
 \section{Discussion}
@@
 
-Most of the stars observed by Kepler are in the thin disk.
-This can be verified by following \citet{Gaia_2018}, and denoting stars with 2D tangential
-velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members, and those
-with 60$<$$v_{\rm T}$$<$150\,\kms as thick disk members.  For all
-stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick disk stars respectively.  
-The thin disk stars have a threefold larger detected rotation period
-fraction: 35{,}675 rotators are in the thin disk, while 7{,}312 are in
-the thick disk.  Further imposing the gyrochronology quality flags
-discussed in Section~\ref{subsec:flags} yields 17{,}755
-thin-disk stars for which gyrochronology is nominally applicable, and
-2{,}462 thick-disk stars.
-
-Classifying Kepler stars in this manner helps connect
-them to the broader literature focused
-on the history of star formation in the Galaxy.  On extragalactic
-scales, the star formation rate in spiral galaxies with mass
-comparable to the Milky Way peaked $\approx$10\,Gyr ago, and has since
-decreased by an order of magnitude
+Most of the stars observed by Kepler are in the thin disk.  This can
+be verified by following \citet{Gaia_2018}, and denoting stars with 2D
+tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
+and those with 60$<$$v_{\rm T}$$<$150\,\kms as thick disk members.
+For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
+disk stars respectively.  The thin disk stars have a threefold larger
+detected rotation period fraction: 35{,}675 rotators are in the thin
+disk, while 7{,}312 are in the thick disk.  Further imposing the
+gyrochronology quality flags discussed in Section~\ref{subsec:flags}
+yields 17{,}755 thin-disk stars for which gyrochronology is nominally
+applicable, and 2{,}462 thick-disk stars.
+
+Classifying Kepler stars in this manner helps connect them to the
+broader literature focused on the history of star formation in the
+Galaxy.  On extragalactic scales, the star formation rate in spiral
+galaxies with mass comparable to the Milky Way peaked $\approx$10\,Gyr
+ago, and has since decreased by an order of magnitude
 \citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
@@
 and by modeling the local white dwarf luminosity function
-\citep[e.g.][]{2019ApJ...878L..11I}.  
-\citet{2019A&A...624L...1M} and \citet{2019ApJ...878L..11I} for
-instance focused on stars (and white dwarfs) within a few hundred
-parsecs, and both found a peak in the local star
-formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R} focused on
-stars in a wider 2\,kpc bubble, and reported three local maxima
-in the thin disk's star formation history 5.7, 1.9, and 1.0\,Gyr ago,
-which they associate with pericenter passages of the Sagittarius satellite
-galaxy.
-
-The right column of Figure~\ref{fig:hist_tgyro} compares our derived age distribution
-against star formation histories (SFHs) reported by
+\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
+\citet{2019ApJ...878L..11I} for instance focused on stars (and white
+dwarfs) within a few hundred parsecs, and both found a peak in the
+local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
+focused on stars in a wider 2\,kpc bubble, and reported three local
+maxima in the thin disk's star formation history 5.7, 1.9, and
+1.0\,Gyr ago, which they associate with pericenter passages of the
+Sagittarius satellite galaxy.
+
+The right column of Figure~\ref{fig:hist_tgyro} compares our derived
+age distribution against star formation histories (SFHs) reported by
 \citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
-CMD fitting.
-The overall slope of both SFHs broadly agrees with what we find from rotation-based ages.
-The \citet{2019A&A...624L...1M} data in particular seem entirely consistent, after
-accounting for the statistical uncertainties reported in that study.
-Regarding the episodic star formation bursts reported by
-\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK star sample (top row).
-However, rotation-based ages are most precise for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
-\citep{Bouma_2023}.
-The lower row of Figure~\ref{fig:hist_tgyro} selects these stars with a temperature cut, and
-may show a hint of the $\approx$2\,Gyr spike reported by 
-\citet{2020NatAs...4..965R}.
-The $\approx$1\,Gyr spike however is not recovered.
+CMD fitting.  The overall slope of both SFHs broadly agrees with what
+we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
+data in particular seem entirely consistent, after accounting for the
+statistical uncertainties reported in that study.  Regarding the
+episodic star formation bursts reported by
+\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
+star sample (top row).  However, rotation-based ages are most precise
+for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
+\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
+selects these stars with a temperature cut, and may show a hint of the
+$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
+$\approx$1\,Gyr spike however is not recovered.
 
@@
 Most of our ages come from one source of information: rotation.
-Factors other than stellar age and mass can influence rotation rates of main sequence stars.
-Binarity is one example: even intermediate-separation ($\sim$10\,AU) binaries are biased
-toward rapid rotation
-\citep[e.g.][and many studies thereafter]{Meibom_2007}. %,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
+Factors other than stellar age and mass can influence rotation rates
+of main sequence stars.  Binarity is one example: even
+intermediate-separation ($\sim$10\,AU) binaries are biased toward
+rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
 Metallicity may also be a second-order effect
-\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
-In this work, we were interested in not only the overall
-age distribution of the Kepler field, but also in the reliability of individual ages for
-young planet hosting stars.
-The quality flagging scheme in Table~\ref{tab:planets} provides one path toward this end.
-The other method is to consider the region in which we do have
-overlapping sensitivity between lithium and rotation -- 
-G and K dwarfs with $t$$\lesssim$1\,Gyr.  
-Our results for such stars include
-\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
-stars for which the two age indicators are both constraining and in agreement
-(with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
-Conversely, we found only one system, Kepler-786, with a 
-precise lithium age (\kepseveneightsix\,Myr), but no rotation signal.
-
-Other age indicators may help in verifying the ages of
-the $\lesssim$3\,Gyr stars that are the focus of this work.
-Specifically, chromospheric emission in the
-X-ray, \ion{Ca}{2} HK line, and the UV can serve as an age tracer
+\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
+were interested in not only the overall age distribution of the Kepler
+field, but also in the reliability of individual ages for young planet
+hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
+provides one path toward this end.  The other method is to consider
+the region in which we do have overlapping sensitivity between lithium
+and rotation -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
+for such stars include \ltonegyrhighqconfirmedtwosided\ planets
+orbiting ``high-quality'' stars for which the two age indicators are
+both constraining and in agreement (with $Q_{\rm star}$ bits zero
+through nine not raised, and no planet quality flags raised).
+Conversely, we found only one system, Kepler-786, with a precise
+lithium age (\kepseveneightsix\,Myr), but no rotation signal.
+
+Other age indicators may help in verifying the ages of the
+$\lesssim$3\,Gyr stars that are the focus of this work.  Specifically,
+chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
+can serve as an age tracer
 \citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
-However, these age indicators are all in a sense ``rotation-powered''.  The dynamo
-converts kinetic energy into magnetic energy, which is emitted through these chromospheric
-pathways.
-We did not attempt to incorporate these indicators into this study
-due to concerns regarding sensitivity, homogeneity, and
-the question of whether they in fact provide age
-information that is truly independent from rotation.
-
-A broader question, beyond the scope of this work, concerns the ages of
-the older $\gtrsim$3\,Gyr stars that we do not treat.
-For age-dating models based only on rotation rates, or else
-on a combination of rotation and spin-modulation amplitude
-\citep{2023ApJ...952..131M}, an important challenge is that the detectability
-of the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
-is at the limits of the current data \citep{2022ApJ...937...94M}.
-In the ``old star'' regime, methods 
+However, these age indicators are all in a sense ``rotation-powered''.
+The dynamo converts kinetic energy into magnetic energy, which is
+emitted through these chromospheric pathways.  We did not attempt to
+incorporate these indicators into this study due to concerns regarding
+sensitivity, homogeneity, and the question of whether they in fact
+provide age information that is truly independent from rotation.
+
+A broader question, beyond the scope of this work, concerns the ages
+of the older $\gtrsim$3\,Gyr stars that we do not treat.  For
+age-dating models based only on rotation rates, or else on a
+combination of rotation and spin-modulation amplitude
+\citep{2023ApJ...952..131M}, an important challenge is that the
+detectability of the rotation signals for 3-10\,Gyr stars, especially
+Sun-like stars, is at the limits of the current data
+\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods
 using either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, or else
-using both the evolution of
-stellar luminosity and stellar rotation \citep{Angus_2019} seem the most
-capable of providing useful age constraints for single stars.
-Incorporating kinematics \citep{2021AJ....161..189L} seems likely to remain
-most useful for stellar ensembles, due to the stochastically of kinematic heating.
+\citep{vanSaders_2016,2024ApJ...962..138S}, or else using both the
+evolution of stellar luminosity and stellar rotation
+\citep{Angus_2019} seem the most capable of providing useful age
+constraints for single stars.  Incorporating kinematics
+\citep{2021AJ....161..189L} seems likely to remain most useful for
+stellar ensembles, due to the stochastically of kinematic heating.
 
 \subsection{Future Directions}
-{\it Occurrence Rates}: Some interesting
-age-evolution trends have been predicted, and can probably be teased
-from our data.  
-Kelvin-Helmholtz cooling for mini-Neptunes should scale $R_{\rm p}$ as
-$t^{-0.1}$ \citep{Gupta_2019}.
-The photoevaporation desert has been predicted to be ``carved'' over the timescale
-for atmospheric loss \citep{Owen_Lai_2018}.
-Mini-Neptunes should be converted into super-Earths \citep{Rogers_2021}.
-And the boundary of the super-Earth population may correspondingly move up in size over
-time \citep{David_2021}.
-We're going to split the two papers.
+{\it Occurrence Rates}:
+Some interesting age-evolution trends have been predicted, and can
+probably be teased from our data.  Kelvin-Helmholtz cooling for
+mini-Neptunes should scale $R_{\rm p}$ as $t^{-0.1}$
+\citep{Gupta_2019}.  The photoevaporation desert has been predicted to
+be ``carved'' over the timescale for atmospheric loss
+\citep{Owen_Lai_2018}.  Mini-Neptunes should be converted into
+super-Earths \citep{Rogers_2021}.  And the boundary of the super-Earth
+population may correspondingly move up in size over time
+\citep{David_2021}.  We're going to split the two papers.
 
 {\it Field star ages from other surveys}:
-In this study we focused on deriving rotation and lithium-based ages for the Kepler
-planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
-Rotation-based age dating of their discoveries could 
-potentially clarify their youth, and thereby help clarify the broader
-issue of their evolution.
+In this study we focused on deriving rotation and lithium-based ages
+for the Kepler planets.  However surveys such as K2, TESS, HAT, and
+WASP have had very different selection functions.  Rotation-based age
+dating of their discoveries could potentially clarify their youth, and
+thereby help clarify the broader issue of their evolution.
 
@@
 Some of the youngest systems in Table~\ref{tab:planets} have been
-connected to their birth clusters.
-Others have not.
-A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
-not been performed.
-The ideal approach for such a search would be to combine spatial and kinematic information
-from Gaia with rotation measurements from TESS.
-Given that the decoherence time of young clusters is $\sim$100\,Myr, one would expect that
-the youngest ``field stars'' may have young neighbors that are relatively easy to
-identify.
+connected to their birth clusters.  Others have not.  A systematic
+search for the birth clusters of the youngest ``field stars'' has, to
+our knowledge, not been performed.  The ideal approach for such a
+search would be to combine spatial and kinematic information from Gaia
+with rotation measurements from TESS.  Given that the decoherence time
+of young clusters is $\sim$100\,Myr, one would expect that the
+youngest ``field stars'' may have young neighbors that are relatively
+easy to identify.
 
@@
 
-We began this article with two questions: how wrong is the
-assumption of a uniform age distribution for stars in the galactic
-thin disk?  And why are only $\approx$50 sub-gigayear transiting
-planets known, rather than the $\approx$500 that would be expected
-under the assumption of a uniform star formation history for the
+We began this article with two questions: how wrong is the assumption
+of a uniform age distribution for stars in the galactic thin disk?
+And why are only $\approx$50 sub-gigayear transiting planets known,
+rather than the $\approx$500 that would be expected under the
+assumption of a uniform star formation history for the
 $\approx$5{,}000 known planets?
@@
 stellar rotation periods, lithium equivalent widths, and temperatures
-using archival and new data.  We derived new ages
-using empirical methods that have been
-calibrated against an extensive set of open cluster data.
+using archival and new data.  We derived new ages using empirical
+methods that have been calibrated against an extensive set of open
+cluster data.
 
@@
 most of the sample, the independent lithium and rotation ages agreed
-for $\gtrsim$90\% of cases.  
-In particular, we reported {\bf XX} confirmed
-$<$500\,Myr planets with strong age constraints from both rotation and
-lithium, including {\bf YY} mini-Neptunes, {\bf ZZ} super-Earths, {\bf
-AA} Earths and sub-Earths, and a Neptune-sized planet.
+for $\gtrsim$90\% of cases.  In particular, we reported {\bf XX}
+confirmed $<$500\,Myr planets with strong age constraints from both
+rotation and lithium, including {\bf YY} mini-Neptunes, {\bf ZZ}
+super-Earths, {\bf AA} Earths and sub-Earths, and a Neptune-sized
+planet.
 
@@
 
-In this section we discuss fifteen systems with confirmed planets that have nominally
-discrepant rotation and lithium-based ages.
-We find that only one of them, Kepler-786, is interesting.
-
- \paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O}
-has a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot Jupiter,
-and has been studied in detail by multiple investigators
+In this section we discuss the systems with confirmed planets that
+have nominally discrepant rotation and lithium-based ages.  We find
+that only one of them, Kepler-786, is interesting.
+
+\paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} has
+a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot
+Jupiter, and has been studied in detail by multiple investigators
 \citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
-Our lithium-based age of this system
-(\trestwotli\,Myr), derive from EW$_{\rm Li}$=81$\pm$3\,m\AA,
-is qualitatively in line with the previously noted lithium abundance
-\citep{2007ApJ...664.1190S}. 
-However, we detect no photometric rotation signal, and the spectroscopic $v\sin i$ is low.
-The lithium age is strongly asymmetric because
-there is significant amounts of scatter in EW$_{\rm Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G dwarfs.
-We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of 2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.
-The non-detection of rotation, while mildly surprising, is not shocking if the system's true
-age is in this long tail.
-
-\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm Li}$=100$\pm$5\,m\AA,
-nominally implying $t_{\rm Li}$$\approx$200-800\,Myr, but no rotation detection.
-Our automated flags noted that this star has a low surface gravity, high luminosity, and is far from the main sequence.
-\citet{2014A&A...572A...2B} concur: this star is a sub-giant slightly more massive than the Sun;
-no rotation detection is expected; the lithium likely simply survived over the star's
-main sequence lifetime.
-
-\paragraph{Kepler-108}---This system of two mutually
-inclined giant planets \citep{2017AJ....153...45M} has a $T_{\rm eff}$$\approx$5600\,K
-host star with a strong lithium detection, and no rotation detection.
-The host star is entering the red giant branch; this has been previously noted
-through asteroseismology \citep{2013ApJ...767..127H}, and our analysis
-of the Keck/HIRES spectrum with \texttt{SpecMatch-Synth} \citep{2017AJ....154..107P} concurs, yielding
-$T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
-[Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.
-Given the star's high mass, the high lithium content and lack of rotation detection
-is not surprising.
+Our lithium-based age of this system (\trestwotli\,Myr), derive from
+EW$_{\rm Li}$=81$\pm$3\,m\AA, is qualitatively in line with the
+previously noted lithium abundance \citep{2007ApJ...664.1190S}.
+However, we detect no photometric rotation signal, and the
+spectroscopic $v\sin i$ is low.  The lithium age is strongly
+asymmetric because there is significant amounts of scatter in EW$_{\rm
+Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G
+dwarfs.  We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of
+2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.  The non-detection
+of rotation, while mildly surprising, is not shocking if the system's
+true age is in this long tail.
+
+\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm
+Li}$=100$\pm$5\,m\AA, nominally implying $t_{\rm
+Li}$$\approx$200-800\,Myr, but no rotation detection.  Our automated
+flags noted that this star has a low surface gravity, high luminosity,
+and is far from the main sequence.  \citet{2014A&A...572A...2B}
+concur: this star is a sub-giant slightly more massive than the Sun;
+no rotation detection is expected; the lithium likely simply survived
+over the star's main sequence lifetime.
+
+\paragraph{Kepler-108}---This system of two mutually inclined giant
+planets \citep{2017AJ....153...45M} has a $T_{\rm
+eff}$$\approx$5600\,K host star with a strong lithium detection, and
+no rotation detection.  The host star is entering the red giant
+branch; this has been previously noted through asteroseismology
+\citep{2013ApJ...767..127H}, and our analysis of the Keck/HIRES
+spectrum with \texttt{SpecMatch-Synth} \citep{2017AJ....154..107P}
+concurs, yielding $T_{\rm eff}$=5668$\pm$100\,K, $\log g$=3.8$\pm$0.1,
+[Fe/H]=0.38$\pm$0.06, and $v\sin i$=3.2$\pm$1.0\kms.  Given the star's
+high mass, the high lithium content and lack of rotation detection is
+not surprising.
 
@@
 
-%
-%Our quality flags noted that this star has
-%a low surface gravity, a crowded field, high RUWE, and an ambiguous Kepler-to-Gaia crossmatch.
-
-\paragraph{Kepler-786}---This K3 V star has a surprisingly 
-high lithium content.
-The spectrum is single-lined, with
- \texttt{SpecMatch-Synth} derived parameters of
- $T_{\rm eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1,
- [Fe/H]=0.11$\pm$0.06, and $v\sin i$=2.4$\pm$0.6\kms.
-Yet with 
-EW$_{\rm Li}$=83$\pm$6\,m\AA,
-the lithium age of \kepseveneightsix\,Myr would predict an obvious rotation signal.
-None is present in the Kepler light curve, consistent with the low
-$v\sin i$.
-Out of all ``apparently discrepant'' lithium and rotation measurements
-discussed in this appendix, this is the only one that seems to remain discrepant after scrutiny.
-The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm HK}=-4.7\pm0.5$, which
-suggests an age at least as old as the Hyades
-\citep{Mamajek_2008}.
-The Balmer lines are in absorption, and display no obvious signatures of youth.
-
-\paragraph{Kepler-1644}---The rotation-based age of this system (\kepsixteenfourfour\,Myr) is nominally
-much lower than the lithium limit ($>$767\,Myr).
-The Kepler light curve shows a $\approx$1\% amplitude 1.4\,day rotation signal with many flares.
-However the automated quality flags note that the star has low (photometric) surface gravity, a high RUWE (${\rm RUWE}_{\rm DR3}$=7.0),
-and that it is far from the main sequence.
-The Keck/HIRES spectrum also shows visually narrow lines, with a
- \texttt{SpecMatch-Synth} $v \sin i \leq 2$\kms. 
-However, we performed a cross-correlation between the HIRES spectrum and 
-the nearest matches in the Keck/HIRES template library \citep{2015AJ....149...18K}, and found that on 31 July 2022 (UT)
-the system displayed
-a broad CCF with a blended second component at $\approx$$+26$\,\kms relative to
-the primary, with a best-fit flux-ratio of $\approx$3.4\%, and a preferred $T_{\rm eff,B}$$\approx$4400\,K.
-The spectrum and astrometric excess noise therefore point to this system being an unresolved
-binary, which calls the
-reliability of the rotation-based age into question.
-The non-detection of the companion
-from Robo-AO imaging at Palomar \citep{2017AJ....153...66Z} suggests that the companion(s)
-are likely within $\rho$$\lesssim$0.3$''$.
-
-\paragraph{Kepler-1699}---This system is in a similar qualitative regime as Kepler-1644, with an
-apparently young $t_{\rm gyro}$=\kepsixteenninenine\,Myr derived from a $\approx$2\% amplitude
-4.2\,day rotation signal, and no evidence for lithium.
-The system has ${\rm RUWE}_{\rm DR3}$=19.1.
-From the same style of CCF analysis from the Keck/HIRES spectrum acquired on
-31 Aug 2022 (UT), we also find a double-peaked CCF, in this case with a secondary
-component at $\approx$$-16$\,\kms relative to
-the primary with $T_{\rm eff,B}$$\approx$4900\,K.
-This putative companion is similarly not detected in high-resolution imaging
- \citep{2017AJ....153...66Z}.
- The rotation-based age is questionable, given that we do not know which source the
- rotation signal is from, or whether the stars have interacted.
-
-\paragraph{Kepler-1943}---This system nominally has a \kepnineteenfourthree\,Myr
-lithium age, and no reported rotation detection.
-However, the star is flagged as being over-luminous, low-surface gravity,
-and far from the main sequence.
-In other words, it is a subgiant.
-
-\paragraph{Kepler-639, Kepler-320, Kepler-1719, Kepler-1876, Kepler-1072, Kepler-1743, Kepler-1929, Kepler-1488}---These eight systems all have nominally two-sided lithium age posteriors between 1 and 3\,Gyr, and yet lack rotation period detections.
-All are subgiants, flagged with $Q_{\rm star}$ under various combinations of bits 1, 2, and 9.
+\paragraph{Kepler-786}---This K3 V star has a surprisingly high
+lithium content.  The spectrum is single-lined, with
+\texttt{SpecMatch-Synth} derived parameters of $T_{\rm
+eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
+$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm Li}$=83$\pm$6\,m\AA, the
+lithium age of \kepseveneightsix\,Myr would predict an obvious
+rotation signal.  None is present in the Kepler light curve,
+consistent with the low $v\sin i$.  Out of all ``apparently
+discrepant'' lithium and rotation measurements discussed in this
+appendix, this is the only one that seems to remain discrepant after
+scrutiny.  The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm
+HK}=-4.7\pm0.5$, which suggests an age at least as old as the Hyades
+\citep{Mamajek_2008}.  The Balmer lines are in absorption, and display
+no obvious signatures of youth.
+
+\paragraph{Kepler-1644}---The rotation-based age of this system
+(\kepsixteenfourfour\,Myr) is nominally much lower than the lithium
+limit ($>$767\,Myr).  The Kepler light curve shows a $\approx$1\%
+amplitude 1.4\,day rotation signal with many flares.  However the
+automated quality flags note that the star has low (photometric)
+surface gravity, a high RUWE (${\rm RUWE}_{\rm DR3}$=7.0), and that it
+is far from the main sequence.  The Keck/HIRES spectrum also shows
+visually narrow lines, with a \texttt{SpecMatch-Synth} $v \sin i \leq
+2$\kms.  However, we performed a cross-correlation between the HIRES
+spectrum and the nearest matches in the Keck/HIRES template library
+\citep{2015AJ....149...18K}, and found that on 31 July 2022 (UT) the
+system displayed a broad CCF with a blended second component at
+$\approx$$+26$\,\kms relative to the primary, with a best-fit
+flux-ratio of $\approx$3.4\%, and a preferred $T_{\rm
+eff,B}$$\approx$4400\,K.  The spectrum and astrometric excess noise
+therefore point to this system being an unresolved binary, which calls
+the reliability of the rotation-based age into question.  The
+non-detection of the companion from Robo-AO imaging at Palomar
+\citep{2017AJ....153...66Z} suggests that the companion(s) are likely
+within $\rho$$\lesssim$0.3$''$.
+
+\paragraph{Kepler-1699}---This system is in a similar qualitative
+regime as Kepler-1644, with an apparently young $t_{\rm
+gyro}$=\kepsixteenninenine\,Myr derived from a $\approx$2\% amplitude
+4.2\,day rotation signal, and no evidence for lithium.  The system has
+${\rm RUWE}_{\rm DR3}$=19.1.  From the same style of CCF analysis from
+the Keck/HIRES spectrum acquired on 31 Aug 2022 (UT), we also find a
+double-peaked CCF, in this case with a secondary component at
+$\approx$$-16$\,\kms relative to the primary with $T_{\rm
+eff,B}$$\approx$4900\,K.  This putative companion is similarly not
+detected in high-resolution imaging \citep{2017AJ....153...66Z}.  The
+rotation-based age is questionable, given that we do not know which
+source the rotation signal is from, or whether the stars have
+interacted.
+
+\paragraph{Kepler-1943}---This system nominally has a
+\kepnineteenfourthree\,Myr lithium age, and no reported rotation
+detection.  However, the star is flagged as being over-luminous,
+low-surface gravity, and far from the main sequence.  In other words,
+it is a subgiant.
+
+\paragraph{Kepler-639, Kepler-320, Kepler-1719, Kepler-1876,
+Kepler-1072, Kepler-1743, Kepler-1929, Kepler-1488}---These eight
+systems all have nominally two-sided lithium age posteriors between 1
+and 3\,Gyr, and yet lack rotation period detections.  All are
+subgiants, flagged with $Q_{\rm star}$ under various combinations of
+bits 1, 2, and 9.
 
@@
 
-%  Kepler-63b is interesting in that it is {\it missing} from our
-%  catalog.  This 6.1\,$R_\oplus$ ``sub-Saturn'' has been shown to be in
-%  a polar orbit based on both the Rossiter-McLaughlin effect and a
-%  starspot-crossing analysis \citep{2013ApJ...775...54S}.  With $T_{\rm
-%  eff}=5576\pm50$\,K, and a rotation period of 5.40\,days, we calculate
-%  a rotation-based age of $t_{\rm gyro}$=$251^{+80}_{-90}$\,Myr,
-%  qualitatively consistent with the estimates from
-%  \citet{2013ApJ...775...54S}, though with larger uncertainties because
-%  our age accounts for the intrinsic scatter in the open cluster
-%  calibration sequences \citep{Bouma_2023}.  
-%  While the usual catalogs (\citetalias{Santos_2019},
-%  \citetalias{Santos_2021}, \citetalias{McQuillan_2014}, and
-%  \citetalias{Mazeh_2015}) do not report this star's rotation period.
-%  \citetalias{Santos_2021} flagged the system as ``potentially'' having
-%  rotation-based modulation.
-
 %\section{Comparison Against Previous Literature Age Scales}

### 2024-05-11 paper/ms.tex
update ms

@@
   We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
-  and \nplwgyroagenograzing\ planets for which we believe
-  gyrochronology is valid; these rotation-based ages are roughly an
-  order of magnitude more precise than isochrone ages in this sample
-  of stars.
-  %
-  The independent lithium ages agree with the rotation-based ages for
+  and \nplwgyroagenograzing\ planets for which our gyrochronology
+  model is applicable; these rotation-based ages are over an order of
+  magnitude more precise than isochrone ages in this sample of stars.
+  %
+  The independent lithium ages agree with the rotation-based ages in
   $\approx$90\% of cases, suggesting that most of the rotation-based
@@
   %
-  More broadly, we find that the star formation rate in the Kepler
-  field has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over
-  the past 3\,Gyr, equally for both planet hosts and for the overall
+  More broadly, we find that the star formation rate in the Kepler field
+  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
+  3,Gyr. This trend holds true for both planet hosts and the overall
   stellar sample.
@@
 \citep{2017MNRAS.470.1750I} are expected to be most efficient over
-$\approx$10\,Myr to $\approx$1\,Gyr timescales; for most known
-exoplanets, most of the action has already happened.
-
-Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
-help build the timeline for exoplanet evolution.  Informative
-individual young systems include V1298~Tau, an adolescent resonant
-chain of four transiting puffy planets that is a likely precusor to
-the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
-Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
-Thao et al.~submitted).  Population-level analyses have similarly
-suggested differences in the size distribution of young exoplanets
-relative to their older counterparts
+$\approx$10,Myr to $\approx$1,Gyr timescales. For most known
+exoplanets, these processes have already run their course.
+
+Young ($<$1\,Gyr) exoplanets are key to building the timeline for
+exoplanet evolution.  Informative individual young systems include
+V1298~Tau, an adolescent resonant chain of four transiting puffy
+planets that is a likely precusor to the compact multiplanet systems
+\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
+Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
+Population-level analyses have similarly suggested differences in the
+size distribution of young exoplanets relative to their older
+counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 K2, TESS, and Gaia
-\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
+\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
 The strategy pursued by most groups during the 2010s was to focus on
@@
 ages.
+
+%FIXME these paragraphs need improvements in how they incorporate
+%Lu+21 (meh), but certainly Mathur+2023.
 
@@
 \subsection{Future Directions}
+
 {\it Occurrence Rates}:
-Some interesting age-evolution trends have been predicted, and can
-probably be teased from our data.  Kelvin-Helmholtz cooling for
-mini-Neptunes should scale $R_{\rm p}$ as $t^{-0.1}$
-\citep{Gupta_2019}.  The photoevaporation desert has been predicted to
-be ``carved'' over the timescale for atmospheric loss
-\citep{Owen_Lai_2018}.  Mini-Neptunes should be converted into
-super-Earths \citep{Rogers_2021}.  And the boundary of the super-Earth
-population may correspondingly move up in size over time
-\citep{David_2021}.  We're going to split the two papers.
+The age-dependent trends predicted for exoplanet populations, such as
+the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
+time-dependent carving of the photoevaporation desert
+\citep{Owen_Lai_2018}, and the time evolution of the radius valley
+\citep{Rogers_2021} can be explored using our data.  We defer this
+analysis to a separate publication.
 
 {\it Field star ages from other surveys}:
-In this study we focused on deriving rotation and lithium-based ages
-for the Kepler planets.  However surveys such as K2, TESS, HAT, and
-WASP have had very different selection functions.  Rotation-based age
-dating of their discoveries could potentially clarify their youth, and
-thereby help clarify the broader issue of their evolution.
+While this study focused on deriving rotation and lithium-based ages
+for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF, and
+ATLAS have different selection functions.
+Applying rotation-based age dating to their discoveries could clarify
+the youth and evolutionary trends of these populations.  Future
+missions like PLATO \citep{Rauer14} and Roman Galactic Plane Time
+Domain Survey  will further expand the sample of stars with precisely
+determined ages, including those known to host planets.
 
@@
 Some of the youngest systems in Table~\ref{tab:planets} have been
-connected to their birth clusters.  Others have not.  A systematic
-search for the birth clusters of the youngest ``field stars'' has, to
-our knowledge, not been performed.  The ideal approach for such a
-search would be to combine spatial and kinematic information from Gaia
-with rotation measurements from TESS.  Given that the decoherence time
-of young clusters is $\sim$100\,Myr, one would expect that the
-youngest ``field stars'' may have young neighbors that are relatively
-easy to identify.
+linked to their birth clusters, while others have not. A systematic
+search for the birth clusters of the youngest ``field stars'' could
+combine spatial and kinematic information from Gaia with rotation
+measurements from TESS. Given the $\sim$100\,Myr decoherence time of
+young clusters, the youngest "field stars" may have easily
+identifiable young neighbors.
 

### 2024-05-13 paper/ms.tex
update ms

@@
 \citep{2017MNRAS.470.1750I} are expected to be most efficient over
-$\approx$10,Myr to $\approx$1,Gyr timescales. For most known
+$\approx$10\,Myr to $\approx$1\,Gyr timescales. For most known
 exoplanets, these processes have already run their course.
 
-Young ($<$1\,Gyr) exoplanets are key to building the timeline for
+Young ($<$1\,Gyr) exoplanets represent one means of building the timeline for
 exoplanet evolution.  Informative individual young systems include
@@
 planet, and measure the star's age.  Each problem admits a range of
-solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q}.  In
+solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.  In
 this article we will consider planets previously known from transits,
@@
 To begin, imagine that the ages of nearby stars in the Galaxy are
-uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
+uniformly distributed from 0--10\,Gyr \citep[e.g.~Fig~17
 of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
@@
 $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
-currently forming protoplanets \citep{2018A&A...617A..44K}, or of
+currently forming protoplanets \citep{2018A&A...617A..44K}, and of
 exoplanets evolving just after disk dispersal
@@
 Different sub-groups of the complex may vary in age by $\approx$50\%
-(\citealt{Bouma_2022b}; Kerr~et.al.~in preparation).  The planets
+(\citealt{Bouma_2022b}; Kerr~et al.,~in preparation).  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
@@
 rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
-Metallicity may also be a second-order effect
+Metallicity may also be relevant
 \citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
@@
 provides one path toward this end.  The other method is to consider
-the region in which we do have overlapping sensitivity between lithium
-and rotation -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
+the cases for which lithium
+and rotation overlap in sensitivity -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
 for such stars include \ltonegyrhighqconfirmedtwosided\ planets
@@
   positive KOIs with MES$>$10.  
-  A shell script to decode the $Q_{\rm star}$ quality flag is {\bf \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available online}}.
+  A bash script to decode any given $Q_{\rm star}$ quality flag is
+  {\bf
+  \href{https://gist.github.com/lgbouma/20368253f1a98da1b39cf32fdda0be13}{available
+  online}}.  A python script to select stars with specific bit flags
+  is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
   \label{tab:planets}}
@@
   \enddata
-  \tablecomments{``Confirmed'' planets appear in the machine-readable version before``candidate'' planets.
-    The bit quality flags for the rotation-based ages,
-    $Q_{\rm star}$, are described in Section~\ref{subsec:flags}.
-    Concisely summarized, they are:
+  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
+  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
+  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
+  ``Confirmed'' planets appear in the machine-readable
+  version before``candidate'' planets.  The bit quality flags for the
+  rotation-based ages, $Q_{\rm star}$, are described in
+  Section~\ref{subsec:flags}.  Concisely summarized, they are:
   Bit 0: $T_{\rm eff}/{\rm K} \in [ 3800-6200]$ ?

### 2024-05-14 paper/ms.tex
update ms

@@
 
-Exoplanet science is in an age of abundance, fueled by the discovery
+Exoplanet science is enjoying an age of abundance, fueled by the discovery
 of thousands of worlds orbiting close to their host stars
@@
 The main problem with the ``cluster-first'' approach is that most
-nearby young stars are in the field.  Only $\approx$1\% of stars
-within the nearest $\approx$500\,pc are associated with their birth
-cluster
+nearby sub-gigayear stars are in the field.  In other words,
+$\lesssim$1\% of stars within the nearest $\approx$500\,pc have been
+associated with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
@@
 or TESS.  However, combined, those surveys have discovered
-$\approx$5{,}000 planets.  Unless there is a major bias against the
-discovery of sub-gigayear old planets with precisely constrained ages,
-a constant star formation history implies that an order of magnitude
+$\approx$5{,}000 planets.  
+Barring a major discovery bias against sub-gigayear planets,
+a constant star formation history would imply that an order of magnitude
 more young transiting planets should be known.
@@
 Second, where are the missing young planets?  Briefly, we will find
-that uniform is wrong by a factor of a few, and that at least one bias
-against the discovery of young planets is in the measurement of their
-ages.
+that uniform is wrong by a factor of a few, and that increased stellar
+activity for young stars may be a less important bias than the
+challenge of determining precise ages for field stars.
 
@@
 improvements in the lithium age scale.  Lithium ages include two
-distinct regimes: depletion boundary ages for M-dwarfs in star
+distinct regimes: depletion boundary ages for M dwarfs and brown dwarfs in star
 clusters, and the less precise decline-based ages for individual field
@@
 are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
-have been reported for $\approx$one in three Kepler targets
+have been reported for roughly one in three Kepler targets
 \citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
@@
 describe the set of stars that we adopt with measured rotation periods
-(Section~\ref{subsec:rotsel}), the set of objects we adopt as planets
-(Section~\ref{subsec:planetsel}), and the set of planets with
+(Section~\ref{subsec:rotsel}), the set of objects we adopt as hosting planets
+(Section~\ref{subsec:planetsel}), and the subset of these with
 high-resolution spectra suitable for lithium analysis
@@
 only homogeneous analysis of both Kepler targets and KOIs were the
-studies by \citet{Santos_2019,Santos_2021} (\citetalias{Santos_2019}
-and \citetalias{Santos_2021}).  \citetalias{Santos_2019} and
+studies by \citet{Santos_2019,Santos_2021}, hereafter \citetalias{Santos_2019}
+and \citetalias{Santos_2021}.  \citetalias{Santos_2019} and
 \citetalias{Santos_2021} combined a wavelet analysis and
@@
 
-We considered planets in the NEA cumulative KOI table, which included
-the best knowledge available on any given planet candidate as of 6
-June 2023, while also incorporating human-based vetting.  These
-planets represent a superset of those in the fully automated Q1-Q17
-DR25 KOI Table \citep{Thompson_2018}, which could be adopted in future
-work for planet occurrence rate calculations.  This version of the
-cumulative KOI table included \nkoisnofp\ objects that are either
-``confirmed'' or ``candidate'' planets, after excluding known false
-positives.
+We considered planets in the NEA cumulative KOI table as of 6 June
+2023, which included the best knowledge available on any given planet
+candidate while also incorporating human-based vetting.  These planets
+represent a superset of those in the fully automated Q1-Q17 DR25 KOI
+Table \citep{Thompson_2018}, which could be adopted in future work for
+planet occurrence rate calculations.  This version of the cumulative
+KOI table included \nkoisnofp\ objects that are either ``confirmed''
+or ``candidate'' planets, after excluding known false positives.
 
@@
 Interest were already analyzed by \citet{2018ApJ...855..115B} for
-$\approx$three quarters of the spectra in our sample.  However, new
+roughly three quarters of the spectra in our sample.  However, new
 spectra have since been acquired, and our approach and selection
-function are different.  We therefore perform our own processing of
+function are different.  We therefore perform our own line width
+measurements on
 the reduced HIRES spectra.
 
-We queried the CPS archive to collect all blaze-corrected HIRES
-spectra that were available for all non-false positive Kepler Objects
-of Interest with a MES$>$10.  This yielded at least one spectrum for
-\nlithiumstars\ stars and \nlithiumplanets\ planets.  For stars with
-multiple spectra available at varying ranges of signal-to-noise (S/N),
-we analyzed only the spectrum with the highest number of counts.  The
-resulting spectra were acquired between 6 September 2009 and 16 May
-2024, and cumulatively comprise \nhireshours\,hours of open-shutter
-time.
+We collected all blaze-corrected HIRES spectra from our group's
+observations of non-false positive Kepler Objects of Interest with a
+``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
+least 10.  This yielded at least one spectrum for \nstwithspec\
+stars hosting \nplwithspec\ planets.  About half of these stars have
+rotation periods (\nstwithspecandprot\ stars and
+\nplwithspecandprot, respectively).   For stars with multiple spectra
+available at varying ranges of signal-to-noise (S/N), we analyzed only
+the spectrum with the highest number of counts.  The resulting spectra
+were acquired between 6 September 2009 and 16 May 2024, and
+cumulatively comprise \nhireshours\,hours of open-shutter time.
 
@@
 differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
-required positive EWs, while we allow for negative ones.  At
+required positive EWs, while we allow for statistically negative ones.  At
 $>$30\,m\AA, there is however a small offset in our respective scales,
@@
 our non-treatment of the iron blend.  We therefore directly subtract
-the constant mean value noted above when calculating lithium-based
+this constant mean value (7.50\,m\AA) when calculating lithium-based
 ages.
@@
     \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
-    isochrones from the same models as in Figure~\ref{fig:models}; the
+    isochrones from the same models as in Figure~\ref{fig:models}.  The
     intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
@@
     \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
-    are the ``mean models'' at various ages; the intrinsic dispersion
+    are the ``mean models'' at various ages.  The intrinsic dispersion
     in rotation and lithium about these mean models, which is what the
@@
 the framework interpolates between reference open clusters using cubic
-hermite polynomials.  We calculate the probability of the
+Hermite polynomials.  We calculate the probability of the
 rotation-based age $t_{\rm gyro}$, given the observed rotation
@@
 \label{subsec:plflags}
-Some planets are more reliable than others.  We use the following
-additional quality indicators to assess the reliability and utility of
-a planet, and assemble them into a separate bit-flag, $Q_{\rm
-planet}$.
+Some planets are more reliably identified than others.  We use the
+following additional quality indicators to assess the reliability and
+utility of a planet, and assemble them into a separate bit-flag,
+$Q_{\rm planet}$.
 
@@
 \nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
-cuts, save for that on RUWE.  Then, require these confirmed and
+cuts, save for that on RUWE.  We can then require these confirmed and
 candidate planets to have MES$>$10.  This combination of selection
@@
 \texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
-a purely empirical interpolation approach.  \texttt{EAGLES} however
+a purely empirical interpolation approach.  \texttt{EAGLES} 
 was calibrated on lithium measurements of stars observed by the
@@
 cluster age out to which our interpolation-based method is calibrated.
+
 For Sun-like stars, the fraction of rotation signals detectable by
-Kepler has been found to be near-unity up to $\approx$3\,Gyr, and it
+Kepler is near unity up to $\approx$3\,Gyr, and it
 drops to almost zero by $\approx$5\,Gyr \citep{2022ApJ...937...94M}.
@@
 Kepler sample yield \{\nplyounggyro, \nplmidgyro, \nploldgyro\}
-detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.  For
-the sub-gigayear bin in particular, which holds extra weight in
-studies of planet evolution, while there are \nplyounggyro\ planets
+detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.  
+The stellar ages and planetary
+properties of all of these systems are given in
+Table~\ref{tab:planets}.
+
+\subsection{Planets Younger Than One Billion Years}
+
+Sub-gigayear planets hold particular weight in studies of planet
+evolution.
+While our catalog has \nplyounggyro\ confirmed planets
 with median ages below 1\,Gyr, requiring $t_{\rm gyro}$$<$1\,Gyr at
 2$\sigma$ confidence yields \nplyounggyrotwosigma\ planets orbiting
-\nplhostsyounggyrotwosigma\ stars.  The stellar and planetary
-properties of this latter sample of young planets are given in
-Table~\ref{tab:planets}.
-
-\subsection{Planets Younger Than One Billion Years}
-
-What are the most notable confirmed planets with well-measured ages?
-These planets can be split into two classes: those in clusters, and
-those with independent and consistent rotation and lithium-based ages.
+\nplhostsyounggyrotwosigma\ stars.  
+The most notable of these planets can be split into two classes: those
+in clusters, and those with independent and consistent rotation and
+lithium-based ages.
 
@@
 
-There are currently four known stellar ensembles in the Kepler field
-that have yielded a total of fourteen transiting planets.  Our
-analysis blindly recovers each of these planets, as well as their
-ages.
+Four stellar ensembles in the Kepler field have to date yielded a
+total of fourteen transiting planets.  Our analysis blindly recovers
+the youth of these planets.
 
@@
 
-We compared the independent rotation and lithium-based ages using the
-point-estimates of the posteriors, when available.  There are three
-cases of interest:
+We compared the independent rotation and lithium-based ages of the
+known planet hosts using the point-estimates of the available
+posteriors.  There are three cases of interest:
 
@@
 
-In the entire sample (i.e.~without imposing any quality cuts) this
+In the entire planet sample (i.e.~without imposing any quality cuts) this
 yielded \allagesyesconsistent\ consistent cases,

### 2024-05-15 paper/ms.tex
update ms

@@
 
-However, beyond rotation-based ages, recent years have also yielded
+Beyond rotation-based ages, recent years have also yielded
 improvements in the lithium age scale.  Lithium ages include two
@@
 Empirical understanding has however improved due to the work by
-\citet{Jeffries_2023}, who built a model for how the equivalent width
-(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
-stellar effective temperature and age for a set of 6{,}200 stars in 52
+\citet{Jeffries_2023}, who modeled the evolution of 
+the \ion{Li}{1} 6708\,\AA\ doublet equivalent width (EW) through
+time using a set of 6{,}200 stars in 52
 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
@@
 for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
-DR2,  and $g$-band photometry from either the KIC or the Kepler-INT
-survey.  The parameters themselves were derived using
+DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
+\citep{2012AJ....144...24G}.  The parameters themselves were derived using
 \texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
@@
   Kepler \citep{Borucki10},
-  Keck:I (HIRES).
+  Keck:I (HIRES),
+	2MASS \citep{Skrutskie06},
+	SDSS \citep{2000AJ....120.1579Y}.
 }
@@
 
+\section{TODO's}
+
+\begin{enumerate}
+  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
+    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
+    Could compare against asteroseis too.
+    %
+  \item Intro needs better discussion of Mathur2023 and Lu2021
+    %
+  \item Consider moving the bit list description to its own appendix.
+    %
+  \item Rerun gyro calculation with denser grid, fixing quantization
+    and extending upper limit to 5gyr.  (Will use updated Teff
+    uncertainties)
+\end{enumerate}
+
 %\section{Comparison Against Previous Literature Age Scales}

### 2024-05-15 paper/ms.tex
update ms

@@
 
-{\it How do we use the bit-flags?} By default, we assume that a star
+{\it How do we use the bitmasks?} By default, we assume that a star
 is suitable for gyrochronology if none of bits zero through nine
@@
 following additional quality indicators to assess the reliability and
-utility of a planet, and assemble them into a separate bit-flag,
+utility of a planet, and assemble them into a separate bitmask,
 $Q_{\rm planet}$.
@@
 
-Figure~\ref{fig:hist_tgyro} shows the gyrochone age distributions of
-stars and known planet-hosts in the Kepler field.  In detail, these
+The stellar ages and planetary properties of the systems discussed in
+the previous sections are given in Tables~\ref{tab:planets}
+and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
+age distributions of the stars and known planet-hosts.  These
 ``histograms'' are sampled from the posterior probability
 distributions by drawing ten random samples from each posterior
-distribution, and then computing the histogram of the resulting
-samples.  We truncated the plots near $4$\,Gyr since this is the
-cluster age out to which our interpolation-based method is calibrated.
-
-For Sun-like stars, the fraction of rotation signals detectable by
-Kepler is near unity up to $\approx$3\,Gyr, and it
-drops to almost zero by $\approx$5\,Gyr \citep{2022ApJ...937...94M}.
-This trend has the opposite functional form from our derived age
-distributions.
-
-To quantify the amount by which young stars are more rare than old, we
-can group stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
+distribution, and then computing the re-weighted histogram of the
+resulting samples.  The uncertainties assume Poisson statistics.  We
+truncated the plots near $4$\,Gyr since this is the cluster age out to
+which our interpolation-based method is calibrated.
+We refer to this as a ``demographic cliff'' in reference to an
+impending crisis expected to befall higher education.
+
+The age distributions of Kepler stars and planet hosts are
+statistically quite similar.  A simple diagnostic of whether the two
+distributions are drawn from the same underlying distribution is the
+Kolmogorov-Smirnov test;  we calculated this statistic in a manner
+that accounts for the Poisson uncertainties by performing 100 random
+draws of 70\% of the stars from each sample.  The resulting median
+$p$-values were $\approx$0.01 for both rows of
+Figure~\ref{fig:hist_tgyro}.  This degree of similarity suggests that
+the ``common wisdom'' that young transiting planets are hard to detect
+because their host stars are photometrically variable requires some
+nuance.  If true, this statement only holds for the very youngest
+stars, since at the age resolution of this study, there appears to be
+little to no selection bias against sub-gigayear planets.
+
+An important separate selection effect concerns Kepler's ability to
+detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
+issue, and found for Sun-like stars that the fraction of rotation
+signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
+that it drops to almost zero by $\approx$5\,Gyr .  This trend is
+opposite in functional form to our derived age distributions.
+
+We can quantify the relative counts of stars as a function of age by
+labelling stars between 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr as ``young'',
 ``intermediate-age'', and ``old''.  A simple counting exercise from
@@
 detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
-The stellar ages and planetary
-properties of all of these systems are given in
-Table~\ref{tab:planets}.
 
@@
 Different sub-groups of the complex may vary in age by $\approx$50\%
-(\citealt{Bouma_2022b}; Kerr~et al.,~in preparation).  The planets
+(\citealt{Bouma_2022b}; Kerr~et al.,~submitted).  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
@@
 
-%Restricting our attention to the
-%$\lesssim$1\,Gyr stars, for which lithium ages are more constraining,
-%the respective counts are \minageltonegyryesconsistent,
-%\minageltonegyrmaybeconsistent, and \minageltonegyrnoconsistent.
-%Rephrased, in the overall sample, \fracconsistentallages\% of stars
-%have consistent rotation and lithium-based ages
-%(\fracpotentiallyconsistentallages\% potentially consistent).  In the
-%sample of sub-gigayear stars, \fracconsistentminageltonegyr\% of the
-%ages are consistent (\fracpotentiallyconsistentminageltonegyr\%
-%potentially consistent).  We feel it necessary to distinguish
-%``maybe'' consistent cases because of lingering uncertainties in how
-%the dispersion of the lithium data evolve over the first gigayear.
-
 \subsubsection{New Young Planets}
-%
-%The astrophysical processes behind rotation and lithium evolution are
-%independent.
-%Therefore 
-%whenever the two age indicators agree within a
-%narrowly-constrained range, confidence in the true age
-%increases.
-%The region of greatest overlap between rotation and lithium ages is
-%dictated by the timescale for lithium depletion: for stars with
-%temperatures between 4500-5500\,K with ages $\lesssim$1\,Gyr, we can
-%hope to obtain independent two-sided constraints
-%\citep{Bouma_2023,Jeffries_2023}.
 
@@
 provides an age constraint while $t_{\rm Li}$ is one-sided.
-%Two of the former systems are in clusters (Kepler-1643 and Kepler-1928);
-%five of the latter systems are in Theia~520.
 
@@
 
-\subsection{Caveats, Limitations}
-
-Most of our ages come from one source of information: rotation.
-Factors other than stellar age and mass can influence rotation rates
-of main sequence stars.  Binarity is one example: even
+\subsection{Caveats \& Limitations}
+
+Our most constraining ages tend to come from one source of
+information: rotation.  Factors other than stellar age and mass can
+influence stellar rotation rates.  Binarity is one example: even
 intermediate-separation ($\sim$10\,AU) binaries are biased toward
@@
 Metallicity may also be relevant
-\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  In this work, we
-were interested in not only the overall age distribution of the Kepler
-field, but also in the reliability of individual ages for young planet
-hosting stars.  The quality flagging scheme in Table~\ref{tab:planets}
-provides one path toward this end.  The other method is to consider
-the cases for which lithium
-and rotation overlap in sensitivity -- G and K dwarfs with $t$$\lesssim$1\,Gyr.  Our results
-for such stars include \ltonegyrhighqconfirmedtwosided\ planets
-orbiting ``high-quality'' stars for which the two age indicators are
-both constraining and in agreement (with $Q_{\rm star}$ bits zero
-through nine not raised, and no planet quality flags raised).
-Conversely, we found only one system, Kepler-786, with a precise
-lithium age (\kepseveneightsix\,Myr), but no rotation signal.
+\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
+is that our rotation-based ages used photometric effective
+temperatures (Section~\ref{sec:stellarprops}), even though
+spectroscopic temperatures are available for the planet-hosting
+subset.  This decision was driven by a desire for homogeneity 
+irrespective of planet-hosting status.  It may however have
+incurred small systematic errors in the implied ages.
+
+In this work, we were interested in not only the overall age
+distribution of the Kepler field, but also in the reliability of
+individual ages for young stars known to host planets.  The quality
+bitmask scheme in Table~\ref{tab:planets} provided one path toward
+this end.  Our other method was to compare the cases for which lithium
+and rotation overlap in sensitivity -- G and K dwarfs with
+$t$$\lesssim$1\,Gyr.  Our results for such stars include
+\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
+stars for which the two age indicators are both constraining and in
+agreement (stars with $Q_{\rm star}$ bits zero through nine not
+raised, and no planet quality flags raised).  Conversely, we found
+only one system, Kepler-786, with a precise lithium age
+(\kepseveneightsix\,Myr), but no rotation signal.
 
@@
 While this study focused on deriving rotation and lithium-based ages
-for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF, and
-ATLAS have different selection functions.
-Applying rotation-based age dating to their discoveries could clarify
-the youth and evolutionary trends of these populations.  Future
-missions like PLATO \citep{Rauer14} and Roman Galactic Plane Time
-Domain Survey  will further expand the sample of stars with precisely
-determined ages, including those known to host planets.
+for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF,
+and ATLAS have different selection functions.  Applying rotation-based
+age dating to their discoveries could clarify the youth and
+evolutionary trends of these populations.  Future prospects like PLATO
+\citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
+Galactic Plane Time Domain Survey seem poisted to further expand the
+sample of stars with precisely determined ages, including those known
+to host planets.
 
@@
   As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
-  The analogous planet quality bit-flag, $Q_{\rm planet}$, has the
+  The analogous planet quality bitmask, $Q_{\rm planet}$, has the
   following meaning.
@@
   shown here to give an idea for form and content.  The quality
-  bit-flag is as in Table~\ref{tab:planets}; requiring bits zero to
+  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
   nine to be null yields \nuniqstarsantosrotgyroappl\ stars for

### 2024-05-15 paper/ms.tex
better words

@@
 and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
-age distributions of the stars and known planet-hosts.  These
-``histograms'' are sampled from the posterior probability
-distributions by drawing ten random samples from each posterior
-distribution, and then computing the re-weighted histogram of the
-resulting samples.  The uncertainties assume Poisson statistics.  We
-truncated the plots near $4$\,Gyr since this is the cluster age out to
-which our interpolation-based method is calibrated.
-We refer to this as a ``demographic cliff'' in reference to an
-impending crisis expected to befall higher education.
+age distributions of all the stars (left column) and the known
+planet-hosts (middle column).  These ``histograms'' are sampled from
+the posterior probability distributions by drawing ten random samples
+from each posterior distribution, and then computing the re-weighted
+histogram of the resulting samples.  The uncertainties assume Poisson
+statistics.  We truncated the plots near $4$\,Gyr since this is the
+cluster age out to which our interpolation-based method is calibrated.
+We refer to the overall slope as a ``demographic cliff'' in reference
+to an impending human phenomenon forecasted to happen in coming years.
 

### 2024-05-16 paper/ms.tex
update ms (wip)

@@
 \appendix
+\section{Literature Age Scale Comparison}
+\label{app:litagecomparison}
+
+Figure~\ref{fig:agescalecompone}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{NGC6811_lit_ages_vs_cluster_ages.pdf}
+		}
+
+		\vspace{-0.35cm}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{Cep-Her_lit_ages_vs_cluster_ages.pdf}
+		}
+
+		\vspace{-0.35cm}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{NGC6819_lit_ages_vs_cluster_ages.pdf}
+		}
+
+		\vspace{-0.35cm}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{Melange3_lit_ages_vs_cluster_ages.pdf}
+	 	}
+
+		\vspace{-0.35cm}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{Theia520_lit_ages_vs_cluster_ages.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Age scale comparison (part 1).}  
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+    lorem ipsum dolur. 
+	}
+	\label{fig:agescalecompone}
+\end{figure*}
+
 \section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}

### 2024-05-16 gyrojo/getters.py

@@ -915,6 +915,7 @@ def get_cleaned_gaiadr3_X_kepler_dataframe():
     positives, fail to yield Gaia DR3 crossmatches.
     """
 
+    # Bedell's list...
     kepler_dr3_path = join(LOCALDIR, "kepler_dr3_4arcsec.fits")
     hdul = fits.open(kepler_dr3_path)
     gk_df = Table(hdul[1].data).to_pandas()

### 2024-05-16 paper/collect.sh

@@ -36,3 +36,9 @@ cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pd
 # appendices (probably omit)
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
+
+cp $r/lit_ages_vs_cluster_ages/NGC6811_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/Cep-Her_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/NGC6819_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/Melange3_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/Theia520_lit_ages_vs_cluster_ages.pdf .

### 2024-05-16 drivers/compare_literature_ages_to_cluster_ages.py
compare lit: add merged

@@ -19,6 +19,7 @@ from gyrojo.getters import select_by_quality_bits
 from gyrointerp.helpers import prepend_colstr, left_merge
 from astroquery.vizier import Vizier
 from aesthetic.plot import set_style, savefig
+from collections import Counter
 
 Vizier.ROW_LIMIT = -1
 
@@ -50,6 +51,7 @@ def get_cluster_dfs():
     # Plus another 200 from Kounkel+2020.
     nsdf = sdf[sdf.cluster.str.contains("6811") &
                ~sdf.cluster.str.contains("kcs20group_6811")]
+    print(Counter(nsdf.reference_bibcode).most_common(n=10))
 
     # Now xmatch against the kic <-> dr2 xmatch...
     ngc6811_df = nsdf.merge(gk_df, how='inner', on='source_id')
@@ -59,6 +61,7 @@ def get_cluster_dfs():
     #
     nsdf = sdf[sdf.cluster.str.contains("6819") &
                ~sdf.cluster.str.contains("kcs20group_6819")]
+    print(Counter(nsdf.reference_bibcode).most_common(n=10))
     ngc6819_df = nsdf.merge(gk_df, how='inner', on='source_id')
 
     # 
@@ -66,6 +69,7 @@ def get_cluster_dfs():
     #
     t520_sdf = sdf[sdf.cluster.str.contains("kcs20group_520") |
                    sdf.cluster.str.contains("UBC_1")]
+    print(Counter(t520_sdf.reference_bibcode).most_common(n=10))
     theia520_df = t520_sdf.merge(gk_df, how='inner', on='source_id')
 
     #
@@ -90,12 +94,14 @@ def get_cluster_dfs():
 
     ##########################################
 
-    cluster_dfs = [ngc6811_df, theia520_df, melange3_df, cepher_df, ngc6819_df]
-    cluster_names = ['NGC6811 (1 Gyr)', 'Theia520 (300 Myr)',
-                     'Melange3 (150 Myr)', # Δv$_\mathrm{T}$<2km/s 
+    cluster_dfs = [ngc6811_df, cepher_df, ngc6819_df, melange3_df, theia520_df]
+    cluster_names = ['NGC6811 (1 Gyr)',
                      'Cep-Her (40 Myr)', #$w$>0.1 
-                     'NGC6819 (2.5 Gyr)']
-    true_ages = [1, 0.3, 0.15, 0.04, 2.5]
[…]

### 2024-05-16 paper/ms.tex
update ms

--- 2024-05-15 paper/ms.tex
@@
 \appendix
+\section{Literature Age Scale Comparison}
+\label{app:litagecomparison}
+
+Figure~\ref{fig:agescalecompone} shows the result of a literature comparison against
+ages reported by \citet{Berger_2020a_catalog}, \citet{2021AJ....161..189L}, 
+ \citet{2023ApJ...952..131M}, and
+\citet{2024AJ....167..159L}.
+We followed any guidance available from these respective studies for removing ages
+that were flagged as unreliable.
+Rather than compare the ages of the stars directly against each other, we then decided
+to compare them against a common benchmark: open clusters.
+
+Cluster memberships are a delicate subject.
+Figure~\ref{fig:agescalecompone} is in fact showing putative ages for a set of putative cluster members.
+For NGC6811 and NGC6819, we adopted all reported members from \citet{2018A&A...618A..93C,CantatGaudin_2020} and \citet{Kounkel_2020}. 
+For Melange-3 we used the members from \citet{2022AJ....164...88B} and required ``offset'' tangential velocities below 2\kms.
+For Cep-Her we used stars from Table~2 of \citet{Bouma_2022b}, requiring ``weights'' exceeding 0.1.
+These cluster lists almost certainly include non-members, in fractions that we do not aim to quantify.
+Even with these contaminants, we can compare the relative age distributions derived by
+the different studies.
+
+There are two main metrics for success in this test:
+the first is the extent to which the reported ages agree with the cluster age (accuracy).
+The second concerns how precise the reported ages are, and whether that precision agrees with the observed
+distribution of ages around unity.
+
+Starting with the isochrone ages from \citet{Berger_2020a_catalog}, the results should not be particularly surprising.
+The isochrone ages are most accurate and precise (in relative terms) for the oldest cluster, NGC6819.
+For the 1\,Gyr NGC6811, about half the stars have ages overestimated by a factor of three.
+For the youngest clusters, the disagreement is greater.
+The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at $\lesssim$1\,Gyr.
+An important virtue of these isochrone ages is that they are calculated for far more stars in the two largest
+clusters than the other methods.
+They are calculable for far more stars than rotation-based ages.
+
+The ``gyro-kinematic'' ages from \citet{2021AJ....161..189L} do...
+{\bf todo explain}
+
+The rotation-only ages calculated using the model described by \citet{2024AJ....167..159L} have some
+peculiarities compared against cluster ages.
+While NGC6811 seems to be in general agreement, NGC6819 has an abnormally large population of
+young reported systems, with apparently precise ages.  All the younger cluster members
+($<$300\,Myr) similarly have reported ages that are biased old, with narrow uncertainties.
+
+\citet{2023ApJ...952..131M} looks great.  {\bf todo explain method}.
+However it still fails for the youngest clusters.
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{merged_lit_ages_vs_cluster_ages.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Literature age scale comparison.}
+    Each point denotes the reported age of a star observed by Kepler,
+    normalized by the age of a stellar ensemble in which the star
+    is a putative member.
+    Rows show the available stellar ensembles, and
+    columns show the various age-measurement techniques described in Appendix~\ref{app:litagecomparison}.
+    Stars are sorted according to their median reported ages.
+    Note that the horizontal axis is logarithmic; the first dotted line
+    to the right of unity corresponds to a reported age ten times older
+    than the cluster age.
+	}
+	\label{fig:agescalecompone}
+\end{figure*}
+
 \section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}

### 2024-05-16 paper/collect.sh

@@ -37,8 +37,4 @@ cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pd
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
 
-cp $r/lit_ages_vs_cluster_ages/NGC6811_lit_ages_vs_cluster_ages.pdf .
-cp $r/lit_ages_vs_cluster_ages/Cep-Her_lit_ages_vs_cluster_ages.pdf .
-cp $r/lit_ages_vs_cluster_ages/NGC6819_lit_ages_vs_cluster_ages.pdf .
-cp $r/lit_ages_vs_cluster_ages/Melange3_lit_ages_vs_cluster_ages.pdf .
-cp $r/lit_ages_vs_cluster_ages/Theia520_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/merged_lit_ages_vs_cluster_ages.pdf .

### 2024-05-16 paper/ms.tex
update ms

@@
 
-Figure~\ref{fig:agescalecompone} shows the result of a literature comparison against
-ages reported by \citet{Berger_2020a_catalog}, \citet{2021AJ....161..189L}, 
- \citet{2023ApJ...952..131M}, and
+Figure~\ref{fig:agescalecompone} shows the result of a literature
+comparison against ages reported by \citet{Berger_2020a_catalog},
+\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
 \citet{2024AJ....167..159L}.
-We followed any guidance available from these respective studies for removing ages
-that were flagged as unreliable.
-Rather than compare the ages of the stars directly against each other, we then decided
-to compare them against a common benchmark: open clusters.
+We followed any guidance available from these respective studies for
+removing ages that were flagged as unreliable.
+Rather than comparing the ages of the stars directly against each other,
+we compared them against a common benchmark: open clusters.
 
 Cluster memberships are a delicate subject.
-Figure~\ref{fig:agescalecompone} is in fact showing putative ages for a set of putative cluster members.
-For NGC6811 and NGC6819, we adopted all reported members from \citet{2018A&A...618A..93C,CantatGaudin_2020} and \citet{Kounkel_2020}. 
-For Melange-3 we used the members from \citet{2022AJ....164...88B} and required ``offset'' tangential velocities below 2\kms.
-For Cep-Her we used stars from Table~2 of \citet{Bouma_2022b}, requiring ``weights'' exceeding 0.1.
-These cluster lists almost certainly include non-members, in fractions that we do not aim to quantify.
-Even with these contaminants, we can compare the relative age distributions derived by
-the different studies.
-
-There are two main metrics for success in this test:
-the first is the extent to which the reported ages agree with the cluster age (accuracy).
-The second concerns how precise the reported ages are, and whether that precision agrees with the observed
-distribution of ages around unity.
-
-Starting with the isochrone ages from \citet{Berger_2020a_catalog}, the results should not be particularly surprising.
-The isochrone ages are most accurate and precise (in relative terms) for the oldest cluster, NGC6819.
-For the 1\,Gyr NGC6811, about half the stars have ages overestimated by a factor of three.
+Figure~\ref{fig:agescalecompone} is showing putative ages for
+a set of putative cluster members.  For NGC6811 and NGC6819, we
+adopted candidate members from
+\citet{2018A&A...618A..93C,CantatGaudin_2020} and
+\citet{Kounkel_2020}. 
+For Melange-3 we used the candidate members reported by
+\citet{2022AJ....164...88B} and required ``offset'' tangential
+velocities below 2\kms.  For Cep-Her we
+used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
+``weights'' exceeding 0.1.
+These lists almost certainly include non-members, in fractions that we
+do not aim to quantify.  Even with these contaminants, we can compare
+the relative age distributions derived by the different studies.
+
+There are two main metrics for success in this test: the first is
+whether the reported ages agree with the cluster age (accuracy).  The
+second is whether the reported age precisions agree with their
+dispersion around the cluster age.
+
+Starting with the isochrone ages from \citet{Berger_2020a_catalog},
+the results should not be particularly surprising.
+The isochrone ages are most accurate and precise (in relative terms)
+for the oldest cluster, NGC6819.
+For the 1\,Gyr NGC6811, about half the stars have ages overestimated
+by a factor of three.
 For the youngest clusters, the disagreement is greater.
-The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at $\lesssim$1\,Gyr.
-An important virtue of these isochrone ages is that they are calculated for far more stars in the two largest
-clusters than the other methods.
+The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at
+$\lesssim$1\,Gyr.
+An important virtue of these isochrone ages is that they are
+calculated for far more stars in the two largest clusters than the
+other methods.
 They are calculable for far more stars than rotation-based ages.
@@
 
-The rotation-only ages calculated using the model described by \citet{2024AJ....167..159L} have some
-peculiarities compared against cluster ages.
-While NGC6811 seems to be in general agreement, NGC6819 has an abnormally large population of
-young reported systems, with apparently precise ages.  All the younger cluster members
-($<$300\,Myr) similarly have reported ages that are biased old, with narrow uncertainties.
+The rotation-only ages calculated using the model described by
+\citet{2024AJ....167..159L} have some peculiarities compared against
+cluster ages.
+While NGC6811 seems to be in general agreement, NGC6819 has an
+abnormally large population of young reported systems, with apparently
+precise ages.  All the younger cluster members ($<$300\,Myr) similarly
+have reported ages that are biased old, with narrow uncertainties.
 
@@
     normalized by the age of a stellar ensemble in which the star
-    is a putative member.
-    Rows show the available stellar ensembles, and
-    columns show the various age-measurement techniques described in Appendix~\ref{app:litagecomparison}.
-    Stars are sorted according to their median reported ages.
+    is a candidate member.
+    Rows show the available stellar ensembles, and columns show the
+    various age-measurement techniques described in
+    Appendix~\ref{app:litagecomparison}.
+    Stars are sorted according to their median reported ages, and
+    clusters are sorted by descending richness.
     Note that the horizontal axis is logarithmic; the first dotted line

### 2024-05-16 drivers/compare_literature_ages_to_cluster_ages.py
better regex

@@ -66,9 +66,12 @@ def get_cluster_dfs():
 
     # 
     # Get Theia520
+    # The regex is a negative lookahead assertion that ensures the character
+    # immediately following the end digit (e.g. UBC_"1") is not a digit.
     #
-    t520_sdf = sdf[sdf.cluster.str.contains("kcs20group_520") |
-                   sdf.cluster.str.contains("UBC_1")]
+    t520_sdf = sdf[sdf.cluster.str.contains("kcs20group_520(?!\d)") |
+                   sdf.cluster.str.contains(r"UBC_1(?!\d)")
+                   ]
     print(Counter(t520_sdf.reference_bibcode).most_common(n=10))
     theia520_df = t520_sdf.merge(gk_df, how='inner', on='source_id')
 

### 2024-05-17 paper/ms.tex
update ms

@@
 Figure~\ref{fig:agescalecompone} is showing putative ages for
-a set of putative cluster members.  For NGC6811 and NGC6819, we
+a set of spatially and kinematically selected candidate cluster members.  For NGC6811 and NGC6819, we
 adopted candidate members from
@@
     is a candidate member.
+    We used memberships from spatio-kinematic estimates in the
+    literature.
     Rows show the available stellar ensembles, and columns show the

### 2024-05-20 paper/ms.tex
update ms

@@
   %
-  Rotation periods are drawn from previous Kepler analyses; lithium is
+  Rotation periods are drawn from previous work; lithium is
   measured from new and archival Keck/HIRES spectra.
@@
   We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
-  and \nplwgyroagenograzing\ planets for which our gyrochronology
-  model is applicable; these rotation-based ages are over an order of
-  magnitude more precise than isochrone ages in this sample of stars.
-  %
-  The independent lithium ages agree with the rotation-based ages in
-  $\approx$90\% of cases, suggesting that most of the rotation-based
-  ages are not only precise, they are also accurate.
+  and \nplwgyroagenograzing\ planets for which our method is
+  applicable.
+  %
+  We find that our rotation-based ages accurately recover ages of
+  known Kepler field clusters; they also agree with
+  $\gtrsim$90\% of the independent lithium ages.
   %
@@
   has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3,Gyr. This trend holds true for both planet hosts and the overall
+  3\,Gyr.  This trend holds true for both planet hosts and the overall
   stellar sample.
@@
 
-Exoplanet science is enjoying an age of abundance, fueled by the discovery
-of thousands of worlds orbiting close to their host stars
+Exoplanet science is enjoying an age of abundance, fueled by the
+discovery of thousands of worlds orbiting close to their host stars
 \citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
 billions of years old.  This fact leaves many gaps in our knowledge of
-the exact physical and dynamical pathways that produce these objects.
-The reason is that processes such as thermal cooling
+the exact physical and dynamical histories of these objects.  The
+reason is that processes such as thermal cooling
 \citep{2007ApJ...659.1661F}, atmospheric loss
@@
 \citep{2017MNRAS.470.1750I} are expected to be most efficient over
-$\approx$10\,Myr to $\approx$1\,Gyr timescales. For most known
-exoplanets, these processes have already run their course.
-
-Young ($<$1\,Gyr) exoplanets represent one means of building the timeline for
-exoplanet evolution.  Informative individual young systems include
-V1298~Tau, an adolescent resonant chain of four transiting puffy
-planets that is a likely precusor to the compact multiplanet systems
-\citep{David_2019}, and HIP~67522, a Jupiter-sized planet with a
-Neptunian mass (\citealt{Rizzuto_2020}; Thao et al.~submitted).
-Population-level analyses have similarly suggested differences in the
-size distribution of young exoplanets relative to their older
-counterparts
+timescales of much less than 1\,Gyr.  For most known exoplanets, these
+processes have run their course.
+
+Young ($<$1\,Gyr) exoplanets therefore represent one means of building
+the timeline for exoplanet evolution.  Informative individual
+exemplars include V1298~Tau, an adolescent resonant chain of four
+transiting puffy planets that is a likely precusor to the compact
+multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
+planet with a Neptunian mass (\citealt{Rizzuto_2020}; Thao et
+al.~submitted).  Population-level analyses have similarly suggested
+differences in the size distribution of young exoplanets relative to
+their older counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 planet, and measure the star's age.  Each problem admits a range of
-solutions \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.  In
-this article we will consider planets previously known from transits,
-and infer new stellar ages using rotation and lithium.
+solutions
+\citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.
+In this article we will consider planets whose existence is previously
+known from transits, and infer new stellar ages using rotation and
+lithium.
 
 To begin, imagine that the ages of nearby stars in the Galaxy are
-uniformly distributed from 0--10\,Gyr \citep[e.g.~Fig~17
-of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
-young stars: under this assumption, $\approx$1\% of nearby stars have
-$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
-currently forming protoplanets \citep{2018A&A...617A..44K}, and of
-exoplanets evolving just after disk dispersal
-\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
-maximum achievable sample sizes by the tyranny of the galactic star
-formation rate.
+uniformly distributed from 0--10\,Gyr
+\citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
+clarifies the rarity of young stars: under this assumption,
+$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
+have $t$$<$1\,Gyr.  Studies of currently forming protoplanets
+\citep{2018A&A...617A..44K}, and of exoplanets evolving just after
+disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped
+in their maximum achievable sample sizes by the tyranny of the
+galactic star formation rate.
 
@@
 K2, TESS, and Gaia
-\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
+\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
 The strategy pursued by most groups during the 2010s was to focus on
@@
 
-The main problem with the ``cluster-first'' approach is that most
-nearby sub-gigayear stars are in the field.  In other words,
-$\lesssim$1\% of stars within the nearest $\approx$500\,pc have been
-associated with their birth cluster
+The ``cluster-first'' approach has a major limitation: most nearby
+sub-gigayear stars are in the field.  Only $\lesssim$1\% of stars
+within $\approx$500\,pc have been associated with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-One can appreciate the problem by querying NASA's Exoplanet Archive
+One can appreciate the problem by querying the NASA Exoplanet Archive
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
-than 1\,Gyr.  Imposing reasonable precision requirements
-(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say S/N$=$3)
-gives $\approx$50 ``young'' transiting planets at the time of writing.
-Most of these planets are in clusters, and were found by Kepler, K2,
-or TESS.  However, combined, those surveys have discovered
-$\approx$5{,}000 planets.  
-Barring a major discovery bias against sub-gigayear planets,
-a constant star formation history would imply that an order of magnitude
-more young transiting planets should be known.
+than 1\,Gyr.  Imposing a precision requirement of say $t$$<$1\,Gyr at
+2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
+the time of writing.  Most of these young planets are in clusters, and
+were found by Kepler, K2, or TESS.  However, these surveys have
+discovered $\approx$5{,}000 planets in total.  Assuming a constant
+star formation history, we would expect an order of magnitude more
+sub-gigayear transiting planets.
 
@@
 to assume a uniform age distribution for transiting planet host stars?
-Second, where are the missing young planets?  Briefly, we will find
-that uniform is wrong by a factor of a few, and that increased stellar
-activity for young stars may be a less important bias than the
-challenge of determining precise ages for field stars.
-
-%FIXME these paragraphs need improvements in how they incorporate
-%Lu+21 (meh), but certainly Mathur+2023.
-
-Given that Kepler's main mission ended in 2013, one might be forgiven
-for assuming that the ages of the Kepler planets are already known.
-Of course, they have been studied: \citet{Berger_2020b_rpage} and
-\citet{Petigura_2022} for instance calculated isochrone ages for
-Kepler Objects of Interest (KOIs).  These ages are most precise for
-stars whose luminosities and temperatures separate them from the
-main-sequence.  \citet{David_2021} leveraged isochronal ages from an
-earlier analogous study \citep{Fulton_2018}, and also used a relative
-gyrochrone age-ranking approach.  These studies were conducted in the
-broader context of investigating trends in exoplanet demographics,
-which also depend on stellar metallicity and mass
-\citep[e.g.][]{Petigura_2018}.  These studies offered limited
-sensitivity to the ages of $\lesssim$1\,Gyr planets, due to the nature
-of isochronal dating.
-
-Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
-between 1-4\,Gyr, and provides more complex asymmetric posteriors at
-earlier times \citep{Bouma_2023}.  The idea of using a star's
-spin-down as a clock has a rich history
-\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
-and physics-based models for the spin-down itself can clarify many
-aspects of how the stellar winds, internal structure, and magnetic
-dynamo all evolve
-\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
-context of Kepler, empirical models have been previously applied to
-derive rotation-based ages for subsets of the Kepler data
-\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic''
-ages have also been reported \citep{2021AJ....161..189L}.  There have
-been major improvements since these pioneering studies in:
+Second, where are the missing young planets?  We will find that
+uniform is wrong by a factor of a few, and that stellar activity may
+be a less significant bias than the challenge of precisely determining
+ages for field stars.
+
+Despite Kepler's main mission ending quite some time ago, the ages of
+many Kepler planets remain uncertain.  While isochrone ages have been
+calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
+Objects of Interest \citep[KOIs;][]{Petigura_2022}, isochrone ages are
+most precise for stars whose luminosities and temperatures separate
+them from the main sequence.  For $\lesssim$1\,Gyr stars, isochrone
+ages are therefore limited.
+
+Stellar rotation periods offer a promising alternative.  The idea of
+using a star's spin-down as a clock has a rich history
+\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
+Modern models yield ages precise to $\lesssim$30\% for FGK stars
+between 1-4\,Gyr, and constraining age posteriors at earlier times
+\citep{Bouma_2023}.  Physics-based models
+\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} are beginning to
+connect these empirical relations to the evolution of stellar winds,
+internal structure, and angular momentum transport.
+
+In the context of Kepler, rotation-based ages have been reported for
+planet-hosting stars since the early data releases
+\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
+focused on rotation-based ages for Kepler have further explored
+incorporating information from stellar kinematics
+\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
+features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
+analysis is motivated by a few factors that we believe constitute
+improvements, particularly for the ages of sub-gigayear stars.   These
+factors are as follows.
 \begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item The extent of the Kepler Object of Interest catalog, and our
-    vetting of false positives within it;
-  %
-  \item Available measurements of stellar rotation periods in open
-    clusters
+  \item The Kepler Object of Interest catalog, and our vetting of
+    false positives within it, has reached maturity
+    \citep[e.g.][]{Thompson_2018};
+  %
+  \item Measured rotation periods of FGK stars in open clusters now
+    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
+    but also how the astrophysical dispersion of stars around this average
+    converges by the $\approx$700\,Myr age of Praesepe
     \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
   %
-  \item Our ability to model not only the mean evolution but also the
-    intrinsic dispersion in rotation through time \citep{Bouma_2023}.
-    For Sun-like stars, the new model is broadly consistent with older
-    calibrations \citep[e.g.][]{Mamajek_2008}.  For late G dwarfs and
-    K dwarfs, the new calibration differs by factors of a few;
-  %
-  \item Our ability to identify binary stars, which can be problematic
-    for estimates of rotation-based ages.
+  \item We can fit the open cluster data, and use these fits to infer
+    ages by marginalizing over the range of ages that might explain
+    any one star's rotation period \citep{Bouma_2023}.
+    For Sun-like stars, this model is consistent with older
+    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
+    of uncertainty propagation.  For late G dwarfs and
+    K dwarfs, the calibration differs by factors of a few;
+  %
+  \item We can identify binary stars at greater fidelity today than
+    in the past, which can clarify otherwise problematic estimates of
+    rotation-based ages.
+  %
+  \item New open clusters in various stages of dissolution have been
+    found in the Kepler field
+    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,2022AJ....164...88B}.
+    These clusters offer a sample of stars for which we can test the
+    reliability of single-star age-dating methods.
 \end{enumerate}
-These improvements form the basis for why we believe the present
-analysis to be necessary.
-
-Beyond rotation-based ages, recent years have also yielded
-improvements in the lithium age scale.  Lithium ages include two
-distinct regimes: depletion boundary ages for M dwarfs and brown dwarfs in star
+
+Independently, recent years have also yielded improvements in the
+lithium age scale.  Lithium ages include two distinct regimes:
+depletion boundary ages for M dwarfs and brown dwarfs in star
 clusters, and the less precise decline-based ages for individual field
@@
 Empirical understanding has however improved due to the work by
-\citet{Jeffries_2023}, who modeled the evolution of 
-the \ion{Li}{1} 6708\,\AA\ doublet equivalent width (EW) through
-time using a set of 6{,}200 stars in 52
-open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
-stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
-spectral type since the K dwarfs lose their surface lithium much
-faster than G dwarfs.  The precision of the ages reported by the
-\citet{Jeffries_2023} method in this regime are in the range of
+\citet{Jeffries_2023}, who modeled the time evolution of the
+\ion{Li}{1} 6708\,\AA\ equivalent width (EW) using a set of 6{,}200
+stars in 52 open clusters.  Two-sided lithium ages are useful for
+Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, with a strong
+dependence on spectral type since the K dwarfs lose their surface
+lithium much faster than G dwarfs.  The precision of the ages reported
+by the \citet{Jeffries_2023} method in this regime are in the range of
 0.3-0.5~dex.
@@
 stellar parameters other than ages in Section~\ref{sec:stellarprops}.
-We describe and validate our age-dating methods in
-Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
+We describe our age-dating methods in
+Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them using
+stars in clusters in the Kepler field.  The population-level
 trends for both stars and planets are described in
@@
 
-{\it Near the main-sequence (bit 9)}---Many stars observed by Kepler
+{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
 are far from the main sequence: this can be seen in both panels of
 Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
-ages because {\it i)} point sources far from the main-sequence include
+ages because {\it i)} point sources far from the main sequence include
 unresolved photometric binaries, and {\it ii}) effects other than
 magnetic braking influence the rotation periods of bona fide single
-stars far from the main-sequence.  The added complications of
+stars far from the main sequence.  The added complications of
 reddening and metallicity further complicate matters.  After exploring
@@
 ages are lower limits.  In such cases,  the inferred age posterior is
-strongly influenced by {\it i)} the assumed intrinsic dispersion in
-the lithium model, and {\it ii)} by the prior.  In these instances, we
+strongly influenced by the assumed intrinsic dispersion in
+the lithium model, and by the prior.  In these instances, we
 quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
@@
 analysis to a separate publication.
+Some care is required since beyond age,
+exoplanet demographics also depend on stellar metallicity and mass
+\citep[e.g.][]{Petigura_2018}.
 
@@
   Bit 8: Crowded?
-  Bit 9: Far from main-sequence?
+  Bit 9: Far from main sequence?
   Bit 10: \citetalias{Santos_2021} CP/CB?

### 2024-05-20 paper/ms.tex
update ms

@@
     found in the Kepler field
-    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,2022AJ....164...88B}.
+    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
     These clusters offer a sample of stars for which we can test the
@@
 to 4\,Gyr, the age of M67
-\citep[see][]{2022ApJ...938..118D,Gruner_2023}.  To estimate the
+\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the
 stellar spin-down rate as a function of time and stellar temperature,
@@
 the posterior probability distributions by drawing ten random samples
-from each posterior distribution, and then computing the re-weighted
+from each posterior, and then computing the re-weighted
 histogram of the resulting samples.  The uncertainties assume Poisson
@@
 The age distributions of Kepler stars and planet hosts are
-statistically quite similar.  A simple diagnostic of whether the two
+nearly indistinguishable.  A simple diagnostic of whether the two
 distributions are drawn from the same underlying distribution is the
@@
 
-{\it MELANGE-3}---\citet{2022AJ....164...88B} reported a
+{\it MELANGE-3}---\citet{Barber_2022} reported a
 $105\pm10$\,Myr association in the Kepler field and derived its age
@@
 Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
-this association \citep{2022AJ....164...88B}.  Our rotation-based
+this association.  Our rotation-based
 age-dating independently recovers these systems, with $t_{\rm
@@
 age being slightly older than the suggested Pleiades age for the
-assocation agrees with Figure~4 of \citet{2022AJ....164...88B}.
+assocation agrees with Figure~4 of \citet{Barber_2022}.
 

### 2024-05-20 paper/ms.tex
ms: WIP cluster age refactor

@@
 quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
+
+\section{Cluster Age Comparison}
+\label{sec:litagecomparison}
+
+Stars that formed in the same birth cluster, at approximately the same time,
+are the gold standard for the astronomical age scale
+\citep{Soderblom10}.
+Before Gaia, only a handful of $\gtrsim$1\,Gyr open clusters
+were known in the Kepler field.
+
+The gold standard for stellar ages comes from open clusters
+
+Figure~\ref{fig:agescalecompone} shows the result of a literature
+comparison against ages reported by \citet{Berger_2020a_catalog},
+\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
+\citet{2024AJ....167..159L}.
+We followed any guidance available from these respective studies for
+removing ages that were flagged as unreliable.
+Rather than comparing the ages of the stars directly against each other,
+we compared them against a common benchmark: open clusters.
+
+Cluster memberships are a delicate subject.
+Figure~\ref{fig:agescalecompone} is showing putative ages for
+a set of spatially and kinematically selected candidate cluster members.  For NGC6811 and NGC6819, we
+adopted candidate members from
+\citet{2018A&A...618A..93C,CantatGaudin_2020} and
+\citet{Kounkel_2020}. 
+For Melange-3 we used the candidate members reported by
+\citet{2022AJ....164...88B} and required ``offset'' tangential
+velocities below 2\kms.  For Cep-Her we
+used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
+``weights'' exceeding 0.1.
+These lists almost certainly include non-members, in fractions that we
+do not aim to quantify.  Even with these contaminants, we can compare
+the relative age distributions derived by the different studies.
+
+There are two main metrics for success in this test: the first is
+whether the reported ages agree with the cluster age (accuracy).  The
+second is whether the reported age precisions agree with their
+dispersion around the cluster age.
+
+Starting with the isochrone ages from \citet{Berger_2020a_catalog},
+the results should not be particularly surprising.
+The isochrone ages are most accurate and precise (in relative terms)
+for the oldest cluster, NGC6819.
+For the 1\,Gyr NGC6811, about half the stars have ages overestimated
+by a factor of three.
+For the youngest clusters, the disagreement is greater.
+The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at
+$\lesssim$1\,Gyr.
+An important virtue of these isochrone ages is that they are
+calculated for far more stars in the two largest clusters than the
+other methods.
+They are calculable for far more stars than rotation-based ages.
+
+The ``gyro-kinematic'' ages from \citet{2021AJ....161..189L} do...
+{\bf todo explain}
+
+The rotation-only ages calculated using the model described by
+\citet{2024AJ....167..159L} have some peculiarities compared against
+cluster ages.
+While NGC6811 seems to be in general agreement, NGC6819 has an
+abnormally large population of young reported systems, with apparently
+precise ages.  All the younger cluster members ($<$300\,Myr) similarly
+have reported ages that are biased old, with narrow uncertainties.
+
+\citet{2023ApJ...952..131M} looks great.  {\bf todo explain method}.
+However it still fails for the youngest clusters.
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.98\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+		{\bf Literature age comparison.}
+		Each point denotes a reported age,
+		normalized by the age of a stellar ensemble in which the star
+		is a candidate member.
+		We used candidate cluster members as reported by previous studies based on spatial and kinematic
+		information.  Although some field interlopers may be present in the 
+		membership lists,
+		any outliers can be compared between different methods on a
+		relative basis.
+		The number of stars in any given cluster follows by crossmatching every study against
+		this fixed cluster list; certain methods report ages for more stars than others.
+		Horizontal scatter is artificially added to visually
+		clarify the statistical age uncertainties.
+	}
+	\label{fig:agescalecompone}
+\end{figure*}
 
@@
 
-\appendix
-\section{Literature Age Scale Comparison}
-\label{app:litagecomparison}
-
-Figure~\ref{fig:agescalecompone} shows the result of a literature
-comparison against ages reported by \citet{Berger_2020a_catalog},
-\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
-\citet{2024AJ....167..159L}.
-We followed any guidance available from these respective studies for
-removing ages that were flagged as unreliable.
-Rather than comparing the ages of the stars directly against each other,
-we compared them against a common benchmark: open clusters.
-
-Cluster memberships are a delicate subject.
-Figure~\ref{fig:agescalecompone} is showing putative ages for
-a set of spatially and kinematically selected candidate cluster members.  For NGC6811 and NGC6819, we
-adopted candidate members from
-\citet{2018A&A...618A..93C,CantatGaudin_2020} and
-\citet{Kounkel_2020}. 
-For Melange-3 we used the candidate members reported by
-\citet{2022AJ....164...88B} and required ``offset'' tangential
-velocities below 2\kms.  For Cep-Her we
-used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
-``weights'' exceeding 0.1.
-These lists almost certainly include non-members, in fractions that we
-do not aim to quantify.  Even with these contaminants, we can compare
-the relative age distributions derived by the different studies.
-
-There are two main metrics for success in this test: the first is
-whether the reported ages agree with the cluster age (accuracy).  The
-second is whether the reported age precisions agree with their
-dispersion around the cluster age.
-
-Starting with the isochrone ages from \citet{Berger_2020a_catalog},
-the results should not be particularly surprising.
-The isochrone ages are most accurate and precise (in relative terms)
-for the oldest cluster, NGC6819.
-For the 1\,Gyr NGC6811, about half the stars have ages overestimated
-by a factor of three.
-For the youngest clusters, the disagreement is greater.
-The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at
-$\lesssim$1\,Gyr.
-An important virtue of these isochrone ages is that they are
-calculated for far more stars in the two largest clusters than the
-other methods.
-They are calculable for far more stars than rotation-based ages.
-
-The ``gyro-kinematic'' ages from \citet{2021AJ....161..189L} do...
-{\bf todo explain}
-
-The rotation-only ages calculated using the model described by
-\citet{2024AJ....167..159L} have some peculiarities compared against
-cluster ages.
-While NGC6811 seems to be in general agreement, NGC6819 has an
-abnormally large population of young reported systems, with apparently
-precise ages.  All the younger cluster members ($<$300\,Myr) similarly
-have reported ages that are biased old, with narrow uncertainties.
-
-\citet{2023ApJ...952..131M} looks great.  {\bf todo explain method}.
-However it still fails for the youngest clusters.
-
-\begin{figure*}[!t]
-	\begin{center}
-		\subfloat{
-			\includegraphics[width=0.98\textwidth]{merged_lit_ages_vs_cluster_ages.pdf}
-		}
-	\end{center}
-	\vspace{-0.5cm}
-	\caption{
-    {\bf Literature age scale comparison.}
-    Each point denotes the reported age of a star observed by Kepler,
-    normalized by the age of a stellar ensemble in which the star
-    is a candidate member.
-    We used memberships from spatio-kinematic estimates in the
-    literature.
-    Rows show the available stellar ensembles, and columns show the
-    various age-measurement techniques described in
-    Appendix~\ref{app:litagecomparison}.
-    Stars are sorted according to their median reported ages, and
-    clusters are sorted by descending richness.
-    Note that the horizontal axis is logarithmic; the first dotted line
-    to the right of unity corresponds to a reported age ten times older
-    than the cluster age.
-	}
-	\label{fig:agescalecompone}
-\end{figure*}
-
 \section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}

### 2024-05-20 paper/collect.sh

@@ -37,4 +37,4 @@ cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pd
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
 
-cp $r/lit_ages_vs_cluster_ages/merged_lit_ages_vs_cluster_ages.pdf .
+cp $r/lit_ages_vs_cluster_ages/olympic_merged_lit_ages_vs_cluster_ages.pdf .

### 2024-05-20 paper/ms.tex
update ms

@@
 \newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
-\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics, 77 Massachusetts Avenue, Cambridge, MA 02139}
-\newcommand{\berkeley}{Astronomy Department, University of California, Berkeley, CA 94720, USA}
+\newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
+77 Massachusetts Avenue, Cambridge, MA 02139, USA}
+\newcommand{\berkeley}{Astronomy Department, University of California,
+Berkeley, CA 94720, USA}
+\newcommand{\cmu}{Department of Statistics \& Data Science, Carnegie
+Mellon University, Pittsburgh, PA 15213, USA}
 
@@
 \author[0000-0002-0514-5538]{Luke~G.~Bouma}
-%\altaffiliation{51 Pegasi b Fellow}
+\altaffiliation{51 Pegasi b Fellow}
 \affiliation{\caltech}
-\affiliation{51 Pegasi b Fellow}
-
-\author{coauthors, including but not limited to
-Hillenbrand, Howard, Isaacson, Palumbo}
+%\affiliation{51 Pegasi b Fellow}
+
+\author{and coauthors, including but not limited to...}
+
+\author{Lynne~A.~Hillenbrand}
+\affiliation{\caltech}
+
+\author[0000-0001-8638-0320]{Andrew W. Howard}
+\affiliation{\caltech}
+
+\author[0000-0002-0531-1073]{Howard Isaacson}
+\affiliation{\berkeley}
+
+\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
+\affiliation{\cmu}
 
@@
   We find that our rotation-based ages accurately recover ages of
-  known Kepler field clusters; they also agree with
+  stars in open clusters; they also agree with
   $\gtrsim$90\% of the independent lithium ages.
   %
-  Our resulting catalog includes \nplyounggyrotwosigma\ planets
+  The results include \nplyounggyrotwosigma\ planets
   younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
@@
   has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3\,Gyr.  This trend holds true for both planet hosts and the overall
+  3\,Gyr.  This trend holds for both planet hosts and for the overall
   stellar sample.
@@
 to a local window centered on the lithium line, using a window width
-of 15\,\AA\ (we considered 10\,\AA\ and 20\,\AA; the results were
+of 15\,\AA\ (we also considered 10\,\AA\ and 20\,\AA; the results were
 consistent).  We continuum-normalized by fitting a third-degree
@@
 estimated uncertainties on the line width through a Monte Carlo
-procedure that bootstraped against the local scatter in the continuum.
+procedure that bootstrapped against the local scatter in the continuum.
 
@@
     for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024 that comprised
-    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
+    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the ``mean''
     isochrones from the same models as in Figure~\ref{fig:models}.  The
-    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
+    intrinsic dispersion at $\gtrsim$1\,Gyr becomes much larger
     than changes in the mean.  Some stars with lithium detections do
@@
 
-\section{Ages From Lithium}
-\label{sec:liage}
-
-Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
-the lithium 6708\AA\ absorption doublet, plotted over the mean
-isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
-declare an upper-limit for plotting purposes if the 1$\sigma$ lower
-limit on the equivalent width is below 10\,m\AA.  Our measured EWs
-span -40 to 250\,m\AA; all negative EWs have uncertainties that are
-statistically consistent with zero.
-
-To calculate lithium ages, we use \texttt{EAGLES} (git commit
-\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
-a purely empirical interpolation approach.  \texttt{EAGLES} 
-was calibrated on lithium measurements of stars observed by the
-Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
-to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
-age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
-uncertainties for purposes of interfacing with \texttt{EAGLES} by
-taking the maximum of the measured positive and negative 1$\sigma$
-uncertainty intervals.
-
-In part because many of our EWs are upper limits, many of the lithium
-ages are lower limits.  In such cases,  the inferred age posterior is
-strongly influenced by the assumed intrinsic dispersion in
-the lithium model, and by the prior.  In these instances, we
-quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
-
-\section{Cluster Age Comparison}
-\label{sec:litagecomparison}
-
-Stars that formed in the same birth cluster, at approximately the same time,
-are the gold standard for the astronomical age scale
-\citep{Soderblom10}.
-Before Gaia, only a handful of $\gtrsim$1\,Gyr open clusters
-were known in the Kepler field.
-
-The gold standard for stellar ages comes from open clusters
-
-Figure~\ref{fig:agescalecompone} shows the result of a literature
-comparison against ages reported by \citet{Berger_2020a_catalog},
-\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
-\citet{2024AJ....167..159L}.
-We followed any guidance available from these respective studies for
-removing ages that were flagged as unreliable.
-Rather than comparing the ages of the stars directly against each other,
-we compared them against a common benchmark: open clusters.
-
-Cluster memberships are a delicate subject.
-Figure~\ref{fig:agescalecompone} is showing putative ages for
-a set of spatially and kinematically selected candidate cluster members.  For NGC6811 and NGC6819, we
-adopted candidate members from
-\citet{2018A&A...618A..93C,CantatGaudin_2020} and
-\citet{Kounkel_2020}. 
-For Melange-3 we used the candidate members reported by
-\citet{2022AJ....164...88B} and required ``offset'' tangential
-velocities below 2\kms.  For Cep-Her we
-used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
-``weights'' exceeding 0.1.
-These lists almost certainly include non-members, in fractions that we
-do not aim to quantify.  Even with these contaminants, we can compare
-the relative age distributions derived by the different studies.
-
-There are two main metrics for success in this test: the first is
-whether the reported ages agree with the cluster age (accuracy).  The
-second is whether the reported age precisions agree with their
-dispersion around the cluster age.
-
-Starting with the isochrone ages from \citet{Berger_2020a_catalog},
-the results should not be particularly surprising.
-The isochrone ages are most accurate and precise (in relative terms)
-for the oldest cluster, NGC6819.
-For the 1\,Gyr NGC6811, about half the stars have ages overestimated
-by a factor of three.
-For the youngest clusters, the disagreement is greater.
-The uncertainties seem accurate at 2.5\,Gyr, but perhaps less so at
-$\lesssim$1\,Gyr.
-An important virtue of these isochrone ages is that they are
-calculated for far more stars in the two largest clusters than the
-other methods.
-They are calculable for far more stars than rotation-based ages.
-
-The ``gyro-kinematic'' ages from \citet{2021AJ....161..189L} do...
-{\bf todo explain}
-
-The rotation-only ages calculated using the model described by
-\citet{2024AJ....167..159L} have some peculiarities compared against
-cluster ages.
-While NGC6811 seems to be in general agreement, NGC6819 has an
-abnormally large population of young reported systems, with apparently
-precise ages.  All the younger cluster members ($<$300\,Myr) similarly
-have reported ages that are biased old, with narrow uncertainties.
-
-\citet{2023ApJ...952..131M} looks great.  {\bf todo explain method}.
-However it still fails for the youngest clusters.
-
 \begin{figure*}[!t]
@@
 
+\section{Ages From Lithium}
+\label{sec:liage}
+
+Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
+the lithium 6708\AA\ absorption doublet, plotted over the mean
+isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
+declare an upper-limit for plotting purposes if the 1$\sigma$ lower
+limit on the equivalent width is below 10\,m\AA.  Our measured EWs
+span -40 to 250\,m\AA; all negative EWs have uncertainties that are
+statistically consistent with zero.
+
+To calculate lithium ages, we use \texttt{EAGLES} (git commit
+\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
+a purely empirical interpolation approach.  \texttt{EAGLES} 
+was calibrated on lithium measurements of stars observed by the
+Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
+to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
+age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
+uncertainties for purposes of interfacing with \texttt{EAGLES} by
+taking the maximum of the measured positive and negative 1$\sigma$
+uncertainty intervals.
+
+In part because many of our EWs are upper limits, many of the lithium
+ages are lower limits.  In such cases,  the inferred age posterior is
+strongly influenced by the assumed intrinsic dispersion in
+the lithium model, and by the prior.  In these instances, we
+quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
+
+\section{Cluster Age Comparison}
+\label{sec:litagecomparison}
+
+Stars that formed in the same birth cluster 
+are the gold standard for the astronomical age scale
+\citep{Soderblom_2010}.
+Before Gaia, the few $\gtrsim$1\,Gyr open clusters
+known in the Kepler field had been visually identified by
+\citet{1864RSPT..154....1H}.
+Gaia proper motions and distances 
+enabled the discovery of stellar ensembles that are typically more diffuse,
+but which nonetheless
+share a common age 
+based on e.g.~isochrone, rotation, and lithium dating
+\cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
+
+In Figure~\ref{fig:agescalecompone}, we compare 
+ages from the literature against the ages of stars in the richest
+clusters in the Kepler field.
+In particular, we drew ages for Kepler stars from \citet{Berger_2020a_catalog},
+\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
+\citet{2024AJ....167..159L}.
+We followed any guidance available from these respective studies for
+removing ages that were explicitly flagged as unreliable.
+
+Cluster memberships are a delicate subject.
+Figure~\ref{fig:agescalecompone} is showing reported ages for
+a set of spatially and kinematically selected set of stars that could be cluster members, or
+they could be field interlopers.  For NGC6811 and NGC6819, we
+adopted candidate members from
+\citet{2018A&A...618A..93C,CantatGaudin_2020} and
+\citet{Kounkel_2020}. 
+For Melange-3 we used the candidates reported by
+\citet{Barber_2022} and required ``offset'' tangential
+velocities below 2\,\kms.  For Cep-Her we
+used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
+``weights'' exceeding 0.1.
+Even with contaminants, we can compare
+the relative age distributions derived by the different studies because in all
+instances we are comparing reported ages against a fixed list of stars.
+
+There are two main metrics for success in this test: accuracy
+in whether the reported ages agree with the cluster age, and precision in
+whether the reported age uncertainties agree with their
+dispersion around the cluster age.
+
+The point of Figure~\ref{fig:agescalecompone} is hopefully clear.
+While previous studies have been able to derive accurate ages for
+stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have historically been
+overestimated by anywhere from 0.5--2\,dex.
+The situation is particularly severe for the 40\,Myr stars in Cep-Her,
+but it extends out to 150--300\,Myr stars in Melange-3 and Theia~520.
+For isochrone and kinematic ages,
+this is not surprising because these methods use stellar parameters that do
+not appreciably change at $t$$\lesssim$1\,Gyr.
+For the rotation-based ages quoted by
+\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the discrepancy
+between their ages and the cluster age scale is likely caused by details of the calibration 
+methodologies.
+The physical spin-down model used by
+\citet{2023ApJ...952..131M},
+and the gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
+may not reproduce
+the rotation sequences of open clusters such as Praesepe, the Hyades, and the Pleiades.
+The statistical age uncertainties yielded by these approaches
+similarly may not marginalize over the intrinsic
+scatter in rotation periods present in stars of a fixed mass at any given age.
+
+%The rotation-based ages that we have calculated have a precision
+%limited by the intrinsic scatter in the cluster rotation sequences.
+%Within the age range accessible through the clusters, they seem to be
+%mostly accurate,
+%with the one exception being that our ages for stars in Cep-Her 
+%have a typical median of $\approx$80\,Myr, rather than 40\,Myr.
+%The reason is that 80\,Myr is the lower bound of 
+%the \texttt{gyro-interp}  age grid; pre-main-sequence
+%spin-up is outside of our calibration regime.
+
 \section{Results}
@@
 
-In this work, we were interested in not only the overall age
+We were interested in not only the overall age
 distribution of the Kepler field, but also in the reliability of
-individual ages for young stars known to host planets.  The quality
-bitmask scheme in Table~\ref{tab:planets} provided one path toward
-this end.  Our other method was to compare the cases for which lithium
-and rotation overlap in sensitivity -- G and K dwarfs with
-$t$$\lesssim$1\,Gyr.  Our results for such stars include
+individual ages for young stars known to host planets.  
+We assessed this reliability using
+quality
+bitmasks (Table~\ref{tab:planets}), 
+a comparison against open cluster ages (Section~\ref{sec:litagecomparison}),
+and a comparison between rotation and lithium ages.
+The open cluster comparison suggested that our ages were generally
+accurate at $\ll$0.5\,dex across 0.04--2.5\,Gyr, comparable to their
+stated precision level.
+Nonetheless,
+outliers do exist in the upper-left panel of Figure~\ref{fig:agescalecompone},
+typically at most 10\% of the population in each cluster.
+These stars could be either field star contaminants, or else anomalous
+stars whose physical rotation histories were altered by processes not
+captured by our statistical uncertainties.
+Similarly, although
+we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
+for 
 \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
-stars for which the two age indicators are both constraining and in
-agreement (stars with $Q_{\rm star}$ bits zero through nine not
-raised, and no planet quality flags raised).  Conversely, we found
-only one system, Kepler-786, with a precise lithium age
+stars, we did find
+one system, Kepler-786, with a precise lithium age
 (\kepseveneightsix\,Myr), but no rotation signal.
@@
 A broader question, beyond the scope of this work, concerns the ages
-of the older $\gtrsim$3\,Gyr stars that we do not treat.  For
+of stars $\gtrsim$4\,Gyr old.  For
 age-dating models based only on rotation rates, or else on a
@@
 \citep{2021AJ....161..189L} seems likely to remain most useful for
-stellar ensembles, due to the stochastically of kinematic heating.
+stellar ensembles, due to the stochasticity of kinematic heating.
 
@@
 
+\appendix
 \section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
@@
 \begin{enumerate}
-  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
-    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
-    Could compare against asteroseis too.
-    %
-  \item Intro needs better discussion of Mathur2023 and Lu2021
+%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
+%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
+%    Could compare against asteroseis too.
+%    %
+%  \item Intro needs better discussion of Mathur2023 and Lu2021
     %

### 2024-05-20 gyrojo/plotting.py
update plot

@@ -475,9 +475,9 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
         color = 'k'
         alpha = 0.9
         if not show_dispersion and not show_dispersionpoints:
-            alpha = 0.4
+            alpha = 0.8
         ax.plot(
-            Teff_model, li_model, color='k', linewidth=1, zorder=10,
+            Teff_model, li_model, color=f'C{ix}', linewidth=1, zorder=10,
             alpha=alpha, ls=ls, label=agestr
         )
 

### 2024-05-21 paper/ms.tex
update ms

@@
 This work is focused on Kepler stars for which ages can be inferred
-using either rotation, lithium, or both.  These ``age-dateable stars''
+using either rotation, lithium, or both.  Such stars
 are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
@@
 reliable'' in their visual analysis of previously reported KOI
-rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
+rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013}, \citet{Mazeh_2015}
 and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
@@
 
-Finally, since our scope is focused on rotation-based age dating, we
+Finally, since our scope is focused on rotation-based ages, we
 restricted our attention to stars with reported $P_{\rm
-rot}$$<$45\,days.  FGK stars with slower rotation periods spin much
-slower than the stars in the oldest open clusters currently used to
-calibrate our gyrochronology model.  Figure~\ref{fig:prot_vs_teff}
+rot}$$<$45\,days.  The slowest-rotating FGK
+stars in the open clusters used to
+calibrate our gyrochronology model have $P_{\rm rot}$$\approx$35\,days.
+Figure~\ref{fig:prot_vs_teff}
 shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
@@
 The final piece of our analysis involves assessing ages based on the
-\ion{Li}{1} 6708\,\AA\ doublet.  Specifically, we analyzed spectra
+\ion{Li}{1} 6708\,\AA\ doublet.  We analyzed spectra
 from the High Resolution Echelle Spectrometer (HIRES;
@@
 spectra have since been acquired, and our approach and selection
-function are different.  We therefore perform our own line width
+function are different.  We therefore performed our own line width
 measurements on
@@
 stars hosting \nplwithspec\ planets.  About half of these stars have
-rotation periods (\nstwithspecandprot\ stars and
+measured rotation periods (\nstwithspecandprot\ stars and
 \nplwithspecandprot, respectively).   For stars with multiple spectra
-available at varying ranges of signal-to-noise (S/N), we analyzed only
+available, we analyzed only
 the spectrum with the highest number of counts.  The resulting spectra
@@
 stars, and so the continuum in the vicinity of the \ion{Li}{1}
-6708\,\AA\ doublet is well-defined.  We Doppler corrected the spectra
+6708\,\AA\ doublet is well-defined.  We Doppler-corrected the spectra
 to a common reference wavelength by cross-correlating against a high
@@
 %crossmatch
-of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3.  The
+of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3 (accessible
+online at \url{https://gaia-kepler.fun}).  The
 separation distribution of the Kepler-Gaia DR3 crossmatches is such
@@
 magnetic braking influence the rotation periods of bona fide single
-stars far from the main sequence.  The added complications of
+stars far from the main sequence.  The added concerns of
 reddening and metallicity further complicate matters.  After exploring
@@
 g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
-the details of how exactly this locus is constructed are arbitrary, 
-%and convoluted,\footnote{In practice, for K dwarfs we fitted an
-%$N^{\rm th}$-order polynomial to the set of stars for which $t_{\rm
-%B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$, where
-%$t_{\rm B20,iso}$ was the isochronal age reported by
-%\citet{Berger_2020a_catalog}. We let the order of the fit vary from
-%$N$=1 to 10, minimized the Bayesian Information Criterion,
-%and found $N$=6.  For F and G dwarfs, this procedure failed to produce
-%a reasonable locus because of shifts in the relative precision of
-%the isochrone ages.  We therefore visually selected KOIs that appeared
-%to be near the ``main sequence'', fitted a separate polynomial to
-%their locations, and then stitched the resulting polynomials from the
-%F and G dwarfs to that of the K dwarfs.}
+the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
+	the method was as follows.
+The orange line in Figure~\ref{fig:stellarprops} is	defined as
+the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
+We defined $f_0$ by fitting an
+$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
+B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
+and $T_{\rm eff}\in[3700,5900]$\,K, where
+$t_{\rm B20,iso}$ was the isochronal age reported by
+\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
+$N$=1 to 10, minimized the Bayesian Information Criterion,
+and found $N$=6.  
+By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
+However for F and G dwarfs, the resulting locus
+allowed for stars that were `too evolved''; the density of KOIs 
+was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
+We therefore visually selected KOIs that appeared
+to be near the main sequence -- i.e.~ most of the blue
+points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
+polynomial $f_1$ through a similar BIC-minimization procedure,
+and found $N$=3.
+The portion of the orange locus from $\approx$5300--6200\,K
+is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
+$\approx$0.1\,dex above the ``KOI main sequence''.
+The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
+We emphasize that for stars
+surface gravities above the orange locus, we do report rotation based ages;
+they are simply flagged as not being near the main sequence.  }
 the general aim is to flag stars for which there is evidence based on
@@
 {\it Field star ages from other surveys}:
-While this study focused on deriving rotation and lithium-based ages
-for Kepler planets, surveys such as K2, TESS, HAT, WASP, NGTS, ZTF,
-and ATLAS have different selection functions.  Applying rotation-based
-age dating to their discoveries could clarify the youth and
-evolutionary trends of these populations.  Future prospects like PLATO
+While this study focused on
+Kepler, other photometric surveys (e.g.~K2, TESS, HAT, WASP, NGTS, ZTF,
+ATLAS) open opportunities for age-dating a far broader set
+of stars and planets.
+Future prospects also include PLATO
 \citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
-Galactic Plane Time Domain Survey seem poisted to further expand the
-sample of stars with precisely determined ages, including those known
-to host planets.
+Galactic Plane Time Domain Survey.
+The long-baseline surveys 
+that specifically target different regions of the Galaxy
+seem most likely to yield new insight into its star
+formation history.
 

### 2024-05-21 paper/ms.tex
update ms

@@
 \defcitealias{Berger_2020a_catalog}{B20}
+
+% Redefine footnote command to not have indent
+%\makeatletter
+%\long\def@makefntext#1{\noindent\footnotesize@makefnmark#1}
+%\makeatother
+\setlength{\footnotesep}{0pt}
 
@@
 Gaia proper motions and distances 
-enabled the discovery of stellar ensembles that are typically more diffuse,
+enabled the discovery of stellar ensembles that are more diffuse,
 but which nonetheless
@@
 We followed any guidance available from these respective studies for
-removing ages that were explicitly flagged as unreliable.
+removing ages that were explicitly flagged as unreliable, and plotted
+stars within $1\sigma$ of zero as being upper limits.
 
@@
 methodologies.
-The physical spin-down model used by
+The physical spin-down models used by
 \citet{2023ApJ...952..131M},

### 2024-05-21 paper/ms.tex
update ms

@@
 
-\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
+%\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
+\title{A Dearth of Young Stars and Planets in the Kepler Field}
 
@@
 
-\author{and coauthors, including but not limited to...}
+%\author{and coauthors, including but not limited to...}
 
@@
 
+\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
+\affiliation{\cmu}
+
 \author[0000-0002-0531-1073]{Howard Isaacson}
 \affiliation{\berkeley}
-
-\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
-\affiliation{\cmu}
 
@@
   We find that our rotation-based ages accurately recover ages of
-  stars in open clusters; they also agree with
+  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
   $\gtrsim$90\% of the independent lithium ages.
@@
   %
-  More broadly, we find that the star formation rate in the Kepler field
+  This is far fewer than expected under the classic assumption of a uniform
+  star formation history: the star formation rate in the Kepler field
   has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
@@
   %
-  This ``demographic cliff'' is a known feature of the Galaxy's star
-  formation history, and it helps clarify the relative balance of
+  This ``demographic cliff'' in the Galaxy's star
+  formation history has been previously reported, and it helps clarify the relative balance of
   young and old planets that have been detected to date using
@@
 the timeline for exoplanet evolution.  Informative individual
-exemplars include V1298~Tau, an adolescent resonant chain of four
-transiting puffy planets that is a likely precusor to the compact
+exemplars include V1298~Tau, a resonant chain of 
+puffy planets that is a likely precusor to the compact
 multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
@@
 \citep[e.g.][]{2008Sci...322.1348M,2012ApJ...756L..33Q,2024AJ....167..193T}.
-In this article we will consider planets whose existence is previously
-known from transits, and infer new stellar ages using rotation and
-lithium.
+In this article we will consider planets whose existence has been
+previously established using transits, and infer new stellar ages
+using rotation and lithium.
 
@@
 assumed planetary) ages are precise at the $\approx$10\% level.  A
-more recent development, facilitated by the second Gaia data release,
-is the idea that finding transiting planets is often easier than
-finding the host stellar association \citep[e.g.][]{Tofflemire_2021}.
+more recent development, facilitated by Gaia, is the idea that finding
+transiting planets is often easier than finding the host stellar
+association \citep[e.g.][]{Tofflemire_2021}.
 
@@
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
-than 1\,Gyr.  Imposing a precision requirement of say $t$$<$1\,Gyr at
+than 1\,Gyr.  Requiring say $t$$<$1\,Gyr at
 2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
@@
 
-This study aims to resolve two main questions.  First, how wrong is it
+This study aims to resolve two questions.  First, how wrong is it
 to assume a uniform age distribution for transiting planet host stars?
@@
 uniform is wrong by a factor of a few, and that stellar activity may
-be a less significant bias than the challenge of precisely determining
-ages for field stars.
+be a less significant bias in young transiting planet detection than
+the challenge of precisely determining ages for field stars.
 
@@
 calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
-Objects of Interest \citep[KOIs;][]{Petigura_2022}, isochrone ages are
+Objects of Interest \citep[KOIs;][]{Petigura_2022}, such ages are
 most precise for stars whose luminosities and temperatures separate
@@
 \citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
-Modern models yield ages precise to $\lesssim$30\% for FGK stars
+Empirical models now yield ages precise to $\lesssim$30\% for FGK stars
 between 1-4\,Gyr, and constraining age posteriors at earlier times
 \citep{Bouma_2023}.  Physics-based models
-\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} are beginning to
+\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} can 
 connect these empirical relations to the evolution of stellar winds,
@@
 
-In the context of Kepler, rotation-based ages have been reported for
-planet-hosting stars since the early data releases
-\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  Recent studies
-focused on rotation-based ages for Kepler have further explored
-incorporating information from stellar kinematics
-\citep{2021AJ....161..189L,2024AJ....167..159L}, and from light curve
-features such as starspot amplitudes \citep{2023ApJ...952..131M}.  Our
-analysis is motivated by a few factors that we believe constitute
-improvements, particularly for the ages of sub-gigayear stars.   These
-factors are as follows.
-\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
+Rotation-based ages have been reported for
+various subsets of Kepler stars since the early data releases
+\citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
+More recent work has further explored incorporating information from
+stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
+and from light curve features such as starspot amplitudes
+\citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
+factors that can yield improvements, particularly for the ages of
+sub-gigayear stars.   These factors are as follows.
+\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
   \item The Kepler Object of Interest catalog, and our vetting of
-    false positives within it, has reached maturity
+    false positives within it, has now reached maturity
     \citep[e.g.][]{Thompson_2018};
@@
     any one star's rotation period \citep{Bouma_2023}.
-    For Sun-like stars, this model is consistent with older
-    calibrations \citep[e.g.][]{Mamajek_2008}, though with an improved treatment
-    of uncertainty propagation.  For late G dwarfs and
-    K dwarfs, the calibration differs by factors of a few;
-  %
-  \item We can identify binary stars at greater fidelity today than
+    This represents an improvement in both accuracy and in realism
+    of uncertainty propagation relative to previous calibrations.
+  %
+  \item We can identify binary stars with greater fidelity than
     in the past, which can clarify otherwise problematic estimates of
@@
     \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
-    These clusters offer a sample of stars for which we can test the
+    These clusters offer a sample of stars with which we can test the
     reliability of single-star age-dating methods.
@@
 We describe our age-dating methods in
-Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them using
-stars in clusters in the Kepler field.  The population-level
-trends for both stars and planets are described in
+Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them in
+Section~\ref{sec:litagecomparison} using
+clusters in the Kepler field.  The population-level
+trends are described in
 Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
@@
 
-\begin{figure*}[!t]
+\begin{figure}[!t]
 	\begin{center}
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
+
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
     with previously reported rotation periods (dark gray points).  The
-    rotation periods were primarily measured by
+    rotation periods are primarily drawn from
     \citet{Santos_2019,Santos_2021}.  About half of these rotators are
@@
     \citet{Berger_2020a_catalog}.
+    MIST isochrones spanning 3--6\,Gyr in 1\,Gyr increments 
+    are shown in the lower panel \citep{Choi_2016} .
 	}
 	\label{fig:stellarprops}
-\end{figure*}
+\end{figure}
 
@@
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
 procedure that bootstrapped against the local scatter in the continuum.
+The resulting EWs are shown in Figure~\ref{fig:li_vs_teff}.
 
@@
 		\leavevmode
-		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
+		\includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
 	\end{center}
-	\vspace{-0.6cm}
+	\vspace{-0.5cm}
 	\caption{
@@
     for planet-hosting stars.} These measurements were made from
-    Keck/HIRES spectra collected between 2009 and 2024.  Lines are the ``mean''
+    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
     isochrones from the same models as in Figure~\ref{fig:models}.  The
-    intrinsic dispersion at $\gtrsim$1\,Gyr becomes much larger
-    than changes in the mean.  Some stars with lithium detections do
+    intrinsic dispersion becomes much larger
+    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
     not have detected rotation periods, and vice-versa.
@@
 parallaxes, and $V$-band extinction from the
-\citet{2018MNRAS.478..651G} reddening map.  The resulting stellar
+\citet{2019ApJ...887...93G} reddening map.  The resulting stellar
 parameters are available for \fracstarswithprotwithbtwenty\ of the
@@
 $R_\star$, [Fe/H]) against the spectroscopic parameters from
-\citet{Fulton_2018}.  The temperature scales did show a few-percent
+\citet{Fulton_2018}.  The temperature scales showed a few-percent
 systematic difference, with \citetalias{Fulton_2018} quoting higher
@@
 and lower temperatures for early F dwarfs.  Our age analysis,
-described below, assumes Gaussian uncertainties in effective
-temperatures; we by default adopt the maximum of the two-sided
-uncertainties reported by \citetalias{Berger_2020a_catalog} as our
-symmetric temperature uncertainty.  Overall, the impact of systematic
-uncertainties in the temperature scale on our age uncertainties is
-generally smaller than that of the intrinsic scatter in the
-calibration sequences of the open clusters.
+described below, adopts the maximum of the two-sided
+uncertainties reported by \citetalias{Berger_2020a_catalog} as a
+symmetric Gaussian temperature uncertainty.  Systematic
+uncertainties in the temperature scale generally influence 
+our age uncertainties at a
+smaller level than the statistical intrinsic scatter in the
+open cluster rotation sequences.
 
@@
 age posterior should therefore incorporate this intrinsic
-population-level scatter into the precision with which the age can be
-inferred.  Figure~\ref{fig:models} highlights this problem,
+population-level scatter into the precision of its reported
+ages.  Figure~\ref{fig:models} highlights this problem,
 particularly in regions where the 0.1-1\,Gyr stars overlap.
@@
 into a single binary number, $Q_{\rm star}$.  When and how the flags
-represented in this should be applied depends on the question being
+represented in this bitmask should be applied depends on the question being
 asked.  If the goal is to construct a false-positive free sample, they
@@
 %crossmatch
-of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3 (accessible
-online at \url{https://gaia-kepler.fun}).  The
+of the NEA \texttt{q1\_q17\_dr25\_stellar} catalog with Gaia DR3
+(available at \url{https://gaia-kepler.fun}).  The
 separation distribution of the Kepler-Gaia DR3 crossmatches is such
@@
 	\caption{
-		{\bf Literature age comparison.}
+		{\bf Literature ages vs.~benchmark open cluster ages.}
 		Each point denotes a reported age,
@@
 		is a candidate member.
-		We used candidate cluster members as reported by previous studies based on spatial and kinematic
+		We used candidate cluster members from previous studies that used spatial and kinematic
 		information.  Although some field interlopers may be present in the 
@@
 		The number of stars in any given cluster follows by crossmatching every study against
-		this fixed cluster list; certain methods report ages for more stars than others.
+		a fixed cluster list; certain methods report ages for more stars than others.
 		Horizontal scatter is artificially added to visually
@@
 Gaia proper motions and distances 
-enabled the discovery of stellar ensembles that are more diffuse,
+have enabled the discovery of stellar ensembles that are more diffuse,
 but which nonetheless
@@
 
-In Figure~\ref{fig:agescalecompone}, we compare 
-ages from the literature against the ages of stars in the richest
+In Figure~\ref{fig:agescalecompone}, we compare our rotation-based
+ages, as well as select ages from the literature, against the ages of stars in the richest
 clusters in the Kepler field.
-In particular, we drew ages for Kepler stars from \citet{Berger_2020a_catalog},
+From previous studies, we show ages from \citet{Berger_2020a_catalog},
 \citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
@@
 The situation is particularly severe for the 40\,Myr stars in Cep-Her,
-but it extends out to 150--300\,Myr stars in Melange-3 and Theia~520.
+but it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
 For isochrone and kinematic ages,
@@
 
-\subsection{The Demographic Cliff}
-
-The stellar ages and planetary properties of the systems discussed in
-the previous sections are given in Tables~\ref{tab:planets}
-and~\ref{tab:stars}.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
+\subsection{A Dearth Of Young Stars And Planets}
+
+The stellar ages 
+are given in Table~\ref{tab:planets} for the known planet-hosts,
+and in Table~\ref{tab:stars} for all Kepler stars.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
 age distributions of all the stars (left column) and the known
@@
 cluster age out to which our interpolation-based method is calibrated.
-We refer to the overall slope as a ``demographic cliff'' in reference
-to an impending human phenomenon forecasted to happen in coming years.
+The overall slope yields a dearth of young stars relative to the expectation
+of a uniform age distribution.
+%In an alternative coinage, this tendency for fewer young
+%stars to be present could also be dubbed a ``demographic cliff'',
+%in reference to the phenomenon from human demography.
 
@@
 else unresolved binaries, and are automatically flagged as such.  The
-two exceptions are Kepler-786 (KOI-1199), and Kepler-1 (TrES-2).
-Kepler-786 is an early K dwarf with $t_{\rm
+most interesting exception is Kepler-786,
+an early K dwarf with $t_{\rm
 Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
@@
 age indicators in the spectrum similarly suggest that the star may not
-in fact be young.  TrES-2 is therefore a ``lithium-young'' system for
-which we believe the high lithium EW may be a statistical fluke.
+in fact be young.
 
@@
 
-(author contribution statements will go here)
-L.G.B.~conceived the project, collected data, 
-inferred ages, and wrote the manuscript.
-E.K.P.~contributed to early iterations of the rotation-based age analysis.
-L.A.H.~contributed to project design.
-H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
-the HIRES data.
-All authors assisted in manuscript revision.
+%(author contribution statements will go here)
+%L.G.B.~conceived the project, collected data, 
+%inferred ages, and wrote the manuscript.
+%E.K.P.~contributed to early iterations of the rotation-based age analysis.
+%L.A.H.~contributed to project design.
+%H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
+%the HIRES data.
+%All authors assisted in manuscript revision.
 

### 2024-05-21 paper/ms.tex
update ms

@@
 sub-gigayear stars.   These factors are as follows.
+
 \begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
@@
     false positives within it, has now reached maturity
-    \citep[e.g.][]{Thompson_2018};
+    \citep[e.g.][]{Thompson_2018}.
   %
@@
     converges by the $\approx$700\,Myr age of Praesepe
-    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
+    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
   %
@@
 		\subfloat{
-			\includegraphics[width=0.98\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
+			\includegraphics[width=0.99\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
 		}
@@
     \subfloat{
-    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
     }
@@
     \subfloat{
-    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
     }
   \end{center}
-  \vspace{-0.6cm}
+  \vspace{-0.66cm}
   \caption{{\bf Kepler's demographic cliff,} visible in the
@@
   temperatures of 3800--6200\,K for which we calculated rotation-based
-  ages.  The darker lines impose quality cuts on factors such as
+  ages.  The opaque lines impose quality cuts on factors such as
   binarity, crowding, and the star's evolutionary state.  The bottom
   row selects only stars with temperatures of 4400-5400\,K, which have
-  relatively precise ages near 2\,Gyr due to their fast spin-down
-  \citep[see][]{Bouma_2023}.  The statistical uncertainty for a mean
+  relatively precise ages near 2\,Gyr due to their fast spin-down.
+  The statistical uncertainty for a mean
   individual star at 1, 2, and 3\,Gyr is shown, and is identical
   across each row.  Finally, the right panel compares the
-  rotation-based age distributions against star formation histories
-  derived using CMD fitting.  Completeness is near unity until
+  rotation-derived ages against star formation histories
+  derived using CMD fitting.  Completeness in Kepler's
+  $P$$_{\rm rot}$ detection sensitivity is near unity until
   $t$$\lesssim$3\,Gyr \citep{2022ApJ...937...94M}.
@@
 
-{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex currently has its
-most constraining age measurement from relative isochrone dating of
+{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex has been
+dated using the isochrone loci of its
 pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
-Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022a,Bouma_2022b}.
+Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
 Different sub-groups of the complex may vary in age by $\approx$50\%
@@
 rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the cluster age reported by \citet{Bouma_2022b} for the RSG-5
+above the cluster age reported in \citet{Bouma_2022b} for the RSG-5
 subgroup.
@@
 Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
-this association.  Our rotation-based
-age-dating independently recovers these systems, with $t_{\rm
+this association.  We find rotation-based ages
+of $t_{\rm
 gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
@@
 nearby stars.  The core of this structure is also known as UBC-1
-\citep{2018A&A...618A..59C}; since both Kepler planet hosts are in
-diffuse regions of the cluster, we use the name from
-\citeauthor{2019AJ....158..122K}.  Based on isochrone-fitting,
+\citep{2018A&A...618A..59C}.  Based on isochrone-fitting,
 Theia~520 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
@@
 asteroseismology, gyrochronology, and isochrone fitting found
-consistent results.  Ongoing work by J.~Curtis et al.~broadly finds
+consistent results.  Ongoing work by \citet{Curtis2024} broadly finds
 that the two planet-hosting stars are indeed members of the diffuse
@@
 this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
-currently known from the primary Kepler mission.
+currently known from the main Kepler mission.
 
@@
 rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
-posterior because the spindown rates of stars this hot slow down after
-the first gigayear.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
+posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
 rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
@@
 
-We compared the independent rotation and lithium-based ages of the
+We compared the independent rotation and lithium ages of the
 known planet hosts using the point-estimates of the available
@@
 else unresolved binaries, and are automatically flagged as such.  The
-most interesting exception is Kepler-786,
+more interesting of the two is Kepler-786,
 an early K dwarf with $t_{\rm
@@
 using archival and new data.  We derived new ages using empirical
-methods that have been calibrated against an extensive set of open
-cluster data.
-
-Our analysis blindly recovered all 14 known Kepler planets in
-clusters, with ages in agreement with the more precise cluster-based
-ages.  Although lithium provided only marginal added information for
-most of the sample, the independent lithium and rotation ages agreed
-for $\gtrsim$90\% of cases.  In particular, we reported {\bf XX}
+interpolation-based methods, and assessed the reliability of these ages
+by comparing them against benchmark open clusters
+(Figure~\ref{fig:agescalecompone}).
+
+Our analysis blindly recovered the youth of all 14 known Kepler planets in
+clusters, with ages that agreed with the more precise ensemble ages.
+While lithium provided only marginal added information for
+most of the sample, in cases for which comparison was possible, the independent lithium and rotation ages agreed
+$\gtrsim$90\% of the time.  In particular, we reported {\bf XX}
 confirmed $<$500\,Myr planets with strong age constraints from both

### 2024-05-21 gyrojo/plotting.py
update plots

@@ -442,12 +442,15 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
         if not show_dispersion and not show_dispersionpoints:
             ages = [100, 500, 2000, 4000]
             agestrs = ['100 Myr', '500 Myr', '2 Gyr', '4 Gyr']
+            colors = ['C0', 'C1', 'purple', 'lime']
         elif show_dispersionpoints:
             ages = [100, 500, 2000]
             agestrs = ['100 Myr', '500 Myr', '2 Gyr']
+            colors = ['C0', 'C1', 'purple', 'whatever']
         else:
             ages = [100, 500, 2000, 4000]
             agestrs = ['100 Myr', '500 Myr', '2 Gyr', '4 Gyr']
+            colors = ['C0', 'C1', 'purple', 'lime']
         lAges = [np.log10(t)+6 for t in ages]  # log age in years
 
         li_models = []
@@ -467,7 +470,6 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
         agestrs = [f"{a}" for a in ages]
 
     linestyles = ['solid', 'dotted', 'dashed', 'dashdot', 'solid']
-    colors = ['C0', 'C1', 'purple', 'whatever']
     yvals = [110, 68.5, 38, 999]
     for ix, (li_model, ls, age, _c, agestr, yval) in enumerate(
         zip(li_models, linestyles, ages, colors, agestrs, yvals)
@@ -477,7 +479,7 @@ def plot_li_vs_teff(outdir, sampleid=None, yscale=None,
         if not show_dispersion and not show_dispersionpoints:
             alpha = 0.8
         ax.plot(
-            Teff_model, li_model, color=f'C{ix}', linewidth=1, zorder=10,
+            Teff_model, li_model, color=_c, linewidth=1, zorder=10,
             alpha=alpha, ls=ls, label=agestr
         )
 

### 2024-05-22 paper/ms.tex
update ms

@@
   %
-  We find that our rotation-based ages accurately recover ages of
+  We find that our rotational ages accurately recover the ages of
   stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
@@
   %
-  The results include \nplyounggyrotwosigma\ planets
+  The resulting yield includes \nplyounggyrotwosigma\ planets
   younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
@@
   This is far fewer than expected under the classic assumption of a uniform
-  star formation history: the star formation rate in the Kepler field
-  has dropped by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3\,Gyr.  This trend holds for both planet hosts and for the overall
+  star formation history.  We find that this dearth can be attributed to the star formation
+  rate in the Kepler field
+  dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
+  3\,Gyr.  This trend is observed both for known planet hosts as well as for the overall
   stellar sample.
@@
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.475\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
 
+		\vspace{-0.3cm}
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.475\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 	\vspace{-0.5cm}
-	\caption{
-    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
-    with previously reported rotation periods (dark gray points).  The
+	\caption{{\bf The stars.}  Our analysis focuses on stars observed by Kepler
+    with previously reported rotation periods (gray points).  The
     rotation periods are primarily drawn from
     \citet{Santos_2019,Santos_2021}.  About half of these rotators are
-    suitable for gyrochronology (black points), based on factors
-    including their temperatures, surface gravities, non-binarity, and
-    proximity to the main sequence (see Section~\ref{subsec:flags}).
-    Some are ``confirmed'' or ``candidate'' Kepler Objects of Interest
-    that meet an additional set of planetary quality criteria (blue
+    suitable for gyrochronology (dark gray points), based on factors
+    including their non-binarity and
+    proximity to the spectroscopic main sequence (orange boundary; see Section~\ref{subsec:flags}).
+    Some are ``confirmed'' or ``candidate'' KOIs
+    that meet an additional set of planetary quality criteria (yellow
     points; Section~\ref{subsec:plflags}).  Surface gravities and
@@
     \citet{Berger_2020a_catalog}.
-    MIST isochrones spanning 3--6\,Gyr in 1\,Gyr increments 
-    are shown in the lower panel \citep{Choi_2016} .
+    Isochrones 
+    in the lower panel are from MIST \citep{Choi_2016} .
 	}
@@
 		\leavevmode
-		\includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
+		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
 	\end{center}
-	\vspace{-0.5cm}
-	\caption{
-    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
+	\vspace{-0.25cm}
+	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
     for planet-hosting stars.} These measurements were made from
@@
 We therefore visually selected KOIs that appeared
-to be near the main sequence -- i.e.~ most of the blue
+to be near the main sequence -- i.e.~ most of the yellow
 points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
 polynomial $f_1$ through a similar BIC-minimization procedure,
-and found $N$=3.
+which yielded $N$=3.
+This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
 The portion of the orange locus from $\approx$5300--6200\,K
 is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
-$\approx$0.1\,dex above the ``KOI main sequence''.
+0.1\,dex above the ``KOI main sequence''.
 The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
-We emphasize that for stars
-surface gravities above the orange locus, we do report rotation based ages;
+We emphasize that we do report rotation-based ages for stars
+above the orange locus;
 they are simply flagged as not being near the main sequence.  }
@@
 exoplanet demographics also depend on stellar metallicity and mass
-\citep[e.g.][]{Petigura_2018}.
+\citep[e.g.][]{Petigura_2018,Miyazaki2023}.
 

### 2024-05-22 paper/ms.tex
add KIC identifer to Table 1 MRT

@@
   gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
-  not be raised, and for a planet to be ``reliabe'', we suggested
+  not be raised, and for a planet to be ``reliable'', we suggested
   $Q_{\rm planet}$ to be zero.
+  In the comment string, a single \checkmark\ denotes systems with no
+  reliability issues for which $t_{\rm gyro}$$<$1\,Gyr but $t_{\rm
+  Li}$ is single-sided; a double \checkmark\checkmark\ denotes
+  analogous systems with two-sided rotation and lithium ages in
+  agreement.
   }

### 2024-05-22 drivers/make_younghighlight_table.py
ditto

@@ -1,4 +1,5 @@
 import os
+from numpy import array as nparr
 from os.path import join
 from copy import deepcopy
 import numpy as np, pandas as pd, matplotlib.pyplot as plt
@@ -348,6 +349,7 @@ def make_table(
             print(f"{s}: {frac_consistent}% are consistent")
             print(10*'-')
 
+    kepids = nparr(pdf['kepid'])
     pdf = pdf.drop(columns=['kepid'])
     mapdict.pop('kepid')
 
@@ -436,6 +438,11 @@ def make_table(
         ulkvp('nnonfopkoissomeageinfo', len(pdf))
 
         pdf = pdf.rename(invdict, axis='columns')
+
+        pdf['kepid'] = kepids
+        selcols = [c for c in pdf.columns]
+        selcols = [selcols[-1]] + selcols[0:-1]
+        pdf = pdf[selcols]
 
         csvpath = join(PAPERDIR, 'table_allageinfo.csv')
         pdf.to_csv(csvpath, index=False)

### 2024-05-22 paper/ms.tex
table clarity

--- 2024-05-22 paper/ms.tex
@@
   \startdata
-  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}$ &
+  KOI & Kepler &  $T_{\rm eff}$ & $P_{\rm rot}$ & EW$_{\rm Li}^{\rm \ast}$ &
   $t_{\rm gyro}$ & $t_{\rm Li} $ & Consistent? & $R_{\rm p}$ & $P_{\rm orb}$ &
@@
   \enddata
-  \tablecomments{Two checkmarks denotes systems with two-sided $t_{\rm
-  gyro}$ and $t_{\rm Li}$ for which no important quality flags are
-  raised.  One checkmark denotes two-sided $t_{\rm gyro}$ only. 
+  \tablecomments{
+  	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
+  	Two checkmarks ($\checkmark\checkmark$) denote systems with two-sided $t_{\rm
+  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are expected to be reliable.  One checkmark ($\checkmark$) denotes two-sided $t_{\rm gyro}$ only. 
   ``Confirmed'' planets appear in the machine-readable
@@
   For a star to have a high likelihood of being ``reliable for
-  gyrochronology'' we suggest $Q_{\rm star}$ bits zero to nine to
-  not be raised, and for a planet to be ``reliabe'', we suggested
+  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--9
+  not raised, and for a planet to be ``reliable'', we suggest
   $Q_{\rm planet}$ to be zero.

### 2024-05-22 drivers/compare_literature_ages_to_cluster_ages.py
add reinhold15 plot

@@ -146,6 +146,11 @@ l21_df = ascii.read(tabpath, format='mrt').to_pandas()
 tabpath = join(DATADIR, "literature", "Lu_2024_ajad28b9t2_mrt.txt")
 l24_df = ascii.read(tabpath, format='mrt').to_pandas()
 
+# Reinhold+2015
+catalogs = _v.get_catalogs("J/A+A/583/A65")
+r15_df = catalogs[1].to_pandas() # output parameters
+r15_df = r15_df[~r15_df['tMH08'].isna()]
+
 def _get_cluster_xm_lit_dfs(cluster_df):
     """
     Given "cluster dataframe" with some set of kepler id's, get the
@@ -216,7 +221,15 @@ def _get_cluster_xm_lit_dfs(cluster_df):
     )
     sl24_df = l24_df[sel].sort_values(by='Age')
 
-    return sb24df, sm23_df, sm23_df3, sb20_df, sl21_df, sl24_df
+    #
+    # Reinhold gyro-only
+    #
+    sel = (
+        (r15_df.KIC.isin(cluster_df.kepid))
+    )
+    sr15_df = r15_df[sel].sort_values(by='tMH08')
+
+    return sb24df, sm23_df, sm23_df3, sb20_df, sl21_df, sl24_df, sr15_df
 
 
 for cluster_df, cluster_full_name, true_age in zip(
@@ -225,7 +238,7 @@ for cluster_df, cluster_full_name, true_age in zip(
 
     cluster_short_name = cluster_full_name.split(" ")[0]
 
-    sb24df, sm23_df, sm23_df3, sb20_df, sl21_df, sl24_df = (
+    sb24df, sm23_df, sm23_df3, sb20_df, sl21_df, sl24_df, sr15_df = (
         _get_cluster_xm_lit_dfs(cluster_df)
     )
 
@@ -236,13 +249,13 @@ for cluster_df, cluster_full_name, true_age in zip(
     plt.close("all")
     set_style("clean")
 
-    fig, axs = plt.subplots(nrows=1, ncols=5, figsize=(0.8*7,0.8*1.8))
+    fig, axs = plt.subplots(nrows=1, ncols=6, figsize=(0.8*7,0.8*1.8))
     axs = axs.flatten()
 
     # STAREVOL from SM23 qualtiatively similar, but mildly worse.
     dfs = [
         sb24df, sm23_df, #sm23_df3,
[…]

### 2024-05-22 paper/ms.tex
update ms

@@
 		\subfloat{
-			\includegraphics[width=0.99\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
+			\includegraphics[width=0.98\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
 		}
@@
 		is a candidate member.
-		We used candidate cluster members from previous studies that used spatial and kinematic
-		information.  Although some field interlopers may be present in the 
+		Cluster membership was assessed in previous studies without knowledge of rotation.
+		Although some field interlopers may be present in the 
 		membership lists,
@@
 
-Stars that formed in the same birth cluster 
-are the gold standard for the astronomical age scale
-\citep{Soderblom_2010}.
-Before Gaia, the few $\gtrsim$1\,Gyr open clusters
-known in the Kepler field had been visually identified by
-\citet{1864RSPT..154....1H}.
-Gaia proper motions and distances 
-have enabled the discovery of stellar ensembles that are more diffuse,
-but which nonetheless
-share a common age 
-based on e.g.~isochrone, rotation, and lithium dating
+Stars that formed in the same birth cluster are the gold standard for
+the astronomical age scale \citep{Soderblom_2010}.  Before Gaia, the
+few $\gtrsim$1\,Gyr open clusters known in the Kepler field had been
+visually identified by \citet{1864RSPT..154....1H}.  Gaia proper
+motions and distances have enabled the discovery of stellar ensembles
+that are more diffuse, but which nonetheless share a common age based
+on e.g.~isochrone, rotation, and lithium dating
 \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
@@
 In Figure~\ref{fig:agescalecompone}, we compare our rotation-based
-ages, as well as select ages from the literature, against the ages of stars in the richest
-clusters in the Kepler field.
-From previous studies, we show ages from \citet{Berger_2020a_catalog},
-\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
-\citet{2024AJ....167..159L}.
-We followed any guidance available from these respective studies for
+ages, as well as select ages from the literature, against the ages of
+stars in the richest clusters in the Kepler field.  From the
+literature, we drew ages from \citet{Reinhold_2015},
+\citet{Berger_2020a_catalog}, \citet{2021AJ....161..189L},
+\citet{2023ApJ...952..131M}, and \citet{2024AJ....167..159L}.  We
+followed any guidance available from these respective studies for
 removing ages that were explicitly flagged as unreliable, and plotted
-stars within $1\sigma$ of zero as being upper limits.
-
-Cluster memberships are a delicate subject.
-Figure~\ref{fig:agescalecompone} is showing reported ages for
-a set of spatially and kinematically selected set of stars that could be cluster members, or
-they could be field interlopers.  For NGC6811 and NGC6819, we
-adopted candidate members from
+stars within $1\sigma$ of zero as being upper limits.  For the
+\citeauthor{Reinhold_2015} ages, we used those that were calculated
+using the \citet{Mamajek_2008} calibration; for the
+\citeauthor{2023ApJ...952..131M} ages, we show those from
+\texttt{kiauhoku} \citep{Claytor2020} rather than \texttt{STAREVOL}
+\citep{Amard2019}, based on the defaults adopted by
+\citeauthor{2023ApJ...952..131M}.
+
+Cluster membership is a delicate subject.
+Figure~\ref{fig:agescalecompone} is showing reported ages for a set of
+spatially and kinematically selected set of stars that could be
+cluster members, or they could be field interlopers.  For NGC6811 and
+NGC6819, we adopted candidate members from
 \citet{2018A&A...618A..93C,CantatGaudin_2020} and
-\citet{Kounkel_2020}. 
-For Melange-3 we used the candidates reported by
-\citet{Barber_2022} and required ``offset'' tangential
-velocities below 2\,\kms.  For Cep-Her we
-used candidate members from Table~2 of \citet{Bouma_2022b}, requiring
-``weights'' exceeding 0.1.
-Even with contaminants, we can compare
-the relative age distributions derived by the different studies because in all
-instances we are comparing reported ages against a fixed list of stars.
-
-There are two main metrics for success in this test: accuracy
-in whether the reported ages agree with the cluster age, and precision in
-whether the reported age uncertainties agree with their
-dispersion around the cluster age.
+\citet{Kounkel_2020}.  For Melange-3 we used the candidates reported
+by \citet{Barber_2022} and required ``offset'' tangential velocities
+below 2\,\kms.  For Cep-Her we used candidate members from Table~2 of
+\citet{Bouma_2022b}, requiring ``weights'' exceeding 0.1.  Even with
+contaminants, we can compare the relative age distributions derived by
+the different studies because in all instances we are comparing
+reported ages against a fixed list of stars.
+
+There are two main metrics for success in this test: accuracy in
+whether the reported ages agree with the cluster age, and precision in
+whether the reported age uncertainties agree with their dispersion
+around the cluster age.
 
@@
 While previous studies have been able to derive accurate ages for
-stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have historically been
-overestimated by anywhere from 0.5--2\,dex.
-The situation is particularly severe for the 40\,Myr stars in Cep-Her,
-but it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
-For isochrone and kinematic ages,
-this is not surprising because these methods use stellar parameters that do
-not appreciably change at $t$$\lesssim$1\,Gyr.
-For the rotation-based ages quoted by
-\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the discrepancy
-between their ages and the cluster age scale is likely caused by details of the calibration 
-methodologies.
-The physical spin-down models used by
-\citet{2023ApJ...952..131M},
-and the gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
-may not reproduce
-the rotation sequences of open clusters such as Praesepe, the Hyades, and the Pleiades.
-The statistical age uncertainties yielded by these approaches
-similarly may not marginalize over the intrinsic
-scatter in rotation periods present in stars of a fixed mass at any given age.
-
-%The rotation-based ages that we have calculated have a precision
-%limited by the intrinsic scatter in the cluster rotation sequences.
-%Within the age range accessible through the clusters, they seem to be
-%mostly accurate,
-%with the one exception being that our ages for stars in Cep-Her 
-%have a typical median of $\approx$80\,Myr, rather than 40\,Myr.
-%The reason is that 80\,Myr is the lower bound of 
-%the \texttt{gyro-interp}  age grid; pre-main-sequence
-%spin-up is outside of our calibration regime.
+stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have
+historically been overestimated by anywhere from 0.3--2\,dex.  The
+situation is particularly severe for the 40\,Myr stars in Cep-Her, but
+it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
+For isochrone and kinematic ages, this is not surprising because these
+methods rely on parameters that do not appreciably change at
+$t$$\lesssim$1\,Gyr.  For the rotation-based ages quoted by
+\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
+discrepancy is likely
+caused by details of the calibration methodologies.  The physical
+spin-down models used by \citet{2023ApJ...952..131M}, and the
+gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
+may not reproduce the rotation sequences of open clusters such as
+Praesepe, the Hyades, and the Pleiades.  The \citet{Mamajek_2008}
+calibration used by \citet{Reinhold_2015} does reproduce such
+sequences for Sun-like stars.  However the
+uncertainties determined using this approach are underestimates, because
+they do not marginalize over
+the intrinsic scatter in rotation periods present for stars of a fixed
+mass and age.
 
@@
     \subfloat{
-    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
     }
@@
     \subfloat{
-    \includegraphics[width=.99\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
     }
@@
 
-\software{
-  astropy \citep{Astropy18},
-  matplotlib \citep{matplotlib},
-  numpy \citep{numpy},
-  scipy \citep{scipy},
-}
+%\software{
+%  astropy \citep{Astropy18},
+%  matplotlib \citep{matplotlib},
+%  numpy \citep{numpy},
+%  scipy \citep{scipy},
+%}
 

### 2024-05-23 paper/ms.tex
update ms

@@
 	\vspace{-0.5cm}
-	\caption{{\bf The stars.}  Our analysis focuses on stars observed by Kepler
-    with previously reported rotation periods (gray points).  The
+  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
+  Kepler with previously reported rotation periods (gray points).  The
   rotation periods are primarily drawn from
@@
   suitable for gyrochronology (dark gray points), based on factors
-    including their non-binarity and
-    proximity to the spectroscopic main sequence (orange boundary; see Section~\ref{subsec:flags}).
-    Some are ``confirmed'' or ``candidate'' KOIs
-    that meet an additional set of planetary quality criteria (yellow
-    points; Section~\ref{subsec:plflags}).  Surface gravities and
-    effective temperatures were derived photometrically by
-    \citet{Berger_2020a_catalog}.
-    Isochrones 
-    in the lower panel are from MIST \citep{Choi_2016} .
+  including their non-binarity and proximity to the spectroscopic main
+  sequence (orange boundary; see Section~\ref{subsec:flags}).  Some
+  are ``confirmed'' or ``candidate'' KOIs that meet an additional set
+  of planetary quality criteria (yellow points;
+  Section~\ref{subsec:plflags}).  Surface gravities and effective
+  temperatures were derived photometrically by
+  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
+  from MIST \citep{Choi_2016}.
 	}
@@
 This work is focused on Kepler stars for which ages can be inferred
-using either rotation, lithium, or both.  Such stars
-are a minority of the \nkeplerstars\ Kepler targets.  Rotation periods
-have been reported for roughly one in three Kepler targets
+using either rotation, lithium, or both.  Such stars are a minority of
+the \nkeplerstars\ Kepler targets.  Rotation periods have been
+reported for roughly one in three Kepler targets
 \citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
@@
 describe the set of stars that we adopt with measured rotation periods
-(Section~\ref{subsec:rotsel}), the set of objects we adopt as hosting planets
-(Section~\ref{subsec:planetsel}), and the subset of these with
+(Section~\ref{subsec:rotsel}), the set of objects we adopt as hosting
+planets (Section~\ref{subsec:planetsel}), and the subset of these with
 high-resolution spectra suitable for lithium analysis
@@
     Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
-    isochrones from the same models as in Figure~\ref{fig:models}.  The
-    intrinsic dispersion becomes much larger
-    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
+    isochrones from \citet{Jeffries_2023}.  The
+    intrinsic dispersion around these isochrones becomes much larger
+    than changes in the mean at $\gtrsim$1\,Gyr (see
+    Figure~\ref{fig:models}).  Some stars with lithium detections do
     not have detected rotation periods, and vice-versa.
@@
 g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
-the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
-	the method was as follows.
-The orange line in Figure~\ref{fig:stellarprops} is	defined as
-the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
-We defined $f_0$ by fitting an
-$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
-B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
-and $T_{\rm eff}\in[3700,5900]$\,K, where
-$t_{\rm B20,iso}$ was the isochronal age reported by
-\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
-$N$=1 to 10, minimized the Bayesian Information Criterion,
-and found $N$=6.  
-By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
-However for F and G dwarfs, the resulting locus
-allowed for stars that were `too evolved''; the density of KOIs 
-was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
-We therefore visually selected KOIs that appeared
-to be near the main sequence -- i.e.~ most of the yellow
-points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
-polynomial $f_1$ through a similar BIC-minimization procedure,
-which yielded $N$=3.
-This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
-The portion of the orange locus from $\approx$5300--6200\,K
-is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
-0.1\,dex above the ``KOI main sequence''.
-The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
-We emphasize that we do report rotation-based ages for stars
-above the orange locus;
-they are simply flagged as not being near the main sequence.  }
+the exact details of how this locus is constructed are arbitrary (see Appendix~\ref{app:linemethod}),
 the general aim is to flag stars for which there is evidence based on
@@
 sequences for Sun-like stars.  However the
-uncertainties determined using this approach are underestimates, because
+uncertainties determined using that method are underestimates because
 they do not marginalize over
@@
 
-Let us restrict our attention to blemish-free stars, with blemish-free
-``confirmed'' planets: $Q_{\rm star}$ must not have any of bits zero
-through nine raised, and $Q_{\rm planet}$ must have no quality flags
-raised at all.  Under this constraint, our results include
+Let us restrict discussion to blemish-free stars with 
+confirmed planets: $Q_{\rm star}$ must not have any of bits zero
+through nine raised, and $Q_{\rm planet}$ must also have no quality flags
+raised.  Under this constraint, our results include
 \ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
@@
 \nminineptuneshighqconfirmed\ mini-Neptunes above the radius valley
-(as parametrized by \citealt{2018MNRAS.479.4786V}), and
+as parametrized by \citealt{2018MNRAS.479.4786V}, and
 \nsuperearthshighqconfirmed\ super-Earth sized planets below the
@@
 
-A few new highlights from the high quality stars and planets include
+A few highlights from these systems include
 Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
@@
 super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
-near-Solar analog with two transiting planets: an Earth-sized planet
+near-Solar analog with an Earth-sized planet
 on a one-day orbit, and a mini-Neptune on a five-day orbit.  The
-Earth-sized planet, Kepler-1312c, along with the near-identical
-Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
+Earth-sized planet, Kepler-1312~c, along with the near-identical
+Kepler-1561~b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
 youngest Earth-sized planets currently known, comparable to systems
@@
 tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
-and those with 60$<$$v_{\rm T}$$<$150\,\kms as thick disk members.
+and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
 For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
@@
 
-% compare against Berger+20, 
-% note their sentence
+% compare against Berger+20, note their sentence
 % "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al.
@@
 time-dependent carving of the photoevaporation desert
-\citep{Owen_Lai_2018}, and the time evolution of the radius valley
+\citep{Owen2018}, and the time evolution of the radius valley
 \citep{Rogers_2021} can be explored using our data.  We defer this
@@
 
-We began this article with two questions: how wrong is the assumption
+We began this study with two questions: how wrong is the assumption
 of a uniform age distribution for stars in the galactic thin disk?
@@
 
-Our analysis blindly recovered the youth of all 14 known Kepler planets in
-clusters, with ages that agreed with the more precise ensemble ages.
-While lithium provided only marginal added information for
-most of the sample, in cases for which comparison was possible, the independent lithium and rotation ages agreed
-$\gtrsim$90\% of the time.  In particular, we reported {\bf XX}
-confirmed $<$500\,Myr planets with strong age constraints from both
-rotation and lithium, including {\bf YY} mini-Neptunes, {\bf ZZ}
-super-Earths, {\bf AA} Earths and sub-Earths, and a Neptune-sized
-planet.
+Our analysis blindly recovered the ages of all 14 known Kepler planets
+in clusters.  While lithium provided only marginal added information
+for most of the sample, the lithium and rotation ages agreed in
+$\gtrsim$90\% of cases for which comparison was possible.
+Our resulting ages included two-sided $t_{\rm gyro}$ and $t_{\rm Li}$
+for \ltonegyrhighqconfirmedtwosided\ sub-gigayear confirmed planets, and
+two-sided $t_{\rm gyro}$ with one-sided $t_{\rm Li}$ for
+\ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
+systems with RUWE$>$1.4 and including ``candidate'' planets
+would increase the counts to
+\ltonegyrmediumqconfirmedtwosided\ and
+\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
+such as on transit geometry, would further increase the counts.  The sizes and
+periods of the most restricted set of planets are shown in
+Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
+mostly mini-Neptunes, some are near the lower boundary of the
+``sub-Jovian desert ''\citep{Owen2018}, and others include Earth-sized 
+planets a few hundred million years old.
 
@@
     three billion years ago.
-    This star formation history broadly agrees with those derived
+    This result from rotation-based ages broadly agrees with recent reports of
+    a declining star-formation rate
     from CMD fitting and white-dwarf chronology, though with
@@
 \appendix
+\section{Defining An ``Isochronally Too Far From The Main Sequence'' Boundary}
+The (arbitrary) method used to construct the orange line in 
+the lower panel of Figure~\ref{fig:stellarprops} was as follows.
+The line is	defined as
+the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
+We defined $f_0$ by fitting an
+$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
+	B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
+and $T_{\rm eff}\in[3700,5900]$\,K, where
+$t_{\rm B20,iso}$ was the isochronal age reported by
+\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
+$N$=1 to 10, minimized the Bayesian Information Criterion,
+and found $N$=6.  
+By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
+However for F and G dwarfs, the resulting locus
+allowed for stars that were ``too evolved''; the density of KOIs 
+was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
+We therefore visually selected KOIs that appeared
+to be near the main sequence -- i.e.~ most of the yellow
+points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
+polynomial $f_1$ through a similar BIC-minimization procedure,
+which yielded $N$=3.
+This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
+The portion of the orange locus from $\approx$5300--6200\,K
+is set as $f_1 + c$, for $c$ a constant offset that we defined to be
+0.1\,dex above the ``KOI main sequence''.
+The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
+We do report rotation-based ages for stars
+above this orange locus;
+they are simply flagged as not being near the main sequence.
+
 \section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
@@
 
-\paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} has
-a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot
-Jupiter, and has been studied in detail by multiple investigators
+\paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} is
+a $\approx$2.5\,day near-grazing hot Jupiter oribiting a G0 V primary
+which has been studied in detail by multiple investigators
 \citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
-Our lithium-based age of this system (\trestwotli\,Myr), derive from
-EW$_{\rm Li}$=81$\pm$3\,m\AA, is qualitatively in line with the
+Our lithium-based age of this system (\trestwotli\,Myr), derived from
+EW$_{\rm Li}^{\ast}$=81$\pm$3\,m\AA, qualitatively agrees with the
 previously noted lithium abundance \citep{2007ApJ...664.1190S}.
@@
 spectroscopic $v\sin i$ is low.  The lithium age is strongly
-asymmetric because there is significant amounts of scatter in EW$_{\rm
-Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G
-dwarfs.  We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of
-2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.  The non-detection
-of rotation, while mildly surprising, is not shocking if the system's
-true age is in this long tail.
+asymmetric because of the scatter in EW$_{\rm Li}$ vs.~$T_{\rm eff}$
+over ages $t$$\gtrsim$1\,Gyr for early G dwarfs.  We calculate a
+$2\sigma$ upper limit on $t_{\rm Li}$ of 2.8\,Gyr, and a $3\sigma$
+upper limit of 6.1\,Gyr.  The non-detection of rotation, while mildly
+surprising, is not shocking if the system's true age is in this long
+tail.
 
 \paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm
-Li}$=100$\pm$5\,m\AA, nominally implying $t_{\rm
+Li}^{\ast}$=100$\pm$5\,m\AA, nominally implying $t_{\rm
 Li}$$\approx$200-800\,Myr, but no rotation detection.  Our automated
@@
 eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
-$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm Li}$=83$\pm$6\,m\AA, the
-lithium age of \kepseveneightsix\,Myr would predict an obvious
-rotation signal.  None is present in the Kepler light curve,
-consistent with the low $v\sin i$.  Out of all ``apparently
-discrepant'' lithium and rotation measurements discussed in this
-appendix, this is the only one that seems to remain discrepant after
-scrutiny.  The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm
-HK}=-4.7\pm0.5$, which suggests an age at least as old as the Hyades
-\citep{Mamajek_2008}.  The Balmer lines are in absorption, and display
-no obvious signatures of youth.
+$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm
+Li}^{\ast}$=83$\pm$6\,m\AA, the lithium age of \kepseveneightsix\,Myr
+would predict an obvious rotation signal.  None is present in the
+Kepler light curve, consistent with the low $v\sin i$.  Out of all
+``apparently discrepant'' lithium and rotation measurements discussed
+in this appendix, this is the only one that seems to remain discrepant
+after scrutiny.  The  \ion{Ca}{2} doublet is in emission, with
+$R'_{\rm HK}=-4.7\pm0.5$, which suggests an age at least as old as the
+Hyades \citep{Mamajek_2008}.  The Balmer lines are in absorption, and
+display no obvious signatures of youth.
 
@@
 
-\section{TODO's}
-
-\begin{enumerate}
-%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
-%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
-%    Could compare against asteroseis too.
+%\section{TODO's}
+%
+%\begin{enumerate}
+%%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
+%%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
+%%    Could compare against asteroseis too.
+%%    %
+%%  \item Intro needs better discussion of Mathur2023 and Lu2021
 %    %
-%  \item Intro needs better discussion of Mathur2023 and Lu2021
-    %
-  \item Consider moving the bit list description to its own appendix.
-    %
-  \item Rerun gyro calculation with denser grid, fixing quantization
-    and extending upper limit to 5gyr.  (Will use updated Teff
-    uncertainties)
-\end{enumerate}
+%  \item Consider moving the bit list description to its own appendix.
+%    %
+%  \item Rerun gyro calculation with denser grid, fixing quantization
+%    and extending upper limit to 5gyr.  (Will use updated Teff
+%    uncertainties)
+%\end{enumerate}
 

### 2024-05-23 paper/ms.tex
all minor

@@
     \subfloat{
-    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
+    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf}
     }
@@
     \subfloat{
-    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
+    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf}
     }
@@
 
-%\software{
-%  astropy \citep{Astropy18},
-%  matplotlib \citep{matplotlib},
-%  numpy \citep{numpy},
-%  scipy \citep{scipy},
-%}
+\software{
+  astropy \citep{Astropy18},
+  matplotlib \citep{matplotlib},
+  numpy \citep{numpy},
+  scipy \citep{scipy},
+}
 
@@
 \section{Defining An ``Isochronally Too Far From The Main Sequence'' Boundary}
+\label{app:linemethod}
+
 The (arbitrary) method used to construct the orange line in 

### 2024-05-23 gyrojo/plotting.py

@@ -1577,7 +1577,7 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
     ##########################################
 
     for ax in axs:
-        add_gradient_patch(ax, 3.4, 4.1, 0, 0.1)
+        add_gradient_patch(ax, 3.1, 4.1, 0, 0.1)
 
     xmin = 0
     xmax = MAXAGE-20 if MAXAGE < 4000 else MAXAGE

### 2024-05-23 paper/collect.sh

@@ -19,8 +19,8 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
 
 # plot_hist_field_gyro_ages 
-cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf .
-cp $r/hist_field_gyro_ages_20240430/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf .
+cp $r/hist_field_gyro_ages_20240521/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf .
+cp $r/hist_field_gyro_ages_20240521/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf .
 
 # plot_li_vs_teff
 #cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .

### 2024-05-27 paper/ms.tex
update ms

@@
   rotation-derived ages of stars (left) and planet hosts (middle) in
-  the Kepler field.  The top row corresponds to all stars with
+  the Kepler field.  The top row shows all stars with
   temperatures of 3800--6200\,K for which we calculated rotation-based
-  ages.  The opaque lines impose quality cuts on factors such as
-  binarity, crowding, and the star's evolutionary state.  The bottom
-  row selects only stars with temperatures of 4400-5400\,K, which have
-  relatively precise ages near 2\,Gyr due to their fast spin-down.
+  ages.  Opaque lines impose quality cuts on 
+  binarity, crowding, and the star's evolutionary state; transparent lines do not.
+  The bottom
+  shows stars with temperatures of 4400--5400\,K, which have
+  more precise ages due to their fast spin-down.
   The statistical uncertainty for a mean
@@
   rotation-derived ages against star formation histories
-  derived using CMD fitting.  Completeness in Kepler's
+  derived using CMD fitting (dashed and dotted lines).  Completeness in Kepler's
   $P$$_{\rm rot}$ detection sensitivity is near unity until
@@
 
-The age distributions of Kepler stars and planet hosts are
-nearly indistinguishable.  A simple diagnostic of whether the two
+The age distributions of Kepler stars and planet hosts are quite
+similar.  A simple diagnostic of whether the two
 distributions are drawn from the same underlying distribution is the
 Kolmogorov-Smirnov test;  we calculated this statistic in a manner
-that accounts for the Poisson uncertainties by performing 100 random
-draws of 70\% of the stars from each sample.  The resulting median
-$p$-values were $\approx$0.01 for both rows of
-Figure~\ref{fig:hist_tgyro}.  This degree of similarity suggests that
+that accounts for the Poisson uncertainties by performing 1000 random
+draws of 70\% of the stars from each sample.  The resulting 
+range of $\log_{\rm 10}p$-values (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentile)
+spanned -3.9 to -1.2 for the K-dwarfs only,
+and -5.2 to -2.2 for all Kepler stars.  This degree of similarity suggests that
 the ``common wisdom'' that young transiting planets are hard to detect
@@
   astropy \citep{Astropy18},
+  claude \citep{claude2024anthropic},
   matplotlib \citep{matplotlib},

### 2024-05-27 gyrojo/plotting.py
tweak histogram

@@ -1590,7 +1590,7 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
         if ix == 0:
             ax.update({
             'ylabel': '$N$/$N_{\mathrm{max}}$\n'+f'{teffstr}',
-            'yticklabels': [0, 0.02, 0.04, 0.06],
+            'yticklabels': [0, 0.02, 0.04, 0.06, 0.08],
             })
         else:
             ax.update({
@@ -1709,7 +1709,7 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
     ###########################################
 
     from scipy import stats
-    sel_age = mdf.age < 3.e9
+    sel_age = mdf.age < 3e9
 
     star_ages = nparr(
         mdf.loc[sel_gyro_ok & sel_age, 'age']
@@ -1726,9 +1726,11 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
     )
 
     txt = (f'preciseagesonly={bool(preciseagesonly)}: '
-           f'med(p)={p_value_median:.1e}, '
-           f'<p value>={p_value_mean:.1e}, σ_pval={p_value_std:.1e}, '
-           f'2.5-97.5% CI: {p_value_ci[0]:.1e} - {p_value_ci[1]:.1e} 2-sample KS')
+           f'med(log10p)={np.log10(p_value_median):.1f}, '
+           f'<log10p value>={np.log10(p_value_mean):.1f}, '
+           f'log10(σ_pval)={np.log10(p_value_std):.1f}, '
+           f'2.5-97.5% CI (log10p): '
+           f'{np.log10(p_value_ci[0]):.1f} - {np.log10(p_value_ci[1]):.1f} 2-sample KS')
     print(42*'-')
     print(txt)
     print(42*'-')

### 2024-05-27 paper/ms.tex
update ms

@@
 are given in Table~\ref{tab:planets} for the known planet-hosts,
-and in Table~\ref{tab:stars} for all Kepler stars.  Figure~\ref{fig:hist_tgyro} shows the gyrochone
+and in Table~\ref{tab:stars} for all Kepler stars.  Figure~\ref{fig:hist_tgyro} shows the rotation-based
 age distributions of all the stars (left column) and the known
@@
 
-The age distributions of Kepler stars and planet hosts are quite
-similar.  A simple diagnostic of whether the two
+The age distributions of Kepler stars and KOI hosts in Figure~\ref{fig:hist_tgyro} appear
+similar.  One diagnostic for whether the two
 distributions are drawn from the same underlying distribution is the
@@
 draws of 70\% of the stars from each sample.  The resulting 
-range of $\log_{\rm 10}p$-values (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentile)
+range of $\log_{\rm 10}p$ (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentile)
 spanned -3.9 to -1.2 for the K-dwarfs only,
-and -5.2 to -2.2 for all Kepler stars.  This degree of similarity suggests that
+and -5.2 to -2.2 for all Kepler stars.  This degree of similarity suggests 
+that although weak differences may be present,
 the ``common wisdom'' that young transiting planets are hard to detect
@@
 nuance.  If true, this statement only holds for the very youngest
-stars, since at the age resolution of this study, there appears to be
-little to no selection bias against sub-gigayear planets.
+stars, where we see only a marginal underabundance in the KOI host age
+distribution relative to that of the overall stellar sample.
+At the age resolution of this study, 
+the selection bias against sub-gigayear planets seems small.
 

### 2024-05-27 drivers/make_star_table.py
star table float precision

@@ -148,7 +148,7 @@ def make_star_table(
         print(f"Wrote {texpath}")
 
     csvpath = join(PAPERDIR, 'table_star_gyro.csv')
-    _sdf.to_csv(csvpath, index=False)
+    _sdf.to_csv(csvpath, index=False, float_format='%.2f')
     print(f'Wrote {csvpath}')
 
     csvpath = join(PAPERDIR, 'table_star_gyro_allcols.csv')

### 2024-05-27 paper/ms.tex
update ms

@@
 
-Using our effective temperatures and rotation periods, we ran
-\texttt{gyro-interp} over an age grid of 0 to 4\,Gyr, linearly spaced
-over 500 grid points.  \texttt{gyro-interp} itself is calibrated out
-to 4\,Gyr, the age of M67
-\citep[see][]{Dungee_2022,Gruner_2023}.  To estimate the
+To estimate the
 stellar spin-down rate as a function of time and stellar temperature,
-the framework interpolates between reference open clusters using cubic
-Hermite polynomials.  We calculate the probability of the
+\texttt{gyro-interp} uses measured rotation periods and effective
+temperatures from reference open clusters, and interpolates between them using cubic
+Hermite polynomials.  We calculated the probability of the
 rotation-based age $t_{\rm gyro}$, given the observed rotation
-periods, effective temperatures, and their uncertainties by
-integrating Equation~1 of \citet{Bouma_2023} at the default grid
+periods and effective temperatures (and their uncertainties) by
+integrating Equation~1 of \citet{Bouma_2023} at the default
 resolution.  This procedure marginalizes over the intrinsic age and
@@
 sequences to yield statistical age uncertainties.
+We adopted a linear age grid spanning 0--5\,Gyr with
+500 grid points, and interpolated the resulting posteriors to determine
+point estimate summary statistics.
+At the old end, 
+\texttt{gyro-interp} is calibrated out
+to 4\,Gyr M67 cluster
+\citep[see][]{Dungee_2022,Gruner_2023}.  
+For Kepler stars with rotation periods above the 
+M67 sequence, our resulting age constraints are quoted as lower limits.
 
@@
 size-impact parameter degeneracy.  We flag planet candidates as
-potentially grazing if $b>0.8$, using the impact parameters reported
+potentially grazing if $b$$>$0.8, using the impact parameters reported
 in the Markov Chain Monte Carlo (MCMC) fits from
@@
 In Figure~\ref{fig:agescalecompone}, we compare our rotation-based
-ages, as well as select ages from the literature, against the ages of
-stars in the richest clusters in the Kepler field.  From the
+ages, and available ages from the literature, against the ages of
+stars in open clusters in the Kepler field.  From the
 literature, we drew ages from \citet{Reinhold_2015},
@@
 removing ages that were explicitly flagged as unreliable, and plotted
-stars within $1\sigma$ of zero as being upper limits.  For the
-\citeauthor{Reinhold_2015} ages, we used those that were calculated
+stars within $1\sigma$ of zero upper limits.  For the
+\citeauthor{Reinhold_2015} ages, we used those calculated
 using the \citet{Mamajek_2008} calibration; for the
@@
 \texttt{kiauhoku} \citep{Claytor2020} rather than \texttt{STAREVOL}
-\citep{Amard2019}, based on the defaults adopted by
-\citeauthor{2023ApJ...952..131M}.
-
-Cluster membership is a delicate subject.
+\citep{Amard2019}, since it showed better agreement with
+the cluster age scale.
+
+Cluster membership is a nuanced subject.
 Figure~\ref{fig:agescalecompone} is showing reported ages for a set of
-spatially and kinematically selected set of stars that could be
+spatially and kinematically selected stars that could be
 cluster members, or they could be field interlopers.  For NGC6811 and
@@
 below 2\,\kms.  For Cep-Her we used candidate members from Table~2 of
-\citet{Bouma_2022b}, requiring ``weights'' exceeding 0.1.  Even with
+\citet{Bouma_2022b} with ``weights'' exceeding 0.1.  Even with
 contaminants, we can compare the relative age distributions derived by
@@
 
-There are two main metrics for success in this test: accuracy in
-whether the reported ages agree with the cluster age, and precision in
-whether the reported age uncertainties agree with their dispersion
-around the cluster age.
+There are two main metrics for success in this test: {\it 1)}
+do the reported ages agree with the cluster age? And {\it 2)} do the
+reported age uncertainties agree with their dispersion
+around the cluster age?
 
@@
 While previous studies have been able to derive accurate ages for
-stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have
-historically been overestimated by anywhere from 0.3--2\,dex.  The
-situation is particularly severe for the 40\,Myr stars in Cep-Her, but
-it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
-For isochrone and kinematic ages, this is not surprising because these
+stars $\gtrsim$1\,Gyr old, available literature ages of sub-gigayear stars are
+usually overestimated by 0.3--2\,dex.  
+The ages of the youngest Kepler stars and planets were not previously well-measured.
+For isochrone and kinematic ages, this is because these
 methods rely on parameters that do not appreciably change at
-$t$$\lesssim$1\,Gyr.  For the rotation-based ages quoted by
+$t$$\lesssim$1\,Gyr.  For the rotation-based ages from
 \citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
-discrepancy is likely
-caused by details of the calibration methodologies.  The physical
-spin-down models used by \citet{2023ApJ...952..131M}, and the
-gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
-may not reproduce the rotation sequences of open clusters such as
-Praesepe, the Hyades, and the Pleiades.  The \citet{Mamajek_2008}
-calibration used by \citet{Reinhold_2015} does reproduce such
+discrepancy is
+caused by details of the calibration methodologies.  The
+spin-down models used in these studies
+may struggle to reproduce the observed rotation
+sequences seen in open clusters younger than a few hundred million years.  The \citet{Mamajek_2008}
+calibration used by \citet{Reinhold_2015} does fit such
 sequences for Sun-like stars.  However the
-uncertainties determined using that method are underestimates because
+uncertainties from all of these methods are underestimated because
 they do not marginalize over
-the intrinsic scatter in rotation periods present for stars of a fixed
-mass and age.
+the range of rotation periods that stars of a fixed
+mass and age might have.
 
@@
 Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
-detectable.  However no photometric rotation signal is detected; other
-age indicators in the spectrum similarly suggest that the star may not
-in fact be young.
+detectable.  However no photometric rotation signal is detected, and other
+age indicators in the spectrum similarly suggest that the star is not
+young.
 

### 2024-05-27 drivers/construct_field_star_gyro_quality_flags.py
adjust santos pipeline quality flag

@@ -195,6 +195,8 @@ def build_gyro_quality_flag(sample='gyro', datestr='20240430'):
             (df.Prot_provenance == 'Santos2019')
             |
             (df.Prot_provenance == 'Santos2021')
+            |
+            (df.Prot_provenance == 'SantosPrivComm')
         )
     elif sample == 'allKIC':
         df['flag_Prot_provenance'] = np.zeros(len(df)).astype(bool)

### 2024-05-27 paper/ms.tex
update ms

@@
 
-\section{Ages From Rotation}
-\label{sec:rotage}
-
 \begin{figure*}[!t]
@@
 
+\section{Ages From Rotation}
+\label{sec:rotage}
+
 We calculate gyrochrone ages using \texttt{gyro-interp}
@@
 We adopted a linear age grid spanning 0--5\,Gyr with
-500 grid points, and interpolated the resulting posteriors to determine
-point estimate summary statistics.
-At the old end, 
-\texttt{gyro-interp} is calibrated out
-to 4\,Gyr M67 cluster
-\citep[see][]{Dungee_2022,Gruner_2023}.  
+500 grid points, and interpolated the resulting posteriors to calculate
+any summary statistics.
+The oldest cluster for which
+\texttt{gyro-interp} is calibrated 
+is the 4\,Gyr M67 cluster
+\citep[][]{Dungee_2022,Gruner_2023}.  
 For Kepler stars with rotation periods above the 
@@
 We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
-stars with reported rotation periods and effective temperatures of
-3800--6200\,K.  To flag stars for which we suspect these ages may not
-in fact be valid, we built a set of quality flags, which we condensed
-into a single binary number, $Q_{\rm star}$.  When and how the flags
-represented in this bitmask should be applied depends on the question being
-asked.  If the goal is to construct a false-positive free sample, they
-could all be applied.  If the goal is to construct a complete sample,
+stars with reported rotation periods that had effective temperatures of
+3800--6200\,K.  To identify stars for which we suspect these ages may not
+be valid, we built a set of quality flags which we then condensed
+into a single binary number: $Q_{\rm star}$.  When and how
+this bitmask should be applied depends on the question being
+asked.  If the goal is to construct a false-positive free sample, all the
+quality flags could be applied.  If the goal is to construct a complete sample,
 then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
@@
 against such cases as a decision to the user.
+For our own analysis, 
+we assume that a star
+is suitable for gyrochronology if none of bits zero through nine
+(inclusive) are raised.  For analyses that require all
+stellar rotation periods to come from the same detection pipeline, we
+further require bit 11 to be zero.
 
@@
 flag is set for the KOIs with rotation periods drawn from a source
-other than \citet{Santos_2019,Santos_2021}.
-
-{\it How do we use the bitmasks?} By default, we assume that a star
-is suitable for gyrochronology if none of bits zero through nine
-(inclusive) are raised.  For analyses that require a homogeneous
-stellar rotation selection function, we also require bit 11 to be
-zero.
+other than the \citeauthor{Santos_2019} pipeline.
 
@@
 
+\section{Ages From Lithium}
+\label{sec:liage} 
 \subsection{Sample Summary}
@@
 
+Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
+the lithium 6708\AA\ absorption doublet, plotted over the mean
+isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
+declare an upper-limit for plotting purposes if the 1$\sigma$ lower
+limit on the equivalent width is below 10\,m\AA.  Our measured EWs
+span -40 to 250\,m\AA; all negative EWs have uncertainties that are
+statistically consistent with zero.
+
+To calculate lithium ages, we use \texttt{EAGLES} (git commit
+\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
+a purely empirical interpolation approach.  \texttt{EAGLES} 
+was calibrated on lithium measurements of stars observed by the
+Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
+to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
+age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
+uncertainties for purposes of interfacing with \texttt{EAGLES} by
+taking the maximum of the measured positive and negative 1$\sigma$
+uncertainty intervals.
+
+In part because many of our EWs are upper limits, many of the lithium
+ages are lower limits.  In such cases,  the inferred age posterior is
+strongly influenced by the assumed intrinsic dispersion in
+the lithium model, and by the prior.  In these instances, we
+quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
+
+\section{Cluster Age Comparison}
+\label{sec:litagecomparison}
+
 \begin{figure*}[!t]
@@
 		\subfloat{
-			\includegraphics[width=0.98\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
+			\includegraphics[width=0.96\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
 		}
@@
 
-\section{Ages From Lithium}
-\label{sec:liage}
-
-Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
-the lithium 6708\AA\ absorption doublet, plotted over the mean
-isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
-declare an upper-limit for plotting purposes if the 1$\sigma$ lower
-limit on the equivalent width is below 10\,m\AA.  Our measured EWs
-span -40 to 250\,m\AA; all negative EWs have uncertainties that are
-statistically consistent with zero.
-
-To calculate lithium ages, we use \texttt{EAGLES} (git commit
-\texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
-a purely empirical interpolation approach.  \texttt{EAGLES} 
-was calibrated on lithium measurements of stars observed by the
-Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
-to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
-age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
-uncertainties for purposes of interfacing with \texttt{EAGLES} by
-taking the maximum of the measured positive and negative 1$\sigma$
-uncertainty intervals.
-
-In part because many of our EWs are upper limits, many of the lithium
-ages are lower limits.  In such cases,  the inferred age posterior is
-strongly influenced by the assumed intrinsic dispersion in
-the lithium model, and by the prior.  In these instances, we
-quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
-
-\section{Cluster Age Comparison}
-\label{sec:litagecomparison}
-
 Stars that formed in the same birth cluster are the gold standard for
 the astronomical age scale \citep{Soderblom_2010}.  Before Gaia, the
-few $\gtrsim$1\,Gyr open clusters known in the Kepler field had been
+few open clusters known in the Kepler field had been
 visually identified by \citet{1864RSPT..154....1H}.  Gaia proper
@@
 \citet{2023ApJ...952..131M}, and \citet{2024AJ....167..159L}.  We
-followed any guidance available from these respective studies for
-removing ages that were explicitly flagged as unreliable, and plotted
+followed any guidance available from each study for
+removing ages that were flagged as unreliable, and plotted
 stars within $1\sigma$ of zero upper limits.  For the
@@
 \texttt{kiauhoku} \citep{Claytor2020} rather than \texttt{STAREVOL}
-\citep{Amard2019}, since it showed better agreement with
+\citep{Amard2019}, since the former showed better agreement with
 the cluster age scale.
@@
 
-The point of Figure~\ref{fig:agescalecompone} is hopefully clear.
-While previous studies have been able to derive accurate ages for
-stars $\gtrsim$1\,Gyr old, available literature ages of sub-gigayear stars are
-usually overestimated by 0.3--2\,dex.  
-The ages of the youngest Kepler stars and planets were not previously well-measured.
+The point of Figure~\ref{fig:agescalecompone} is hopefully clear:
+until this study, the ages of the youngest Kepler stars and planets were not 
+well-measured.
+While previous studies reported accurate ages for
+$\gtrsim$1\,Gyr stars, ages of sub-gigayear stars have historically been
+overestimated by 0.3--2\,dex, with severely understated uncertainties.  
 For isochrone and kinematic ages, this is because these
 methods rely on parameters that do not appreciably change at
-$t$$\lesssim$1\,Gyr.  For the rotation-based ages from
+$t$$\lesssim$1\,Gyr.  For the rotation-based ages in
 \citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
@@
 caused by details of the calibration methodologies.  The
-spin-down models used in these studies
-may struggle to reproduce the observed rotation
-sequences seen in open clusters younger than a few hundred million years.  The \citet{Mamajek_2008}
-calibration used by \citet{Reinhold_2015} does fit such
-sequences for Sun-like stars.  However the
-uncertainties from all of these methods are underestimated because
+\texttt{kiauhoku} spin-down model for instance has known
+``stretches'' and ``compressions'' in its age scale relative to observed open clusters
+\citep[see][Sec. 7.3]{2023ApJ...952..131M}.
+The \citet{Mamajek_2008}
+calibration used by \citet{Reinhold_2015} appears more accurate than the former
+two models, because it 
+was directly fit to reproduce the cluster sequences known at the time.  However the
+uncertainties from all of these methods appear to be underestimated.  This
+is likely because
 they do not marginalize over
@@
 Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
-Different sub-groups of the complex may vary in age by $\approx$50\%
-(\citealt{Bouma_2022b}; Kerr~et al.,~submitted).  The planets
+Different associations in the complex vary in age by $\approx$50\%
+(Kerr~et al.,~submitted).  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
@@
 rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the cluster age reported in \citet{Bouma_2022b} for the RSG-5
-subgroup.
+above the cluster age reported in \citet{Bouma_2022b} for RSG-5.
 
@@
 Most of the stars observed by Kepler are in the thin disk.  This can
-be verified by following \citet{Gaia_2018}, and denoting stars with 2D
+be verified by following \citet{Gaia_2018}, and labeling stars with 2D
 tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
@@
 Galaxy.  On extragalactic scales, the star formation rate in spiral
-galaxies with mass comparable to the Milky Way peaked $\approx$10\,Gyr
+galaxies with mass similar to the Milky Way peaked $\approx$10\,Gyr
 ago, and has since decreased by an order of magnitude
@@
 maxima in the thin disk's star formation history 5.7, 1.9, and
-1.0\,Gyr ago, which they associate with pericenter passages of the
+1.0\,Gyr ago, which they associated with pericenter passages of the
 Sagittarius satellite galaxy.
@@
 we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
-data in particular seem entirely consistent, after accounting for the
+data in particular are consistent, particularly after accounting for the
 statistical uncertainties reported in that study.  Regarding the
@@
 such as on transit geometry, would further increase the counts.  The sizes and
-periods of the most restricted set of planets are shown in
+periods for the most secure set of planets are shown in
 Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
 mostly mini-Neptunes, some are near the lower boundary of the
-``sub-Jovian desert ''\citep{Owen2018}, and others include Earth-sized 
+``sub-Jovian desert'' \citep{Owen2018}, and others include Earth-sized 
 planets a few hundred million years old.
 
-Our three main conclusions with respect to our original questions are
+Our main conclusions with respect to our original questions are
 as follows.
@@
     known Kepler planet-hosts show a demographic cliff.  There are
-    \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr) stars in the
+    twice as many ``old'' (2-3\,Gyr) stars in the
     Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation

### 2024-05-27 paper/ms.tex
update ms

@@
 few open clusters known in the Kepler field had been
-visually identified by \citet{1864RSPT..154....1H}.  Gaia proper
-motions and distances have enabled the discovery of stellar ensembles
+visually identified by \citet{1864RSPT..154....1H}.  Gaia
+has enabled the discovery of stellar ensembles
 that are more diffuse, but which nonetheless share a common age based
@@
 uncertainties from all of these methods appear to be underestimated.  This
-is likely because
+is probably because
 they do not marginalize over
-the range of rotation periods that stars of a fixed
-mass and age might have.
+the range of rotation periods accessible to stars of a fixed
+mass and age.
 
@@
 range of $\log_{\rm 10}p$ (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentile)
-spanned -3.9 to -1.2 for the K-dwarfs only,
+spanned -3.9 to -1.2 for the K dwarfs,
 and -5.2 to -2.2 for all Kepler stars.  This degree of similarity suggests 
@@
 the ``common wisdom'' that young transiting planets are hard to detect
-because their host stars are photometrically variable requires some
+due to the variability of their host stars requires some
 nuance.  If true, this statement only holds for the very youngest
-stars, where we see only a marginal underabundance in the KOI host age
+stars, where we see a marginal deficit in the KOI host age
 distribution relative to that of the overall stellar sample.
-At the age resolution of this study, 
-the selection bias against sub-gigayear planets seems small.
+However at the age resolution of this study, 
+this selection bias against sub-gigayear planets seems small.
 
@@
 \ratiosfr$\pm$\uncratiosfr\ over the past three billion years.
+If this decrease continues linearly into the future, then
+a simple extrapolation would imply
+that in $\approx$1.4$\pm$0.3\,Gyr the rate of new stars forming in the
+Galaxy will reach zero.
 

### 2024-05-28 paper/ms.tex
update ms

@@
 
-The point of Figure~\ref{fig:agescalecompone} is hopefully clear:
-until this study, the ages of the youngest Kepler stars and planets were not 
-well-measured.
-While previous studies reported accurate ages for
-$\gtrsim$1\,Gyr stars, ages of sub-gigayear stars have historically been
+Figure~\ref{fig:agescalecompone} show that although
+previous studies reported ages that agree with the cluster scale for
+$\gtrsim$1\,Gyr stars, sub-gigayear stars have historically had their ages 
 overestimated by 0.3--2\,dex, with severely understated uncertainties.  
@@
 The stellar ages 
-are given in Table~\ref{tab:planets} for the known planet-hosts,
+are given in Table~\ref{tab:planets} for the known planet hosts,
 and in Table~\ref{tab:stars} for all Kepler stars.  Figure~\ref{fig:hist_tgyro} shows the rotation-based
-age distributions of all the stars (left column) and the known
-planet-hosts (middle column).  These ``histograms'' are sampled from
+age distributions of all the stars (left column) and the
+KOI hosts (middle column).  These ``histograms'' are sampled from
 the posterior probability distributions by drawing ten random samples
@@
 draws of 70\% of the stars from each sample.  The resulting 
-range of $\log_{\rm 10}p$ (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentile)
-spanned -3.9 to -1.2 for the K dwarfs,
-and -5.2 to -2.2 for all Kepler stars.  This degree of similarity suggests 
-that although weak differences may be present,
-the ``common wisdom'' that young transiting planets are hard to detect
-due to the variability of their host stars requires some
-nuance.  If true, this statement only holds for the very youngest
-stars, where we see a marginal deficit in the KOI host age
+$\log_{\rm 10}p$ values spanned (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentiles)
+-3.9 to -1.2 for the K dwarfs,
+and -5.2 to -2.2 for all Kepler stars.  This suggests 
+that while small differences may be significant, they are not drastic.
+This similarity
+contradicts
+the common-sense argument that young transiting planets are hard to detect
+due to the photometric variability of their host stars.  If true, this statement only holds for the very youngest
+($<$0.4\,Gyr)
+stars, where there is a marginal deficit in the KOI host age
 distribution relative to that of the overall stellar sample.
-However at the age resolution of this study, 
-this selection bias against sub-gigayear planets seems small.
 
@@
 signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
-that it drops to almost zero by $\approx$5\,Gyr .  This trend is
+that it drops to almost zero by $\approx$5\,Gyr.  This trend is
 opposite in functional form to our derived age distributions.

### 2024-05-28 drivers/estimate_monte_carlo_age_distribution.py
minor

@@ -31,7 +31,7 @@ def draw_star_ages(N):
             age_break = 3
             if age <= age_break:
                 # If age is between 0 and 3 billion years, apply the linear increase factor
-                factor = (1 / 2.75) + (age / age_break) * (1 - (1 / 2.75))
+                factor = (1 / 2.49) + (age / age_break) * (1 - (1 / 2.49))
                 if np.random.random() < factor:
                     ages[i] = age
                     break

### 2024-05-28 drivers/plot_param_vs_age.py

@@ -10,7 +10,9 @@ def plot_param_vs_age(param='koi_max_mult_ev'):
     sel = (
         (df.flag_is_gyro_applicable.astype(bool))
         &
-        ~(df.gyro_median == '--')
+        ( ~(df.gyro_median == '--') )
+        &
+        (df.flag_is_ok_planetcand.astype(bool))
     )
     df = df[sel]
 

### 2024-05-28 paper/ms.tex
update ms

@@
   %
-  The resulting yield includes \nplyounggyrotwosigma\ planets
-  younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\ planets
-  at 3$\sigma$).
-  %
-  This is far fewer than expected under the classic assumption of a uniform
-  star formation history.  We find that this dearth can be attributed to the star formation
-  rate in the Kepler field
+  The resulting yield includes \nplyounggyrotwosigma\ planets younger
+  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
+  1\,Gyr.
+  %
+  This is about half the number expected under the classic assumption
+  of a uniform star formation history.  We find that this dearth can
+  be attributed to the star formation rate in the Kepler field
   dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3\,Gyr.  This trend is observed both for known planet hosts as well as for the overall
-  stellar sample.
-  %
-  This ``demographic cliff'' in the Galaxy's star
-  formation history has been previously reported, and it helps clarify the relative balance of
-  young and old planets that have been detected to date using
-  transits.
+  3\,Gyr.  This trend is observed both for known planet hosts as well
+  as for the overall stellar sample.
+  %
+  This ``demographic cliff'' in the Galaxy's star formation history
+  has been previously reported, and it helps clarify the relative
+  balance of young and old planets that have been detected to date
+  using transits.
   %
@@
 
-Young ($<$1\,Gyr) exoplanets therefore represent one means of building
+Young ($<$1\,Gyr) exoplanets represent one means of building
 the timeline for exoplanet evolution.  Informative individual
@@
 multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
-planet with a Neptunian mass (\citealt{Rizzuto_2020}; Thao et
+planet with a sub-Neptunian mass (\citealt{Rizzuto_2020}; Thao et
 al.~submitted).  Population-level analyses have similarly suggested
@@
 assumed planetary) ages are precise at the $\approx$10\% level.  A
-more recent development, facilitated by Gaia, is the idea that finding
+recent development, facilitated by Gaia, is the idea that finding
 transiting planets is often easier than finding the host stellar
@@
 stellar kinematics \citep{2021AJ....161..189L,2024AJ....167..159L},
-and from light curve features such as starspot amplitudes
+and from stellar colors, luminosities, and starspot amplitudes
 \citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
@@
     any one star's rotation period \citep{Bouma_2023}.
-    This represents an improvement in both accuracy and in realism
+    This represents an improvement in the accuracy and realism
     of uncertainty propagation relative to previous calibrations.
@@
     \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
-    These clusters offer a sample of stars with which we can test the
-    reliability of single-star age-dating methods.
+    These clusters offer a sample of stars that we can use to test
+    the reliability of the available single-star age-dating methods.
 \end{enumerate}
@@
 periods for 55{,}232 main-sequence and subgiant FGKM stars.  They
-included known KOIs and binaries in their analysis, and assigned them
-specific quality flags, while removing transits and eclipses during
-the stellar rotation measurement process.
+included known KOIs and binaries in their analysis, and removed
+transits and eclipses during the stellar rotation measurement process.
 
@@
 noticed that some known KOIs with rotation periods were missing.  This
-is not surprising, in that the rotation periods of KOIs have received
+is not surprising, since the rotation periods of KOIs have received
 considerably more scrutiny than those of ordinary Kepler stars.  We
@@
 
-We considered planets in the NEA cumulative KOI table as of 6 June
-2023, which included the best knowledge available on any given planet
+We considered planets in the NEA cumulative KOI table as of 2023 June
+6, which included the best knowledge available on any given planet
 candidate while also incorporating human-based vetting.  These planets
@@
 the spectrum with the highest number of counts.  The resulting spectra
-were acquired between 6 September 2009 and 16 May 2024, and
+were acquired between 2009 September 6 and 2024 May 16, and
 cumulatively comprise \nhireshours\,hours of open-shutter time.
@@
 
-Our EW measurement approach does not correct for the neighboring
+Our EW measurement approach did not correct for the neighboring
 \ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
-precision of our measurement technique, after applying an initial
+precision of our method, after applying an initial
 iteration on the \nlithiumstars\ KOIs with spectra, we compared our
@@
 differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
-required positive EWs, while we allow for statistically negative ones.  At
-$>$30\,m\AA, there is however a small offset in our respective scales,
-such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  This offset is caused by
-our non-treatment of the iron blend.  We therefore directly subtract
+required positive EWs, while we allowed for statistically negative ones.  At
+$>$30\,m\AA, there is however a small offset in our respective scales caused
+by our non-treatment of the iron blend,
+such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.
+We therefore directly subtracted
 this constant mean value (7.50\,m\AA) when calculating lithium-based
-ages.
+ages.  This offset is at worst five times smaller than the astrophysical
+scatter present in lithium EWs for calibration clusters at any given age
+\citep[see][]{Jeffries_2023}.
 
@@
 
-Our default source for stellar temperatures and surface gravities is
+Our default source for stellar temperatures and surface gravities was
 the Gaia-Kepler Stellar Properties Catalog (GKSP;
@@
 temperatures and surface gravities from
-\citetalias{Berger_2020a_catalog}, we adopt the values reported by
+\citetalias{Berger_2020a_catalog}, we adopted the values reported by
 \citet{Santos_2019} and \citet{Santos_2021}, which for these cases
@@
 non-false-positive KOIs with rotation periods have parameters from
-\citet{Berger_2020a_catalog}, and the remainder similarly are drawn
+\citet{Berger_2020a_catalog}, and the remainder are drawn
 from DR25.
@@
 
-We calculate gyrochrone ages using \texttt{gyro-interp}
+We calculated gyrochrone ages using \texttt{gyro-interp}
 \citep{Bouma_2023}, which is a method designed to address the fact
-that stars with the same mass and same rotation period can have very
-different ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
-age posterior should therefore incorporate this intrinsic
-population-level scatter into the precision of its reported
-ages.  Figure~\ref{fig:models} highlights this problem,
-particularly in regions where the 0.1-1\,Gyr stars overlap.
+that stars with the same mass and same rotation period can have a wide
+range of ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
+age posterior should therefore incorporate the intrinsic
+population-level scatter into its statistical precision.  Figure~\ref{fig:models} highlights this problem,
+particularly in regions where the 0.1--1\,Gyr stars overlap.
 
@@
 Hermite polynomials.  We calculated the probability of the
-rotation-based age $t_{\rm gyro}$, given the observed rotation
-periods and effective temperatures (and their uncertainties) by
-integrating Equation~1 of \citet{Bouma_2023} at the default
-resolution.  This procedure marginalizes over the intrinsic age and
-temperature dependent scatter that is observed in the cluster
-sequences to yield statistical age uncertainties.
+rotation-based age, $t_{\rm gyro}$, given the observed 
+periods and temperatures (and their uncertainties) by
+integrating Equation~1 of \citet{Bouma_2023}.  This procedure marginalizes over the 
+astrophysical scatter that is observed in the open cluster
+sequences.
 We adopted a linear age grid spanning 0--5\,Gyr with
@@
 potentially grazing if $b$$>$0.8, using the impact parameters reported
-in the Markov Chain Monte Carlo (MCMC) fits from
-\citet{Thompson_2018}.
+by \citet{Thompson_2018}.
 
@@
 Figure~\ref{fig:li_vs_teff} shows our measured equivalent widths for
-the lithium 6708\AA\ absorption doublet, plotted over the mean
+the lithium 6708\,\AA\ absorption doublet, plotted over the mean
 isochrone models from \texttt{EAGLES} \citep{Jeffries_2023}.  We
-declare an upper-limit for plotting purposes if the 1$\sigma$ lower
+show an upper-limit for plotting purposes if the 1$\sigma$ lower
 limit on the equivalent width is below 10\,m\AA.  Our measured EWs
@@
 \citet{2018A&A...618A..93C,CantatGaudin_2020} and
-\citet{Kounkel_2020}.  For Melange-3 we used the candidates reported
+\citet{Kounkel_2020}.  
+For Theia~520, we used candidate members from \citet{Kounkel_2020}.
+For Melange-3 we used the candidates reported
 by \citet{Barber_2022} and required ``offset'' tangential velocities
@@
 
-There are two main metrics for success in this test: {\it 1)}
-do the reported ages agree with the cluster age? And {\it 2)} do the
+There are two main metrics for success in this test. {\it 1)}
+Do the reported ages agree with the cluster age? {\it 2)} Do the
 reported age uncertainties agree with their dispersion
@@
 two models, because it 
-was directly fit to reproduce the cluster sequences known at the time.  However the
+was fitted to reproduce the open cluster sequences known at the time.  However the
 uncertainties from all of these methods appear to be underestimated.  This
@@
 
-The age distributions of Kepler stars and KOI hosts in Figure~\ref{fig:hist_tgyro} appear
-similar.  One diagnostic for whether the two
-distributions are drawn from the same underlying distribution is the
-Kolmogorov-Smirnov test;  we calculated this statistic in a manner
-that accounts for the Poisson uncertainties by performing 1000 random
-draws of 70\% of the stars from each sample.  The resulting 
-$\log_{\rm 10}p$ values spanned (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentiles)
--3.9 to -1.2 for the K dwarfs,
-and -5.2 to -2.2 for all Kepler stars.  This suggests 
-that while small differences may be significant, they are not drastic.
-This similarity
-contradicts
-the common-sense argument that young transiting planets are hard to detect
-due to the photometric variability of their host stars.  If true, this statement only holds for the very youngest
-($<$0.4\,Gyr)
-stars, where there is a marginal deficit in the KOI host age
-distribution relative to that of the overall stellar sample.
+The age distributions of Kepler stars and KOI hosts in
+Figure~\ref{fig:hist_tgyro} appear similar.  One diagnostic for
+whether the two distributions are drawn from the same underlying
+distribution is the Kolmogorov-Smirnov test;  we calculated this
+statistic in a manner that accounts for the Poisson uncertainties by
+performing 1000 random draws of 70\% of the stars from each sample
+when requiring $t_{\rm gyro}$$<$3\,Gyr.  The resulting $\log_{\rm
+10}p$ values spanned (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentiles)
+-3.9 to -1.2 for the K dwarfs, and -5.2 to -2.2 for all Kepler stars.
+This agrees with the visual impression that while small differences
+may be present, they are not drastic.
+
+The similarity of the age distributions between KOI hosts and Kepler
+stars adds some nuance to the argument that young transiting planets
+are hard to detect due to the photometric variability of their host
+stars.  If true, this statement seems to only hold for the very youngest
+($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the KOI host
+age distribution relative to that of the overall stellar sample.
+Examining the planet properties as a function of time, we similary
+find that fewer systems at $<$0.4\,Gyr are detected with $P_{\rm
+orb}$$\gtrsim$30\,days than at $>$0.4\,Gyr.
 
@@
 
-{\it Theia~520}---We derive rotation-based ages of 330-350\,Myr for
-Kepler-968 and Kepler-52.  These stars were reported by
-\citet{2019AJ....158..122K} to be members of an association called
-Theia~520 based on their spatial and kinematic correlations with
-nearby stars.  The core of this structure is also known as UBC-1
-\citep{2018A&A...618A..59C}.  Based on isochrone-fitting,
-Theia~520 is $247^{+64}_{-51}$\,Myr old \citep{2019AJ....158..122K}.
-An independent analysis by \citet{2024A&A...681A..13F} using g-mode
-asteroseismology, gyrochronology, and isochrone fitting found
-consistent results.  Ongoing work by \citet{Curtis2024} broadly finds
-that the two planet-hosting stars are indeed members of the diffuse
+{\it Theia~520}---We derive rotation-based ages of $\approx$350\,Myr
+for Kepler-968 and Kepler-52.  Based on spatial and kinematic
+clustering, these stars were reported by \citet{2019AJ....158..122K}
+to be members of an association called Theia~520.  The core of this
+structure is also known as UBC-1 \citep{2018A&A...618A..59C}.  Based
+on isochrone and rotation-based age-dating, Theia~520 seems to be
+$\approx$230$\pm$70\,Myr old
+\citet{2019AJ....158..122K,2024A&A...681A..13F}, which is within the
+1$\sigma$ statistical uncertainties of our $t_{\rm gyro}$
+measurements.  Ongoing work by \citet{Curtis2024} broadly finds that
+the two planet-hosting stars are indeed members of this diffuse
 population.  Given the cluster-level and individual-star evidence,
@@
 rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
-posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm eff}$$\approx$5100\,K, $P_{\rm
-rot}$$\approx$10.3\,days) is marginally hotter than the later K dwarfs
-that have ``stalled'' spindown from 0.7 to 1\,Gyr, enabling its
-rotation period to be diagnostic of the stellar age.
+posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm
+eff}$$\approx$5100\,K, $P_{\rm rot}$$\approx$10.3\,days) is marginally
+hotter than the later K dwarfs that are in ``stalled'' spindown at
+this time, enabling its rotation period to be diagnostic of its age.
 
@@
 yielded \allagesyesconsistent\ consistent cases,
-\allagesmaybeconsistent\ ``maybe consistent'' cases, and
+\allagesmaybeconsistent\ maybe consistent cases, and
 \allagesnoconsistent\ inconsistent cases.  Rephrased, in the overall
@@
 
-Classifying Kepler stars in this manner helps connect them to the
-broader literature focused on the history of star formation in the
+Classifying Kepler stars as thin vs.~thick disk members helps connect them to 
+previous work on the star formation history of the
 Galaxy.  On extragalactic scales, the star formation rate in spiral
@@
 ago, and has since decreased by an order of magnitude
-\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  For our own
+\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
 galaxy, many investigators have worked to measure similar star
-formation histories, through both CMD fitting of resolved stellar
+formation histories through CMD fitting of resolved stellar
 populations
@@
 subset.  This decision was driven by a desire for homogeneity 
-irrespective of planet-hosting status.  It may however have
-incurred small systematic errors in the implied ages.
-
-We were interested in not only the overall age
-distribution of the Kepler field, but also in the reliability of
-individual ages for young stars known to host planets.  
-We assessed this reliability using
-quality
-bitmasks (Table~\ref{tab:planets}), 
-a comparison against open cluster ages (Section~\ref{sec:litagecomparison}),
-and a comparison between rotation and lithium ages.
-The open cluster comparison suggested that our ages were generally
-accurate at $\ll$0.5\,dex across 0.04--2.5\,Gyr, comparable to their
-stated precision level.
-Nonetheless,
-outliers do exist in the upper-left panel of Figure~\ref{fig:agescalecompone},
-typically at most 10\% of the population in each cluster.
-These stars could be either field star contaminants, or else anomalous
-stars whose physical rotation histories were altered by processes not
-captured by our statistical uncertainties.
-Similarly, although
-we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
-for 
-\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
-stars, we did find
-one system, Kepler-786, with a precise lithium age
-(\kepseveneightsix\,Myr), but no rotation signal.
+irrespective of planet-hosting status.  However it implies that the
+implied planet ages might have shift if one were to account for the
+spectroscopic information.
+
+We were interested in not only the overall age distribution of the
+Kepler field, but also in the reliability of individual ages for young
+stars known to host planets.  We assessed this reliability using
+quality bitmasks (Table~\ref{tab:planets}), a comparison against open
+cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
+between rotation and lithium ages.  The open cluster comparison
+suggested that our ages were generally accurate at $\ll$0.5\,dex
+across 0.04--2.5\,Gyr, comparable to their stated precision level.
+Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
+shows that outliers do exist, typically at most 10\% of the population
+in each cluster.  These stars could be either field star contaminants,
+or else anomalous stars whose physical rotation histories were altered
+by processes not captured by our statistical uncertainties.
+Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
+for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
+stars, we did find one system, Kepler-786, with a precise lithium age
+(\kepseveneightsix\,Myr), but no rotation signal.  Planetary mergers
+are an example of a process that might produce such a signal, but
+testing this would require a way of measuring differential abundances
+across a large number of elements.
 
@@
 A broader question, beyond the scope of this work, concerns the ages
-of stars $\gtrsim$4\,Gyr old.  For
-age-dating models based only on rotation rates, or else on a
-combination of rotation and spin-modulation amplitude
-\citep{2023ApJ...952..131M}, an important challenge is that the
-detectability of the rotation signals for 3-10\,Gyr stars, especially
-Sun-like stars, is at the limits of the current data
-\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods
-using either asteroseismology
+of stars $\gtrsim$4\,Gyr old.  For age-dating models based only on
+rotation rates, an important challenge is that the detectability of
+the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
+is at the limits of the Kepler data \citep{2022ApJ...937...94M}.  In
+the ``old star'' regime, methods using either asteroseismology
 \citep{vanSaders_2016,2024ApJ...962..138S}, or else using both the
 evolution of stellar luminosity and stellar rotation
-\citep{Angus_2019} seem the most capable of providing useful age
-constraints for single stars.  Incorporating kinematics
-\citep{2021AJ....161..189L} seems likely to remain most useful for
-stellar ensembles, due to the stochasticity of kinematic heating.
+\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most capable of
+providing useful age constraints for single stars.  Incorporating
+kinematics \citep{2021AJ....161..189L} seems likely to remain most
+useful for stellar ensembles, due to the stochasticity of kinematic
+heating.
 
@@
 Galactic Plane Time Domain Survey.
-The long-baseline surveys 
-that specifically target different regions of the Galaxy
-seem most likely to yield new insight into its star
-formation history.
+The surveys with good coverage of different regions of the Galaxy seem
+most likely to yield new insight into whether the star formation
+history we see in the Kepler field is universal across the thin disk.
 
 {\it Searches for new associations}:
-Some of the youngest systems in Table~\ref{tab:planets} have been
-linked to their birth clusters, while others have not. A systematic
-search for the birth clusters of the youngest ``field stars'' could
-combine spatial and kinematic information from Gaia with rotation
-measurements from TESS. Given the $\sim$100\,Myr decoherence time of
-young clusters, the youngest "field stars" may have easily
-identifiable young neighbors.
+Some of the youngest stars in Tables~\ref{tab:planets}
+and~\ref{tab:stars} have been linked to their birth clusters, while
+others have not. A systematic search for the birth clusters of the
+youngest ``field stars'' could combine spatial and kinematic
+information from Gaia with rotation measurements from TESS. Given the
+$\sim$100\,Myr decoherence time of young clusters, the youngest ``field
+stars'' may have easily identifiable young neighbors.
 
@@
 
-We began this study with two questions: how wrong is the assumption
+We began this work with two questions: how wrong is the assumption
 of a uniform age distribution for stars in the galactic thin disk?
@@
 We therefore visually selected KOIs that appeared
-to be near the main sequence -- i.e.~ most of the yellow
+to be near the main sequence -- i.e.~most of the yellow
 points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
@@
 above this orange locus;
-they are simply flagged as not being near the main sequence.
+they are simply flagged as not being near the main sequence in the $\log g$ vs.~$T_{\rm eff}$ plane.
 

### 2024-05-28 drivers/construct_field_star_gyro_quality_flags.py
HR diagram outliers

@@ -181,7 +181,23 @@ def build_gyro_quality_flag(sample='gyro', datestr='20240430'):
         _y0 = p - 0.12
         # 'top'
         y0 = np.maximum(_y, _y0)
-        df['flag_farfrommainsequence'] = df['adopted_logg'] < y0
+
+        sel_logg_teff = df['adopted_logg'] < y0
+
+        # from tests/test_MG_bprp_locus.py
+        csvpath = join(DATADIR, "interim", f"MG_bprp_locus_coeffs_poly.csv")
+        coeffs = pd.read_csv(csvpath).values.flatten()
+
+        _bprp = df['dr3_bp_rp']
+        poly_vals = np.polyval(coeffs, _bprp)
+
+        sel_MG_bprp = ~(
+            (df['M_G'] > poly_vals - 1)
+            &
+            (df['M_G'] < poly_vals + 1)
+        )
+
+        df['flag_farfrommainsequence'] = sel_logg_teff | sel_MG_bprp
 
     # calculate for ater
     df['b20t2_rel_E_Age'] = np.abs(df['b20t2_E_Age'])/df['b20t2_Age']

### 2024-05-28 gyrojo/locus_definer.py

@@ -60,7 +60,7 @@ def detect_outliers_windowed(x, y, window_size=10, threshold=2):
     return outliers
 
 
-def fit_polynomial(data, N):
+def fit_polynomial(data, N, xkey='adopted_Teff', ykey='adopted_logg'):
     """
     Fit an Nth-order polynomial to the given data points.
 
@@ -72,7 +72,7 @@ def fit_polynomial(data, N):
     tuple: The polynomial coefficients and the mean squared error of the fit.
     """
 
-    xvals, yvals = nparr(data['adopted_Teff']), nparr(data['adopted_logg'])
+    xvals, yvals = nparr(data[xkey]), nparr(data[ykey])
     inds = np.argsort(xvals)
     xvals, yvals = xvals[inds], yvals[inds]
 
@@ -86,7 +86,7 @@ def fit_polynomial(data, N):
     p = np.poly1d(coeffs)
 
     # Calculate the mean squared error
-    mse = mean_squared_error(data['adopted_logg'], p(data['adopted_Teff']))
+    mse = mean_squared_error(data[ykey], p(data[xkey]))
 
     return p, mse, coeffs
 
@@ -106,7 +106,7 @@ def bic(mse, n, k):
     return n * np.log(mse) + k * np.log(n)
 
 
-def evaluate_models(data, max_degree):
+def evaluate_models(data, max_degree, xkey='adopted_Teff', ykey='adopted_logg'):
     """
     Evaluate polynomial models of various degrees and calculate their BIC scores.
 
@@ -119,7 +119,7 @@ def evaluate_models(data, max_degree):
     """
     results = []
     for N in range(1, max_degree + 1):
-        p, mse, _ = fit_polynomial(data, N)
+        p, mse, _ = fit_polynomial(data, N, xkey=xkey, ykey=ykey)
         k = N + 1  # Number of parameters is degree + 1 (including the constant term)
         bic_score = bic(mse, len(data), k)
         results.append({
@@ -130,19 +130,19 @@ def evaluate_models(data, max_degree):
 
         # Plot the data and the fitted polynomial
         plt.figure(figsize=(10, 6))
-        plt.scatter(data['adopted_Teff'], data['adopted_logg'], s=10,
+        plt.scatter(data[xkey], data[ykey], s=10,
                     label='Data Points')
-        plt.plot(np.sort(data['adopted_Teff']),
-                 p(np.sort(data['adopted_Teff'])), color='red',
[…]

### 2024-05-28 gyrojo/plotting.py

@@ -2286,6 +2286,18 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G', vtangcut=None):
                 edgecolors='k'
             )
 
+    if ykey == 'M_G' and xkey == 'dr3_bp_rp':
+        csvpath = join(DATADIR, "interim", f"MG_bprp_locus_coeffs_poly.csv")
+        coeffs = pd.read_csv(csvpath).values.flatten()
+        _bprp = np.linspace(0.5, 3, 1000)
+        poly_vals = np.polyval(coeffs, _bprp)
+        #ax.plot(bprp, poly_vals, zorder=5, c='C1', lw=0.5)
+        show_bounds = 0
+        if show_bounds:
+            ax.plot(_bprp, poly_vals-1, zorder=5, c='C1', lw=0.5)
+            ax.plot(_bprp, poly_vals+1, zorder=5, c='C1', lw=0.5)
+
+
     if ykey == 'adopted_logg' and xkey == 'adopted_Teff':
         xerr = dfs[1].adopted_Teff_err.mean()
         yerr = dfs[1].adopted_logg_err.mean()

### 2024-05-28 paper/ms.tex
update ms

@@
   suitable for gyrochronology (dark gray points), based on factors
-  including their non-binarity and proximity to the spectroscopic main
-  sequence (orange boundary; see Section~\ref{subsec:flags}).  Some
+  including their non-binarity and proximity to the main
+  sequence (see Section~\ref{subsec:flags}).  Some
   are ``confirmed'' or ``candidate'' KOIs that meet an additional set
@@
 but the currently available open cluster data have yet to clarify
-exactly when the intrinsic scatter in the population decreases.
+when the intrinsic scatter in the population decreases.
 
@@
 are far from the main sequence: this can be seen in both panels of
-Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
-ages because {\it i)} point sources far from the main sequence include
-unresolved photometric binaries, and {\it ii}) effects other than
-magnetic braking influence the rotation periods of bona fide single
-stars far from the main sequence.  The added concerns of
-reddening and metallicity further complicate matters.  After exploring
-various options, we settled on the orange locus in the $\log
-g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
-the exact details of how this locus is constructed are arbitrary (see Appendix~\ref{app:linemethod}),
-the general aim is to flag stars for which there is evidence based on
-their location in the $\log g$--$T_{\rm eff}$ plane that our
-gyrochronology model may not be reliable.  
+Figure~\ref{fig:stellarprops}.  Some of the challenges this introduces
+for rotation-based ages include unresolved binaries, metallicity,
+reddening, and spin-evolution effects other than magnetic braking.
+After exploring various options, we settled on the orange locus in the
+$\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
+as a way of flagging unresolved binaries, as well as mildly evolved
+late-F and early-G stars.  While the exact details of how this locus
+is constructed are arbitrary (see Appendix~\ref{app:linemethod}), the
+general aim is to flag stars for which there is evidence based on
+their location in the HR diagram that our gyrochronology model may not
+be reliable.  While this $\log g$--$T_{\rm eff}$ cut flags about half
+of Kepler stars with rotation periods as potentially not being
+suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
+BP}$-$G_{\rm RP}$ remained after applying this cut.  Such stars are
+likely also questionable.  We therefore also
+fitted a polynomial to the KOI main sequence, and flagged stars more
+than 1\,magnitude from this locus as part of the same bitflag.
+
 %The rotation period
@@
 %likely due to a combination of astrophysical and systematic effects.
-See for instance Figure~8a of \citet{2022AJ....164..137K} or
-Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
-the period distribution of this population differs from stars near the
-main sequence.
+%See for instance Figure~8a of \citet{2022AJ....164..137K} or
+%Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
+%the period distribution of this population differs from stars near the
+%main sequence.
 

### 2024-05-28 drivers/plot_st_params.py
minor

@@ -5,14 +5,16 @@ import gyrojo.plotting as ap
 outdir = os.path.join(RESULTSDIR, 'st_params')
 if not os.path.exists(outdir): os.mkdir(outdir)
 
-# logg vs Teff
+# in ms
+ap.plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G')
 ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='adopted_logg')
+
+# logg vs Teff
 for vtangcut in ['thindisk', 'thickdisk', 'halo']:
     ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='adopted_logg',
                       vtangcut=vtangcut)
 
 # CAMDs
-ap.plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G')
 ap.plot_st_params(outdir, xkey='adopted_Teff', ykey='M_G')
 
 # mKep vs Teff

### 2024-05-28 paper/ms.tex
update ms

@@
 The ``cluster-first'' approach has a major limitation: most nearby
-sub-gigayear stars are in the field.  Only $\lesssim$1\% of stars
+stars are in the field.  Only $\lesssim$1\% of stars
 within $\approx$500\,pc have been associated with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-One can appreciate the problem by querying the NASA Exoplanet Archive
+The implication can appreciated by querying the NASA Exoplanet Archive
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
 than 1\,Gyr.  Requiring say $t$$<$1\,Gyr at
-2$\sigma$ precision gives $\approx$50 ``young'' transiting planets at
+2$\sigma$ precision gives $\approx$50 such planets at
 the time of writing.  Most of these young planets are in clusters, and
@@
     any one star's rotation period \citep{Bouma_2023}.
-    This represents an improvement in the accuracy and realism
+    This represents an improvement in the accuracy 
     of uncertainty propagation relative to previous calibrations.
@@
 Independently, recent years have also yielded improvements in the
-lithium age scale.  Lithium ages include two distinct regimes:
+lithium age scale.  Lithium ages include 
 depletion boundary ages for M dwarfs and brown dwarfs in star
-clusters, and the less precise decline-based ages for individual field
-FGK stars \citep{Soderblom_2010}.  Here we focus on the latter
-approach, which relies on the observation that the Li abundances of
-partially-convective stars gradually decline as they age
+clusters, and decline-based ages for individual field
+FGK stars \citep{Soderblom_2010}.  The latter
+approach relies on the observed decline of Li abundances in
+partially-convective stars as they age
 \citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
@@
 stars in 52 open clusters.  Two-sided lithium ages are useful for
-Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, with a strong
-dependence on spectral type since the K dwarfs lose their surface
-lithium much faster than G dwarfs.  The precision of the ages reported
-by the \citet{Jeffries_2023} method in this regime are in the range of
-0.3-0.5~dex.
+Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, though with a strong
+dependence on spectral type.  The precision of lithium ages 
+in this regime are now in the range of 0.3-0.5~dex.
 
@@
 		\subfloat{
-			\includegraphics[width=0.475\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.485\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.475\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.485\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 
-{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
-are far from the main sequence: this can be seen in both panels of
-Figure~\ref{fig:stellarprops}.  Some of the challenges this introduces
+{\it Near the main sequence (bit 9)}---Figure~\ref{fig:stellarprops} shows
+that many stars observed by Kepler
+are far from the main sequence.  Some of the challenges this introduces
 for rotation-based ages include unresolved binaries, metallicity,
-reddening, and spin-evolution effects other than magnetic braking.
+reddening, and drivers of rotation other than magnetic braking.
 After exploring various options, we settled on the orange locus in the
 $\log g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}
-as a way of flagging unresolved binaries, as well as mildly evolved
+as a way of flagging unresolved binaries, as well as evolved
 late-F and early-G stars.  While the exact details of how this locus
@@
 their location in the HR diagram that our gyrochronology model may not
-be reliable.  While this $\log g$--$T_{\rm eff}$ cut flags about half
+be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
 of Kepler stars with rotation periods as potentially not being
 suitable for gyrochronology, a few outliers in $M_{\rm G}$ vs.~$G_{\rm
-BP}$-$G_{\rm RP}$ remained after applying this cut.  Such stars are
-likely also questionable.  We therefore also
-fitted a polynomial to the KOI main sequence, and flagged stars more
-than 1\,magnitude from this locus as part of the same bitflag.
+BP}$-$G_{\rm RP}$ remained after applying it.  Such outliers are
+likely also questionable.  We therefore also fitted a polynomial to
+the KOI main sequence in $M_{\rm G}$ vs.~$G_{\rm
+	BP}$-$G_{\rm RP}$, and flagged stars more than 1\,magnitude from
+this locus as part of the same bitflag.
 

### 2024-05-29 paper/ms.tex
update ms

@@
 
-%\author{and coauthors, including but not limited to...}
+\author{and coauthors, including but not limited to...}
 
@@
 \author[0000-0001-8638-0320]{Andrew W. Howard}
+\affiliation{\caltech}
+
+\author{and to be confirmed:}
 \affiliation{\caltech}
@@
 		\subfloat{
-			\includegraphics[width=0.485\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.485\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 	\caption{
-		{\bf Literature ages vs.~benchmark open cluster ages.}
+		{\bf Reported ages of individual stars in benchmark open clusters.}
 		Each point denotes a reported age,
@@
 		membership lists,
-		any outliers can be compared between different methods on a
+		outliers can be compared between different methods on a
 		relative basis.
-		The number of stars in any given cluster follows by crossmatching every study against
-		a fixed cluster list; certain methods report ages for more stars than others.
+    Each study was cross-matched against the same cluster list;
+    certain methods report ages for more stars than others.
 		Horizontal scatter is artificially added to visually

### 2024-05-29 drivers/compare_literature_ages_to_cluster_ages.py
minor

@@ -579,16 +579,27 @@ for ix, (cluster_df, cluster_full_name, true_age) in enumerate(zip(
                 ax.errorbar(
                     eps+ix*np.ones_like(yval[~upperlim]), yval[~upperlim],
                     yerr=yerr[:, ~upperlim],
-                    marker='o', elinewidth=0.2, capsize=0, lw=0, mew=0.5, color=f'C{iy}',
-                    markersize=0.5, zorder=5
+                    marker='o', elinewidth=0.1, capsize=0, lw=0, mew=0.5, color=f'C{iy}',
+                    markersize=0, zorder=5, alpha=0.6
+                )
+                ax.errorbar(
+                    eps+ix*np.ones_like(yval[~upperlim]), yval[~upperlim],
+                    yerr=yerr[:, ~upperlim],
+                    marker='o', elinewidth=0., capsize=0, lw=0, mew=0.5, color=f'C{iy}',
+                    markersize=0.5, zorder=6
                 )
 
         else:
             eps = np.random.normal(loc=0, scale=0.12, size=len(yval))
             ax.errorbar(
                 eps+ix*np.ones_like(yval), yval, yerr=yerr,
-                marker='o', elinewidth=0.2, capsize=0, lw=0, mew=0.5, color=f'C{iy}',
-                markersize=0.5, zorder=5
+                marker='o', elinewidth=0.1, capsize=0, lw=0, mew=0.5, color=f'C{iy}',
+                markersize=0., zorder=5, alpha=0.6
+            )
+            ax.errorbar(
+                eps+ix*np.ones_like(yval), yval, yerr=yerr,
+                marker='o', elinewidth=0., capsize=0, lw=0, mew=0.5, color=f'C{iy}',
+                markersize=0.5, zorder=6
             )
 
         for lval in [0.01, 0.1, 1, 10, 100]:

### 2024-06-02 paper/ms.tex
minor
update ms

@@
 few open clusters known in the Kepler field had been
-visually identified by \citet{1864RSPT..154....1H}.  Gaia
+cataloged by \citet{1864RSPT..154....1H}.  Gaia
 has enabled the discovery of stellar ensembles
@@
 by \citet{Barber_2022} and required ``offset'' tangential velocities
-below 2\,\kms.  For Cep-Her we used candidate members from Table~2 of
-\citet{Bouma_2022b} with ``weights'' exceeding 0.1.  Even with
+below 2\,\kms.  For Cep-Her we used candidate members from Kerr et al.~2024 (submitted)
+with $P_{\rm fin}$$>$0.8.  Even with
 contaminants, we can compare the relative age distributions derived by
@@
 %    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
-    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf}
+%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf}
+        \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
     }
@@
 %    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
-    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf}
+%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf}
+        \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
     }
@@
 10}p$ values spanned (2.5$^{\rm th}$ to 97.5$^{\rm th}$ percentiles)
--3.9 to -1.2 for the K dwarfs, and -5.2 to -2.2 for all Kepler stars.
+-4.5 to -1.8 for the K dwarfs, and -4.5 to -1.9 for all Kepler stars.
 This agrees with the visual impression that while small differences
@@
 
-%\section{Figure dump}
-%
-%\begin{figure*}[!t]
-%  \begin{center}
-%    \leavevmode
-%    \includegraphics[width=0.5\textwidth]{gyroage_vs_teff_errs_showplanets_linear.png}
-%  \end{center}
-%  \vspace{-0.6cm}
-%  \caption{
-%    {\bf Gyro-age vs Teff}.
-%    $\pm$1$\sigma$ uncertainties are plotted around the median for Kepler stars (black) and KOIs (blue).
-%    Visual pileup on the left is b/c i) uncertain ages, and ii) lots of stars there.  Precision-land is the other portion.
-%    \label{fig:gyroage_vs_teff}
-%  }
-%\end{figure*}
+\section{Age Diagnostics}
+\label{app:age_diagnostic}
+
+Figure~\ref{fig:gyroage_vs_teff} is a visualization of $t_{\rm gyro}$
+vs.~$T_{\rm eff}$ for our rotation-based age catalog. 
+Each bar denotes the $\pm$1$\sigma$ uncertainty for a star's
+rotation-based age:
+this plot can be viewed as the transformation of
+the left panel of Figure~\ref{fig:prot_vs_teff}.
+The overdensity of $P_{\rm rot}$$\approx$20\,day G dwarfs corresponds
+to the large overdensity at $t_{\rm gyro}$$\approx$3\,Gyr.
+The (non-physical) deficit of $\sim$22\,day rotators appears as a
+deficit between 5500--6000\,K.
+A second deficit is also visible from 3800--5000\,K;
+it is associated with the ``intermediate period gap'', which may be
+associated with a transition from spot-dominated to faculae-dominated
+light curves \citep[e.g.][]{Reinhold2019}.
+At $\approx$1\,Gyr, stalled spin-down yields larger uncertainties for
+K dwarfs.
+The pile-up near $\approx$5\,Gyr is imposed by our choice of prior;
+given that \texttt{gyro-interp} returns non-calibrated ages in this
+regime, we truncated our analysis at 4\,Gyr.
+
+\begin{figure*}[!t]
+  \begin{center}
+    \leavevmode
+        \includegraphics[width=0.48\textwidth]{gyroage_vs_teff_errs_linear.png}
+    \includegraphics[width=0.48\textwidth]{gyroage_vs_teff_errs_showplanets_linear.png}
+  \end{center}
+  \vspace{-0.6cm}
+  \caption{
+    {\bf Rotation-based age as a function of effective temperature}.
+    The entire Kepler sample is shown on the left.
+    The KOI hosts are shown on the right.
+    Each bar shows the $\pm$1$\sigma$ uncertainty for a star's age as
+    derived from its rotation period.  
+    Individual features are discussed in
+    Appendix~\ref{app:age_diagnostic}.
+    \label{fig:gyroage_vs_teff}
+  }
+\end{figure*}
 

### 2024-06-02 paper/collect.sh

@@ -19,8 +19,8 @@ cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
 cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
 
 # plot_hist_field_gyro_ages 
-cp $r/hist_field_gyro_ages_20240521/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf .
-cp $r/hist_field_gyro_ages_20240521/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf .
+cp $r/hist_field_gyro_ages_20240530/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf .
+cp $r/hist_field_gyro_ages_20240530/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf .
 
 # plot_li_vs_teff
 #cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
@@ -36,5 +36,6 @@ cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pd
 # appendices (probably omit)
 r="/Users/luke/Dropbox/proj/gyrojo/results"
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
+cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_linear.png .
 
 cp $r/lit_ages_vs_cluster_ages/olympic_merged_lit_ages_vs_cluster_ages.pdf .

### 2024-06-03 paper/ms.tex
further table formatting

@@
   \tablecaption{Ages of Kepler planets and planet candidates.  This
-  version of the table is truncated to include only the youngest
+  version of the table is truncated to include the youngest
   confirmed Kepler planets, sorted by the minimum of either the
@@
   positive KOIs with MES$>$10.  
-  A bash script to decode any given $Q_{\rm star}$ quality flag is
+  A bash script to decode the $Q_{\rm star}$ quality flag is
   {\bf
@@
   is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
+  All quoted uncertainties are statistical; they do not incorporate
+  possible systematic uncertainties, e.g.~in the absolute ages of open clusters.
   \label{tab:planets}}
@@
 \appendix
-\section{Defining An ``Isochronally Too Far From The Main Sequence'' Boundary}
+\section{Defining An ``Isochronally Far From The Main Sequence'' Boundary}
 \label{app:linemethod}
@@
   \caption{
-    {\bf Rotation-based age as a function of effective temperature}.
+    {\bf Rotation-based age vs.~effective temperature}.
     The entire Kepler sample is shown on the left.
-    The KOI hosts are shown on the right.
+    KOI hosts are shown on the right.
     Each bar shows the $\pm$1$\sigma$ uncertainty for a star's age as
     derived from its rotation period.  
-    Individual features are discussed in
+    Notable features are discussed in
     Appendix~\ref{app:age_diagnostic}.

### 2024-06-03 drivers/make_star_table.py

@@ -129,6 +129,10 @@ def make_star_table(
 
     latexpath = join(PAPERDIR, 'table_star_gyro.tex')
     np.random.seed(42)
+
+    # format ages
+    sel = sdf.gyro_median.astype(float) > 4000
+    _sdf.loc[sel, 't_gyro'] = '$> 4000$'
     _sdf.sample(n=10).to_latex(
         latexpath, index=False, na_rep='--', escape=False, formatters=formatters
     )
@@ -166,7 +170,7 @@ def make_star_table(
     sdf.loc[sel, 'gyro_median'] = np.nan
     sdf.loc[sel, 'gyro_+1sigma'] = np.nan
     sdf.loc[sel, 'gyro_-1sigma'] = np.nan
-    sdf.to_csv(csvpath, index=False)
+    sdf.to_csv(csvpath, index=False, na_rep='')
     print(f'Wrote {csvpath}')
 
 

### 2024-06-03 drivers/make_younghighlight_table.py

@@ -506,7 +506,7 @@ def make_table(
                     outdf[column] = outdf[column].str.replace(key, value)
 
         csvpath = join(TABLEDIR, 'table_allageinfo.csv')
-        outdf.to_csv(csvpath, index=False, na_rep='--')
+        outdf.to_csv(csvpath, index=False)
         print(f'Wrote {csvpath}')
 
         # format ages
@@ -522,11 +522,11 @@ def make_table(
         outdf.loc[sel, 'gyro_-1sigma'] = np.nan
 
         csvpath = join(PAPERDIR, 'table_allageinfo_agesformatted.csv')
-        outdf.to_csv(csvpath, index=False, na_rep='--')
+        outdf.to_csv(csvpath, index=False)
         print(f'Wrote {csvpath}')
 
         csvpath = join(TABLEDIR, 'table_allageinfo_allcols.csv')
-        longdf.to_csv(csvpath, index=False, na_rep='--')
+        longdf.to_csv(csvpath, index=False)
         print(f'Wrote {csvpath}')
 
 

### 2024-06-03 paper/ms.tex
update ms

@@
 are given in Table~\ref{tab:planets} for the known planet hosts,
-and in Table~\ref{tab:stars} for all Kepler stars.  Figure~\ref{fig:hist_tgyro} shows the rotation-based
+and in Table~\ref{tab:stars} for all Kepler stars.  
+Figure~\ref{fig:hist_tgyro} shows the rotation-based
 age distributions of all the stars (left column) and the
@@
 statistics.  We truncated the plots near $4$\,Gyr since this is the
-cluster age out to which our interpolation-based method is calibrated.
+cluster age out to which our interpolation-based method is well-calibrated.
 The overall slope yields a dearth of young stars relative to the expectation
@@
 
-The similarity of the age distributions between KOI hosts and Kepler
-stars adds some nuance to the argument that young transiting planets
+The similarity of the KOI host and Kepler
+star age distributions adds some nuance to the argument that young transiting planets
 are hard to detect due to the photometric variability of their host
-stars.  If true, this statement seems to only hold for the very youngest
+stars.  If true, this statement seems to hold only for the very youngest
 ($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the KOI host
 age distribution relative to that of the overall stellar sample.
-Examining the planet properties as a function of time, we similary
-find that fewer systems at $<$0.4\,Gyr are detected with $P_{\rm
-orb}$$\gtrsim$30\,days than at $>$0.4\,Gyr.
+Examining scatter plots of the planet properties as a function of time, we similarly
+find that fewer systems at $t_{\rm gyro}$$<$0.4\,Gyr are detected with $P_{\rm
+orb}$$\gtrsim$30\,days than at $t_{\rm gyro}$$>$0.4\,Gyr.
 
@@
   ``Confirmed'' planets appear in the machine-readable
-  version before``candidate'' planets.  The bit quality flags for the
+  version before``candidate'' planets.  
+  Planetary sizes are mostly drawn from (in order of precedence)
+  \citet{Petigura_2022}, \citet{Berger_2020b_rpage}, and \citet{Thompson_2018}, and might be unphysical
+  for grazing planets.
+  The bit quality flags for the
   rotation-based ages, $Q_{\rm star}$, are described in

### 2024-06-03 paper/ms.tex
update ms

@@
 
-The stellar ages 
-are given in Table~\ref{tab:planets} for the known planet hosts,
-and in Table~\ref{tab:stars} for all Kepler stars.  
-Figure~\ref{fig:hist_tgyro} shows the rotation-based
-age distributions of all the stars (left column) and the
-KOI hosts (middle column).  These ``histograms'' are sampled from
-the posterior probability distributions by drawing ten random samples
-from each posterior, and then computing the re-weighted
-histogram of the resulting samples.  The uncertainties assume Poisson
-statistics.  We truncated the plots near $4$\,Gyr since this is the
-cluster age out to which our interpolation-based method is well-calibrated.
-The overall slope yields a dearth of young stars relative to the expectation
-of a uniform age distribution.
+The stellar ages are given in Table~\ref{tab:planets} for the known
+planet hosts, and in Table~\ref{tab:stars} for all Kepler stars.
+Figure~\ref{fig:hist_tgyro} shows the rotation-based age distributions
+of all the stars (left column) and the KOI hosts (middle column).
+These ``histograms'' are sampled from the posterior probability
+distributions by drawing ten random samples from each posterior, and
+then computing the re-weighted histogram of the resulting samples.
+The uncertainties assume Poisson statistics.  We truncated the plots
+near $4$\,Gyr since this is the cluster age out to which our
+interpolation-based method is well-calibrated.  The overall slope
+yields a dearth of young stars relative to the expectation of a
+uniform age distribution.
 %In an alternative coinage, this tendency for fewer young
@@
 
-The similarity of the KOI host and Kepler
-star age distributions adds some nuance to the argument that young transiting planets
-are hard to detect due to the photometric variability of their host
-stars.  If true, this statement seems to hold only for the very youngest
-($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the KOI host
-age distribution relative to that of the overall stellar sample.
-Examining scatter plots of the planet properties as a function of time, we similarly
-find that fewer systems at $t_{\rm gyro}$$<$0.4\,Gyr are detected with $P_{\rm
-orb}$$\gtrsim$30\,days than at $t_{\rm gyro}$$>$0.4\,Gyr.
+The similarity of the KOI host and Kepler star age distributions adds
+some nuance to the argument that young transiting planets are hard to
+detect due to the photometric variability of their host stars.  If
+true, this statement seems to hold only for the very youngest
+($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
+KOI host age distribution relative to that of the overall stellar
+sample.  Examining scatter plots of the planet properties as a
+function of time, we similarly find that fewer systems at $t_{\rm
+gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
+than at $t_{\rm gyro}$$>$0.4\,Gyr.
 
@@
 formation rate of stars in the Kepler field decreased by a factor of
-\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.
-If this decrease continues linearly into the future, then
-a simple extrapolation would imply
-that in $\approx$1.4$\pm$0.3\,Gyr the rate of new stars forming in the
-Galaxy will reach zero.
+\ratiosfr$\pm$\uncratiosfr\ over the past three billion years.  If
+this decrease continues linearly into the future, then a simple
+extrapolation would imply that in $\approx$1.4$\pm$0.3\,Gyr the rate
+of new stars forming in the Galaxy will reach zero.
 
@@
 
-Sub-gigayear planets hold particular weight in studies of planet
+Sub-gigayear planets can be particularly informative for studies of planet
 evolution.
-While our catalog has \nplyounggyro\ confirmed planets
-with median ages below 1\,Gyr, requiring $t_{\rm gyro}$$<$1\,Gyr at
-2$\sigma$ confidence yields \nplyounggyrotwosigma\ planets orbiting
+Our catalog has \nplyounggyro\ confirmed planets
+with median ages below 1\,Gyr.  Requiring $t_{\rm gyro}$$<$1\,Gyr at
+2$\sigma$ yields \nplyounggyrotwosigma\ planets orbiting
 \nplhostsyounggyrotwosigma\ stars.  
-The most notable of these planets can be split into two classes: those
-in clusters, and those with independent and consistent rotation and
+The youth claim is most secure for the subsets that are either
+in clusters, or that have independent rotation and
 lithium-based ages.
@@
 total of fourteen transiting planets.  Our analysis blindly recovers
-the youth of these planets.
-
-{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex has been
-dated using the isochrone loci of its
-pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
+the youth of all of these planets.
+
+{\it Cep-Her}---The Cep-Her complex ($\approx$40\,Myr)
+contains Kepler-1627Ab, Kepler-1643b,
 Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
-Different associations in the complex vary in age by $\approx$50\%
+Different sub-groups of the complex vary in age by $\approx$50\%
 (Kerr~et al.,~submitted).  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
-2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
+2-4\,$R_\oplus$.  Our rotation and lithium ages agree with the cluster age
 for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
@@
 {\it MELANGE-3}---\citet{Barber_2022} reported a
-$105\pm10$\,Myr association in the Kepler field and derived its age
-using rotation, lithium, and the isochronal pre-main-sequence.
-Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
-this association.  We find rotation-based ages
+$105\pm10$\,Myr association in the Kepler field
+containing two transiting mini-Neptunes:
+Kepler-970 and Kepler-1928.  We find rotation-based ages
 of $t_{\rm
@@
 age being slightly older than the suggested Pleiades age for the
-assocation agrees with Figure~4 of \citet{Barber_2022}.
+association agrees with Figure~4 of \citet{Barber_2022}.
 
@@
 clustering, these stars were reported by \citet{2019AJ....158..122K}
-to be members of an association called Theia~520.  The core of this
+to be members of Theia~520.  The core of this
 structure is also known as UBC-1 \citep{2018A&A...618A..59C}.  Based
@@
 $\approx$230$\pm$70\,Myr old
-\citet{2019AJ....158..122K,2024A&A...681A..13F}, which is within the
-1$\sigma$ statistical uncertainties of our $t_{\rm gyro}$
+\citep{2019AJ....158..122K,2024A&A...681A..13F}, which is within 
+1$\sigma$ of our $t_{\rm gyro}$
 measurements.  Ongoing work by \citet{Curtis2024} broadly finds that
@@
 Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
-We report rotation-based ages for these systems of
+We derive rotation-based ages for these systems of
 \kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
@@
 
-We compared the independent rotation and lithium ages of the
-known planet hosts using the point-estimates of the available
-posteriors.  There are three cases of interest:
-
-\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
-    if their values were consistent within 2$\sigma$, we judged them
+The comment column of Table~\ref{tab:planets} compares
+the independent rotation and lithium ages of the
+known planet hosts using the point-estimates of their age posteriors.
+There are three cases of interest.
+{\it 1i)} If two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$ existed,
+and their values were consistent within 2$\sigma$, we judged them
 to be consistent; if they were consistent within 2-3$\sigma$, we
-    suggest them to be ``maybe'' consistent.  Otherwise, they
-    disagree.
-    %
-  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
+listed them as ``maybe'' consistent.  Otherwise, they
+disagreed.
+{\it ii)}
+If $t_{\rm Li}$ was a lower limit, we compared this lower limit
 with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
-    overlapped, we judge the age estimates to be consistent.
-    %
-  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
-    rotation is reported, we judge the age estimates to be
-    inconsistent.  This is because this lithium constraint can only be
+overlapped, we judged the age estimates to be consistent.
+{\it iii)}
+If $t_{\rm Li}$ provided a two-sided posterior, and no
+rotation period was found, we judged the age estimates to be
+inconsistent.  This is because a two-sided lithium constraint can only be
 provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
 have been sensitive to the rotation periods of such stars.
-\end{enumerate}
+
+%\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
+%  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
+%    if their values were consistent within 2$\sigma$, we judged them
+%    to be consistent; if they were consistent within 2-3$\sigma$, we
+%    suggest them to be ``maybe'' consistent.  Otherwise, they
+%    disagree.
+%    %
+%  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
+%    with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
+%    overlapped, we judge the age estimates to be consistent.
+%    %
+%  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
+%    rotation is reported, we judge the age estimates to be
+%    inconsistent.  This is because this lithium constraint can only be
+%    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
+%    have been sensitive to the rotation periods of such stars.
+%\end{enumerate}
 
@@
 
-Classifying Kepler stars as thin vs.~thick disk members helps connect them to 
-previous work on the star formation history of the
-Galaxy.  On extragalactic scales, the star formation rate in spiral
-galaxies with mass similar to the Milky Way peaked $\approx$10\,Gyr
-ago, and has since decreased by an order of magnitude
+Classifying Kepler stars as thin vs.\ thick disk members helps connect
+them to previous work on the star formation history of the Galaxy.  On
+extragalactic scales, the star formation rate in spiral galaxies with
+mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
+since decreased by an order of magnitude
 \citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
@@
 we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
-data in particular are consistent, particularly after accounting for the
-statistical uncertainties reported in that study.  Regarding the
+data in particular are consistent, particularly after accounting for
+the statistical uncertainties reported in that study.  Regarding the
 episodic star formation bursts reported by
@@
 $\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
-$\approx$1\,Gyr spike however is not recovered.
-
-% compare against Berger+20, note their sentence
-% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al.
-(2018) and Pinsonneault et al.
-(2018), as well as the rotation-based ages in Claytor et al.
-(2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al.
-(2019)."
-%
-% also compare against Mor2019 and their literature assessment
-% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al.
-(2015), and Haywood et al.
-(2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation.
-Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly.
-The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al.
-(2002) and Cignoni et al.
-(2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively.
-Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."
-
-%\subsection{Isochrone Age Comparison}
-
-%\subsection{Lithium Age Comparison}
+$\approx$1\,Gyr spike however is not recovered.  This
+$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
+ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
+in the red giant asteroseismic ages derived by
+\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
+those studies is their accuracy at $<$1\,Gyr.
 
@@
   finite reported gyrochrone ages.  A random set of these stars is
-  shown here to give an idea for form and content.  The quality
+  shown to give an idea for form and content.  The quality
   bitmask is as in Table~\ref{tab:planets}; requiring bits zero to

### 2024-06-03 drivers/make_star_table.py
minor

@@ -128,12 +128,12 @@ def make_star_table(
     _sdf = sdf[mapdict.values()]
 
     latexpath = join(PAPERDIR, 'table_star_gyro.tex')
-    np.random.seed(42)
+    np.random.seed(123)
 
     # format ages
     sel = sdf.gyro_median.astype(float) > 4000
     _sdf.loc[sel, 't_gyro'] = '$> 4000$'
-    _sdf.sample(n=10).to_latex(
+    _sdf.sample(n=5).to_latex(
         latexpath, index=False, na_rep='--', escape=False, formatters=formatters
     )
 

### 2024-06-03 gyrojo/plotting.py

@@ -67,6 +67,10 @@ from aesthetic.plot import set_style, savefig
 
 from cdips.utils.gaiaqueries import propermotion_to_kms
 
+from astroquery.vizier import Vizier
+Vizier.ROW_LIMIT = -1
+
+
 ###########
 # helpers #
 ###########
@@ -1693,6 +1697,20 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
         axs[2].plot(
             x_new, y_new/factor, c='k', lw=0.5, zorder=-10, ls=':',
             label=l2_4, alpha=0.7
+        )
+
+    SHOW_BERGER2020 = 0
+    if SHOW_BERGER2020:
+        # Berger+2020: Gaia-Kepler stellar properties catalog.
+        fitspath = join(DATADIR, 'literature', 'Berger_2020_t2_secret.fits')
+        hl = fits.open(fitspath)
+        b20t2_df = Table(hl[1].data).to_pandas()
+        #NOTE: might want to resample uncs???
+        heights, _, _ = axs[2].hist(
+            b20t2_df.Age, bins=bins, histtype='step',
+            weights=np.ones(len(b20t2_df))*0.06/5531,
+            color='gray', alpha=0.9,
+            zorder=-2, label='Berger+20'
         )
 
 

### 2024-06-03 paper/ms.tex
update ms

@@
 
-\subsubsection{Comparing Rotation and Lithium Ages}
-
-The comment column of Table~\ref{tab:planets} compares
-the independent rotation and lithium ages of the
-known planet hosts using the point-estimates of their age posteriors.
-There are three cases of interest.
-{\it 1i)} If two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$ existed,
-and their values were consistent within 2$\sigma$, we judged them
-to be consistent; if they were consistent within 2-3$\sigma$, we
+\subsubsection{Rotation vs.~Lithium Ages}
+
+We compared 
+the rotation and lithium ages for the
+known planet hosts 
+in the ``comment'' column of Table~\ref{tab:planets}.
+There were three cases of interest.
+{\it i)} If two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$ existed,
+and their median values were consistent within 2$\sigma$, we listed them
+as consistent; if they were consistent within 2-3$\sigma$, we
 listed them as ``maybe'' consistent.  Otherwise, they
@@
 inconsistent.  This is because a two-sided lithium constraint can only be
-provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
+provided for G and K dwarfs $\lesssim$2\,Gyr old, and Kepler should
 have been sensitive to the rotation periods of such stars.
@@
 
-Let us restrict discussion to blemish-free stars with 
+Let us focus discussion on blemish-free stars with 
 confirmed planets: $Q_{\rm star}$ must not have any of bits zero

### 2024-06-04 paper/ms.tex
update ms

@@
 and also for the broader set of all Kepler target stars
-\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.  These
-studies used a range of detection methods and selection functions.  We
-are interested in understanding the age distribution of all Kepler
-target stars irrespective of KOI status.  Out of these studies, the
-only homogeneous analysis of both Kepler targets and KOIs were the
-studies by \citet{Santos_2019,Santos_2021}, hereafter \citetalias{Santos_2019}
-and \citetalias{Santos_2021}.  \citetalias{Santos_2019} and
+\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021,Reinhold2023}.
+These studies used a range of detection methods and selection
+functions.  We are interested in understanding the age distribution of
+all Kepler target stars irrespective of KOI status.  Out of these
+studies, the most homogeneous and precise analyses of both Kepler
+targets and KOIs appear to be those by
+\citet{Santos_2019,Santos_2021}, hereafter \citetalias{Santos_2019}
+and \citetalias{Santos_2021}.  
+For a discussion of the work by \citet{Reinhold2023}, please see
+Appendix~\ref{app:reinhold23}.
+\citetalias{Santos_2019} and
 \citetalias{Santos_2021} combined a wavelet analysis and
@@
 \appendix
+
+\section{Regarding The Preferred Rotation Period Catalog}
+\label{app:reinhold23}
+
+While we opted for our default stellar rotation periods to be those
+reported by \citetalias{Santos_2019} and \citetalias{Santos_2021},
+the more recent analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023} used a
+similar selection function as \citetauthor{Santos_2021}, and
+considered a novel period measurement approach based on gradients of
+power spectra (GPS).
+Their method likely provides greater completeness, due
+to its greater sensitivity to signals that are only weakly periodic.
+However,
+ comparing 
+overlapping stars from
+\citet{McQuillan_2014}, \citetalias{Santos_2019},
+\citetalias{Santos_2021}, and \citet{Reinhold2023} suggested that for
+stellar signals that are strongly periodic, 
+the GPS method is likely a factor of a few times less precise than 
+the more traditional approaches (ACF, Lomb-Scargle, or wavelet
+analysis).
+
+{\bf TODO: finish writing!}
+{\bf TODO: finish writing!}
+{\bf TODO: finish writing!}
+
+%TODO: how many "new rotatoin periods" would this reinhold catalog
+%yield?
+%TODO: maybe you should just report ages for them!!
+
 \section{Defining An ``Isochronally Far From The Main Sequence'' Boundary}

### 2024-06-04 paper/ms.tex
update ms

@@
 
+\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
+\affiliation{\cmu}
+
 \author{and to be confirmed:}
 \affiliation{\caltech}
-
-\author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
-\affiliation{\cmu}
 
@@
 
-\section{Regarding The Preferred Rotation Period Catalog}
+\section{The Preferred Rotation Period Catalog}
 \label{app:reinhold23}
 
-While we opted for our default stellar rotation periods to be those
-reported by \citetalias{Santos_2019} and \citetalias{Santos_2021},
+While we opted for the 
+\citetalias{Santos_2019} and \citetalias{Santos_2021}
+rotation period catalogs,
 the more recent analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023} used a
-similar selection function as \citetauthor{Santos_2021}, and
+similar selection function as \citeauthor{Santos_2021}, and
 considered a novel period measurement approach based on gradients of
@@
 stellar signals that are strongly periodic, 
-the GPS method is likely a factor of a few times less precise than 
+the GPS method is a few times less precise than 
 the more traditional approaches (ACF, Lomb-Scargle, or wavelet
@@
 The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
-We do report rotation-based ages for stars
-above this orange locus;
-they are simply flagged as not being near the main sequence in the $\log g$ vs.~$T_{\rm eff}$ plane.
+While we do report rotation-based ages for stars
+above this orange locus,
+they are flagged as not being near the main sequence in the $\log g$ vs.~$T_{\rm eff}$ plane.
 
@@
 
-Kepler~411, a $\approx$780$\pm$170\,Myr early K dwarf with three
+Kepler~411, a $t_{\rm gyro}$$\approx$780$\pm$170\,Myr early K dwarf with three
 transiting planets at 3.0, 7.8, and 58\,days, has a rare

### 2024-06-04 paper/ms.tex
update ms

@@
 \citetalias{Santos_2021} periods with those reported by
-\citet{McQuillan_2014}.  We found that for $P_{\rm
-rot}\lesssim$15\,days, the two datasets agree at a precision of
-$\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
-rot}\approx$30\,days, the agreement was typically at the
-$\lesssim$0.03$P_{\rm rot}$ level, and the upper envelope of the
+\citet{McQuillan_2014} (see Appendix~\ref{app:reinhold23}).  We found
+that for $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a
+precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of
+$P_{\rm rot}\approx$30\,days, the agreement was typically at the
+$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the
 period difference increased roughly linearly with period.  Based on
@@
 
-While we opted for the 
-\citetalias{Santos_2019} and \citetalias{Santos_2021}
-rotation period catalogs,
-the more recent analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023} used a
-similar selection function as \citeauthor{Santos_2021}, and
-considered a novel period measurement approach based on gradients of
-power spectra (GPS).
-Their method likely provides greater completeness, due
-to its greater sensitivity to signals that are only weakly periodic.
-However,
- comparing 
-overlapping stars from
-\citet{McQuillan_2014}, \citetalias{Santos_2019},
-\citetalias{Santos_2021}, and \citet{Reinhold2023} suggested that for
-stellar signals that are strongly periodic, 
-the GPS method is a few times less precise than 
-the more traditional approaches (ACF, Lomb-Scargle, or wavelet
-analysis).
-
-{\bf TODO: finish writing!}
-{\bf TODO: finish writing!}
-{\bf TODO: finish writing!}
-
-%TODO: how many "new rotatoin periods" would this reinhold catalog
-%yield?
-%TODO: maybe you should just report ages for them!!
+\begin{figure*}[!b]
+  \begin{center}
+    \leavevmode
+      \includegraphics[width=0.48\textwidth]{perioddiff_vs_period_diffProt-m14_Prot_vs_Prot.pdf}
+      \includegraphics[width=0.48\textwidth]{perioddiff_vs_period_diffProt-r23_ProtGPS_vs_Prot.pdf}
+  \end{center}
+  \vspace{-0.6cm}
+  \caption{
+    {\bf Comparison of reported literature stellar rotation periods}.
+    ``Santos'' refers to the concatenation of \citet{Santos_2019} and
+    \citet{Santos_2021}.  ``M14'' refers to \citet{McQuillan_2014}.
+    ``R23$_{\rm GPS}$'' refers to the Gradient of Power Spectrum
+    periods from \citet{Reinhold2023}.  The background is a 2-D
+    histogram with a logarithmic color stretch that counts overlapping
+    stars between these studies.  The black errorbars show the median
+    and $\pm$1$\sigma$ range of period the differences in 1-day bins.
+    The
+    solid gray ``empirical $\sigma_{\rm P}/P$'' line in the left panel
+    shows the expected $\pm 1\sigma$ range that would be spanned if
+    the indepndent $P_{\rm rot}$ measurements had gaussian
+    uncertainties drawn following the empirical estimate described in
+    Section~\ref{subsec:rotsel}.
+    Note the different vertical scales.
+    \label{fig:perioddiff}
+  }
+\end{figure*}
+
+While we opted for the \citetalias{Santos_2019} and
+\citetalias{Santos_2021} rotation period catalogs, the more recent
+analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
+used a similar selection function as \citeauthor{Santos_2021}
+(i.e.~all Kepler stars), and considered a novel period measurement
+approach based on gradients of power spectra (GPS).  Their method
+likely provides greater completeness, due to its greater sensitivity
+to signals that are only weakly periodic.
+
+Figure~\ref{fig:perioddiff} shows histograms of the differences
+between reported periods from these various studies.  The left panel
+compares overlapping stars from \citet{McQuillan_2014} and
+the \citeauthor{Santos_2021} studies.  The periods agree at a precision of
+$\lesssim$0.01$P_{\rm rot}$ for $P_{\rm rot}\lesssim$15\,days, and at
+$\lesssim$0.03$P_{\rm rot}$ for $P_{\rm rot}\approx$30\,days.  
+A small bias exists at $P_{\rm rot}$$\approx$10\,days,
+in the sense that the Santos periods tend to be $\approx$1\% faster
+for such stars.
+
+The right panel compares overlapping periods from the
+\citet{Reinhold2023} GPS method and the \citeauthor{Santos_2021} studies.  The
+periods agree at a precision of $\lesssim$0.20$P_{\rm rot}$ for
+$P_{\rm rot}\lesssim$15\,days, and at $\lesssim$0.30$P_{\rm rot}$ for
+$P_{\rm rot}\approx$30\,days.  A  bias develops past $P_{\rm
+rot}\gtrsim$30\,days, in the sense that the \citeauthor{Santos_2021}
+periods tend to be 10-20\% longer than the GPS periods.  The origin of
+this larger scatter could be connected to the need for the GPS method
+to determine the ``$\alpha$ calibration factor'', which is known only
+at a statistical level for large stellar populations
+\citep[see][]{Reinhold2023}.  While we considered deriving independent
+rotation-based ages using the \citet{Reinhold2023} GPS periods with an
+inflated empirical uncertainty, we ultimately opted against this path.
+A rotation period difference of 20-30\% away from the ``true period''
+could sufficiently skew a star's age that we prefer to restrict our
+attention to the stars for which precision is a possible
+outcome.
 

### 2024-06-04 paper/collect.sh

@@ -39,3 +39,6 @@ cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
 cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_linear.png .
 
 cp $r/lit_ages_vs_cluster_ages/olympic_merged_lit_ages_vs_cluster_ages.pdf .
+
+cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-m14_Prot_vs_Prot.pdf .
+cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-r23_ProtGPS_vs_Prot.pdf .

### 2024-06-04 drivers/plot_perioddiff_vs_period.py
add drivers/plot_perioddiff_vs_period.py

@@ -1,0 +1,14 @@
+import os
+from gyrojo.paths import DATADIR, RESULTSDIR
+import gyrojo.plotting as ap
+
+outdir = os.path.join(RESULTSDIR, 'perioddiff_vs_period')
+if not os.path.exists(outdir): os.mkdir(outdir)
+
+ap.plot_perioddiff_vs_period(outdir, ykey='m14_Prot', ylim=[-4.3, 4.3],
+                             dx=0.25, dy=0.1)
+
+ap.plot_perioddiff_vs_period(outdir, ykey='r23_ProtGPS', dx=0.5, dy=0.2)
+
+ap.plot_perioddiff_vs_period(outdir, ykey='r23_ProtFin', dx=0.5, dy=0.2)
+

### 2024-06-05 paper/ms.tex
update ms

@@
   sequence (see Section~\ref{subsec:flags}).  Some
-  are ``confirmed'' or ``candidate'' KOIs that meet an additional set
-  of planetary quality criteria (yellow points;
+  host ``confirmed'' or ``candidate'' KOIs that meet additional
+  planetary quality criteria (yellow points;
   Section~\ref{subsec:plflags}).  Surface gravities and effective
@@
 To assess the statistical uncertainties of these rotation periods, we
-compared our adopted \citetalias{Santos_2019} and
-\citetalias{Santos_2021} periods with those reported by
-\citet{McQuillan_2014} (see Appendix~\ref{app:reinhold23}).  We found
+compared our adopted periods with those reported by
+\citet{McQuillan_2014}.  The details are in Appendix~\ref{app:reinhold23}.  We found
 that for $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a
@@
 \label{sec:liage}
-\subsection{Sample Summary}
-\label{subsec:tally}
-
-The rotation periods for the star and planet samples are shown in
-Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
-observed by Kepler, requiring reported rotation periods and a reliable
-DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
-these stars to have temperatures of 3800--6200\,K, yields
-\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
-apparently single, near the main sequence, with $\log g$$>$4.2,
-$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
-\nuniqstarsantosallbutruwe\ rotators potentially amenable for
-gyrochronology.  Finally, imposing RUWE$<$1.4 and requiring no
-possible blends ($\Delta G$<2.5) within 4$''$ leaves
-\nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
-to be valid.
-
-In the planet sample, the cumulative KOI table included
-\nnonfpcumkois\ confirmed and candidate planets, orbiting
-\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
-cuts, save for that on RUWE.  We can then require these confirmed and
-candidate planets to have MES$>$10.  This combination of selection
-criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
-rotation-based ages orbiting
-\nplhoststarwgyroagewithgrazingandhighruwe\ stars
-(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
-If we additionally require the planets to be non-grazing ($b<0.8$), we
-are left with \nplwgyroagenograzing\ planets orbiting
-\nplhoststarwgyroagenograzing\ stars.
 
@@
 \texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
-a purely empirical interpolation approach.  \texttt{EAGLES} 
+an empirical interpolation approach.  \texttt{EAGLES} 
 was calibrated on lithium measurements of stars observed by the
@@
 quote the 3$\sigma$ (99.7$^{\rm th}$ percentile) limits.
+
+%\subsection{Sample Summary}
+%\label{subsec:tally}
+%
+%The rotation periods for the star and planet samples are shown in
+%Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
+%observed by Kepler, requiring reported rotation periods and a reliable
+%DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
+%these stars to have temperatures of 3800--6200\,K yields
+%\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
+%apparently single, near the main sequence, with $\log g$$>$4.2,
+%$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
+%\nuniqstarsantosallbutruwe\ rotators.
+%Finally, imposing the full set of quality cuts,
+%i.e.~further requiring RUWE$<$1.4 and requiring no
+%resolved blends ($\Delta G$<2.5) within 4$''$ leaves
+%\nuniqstarsantosrotgyroappl\ stars for which we expect rotation-based
+%age dating to be secure.
+%
+%
+%In the planet sample, the cumulative KOI table included
+%\nnonfpcumkois\ confirmed and candidate planets, orbiting
+%\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
+%cuts, save for that on RUWE.  We can then require these confirmed and
+%candidate planets to have MES$>$10.  This combination of selection
+%criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
+%rotation-based ages orbiting
+%\nplhoststarwgyroagewithgrazingandhighruwe\ stars
+%(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
+%If we additionally require the planets to be non-grazing ($b<0.8$), we
+%are left with \nplwgyroagenograzing\ planets orbiting
+%\nplhoststarwgyroagenograzing\ stars with rotatio-based ages.
+%This cut drops $\approx$41 more planets than one would expect if the initial sample
+%of \nplwgyroagewithgrazingandhighruwe\ planets had zero false positives.
+%The binomial probability that this could happen by coincidence is 0.1\%.
 
@@
 provides an age constraint while $t_{\rm Li}$ is one-sided.
+%FIXME TODO GET MORE PRECISE LANGUAGE -- SYSTEMS OR PLANETS OR STARS?
 
@@
 stellar rotation periods, lithium equivalent widths, and temperatures
-using archival and new data.  We derived new ages using empirical
+using archival and new data from the Kepler field.  We derived new ages using empirical
 interpolation-based methods, and assessed the reliability of these ages
@@
 mostly mini-Neptunes, some are near the lower boundary of the
-``sub-Jovian desert'' \citep{Owen2018}, and others include Earth-sized 
-planets a few hundred million years old.
+``sub-Jovian desert'' \citep{Owen2018}, which could have an evolutionary
+connection to their youth.
+Other discoveries include Earth-sized 
+planets with new ages of only a few hundred million years.
 
@@
     from CMD fitting and white-dwarf chronology, though with
-    important nuances.
+    some nuances.
+    One important advance is that our
+    rotation-based ages offer better accuracy at $t$$<$1\,Gyr
+    than current isochronal methods (Figure~\ref{fig:agescalecompone}).
   %
@@
   %FIXME: 120 number
-  \item Using rotation-based ages, we have derived ages for
+  \item We have derived rotation-based ages for
     \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
@@
 
-\section{The Preferred Rotation Period Catalog}
+\section{Rotation Period Catalog Comparison}
 \label{app:reinhold23}
@@
 While we opted for the \citetalias{Santos_2019} and
-\citetalias{Santos_2021} rotation period catalogs, the more recent
+\citetalias{Santos_2021} rotation period catalogs, the recent
 analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
 used a similar selection function as \citeauthor{Santos_2021}
-(i.e.~all Kepler stars), and considered a novel period measurement
-approach based on gradients of power spectra (GPS).  Their method
+(all Kepler stars), and considered a novel period measurement
+approach based on the gradient of the power spectrum (GPS).  Their method
 likely provides greater completeness, due to its greater sensitivity

### 2024-06-05 gyrojo/plotting.py
fig1: bugfix numbers

@@ -56,7 +56,8 @@ from astropy.io import fits
 from gyrojo.paths import DATADIR, RESULTSDIR, LOCALDIR, CACHEDIR
 from gyrojo.getters import (
     get_gyro_data, get_li_data, get_age_results,
-    get_kicstar_data, get_koi_data, get_prot_metacatalog
+    get_kicstar_data, get_koi_data, get_prot_metacatalog,
+    select_by_quality_bits
 )
 from gyrojo.papertools import update_latex_key_value_pair as ulkvp
 
@@ -2241,9 +2242,12 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G', vtangcut=None):
 
     fig, ax = plt.subplots(figsize=(2.5,2.7))
 
+    # Stars with Santos+ rotation period reported (...and good Gaia matches)
+    sel = select_by_quality_bits(kicdf, [4, 5], [0, 0])
+
     dfs = [
        #cgk_df,
-        kicdf,
+        kicdf[sel],
         kicdf[kicdf['flag_is_gyro_applicable']],
         koidf
     ]
@@ -2270,7 +2274,7 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G', vtangcut=None):
         #'KIC',
         'Has Prot',
         '...& gyro applicable',
-        '...& KOI',
+        '...& KOI host',
     ]
     rasterized = [
         #True,
@@ -2294,17 +2298,18 @@ def plot_st_params(outdir, xkey='dr3_bp_rp', ykey='M_G', vtangcut=None):
 
         yval = nparr(df[ykey])
         xval = nparr(df[xkey])
+        kicids = nparr(df['KIC'])
         if _i != len(dfs) - 1:
             ax.scatter(
                 xval, yval,
                 marker='o', c=c, zorder=z, s=s, linewidths=0,
-                label=l+f" ($N$={len(xval)})", rasterized=r
+                label=l+f" ($N$={len(np.unique(kicids))})", rasterized=r
             )
         else:
             ax.scatter(
                 xval, yval,
                 marker='o', c=c, zorder=z, s=s, linewidths=0.1,
-                label=l+f" ($N$={len(xval)})", rasterized=r,
+                label=l+f" ($N$={len(np.unique(kicids))})", rasterized=r,
                 edgecolors='k'
             )
 

### 2024-06-05 paper/ms.tex
update ms

@@
   This is about half the number expected under the classic assumption
-  of a uniform star formation history.  We find that this dearth can
+  of a uniform star formation history.  We find that the dearth can
   be attributed to the star formation rate in the Kepler field
@@
     exoplanets younger than one billion years.} Stars denote
-    \nplyounggyrotwosigmanograzingnoruwe\ planets that are newly
+    \nplyounggyrotwosigmanograzingnoruwe\ planets that are 
     characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
@@
     0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-    hosts to have no quality flags raised, and for the planets to
-    similarly no have quality flags raised. 
+		hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the planets 
+		to similarly no have quality flags raised. 
     \label{fig:rp_period_age_results}
@@
 else unresolved binaries, and are automatically flagged as such.  The
-more interesting of the two is Kepler-786,
-an early K dwarf with $t_{\rm
-Li}$=\kepseveneightsix\,Myr, which would imply that rotation should be
-detectable.  However no photometric rotation signal is detected, and other
-age indicators in the spectrum similarly suggest that the star is not
-young.
+more interesting of the two is Kepler-786, an early K dwarf with
+$t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
+rotaton period that would imply $t_{\rm
+gyro}$=\kepseveneightsixgyro\,Myr.  Other age indicators in the
+spectrum similarly suggest that the star is not young.
 
@@
 
-Our analysis blindly recovered the ages of all 14 known Kepler planets
+Our analysis recovered the ages of all 14 known Kepler planets
 in clusters.  While lithium provided only marginal added information

### 2024-06-05 paper/ms.tex
number fixing

@@
 Let us focus discussion on blemish-free stars with 
-confirmed planets: $Q_{\rm star}$ must not have any of bits zero
-through nine raised, and $Q_{\rm planet}$ must also have no quality flags
+blemish-free planets: $Q_{\rm star}$ must not have any of bits zero
+through nine raised, and $Q_{\rm planet}$ must similarly have no quality flags
 raised.  Under this constraint, our results include
-\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
+\ltonegyrhighqconfirmedtwosided\ confirmed planets younger than 1\,Gyr with
 two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
-\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
+\ltonegyrhighqconfirmedonesided\ confirmed planets with comparable
+ages for which $t_{\rm gyro}$
 provides an age constraint while $t_{\rm Li}$ is one-sided.
-%FIXME TODO GET MORE PRECISE LANGUAGE -- SYSTEMS OR PLANETS OR STARS?
-
-These planets are shown in Figure~\ref{fig:rp_period_age_results}.
-They include \nearthshighqconfirmed\ Earth-sized or smaller planets,
-no Jupiter-sized or larger planets, \nsubsaturnshighqconfirmed\
+
+Figure~\ref{fig:rp_period_age_results} shows
+the planets and planet candidates with $t_{\rm gyro}$$<$1\,Gyr at
+2$\sigma$.  
+This figure includes \nconfirmedplyounggyrotwosigmanograzingnoruwe\
+confirmed Kepler planets, and
+\ncandidateplyounggyrotwosigmanograzingnoruwe\ candidate planets, all of which
+are listed in Table~\ref{tab:planets}.
+These include
+\nearthshighqconfirmed\ Earth-sized or smaller planets,
+one Jovian planet candidate, \nsubsaturnshighqconfirmed\
 ``super-Neptune'' ($>$4\,$R_\oplus$) sized planets,

### 2024-06-05 drivers/get_values_for_manuscript.py

@@ -236,6 +236,22 @@ sel = (df['gyro_median'] + df['gyro_+2sigma']  <= 1000)
 N = len(df[sel])
 ulkvp('nplyounggyrotwosigmanograzingnoruwe', N)
 
+sel = (
+	(df['gyro_median'] + df['gyro_+2sigma']  <= 1000)
+	&
+	(df.koi_disposition == 'CONFIRMED')
+)
+N = len(df[sel])
+ulkvp('nconfirmedplyounggyrotwosigmanograzingnoruwe', N)
+
+sel = (
+	(df['gyro_median'] + df['gyro_+2sigma']  <= 1000)
+	&
+	(df.koi_disposition == 'CANDIDATE')
+)
+N = len(df[sel])
+ulkvp('ncandidateplyounggyrotwosigmanograzingnoruwe', N)
+
 
 
 ##########################################

### 2024-06-05 paper/ms.tex
update ms

@@
 %\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
-\title{A Dearth of Young Stars and Planets in the Kepler Field}
+%\title{A Cliff In The Age Distribution of Young Stars and Planets in the Kepler Field}
+\title{A Paucity of Young Stars and Planets in the Kepler Field}
 
@@
   This is about half the number expected under the classic assumption
-  of a uniform star formation history.  We find that the dearth can
+  of a uniform star formation history.  We find that the scarcity of sub-gigayear
+  systems can
   be attributed to the star formation rate in the Kepler field
@@
   This ``demographic cliff'' in the Galaxy's star formation history
-  has been previously reported, and it helps clarify the relative
-  balance of young and old planets that have been detected to date
+  has been previously reported, and it helps clarify the abundance
+  of young and old planets that have been detected to date
   using transits.
@@
 
-Exoplanet science is enjoying an age of abundance, fueled by the
+Exoplanet science is thriving, fueled by the
 discovery of thousands of worlds orbiting close to their host stars
-\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
+\citep{Borucki10,2015JATIS...1a4003R}.  However, most known exoplanets are
 billions of years old.  This fact leaves many gaps in our knowledge of
-the exact physical and dynamical histories of these objects.  The
+the exact physical and dynamical origins of these objects.  The
 reason is that processes such as thermal cooling

### 2024-06-05 paper/ms.tex
update ms

@@
 
-\subsection{A Dearth Of Young Stars And Planets}
+\subsection{A Paucity Of Young Stars And Planets}
 
@@
 interpolation-based method is well-calibrated.  The overall slope
-yields a dearth of young stars relative to the expectation of a
+yields a paucity of young stars relative to the expectation of a
 uniform age distribution.
@@
 
-Let us focus discussion on blemish-free stars with 
+Let us focus on blemish-free stars with 
 blemish-free planets: $Q_{\rm star}$ must not have any of bits zero
@@
 ages for which $t_{\rm gyro}$
-provides an age constraint while $t_{\rm Li}$ is one-sided.
+is two-sided while $t_{\rm Li}$ is one-sided.
 
@@
 These include
-\nearthshighqconfirmed\ Earth-sized or smaller planets,
-one Jovian planet candidate, \nsubsaturnshighqconfirmed\
-``super-Neptune'' ($>$4\,$R_\oplus$) sized planets,
-\nminineptuneshighqconfirmed\ mini-Neptunes above the radius valley
+\nearthshighq\ objects Earth-sized or smaller,
+\njupitershighq\ Jovian-sized planet candidates, \nsubsaturnshighq\
+super-Neptunes (4--10\,$R_\oplus$),
+\nminineptuneshighq\ mini-Neptunes above the radius valley
 as parametrized by \citealt{2018MNRAS.479.4786V}, and
-\nsuperearthshighqconfirmed\ super-Earth sized planets below the
+\nsuperearthshighq\ super-Earth sized planets below the
 radius valley.  While most of these planets are on orbital periods
-below 50\,days, \nlongperiodhighqconfirmed\ are on more distant
+below 50\,days, \nlongperiodhighq\ are on more distant
 orbits.
 
-A few highlights from these systems include
+Highlights from these systems include
 Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
@@
 Kepler-1561~b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
-youngest Earth-sized planets currently known, comparable to systems
+youngest Earth-sized planets currently known, comparable to planets
 such as HD~63433d \citep[1.1\,$R_\oplus$,

### 2024-06-05 paper/ms.tex
tweak table 1

@@
   version of the table is truncated to include the youngest
-  confirmed Kepler planets, sorted by the minimum of either the
+  systems, sorted by the minimum of either the
   rotation or lithium-based age.  The full machine-readable table
@@
   is {\bf \href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{also available}}.
-  All quoted uncertainties are statistical; they do not incorporate
-  possible systematic uncertainties, e.g.~in the absolute ages of open clusters.
+  All quoted age uncertainties are statistical.
   \label{tab:planets}}
@@
 % \rule
-%  \input{table_subgyr_candidate}
+  \input{table_subgyr_candidate}
   \enddata
@@
   	EW$_{\rm Li}^{\rm \ast}$ is the lithium equivalent width {\it after} subtracting a constant 7.5\,m\AA\ to account for the \ion{Fe}{1} 6707.44\,\AA\ blend (see Section~\ref{subsec:lithiumsel}).
-  	Two checkmarks ($\checkmark\checkmark$) denote systems with two-sided $t_{\rm
-  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are expected to be reliable.  One checkmark ($\checkmark$) denotes two-sided $t_{\rm gyro}$ only. 
+  	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
+  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are expected to be reliable.  One checkmark ($\checkmark$) denotes confirmed planets iwth two-sided $t_{\rm gyro}$ only. 
   ``Confirmed'' planets appear in the machine-readable

### 2024-06-05 drivers/make_younghighlight_table.py

@@ -400,13 +400,13 @@ def make_table(
     if SELECT_YOUNG:
         sel = (pdf.koi_disposition == 'CONFIRMED')
         latexpath1 = join(PAPERDIR, 'table_subgyr_confirmed.tex')
-        pdf[sel][mapdict.values()].head(n=95).to_latex(
+        pdf[sel][mapdict.values()].head(n=85).to_latex(
             latexpath1, index=False, na_rep='--', escape=False, formatters=formatters
         )
 
         sel = (pdf.koi_disposition == 'CANDIDATE')
         latexpath2 = join(PAPERDIR, 'table_subgyr_candidate.tex')
-        t2 = pdf[sel][mapdict.values()].head(n=20).to_latex(
+        t2 = pdf[sel][mapdict.values()].head(n=10).to_latex(
             latexpath2, index=False, na_rep='--', escape=False, formatters=formatters
         )
 

### 2024-06-05 paper/ms.tex
update ms

@@
 
-Let us focus on blemish-free stars with 
-blemish-free planets: $Q_{\rm star}$ must not have any of bits zero
-through nine raised, and $Q_{\rm planet}$ must similarly have no quality flags
-raised.  Under this constraint, our results include
-\ltonegyrhighqconfirmedtwosided\ confirmed planets younger than 1\,Gyr with
-two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
-\ltonegyrhighqconfirmedonesided\ confirmed planets with comparable
-ages for which $t_{\rm gyro}$
-is two-sided while $t_{\rm Li}$ is one-sided.
-
-Figure~\ref{fig:rp_period_age_results} shows
-the planets and planet candidates with $t_{\rm gyro}$$<$1\,Gyr at
-2$\sigma$.  
-This figure includes \nconfirmedplyounggyrotwosigmanograzingnoruwe\
-confirmed Kepler planets, and
-\ncandidateplyounggyrotwosigmanograzingnoruwe\ candidate planets, all of which
-are listed in Table~\ref{tab:planets}.
-These include
-\nearthshighq\ objects Earth-sized or smaller,
+Let us focus on blemish-free stars with blemish-free planets: $Q_{\rm
+star}$ must not have any of bits zero through nine raised, and $Q_{\rm
+planet}$ must similarly have no quality flags raised.  Under this
+constraint, our results include \ltonegyrhighqconfirmedtwosided\
+confirmed planets younger than 1\,Gyr with two-sided $t_{\rm gyro}$
+and $t_{\rm Li}$, and \ltonegyrhighqconfirmedonesided\ confirmed
+planets with comparable ages for which $t_{\rm gyro}$ is two-sided
+while $t_{\rm Li}$ is one-sided.
+
+Figure~\ref{fig:rp_period_age_results} shows the planets and planet
+candidates with $t_{\rm gyro}$$<$1\,Gyr at 2$\sigma$.  This figure
+includes \nconfirmedplyounggyrotwosigmanograzingnoruwe\ confirmed
+Kepler planets, and \ncandidateplyounggyrotwosigmanograzingnoruwe\
+candidate planets, all of which are listed in Table~\ref{tab:planets}.
+These include \nearthshighq\ objects Earth-sized or smaller,
 \njupitershighq\ Jovian-sized planet candidates, \nsubsaturnshighq\
-super-Neptunes (4--10\,$R_\oplus$),
-\nminineptuneshighq\ mini-Neptunes above the radius valley
-as parametrized by \citealt{2018MNRAS.479.4786V}, and
-\nsuperearthshighq\ super-Earth sized planets below the
-radius valley.  While most of these planets are on orbital periods
-below 50\,days, \nlongperiodhighq\ are on more distant
-orbits.
+super-Neptunes (4--10\,$R_\oplus$), \nminineptuneshighq\ mini-Neptunes
+above the radius valley as parametrized by
+\citealt{2018MNRAS.479.4786V}, and \nsuperearthshighq\ super-Earth
+sized planets below the radius valley.  While most of these planets
+are on orbital periods below 50\,days, \nlongperiodhighq\ are on more
+distant orbits.
 
@@
 CMD fitting.  The overall slope of both SFHs broadly agrees with what
-we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
-data in particular are consistent, particularly after accounting for
-the statistical uncertainties reported in that study.  Regarding the
+we find from rotation-based ages.  The SFH from \citet{2019A&A...624L...1M}
+seems entirely consistent, particularly after accounting for
+the statistical uncertainties of that study.  Regarding the
 episodic star formation bursts reported by
@@
 \citet{SilvaAguirre2018}.  The main novelty of our ages relative to
-those studies is their accuracy at $<$1\,Gyr.
+those studies is that they are accurate at $<$1\,Gyr.
 

### 2024-06-05 paper/ms.tex
update details

@@
 puffy planets that is a likely precusor to the compact
-multiplanet systems \citep{David_2019}, and HIP~67522, a Jupiter-sized
+multiplanet systems \citep{David_2019}, and HIP~67522b, a Jupiter-sized
 planet with a sub-Neptunian mass (\citealt{Rizzuto_2020}; Thao et
@@
 most precise for stars whose luminosities and temperatures separate
-them from the main sequence.  For $\lesssim$1\,Gyr stars, isochrone
-ages are therefore limited.
+them from the main sequence.  For 0.1--1\,Gyr stars that are on
+the main sequence, isochrone ages are therefore limited.
 
@@
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.48\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
   rotation periods are primarily drawn from
-  \citet{Santos_2019,Santos_2021}.  About half of these rotators are
+  \citet{Santos_2019,Santos_2021}.  About half of the stars with
+  rotation periods are
   suitable for gyrochronology (dark gray points), based on factors
   including their non-binarity and proximity to the main
-  sequence (see Section~\ref{subsec:flags}).  Some
+  sequence (orange line in lower panel; see Section~\ref{subsec:flags}).  Some
   host ``confirmed'' or ``candidate'' KOIs that meet additional
@@
     single stars near the main sequence with $\log g$$>$4.2,
-    RUWE$<$1.4, and temperatures of 3800--6200\,K.  The color map has
-    a linear stretch.  The planet-hosts (right) require the same
+    RUWE$<$1.4, and temperatures of 3800--6200\,K.  
+    The planet-hosts (right) require the same
     stellar cuts, and include only the confirmed and candidate planets
-    described in Section~\ref{subsec:plflags}.  An order of magnitude
+    described in Section~\ref{subsec:plflags}.  The number of planets $N_{\rm p}$
+    and number of stars $N_{\rm s}$ are noted in each panel.
+    An order of magnitude
     estimate for the number of young planets discovered by Kepler

### 2024-06-05 gyrojo/plotting.py

@@ -141,7 +141,7 @@ def plot_koi_mean_prot_teff(outdir, sampleid='koi_X_S19S21dquality',
                             grazing_is_ok=0):
     # For KOIs
 
-    df = get_gyro_data(sampleid, grazing_is_ok=grazing_is_ok)
+    df = get_gyro_data(sampleid, grazing_is_ok=grazing_is_ok, drop_highruwe=0)
 
     if 'deprecated' not in sampleid:
         assert len(df) == df['flag_is_gyro_applicable'].sum()
@@ -167,7 +167,7 @@ def plot_koi_mean_prot_teff(outdir, sampleid='koi_X_S19S21dquality',
     yvals = [9.8,14.8,16.7,20,28]
 
     _Teff = np.linspace(3800, 6200, int(1e3))
-    linestyles = ['solid', 'dotted', 'dashed', 'dashdot', 'solid']
+    linestyles = ['dotted', 'solid', 'dashed', 'dashdot', 'solid']
     for model_id, ls, age, yval in zip(model_ids, linestyles, ages, yvals):
         color = 'k'
         poly_order = 7
@@ -181,7 +181,7 @@ def plot_koi_mean_prot_teff(outdir, sampleid='koi_X_S19S21dquality',
             )
 
         ax.plot(
-            _Teff, _Prot, color=color, linewidth=1, zorder=10, alpha=0.4, ls=ls
+            _Teff, _Prot, color=color, linewidth=0.8, zorder=10, alpha=0.6, ls=ls
         )
 
         bbox = dict(facecolor='white', alpha=1, pad=0, edgecolor='white')
@@ -282,7 +282,7 @@ def plot_star_Prot_Teff(outdir, sampleid):
     yvals = [9.8,14.8,16.7,20,28]
 
     _Teff = np.linspace(3800, 6200, int(1e3))
-    linestyles = ['solid', 'dotted', 'dashed', 'dashdot', 'solid']
+    linestyles = ['dotted', 'solid', 'dashed', 'dashdot', 'solid']
     for model_id, ls, age, yval in zip(model_ids, linestyles, ages, yvals):
         color = 'k'
         poly_order = 7
@@ -296,7 +296,7 @@ def plot_star_Prot_Teff(outdir, sampleid):
             )
 
         ax.plot(
-            _Teff, _Prot, color=color, linewidth=1, zorder=10, alpha=0.4, ls=ls
+            _Teff, _Prot, color=color, linewidth=0.8, zorder=10, alpha=0.7, ls=ls
[…]

### 2024-06-06 paper/ms.tex
update ms

@@
 
-Discovering a young planet requires solving two problems: find the
-planet, and measure the star's age.  Each problem admits a range of
+Discovering a young planet requires solving two problems: finding the planet and measuring the star's age.
+Each problem admits a range of
 solutions
@@
 most precise for stars whose luminosities and temperatures separate
-them from the main sequence.  For 0.1--1\,Gyr stars that are on
+them from the main sequence.  For sub-gigayear stars on
 the main sequence, isochrone ages are therefore limited.
@@
 Empirical models now yield ages precise to $\lesssim$30\% for FGK stars
-between 1-4\,Gyr, and constraining age posteriors at earlier times
+between 1-4\,Gyr, and constraining age posteriors for younger ages
 \citep{Bouma_2023}.  Physics-based models
@@
     The gray lines are ``mean fits'' to the slow rotation sequences of
-    open clusters.  The stellar sample (left) includes only apparently
+    open clusters.  The 2-D histogram of the stellar sample (left) includes only apparently
     single stars near the main sequence with $\log g$$>$4.2,
@@
     stellar cuts, and include only the confirmed and candidate planets
-    described in Section~\ref{subsec:plflags}.  The number of planets $N_{\rm p}$
-    and number of stars $N_{\rm s}$ are noted in each panel.
-    An order of magnitude
+    described in Section~\ref{subsec:plflags}.  The number of stars $N_{\rm s}$ and
+    planets $N_{\rm p}$are noted in 
+    the relevant panels.
+    A simple
     estimate for the number of young planets discovered by Kepler
@@
 		{\bf Reported ages of individual stars in benchmark open clusters.}
-		Each point denotes a reported age,
+		Each point denotes a star's reported age,
 		normalized by the age of a stellar ensemble in which the star
 		is a candidate member.
-		Cluster membership was assessed in previous studies without knowledge of rotation.
+		Cluster membership was evaluated without knowledge of rotation.
 		Although some field interlopers may be present in the 
@@
     certain methods report ages for more stars than others.
-		Horizontal scatter is artificially added to visually
+		Horizontal scatter is added to visually
 		clarify the statistical age uncertainties.
+		At $<$1\,Gyr,
+		the rotation-based ages from our work provide better accuracy 
+		and better-quantified uncertainties than previous analyses.
 	}
@@
   \caption{{\bf Kepler's demographic cliff,} visible in the
-  rotation-derived ages of stars (left) and planet hosts (middle) in
-  the Kepler field.  The top row shows all stars with
-  temperatures of 3800--6200\,K for which we calculated rotation-based
+  rotation-derived age distributions of its stars (left) and planet hosts (middle).  The top row shows all stars with
+  temperatures of 3800--6200\,K for which we calculated rotational
   ages.  Opaque lines impose quality cuts on 
   binarity, crowding, and the star's evolutionary state; transparent lines do not.
-  The bottom
+  The bottom row
   shows stars with temperatures of 4400--5400\,K, which have
@@
   The statistical uncertainty for a mean
-  individual star at 1, 2, and 3\,Gyr is shown, and is identical
+  star at 1, 2, and 3\,Gyr is shown, and is identical
   across each row.  Finally, the right panel compares the
@@
     {\bf Sizes, orbital periods, and rotation-based ages of transiting
-    exoplanets younger than one billion years.} Stars denote
+    exoplanets younger than one billion years.} Star symbols denote
     \nplyounggyrotwosigmanograzingnoruwe\ planets that are 
@@
 gyrochronology quality flags discussed in Section~\ref{subsec:flags}
-yields 17{,}755 thin-disk stars for which gyrochronology is nominally
+yields 17{,}755 thin-disk stars for which gyrochronology is 
 applicable, and 2{,}462 thick-disk stars.
@@
 Classifying Kepler stars as thin vs.\ thick disk members helps connect
-them to previous work on the star formation history of the Galaxy.  On
+them to previous work on the Galaxy's star formation history.  On
 extragalactic scales, the star formation rate in spiral galaxies with
@@
 \citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
-galaxy, many investigators have worked to measure similar star
+galaxy, previous studies have measured star
 formation histories through CMD fitting of resolved stellar
@@
 local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
-focused on stars in a wider 2\,kpc bubble, and reported three local
-maxima in the thin disk's star formation history 5.7, 1.9, and
-1.0\,Gyr ago, which they associated with pericenter passages of the
+focused on stars in a wider 2\,kpc bubble, and additionally
+reported three local
+maxima in the galactic star formation rate, which they associated with pericenter passages of the
 Sagittarius satellite galaxy.
@@
 irrespective of planet-hosting status.  However it implies that the
-implied planet ages might have shift if one were to account for the
+estimated planet ages might shift if one were to account for the
 spectroscopic information.

### 2024-06-06 paper/ms.tex
update ms

@@
 Mellon University, Pittsburgh, PA 15213, USA}
+\newcommand{\osaka}{Department of Earth and Space Science, Osaka University, Osaka 560-0043, Japan}
 
@@
 %\title{A Cliff In The Age Distribution of Young Stars and Planets in the Kepler Field}
-\title{A Paucity of Young Stars and Planets in the Kepler Field}
+%\title{A Paucity of Young Stars and Planets in the Kepler Field}
+\title{Ages of Young Stars and Planets in the Kepler Field}
 
 \correspondingauthor{Luke G. Bouma}
-\email{luke@astro.caltech.edu}
+\email{bouma.luke@gmail.com}
 
@@
 
+%\author{and to be confirmed:}
+%\affiliation{\caltech}
+
+\author[0000-0002-0531-1073]{Howard Isaacson}
+\affiliation{\berkeley}
+
+%  \author[0000-0003-1298-9699]{Kento Masuda} ?
+%  \affiliation{\osaka}
+
 \author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
 \affiliation{\cmu}
-
-\author{and to be confirmed:}
-\affiliation{\caltech}
-
-\author[0000-0002-0531-1073]{Howard Isaacson}
-\affiliation{\berkeley}
 
@@
   dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3\,Gyr.  This trend is observed both for known planet hosts as well
-  as for the overall stellar sample.
+  3\,Gyr.  We observe this trend both for known planet hosts and also
+  for the parent stellar sample.
   %
   This ``demographic cliff'' in the Galaxy's star formation history
-  has been previously reported, and it helps clarify the abundance
-  of young and old planets that have been detected to date
-  using transits.
+  has been previously reported, and its confirmation helps clarify the
+  age distribution of the known transiting exoplanets.
   %
@@
 
-Exoplanet science is thriving, fueled by the
-discovery of thousands of worlds orbiting close to their host stars
-\citep{Borucki10,2015JATIS...1a4003R}.  However, most known exoplanets are
-billions of years old.  This fact leaves many gaps in our knowledge of
-the exact physical and dynamical origins of these objects.  The
-reason is that processes such as thermal cooling
+Exoplanet science is thriving, fueled by the discovery of thousands of
+worlds orbiting close to their host stars
+\citep{Borucki10,2015JATIS...1a4003R}.  However, most known exoplanets
+are billions of years old.  This fact leaves many gaps in our
+knowledge of the exact physical and dynamical origins of these
+objects.  The reason is that processes such as thermal cooling
 \citep{2007ApJ...659.1661F}, atmospheric loss
@@
 
-Young ($<$1\,Gyr) exoplanets represent one means of building
-the timeline for exoplanet evolution.  Informative individual
-exemplars include V1298~Tau, a resonant chain of 
-puffy planets that is a likely precusor to the compact
-multiplanet systems \citep{David_2019}, and HIP~67522b, a Jupiter-sized
-planet with a sub-Neptunian mass (\citealt{Rizzuto_2020}; Thao et
-al.~submitted).  Population-level analyses have similarly suggested
-differences in the size distribution of young exoplanets relative to
-their older counterparts
+Young ($<$1\,Gyr) exoplanets represent one means of building the
+timeline for exoplanet evolution.  Informative individual exemplars
+include HIP~67522b, a Jupiter-sized planet with a sub-Neptunian mass
+(\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
+resonant chain of puffy planets that is a likely precusor to the
+compact multiplanet systems \citep{David_2019}.  Population-level
+analyses have similarly suggested differences in the size distribution
+of young exoplanets relative to their older counterparts
 \citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
@@
 \citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
-clarifies the rarity of young stars: under this assumption,
+suggests that
 $\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
@@
 stars with known ages---obvious members of open clusters---and to
-search them for transiting planets.  The resulting stellar (and
-assumed planetary) ages are precise at the $\approx$10\% level.  A
-recent development, facilitated by Gaia, is the idea that finding
-transiting planets is often easier than finding the host stellar
+search them for transiting planets.  The resulting stellar, and
+assumed planetary, ages are precise at the $\approx$10\% level.  A
+recent development facilitated by Gaia is the idea that finding
+transiting planets is often easier than finding the host star's birth
 association \citep[e.g.][]{Tofflemire_2021}.
@@
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
-than 1\,Gyr.  Requiring say $t$$<$1\,Gyr at
+than 1\,Gyr.  Requiring $t$$<$1\,Gyr at
 2$\sigma$ precision gives $\approx$50 such planets at
@@
 \citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
-factors that can yield improvements, particularly for the ages of
+factors that can yield improvements, particularly for 
 sub-gigayear stars.   These factors are as follows.
@@
   %
-  \item We can fit the open cluster data, and use these fits to infer
-    ages by marginalizing over the range of ages that might explain
-    any one star's rotation period \citep{Bouma_2023}.
-    This represents an improvement in the accuracy 
-    of uncertainty propagation relative to previous calibrations.
+  \item Using open cluster data, we can marginalize over the range of
+    ages that might explain any one star's rotation period
+    \citep{Bouma_2023}.  This represents an improvement in the
+    accuracy of uncertainty propagation relative to previous
+    calibrations.
   %
@@
     \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
-    These clusters offer a sample of stars that we can use to test
+    These clusters offer stars that we can use to test
     the reliability of the available single-star age-dating methods.
@@
 \citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
-Empirical understanding has however improved due to the work by
+Empirical understanding however has improved due to the work by
 \citet{Jeffries_2023}, who modeled the time evolution of the
@@
 Section~\ref{sec:litagecomparison} using
-clusters in the Kepler field.  The population-level
-trends are described in
-Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
-are offered in Section~\ref{sec:conclusions}.
+clusters in the Kepler field.  We discuss population-level
+trends in Sections~\ref{sec:results} and~\ref{sec:disc}, and offer a
+few conclusions in Section~\ref{sec:conclusions}.
 
@@
 		\subfloat{
-			\includegraphics[width=0.47\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.43\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.47\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.43\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
     planet hosts (right)}.  
-    The gray lines are ``mean fits'' to the slow rotation sequences of
+    The gray lines are ``mean fits'' to the rotation sequences of
     open clusters.  The 2-D histogram of the stellar sample (left) includes only apparently
@@
     described in Section~\ref{subsec:plflags}.  The number of stars $N_{\rm s}$ and
-    planets $N_{\rm p}$are noted in 
+    planets $N_{\rm p}$ are noted in 
     the relevant panels.
@@
     estimate for the number of young planets discovered by Kepler
-    follows by counting planets below the mean rotational isochrones.
+    follows by counting points below the mean rotational isochrones.
 	}
@@
 is not surprising, since the rotation periods of KOIs have received
-considerably more scrutiny than those of ordinary Kepler stars.  We
+more scrutiny than those of ordinary Kepler stars.  We
 therefore decided to split our subsequent analysis into a homogeneous
@@
 additional \nnewdavidtwentyone\ rotation periods for KOI hosts that
-\citet{David_2021} had described as either ``reliable'' or ``highly
+\citet{David_2021} described as either ``reliable'' or ``highly
 reliable'' in their visual analysis of previously reported KOI
@@
 Figure~\ref{fig:prot_vs_teff}
-shows the resulting \nuniqstarsantosrot\ Kepler stars with rotation
+shows a subset of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
 periods compiled from \citetalias{Santos_2019},
@@
 ``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
-least 10.  This yielded at least one spectrum for \nstwithspec\
-stars hosting \nplwithspec\ planets.  About half of these stars have
+least 10.  This yielded at least one spectrum for 1{,}464
+stars hosting 2{,}174 planets.  About half of these stars have
 measured rotation periods (\nstwithspecandprot\ stars and
-\nplwithspecandprot, respectively).   For stars with multiple spectra
+\nplwithspecandprot\ planets, respectively).   For stars with multiple spectra
 available, we analyzed only
@@
 precision of our method, after applying an initial
-iteration on the \nlithiumstars\ KOIs with spectra, we compared our
+iteration on the \nlithiumstars\ stars with spectra, we compared our
 lithium equivalent widths with those reported by
-\citet{2018ApJ...855..115B}.  For the \nbergeroverlap\ stars in both
+\citet{2018ApJ...855..115B}.  For the stars in both
 samples, we found broad agreement at $>$30\,m\AA, and significant
@@
 \citetalias{Berger_2020a_catalog}, we adopted the values reported by
-\citet{Santos_2019} and \citet{Santos_2021}, which for these cases
-primarily derive from the \citet{Mathur_2017} DR25 Kepler Stellar
-Properties Catalog, and are mostly derived from photometry.  In the
+\citet{Santos_2019} and \citet{Santos_2021}, which are
+primarily derived from the photometric \citet{Mathur_2017} DR25 Kepler Stellar
+Properties Catalog.  In the
 planet sample, \frackoisnofpwithprotwithbtwenty\ of the
@@
 		are the ``mean models'' at various ages.  The intrinsic dispersion
-		in rotation and lithium about these mean models, which is what the
-		models fit, sets the theoretical precision floor for the
+    around the mean, which is what the models fit, sets the
+    theoretical precision floor for the
 		age-dating methods.    Additional sources of uncertainty,
 		including measurement uncertainty, impose further limits on
-		achievable precision.  These models are calibrated using open
-		clusters younger than 4\,Gyr.  The displayed points assume a
-		uniform distribution in temperature for visual clarity.  The sizes
+		achievable precision.  These models are calibrated using data from open
+		clusters.  The sizes
 		of the points are the same in each panel, so that low apparent
@@
 3800--6200\,K.  To identify stars for which we suspect these ages may not
-be valid, we built a set of quality flags which we then condensed
+be valid, we then built a set of quality flags which we condensed
 into a single binary number: $Q_{\rm star}$.  When and how
@@
 state of the star must be well-specified by its temperature and its
-age, {\it 2)} the star's spindown must not be influenced by binary
+age, {\it 2)} the star's spin-down must not be influenced by binary
 companions, and {\it 3)} the rotation period distribution for field
@@
 {\it Surface gravity (bit 1)}---We flagged stars as possible subgiants
-if they had $\log g < 4.2$.  We set this threshold by examining the
-relevant flag from \citet{berger_2018_radii_evolnstates},
-cross-matched against \citet{Berger_2020a_catalog}.
+if they had $\log g < 4.2$. 
+%We set this threshold by examining the
+%relevant flag from \citet{berger_2018_radii_evolnstates},
+%cross-matched against \citet{Berger_2020a_catalog}.
 
@@
 DR3 $G$-band luminosity, ignoring reddening, using the reported
-apparent $G$-band magnitude and parallaxes.  We flag stars with
+apparent $G$-band magnitude and parallaxes.  We flagged stars with
 $M_{\rm G} < 3.9$ or $M_{\rm G} > 8.5$, corresponding to main-sequence
@@
 
-{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- In situations
-where we wanted to leverage Gaia DR3 data, we used M.~Bedell's 4$''$
+{\it Kepler-Gaia crossmatch quality (bits 4 and 5)}--- To leverage 
+Gaia DR3 data, we used M.~Bedell's 4$''$
 Kepler-to-Gaia crossmatch %\footnote{\url{https://gaia-kepler.fun}, last
@@
 renormalized unit weight error (RUWE) as a function of other stellar
-parameters, and flag stars with RUWE$>$1.4 as possible binaries.  Such
+parameters, and flagged stars with RUWE$>$1.4 as possible binaries.  Such
 astrometric outliers can be either bona fide astrometric binaries, or
@@
 {\it Not in the homogeneous stellar rotation sample (bit 11)}--- This
-flag is set for the KOIs with rotation periods drawn from a source
+flag was set for the KOIs with rotation periods drawn from a source
 other than the \citeauthor{Santos_2019} pipeline.
@@
 \label{subsec:plflags}
-Some planets are more reliably identified than others.  We use the
+Some planets are more reliably identified than others.  We used the
 following additional quality indicators to assess the reliability and
-utility of a planet, and assemble them into a separate bitmask,
+utility of a planet, and assembled them into a separate bitmask,
 $Q_{\rm planet}$.
@@
 status ({\it koi\_disposition}), as well as a disposition based only
-on the Kepler data ({\it koi\_pdisposition}).  We require both to
+on the Kepler data ({\it koi\_pdisposition}).  We required both to
 include only ``planet candidates'' and ``confirmed planets''. 
@@
 higher rates \citep{2016ApJ...822...86M}, in part due to the
-size-impact parameter degeneracy.  We flag planet candidates as
+size-impact parameter degeneracy.  We flagged planet candidates as
 potentially grazing if $b$$>$0.8, using the impact parameters reported
@@
 
-To calculate lithium ages, we use \texttt{EAGLES} (git commit
+To calculate lithium ages, we used \texttt{EAGLES} (git commit
 \texttt{ac09637}), which, similar to \texttt{gyro-interp}, is based on
@@
 Gaia-ESO spectroscopic survey in 52 open clusters with ages spanning 2
-to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopt a linear prior in
+to 6{,}000\,Myr \citep{Jeffries_2023}.  We adopted a linear prior in
 age, spanning $10^6$ to $10^{10}$\,yr, and symmetrize our EW
@@
 		At $<$1\,Gyr,
-		the rotation-based ages from our work provide better accuracy 
-		and better-quantified uncertainties than previous analyses.
+    our ages are generally accurate, and their statistical
+    uncertainties match their dispersion.
 	}
@@
 that are more diffuse, but which nonetheless share a common age based
-on e.g.~isochrone, rotation, and lithium dating
+on isochrone, rotation, and lithium dating
 \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
@@
 followed any guidance available from each study for
-removing ages that were flagged as unreliable, and plotted
-stars within $1\sigma$ of zero upper limits.  For the
+removing ages that were unreliable, and plotted
+stars within $1\sigma$ of zero as upper limits.  For the
 \citeauthor{Reinhold_2015} ages, we used those calculated
@@
 they do not marginalize over
-the range of rotation periods accessible to stars of a fixed
+the range of rotation periods accessible to sub-gigayear stars of a fixed
 mass and age.
@@
 near $4$\,Gyr since this is the cluster age out to which our
-interpolation-based method is well-calibrated.  The overall slope
+rotational ages are calibrated.  The overall slope
 yields a paucity of young stars relative to the expectation of a
@@
 ($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
-KOI host age distribution relative to that of the overall stellar
+KOI host age distribution relative to the parent stellar
 sample.  Examining scatter plots of the planet properties as a
@@
 ``intermediate-age'', and ``old''.  A simple counting exercise from
-Figure~\ref{fig:hist_tgyro} tells us that there are \ratioobtoybstars\
-times as many old stars in the Kepler field as young stars.
+Figure~\ref{fig:hist_tgyro} tells us that there are
+\ratioobtoybstars\ times as many old stars in the Kepler field as young stars.
 Similarly, there are \ratioobtoybplanets\ times as many old planet
@@
 evolution.
-Our catalog has \nplyounggyro\ confirmed planets
+Our catalog has \nplyounggyro\ confirmed and candidate planets
 with median ages below 1\,Gyr.  Requiring $t_{\rm gyro}$$<$1\,Gyr at
@@
 rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the cluster age reported in \citet{Bouma_2022b} for RSG-5.
+above the cluster age reported by \citet{Bouma_2022b} for RSG-5.
 
@@
 rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
-posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm
+posterior because of its slow spin-down rate.  Kepler-67 ($T_{\rm
 eff}$$\approx$5100\,K, $P_{\rm rot}$$\approx$10.3\,days) is marginally
-hotter than the later K dwarfs that are in ``stalled'' spindown at
+hotter than the later K dwarfs that are in ``stalled'' spin-down at
 this time, enabling its rotation period to be diagnostic of its age.
@@
 \end{figure}
+%\begin{figure*}[!t]
+%	\begin{center}
+%		\includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+%	\end{center}
+%	\vspace{-0.5cm}
+%	\caption{
+%		{\bf Sizes, orbital periods, and rotation-based ages of transiting
+%			exoplanets younger than one billion years.} Star symbols denote
+%		\nplyounggyrotwosigmanograzingnoruwe\ confirmed and candidate planets that are 
+%		characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
+%		$2$$\sigma$; circles are planets from the literature meeting the
+%		same age requirement.  Gray points are transiting planets from the NASA
+%		Exoplanet Archive older than 1\,Gyr.  If one were to instead
+%		select for precise ages ($t/\sigma_{\rm t}$$>$3), this would hide
+%		ten $<$0.5\,Gyr planets, and add fifteen
+%		0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
+%		hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the planets 
+%		to similarly no have quality flags raised. 
+%		\label{fig:rp_period_age_results}
+%	}
+%\end{figure*}
 
@@
 known planet hosts 
-in the ``comment'' column of Table~\ref{tab:planets}.
+in the ``Consistent?'' column of Table~\ref{tab:planets}.
 There were three cases of interest.
@@
 $t_{\rm Li}$=\kepseveneightsix\,Myr, but with a $\approx$33\,day
-rotaton period that would imply $t_{\rm
+rotation period that implies $t_{\rm
 gyro}$=\kepseveneightsixgyro\,Myr.  Other age indicators in the
@@
 \citet{SilvaAguirre2018}.  The main novelty of our ages relative to
-those studies is that they are accurate at $<$1\,Gyr.
+those studies is their improved accuracy at $<$1\,Gyr.
 
@@
 for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
-stars, we did find one system, Kepler-786, with a precise lithium age
-(\kepseveneightsix\,Myr), but no rotation signal.  Planetary mergers
-are an example of a process that might produce such a signal, but
+stars, we did find one system, Kepler-786, with radically different
+lithium (\kepseveneightsix\,Myr) and rotational
+(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
+one process that could produce such a signal, but
 testing this would require a way of measuring differential abundances
@@
 \citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
-Galactic Plane Time Domain Survey.
+Galactic Bulge Time Domain Survey.
 The surveys with good coverage of different regions of the Galaxy seem
@@
 others have not. A systematic search for the birth clusters of the
-youngest ``field stars'' could combine spatial and kinematic
-information from Gaia with rotation measurements from TESS. Given the
+youngest ``field stars'' could combine positions and velocities
+from Gaia with rotation measurements from TESS. Given the
 $\sim$100\,Myr decoherence time of young clusters, the youngest ``field
@@
     from CMD fitting and white-dwarf chronology, though with
-    some nuances.
-    One important advance is that our
-    rotation-based ages offer better accuracy at $t$$<$1\,Gyr
-    than current isochronal methods (Figure~\ref{fig:agescalecompone}).
+    the advantage of better accuracy for FGK stars at $t$$<$1\,Gyr
+    (e.g.~Figure~\ref{fig:agescalecompone}).
   %
@@
     Concatenating these planets against the existing literature yields
-    $\approx$120-170 known sub-gigayear planets with well-constrained
-    ages.  This lessens the original factor of ten discrepancy to
+    $\approx$170 known sub-gigayear planets.
+    This lessens the original factor of ten discrepancy to
     a factor of at most two.
@@
 This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
-(LGB)
+(LGB, EKP)
 and the Arthur R.~Adams SURF Fellowship (EKP).
 
-%(author contribution statements will go here)
-%L.G.B.~conceived the project, collected data, 
-%inferred ages, and wrote the manuscript.
-%E.K.P.~contributed to early iterations of the rotation-based age analysis.
-%L.A.H.~contributed to project design.
-%H.I. and A.W.H~contributed to acquisition, reduction, and analysis of
-%the HIRES data.
-%All authors assisted in manuscript revision.
+{\it \large Contributions}: Per \url{https://credit.niso.org/}:
+Conceptualization: LGB.
+Data curation: LGB, AWH, HI.
+Formal analysis: LGB.
+Funding acquisition: LGB.
+Investigation: LGB, EKP.
+Methodology: LGB, EKP.
+Project administration: LGB, LAH, AWH.
+Resources: LGB, AWH.
+Software: LGB.
+Supervision: LGB, LAH.
+Validation: LGB.
+Visualization: LGB.
+Writing – original draft: LGB.
+Writing – review \& editing: all co-authors.
 
@@
   Kepler \citep{Borucki10},
-  Keck:I (HIRES),
+  Keck:I (HIRES) \citep{Vogt94},
 	2MASS \citep{Skrutskie06},
@@
   claude \citep{claude2024anthropic},
+  eagles \citep{Jeffries_2023},
+  gyro-interp \citep{Bouma_2023},
   matplotlib \citep{matplotlib},
   numpy \citep{numpy},
-  scipy \citep{scipy},
+  scipy \citep{scipy}.
 }
 
-%\clearpage
+\clearpage
 
@@
 \paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} is
-a $\approx$2.5\,day near-grazing hot Jupiter oribiting a G0 V primary
+a $\approx$2.5\,day near-grazing hot Jupiter orbiting a G0 V primary
 which has been studied in detail by multiple investigators
@@
 system displayed a broad CCF with a blended second component at
-$\approx$$+26$\,\kms relative to the primary, with a best-fit
+$\approx$$+26$\,\kms\ relative to the primary, with a best-fit
 flux-ratio of $\approx$3.4\%, and a preferred $T_{\rm
@@
 double-peaked CCF, in this case with a secondary component at
-$\approx$$-16$\,\kms relative to the primary with $T_{\rm
+$\approx$$-16$\,\kms\ relative to the primary with $T_{\rm
 eff,B}$$\approx$4900\,K.  This putative companion is similarly not
@@
 
-\section{Notes on Individual Systems}
-
-%Kepler~221 is a 633$^{+120}_{-112}$\,Myr four-planet system.
-%
-%Kepler~51 (CITE Masuda2014) has previously been noted for its bizarre architure.
-%The rotation-based age we find here is higher/lower than previously noted.
-
-Kepler~411, a $t_{\rm gyro}$$\approx$780$\pm$170\,Myr early K dwarf with three
-transiting planets at 3.0, 7.8, and 58\,days, has a rare
+\section{Kepler-411}
+
+Kepler-411, a $t_{\rm gyro}$$\approx$780$\pm$170\,Myr early K dwarf
+with three transiting planets at 3.0, 7.8, and 58\,days, has a rare
 architecture.  \citet{2019A&A...624A..15S} conducted a transiting
@@
 \citeauthor{2007ApJ...669.1167B} relation has been known to not fit
-observations for a while
-\citep[e.g.][Fig.~9]{Mamajek_2008}.  Our adopted age of
-$\approx$780$\pm$170\,Myr agrees with the open cluster calibrators.
-
-%\section{TODO's}
-%
-%\begin{enumerate}
-%%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
-%%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
-%%    Could compare against asteroseis too.
-%%    %
-%%  \item Intro needs better discussion of Mathur2023 and Lu2021
-%    %
-%  \item Consider moving the bit list description to its own appendix.
-%    %
-%  \item Rerun gyro calculation with denser grid, fixing quantization
-%    and extending upper limit to 5gyr.  (Will use updated Teff
-%    uncertainties)
-%\end{enumerate}
-
-%\section{Comparison Against Previous Literature Age Scales}
-%Searching the literature for gyrochronal analyses of the Kepler field,
-%the most relevant studies seemed to be those of
-%\citet{Walkowicz_2013}, \citet{Reinhold_2015}, and 
-%\citet{David_2021}.
-%
-%Also Mathur2023, which does ``magneto-gyrochronology'', including the
-%Sph indicator in a model of the ages.
-%
-%
-%\subsection{Asteroseismic Age Comparison}
-%T Ceillier, J Van Saders et al 2016 MNRAS...
+observations for a while \citep[e.g.][Fig.~9]{Mamajek_2008}.  Our
+adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
+calibrators.
 

### 2024-06-06 paper/ms.tex
update ms

@@
     found in the Kepler field
-    \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
+    \cite[e.g.][]{2019AJ....158..122K,Kerr_2021,Barber_2022}.
     These clusters offer stars that we can use to test
@@
 
-Our analysis recovered the ages of all 14 known Kepler planets
-in clusters.  While lithium provided only marginal added information
-for most of the sample, the lithium and rotation ages agreed in
+Tables~\ref{tab:planets} and~\ref{tab:stars} summarize the results for planets and stars,
+respectively.
+Our recovered ages are accurate for all 14 known Kepler planets
+in clusters.  While lithium provided minimal added information
+for most of the sample, $t_{\rm Li}$ and $t_{\rm gyro}$ agreed in
 $\gtrsim$90\% of cases for which comparison was possible.
@@
 \ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
-systems with RUWE$>$1.4 and including ``candidate'' planets
-would increase the counts to
-\ltonegyrmediumqconfirmedtwosided\ and
-\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
-such as on transit geometry, would further increase the counts.  The sizes and
+``candidate'' planets, grazing transit geometries, and stars with RUWE far from unity
+expands the counts by a factor of about two.
+%would increase the counts to
+%\ltonegyrmediumqconfirmedtwosided\ and
+%\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
+%such as on transit geometry, would further increase the counts.
+
+The sizes and
 periods for the most secure set of planets are shown in
@@
   %
-  \item The age distributions of both the Kepler target stars and the
+  \item Rather than being uniform, the age distributions of both the Kepler target stars and the
     known Kepler planet-hosts show a demographic cliff.  There are
@@
     from CMD fitting and white-dwarf chronology, though with
-    the advantage of better accuracy for FGK stars at $t$$<$1\,Gyr
+    the advantage of better sensitivity for FGK stars at $t$$<$1\,Gyr
     (e.g.~Figure~\ref{fig:agescalecompone}).
@@
   %
-  %FIXME: 120 number
   \item We have derived rotation-based ages for
@@
 Writing – original draft: LGB.
-Writing – review \& editing: all co-authors.
+Writing – review \& editing: all authors.
 

### 2024-06-06 paper/ms.tex
update ms

@@
 % upon AAS submission
-%\documentclass[12pt,twocolumn,tighten,linenumbers]{aastex63}
+\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
 %\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
-\documentclass[11pt,twocolumn,tighten]{aastex63}
+%\documentclass[11pt,twocolumn,tighten]{aastex63}
 \turnoffedit
@@
   %
-  Rotation periods are drawn from previous work; lithium is
-  measured from new and archival Keck/HIRES spectra.
+  Rotation periods are drawn from previous work; lithium is measured
+  from new and archival Keck/HIRES spectra.
   %
@@
   This is about half the number expected under the classic assumption
-  of a uniform star formation history.  We find that the scarcity of sub-gigayear
-  systems can
-  be attributed to the star formation rate in the Kepler field
-  dropping by a factor of \ratiosfr$\pm$\uncratiosfr\ over the past
-  3\,Gyr.  We observe this trend both for known planet hosts and also
-  for the parent stellar sample.
+  of a uniform star formation history.  We find that the scarcity of
+  sub-gigayear systems can be attributed to the star formation rate in
+  the Kepler field dropping by a factor of \ratiosfr$\pm$\uncratiosfr\
+  over the past 3\,Gyr.  We observe this trend both in known planet
+  hosts and in the parent stellar sample.
   %
@@
 
-Discovering a young planet requires solving two problems: finding the planet and measuring the star's age.
-Each problem admits a range of
+Discovering a young planet requires solving two problems: finding the
+planet and measuring the star's age.  Each problem admits a range of
 solutions
@@
 \citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
-suggests that
-$\approx$1\% of nearby stars have $t$$<$100\,Myr, and $\approx$10\%
-have $t$$<$1\,Gyr.  Studies of currently forming protoplanets
-\citep{2018A&A...617A..44K}, and of exoplanets evolving just after
-disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped
-in their maximum achievable sample sizes by the tyranny of the
-galactic star formation rate.
+suggests that $\approx$1\% of nearby stars have $t$$<$100\,Myr, and
+$\approx$10\% have $t$$<$1\,Gyr.  Studies of currently forming
+protoplanets \citep{2018A&A...617A..44K}, and of exoplanets evolving
+just after disk dispersal \citep[e.g.][]{2022MNRAS.512.5067K}, are
+thereby capped in their maximum achievable sample sizes by the tyranny
+of the galactic star formation rate.
 
@@
 The ``cluster-first'' approach has a major limitation: most nearby
-stars are in the field.  Only $\lesssim$1\% of stars
-within $\approx$500\,pc have been associated with their birth cluster
+stars are in the field.  Only $\lesssim$1\% of stars within
+$\approx$500\,pc have been associated with their birth cluster
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
@@
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
-than 1\,Gyr.  Requiring $t$$<$1\,Gyr at
-2$\sigma$ precision gives $\approx$50 such planets at
-the time of writing.  Most of these young planets are in clusters, and
-were found by Kepler, K2, or TESS.  However, these surveys have
-discovered $\approx$5{,}000 planets in total.  Assuming a constant
-star formation history, we would expect an order of magnitude more
-sub-gigayear transiting planets.
-
-This study aims to resolve two questions.  First, how wrong is it
-to assume a uniform age distribution for transiting planet host stars?
+than 1\,Gyr.  Requiring $t$$<$1\,Gyr at 2$\sigma$ precision gives
+$\approx$50 such planets at the time of writing.  Most of these young
+planets are in clusters, and were found by Kepler, K2, or TESS.
+However, these surveys have discovered $\approx$5{,}000 planets in
+total.  Assuming a constant star formation history, we would expect an
+order of magnitude more sub-gigayear transiting planets.
+
+This study aims to resolve two questions.  First, how wrong is it to
+assume a uniform age distribution for transiting planet host stars?
 Second, where are the missing young planets?  We will find that
@@
 calculated for Kepler stars \citep{Berger_2020b_rpage} and Kepler
-Objects of Interest \citep[KOIs;][]{Petigura_2022}, such ages are
-most precise for stars whose luminosities and temperatures separate
-them from the main sequence.  For sub-gigayear stars on
-the main sequence, isochrone ages are therefore limited.
+Objects of Interest \citep[KOIs;][]{Petigura_2022}, such ages are most
+precise for stars whose luminosities and temperatures separate them
+from the main sequence.  For sub-gigayear stars on the main sequence,
+isochrone ages are therefore limited.
 
@@
 \citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
-Empirical models now yield ages precise to $\lesssim$30\% for FGK stars
-between 1-4\,Gyr, and constraining age posteriors for younger ages
-\citep{Bouma_2023}.  Physics-based models
-\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} can 
-connect these empirical relations to the evolution of stellar winds,
-internal structure, and angular momentum transport.
-
-Rotation-based ages have been reported for
-various subsets of Kepler stars since the early data releases
+Empirical models now yield ages precise to $\lesssim$30\% for FGK
+stars between 1-4\,Gyr, and constraining age posteriors for younger
+ages \citep{Bouma_2023}.  Physics-based models
+\citep{Matt_2015,Gallet_Bouvier_2015,Spada_2020} can connect these
+empirical relations to the evolution of stellar winds, internal
+structure, and angular momentum transport.
+
+Rotation-based ages have been reported for various subsets of Kepler
+stars since the early data releases
 \citep[e.g.][]{Walkowicz_2013,McQuillan_2014,Reinhold_2015,Angus_2018}.
@@
 \citep{2023ApJ...952..131M}.  Our analysis is motivated by a few
-factors that can yield improvements, particularly for 
-sub-gigayear stars.   These factors are as follows.
+factors that can yield improvements, particularly for sub-gigayear
+stars.   These factors are as follows.
 
@@
   \item Measured rotation periods of FGK stars in open clusters now
-    show not only the average evolution of $P_{\rm rot}(T_{\rm eff},t)$,
-    but also how the astrophysical dispersion of stars around this average
-    converges by the $\approx$700\,Myr age of Praesepe
+    show not only the average evolution of $P_{\rm rot}(T_{\rm
+    eff},t)$, but also how the astrophysical dispersion of stars
+    around this average converges by the $\approx$700\,Myr age of
+    Praesepe
     \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
@@
 Independently, recent years have also yielded improvements in the
-lithium age scale.  Lithium ages include 
-depletion boundary ages for M dwarfs and brown dwarfs in star
-clusters, and decline-based ages for individual field
-FGK stars \citep{Soderblom_2010}.  The latter
+lithium age scale.  Lithium ages include depletion boundary ages for M
+dwarfs and brown dwarfs in star clusters, and decline-based ages for
+individual field FGK stars \citep{Soderblom_2010}.  The latter
 approach relies on the observed decline of Li abundances in
@@
 stars in 52 open clusters.  Two-sided lithium ages are useful for
-Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, though with a strong
-dependence on spectral type.  The precision of lithium ages 
-in this regime are now in the range of 0.3-0.5~dex.
+Kepler (FGK) stars between $\approx$0.03-0.5\,Gyr, though with a
+strong dependence on spectral type.  The precision of lithium ages in
+this regime are now in the range of 0.3-0.5~dex.
 
@@
 stellar parameters other than ages in Section~\ref{sec:stellarprops}.
-We describe our age-dating methods in
-Sections~\ref{sec:rotage} and~\ref{sec:liage}, and test them in
-Section~\ref{sec:litagecomparison} using
-clusters in the Kepler field.  We discuss population-level
-trends in Sections~\ref{sec:results} and~\ref{sec:disc}, and offer a
-few conclusions in Section~\ref{sec:conclusions}.
+We describe our age-dating methods in Sections~\ref{sec:rotage}
+and~\ref{sec:liage}, and test them in
+Section~\ref{sec:litagecomparison} using clusters in the Kepler field.
+We discuss population-level trends in Sections~\ref{sec:results}
+and~\ref{sec:disc}, and offer a few conclusions in
+Section~\ref{sec:conclusions}.
 
@@
   \citet{Santos_2019,Santos_2021}.  About half of the stars with
-  rotation periods are
-  suitable for gyrochronology (dark gray points), based on factors
-  including their non-binarity and proximity to the main
-  sequence (orange line in lower panel; see Section~\ref{subsec:flags}).  Some
-  host ``confirmed'' or ``candidate'' KOIs that meet additional
-  planetary quality criteria (yellow points;
-  Section~\ref{subsec:plflags}).  Surface gravities and effective
-  temperatures were derived photometrically by
+  rotation periods are suitable for gyrochronology (dark gray points),
+  based on factors including their non-binarity and proximity to the
+  main sequence (orange line in lower panel; see
+  Section~\ref{subsec:flags}).  Some host ``confirmed'' or
+  ``candidate'' KOIs that meet additional planetary quality criteria
+  (yellow points; Section~\ref{subsec:plflags}).  Surface gravities
+  and effective temperatures were derived photometrically by
   \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
@@
     {\bf Rotation periods for Kepler target stars (left) and known
-    planet hosts (right)}.  
-    The gray lines are ``mean fits'' to the rotation sequences of
-    open clusters.  The 2-D histogram of the stellar sample (left) includes only apparently
-    single stars near the main sequence with $\log g$$>$4.2,
-    RUWE$<$1.4, and temperatures of 3800--6200\,K.  
-    The planet-hosts (right) require the same
-    stellar cuts, and include only the confirmed and candidate planets
-    described in Section~\ref{subsec:plflags}.  The number of stars $N_{\rm s}$ and
-    planets $N_{\rm p}$ are noted in 
-    the relevant panels.
-    A simple
-    estimate for the number of young planets discovered by Kepler
-    follows by counting points below the mean rotational isochrones.
+    planet hosts (right)}.  The gray lines are ``mean fits'' to the
+    rotation sequences of open clusters.  The 2-D histogram of the
+    stellar sample (left) includes only apparently single stars near
+    the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
+    temperatures of 3800--6200\,K.  The planet-hosts (right) require
+    the same stellar cuts, and include only the confirmed and
+    candidate planets described in Section~\ref{subsec:plflags}.  The
+    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
+    the relevant panels.  A simple estimate for the number of young
+    planets discovered by Kepler follows by counting points below the
+    mean rotational isochrones.
 	}
@@
 \citet{Santos_2019,Santos_2021}, hereafter \citetalias{Santos_2019}
-and \citetalias{Santos_2021}.  
-For a discussion of the work by \citet{Reinhold2023}, please see
-Appendix~\ref{app:reinhold23}.
-\citetalias{Santos_2019} and
-\citetalias{Santos_2021} combined a wavelet analysis and
-autocorrelation-based approach, and cumulatively reported rotation
-periods for 55{,}232 main-sequence and subgiant FGKM stars.  They
-included known KOIs and binaries in their analysis, and removed
-transits and eclipses during the stellar rotation measurement process.
-
-%Studies of stellar rotation across the entire Kepler field include
-%those by \citet{McQuillan_2014} (\citetalias{McQuillan_2014}),
-%\citet{Reinhold_2015}, \citet{Santos_2019} (\citetalias{Santos_2019}),
-%and \citet{Santos_2021} (\citetalias{Santos_2021}).
-%\citetalias{McQuillan_2014} used an approach based on the
-%autocorrelation function to detect 34{,}030 rotation periods for
-%main-sequence Kepler targets cooler than 6500\,K, and excluded known
-%eclipsing binaries and KOIs.  \citet{Reinhold_2015} used an iterative
-%Lomb-Scargle based approach to analyze $\approx$40{,}000 stars with a
-%variability range $R_{\rm var}>0.3$\% that were not known EBs, planet
-%candidates, or pulsators; they reported primary rotation periods for
-%24{,}124 of these stars.  \citet{Reinhold_2015} also reported
-%secondary periods for two thirds of the stars with primary periods,
-%which could be caused by differential rotation or finite spot
-%lifetimes.  Finally, \citetalias{Santos_2019} and
-%\citetalias{Santos_2021} combined a wavelet analysis and
-%autocorrelation-based approach, and cumulatively reported rotation
-%periods for 55{,}232 main-sequence and subgiant FGKM stars.
-%\citetalias{Santos_2019} and \citetalias{Santos_2021} included known
-%KOIs and binaries, and assigned them specific quality flags.  The
-%rotation periods of KOIs have received considerable additional
-%scrutiny
-%\citep[e.g.][]{Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021}.
+and \citetalias{Santos_2021}.  For a discussion of the work by
+\citet{Reinhold2023}, please see Appendix~\ref{app:reinhold23}.
+\citetalias{Santos_2019} and \citetalias{Santos_2021} combined a
+wavelet analysis and autocorrelation-based approach, and cumulatively
+reported rotation periods for 55{,}232 main-sequence and subgiant FGKM
+stars.  They included known KOIs and binaries in their analysis, and
+removed transits and eclipses during the stellar rotation measurement
+process.
 
@@
 is not surprising, since the rotation periods of KOIs have received
-more scrutiny than those of ordinary Kepler stars.  We
-therefore decided to split our subsequent analysis into a homogeneous
-portion that used only the \citetalias{Santos_2019} and
+more scrutiny than those of ordinary Kepler stars.  We therefore
+decided to split our subsequent analysis into a homogeneous portion
+that used only the \citetalias{Santos_2019} and
 \citetalias{Santos_2021} data, and an inhomogeneous portion that also
@@
 reliable'' in their visual analysis of previously reported KOI
-rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013}, \citet{Mazeh_2015}
-and \citet{Angus_2018}.  Inclusion of these additional KOI rotation
-periods is a supplementary measure aimed at completeness in our final
-KOI age catalog; the provenances of the individual adopted periods are
-noted in the relevant tables.
+rotation periods from \citet{McQuillan_2013}, \citet{Walkowicz_2013},
+\citet{Mazeh_2015} and \citet{Angus_2018}.  Inclusion of these
+additional KOI rotation periods is a supplementary measure aimed at
+completeness in our final KOI age catalog; the provenances of the
+individual adopted periods are noted in the relevant tables.
 
@@
 restricted our attention to stars with reported $P_{\rm
-rot}$$<$45\,days.  The slowest-rotating FGK
-stars in the open clusters used to
-calibrate our gyrochronology model have $P_{\rm rot}$$\approx$35\,days.
-Figure~\ref{fig:prot_vs_teff}
-shows a subset of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
+rot}$$<$45\,days.  The slowest-rotating FGK stars in the open clusters
+used to calibrate our gyrochronology model have $P_{\rm
+rot}$$\approx$35\,days.  Figure~\ref{fig:prot_vs_teff} shows a subset
+of the resulting \nuniqstarsantosrot\ Kepler stars with rotation
 periods compiled from \citetalias{Santos_2019},
@@
 compared our adopted periods with those reported by
-\citet{McQuillan_2014}.  The details are in Appendix~\ref{app:reinhold23}.  We found
-that for $P_{\rm rot}\lesssim$15\,days, the two datasets agree at a
-precision of $\lesssim$0.01$P_{\rm rot}$.  At longer periods of
-$P_{\rm rot}\approx$30\,days, the agreement was typically at the
-$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the
-period difference increased roughly linearly with period.  Based on
-this comparison, we adopted a simple prescription for the period
+\citet{McQuillan_2014}.  The details are in
+Appendix~\ref{app:reinhold23}.  We found that for $P_{\rm
+rot}\lesssim$15\,days, the two datasets agree at a precision of
+$\lesssim$0.01$P_{\rm rot}$.  At longer periods of $P_{\rm
+rot}\approx$30\,days, the agreement was typically at the
+$\lesssim$0.03$P_{\rm rot}$ level, and the envelope of the period
+difference increased roughly linearly with period.  Based on this
+comparison, we adopted a simple prescription for the period
 uncertainties, such that there are 1\% relative uncertainties below
@@
 The final piece of our analysis involves assessing ages based on the
-\ion{Li}{1} 6708\,\AA\ doublet.  We analyzed spectra
-from the High Resolution Echelle Spectrometer (HIRES;
-\citealt{vogt_hires_1994}) on the Keck I 10m telescope.  These spectra
-were primarily collected through the California Kepler Survey
+\ion{Li}{1} 6708\,\AA\ doublet.  We analyzed spectra from the High
+Resolution Echelle Spectrometer (HIRES; \citealt{vogt_hires_1994}) on
+the Keck I 10m telescope.  These spectra were primarily collected
+through the California Kepler Survey
 \citep{2017AJ....154..107P,2017AJ....154..108J,2017AJ....154..109F}.
@@
 function are different.  We therefore performed our own line width
-measurements on
-the reduced HIRES spectra.
+measurements on the reduced HIRES spectra.
 
@@
 ``multiple event statistic'' (MES, {\it koi\_max\_mult\_ev}) of at
-least 10.  This yielded at least one spectrum for 1{,}464
-stars hosting 2{,}174 planets.  About half of these stars have
-measured rotation periods (\nstwithspecandprot\ stars and
-\nplwithspecandprot\ planets, respectively).   For stars with multiple spectra
-available, we analyzed only
-the spectrum with the highest number of counts.  The resulting spectra
-were acquired between 2009 September 6 and 2024 May 16, and
-cumulatively comprise \nhireshours\,hours of open-shutter time.
+least 10.  This yielded at least one spectrum for 1{,}464 stars
+hosting 2{,}174 planets.  About half of these stars have measured
+rotation periods (\nstwithspecandprot\ stars and \nplwithspecandprot\
+planets, respectively).   For stars with multiple spectra available,
+we analyzed only the spectrum with the highest number of counts.  The
+resulting spectra were acquired between 2009 September 6 and 2024 May
+16, and cumulatively comprise \nhireshours\,hours of open-shutter
+time.
 
@@
 estimated uncertainties on the line width through a Monte Carlo
-procedure that bootstrapped against the local scatter in the continuum.
-The resulting EWs are shown in Figure~\ref{fig:li_vs_teff}.
+procedure that bootstrapped against the local scatter in the
+continuum.  The resulting EWs are shown in
+Figure~\ref{fig:li_vs_teff}.
 
@@
 \ion{Fe}{1} 6707.44\,\AA\ blend.  To evaluate the accuracy and
-precision of our method, after applying an initial
-iteration on the \nlithiumstars\ stars with spectra, we compared our
-lithium equivalent widths with those reported by
-\citet{2018ApJ...855..115B}.  For the stars in both
-samples, we found broad agreement at $>$30\,m\AA, and significant
-differences at $<$30\,m\AA\ because \citet{2018ApJ...855..115B}
-required positive EWs, while we allowed for statistically negative ones.  At
-$>$30\,m\AA, there is however a small offset in our respective scales caused
-by our non-treatment of the iron blend,
-such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.
-We therefore directly subtracted
-this constant mean value (7.50\,m\AA) when calculating lithium-based
-ages.  This offset is at worst five times smaller than the astrophysical
-scatter present in lithium EWs for calibration clusters at any given age
+precision of our method, after applying an initial iteration on the
+\nlithiumstars\ stars with spectra, we compared our lithium equivalent
+widths with those reported by \citet{2018ApJ...855..115B}.  For the
+stars in both samples, we found broad agreement at $>$30\,m\AA, and
+significant differences at $<$30\,m\AA\ because
+\citet{2018ApJ...855..115B} required positive EWs, while we allowed
+for statistically negative ones.  At $>$30\,m\AA, there is however a
+small offset in our respective scales caused by our non-treatment of
+the iron blend, such that (B24-B18)$=$7.50$\pm$8.55\,m\AA.  We
+therefore directly subtracted this constant mean value (7.50\,m\AA)
+when calculating lithium-based ages.  This offset is at worst five
+times smaller than the astrophysical scatter present in lithium EWs
+for calibration clusters at any given age
 \citep[see][]{Jeffries_2023}.
@@
     Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
-    isochrones from \citet{Jeffries_2023}.  The
-    intrinsic dispersion around these isochrones becomes much larger
-    than changes in the mean at $\gtrsim$1\,Gyr (see
-    Figure~\ref{fig:models}).  Some stars with lithium detections do
-    not have detected rotation periods, and vice-versa.
+    isochrones from \citet{Jeffries_2023}.  The intrinsic dispersion
+    around these isochrones becomes much larger than changes in the
+    mean at $\gtrsim$1\,Gyr (see Figure~\ref{fig:models}).  Some stars
+    with lithium detections do not have detected rotation periods, and
+    vice-versa.
 		\label{fig:li_vs_teff}
@@
 for stars with ``AAA'' 2MASS photometry, measured parallaxes in Gaia
-DR2,  and $g$-band photometry available from either SDSS or else Kepler-INT
-\citep{2012AJ....144...24G}.  The parameters themselves were derived using
-\texttt{isoclassify} \citep{2017ApJ...844..102H} to interpolate over
-the MIST isochrone grids \citep{Choi_2016,2016ApJS..222....8D}, given
-the SDSS $g$ and 2MASS $K_{\rm s}$ photometry, the Gaia DR2
-parallaxes, and $V$-band extinction from the
-\citet{2019ApJ...887...93G} reddening map.  The resulting stellar
-parameters are available for \fracstarswithprotwithbtwenty\ of the
-\nuniqstarsantosrot\ Kepler stars with rotation periods.  For the
-remaining \fracstarswithprotwithoutbtwenty\ of stars that lack
-temperatures and surface gravities from
-\citetalias{Berger_2020a_catalog}, we adopted the values reported by
-\citet{Santos_2019} and \citet{Santos_2021}, which are
-primarily derived from the photometric \citet{Mathur_2017} DR25 Kepler Stellar
-Properties Catalog.  In the
-planet sample, \frackoisnofpwithprotwithbtwenty\ of the
-non-false-positive KOIs with rotation periods have parameters from
-\citet{Berger_2020a_catalog}, and the remainder are drawn
-from DR25.
+DR2,  and $g$-band photometry available from either SDSS or else
+Kepler-INT \citep{2012AJ....144...24G}.  The parameters themselves
+were derived using \texttt{isoclassify} \citep{2017ApJ...844..102H} to
+interpolate over the MIST isochrone grids
+\citep{Choi_2016,2016ApJS..222....8D}, given the SDSS $g$ and 2MASS
+$K_{\rm s}$ photometry, the Gaia DR2 parallaxes, and $V$-band
+extinction from the \citet{2019ApJ...887...93G} reddening map.  The
+resulting stellar parameters are available for
+\fracstarswithprotwithbtwenty\ of the \nuniqstarsantosrot\ Kepler
+stars with rotation periods.  For the remaining
+\fracstarswithprotwithoutbtwenty\ of stars that lack temperatures and
+surface gravities from \citetalias{Berger_2020a_catalog}, we adopted
+the values reported by \citet{Santos_2019} and \citet{Santos_2021},
+which are primarily derived from the photometric \citet{Mathur_2017}
+DR25 Kepler Stellar Properties Catalog.  In the planet sample,
+\frackoisnofpwithprotwithbtwenty\ of the non-false-positive KOIs with
+rotation periods have parameters from \citet{Berger_2020a_catalog},
+and the remainder are drawn from DR25.
 
@@
 and lower temperatures for early F dwarfs.  Our age analysis,
-described below, adopts the maximum of the two-sided
-uncertainties reported by \citetalias{Berger_2020a_catalog} as a
-symmetric Gaussian temperature uncertainty.  Systematic
-uncertainties in the temperature scale generally influence 
-our age uncertainties at a
-smaller level than the statistical intrinsic scatter in the
-open cluster rotation sequences.
+described below, adopts the maximum of the two-sided uncertainties
+reported by \citetalias{Berger_2020a_catalog} as a symmetric Gaussian
+temperature uncertainty.  Systematic uncertainties in the temperature
+scale generally influence our age uncertainties at a smaller level
+than the statistical intrinsic scatter in the open cluster rotation
+sequences.
 
@@
 	\vspace{-0.6cm}
-	\caption{
-		{\bf The models.}
+  \caption{{\bf The models.}
     Points represent $10^4$ draws from models that have been fitted to
@@
     around the mean, which is what the models fit, sets the
-    theoretical precision floor for the
-		age-dating methods.    Additional sources of uncertainty,
-		including measurement uncertainty, impose further limits on
-		achievable precision.  These models are calibrated using data from open
-		clusters.  The sizes
-		of the points are the same in each panel, so that low apparent
+    theoretical precision floor for the age-dating methods.
+    Additional sources of uncertainty, including measurement
+    uncertainty, impose further limits on achievable precision.  These
+    models are calibrated using data from open clusters.  The sizes of
+    the points are the same in each panel, so that low apparent
     density signifies greater dispersion around the mean.
@@
 that stars with the same mass and same rotation period can have a wide
-range of ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone
-age posterior should therefore incorporate the intrinsic
-population-level scatter into its statistical precision.  Figure~\ref{fig:models} highlights this problem,
-particularly in regions where the 0.1--1\,Gyr stars overlap.
-
-To estimate the
-stellar spin-down rate as a function of time and stellar temperature,
-\texttt{gyro-interp} uses measured rotation periods and effective
-temperatures from reference open clusters, and interpolates between them using cubic
-Hermite polynomials.  We calculated the probability of the
-rotation-based age, $t_{\rm gyro}$, given the observed 
-periods and temperatures (and their uncertainties) by
-integrating Equation~1 of \citet{Bouma_2023}.  This procedure marginalizes over the 
-astrophysical scatter that is observed in the open cluster
-sequences.
-We adopted a linear age grid spanning 0--5\,Gyr with
-500 grid points, and interpolated the resulting posteriors to calculate
-any summary statistics.
-The oldest cluster for which
-\texttt{gyro-interp} is calibrated 
-is the 4\,Gyr M67 cluster
-\citep[][]{Dungee_2022,Gruner_2023}.  
-For Kepler stars with rotation periods above the 
-M67 sequence, our resulting age constraints are quoted as lower limits.
+range of ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
+posterior should therefore incorporate the intrinsic population-level
+scatter into its statistical precision.  Figure~\ref{fig:models}
+highlights this problem, particularly in regions where the 0.1--1\,Gyr
+stars overlap.
+
+To estimate the stellar spin-down rate as a function of time and
+stellar temperature, \texttt{gyro-interp} uses measured rotation
+periods and effective temperatures from reference open clusters, and
+interpolates between them using cubic Hermite polynomials.  We
+calculated the probability of the rotation-based age, $t_{\rm gyro}$,
+given the observed periods and temperatures (and their uncertainties)
+by integrating Equation~1 of \citet{Bouma_2023}.  This procedure
+marginalizes over the astrophysical scatter that is observed in the
+open cluster sequences.  We adopted a linear age grid spanning
+0--5\,Gyr with 500 grid points, and interpolated the resulting
+posteriors to calculate any summary statistics.  The oldest cluster
+for which \texttt{gyro-interp} is calibrated is the 4\,Gyr M67 cluster
+\citep[][]{Dungee_2022,Gruner_2023}.  For Kepler stars with rotation
+periods above the M67 sequence, our resulting age constraints are
+quoted as lower limits.
 
@@
 We calculated gyrochrone ages for all \nuniqstarsantosrotteffcut\
-stars with reported rotation periods that had effective temperatures of
-3800--6200\,K.  To identify stars for which we suspect these ages may not
-be valid, we then built a set of quality flags which we condensed
-into a single binary number: $Q_{\rm star}$.  When and how
-this bitmask should be applied depends on the question being
-asked.  If the goal is to construct a false-positive free sample, all the
-quality flags could be applied.  If the goal is to construct a complete sample,
-then consider the examples of Kepler~1627Ab ($\approx$40\,Myr) and
-Kepler~51c ($\approx$625\,Myr).  The former has a high RUWE due to a
-resolved binary companion \citep{Bouma_2022a}; the latter is on a
-grazing orbit \citep{2014ApJ...783...53M}.  We leave selection for or
-against such cases as a decision to the user.
-For our own analysis, 
-we assume that a star
-is suitable for gyrochronology if none of bits zero through nine
-(inclusive) are raised.  For analyses that require all
-stellar rotation periods to come from the same detection pipeline, we
-further require bit 11 to be zero.
+stars with reported rotation periods that had effective temperatures
+of 3800--6200\,K.  To identify stars for which we suspect these ages
+may not be valid, we then built a set of quality flags which we
+condensed into a single binary number: $Q_{\rm star}$.  When and how
+this bitmask should be applied depends on the question being asked.
+If the goal is to construct a false-positive free sample, all the
+quality flags could be applied.  If the goal is to construct a
+complete sample, then consider the examples of Kepler~1627Ab
+($\approx$40\,Myr) and Kepler~51c ($\approx$625\,Myr).  The former has
+a high RUWE due to a resolved binary companion \citep{Bouma_2022a};
+the latter is on a grazing orbit \citep{2014ApJ...783...53M}.  We
+leave selection for or against such cases as a decision to the user.
+For our own analysis, we assume that a star is suitable for
+gyrochronology if none of bits zero through nine (inclusive) are
+raised.  For analyses that require all stellar rotation periods to
+come from the same detection pipeline, we further require bit 11 to be
+zero.
 
@@
 \citep{2016ApJ...821...93N,2023ApJ...954L..50E,2024arXiv240312129C},
-but the currently available open cluster data have yet to clarify
-when the intrinsic scatter in the population decreases.
+but the currently available open cluster data have yet to clarify when
+the intrinsic scatter in the population decreases.
 
@@
 if they had $\log g < 4.2$. 
-%We set this threshold by examining the
-%relevant flag from \citet{berger_2018_radii_evolnstates},
-%cross-matched against \citet{Berger_2020a_catalog}.
+%We set this threshold by examining the relevant flag from
+%\citet{berger_2018_radii_evolnstates}, cross-matched against
+%\citet{Berger_2020a_catalog}.
 
@@
 
-\subsection{Planet Quality Flags}
-\label{subsec:plflags}
-Some planets are more reliably identified than others.  We used the
-following additional quality indicators to assess the reliability and
-utility of a planet, and assembled them into a separate bitmask,
-$Q_{\rm planet}$.
-
-% see Erik's 2022 discussion; might want to shift to include a cut on
-% disposition score?
+\subsection{Planet Quality Flags} \label{subsec:plflags} Some planets
+are more reliably identified than others.  We used the following
+additional quality indicators to assess the reliability and utility of
+a planet, and assembled them into a separate bitmask, $Q_{\rm
+planet}$.
+
 {\it Candidate reliability (bit 0)}---The NEA's Kepler Objects of
@@
 include only ``planet candidates'' and ``confirmed planets''.
-%; we are generally less interested
-%in the age distribution of planet candidates that are known false positives.
 
@@
 		{\bf Reported ages of individual stars in benchmark open clusters.}
-		Each point denotes a star's reported age,
-		normalized by the age of a stellar ensemble in which the star
-		is a candidate member.
+    Each point denotes a star's reported age, normalized by the age of
+    a stellar ensemble in which the star is a candidate member.
     Cluster membership was evaluated without knowledge of rotation.
-		Although some field interlopers may be present in the 
-		membership lists,
-		outliers can be compared between different methods on a
-		relative basis.
-    Each study was cross-matched against the same cluster list;
-    certain methods report ages for more stars than others.
-		Horizontal scatter is added to visually
-		clarify the statistical age uncertainties.
-		At $<$1\,Gyr,
-    our ages are generally accurate, and their statistical
-    uncertainties match their dispersion.
+    Although some field interlopers may be present in the membership
+    lists, outliers can be compared between different methods on a
+    relative basis.  Each study was cross-matched against the same
+    cluster list; certain methods report ages for more stars than
+    others.  Horizontal scatter is added to visually clarify the
+    statistical age uncertainties.  At $<$1\,Gyr, our ages are
+    generally accurate, and their statistical uncertainties match
+    their dispersion.
 	}
@@
 the astronomical age scale \citep{Soderblom_2010}.  Before Gaia, the
-few open clusters known in the Kepler field had been
-cataloged by \citet{1864RSPT..154....1H}.  Gaia
-has enabled the discovery of stellar ensembles
-that are more diffuse, but which nonetheless share a common age based
-on isochrone, rotation, and lithium dating
+few open clusters known in the Kepler field had been cataloged by
+\citet{1864RSPT..154....1H}.  Gaia has enabled the discovery of
+stellar ensembles that are more diffuse, but which nonetheless share a
+common age based on isochrone, rotation, and lithium dating
 \cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
@@
 ages, and available ages from the literature, against the ages of
-stars in open clusters in the Kepler field.  From the
-literature, we drew ages from \citet{Reinhold_2015},
-\citet{Berger_2020a_catalog}, \citet{2021AJ....161..189L},
-\citet{2023ApJ...952..131M}, and \citet{2024AJ....167..159L}.  We
-followed any guidance available from each study for
-removing ages that were unreliable, and plotted
-stars within $1\sigma$ of zero as upper limits.  For the
-\citeauthor{Reinhold_2015} ages, we used those calculated
-using the \citet{Mamajek_2008} calibration; for the
+stars in open clusters in the Kepler field.  From the literature, we
+drew ages from \citet{Reinhold_2015}, \citet{Berger_2020a_catalog},
+\citet{2021AJ....161..189L}, \citet{2023ApJ...952..131M}, and
+\citet{2024AJ....167..159L}.  We followed any guidance available from
+each study for removing ages that were unreliable, and plotted stars
+within $1\sigma$ of zero as upper limits.  For the
+\citeauthor{Reinhold_2015} ages, we used those calculated using the
+\citet{Mamajek_2008} calibration; for the
 \citeauthor{2023ApJ...952..131M} ages, we show those from
 \texttt{kiauhoku} \citep{Claytor2020} rather than \texttt{STAREVOL}
-\citep{Amard2019}, since the former showed better agreement with
-the cluster age scale.
+\citep{Amard2019}, since the former showed better agreement with the
+cluster age scale.
 
@@
 Figure~\ref{fig:agescalecompone} is showing reported ages for a set of
-spatially and kinematically selected stars that could be
-cluster members, or they could be field interlopers.  For NGC6811 and
-NGC6819, we adopted candidate members from
+spatially and kinematically selected stars that could be cluster
+members, or they could be field interlopers.  For NGC6811 and NGC6819,
+we adopted candidate members from
 \citet{2018A&A...618A..93C,CantatGaudin_2020} and
-\citet{Kounkel_2020}.  
-For Theia~520, we used candidate members from \citet{Kounkel_2020}.
-For Melange-3 we used the candidates reported
+\citet{Kounkel_2020}.  For Theia~520, we used candidate members from
+\citet{Kounkel_2020}.  For Melange-3 we used the candidates reported
 by \citet{Barber_2022} and required ``offset'' tangential velocities
-below 2\,\kms.  For Cep-Her we used candidate members from Kerr et al.~2024 (submitted)
-with $P_{\rm fin}$$>$0.8.  Even with
+below 2\,\kms.  For Cep-Her we used candidate members from Kerr et
+al.~2024 (submitted) with $P_{\rm fin}$$>$0.8.  Even with
 contaminants, we can compare the relative age distributions derived by
@@
 
-There are two main metrics for success in this test. {\it 1)}
-Do the reported ages agree with the cluster age? {\it 2)} Do the
-reported age uncertainties agree with their dispersion
-around the cluster age?
-
-Figure~\ref{fig:agescalecompone} show that although
-previous studies reported ages that agree with the cluster scale for
-$\gtrsim$1\,Gyr stars, sub-gigayear stars have historically had their ages 
+There are two main metrics for success in this test. {\it 1)} Do the
+reported ages agree with the cluster age? {\it 2)} Do the reported age
+uncertainties agree with their dispersion around the cluster age?
+
+Figure~\ref{fig:agescalecompone} show that although previous studies
+reported ages that agree with the cluster scale for $\gtrsim$1\,Gyr
+stars, sub-gigayear stars have historically had their ages
 overestimated by 0.3--2\,dex, with severely understated uncertainties.
-For isochrone and kinematic ages, this is because these
-methods rely on parameters that do not appreciably change at
-$t$$\lesssim$1\,Gyr.  For the rotation-based ages in
-\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
-discrepancy is
-caused by details of the calibration methodologies.  The
-\texttt{kiauhoku} spin-down model for instance has known
-``stretches'' and ``compressions'' in its age scale relative to observed open clusters
-\citep[see][Sec. 7.3]{2023ApJ...952..131M}.
-The \citet{Mamajek_2008}
-calibration used by \citet{Reinhold_2015} appears more accurate than the former
-two models, because it 
-was fitted to reproduce the open cluster sequences known at the time.  However the
-uncertainties from all of these methods appear to be underestimated.  This
-is probably because
-they do not marginalize over
-the range of rotation periods accessible to sub-gigayear stars of a fixed
-mass and age.
+For isochrone and kinematic ages, this is because these methods rely
+on parameters that do not appreciably change at $t$$\lesssim$1\,Gyr.
+For the rotation-based ages in \citet{2023ApJ...952..131M} and
+\citet{2024AJ....167..159L}, the discrepancy is caused by details of
+the calibration methodologies.  The \texttt{kiauhoku} spin-down model
+for instance has known ``stretches'' and ``compressions'' in its age
+scale relative to observed open clusters \citep[see][Sec.
+7.3]{2023ApJ...952..131M}.  The \citet{Mamajek_2008} calibration used
+by \citet{Reinhold_2015} appears more accurate than the former two
+models, because it was fitted to reproduce the open cluster sequences
+known at the time.  However the uncertainties from all of these
+methods appear to be underestimated.  This is probably because they do
+not marginalize over the range of rotation periods accessible to
+sub-gigayear stars of a fixed mass and age.
 
@@
     \subfloat{
-%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000.pdf}
-%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000.pdf}
         \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
@@
     \subfloat{
-%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
-%    \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240521_maxage4000_preciseagesonly.pdf}
         \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
@@
   \caption{{\bf Kepler's demographic cliff,} visible in the
-  rotation-derived age distributions of its stars (left) and planet hosts (middle).  The top row shows all stars with
-  temperatures of 3800--6200\,K for which we calculated rotational
-  ages.  Opaque lines impose quality cuts on 
-  binarity, crowding, and the star's evolutionary state; transparent lines do not.
-  The bottom row
-  shows stars with temperatures of 4400--5400\,K, which have
-  more precise ages due to their fast spin-down.
-  The statistical uncertainty for a mean
-  star at 1, 2, and 3\,Gyr is shown, and is identical
-  across each row.  Finally, the right panel compares the
-  rotation-derived ages against star formation histories
-  derived using CMD fitting (dashed and dotted lines).  Completeness in Kepler's
-  $P$$_{\rm rot}$ detection sensitivity is near unity until
-  $t$$\lesssim$3\,Gyr \citep{2022ApJ...937...94M}.
+  rotation-derived age distributions of its stars (left) and planet
+  hosts (middle).  The top row shows all stars with temperatures of
+  3800--6200\,K for which we calculated rotational ages.  Opaque lines
+  impose quality cuts on binarity, crowding, and the star's
+  evolutionary state; transparent lines do not.  The bottom row shows
+  stars with temperatures of 4400--5400\,K, which have more precise
+  ages due to their fast spin-down.  The statistical uncertainty for a
+  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
+  row.  Finally, the right panel compares the rotation-derived ages
+  against star formation histories derived using CMD fitting (dashed
+  and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
+  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
+  \citep{2022ApJ...937...94M}.
   \label{fig:hist_tgyro}
@@
 near $4$\,Gyr since this is the cluster age out to which our
-rotational ages are calibrated.  The overall slope
-yields a paucity of young stars relative to the expectation of a
-uniform age distribution.
-%In an alternative coinage, this tendency for fewer young
-%stars to be present could also be dubbed a ``demographic cliff'',
-%in reference to the phenomenon from human demography.
+rotational ages are calibrated.  The overall slope yields a paucity of
+young stars relative to the expectation of a uniform age distribution.
 
@@
 ($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
-KOI host age distribution relative to the parent stellar
-sample.  Examining scatter plots of the planet properties as a
-function of time, we similarly find that fewer systems at $t_{\rm
+KOI host age distribution relative to the parent stellar sample.
+Examining scatter plots of the planet properties as a function of
+time, we similarly find that fewer systems at $t_{\rm
 gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
@@
 
-Sub-gigayear planets can be particularly informative for studies of planet
-evolution.
-Our catalog has \nplyounggyro\ confirmed and candidate planets
-with median ages below 1\,Gyr.  Requiring $t_{\rm gyro}$$<$1\,Gyr at
-2$\sigma$ yields \nplyounggyrotwosigma\ planets orbiting
-\nplhostsyounggyrotwosigma\ stars.  
-The youth claim is most secure for the subsets that are either
-in clusters, or that have independent rotation and
-lithium-based ages.
+Sub-gigayear planets can be particularly informative for studies of
+planet evolution.  Our catalog has \nplyounggyro\ confirmed and
+candidate planets with median ages below 1\,Gyr.  Requiring $t_{\rm
+gyro}$$<$1\,Gyr at 2$\sigma$ yields \nplyounggyrotwosigma\ planets
+orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim is most
+secure for the subsets that are either in clusters, or that have
+independent rotation and lithium-based ages.
 
@@
 
-{\it Cep-Her}---The Cep-Her complex ($\approx$40\,Myr)
-contains Kepler-1627Ab, Kepler-1643b,
-Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
-Different sub-groups of the complex vary in age by $\approx$50\%
-(Kerr~et al.,~submitted).  The planets
+{\it Cep-Her}---The Cep-Her complex ($\approx$40\,Myr) contains
+Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
+\citep{Bouma_2022b,Bouma_2022a}.  Different sub-groups of the complex
+vary in age by $\approx$50\% (Kerr~et al.,~submitted).  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
-2-4\,$R_\oplus$.  Our rotation and lithium ages agree with the cluster age
-for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
-rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
-above the cluster age reported by \citet{Bouma_2022b} for RSG-5.
-
-{\it MELANGE-3}---\citet{Barber_2022} reported a
-$105\pm10$\,Myr association in the Kepler field
-containing two transiting mini-Neptunes:
-Kepler-970 and Kepler-1928.  We find rotation-based ages
-of $t_{\rm
-gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
-Kepler-970 and Kepler-1928 respectively.  The former rotation-based
-age being slightly older than the suggested Pleiades age for the
-association agrees with Figure~4 of \citet{Barber_2022}.
+2-4\,$R_\oplus$.  Our rotation and lithium ages agree with the cluster
+age for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643,
+the rotation and cluster ages agree, and the lithium age is
+2.0$\sigma$ above the cluster age reported by \citet{Bouma_2022b} for
+RSG-5.
+
+{\it MELANGE-3}---\citet{Barber_2022} reported a $105\pm10$\,Myr
+association in the Kepler field containing two transiting
+mini-Neptunes: Kepler-970 and Kepler-1928.  We find rotation-based
+ages of $t_{\rm gyro}=176_{-40}^{+120}$\,Myr and
+$144_{-88}^{+104}$\,Myr for Kepler-970 and Kepler-1928 respectively.
+The former rotation-based age being slightly older than the suggested
+Pleiades age for the association agrees with Figure~4 of
+\citet{Barber_2022}.
 
@@
 clustering, these stars were reported by \citet{2019AJ....158..122K}
-to be members of Theia~520.  The core of this
-structure is also known as UBC-1 \citep{2018A&A...618A..59C}.  Based
-on isochrone and rotation-based age-dating, Theia~520 seems to be
+to be members of Theia~520.  The core of this structure is also known
+as UBC-1 \citep{2018A&A...618A..59C}.  Based on isochrone and
+rotation-based age-dating, Theia~520 seems to be
 $\approx$230$\pm$70\,Myr old
 \citep{2019AJ....158..122K,2024A&A...681A..13F}, which is within
-1$\sigma$ of our $t_{\rm gyro}$
-measurements.  Ongoing work by \citet{Curtis2024} broadly finds that
-the two planet-hosting stars are indeed members of this diffuse
-population.  Given the cluster-level and individual-star evidence,
-this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
-currently known from the main Kepler mission.
+1$\sigma$ of our $t_{\rm gyro}$ measurements.  Ongoing work by
+\citet{Curtis2024} broadly finds that the two planet-hosting stars are
+indeed members of this diffuse population.  Given the cluster-level
+and individual-star evidence, this makes Kepler-968 and Kepler-52 the
+youngest multiplanet systems currently known from the main Kepler
+mission.
 
@@
     0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-		hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the planets 
-		to similarly no have quality flags raised. 
+    hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the
+    planets to similarly no have quality flags raised. 
     \label{fig:rp_period_age_results}
@@
 \end{figure}
-%\begin{figure*}[!t]
-%	\begin{center}
-%		\includegraphics[width=0.7\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
-%	\end{center}
-%	\vspace{-0.5cm}
-%	\caption{
-%		{\bf Sizes, orbital periods, and rotation-based ages of transiting
-%			exoplanets younger than one billion years.} Star symbols denote
-%		\nplyounggyrotwosigmanograzingnoruwe\ confirmed and candidate planets that are 
-%		characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
-%		$2$$\sigma$; circles are planets from the literature meeting the
-%		same age requirement.  Gray points are transiting planets from the NASA
-%		Exoplanet Archive older than 1\,Gyr.  If one were to instead
-%		select for precise ages ($t/\sigma_{\rm t}$$>$3), this would hide
-%		ten $<$0.5\,Gyr planets, and add fifteen
-%		0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-%		hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the planets 
-%		to similarly no have quality flags raised. 
-%		\label{fig:rp_period_age_results}
-%	}
-%\end{figure*}
 
@@
 
-We compared 
-the rotation and lithium ages for the
-known planet hosts 
-in the ``Consistent?'' column of Table~\ref{tab:planets}.
-There were three cases of interest.
-{\it i)} If two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$ existed,
-and their median values were consistent within 2$\sigma$, we listed them
-as consistent; if they were consistent within 2-3$\sigma$, we
-listed them as ``maybe'' consistent.  Otherwise, they
-disagreed.
-{\it ii)}
-If $t_{\rm Li}$ was a lower limit, we compared this lower limit
-with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
-overlapped, we judged the age estimates to be consistent.
-{\it iii)}
-If $t_{\rm Li}$ provided a two-sided posterior, and no
-rotation period was found, we judged the age estimates to be
-inconsistent.  This is because a two-sided lithium constraint can only be
-provided for G and K dwarfs $\lesssim$2\,Gyr old, and Kepler should
-have been sensitive to the rotation periods of such stars.
-
-%\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
-%  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
-%    if their values were consistent within 2$\sigma$, we judged them
-%    to be consistent; if they were consistent within 2-3$\sigma$, we
-%    suggest them to be ``maybe'' consistent.  Otherwise, they
-%    disagree.
-%    %
-%  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
-%    with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
-%    overlapped, we judge the age estimates to be consistent.
-%    %
-%  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
-%    rotation is reported, we judge the age estimates to be
-%    inconsistent.  This is because this lithium constraint can only be
-%    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
-%    have been sensitive to the rotation periods of such stars.
-%\end{enumerate}
-
-In the entire planet sample (i.e.~without imposing any quality cuts) this
-yielded \allagesyesconsistent\ consistent cases,
+We compared the rotation and lithium ages for the known planet hosts
+in the ``Consistent?'' column of Table~\ref{tab:planets}.  There were
+three cases of interest.  {\it i)} If two-sided posteriors for both
+$t_{\rm gyro}$ and $t_{\rm Li}$ existed, and their median values were
+consistent within 2$\sigma$, we listed them as consistent; if they
+were consistent within 2-3$\sigma$, we listed them as ``maybe''
+consistent.  Otherwise, they disagreed.  {\it ii)} If $t_{\rm Li}$ was
+a lower limit, we compared this lower limit with the $1\sigma$ upper
+limit from $t_{\rm gyro}$; if the two overlapped, we judged the age
+estimates to be consistent.  {\it iii)} If $t_{\rm Li}$ provided a
+two-sided posterior, and no rotation period was found, we judged the
+age estimates to be inconsistent.  This is because a two-sided lithium
+constraint can only be provided for G and K dwarfs $\lesssim$2\,Gyr
+old, and Kepler should have been sensitive to the rotation periods of
+such stars.
+
+In the entire planet sample (i.e.~without imposing any quality cuts)
+this yielded \allagesyesconsistent\ consistent cases,
 \allagesmaybeconsistent\ maybe consistent cases, and
@@
 
-Highlights from these systems include
-Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
-is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation
-and lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
+Highlights from these systems include Kepler-1529, Kepler-1565,
+Kepler-1312, and Kepler-1629.  Kepler-1529b is a
+$\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation and
+lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
 super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
-near-Solar analog with an Earth-sized planet
-on a one-day orbit, and a mini-Neptune on a five-day orbit.  The
-Earth-sized planet, Kepler-1312~c, along with the near-identical
-Kepler-1561~b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
-youngest Earth-sized planets currently known, comparable to planets
-such as HD~63433d \citep[1.1\,$R_\oplus$,
-$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b
-\citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
-Kepler-1629~b, with a size just two-thirds that of Earth
-(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on
-rotation (\kepsixteentwonine\,Myr).  The lithium age from a
-reconnaissance TRES spectrum confirms the point.
-
-%  Kepler-1529b, a 5.3\,day 2.0\,$R_\oplus$ mini-Neptune is $\approx$100\,Myr old.
-%  Kepler-1930b, a 13.0\,day 1.5\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$120-180\,Myr old.
-%  Kepler-1313b, doubly unlucky, a 3.8\,day 1.7\,$R_\oplus$ super-Earth on a grazing orbit is $\approx$185\,Myr old.
-%  Kepler-1521b, a 47.2\,day 2.3\,$R_\oplus$ mini-Neptune is 180-210\,Myr old.
-%  Kepler-1467b, a 47.1\,day 3.1\,$R_\oplus$ mini-Neptune is 230-240\,Myr old.
-%  Kepler-1565b, a 1.5\,day 1.3\,$R_\oplus$ super-Earth is 180-230\,Myr old.
-%  Kepler-971b, a 9.6\,day 3.9\,$R_\oplus$ Neptune-sized planet is 300-370\,Myr old.
-%  
-%  Kepler-1933b is a 4.9\,day 1.0\,$R_\oplus$ Earth-sized planet.  Its
-%  rotation-based age of $304^{+96}_{-112}$\,Myr agrees within $1\sigma$ of
-%  its lithium age, though at half the relative uncertainty.  Based on
-%  these lines of evidence, this system may be the youngest Earth-sized
-%  planet known, comparable to systems such as HD~63433d
-%  \citep[1.1\,$R_\oplus$, $414\pm23$\,Myr][]{2024AJ....167...54C}.
-%  
-%  Kepler-1532b is a 1.1\,day 1.3\,$R_\oplus$ super-Earth.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
-%  Kepler-955b is a 14.5\,day 2.7\,$R_\oplus$ mini-Neptune.  Rotation points to a $\approx$400\,Myr age, and lithium is consistent.
-%  Kepler-1481b, a 5.9\,day 1.1\,$R_\oplus$ super-Earth has a
-%  rotation-based age of $\approx$400\,Myr, and a $2\sigma$ lithium
-%  detection that, upon visual inspection of the spectrum, is
-%  believeable.  The implied lithium age is marginally consistent with
-%  the rotation-based age, although perhaps with some tension.
-%  Kepler-447b is supposedly a 32\,$R_\oplus$ planet (grazing) on a
-%  7.8\,day grazing orbit.  Rotation and lithium independently find a
-%  420-440\,Myr age.
-%  Kepler-1561b is a 0.9\,$R_\oplus$ sub-Earth, on a 1.0\,day orbit.
-%  Rotation points to a 420\,Myr age.  The detected EW$_{\rm Li}$ is
-%  consistent, although given the star's temperature, not particularly
-%  constraining.
-%  The same is true for the 1.9\,$R_\oplus$ and 5.8\,day super-Earth Kepler-523b, and the 
-%  1.0\,$R_\oplus$, 12.2\,day Earth-sized Kepler-1482b.
+near-Solar analog with an Earth-sized planet on a one-day orbit, and a
+mini-Neptune on a five-day orbit.  The Earth-sized planet,
+Kepler-1312~c, along with the near-identical Kepler-1561~b ($t_{\rm
+gyro}$=\kepfifteensixone\,Myr), rank among the youngest Earth-sized
+planets currently known, comparable to planets such as HD~63433d
+\citep[1.1\,$R_\oplus$, $414$$\pm$$23$\,Myr;][]{2024AJ....167...54C}
+and TOI-1807~b \citep[1.3\,$R_\oplus$,
+180$\pm$40\,Myr;][]{2021AJ....162...54H}.  Kepler-1629~b, with a size
+just two-thirds that of Earth (0.67$\pm$0.05\,$R_\oplus$), is
+marginally older, also based on rotation (\kepsixteentwonine\,Myr).
+The lithium age from a reconnaissance TRES spectrum confirms the
+point.
 
@@
 \citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
-galaxy, previous studies have measured star
-formation histories through CMD fitting of resolved stellar
-populations
+galaxy, previous studies have measured star formation histories
+through CMD fitting of resolved stellar populations
 \citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
@@
 local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
-focused on stars in a wider 2\,kpc bubble, and additionally
-reported three local
-maxima in the galactic star formation rate, which they associated with pericenter passages of the
-Sagittarius satellite galaxy.
+focused on stars in a wider 2\,kpc bubble, and additionally reported
+three local maxima in the galactic star formation rate, which they
+associated with pericenter passages of the Sagittarius satellite
+galaxy.
 
@@
 CMD fitting.  The overall slope of both SFHs broadly agrees with what
-we find from rotation-based ages.  The SFH from \citet{2019A&A...624L...1M}
-seems entirely consistent, particularly after accounting for
-the statistical uncertainties of that study.  Regarding the
-episodic star formation bursts reported by
+we find from rotation-based ages.  The SFH from
+\citet{2019A&A...624L...1M} seems entirely consistent, particularly
+after accounting for the statistical uncertainties of that study.
+Regarding the episodic star formation bursts reported by
 \citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
@@
 \citep{Rogers_2021} can be explored using our data.  We defer this
-analysis to a separate publication.
-Some care is required since beyond age,
-exoplanet demographics also depend on stellar metallicity and mass
-\citep[e.g.][]{Petigura_2018,Miyazaki2023}.
+analysis to a separate publication.  Some care is required since
+beyond age, exoplanet demographics also depend on stellar metallicity
+and mass \citep[e.g.][]{Petigura_2018,Miyazaki2023}.
 
 {\it Field star ages from other surveys}:
-While this study focused on
-Kepler, other photometric surveys (e.g.~K2, TESS, HAT, WASP, NGTS, ZTF,
-ATLAS) open opportunities for age-dating a far broader set
-of stars and planets.
-Future prospects also include PLATO
-\citep{Rauer14}, Earth 2.0 \citep{2022arXiv220606693G}, and the Roman
-Galactic Bulge Time Domain Survey.
-The surveys with good coverage of different regions of the Galaxy seem
-most likely to yield new insight into whether the star formation
-history we see in the Kepler field is universal across the thin disk.
+While this study focused on Kepler, other photometric surveys
+(e.g.~K2, TESS, HAT, WASP, NGTS, ZTF, ATLAS) open opportunities for
+age-dating a far broader set of stars and planets.  Future prospects
+also include PLATO \citep{Rauer14}, Earth 2.0
+\citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
+Survey.  The surveys with good coverage of different regions of the
+Galaxy seem most likely to yield new insight into whether the star
+formation history we see in the Kepler field is universal across the
+thin disk.
 
@@
 others have not. A systematic search for the birth clusters of the
-youngest ``field stars'' could combine positions and velocities
-from Gaia with rotation measurements from TESS. Given the
-$\sim$100\,Myr decoherence time of young clusters, the youngest ``field
-stars'' may have easily identifiable young neighbors.
+youngest ``field stars'' could combine positions and velocities from
+Gaia with rotation measurements from TESS. Given the $\sim$100\,Myr
+decoherence time of young clusters, the youngest ``field stars'' may
+have easily identifiable young neighbors.
 
@@
 
-We began this work with two questions: how wrong is the assumption
-of a uniform age distribution for stars in the galactic thin disk?
-And why are only $\approx$50 sub-gigayear transiting planets known,
-rather than the $\approx$500 that would be expected under the
-assumption of a uniform star formation history for the
-$\approx$5{,}000 known planets?
+We began this work with two questions: how wrong is the assumption of
+a uniform age distribution for stars in the galactic thin disk?  And
+why are only $\approx$50 sub-gigayear transiting planets known, rather
+than the $\approx$500 that would be expected under the assumption of a
+uniform star formation history for the $\approx$5{,}000 known planets?
 
@@
 stellar rotation periods, lithium equivalent widths, and temperatures
-using archival and new data from the Kepler field.  We derived new ages using empirical
-interpolation-based methods, and assessed the reliability of these ages
-by comparing them against benchmark open clusters
-(Figure~\ref{fig:agescalecompone}).
-
-Tables~\ref{tab:planets} and~\ref{tab:stars} summarize the results for planets and stars,
-respectively.
-Our recovered ages are accurate for all 14 known Kepler planets
-in clusters.  While lithium provided minimal added information
-for most of the sample, $t_{\rm Li}$ and $t_{\rm gyro}$ agreed in
-$\gtrsim$90\% of cases for which comparison was possible.
-Our resulting ages included two-sided $t_{\rm gyro}$ and $t_{\rm Li}$
-for \ltonegyrhighqconfirmedtwosided\ sub-gigayear confirmed planets, and
-two-sided $t_{\rm gyro}$ with one-sided $t_{\rm Li}$ for
-\ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
-``candidate'' planets, grazing transit geometries, and stars with RUWE far from unity
-expands the counts by a factor of about two.
-%would increase the counts to
-%\ltonegyrmediumqconfirmedtwosided\ and
-%\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
-%such as on transit geometry, would further increase the counts.
-
-The sizes and
-periods for the most secure set of planets are shown in
-Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
-mostly mini-Neptunes, some are near the lower boundary of the
-``sub-Jovian desert'' \citep{Owen2018}, which could have an evolutionary
-connection to their youth.
-Other discoveries include Earth-sized 
-planets with new ages of only a few hundred million years.
-
-Our main conclusions with respect to our original questions are
-as follows.
+using archival and new data from the Kepler field.  We derived new
+ages using empirical interpolation-based methods, and assessed the
+reliability of these ages by comparing them against benchmark open
+clusters (Figure~\ref{fig:agescalecompone}).
+
+Tables~\ref{tab:planets} and~\ref{tab:stars} summarize the results for
+planets and stars, respectively.  Our recovered ages are accurate for
+all 14 known Kepler planets in clusters.  While lithium provided
+minimal added information for most of the sample, $t_{\rm Li}$ and
+$t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
+was possible.  Our resulting ages included two-sided $t_{\rm gyro}$
+and $t_{\rm Li}$ for \ltonegyrhighqconfirmedtwosided\ sub-gigayear
+confirmed planets, and two-sided $t_{\rm gyro}$ with one-sided $t_{\rm
+Li}$ for \ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed
+planets.  Allowing for ``candidate'' planets, grazing transit
+geometries, and stars with RUWE far from unity expands the counts by a
+factor of two.
+
+The sizes and periods for the most secure set of planets are shown in
+Figure~\ref{fig:rp_period_age_results}.   While the new young planets
+are mostly mini-Neptunes, some are near the lower boundary of the
+``sub-Jovian desert'' \citep{Owen2018}, which could have an
+evolutionary connection to their youth.  Other discoveries include
+Earth-sized planets with new ages of only a few hundred million years.
+
+Our main conclusions with respect to our original questions are as
+follows.
 
@@
   %
-  \item Rather than being uniform, the age distributions of both the Kepler target stars and the
-    known Kepler planet-hosts show a demographic cliff.  There are
-    twice as many ``old'' (2-3\,Gyr) stars in the
-    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
-    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
-    three billion years ago.
-    This result from rotation-based ages broadly agrees with recent reports of
-    a declining star-formation rate
-    from CMD fitting and white-dwarf chronology, though with
-    the advantage of better sensitivity for FGK stars at $t$$<$1\,Gyr
-    (e.g.~Figure~\ref{fig:agescalecompone}).
+  \item Rather than being uniform, the age distributions of both the
+    Kepler target stars and the known Kepler planet-hosts show a
+    demographic cliff.  There are twice as many ``old'' (2-3\,Gyr)
+    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
+    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
+    than it was three billion years ago.  This result from
+    rotation-based ages broadly agrees with recent reports of a
+    declining star-formation rate from CMD fitting and white-dwarf
+    chronology, though with the advantage of better sensitivity for
+    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
   %
   \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, the age distribution of stars in the
-    Kepler field implies that we should instead expect $\approx$250.
+    billion years, the age distribution of stars in the Kepler field
+    implies that we should instead expect $\approx$250.
   %
-  \item We have derived rotation-based ages for
-    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
-    and \nplyounggyro\ planets with median ages below 1\,Gyr.
+  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
+    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
+    \nplyounggyro\ planets with median ages below 1\,Gyr.
     Concatenating these planets against the existing literature yields
-    $\approx$170 known sub-gigayear planets.
-    This lessens the original factor of ten discrepancy to
-    a factor of at most two.
+    $\approx$170 known sub-gigayear planets.  This lessens the
+    original factor of ten discrepancy to a factor of at most two.
 \end{enumerate}

### 2024-06-06 paper/ms.tex
update ms

@@
 \acknowledgements
-This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
+
+We are grateful to Kento Masuda for thought-provoking discussions 
+which significantly influenced our thinking about this analysis.
+The work itself was supported by the Heising-Simons 51~Pegasi~b Fellowship
 (LGB, EKP)

### 2024-06-07 paper/ms.tex
update ms

@@
 \affiliation{\caltech}
-%\affiliation{51 Pegasi b Fellow}
-
-%\author{and coauthors, including but not limited to...}
 
@@
 
-%\author{and to be confirmed:}
-%\affiliation{\caltech}
-
 \author[0000-0002-0531-1073]{Howard Isaacson}
@@
 
-%  \author[0000-0003-1298-9699]{Kento Masuda} ?
-%  \affiliation{\osaka}
+\author[0000-0003-1298-9699]{Kento Masuda}
+\affiliation{\osaka}
 
@@
 \citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
-Survey.  The surveys with good coverage of different regions of the
-Galaxy seem most likely to yield new insight into whether the star
-formation history we see in the Kepler field is universal across the
-thin disk.
+Survey \citep{Wilson2023}.  Rotation-based age-dating from these surveys could
+yield new insight
+into whether the star formation history that we see in the Kepler field is
+universal across the thin disk.
 
@@
 
-We are grateful to Kento Masuda for thought-provoking discussions 
-which significantly influenced our thinking about this analysis.
 The work itself was supported by the Heising-Simons 51~Pegasi~b Fellowship
-(LGB, EKP)
-and the Arthur R.~Adams SURF Fellowship (EKP).
+(LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
+The HIRES spectra were obtained at the W.~M.~Keck Observatory.
+We recognize the importance that the summit of Maunakea has always had
+within the indigenous Hawaiian community, and we are deeply grateful 
+for the opportunity to conduct observations from this mountain.
 
@@
 Data curation: LGB, AWH, HI.
-Formal analysis: LGB.
+Formal analysis: LGB, KM.
 Funding acquisition: LGB.
 Investigation: LGB, EKP.
-Methodology: LGB, EKP.
+Methodology: LGB, EKP, KM.
 Project administration: LGB, LAH, AWH.
@@
 Supervision: LGB, LAH.
-Validation: LGB.
+Validation: LGB, KM.
 Visualization: LGB.

### 2024-06-07 paper/ms.tex
update ms

@@
 orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim is most
-secure for the subsets that are either in clusters, or that have
-independent rotation and lithium-based ages.
+secure for the systems that are either in clusters (Section~\ref{subsec:clusterplanets}), or that have
+independent rotation and lithium-based ages (Section~\ref{subsec:rotnlithiumcomp}).
+These samples provide broader context for the planets around field
+stars that push the boundaries of the current young exoplanet census
+(Section~\ref{subsec:notables}).
 
@@
 \subsubsection{Rotation vs.~Lithium Ages}
+\label{subsec:rotnlithiumcomp}
 
@@
 
-\subsubsection{New Young Planets}
+\subsubsection{Notable New Young Planets}
+\label{subsec:notables}
 
@@
 
-Highlights from these systems include Kepler-1529, Kepler-1565,
+Among these systems, four highlights are Kepler-1529, Kepler-1565,
 Kepler-1312, and Kepler-1629.  Kepler-1529b is a

### 2024-06-07 paper/ms.tex
update ms

@@
 distributions by drawing ten random samples from each posterior, and
-then computing the re-weighted histogram of the resulting samples.
+then computing the normalized histogram of the resulting samples.
+We considered an alternative approach for constructing these
+plots using hierarchical Bayesian deconvolution, and found similar results.
 The uncertainties assume Poisson statistics.  We truncated the plots

### 2024-06-10 paper/ms.tex
update ms [EKP edits]

@@
 \citetalias{Santos_2021} provided a comparison against
-\citetalias{McQuillan_2014}; the brief summary is that the periods
+\citet{McQuillan_2014} (hereafter \citetalias{McQuillan_2014}); the brief summary is that the periods
 agree for 99.0\% of the 31{,}038 period detections in common between
@@
 star, their presence can confuse rotation period measurements.  We
-therefore flagged any stars with neighbors down to 10$\times$ the
+therefore flagged any stars with neighbors down to 1/10$^{\rm th}$ the
 brightness of the target star within this region ($\Delta G < 2.5$).
@@
 Visual inspection shows that although this flag selects many bona fide
-objects in these classes, it also selects most $P_{\rm
-rot}$$\lesssim$5\,day bona fide young planets.  These objects are
+objects in these classes, it also selects most known young planets
+with $P_{\rm rot}$$\lesssim$5\,day.  These objects are
 neither classical pulsators nor close binaries.  While we propagated
@@
 on the Kepler data ({\it koi\_pdisposition}).  We required both to
-include only ``planet candidates'' and ``confirmed planets''.
+include only ``planet candidates'' and ``confirmed planets.''
 
@@
 \citet{Kounkel_2020}.  For Theia~520, we used candidate members from
-\citet{Kounkel_2020}.  For Melange-3 we used the candidates reported
+\citet{Kounkel_2020}.  For Melange-3, we used the candidates reported
 by \citet{Barber_2022} and required ``offset'' tangential velocities
-below 2\,\kms.  For Cep-Her we used candidate members from Kerr et
+below 2\,\kms.  For Cep-Her, we used candidate members from Kerr et
 al.~2024 (submitted) with $P_{\rm fin}$$>$0.8.  Even with
@@
 models, because it was fitted to reproduce the open cluster sequences
-known at the time.  However the uncertainties from all of these
+known at the time.  However, the uncertainties from all of these
 methods appear to be underestimated.  This is probably because they do
@@
 Kepler-1627Ab, Kepler-1643b, Kepler-1974b and Kepler-1975\,Ab
-\citep{Bouma_2022b,Bouma_2022a}.  Different sub-groups of the complex
+\citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
 vary in age by $\approx$50\% (Kerr~et al.,~submitted).  The planets
@@
   	Two checkmarks ($\checkmark\checkmark$) denote confirmed planets with two-sided $t_{\rm
-  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are expected to be reliable.  One checkmark ($\checkmark$) denotes confirmed planets iwth two-sided $t_{\rm gyro}$ only. 
+  gyro}$ and $t_{\rm Li}$ for which both the age and the planet are
+  expected to be reliable.  One checkmark ($\checkmark$) denotes
+  confirmed planets with two-sided $t_{\rm gyro}$ only. 
   ``Confirmed'' planets appear in the machine-readable
@@
     shows the expected $\pm 1\sigma$ range that would be spanned if
-    the indepndent $P_{\rm rot}$ measurements had gaussian
+    the independent $P_{\rm rot}$ measurements had gaussian
     uncertainties drawn following the empirical estimate described in
@@
 By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
-However for F and G dwarfs, the resulting locus
+However, for F and G dwarfs, the resulting locus
 allowed for stars that were ``too evolved''; the density of KOIs 

### 2024-06-17 paper/ms.tex
update ms (add mcquillan only appendix)

@@
 
+\section{What if we only considered McQuillan's periods?}
+
+We argued in Section~\ref{subsec:rotsel} that our choice of adopting
+the periods from \citet{Santos_2019,Santos_2021} yielded
+the best possible balance between homogeneity and sensitivity for 
+both Kepler targets and for KOIs.
+Nonetheless, it is interesting to consider whether our
+analysis of the Kepler field's rotation-based age distribution is robust to
+this choice, for instance compared to adopting only the rotation periods from
+\citet{McQuillan_2014}.  
+Figures~\ref{fig:prot_vs_teff_mcq} and~\ref{fig:hist_tgyro_mcq}
+repeat the analysis, but make this alternate choice.
+Figure~\ref{fig:prot_vs_teff_mcq} shows that while the overall contours
+of the $P_{\rm rot}$ vs.~$T_{\rm eff}$ distribution are similar between
+the two samples, the \citeauthor{Santos_2019} distribution includes more stars,
+particularly at longer rotation periods.
+The impact of this on the inferred rotation-based age distributions
+(Figure~\ref{fig:hist_tgyro_mcq}) is that while the
+\citeauthor{Santos_2019} distribution peaks at 2.8-3.0\,Gyr,
+the \citeauthor{McQuillan_2014} distribution peaks at 2.4-2.6\,Gyr.
+A number of metrics shift.
+For instance, the ratio of the star formation rate 3\,Gyr age to today
+is \ratiosfr$\pm$\uncratiosfr\ assuming the \citeauthor{Santos_2019} periods,
+and \mcquillanonlyratiosfr$\pm$\mcquillanonlyuncratiosfr\ assuming
+the \citeauthor{McQuillan_2014} periods.
+Similarly, the ratio of ``old'' (2-3\,Gyr) to ``young'' (0-1\,Gyr) stars
+shifts from \ratioobtoybstars\ in the \citeauthor{Santos_2019} case
+to \mcquillanonlyratioobtoybstars\ in the \citeauthor{McQuillan_2014}
+case.
+The K dwarf age distributions (Figure~\ref{fig:hist_tgyro_mcq}) however appear 
+quite similar.
+Generally, we believe these plots support the conclusion that the $t_{\rm gyro}$
+age distribution in the Kepler field is highly non-uniform, and likely peaks in the 2.5-3\,Gyr range.
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
+    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
+    against the \citet{McQuillan_2014} rotation periods ({\it right}).
+    All stars are required to be amenable to rotation-based age dating
+    (i.e.~none of bit flags zero through nine raised).
+	}
+	\label{fig:prot_vs_teff_mcq}
+\end{figure*}
+
+\begin{figure*}[!th]
+  \begin{center}
+    \leavevmode
+    \subfloat{
+        \includegraphics[width=.85\textwidth]{mcq_santos_comp_top.pdf}
+    }
+
+	\vspace{-0.35cm}
+    \subfloat{
+        \includegraphics[width=.85\textwidth]{mcq_santos_comp_bottom.pdf}
+    }
+  \end{center}
+  \vspace{-0.66cm}
+  \caption{{\bf Variant of Figure~\ref{fig:hist_tgyro},} comparing
+  our default adopted catalog of \citet{Santos_2019,Santos_2021}
+  rotation periods against the rotation periods
+  reported by \citet{McQuillan_2014}.
+  \label{fig:hist_tgyro_mcq}
+  }
+\end{figure*}
+
 \clearpage

### 2024-06-17 paper/collect.sh

@@ -42,3 +42,9 @@ cp $r/lit_ages_vs_cluster_ages/olympic_merged_lit_ages_vs_cluster_ages.pdf .
 
 cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-m14_Prot_vs_Prot.pdf .
 cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-r23_ProtGPS_vs_Prot.pdf .
+
+cp $r/star_Prot_Teff/prot_teff_McQuillan2014only_dquality.pdf .
+
+
+cp $r/mcq14_vs_santos_age_histograms_Santos20240530_vs_McQMcQ14_20240613/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf mcq_santos_comp_top.pdf
+cp $r/mcq14_vs_santos_age_histograms_Santos20240530_vs_McQMcQ14_20240613/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf mcq_santos_comp_bottom.pdf

### 2024-06-17 paper/ms.tex
update appendices

--- 2024-06-10 paper/ms.tex
@@
 \subsection{Caveats \& Limitations}
+\label{subsec:caveats}
 
@@
 Kepler field, but also in the reliability of individual ages for young
-stars known to host planets.  We assessed this reliability using
+stars known to host planets.  We attempted to assess this reliability using
 quality bitmasks (Table~\ref{tab:planets}), a comparison against open
@@
 suggested that our ages were generally accurate at $\ll$0.5\,dex
-across 0.04--2.5\,Gyr, comparable to their stated precision level.
+across 0.04--2.5\,Gyr, comparable to their quoted precision.
 Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
-shows that outliers do exist, typically at most 10\% of the population
-in each cluster.  These stars could be either field star contaminants,
+shows that outliers do exist, typically $\lesssim$10\% of the population
+in any cluster.  These stars could be either field star contaminants,
 or else anomalous stars whose physical rotation histories were altered
@@
 (\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
-one process that could produce such a signal, but
-testing this would require a way of measuring differential abundances
+are one process that could produce such a signal, but
+testing this would require a method for measuring differential abundances
 across a large number of elements.
@@
 is at the limits of the Kepler data \citep{2022ApJ...937...94M}.  In
-the ``old star'' regime, methods using either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, or else using both the
+the ``old star'' regime, methods that leverage either asteroseismology
+\citep{vanSaders_2016,2024ApJ...962..138S}, or else that use both the
 evolution of stellar luminosity and stellar rotation
@@
 
-\section{Kepler-411}
-
-Kepler-411, a $t_{\rm gyro}$$\approx$780$\pm$170\,Myr early K dwarf
-with three transiting planets at 3.0, 7.8, and 58\,days, has a rare
-architecture.  \citet{2019A&A...624A..15S} conducted a transiting
-timing analysis of the system, confirming Kepler-411d and arguing for
-the presence of a fourth non-transiting planet, Kepler-411e, based on
-the TTVs seen in Kepler-411d and assuming that the system is nearly
-coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
-the \citet{2007ApJ...669.1167B} relation and the measured rotation
-period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
-star's temperature ($\approx$4900\,K) puts it in a regime where the
-\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
-observations for a while \citep[e.g.][Fig.~9]{Mamajek_2008}.  Our
-adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
-calibrators.
+%\section{Kepler-411}
+%
+%Kepler-411, a $t_{\rm gyro}$$\approx$780$\pm$170\,Myr early K dwarf
+%with three transiting planets at 3.0, 7.8, and 58\,days, has a rare
+%architecture.  \citet{2019A&A...624A..15S} conducted a transiting
+%timing analysis of the system, confirming Kepler-411d and arguing for
+%the presence of a fourth non-transiting planet, Kepler-411e, based on
+%the TTVs seen in Kepler-411d and assuming that the system is nearly
+%coplanar.  Regarding the stellar age, \citet{2019A&A...624A..15S} used
+%the \citet{2007ApJ...669.1167B} relation and the measured rotation
+%period, which yielded an estimate of 212$\pm$31\,Myr.  However, the
+%star's temperature ($\approx$4900\,K) puts it in a regime where the
+%\citeauthor{2007ApJ...669.1167B} relation has been known to not fit
+%observations for a while \citep[e.g.][Fig.~9]{Mamajek_2008}.  Our
+%adopted age of $\approx$780$\pm$170\,Myr agrees with the open cluster
+%calibrators.
 
@@
 
+\section{What if we only considered McQuillan's periods?}
+
+We argued in Section~\ref{subsec:rotsel} that adopting
+the periods from \citet{Santos_2019,Santos_2021} yielded the best
+possible balance between homogeneity and sensitivity for both KOIs and
+for the broader Kepler sample.  Nonetheless, it is interesting to
+consider what would happen if we were to analyze only the rotation
+periods from \citet{McQuillan_2014}.
+Figures~\ref{fig:prot_vs_teff_mcq} and~\ref{fig:hist_tgyro_mcq} repeat
+the analysis, but make this alternate choice.
+Figure~\ref{fig:prot_vs_teff_mcq} shows that while the overall
+contours of the $P_{\rm rot}$ vs.~$T_{\rm eff}$ distributions are
+similar, the \citeauthor{Santos_2019} distribution includes more
+stars, particularly at longer rotation periods.  This is connected to
+a stronger noise-dependent cutoff in the \citet{McQuillan_2014} sample
+than in the \citeauthor{Santos_2019} samples (see
+\citet{Masuda2022_amplevoln} Figures~10 and 15).
+Figure~\ref{fig:hist_tgyro_mcq} shows the impact on the inferred
+rotation-based age histograms: while the \citeauthor{Santos_2019}
+distribution peaks at 2.8-3.0\,Gyr, the \citeauthor{McQuillan_2014}
+distribution peaks at 2.4-2.6\,Gyr.  A number of metrics shift.  For
+instance, the ratio of the star formation rate 3\,Gyr age to today is
+\ratiosfr$\pm$\uncratiosfr\ assuming the \citeauthor{Santos_2019}
+periods, and \mcquillanonlyratiosfr$\pm$\mcquillanonlyuncratiosfr\
+assuming the \citeauthor{McQuillan_2014} periods.  Similarly, the
+ratio of ``old'' (2-3\,Gyr) to ``young'' (0-1\,Gyr) stars shifts from
+\ratioobtoybstars\ in the \citeauthor{Santos_2019} case to
+\mcquillanonlyratioobtoybstars\ in the \citeauthor{McQuillan_2014}
+case.  The K dwarf age distributions (Figure~\ref{fig:hist_tgyro_mcq}
+lower panel) however appear quite similar.  Generally, these plots
+support the conclusion that the $t_{\rm gyro}$ age distribution in the
+Kepler field is non-uniform, with a peak in the 2.5-3\,Gyr range.
+Nonetheless, some of the details do shift depending on the choice of
+rotation detection pipeline.
+
+This analysis is particularly interesting in relation to the question
+of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
+universe in which stellar rotation rates and spot-induced photometric
+amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
+stars would tend to contaminate the 2.5-3\,Gyr peaks in
+Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
+scenario deserves consideration, given evidence from asteroseismic
+rotation periods that has pointed to slowed spin-down rates after
+$\approx$4\,Gyr for Sun-like stars
+\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}.  However, because
+the \citeauthor{McQuillan_2014} sample does not include many stars
+older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis
+would argue that the 2.5-3\,Gyr peak cannot be caused by a pile-up of
+much older stars with especially lethargic spin-down rates.
+
+\begin{figure*}[!t]
+	\begin{center}
+		\subfloat{
+			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+		}
+		\subfloat{
+			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{
+    {\bf Variant of Figure~\ref{fig:prot_vs_teff}}, comparing the default
+    rotation periods ({\it left}) from \citet{Santos_2019,Santos_2021}
+    against the \citet{McQuillan_2014} rotation periods ({\it right}).
+    All stars are required to be amenable to rotation-based age dating
+    (i.e.~none of bit flags zero through nine raised).
+	}
+	\label{fig:prot_vs_teff_mcq}
+\end{figure*}
+
+\begin{figure*}[!th]
+  \begin{center}
+    \leavevmode
+    \subfloat{
+        \includegraphics[width=.85\textwidth]{mcq_santos_comp_top.pdf}
+    }
+
+	\vspace{-0.35cm}
+    \subfloat{
+        \includegraphics[width=.85\textwidth]{mcq_santos_comp_bottom.pdf}
+    }
+  \end{center}
+  \vspace{-0.66cm}
+  \caption{{\bf Variant of Figure~\ref{fig:hist_tgyro},} comparing
+  our default adopted catalog of \citet{Santos_2019,Santos_2021}
+  rotation periods against the rotation periods
+  reported by \citet{McQuillan_2014}.
+  \label{fig:hist_tgyro_mcq}
+  }
+\end{figure*}
+
 \clearpage

### 2024-06-17 paper/ms.tex
update ms

@@
 then computing the normalized histogram of the resulting samples.
-We considered an alternative approach for constructing these
-plots using hierarchical Bayesian deconvolution, and found similar results.
-The uncertainties assume Poisson statistics.  We truncated the plots
-near $4$\,Gyr since this is the cluster age out to which our
-rotational ages are calibrated.  The overall slope yields a paucity of
+We used this sampling approach, rather than taking a histogram of
+only the median values, because it better captures the asymmetric posteriors
+of the youngest stars.
+We also considered an alternative approach for constructing these
+plots using hierarchical Bayesian deconvolution \citep{Masuda2022infer}, and found similar results.
+The plotted uncertainties assume Poisson statistics.  We truncated the plots
+at $4$\,Gyr, which is the upper bound for our
+rotational age calibration.  The overall slope yields a paucity of
 young stars relative to the expectation of a uniform age distribution.

### 2024-06-21 paper/ms.tex
update ms

@@
 then computing the normalized histogram of the resulting samples.
-We used this sampling approach, rather than taking a histogram of
-only the median values, because it better captures the asymmetric posteriors
-of the youngest stars.
 We also considered an alternative approach for constructing these
-plots using hierarchical Bayesian deconvolution \citep{Masuda2022infer}, and found similar results.
-The plotted uncertainties assume Poisson statistics.  We truncated the plots
+plots using hierarchical Bayesian deconvolution
+\citep{Masuda2022infer}, and found similar results.  The plotted
+uncertainties assume Poisson statistics.  We truncated the plots
 at $4$\,Gyr, which is the upper bound for our

### 2024-06-21 paper/ms.tex
aas submitted

@@
 %\title{A Paucity of Young Stars and Planets in the Kepler Field}
-\title{Ages of Young Stars and Planets in the Kepler Field}
+%\title{A Non-Uniform Age Distribution of Stars and Planets in the Kepler Field}
+%\title{}
+% Rotational and Lithium-Based Ages of Stars and Planets in the Kepler Field
+% Stars And Planets Younger Than Four Billion Years In The Kepler Field
+%\title{Sub-Solar Aged Stars and Planets in the Kepler Field}
+%From Youth to Maturity: Ages of Stars and Planets in the Kepler Field
+%Ages of Stars and Planets in the Kepler Field Over The First Four Billion Years
+
+\title{Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years}
 
@@
 
-\received{---}
+\received{2024 June 21}
 \revised{---}
@@
 \author[0000-0001-7967-1795]{Elsa~K.~Palumbo}
+\affiliation{\caltech}
 \affiliation{\cmu}
@@
   We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
-  and \nplwgyroagenograzing\ planets for which our method is
+  (harboring \nplwgyroagenograzing\ known planets) for which our method is
   applicable.
@@
 search them for transiting planets.  The resulting stellar, and
-assumed planetary, ages are precise at the $\approx$10\% level.  A
-recent development facilitated by Gaia is the idea that finding
-transiting planets is often easier than finding the host star's birth
-association \citep[e.g.][]{Tofflemire_2021}.
-
-The ``cluster-first'' approach has a major limitation: most nearby
+assumed planetary, ages are precise at the $\approx$10\% level.  
+An alternative strategy, facilitated by Gaia, is to select transiting
+planets based on empirical youth indicators, and to then search their local
+neighborhoods for comoving companions \citep[e.g.][]{Tofflemire_2021}.
+
+While stellar ensembles provide the gold standard for astronomical ages,
+this ``cluster-first'' approach has a major limitation: most 
 stars are in the field.  Only $\lesssim$1\% of stars within
@@
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-The implication can appreciated by querying the NASA Exoplanet Archive
+The implication for planetary ages can appreciated by querying the NASA Exoplanet Archive
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
@@
 planets are in clusters, and were found by Kepler, K2, or TESS.
-However, these surveys have discovered $\approx$5{,}000 planets in
-total.  Assuming a constant star formation history, we would expect an
+However, these surveys have cumulatively discovered a total of $\approx$5{,}000 planets.
+Assuming a constant star formation history, we would expect an
 order of magnitude more sub-gigayear transiting planets.
@@
 
-Independently, recent years have also yielded improvements in the
+Recent years have also yielded improvements in the
 lithium age scale.  Lithium ages include depletion boundary ages for M
@@
 individual field FGK stars \citep{Soderblom_2010}.  The latter
-approach relies on the observed decline of Li abundances in
+approach relies on the observed decrease of Li abundances in
 partially-convective stars as they age
 \citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
-this decline is debated
+these observations is debated
 \citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}.
-Empirical understanding however has improved due to the work by
+Empirical understanding however has improved due to work by
 \citet{Jeffries_2023}, who modeled the time evolution of the
@@
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+			\includegraphics[width=0.47\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+			\includegraphics[width=0.47\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.96\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
+			\includegraphics[width=0.95\textwidth]{olympic_merged_lit_ages_vs_cluster_ages.pdf}
 		}
@@
 \citet{1864RSPT..154....1H}.  Gaia has enabled the discovery of
-stellar ensembles that are more diffuse, but which nonetheless share a
-common age based on isochrone, rotation, and lithium dating
-\cite[e.g.][]{2019AJ....158..122K,Bouma_2022b,Barber_2022}.
+new stellar ensembles that are more diffuse, but which nonetheless share a
+common age based on isochrone, rotation, and lithium dating.
+Specifically in the Kepler field,
+NGC6811 (1\,Gyr; \citealt{Curtis_2019_ngc6811}) and NGC6819 (2.5\,Gyr; \citealt{Meibom_2015}) 
+have been known for centuries,
+while 
+Theia~520 ($\approx$300\,Myr; \citealt{2019AJ....158..122K}), Melange-3 ($\approx$150\,Myr; \citealt{Barber_2022}), and Cep-Her ($\approx$40\,Myr; \citealt{Bouma_2022b} and Kerr et al.~2024 submitted) are more recent discoveries.
 
@@
 ages, and available ages from the literature, against the ages of
-stars in open clusters in the Kepler field.  From the literature, we
+stars in these open clusters. 
+From the literature, we
 drew ages from \citet{Reinhold_2015}, \citet{Berger_2020a_catalog},
@@
     \subfloat{
-        \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
+        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
     }
@@
     \subfloat{
-        \includegraphics[width=.96\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
+        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
     }
   \end{center}
-  \vspace{-0.66cm}
+  \vspace{-0.6cm}
   \caption{{\bf Kepler's demographic cliff,} visible in the
@@
 
-{\it Theia~520}---We derive rotation-based ages of $\approx$350\,Myr
+{\it Theia~520}---We find rotation-based ages of $\approx$350\,Myr
 for Kepler-968 and Kepler-52.  Based on spatial and kinematic
-clustering, these stars were reported by \citet{2019AJ....158..122K}
-to be members of Theia~520.  The core of this structure is also known
+clustering, these stars are candidate 
+members of Theia~520 \citep{2019AJ....158..122K}.  The core of this cluster is also known
 as UBC-1 \citep{2018A&A...618A..59C}.  Based on isochrone and
@@
 $\approx$230$\pm$70\,Myr old
-\citep{2019AJ....158..122K,2024A&A...681A..13F}, which is within
+\citep{2024A&A...681A..13F}, which is within
 1$\sigma$ of our $t_{\rm gyro}$ measurements.  Ongoing work by
@@
 gyro}$=\kepseveneightsixgyro\,Myr.  Other age indicators in the
-spectrum similarly suggest that the star is not young.
+spectrum similarly suggest that the star is not young.  It 
+therefore appears to be anomalously lithium-rich.
 
@@
 Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
-shows that outliers do exist, typically $\lesssim$10\% of the population
+does show outliers, typically $\lesssim$10\% of the population
 in any cluster.  These stars could be either field star contaminants,
@@
 the ``old star'' regime, methods that leverage either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, or else that use both the
+\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both the
 evolution of stellar luminosity and stellar rotation
 \citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most capable of
-providing useful age constraints for single stars.  Incorporating
-kinematics \citep{2021AJ....161..189L} seems likely to remain most
-useful for stellar ensembles, due to the stochasticity of kinematic
-heating.
+providing useful age constraints for single stars over the full
+span of their main-sequence lifetimes.
+%Incorporating
+%kinematics \citep{2021AJ....161..189L} seems likely to remain most
+%useful for stellar ensembles, due to the stochasticity of kinematic
+%heating.
 
@@
 \acknowledgements
-
-The work itself was supported by the Heising-Simons 51~Pegasi~b Fellowship
+This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
 (LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
-The HIRES spectra were obtained at the W.~M.~Keck Observatory.
+The HIRES data were obtained at the Keck Observatory.
 We recognize the importance that the summit of Maunakea has always had
@@
   Ages of Kepler target stars derived from rotation periods.  The
-  full machine-readable table includes \nuniqstarsantosrot\ stars
-  with reported rotation periods from \citetalias{Santos_2019} and
-  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
-  finite reported gyrochrone ages.  A random set of these stars is
-  shown to give an idea for form and content.  The quality
+  full machine-readable table includes
+  \nuniqstarfinitegyroage\ stars with
+  finite reported gyrochrone ages.  The quality
   bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
@@
 \end{deluxetable*}
+
+%\vspace{-5cm}
 

### 2024-06-21 paper/collect.sh

@@ -6,11 +6,11 @@ cp $r/gyromodeldispersion/gyromodeldispersion.pdf .
 # plot_st_params
 # ...CAMD:
 cp $r/st_params/st_params_M_G_vs_dr3_bp_rp.pdf .
-cp $r/st_params/st_params_M_G_vs_adopted_Teff.pdf .
+#cp $r/st_params/st_params_M_G_vs_adopted_Teff.pdf .
 # ...logg vs Teff
 cp $r/st_params/st_params_adopted_logg_vs_adopted_Teff.pdf .
 # ...G vs Teff
-cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
+#cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
 
 # plot_star_Prot_Teff
 cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .

### 2024-08-16 papers/paper_submitted/collect.sh
re-organize for revisions

@@ -1,0 +1,50 @@
+r="/Users/luke/Dropbox/proj/gyrojo/results"
+
+# plot_gyromodeldispersion
+cp $r/gyromodeldispersion/gyromodeldispersion.pdf .
+
+# plot_st_params
+# ...CAMD:
+cp $r/st_params/st_params_M_G_vs_dr3_bp_rp.pdf .
+cp $r/st_params/st_params_M_G_vs_adopted_Teff.pdf .
+# ...logg vs Teff
+cp $r/st_params/st_params_adopted_logg_vs_adopted_Teff.pdf .
+# ...G vs Teff
+cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
+
+# plot_star_Prot_Teff
+cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
+
+# plot_koi_mean_prot_teff
+cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
+
+# plot_hist_field_gyro_ages 
+cp $r/hist_field_gyro_ages_20240530/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf .
+cp $r/hist_field_gyro_ages_20240530/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf .
+
+# plot_li_vs_teff
+#cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_logy.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_JUMP_eagles.pdf .
+cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
+
+# planets
+r="/Users/luke/Dropbox/proj/NEA_age_plots/results/rp_vs_period_scatter"
+cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf .
+
+
+########## 
+# appendices (probably omit)
+r="/Users/luke/Dropbox/proj/gyrojo/results"
+cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_showplanets_linear.png .
+cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_linear.png .
+
[…]

### 2024-08-19 papers/paper/ms.tex
update ms

--- 2024-06-21 paper/ms.tex
@@
 \documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
-%\documentclass[12pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
+%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
@@
     {\bf Rotation periods for Kepler target stars (left) and known
-    planet hosts (right)}.  The gray lines are ``mean fits'' to the
-    rotation sequences of open clusters.  The 2-D histogram of the
+    planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
+    rotation sequences of open clusters}{seventh-order polynomials
+    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
     stellar sample (left) includes only apparently single stars near
@@
     planets discovered by Kepler follows by counting points below the
-    mean rotational isochrones.
+    \deleted{mean }rotational isochrones.
 	}
@@
 periods compiled from \citetalias{Santos_2019},
-\citetalias{Santos_2021}, and our extended KOI list.
-
-To assess the statistical uncertainties of these rotation periods, we
-compared our adopted periods with those reported by
+\citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
+lines in this figure are the polynomials described by \citet{Bouma_2023}
+that were
+fitted to the slow rotation sequences of the Pleiades \citep{Rebull_2016},
+Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
+Ruprecht-147 \citep{Curtis_2020}, and M67 \citep{Barnes2016,Dungee_2022,Gruner_2023}.}
+
+To assess the statistical uncertainties of \replaced{these}{our adopted} rotation periods, we
+compared our \deleted{adopted} periods with those reported by
 \citet{McQuillan_2014}.  The details are in
@@
 that stars with the same mass and same rotation period can have a wide
-range of ages \citep[e.g.][]{Curtis_2019_ngc6811}.  The gyrochrone age
+range of ages \citep[e.g.][]{Gallet_Bouvier_2015}.  The gyrochrone age
 posterior should therefore incorporate the intrinsic population-level
@@
 general aim is to flag stars for which there is evidence based on
-their location in the HR diagram that our gyrochronology model may not
+their location in the \replaced{HR}{Kiel} diagram that our gyrochronology model may not
 be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
@@
 stars known to host planets.  We attempted to assess this reliability using
-quality bitmasks (Table~\ref{tab:planets}), a comparison against open
+quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
 cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
-between rotation and lithium ages.  The open cluster comparison
+between rotation and lithium ages.  The \deleted{open }cluster comparison
 suggested that our ages were generally accurate at $\ll$0.5\,dex
@@
 than in the \citeauthor{Santos_2019} samples (see
-\citet{Masuda2022_amplevoln} Figures~10 and 15).
+\citealt{Masuda2022_amplevoln} Figures~10 and 15).
 Figure~\ref{fig:hist_tgyro_mcq} shows the impact on the inferred
@@
 support the conclusion that the $t_{\rm gyro}$ age distribution in the
-Kepler field is non-uniform, with a peak in the 2.5-3\,Gyr range.
+Kepler field is non-uniform, with a peak in the \replaced{2.5}{2.3}-3\,Gyr range.
 Nonetheless, some of the details do shift depending on the choice of

### 2024-08-19 drivers/make_star_table.py
add gyro uncertainty floor

@@ -74,6 +74,15 @@ def make_star_table(
         ascending=[True, True]
     ).reset_index(drop=1)
 
+    # impose 10% gyro uncertainty floor
+    rel_punc = sdf['gyro_+1sigma'].astype(float) / sdf['gyro_median'].astype(float)
+    rel_munc = sdf['gyro_-1sigma'].astype(float) / sdf['gyro_median'].astype(float)
+    FLOOR = 0.1
+    sel = rel_punc < FLOOR
+    sdf.loc[sel, 'gyro_+1sigma'] = sdf.loc[sel, 'gyro_median'] * FLOOR
+    sel = rel_munc < FLOOR
+    sdf.loc[sel, 'gyro_-1sigma'] = sdf.loc[sel, 'gyro_median'] * FLOOR
+
     for c in sdf.columns:
         if 'gyro_' in c or c == 'adopted_Teff':
             sdf[c] = sdf[c].apply(cast_to_int_string)
@@ -173,6 +182,7 @@ def make_star_table(
     sdf.to_csv(csvpath, index=False, na_rep='')
     print(f'Wrote {csvpath}')
 
+    import IPython; IPython.embed()
 
     csvpath = join(TABLEDIR, 'table_star_gyro_allcols.csv')
     odf = df.sort_values(

### 2024-08-19 drivers/make_younghighlight_table.py

@@ -137,6 +137,15 @@ def make_table(
         N = len(_df[sel])
         df = _df[sel]
 
+    # impose 10% gyro uncertainty floor
+    rel_punc = df['gyro_+1sigma'].astype(float) / df['gyro_median'].astype(float)
+    rel_munc = df['gyro_-1sigma'].astype(float) / df['gyro_median'].astype(float)
+    FLOOR = 0.1
+    sel = rel_punc < FLOOR
+    df.loc[sel, 'gyro_+1sigma'] = df.loc[sel, 'gyro_median'] * FLOOR
+    sel = rel_munc < FLOOR
+    df.loc[sel, 'gyro_-1sigma'] = df.loc[sel, 'gyro_median'] * FLOOR
+
     selcols = (
         "kepid,kepoi_name,kepler_name,koi_disposition,"
         "min_age,"

### 2024-08-19 papers/paper/ms.tex
update ms

@@
 % upon AAS submission
-\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
-%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
+%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
+\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
@@
 all Kepler target stars irrespective of KOI status.  Out of these
-studies, the most homogeneous and precise analyses of both Kepler
+studies, the most homogeneous \deleted{and precise }analyses of both Kepler
 targets and KOIs appear to be those by
@@
 for which \texttt{gyro-interp} is calibrated is the 4\,Gyr M67 cluster
-\citep[][]{Dungee_2022,Gruner_2023}.  For Kepler stars with rotation
-periods above the M67 sequence, our resulting age constraints are
+\citep[][]{Dungee_2022,Gruner_2023}.  For \deleted{Kepler }stars with rotation
+periods above the M67 sequence, our resulting age constraints are\added{ therefore }
 quoted as lower limits.
+\added{Finally, we opted to impose a $t_{\rm gyro}$ uncertainty floor of 10\%. 
+This seemed warranted given the expected amplitude of differential rotation \citep[e.g.][]{Epstein_2014}, and
+given the accuracy with which the absolute cluster age scale is currently known \citep[e.g.][]{Dahm_2015}. }
 
@@
 
+\added{A separate plausible origin for rotational outliers is that they come from
+erroneously reported rotation periods.  While
+the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
+the referee correctly noted that this only a comparative test of methods,
+rather than an ability to actually recover ground truth \citep[e.g.][]{Aigrain2015}.
+One could imagine an adversarial scenario in which, say, 10-20\% of stars
+with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
+longer rotation periods.  In this scenario, our analysis would overestimate the
+rate by which the SFR in the Kepler field has decreased, by 7-14\%. }
+
 Other age indicators may help in verifying the ages of the
-$\lesssim$3\,Gyr stars that are the focus of this work.  Specifically,
+$\lesssim$3\,Gyr stars that are the \added{main }focus of this work.  Specifically,
 chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
@@
 While we opted for the \citetalias{Santos_2019} and
-\citetalias{Santos_2021} rotation period catalogs, the recent
-analysis by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
+\citetalias{Santos_2021} rotation period catalogs, the \deleted{recent }analysis 
+by \citet{Reinhold2023} bears comment.  \citet{Reinhold2023}
 used a similar selection function as \citeauthor{Santos_2021}
@@
 \section{What if we only considered McQuillan's periods?}
+\label{app:mcqonly}
 

### 2024-08-20 papers/paper/ms.tex
[WIP] update ms

@@
 lines in this figure are the polynomials described by \citet{Bouma_2023}
-that were
-fitted to the slow rotation sequences of the Pleiades \citep{Rebull_2016},
+that 
+fit the slow rotation sequences of the Pleiades \citep{Rebull_2016},
 Praesepe \citep{Rampalli_2021}, NGC-6811 \citep{Curtis_2019_ngc6811}, NGC-6819 \citep{Meibom_2015},
@@
 For our own analysis, we assume that a star is suitable for
-gyrochronology if none of bits zero through nine (inclusive) are
+gyrochronology if none of bits zero through \replaced{nine}{ten} (inclusive) are
 raised.  For analyses that require all stellar rotation periods to
-come from the same detection pipeline, we further require bit 11 to be
+come from the same detection pipeline, we further require bit \replaced{11}{12} to be
 zero.
@@
 %main sequence.
+\added{{\it Metallicity (bit 10)}---Theoretical models
+	suggest that stars with non-solar metallicities
+	spin down at different rates than solar-metallicity stars 
+	\citep{Amard2020}.   This expectation has been hard
+	to verify due to a dearth of nearby non-solar metallicity open clusters.
+	However, if true, this effect could cause systematic
+	biases for gyrochrone ages \citep[e.g.~Figure~9 of][]{Claytor2020}.
+	We therefore flagged stars with spectroscopic $|[{\rm Fe/H}]|$$>$0.3
+	from either LAMOST DR5 \citep{Zong2018} or CKS \citep{Petigura_2022}
+	as being outside the range of our gyrochronology calibration.
+	We discuss limitations of this approach in Section~\ref{sec:disc}.
+}
 
 {\it Candidate pulsators and close-in binaries (bit
-10)}---\citeauthor{Santos_2021} included in a flag for candidate
+\replaced{10}{11})}---\citeauthor{Santos_2021} included in a flag for candidate
 ``classical pulsators'' (e.g.\ Cepheids) and close-in binaries.
@@
 
-{\it Not in the homogeneous stellar rotation sample (bit 11)}--- This
+{\it Not in the homogeneous stellar rotation sample (bit \replaced{11}{12})}--- This
 flag was set for the KOIs with rotation periods drawn from a source
@@
   3800--6200\,K for which we calculated rotational ages.  Opaque lines
-  impose quality cuts on binarity, crowding, and the star's
+  impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
   evolutionary state; transparent lines do not.  The bottom row shows
-  stars with temperatures of 4400--5400\,K, which have more precise
-  ages due to their fast spin-down.  The statistical uncertainty for a
+  stars with temperatures of 4400--5400\,K, which \replaced{have more precise
+  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
   mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
@@
 Examining scatter plots of the planet properties as a function of
-time, we similarly find that fewer systems at $t_{\rm
+age, we similarly find that fewer systems at $t_{\rm
 gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
@@
 that it drops to almost zero by $\approx$5\,Gyr.  This trend is
-opposite in functional form to our derived age distributions.
+opposite in functional form to our derived age distributions\replaced{.}{and
+	most of the stars in our sample are indeed G dwarfs (see e.g.,~the $t_{\rm gyro}$--$T_{\rm eff}$ distributions 
+	in Appendix~\ref{app:age_diagnostic}).}
 
@@
 Let us focus on blemish-free stars with blemish-free planets: $Q_{\rm
-star}$ must not have any of bits zero through nine raised, and $Q_{\rm
+star}$ must not have any of bits zero through \replaced{nine}{ten} raised, and $Q_{\rm
 planet}$ must similarly have no quality flags raised.  Under this
@@
 suggested that our ages were generally accurate at $\ll$0.5\,dex
-across 0.04--2.5\,Gyr, comparable to their quoted precision.
+across 0.04--2.5\,Gyr, similar to their quoted precision.
 Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
@@
 the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
-the referee correctly noted that this only a comparative test of methods,
-rather than an ability to actually recover ground truth \citep[e.g.][]{Aigrain2015}.
-One could imagine an adversarial scenario in which, say, 10-20\% of stars
+the referee correctly pointed out that this only a comparative test of methods,
+rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
+One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
 with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
@@
   bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
-  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
+  ten to be null yields \nuniqstarsantosrotgyroappl\ stars for
   which gyrochronology is likely to be valid.  \label{tab:stars}}
@@
 amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
-stars would tend to contaminate the 2.5-3\,Gyr peaks in
+stars \replaced{would tend to }{could }contaminate the 2.5-3\,Gyr peaks in
 Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
-scenario deserves consideration, given evidence from asteroseismic
-rotation periods that has pointed to slowed spin-down rates after
+scenario deserves consideration, given\replaced{ evidence from asteroseismic
+rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
 $\approx$4\,Gyr for Sun-like stars
-\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}.  However, because
+\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
+in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
+may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
 the \citeauthor{McQuillan_2014} sample does not include many stars
-older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis
-would argue that the 2.5-3\,Gyr peak cannot be caused by a pile-up of
+older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis}{the analysis
+by \citet{2022ApJ...937...94M} fitted isochrones to a high-quality subset of Kepler data,
+and found that the
+sample of detected rotators from \citeauthor{McQuillan_2014} does not include many
+stars older than $\approx$\,4\,Gyr.}
+\replaced{would argue}{This suggests} that the 2.5-3\,Gyr peak \replaced{cannot}{is unlikely to} be caused by a pile-up of
 much older stars with especially lethargic spin-down rates.
@@
     All stars are required to be amenable to rotation-based age dating
-    (i.e.~none of bit flags zero through nine raised).
+    (i.e.~none of bit flags zero through \replaced{nine}{ten} raised).
 	}

### 2024-08-20 papers/paper/ms.tex
update ms

@@
 lithium (\kepseveneightsix\,Myr) and rotational
-(\kepseveneightsixgyro\,Myr) ages.  Planetary mergers
+($\approx$4.4\,Gyr) ages.  Planetary mergers
 are one process that could produce such a signal, but

### 2024-08-21 papers/paper/ms.tex
update ms

@@
   sub-gigayear systems can be attributed to the star formation rate in
-  the Kepler field dropping by a factor of \ratiosfr$\pm$\uncratiosfr\
+  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \ratiosfr$\pm$\uncratiosfr\
   over the past 3\,Gyr.  We observe this trend both in known planet
@@
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-The implication for planetary ages can appreciated by querying the NASA Exoplanet Archive
+The implication for known planetary ages can appreciated by querying the NASA Exoplanet Archive
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
@@
 planets are in clusters, and were found by Kepler, K2, or TESS.
-However, these surveys have cumulatively discovered a total of $\approx$5{,}000 planets.
+However, these surveys have cumulatively discovered \deleted{a total of }$\approx$5{,}000 planets.
 Assuming a constant star formation history, we would expect an
-order of magnitude more sub-gigayear transiting planets.
+order of magnitude more sub-gigayear transiting planets\added{ than the 50 currently known}.
 
@@
 We calculated gyrochrone ages using \texttt{gyro-interp}
-\citep{Bouma_2023}, which is a method designed to address the fact
+\citep[\added{\texttt{v0.6};}][]{Bouma_2023}, which is a method designed to address the fact
 that stars with the same mass and same rotation period can have a wide
@@
 posteriors to calculate any summary statistics.  The oldest cluster
-for which \texttt{gyro-interp} is calibrated is the 4\,Gyr M67 cluster
+for which \texttt{gyro-interp} is calibrated\added{ in \texttt{v0.6}} is the 4\,Gyr M67 cluster
 \citep[][]{Dungee_2022,Gruner_2023}.  For \deleted{Kepler }stars with rotation
-periods above the M67 sequence, our resulting age constraints are\added{ therefore }
-quoted as lower limits.
-\added{Finally, we opted to impose a $t_{\rm gyro}$ uncertainty floor of 10\%. 
-This seemed warranted given the expected amplitude of differential rotation \citep[e.g.][]{Epstein_2014}, and
-given the accuracy with which the absolute cluster age scale is currently known \citep[e.g.][]{Dahm_2015}. }
+periods above the M67 sequence, our resulting age constraints are\added{ therefore} quoted as lower limits.
+
+\added{Finally, we opted to modify \texttt{gyro-interp} in \texttt{v0.6} to
+incorporate an age-dependent population-level Gaussian scatter, as detailed in the change log at \url{gyro-interp.readthedocs.io}.
+While age precisions at $\leq$1\,Gyr are unaffected, this update
+accounts for the observed intrinsic scatter in
+Ruprecht-147 and M67 being larger than in Praesepe or NGC-6811,
+potentially due to differential rotation \citep[e.g.,][]{Epstein_2014}.  
+For $\approx$2-4\,Gyr old stars with rotation periods measured at $\approx$5\% precision, 
+this change yields a $\lesssim$50\% increase in age uncertainties.}
 
@@
 rotation period that implies $t_{\rm
-gyro}$=\kepseveneightsixgyro\,Myr.  Other age indicators in the
+gyro}$$\approx$4.4,Gyr.  Other age indicators in the
 spectrum similarly suggest that the star is not young.  It 
@@
 the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
-the referee correctly pointed out that this only a comparative test of methods,
+the referee correctly pointed out that this is only a comparative test of methods,
 rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
@@
 longer rotation periods.  In this scenario, our analysis would overestimate the
-rate by which the SFR in the Kepler field has decreased, by 7-14\%. }
+degree by which the SFR in the Kepler field has decreased, by 7-14\%. }
 
@@
     billion years, the age distribution of stars in the Kepler field
-    implies that we should instead expect $\approx$250.
+    \replaced{implies}{suggests} that we should instead expect $\approx$250.
   %

### 2024-08-21 drivers/make_star_table.py

@@ -184,8 +184,6 @@ def make_star_table(
     sdf.to_csv(csvpath, index=False, na_rep='')
     print(f'Wrote {csvpath}')
 
-    import IPython; IPython.embed()
-
     csvpath = join(TABLEDIR, 'table_star_gyro_allcols.csv')
     odf = df.sort_values(
         by=['gyro_median','KIC'],

### 2024-08-22 papers/paper/ms.tex
update ms

@@
 general aim is to flag stars for which there is evidence based on
-their location in the \replaced{HR}{Kiel} diagram that our gyrochronology model may not
+their location in the \replaced{HR}{HR or Kiel} diagram\added{s} that our gyrochronology model may not
 be reliable.  While the $\log g$--$T_{\rm eff}$ cut flags about half
@@
   Bit 9: Far from main sequence?
-  Bit 10: \citetalias{Santos_2021} CP/CB?
-  Bit 11: $P_{\rm rot}$ not in the homogeneous
+  Bit 10: Spectroscopic $|[{\rm Fe/H}]|$$>$0.3?
+  Bit 11: \citetalias{Santos_2021} CP/CB?
+  Bit 12: $P_{\rm rot}$ not in the homogeneous
   \citetalias{Santos_2019} \citetalias{Santos_2021} sample?
-  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 10 and bit 7.
+  As an example, Kepler-1627 has $Q_{\rm star}$ flagged with bit 11 and bit 7.
   The analogous planet quality bitmask, $Q_{\rm planet}$, has the
@@
   For a star to have a high likelihood of being ``reliable for
-  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--9
+  gyrochronology'' we suggest a $Q_{\rm star}$ bitmask with bits 0--10
   not raised, and for a planet to be ``reliable'', we suggest

### 2024-08-23 drivers/plot_hist_field_gyro_ages.py
age hist: add dropfracshortrot

@@ -21,6 +21,10 @@ maxages = [4000]
 require_santosonly = [False]
 
 ap.plot_hist_field_gyro_ages(
+    outdir, cache_id, MAXAGE=4000, s19s21only=False,
+    datestr=datestr, dropfracshortrot=1
+)
+ap.plot_hist_field_gyro_ages(
     outdir, cache_id, MAXAGE=4000, s19s21only=0, preciseagesonly=1,
     datestr=datestr
 )

### 2024-08-23 gyrojo/plotting.py

@@ -1379,7 +1379,7 @@ def fit_line_and_print_results(bin_centers, heights, poisson_uncertainties):
 def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
                               datestr='20240821', s19s21only=0,
                               preciseagesonly=0, cache_id1=None, datestr1=None,
-                              dropfraclongrot=0):
+                              dropfraclongrot=0, dropfracshortrot=0):
 
     from gyrointerp.paths import CACHEDIR
     csvdir = join(CACHEDIR, f"samples_field_gyro_posteriors_{datestr}")
@@ -1457,6 +1457,19 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
         santosstr = ''
     if flag_mcq14_comp:
         skdf1 = kdf1[(kdf1.flag_is_gyro_applicable)]
+
+    if dropfracshortrot:
+        N_before = len(skdf)
+        # Randomly drop 5% of total star count, from the subset with Prot<10.
+        N_to_drop = int(0.05 * N_before)
+        filtered_skdf = skdf[skdf['Prot'] < 10]
+        print(N_before, N_to_drop, len(filtered_skdf))
+        frac = N_to_drop / len(filtered_skdf)
+        drop_indices = filtered_skdf.sample(frac=frac, random_state=42).index
+        skdf = skdf.drop(drop_indices).reset_index(drop=True)
+        N_after = len(skdf)
+        print(f'Droping {N_to_drop} stars, after reuqiring Prot<10days... N_after={N_after}')
+
 
     skdf['KIC'] = skdf.KIC.astype(str)
     mdf['KIC'] = mdf.KIC.astype(str)
@@ -1620,7 +1633,7 @@ def plot_hist_field_gyro_ages(outdir, cache_id, MAXAGE=4000,
     n_ob = len(mdf[sel_gyro_ok][
         (mdf[sel_gyro_ok].age > 2000) & (mdf[sel_gyro_ok].age <= 3000)
     ])/10
-    if not preciseagesonly and not flag_mcq14_comp and not dropfraclongrot:
+    if not preciseagesonly and not flag_mcq14_comp and not dropfraclongrot and not dropfracshortrot:
[…]

### 2024-08-23 papers/paper/ms.tex
update ms

@@
 % upon AAS submission
-%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
-\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
+\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
+%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
@@
 therefore flagged any stars with neighbors down to 1/10$^{\rm th}$ the
-brightness of the target star within this region ($\Delta G < 2.5$).
+brightness of the target star within this region ($\Delta G$$<$2.5).\added{ We also
+considered a deeper cut ($\Delta G$$<$5), and found it had negligible impact
+on our conclusions.}
 
@@
 associated with a transition from spot-dominated to faculae-dominated
-light curves \citep[e.g.][]{Reinhold2019}.
+light curves \citep[e.g.][]{Reinhold2019}\added{, or with rapid spin-down at Rossby numbers near 0.5 \citep{Lu2022}}.
 At $\approx$1\,Gyr, stalled spin-down yields larger uncertainties for
@@
 amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
-stars \replaced{would tend to }{could }contaminate the 2.5-3\,Gyr peaks in
+stars \replaced{would tend to }{could }contaminate the 2.3-3\,Gyr peaks in
 Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
@@
 the \citeauthor{McQuillan_2014} sample does not include many stars
-older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this analysis}{the analysis
-by \citet{2022ApJ...937...94M} fitted isochrones to a high-quality subset of Kepler data,
-and found that the
-sample of detected rotators from \citeauthor{McQuillan_2014} does not include many
-stars older than $\approx$\,4\,Gyr.}
-\replaced{would argue}{This suggests} that the 2.5-3\,Gyr peak \replaced{cannot}{is unlikely to} be caused by a pile-up of
-much older stars with especially lethargic spin-down rates.
+older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this
+analysis}{this pileup is thought to happen at $>$5\,Gyr for
+late-G dwarfs and early K dwarfs \citep[see Figure~4
+by][]{2024ApJ...962..138S}.  This means that it cannot affect the
+2.3\,Gyr peak in the age distribution that we see for 4400-5400\,K stars.   
+Moreover, while some contamination by older stars may be present in the peaks,
+the analysis by \citet{2022ApJ...937...94M} fitted
+isochrones to a high-quality subset of Kepler data, and found that the
+\citeauthor{McQuillan_2014} sample does not
+include many stars older than $\approx$\,4\,Gyr.}
+\replaced{would argue}{These lines of evidence suggest} that the peak
+\replaced{cannot be}{in the age distribution is not} caused by a pile-up of
+much older stars with especially lethargic spin-down rates\added{, but that it could be
+contaminated by them, especially for stars hotter than the Sun}.
 

### 2024-08-26 papers/paper/ms.tex
update ms

@@
     \subfloat{
-        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000.pdf}
+        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
     }
@@
     \subfloat{
-        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240530_maxage4000_preciseagesonly.pdf}
-    }
+        \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
+    }
+%
+%	\vspace{-0.35cm}
+%	\subfloat{
+%		\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
+%	}
   \end{center}
@@
   impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
-  evolutionary state; transparent lines do not.  The bottom row shows
+  evolutionary state; transparent lines do not\added{ (see
+  Section~\ref{subsec:flags})}.  The \replaced{bottom}{middle} row shows
   stars with temperatures of 4400--5400\,K, which \replaced{have more precise
@@
   mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
-  row.  Finally, the right panel compares the rotation-derived ages
+  row.  
+  Finally, the right panel compares the rotation-derived ages
   against star formation histories derived using CMD fitting (dashed
@@
 
-\subsection{A Paucity Of Young Stars And Planets}
+\subsection{Rotation-Based Age Distribution}
+
+\subsubsection{A Paucity Of Young Stars And Planets}
 
@@
 detect due to the photometric variability of their host stars.  If
-true, this statement seems to hold only for the very youngest
+true, this statement \replaced{seems to}{might} hold only for the very youngest
 ($\lesssim$0.4\,Gyr) stars, where there is a marginal deficit in the
@@
 than at $t_{\rm gyro}$$>$0.4\,Gyr.
-
-An important separate selection effect concerns Kepler's ability to
-detect the rotation signals.  \citet{2022ApJ...937...94M} studied this
-issue, and found for Sun-like stars that the fraction of rotation
-signals detectable by Kepler is near unity up to $\approx$3\,Gyr, and
-that it drops to almost zero by $\approx$5\,Gyr.  This trend is
-opposite in functional form to our derived age distributions\replaced{.}{and
-	most of the stars in our sample are indeed G dwarfs (see e.g.,~the $t_{\rm gyro}$--$T_{\rm eff}$ distributions 
-	in Appendix~\ref{app:age_diagnostic}).}
 
@@
 detected planets in the 0-1\,Gyr, 1-2\,Gyr, and 2-3\,Gyr bins.
+
+\subsubsection{Nuances in the Age Distribution}
+
+\added{{\it Completeness}---Figure~\ref{fig:hist_tgyro} has a gray
+overlay beyond $\gtrsim$3\,Gyr because Kepler's sensitivity to
+rotation signals diminishes at old ages.} \citet{2022ApJ...937...94M}
+studied \deleted{this issue}\added{Kepler's completeness to rotation
+signals using MIST-derived isochrone ages}, and found for Sun-like
+stars that the fraction of rotation signals \added{that are
+}detectable \deleted{by Kepler }is near unity up to $\approx$3\,Gyr,
+and that it drops to almost zero by $\approx$5\,Gyr.  \deleted{This trend is
+opposite in functional form to our derived age
+distributions.}\added{This estimate suggests that while aspects of
+Figure~\ref{fig:hist_tgyro} might be interpretable in terms of the Galaxy's star
+formation history at $\lesssim$3\,Gyr, at older ages incompleteness
+likely precludes any such interpretation.}
+
+\added{{\it Temperature sensitivity}---There are quantitative differences
+in Figure~\ref{fig:hist_tgyro} between the
+K dwarfs (4400--5400\,K) and the FGK stars (3800--6200\,K).  For instance,
+the distribution peaks at 2.3\,Gyr for K dwarfs, and at 2.9\,Gyr for
+the FGK sample.  The peak is also sharper for the K dwarfs;
+its decrease from 2.3-3\,Gyr is in an age range in which we believe the
+rotation period catalog completeness to be near unity.  We
+discuss further details of the $t_{\rm gyro}$--$T_{\rm eff}$
+distributions in Appendix~\ref{app:age_diagnostic}, and propose possible
+explanations for these differences in Section~\ref{subsec:caveats}. }
 
@@
 
-In the entire planet sample (i.e.~without imposing any quality cuts)
+In the entire planet sample (i.e.,~without imposing any quality cuts)
 this yielded \allagesyesconsistent\ consistent cases,
@@
 rotation period that implies $t_{\rm
-gyro}$$\approx$4.4,Gyr.  Other age indicators in the
+gyro}$$\approx$4.4\,Gyr.  Other age indicators in the
 spectrum similarly suggest that the star is not young.  It 
@@
 Our most constraining ages tend to come from one source of
-information: rotation.  Factors other than stellar age and mass can
-influence stellar rotation rates.  Binarity is one example: even
-intermediate-separation ($\sim$10\,AU) binaries are biased toward
-rapid rotation \citep[e.g.][and many studies thereafter]{Meibom_2007}.
+information: rotation.  Factors other than \deleted{stellar }age and mass can
+influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
+metallicity, differential rotation, and tidal spin-up, while observational
+challenges include the difficulty of measuring small-amplitude,
+long-period signals that are only quasi-coherent due to the starspot
+lifetime approaching the stellar rotation period.  In the following, we attempt to
+quantify the importance of these effects in our analysis, ordered by
+what we perceive to be their descending importance.}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+%		\subfloat{
+%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
+%		}
+%		
+%		\vspace{-0.35cm}
+%		\subfloat{
+%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
+%		}
+		%
+		%	\vspace{-0.35cm}
+			\subfloat{
+				\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
+			}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{{\bf Variant of Figure~\ref{fig:hist_tgyro}}, with rotation-derived age distributions that
+			correct for the estimated (estm.) effect of unresolved $\lesssim$0.5\,AU binaries. 
+		Even after applying our quality cuts (Section~\ref{subsec:flags}), up to
+		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
+		$P_{\rm rot}$$<$10\,days could be unresolved binaries.
+		These distributions correct for this possibility by randomly removing
+		what we expect to be the maximum possible number of such systems; the actual
+		correction relative to Figure~\ref{fig:hist_tgyro} might be smaller.
+		\label{fig:hist_tgyro_corr}
+	}
+\end{figure*}
+
+{\it Binaries}:
+\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
+are biased toward rapid rotation \citep[e.g.][and many studies
+thereafter]{Meibom_2007}.}{Binaries are often biased
+toward rapid rotation, even at intermediate separations of $\lesssim$10\,AU \citep[e.g.][]{Meibom_2007,Gruner_2023}.}
+\added{\citet{2019ApJ...871..174S} for instance found that 59\% of
+Kepler stars with rotation periods between 1.5 and 7\,days lie
+0.3\,mag above the main sequence, compared with 28\% of their full
+rotation sample.
+We attempted to mitigate the effects of binarity with five distinct
+quality flags (Section~\ref{subsec:flags}), and this attempt was at least
+somewhat effective, based on the peak near zero in the raw
+distributions of Figure~\ref{fig:hist_tgyro} being flattened in the
+cleaned distributions.
+However, the sensitivity of our quality flags is hard to assess.  One specific concern is that these
+quality flags might have minimal sensitivity to binaries with
+separations of 0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
+of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
+orbital periods of astrometric binaries in Gaia DR3 \citep{GaiaCollaboration2023}.  }
+
+\added{ If we were not at all sensitive to
+binaries with separations of 0.05-0.5\,AU, then based on
+the binary fraction and separation distribution from
+\citet{Raghavan2010}, as much as $\approx$5\% of our ``gyro
+applicable'' sample
+(\fivepctnuniqstarsantosrotgyroappl/\nuniqstarsantosrotgyroappl\
+stars) might be contaminated.  In a worst case scenario, all of
+these unidentified binaries might be able to induce $P_{\rm rot}$$<$10
+days, which would then be mixed in with the bona fide young, single stars.
+In our actual sample of apparently clean stars,
+\ngyroappllttendays/\nuniqstarsantosrotgyroappl\ had $P_{\rm
+rot}$$<$10\,days.
+
+Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
+calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
+$P_{\rm rot}$$<$10\,day stars.
+There are three important aspects of this result, relative to the original
+distributions in Figure~\ref{fig:hist_tgyro}.
+The first is that it would imply a much larger drop in the star formation rate,
+by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr,
+over the past three gigayears.
+The second is that it allows for a constant rate of decline in star formation
+since $\approx$2.8\,Gyr ago, rather than having a ``kink'' at 1.5\,Gyr,
+as in Figure~\ref{fig:hist_tgyro}.
+And finally, these corrected distributions show closer agreement with SFHs
+derived from CMD fitting.
+We caution however that this correction assumes that {\it none} of our
+quality flags were able to identify any of these unresolved binaries.
+Since we did in fact include flags based on e.g.~location in the HR diagram,
+this correction is best interpreted as an upper limit on the possible impact of binarity.}
+
+%TODO TODO TODO
+\added{{\it Erroneous period measurements:} }
+
+\added{
+{\it Metallicity:}
+}
+
+\added{
+{\it Differential rotation:}
+}
+
+%TODO TODO TODO
+%TODO TODO TODO
+%TODO TODO TODO
+
 Metallicity may also be relevant
@@
 
-A broader question, beyond the scope of this work, concerns the ages
-of stars $\gtrsim$4\,Gyr old.  For age-dating models based only on
-rotation rates, an important challenge is that the detectability of
-the rotation signals for 3-10\,Gyr stars, especially Sun-like stars,
-is at the limits of the Kepler data \citep{2022ApJ...937...94M}.  In
-the ``old star'' regime, methods that leverage either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both the
-evolution of stellar luminosity and stellar rotation
-\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most capable of
-providing useful age constraints for single stars over the full
-span of their main-sequence lifetimes.
 %Incorporating
@@
 
-{\it Occurrence Rates}:
+\added{{\it Ages of stars older than the Sun}: }\deleted{A broader question, beyond
+the scope of this work, concerns the ages of stars $\gtrsim$4\,Gyr
+old.}\added{This work was limited to stars younger than 4\,Gyr}. For
+age-dating models based only on rotation rates, an important challenge
+is that the detectability of the rotation signals for 3-10\,Gyr stars,
+especially Sun-like stars, is at the limits of the Kepler data
+\citep{2022ApJ...937...94M}.  In the ``old star'' regime, methods that
+leverage either asteroseismology
+\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
+the evolution of stellar luminosity and stellar rotation
+\citep{Angus_2019,Claytor2020,2023ApJ...952..131M} seem the most
+capable of providing useful age constraints for single stars over the
+full span of their main-sequence lifetimes.
+
+{\it \added{Planet} occurrence Rates}:
 The age-dependent trends predicted for exoplanet populations, such as

### 2024-08-26 papers/paper/ms.tex
update ms (caveat discn rework)

@@
 
-\subsubsection{A Paucity Of Young Stars And Planets}
+\subsubsection{Kepler's Demographic Cliff}
 
@@
 \citep{Masuda2022infer}, and found similar results.  The plotted
-uncertainties assume Poisson statistics.  We truncated the plots
+uncertainties \replaced{assume Poisson statistics}{are Poissonian}.  We truncated the plots
 at $4$\,Gyr, which is the upper bound for our
-rotational age calibration.  The overall slope yields a paucity of
+rotational age calibration.  The overall slope \replaced{yields}{shows} a paucity of
 young stars relative to the expectation of a uniform age distribution.
@@
 rotation period catalog completeness to be near unity.  We
-discuss further details of the $t_{\rm gyro}$--$T_{\rm eff}$
+show further details of the $t_{\rm gyro}$--$T_{\rm eff}$
 distributions in Appendix~\ref{app:age_diagnostic}, and propose possible
@@
 
-Most of the stars observed by Kepler are in the thin disk.  This can
+Most \deleted{of the }stars observed by Kepler are in the thin disk.  This can
 be verified by following \citet{Gaia_2018}, and labeling stars with 2D
@@
 and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
-For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
-disk stars respectively.  The thin disk stars have a threefold larger
+\replaced{For all stars in the KIC, this yields }{These criteria yield }95{,}694 thin and 54{,}039 thick
+disk stars \replaced{respectively}{observed by Kepler}.  The thin disk stars have a threefold larger
 detected rotation period fraction: 35{,}675 rotators are in the thin
 disk, while 7{,}312 are in the thick disk.  Further imposing the
-gyrochronology quality flags discussed in Section~\ref{subsec:flags}
-yields 17{,}755 thin-disk stars for which gyrochronology is
-applicable, and 2{,}462 thick-disk stars.
+\deleted{gyrochronology }quality flags discussed in Section~\ref{subsec:flags}
+yields 17{,}755 thin disk stars for which gyrochronology is
+applicable, and 2{,}462 thick disk stars.
 
@@
 \citet{SilvaAguirre2018}.  The main novelty of our ages relative to
-those studies is their improved accuracy at $<$1\,Gyr.
+\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.
 
@@
 \label{subsec:caveats}
+
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\subfloat{
+			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
+		}
+	\end{center}
+	\vspace{-0.6cm}
+	\caption{{\bf Potential impact of unresolved binaries.}   This figure shows
+		the same raw (thin blue line) and quality-flag cleaned (``gyro applicable'') distributions
+		as in Figure~\ref{fig:hist_tgyro}.
+		However, even after applying our quality cuts (Section~\ref{subsec:flags}), up to
+		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
+		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries.
+		The thick blue ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
+		what we estimate (estm.) to be the maximum possible number of such systems; the true
+		correction might be smaller.  We expect such a correction to be more important in
+		the sample of all Kepler stars than in the KOI host sample (see text).
+		\label{fig:hist_tgyro_corr}
+	}
+\end{figure*}
 
@@
 influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
-metallicity, differential rotation, and tidal spin-up, while observational
-challenges include the difficulty of measuring small-amplitude,
+metallicity, differential rotation, mergers, and tidal spin-up.  Observational
+limitations include the difficulty of measuring small-amplitude,
 long-period signals that are only quasi-coherent due to the starspot
-lifetime approaching the stellar rotation period.  In the following, we attempt to
+lifetime approaching the stellar rotation period.  There is also the issue that
+our age-dating approach is not calibrated beyond 4\,Gyr.
+In the following, we attempt to
 quantify the importance of these effects in our analysis, ordered by
 what we perceive to be their descending importance.}
-
-\begin{figure*}[!t]
-	\begin{center}
-		\leavevmode
-%		\subfloat{
-%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf}
-%		}
-%		
-%		\vspace{-0.35cm}
-%		\subfloat{
-%			\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_preciseagesonly.pdf}
-%		}
-		%
-		%	\vspace{-0.35cm}
-			\subfloat{
-				\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
-			}
-	\end{center}
-	\vspace{-0.6cm}
-	\caption{{\bf Variant of Figure~\ref{fig:hist_tgyro}}, with rotation-derived age distributions that
-			correct for the estimated (estm.) effect of unresolved $\lesssim$0.5\,AU binaries. 
-		Even after applying our quality cuts (Section~\ref{subsec:flags}), up to
-		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
-		$P_{\rm rot}$$<$10\,days could be unresolved binaries.
-		These distributions correct for this possibility by randomly removing
-		what we expect to be the maximum possible number of such systems; the actual
-		correction relative to Figure~\ref{fig:hist_tgyro} might be smaller.
-		\label{fig:hist_tgyro_corr}
-	}
-\end{figure*}
 
@@
 are biased toward rapid rotation \citep[e.g.][and many studies
-thereafter]{Meibom_2007}.}{Binaries are often biased
-toward rapid rotation, even at intermediate separations of $\lesssim$10\,AU \citep[e.g.][]{Meibom_2007,Gruner_2023}.}
+thereafter]{Meibom_2007}.}{Stars in binaries are often biased
+toward rapid rotation, even beyond the typical
+realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.}
 \added{\citet{2019ApJ...871..174S} for instance found that 59\% of
@@
 We attempted to mitigate the effects of binarity with five distinct
-quality flags (Section~\ref{subsec:flags}), and this attempt was at least
+quality flags (Section~\ref{subsec:flags}).   This attempt was at least
 somewhat effective, based on the peak near zero in the raw
@@
 cleaned distributions.
-However, the sensitivity of our quality flags is hard to assess.  One specific concern is that these
-quality flags might have minimal sensitivity to binaries with
-separations of 0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
+However, the completeness of our quality flags is hard to assess.  One specific concern is that these
+flags might have minimal sensitivity to binaries with
+separations of $\approx$0.05-0.5\,AU, for all but the highest mass ratios.  This inner limit is set by the onset
 of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
-orbital periods of astrometric binaries in Gaia DR3 \citep{GaiaCollaboration2023}.  }
+orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.  }
 
@@
 stars) might be contaminated.  In a worst case scenario, all of
-these unidentified binaries might be able to induce $P_{\rm rot}$$<$10
+these unidentified binaries might induce $P_{\rm rot}$$<$10
 days, which would then be mixed in with the bona fide young, single stars.
-In our actual sample of apparently clean stars,
-\ngyroappllttendays/\nuniqstarsantosrotgyroappl\ had $P_{\rm
+This would constitute a major source of contamination,
+given that our nominal sample of apparently clean stars had
+\ngyroappllttendays\ stars with $P_{\rm
 rot}$$<$10\,days.
@@
 $P_{\rm rot}$$<$10\,day stars.
-There are three important aspects of this result, relative to the original
-distributions in Figure~\ref{fig:hist_tgyro}.
-The first is that it would imply a much larger drop in the star formation rate,
-by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr,
-over the past three gigayears.
-The second is that it allows for a constant rate of decline in star formation
-since $\approx$2.8\,Gyr ago, rather than having a ``kink'' at 1.5\,Gyr,
-as in Figure~\ref{fig:hist_tgyro}.
-And finally, these corrected distributions show closer agreement with SFHs
+Relative to the nominal ``gyro applicable'' age distribution, there are three important differences.
+There is a much larger drop in the star formation rate over the past three gigayears,
+by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr.
+The rate of decline in star formation
+since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
+And finally, the corrected distributions show closer agreement with SFHs
 derived from CMD fitting.
@@
 quality flags were able to identify any of these unresolved binaries.
-Since we did in fact include flags based on e.g.~location in the HR diagram,
+Since we did in fact include flags based on e.g.,~location in the HR diagram,
 this correction is best interpreted as an upper limit on the possible impact of binarity.}
 
-%TODO TODO TODO
-\added{{\it Erroneous period measurements:} }
-
-\added{
-{\it Metallicity:}
-}
-
-\added{
-{\it Differential rotation:}
-}
-
-%TODO TODO TODO
-%TODO TODO TODO
-%TODO TODO TODO
-
-Metallicity may also be relevant
+\added{{\it Rotational outliers:} }
+We were interested in not only the overall age distribution of the
+Kepler field, but also in the reliability of individual ages for \replaced{young
+stars known to host planets}{each star in our sample}.  We attempted to assess this reliability using
+quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
+between rotation and lithium ages.  The \deleted{open }cluster comparison
+suggested that our ages were generally accurate at $\ll$0.5\,dex
+across 0.04--2.5\,Gyr, similar to their quoted precision.
+Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
+does show outliers, typically $\lesssim$10\% of the population
+in any cluster.  \added{While most of these outliers are known to have bad quality flags, some do not.} 
+These stars could be either field star contaminants,
+or else anomalous stars whose physical rotation histories were altered
+by processes not captured by our statistical uncertainties.
+Similarly, although we found consistent $t_{\rm gyro}$ and $t_{\rm Li}$
+for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
+stars, we did find one system, Kepler-786, with radically different
+lithium (\kepseveneightsix\,Myr) and rotational
+($\approx$4.4\,Gyr) ages.  Planetary mergers
+are one process that could produce such a signal, but
+testing this would require a method for measuring differential abundances
+across a large number of elements.
+
+\added{A separate plausible origin for rotational outliers is that they come from
+	erroneously reported rotation periods.  While
+	the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
+	the referee correctly pointed out that this is only a comparative test of methods,
+	rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
+	One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
+	with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
+	longer rotation periods.  In this scenario, a similar analysis to that in Figure~\ref{fig:hist_tgyro_corr} showed
+	that our analysis would overestimate the
+	degree by which the SFR in the Kepler field has decreased, by 7-14\%. }
+
+\added{Other age indicators may help in verifying the ages of the
+$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
+chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
+can serve as an age tracer
+\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
+However, these age indicators are all in a sense ``rotation-powered''.
+The dynamo converts kinetic energy into magnetic energy, which is
+emitted through these chromospheric pathways.  We did not attempt to
+incorporate these indicators into this study due to concerns regarding
+sensitivity, homogeneity, and the question of whether they in fact
+provide age information that is truly independent from rotation.}
+
+\added{{\it Weakened magnetic braking:} 
+If we lived in a
+universe in which stellar rotation rates and spot-induced photometric
+amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
+stars could contaminate the peaks in
+Figure~\ref{fig:hist_tgyro}.  This
+scenario deserves consideration, given asteroseismic evidence for slowed spin-down rates after
+$\approx$4\,Gyr for Sun-like stars
+\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the observation of a ``long-period edge''
+	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
+	may pile up over similar timescales \citep{David_2022}.  However, 
+	this pileup is thought to happen at $>$5\,Gyr for
+	late-G dwarfs and early K dwarfs \citep[see Figure~4
+	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
+	2.3\,Gyr peak that we see in the 4400-5400\,K age distribution in Figure~\ref{fig:hist_tgyro}. 
+	Independently,
+	the analysis by \citet{2022ApJ...937...94M} fitted
+	isochrones to a high-quality subset of Kepler data, and found that the
+	\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly}, does not
+	include many stars older than $\approx$\,4\,Gyr.
+These lines of evidence suggest that the peak
+in the observed age distribution is not caused by a pile-up of
+much older stars with especially lethargic spin-down rates, but that it could be
+	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
+}
+
+\added{{\it Metallicity:}  
+Stars with higher metallicity are thought to have longer convective turnover
+times, which may slow their rate of spin-down
+	\citep{Amard2020}.
+While observational studies of this effect have been suggestive, further work
+remains to definitively establish the importance of the effect \citep[e.g.][]{Amard2020b,2024arXiv240500779S}
+Nonetheless, based on the theoretical predictions, we flagged stars with known
+	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, based on either LAMOST or Keck/HIRES
+	spectra \citep{Zong2018,Petigura_2022}.
+The main limitation of this approach is that only 53.3\% of the stars in Table~\ref{tab:stars} with
+$t_{\rm gyro}$$<$4\,Gyr had such spectra available.
+For the rest of the sample, primarily composed of fainter stars,
+this means that our quality flags were not sensitive to metallicity outliers.
+Based on the metallicity distribution of stars for which we do have spectra,
+this 
+implies that at most $\approx$5\% of stars in our sample might have $|[{\rm Fe/H}]|$$>$0.3,
+with a slight bias toward lower metallicity stars over higher metallicity.
+Based on the results by \citet{Claytor2020},  this will be most important
+for stars near the Kraft break, which could have ages over-estimated
+by $\approx$50-100\% if they are in fact metal-poor, and under-estimated in the converse case.
+}
+
+\added{{\it Differential rotation:}
+The Sun's surface rotates once per $\approx$25-30\,days at latitudes between the equator
+and $\approx$$\pm$60$^\circ$ \citep{Snodgrass1990}.
+This latitudinal surface differential rotation yields two separate limitations on our rotational ages.
+The first is
+on the accuracy with which any one star's rotation period can be measured over a few-year mission like Kepler;
+the second is on the degree to which the intrinsic scatter in open cluster rotation sequences can be measured \citep[e.g.][]{Epstein_2014}.
+In this study, we accounted for the latter intrinsic scatter based on available measurements
+from Ruprecht-147 \citep{Curtis_2020} and M67 data
+\citep{Barnes2016,Dungee_2022,Gruner_2023}.
+This means that irrespective of the assumed $P_{\rm rot}$ measurement precision,
+our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter (perhaps differential-rotation induced)
+at old ages.
+While this accounts for the statistical impact of differential rotation on the inferred
+age distribution,
+the accuracy issue remains;
+any one individual star's observed rotation period might be faster
+or slower than average stars of the same mass and age.  
+This effect can yield a $\approx$20\% age bias for individual Sun-like stars with ages
+of a few billion years; the statistical uncertainties in Table~\ref{tab:stars} should be
+treated with appropriate caution.
+}
+
+\deleted{Metallicity may also be relevant
 \citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
@@
 estimated planet ages might shift if one were to account for the
-spectroscopic information.
-
-We were interested in not only the overall age distribution of the
-Kepler field, but also in the reliability of individual ages for young
-stars known to host planets.  We attempted to assess this reliability using
-quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }
-cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
-between rotation and lithium ages.  The \deleted{open }cluster comparison
-suggested that our ages were generally accurate at $\ll$0.5\,dex
-across 0.04--2.5\,Gyr, similar to their quoted precision.
-Nonetheless, the upper-left panel of Figure~\ref{fig:agescalecompone}
-does show outliers, typically $\lesssim$10\% of the population
-in any cluster.  These stars could be either field star contaminants,
-or else anomalous stars whose physical rotation histories were altered
-by processes not captured by our statistical uncertainties.
-Similarly, although we found $t_{\rm gyro}$ and $t_{\rm Li}$ to agree
-for \ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
-stars, we did find one system, Kepler-786, with radically different
-lithium (\kepseveneightsix\,Myr) and rotational
-($\approx$4.4\,Gyr) ages.  Planetary mergers
-are one process that could produce such a signal, but
-testing this would require a method for measuring differential abundances
-across a large number of elements.
-
-\added{A separate plausible origin for rotational outliers is that they come from
-erroneously reported rotation periods.  While
-the large surveys do show good internal consistency (Appendix~\ref{app:mcqonly}),
-the referee correctly pointed out that this is only a comparative test of methods,
-rather than an ability to recover ground truth \citep[e.g.][]{Aigrain2015}.
-One could imagine an adversarial scenario in which, say, 10-20\%\ of stars
-with $P_{\rm rot}$$>$20\,days are erroneously reported, and in fact have much
-longer rotation periods.  In this scenario, our analysis would overestimate the
-degree by which the SFR in the Kepler field has decreased, by 7-14\%. }
-
-Other age indicators may help in verifying the ages of the
+spectroscopic information.}
+
+\deleted{Other age indicators may help in verifying the ages of the
 $\lesssim$3\,Gyr stars that are the \added{main }focus of this work.  Specifically,
@@
 sensitivity, homogeneity, and the question of whether they in fact
-provide age information that is truly independent from rotation.
+provide age information that is truly independent from rotation.}
 
@@
 old.}\added{This work was limited to stars younger than 4\,Gyr}. For
-age-dating models based only on rotation rates, an important challenge
+age-dating methods based only on rotation rates, an important challenge
 is that the detectability of the rotation signals for 3-10\,Gyr stars,
@@
 Some of the youngest stars in Tables~\ref{tab:planets}
-and~\ref{tab:stars} have been linked to their birth clusters, while
-others have not. A systematic search for the birth clusters of the
+and~\ref{tab:stars} have been linked to their birth clusters.
+Others have not. A systematic search for the birth clusters of the
 youngest ``field stars'' could combine positions and velocities from
@@
 all 14 known Kepler planets in clusters.  While lithium provided
-minimal added information for most of the sample, $t_{\rm Li}$ and
+minimal added information for most of the sample\added{ due to its astrophysical
+rate of depletion}, $t_{\rm Li}$ and
 $t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
-was possible.  Our resulting ages included two-sided $t_{\rm gyro}$
+was possible.  Our \replaced{resulting ages}{results} included two-sided $t_{\rm gyro}$
 and $t_{\rm Li}$ for \ltonegyrhighqconfirmedtwosided\ sub-gigayear
@@
 evolutionary connection to their youth.  Other discoveries include
-Earth-sized planets with new ages of only a few hundred million years.
+Earth-sized planets \added{such as Kepler-1312~c and Kepler-1561~b }with new ages of only a few hundred million years.
 
@@
     Kepler target stars and the known Kepler planet-hosts show a
-    demographic cliff.  There are twice as many ``old'' (2-3\,Gyr)
+    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
     stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
-    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
-    than it was three billion years ago.  This result from
-    rotation-based ages broadly agrees with recent reports of a
+    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
+    than it was three billion years ago\added{.
+The statistical uncertainty on this figure is far smaller than the systematic uncertainty caused by the potential presence of unresolved photometric binaries (Section~\ref{subsec:caveats}); the star formation rate could have decreased by a factor of as much as 4.81$\pm$0.26}.
+\replaced{This}{Either} result \deleted{from
+    rotation-based ages }broadly agrees with recent reports of a
     declining star-formation rate from CMD fitting and white-dwarf
-    chronology, though with the advantage of better sensitivity for
+    chronology, though with the \added{potential }advantage of better sensitivity for
     FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
@@
 \acknowledgements
-This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
+\added{We thank the reviewer for their suggestions, which significantly improved the 
+quality of the manuscript.} This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
 (LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
@@
 
-This analysis is particularly interesting in relation to the question
-of how fast stars $\gtrsim$4\,Gyr old spin down.  If we lived in a
-universe in which stellar rotation rates and spot-induced photometric
-amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
-stars \replaced{would tend to }{could }contaminate the 2.3-3\,Gyr peaks in
-Figures~\ref{fig:hist_tgyro} and~\ref{fig:hist_tgyro_mcq}.  This
-scenario deserves consideration, given\replaced{ evidence from asteroseismic
-rotation periods that has pointed to}{ {\it i)} asteroseismic evidence for} slowed spin-down rates after
-$\approx$4\,Gyr for Sun-like stars
-\citep[e.g.][]{vanSaders_2016,2024ApJ...962..138S}\added{, and {\it ii)} the observation of a ``long-period edge''
-in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
-may pile up over similar timescales \citep{David_2022}}.  However, \replaced{because
-the \citeauthor{McQuillan_2014} sample does not include many stars
-older than $\approx$4\,Gyr \citep{2022ApJ...937...94M}, this
-analysis}{this pileup is thought to happen at $>$5\,Gyr for
-late-G dwarfs and early K dwarfs \citep[see Figure~4
-by][]{2024ApJ...962..138S}.  This means that it cannot affect the
-2.3\,Gyr peak in the age distribution that we see for 4400-5400\,K stars.   
-Moreover, while some contamination by older stars may be present in the peaks,
-the analysis by \citet{2022ApJ...937...94M} fitted
-isochrones to a high-quality subset of Kepler data, and found that the
-\citeauthor{McQuillan_2014} sample does not
-include many stars older than $\approx$\,4\,Gyr.}
-\replaced{would argue}{These lines of evidence suggest} that the peak
-\replaced{cannot be}{in the age distribution is not} caused by a pile-up of
-much older stars with especially lethargic spin-down rates\added{, but that it could be
-contaminated by them, especially for stars hotter than the Sun}.
-
 \begin{figure*}[!t]

### 2024-08-26 papers/paper/ms.tex
update ms

@@
   sub-gigayear systems can be attributed to the star formation rate in
-  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \ratiosfr$\pm$\uncratiosfr\
+  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{a few}
   over the past 3\,Gyr.  We observe this trend both in known planet
@@
 		\subfloat{
-			\includegraphics[width=0.47\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
 		}
 		\subfloat{
-			\includegraphics[width=0.47\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png}
 		}
@@
 rotation periods based on updated knowledge of contaminants
-(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
+(such as giant stars and eclipsing binaries) and visual inspection.  In
 addition, \citetalias{Santos_2021} reported rotation periods for
@@
 as periodic by \citetalias{McQuillan_2014}.  Many of these reported
-detections have lower variability amplitudes and longer periods than
+signals have lower variability amplitudes and longer periods than
 those reported by \citetalias{McQuillan_2014}.
@@
     a stellar ensemble in which the star is a candidate member.
-    Cluster membership was evaluated without knowledge of rotation.
+    Cluster membership was evaluated without knowledge of rotation\added{, and for this plot,
+    no attempt was made to exclude known binaries}.
     Although some field interlopers may be present in the membership
@@
 while 
-Theia~520 ($\approx$300\,Myr; \citealt{2019AJ....158..122K}), Melange-3 ($\approx$150\,Myr; \citealt{Barber_2022}), and Cep-Her ($\approx$40\,Myr; \citealt{Bouma_2022b} and Kerr et al.~2024 submitted) are more recent discoveries.
+Theia~520 ($\approx$300\,Myr; \citealt{2019AJ....158..122K}), Melange-3 ($\approx$150\,Myr; \citealt{Barber_2022}), and Cep-Her ($\approx$40\,Myr; \citealt{Bouma_2022b} and \citealt{Kerr2024}) are more recent discoveries.
 
@@
   rotation-derived age distributions of its stars (left) and planet
-  hosts (middle).  The top row shows all stars with temperatures of
-  3800--6200\,K for which we calculated rotational ages.  Opaque lines
+  hosts (middle).  The top row shows \replaced{all stars with temperatures of
+  3800--6200\,K for which we calculated rotational ages}{stars with temperatures of 3800--6200\,K}.  Opaque lines
   impose quality cuts on binarity, \added{metallicity, }crowding, and the star's
   evolutionary state; transparent lines do not\added{ (see
-  Section~\ref{subsec:flags})}.  The \replaced{bottom}{middle} row shows
+  Section~\ref{subsec:flags})}.  The bottom row shows
   stars with temperatures of 4400--5400\,K, which \replaced{have more precise
-  ages due to their fast spin-down}{often have more precise ages}.  The statistical uncertainty for a
+  ages due to their fast spin-down}{usually have more precise ages}.  The statistical uncertainty for a
   mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
@@
   and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$
-  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr
+  detection sensitivity is near unity until $t$$\lesssim$3\,Gyr\added{, and decreases rapidly at older ages}
   \citep{2022ApJ...937...94M}.
@@
     0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-    hosts to not have $Q_{\rm star}$ bits 0--9 raised, and for the
+    hosts to not have $Q_{\rm star}$ bits 0--\replaced{9}{10} raised, and for the
     planets to similarly no have quality flags raised. 
@@
 \citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
-selects these stars with a temperature cut, and may show a hint of the
-$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
-$\approx$1\,Gyr spike however is not recovered.  This
+selects these stars with a temperature cut, and may show a \added{weak }hint of the
+\added{$\approx$1\,Gyr and }$\approx$2\,Gyr spike\added{s} reported by \citet{2020NatAs...4..965R}.  \deleted{The
+$\approx$1\,Gyr spike however is not recovered. } This
 $\approx$2.5\,Gyr local maximum is similarly present in the isochrone
@@
 of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
-orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.  }
+orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.}
 
@@
 since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
-And finally, the corrected distributions show closer agreement with SFHs
-derived from CMD fitting.
+And finally, the corrected distributions show closer agreement with both the SFHs
+derived from CMD fitting, and also with the KOI host distribution,
+which is almost devoid of close binaries within 1\,AU \citep[e.g.][and references therein]{Moe2021}.
 We caution however that this correction assumes that {\it none} of our
@@
 	include many stars older than $\approx$\,4\,Gyr.
-These lines of evidence suggest that the peak
-in the observed age distribution is not caused by a pile-up of
-much older stars with especially lethargic spin-down rates, but that it could be
+These arguments suggest that the peak
+in the age distribution is not caused by a pile-up of
+much older stars with lethargic spin-down rates, but that it could be
 	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
@@
 evolutionary connection to their youth.  Other discoveries include
-Earth-sized planets \added{such as Kepler-1312~c and Kepler-1561~b }with new ages of only a few hundred million years.
+Earth-sized planets \added{(e.g.~Kepler-1312~c and Kepler-1561~b) }with \deleted{new }ages of only a few hundred million years.
 
@@
 The statistical uncertainty on this figure is far smaller than the systematic uncertainty caused by the potential presence of unresolved photometric binaries (Section~\ref{subsec:caveats}); the star formation rate could have decreased by a factor of as much as 4.81$\pm$0.26}.
-\replaced{This}{Either} result \deleted{from
-    rotation-based ages }broadly agrees with recent reports of a
+\replaced{This}{Either} result \replaced{from
+    rotation-based ages broadly agrees}{is generally consistent} with recent reports of a
     declining star-formation rate from CMD fitting and white-dwarf
-    chronology, though with the \added{potential }advantage of better sensitivity for
+    chronology, though with the \added{potential }advantage of better \replaced{sensitivity}{accuracy} for
     FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
@@
 \added{We thank the reviewer for their suggestions, which significantly improved the 
-quality of the manuscript.} This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
+quality of the manuscript.}
+This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
 (LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
@@
 Li}^{\ast}$=83$\pm$6\,m\AA, the lithium age of \kepseveneightsix\,Myr
-would predict an obvious rotation signal.  None is present in the
-Kepler light curve, consistent with the low $v\sin i$.  Out of all
+would predict \replaced{an obvious}{a short-period} rotation signal.  \replaced{None is present in the
+Kepler light curve}{The actual signal that is present has an amplitude of $\approx$0.1-0.3\% and a period of $\approx$33\,days}, consistent with the low $v\sin i$.  Out of all
 ``apparently discrepant'' lithium and rotation measurements discussed
@@
 		\subfloat{
-			\includegraphics[width=0.43\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+			\includegraphics[width=0.44\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
 		}
 		\subfloat{
-			\includegraphics[width=0.43\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
+			\includegraphics[width=0.44\textwidth]{prot_teff_McQuillan2014only_dquality.png}
 		}
 	\end{center}
-	\vspace{-0.5cm}
+	\vspace{-0.6cm}
 	\caption{
@@
   \caption{{\bf Variant of Figure~\ref{fig:hist_tgyro},} comparing
-  our default adopted catalog of \citet{Santos_2019,Santos_2021}
-  rotation periods against the rotation periods
+  our default adopted age distribution (left) based on the \citet{Santos_2019,Santos_2021}
+  rotation periods against the age distribution (middle) based only on rotation periods
   reported by \citet{McQuillan_2014}.

### 2024-08-26 papers/paper/collect.sh

@@ -12,11 +12,13 @@ cp $r/st_params/st_params_adopted_logg_vs_adopted_Teff.pdf .
 # ...G vs Teff
 #cp $r/st_params/st_params_dr3_phot_g_mean_mag_vs_adopted_Teff.pdf .
 
-# plot_star_Prot_Teff
-cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
+# mpl>3.6 plot_star_Prot_Teff (pdf rendering failing)
+#cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.pdf .
+cp $r/star_Prot_Teff/prot_teff_Santos19_Santos21_dquality.png .
 
-# plot_koi_mean_prot_teff
-cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
+# plot_koi_mean_prot_teff (to match pdf compat...)
+#cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf .
+cp $r/koi_mean_prot_teff/koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png .
 
 # plot_hist_field_gyro_ages 
 cp $r/hist_field_gyro_ages_20240821/comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000.pdf .
@@ -41,10 +43,12 @@ cp $r/gyroage_vs_teff/gyroage_vs_teff_errs_linear.png .
 
 cp $r/lit_ages_vs_cluster_ages/olympic_merged_lit_ages_vs_cluster_ages.pdf .
 
-cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-m14_Prot_vs_Prot.pdf .
-cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-r23_ProtGPS_vs_Prot.pdf .
+# mpl>3.6 pdf rendering failing
+#cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-m14_Prot_vs_Prot.pdf .
+#cp $r/perioddiff_vs_period/perioddiff_vs_period_diffProt-r23_ProtGPS_vs_Prot.pdf .
 
-cp $r/star_Prot_Teff/prot_teff_McQuillan2014only_dquality.pdf .
[…]

### 2024-08-27 papers/paper/ms.tex
update ms

@@
   the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{a few}
-  over the past 3\,Gyr.  We observe this trend both in known planet
-  hosts and in the parent stellar sample.
+  over the past 3\,Gyr.  We observe this trend both for known planet
+  hosts and for the parent stellar sample.
   %
@@
   %
-  \item We can identify binary stars with greater fidelity than
+  \item We can identify binary stars \added{and high-metallicity outliers }with greater fidelity than
     in the past, which can clarify otherwise problematic estimates of
@@
     the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
-    temperatures of 3800--6200\,K.  The planet-hosts (right) require
+    temperatures of 3800--6200\,K.  The planet hosts (right) require
     the same stellar cuts, and include only the confirmed and
@@
   \begin{center}
-    \includegraphics[width=0.48\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf}
+    \includegraphics[width=0.48\textwidth]{rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung_allowsgrazingandhighruwe.pdf}
   \end{center}
@@
     {\bf Sizes, orbital periods, and rotation-based ages of transiting
-    exoplanets younger than one billion years.} Star symbols denote
-    \nplyounggyrotwosigmanograzingnoruwe\ planets that are
-    characterized in this work to have $t_{\rm gyro}$<1\,Gyr at
-    $2$$\sigma$; circles are planets from the literature meeting the
-    same age requirement, but with a heterogeneous set of age
+    exoplanets younger than one billion years.}
+    \replaced{Star symbols denote \nplyounggyrotwosigmanograzingnoruwe\ planets
+    that are characterized in this work to have $t_{\rm gyro}$<1\,Gyr
+    at $2$$\sigma$; circles are planets from the literature meeting
+    the same age requirement, but with a heterogeneous set of age
     provenances.  Gray points are transiting planets from the NASA
@@
     0.5-1\,Gyr planets.  For this plot, we require our Kepler stellar
-    hosts to not have $Q_{\rm star}$ bits 0--\replaced{9}{10} raised, and for the
-    planets to similarly no have quality flags raised. 
+    hosts to not have $Q_{\rm star}$ bits 0--9 raised,
+    and for the planets to similarly no have quality flags
+    raised.}{Transiting systems are selected to have $t$<1\,Gyr at $2$$\sigma$.
+    Star symbols denote \nplyounggyrotwosigmanograzingnoruwe\ planets
+    characterized in this work to have such ages at the highest
+    confidence, meaning that none of $Q_{\rm star}$ bits 0--10 are raised,
+    and the planets have no quality flags
+    raised.  Transparent crosses denote \nplyounggrazingorhighruwe\
+    planets for which either the planet is grazing, or the star has
+    RUWE$>$1.4.  Gray points are from the NASA Exoplanet Archive. }
     \label{fig:rp_period_age_results}
@@
 Figure~\ref{fig:rp_period_age_results} shows the planets and planet
-candidates with $t_{\rm gyro}$$<$1\,Gyr at 2$\sigma$.  This figure
-includes \nconfirmedplyounggyrotwosigmanograzingnoruwe\ confirmed
+candidates with $t_{\rm gyro}$$<$1\,Gyr at 2$\sigma$.  \replaced{This}{The stars in this} figure
+include\deleted{s} \nconfirmedplyounggyrotwosigmanograzingnoruwe\ confirmed
 Kepler planets, and \ncandidateplyounggyrotwosigmanograzingnoruwe\
@@
 		as in Figure~\ref{fig:hist_tgyro}.
-		However, even after applying our quality cuts (Section~\ref{subsec:flags}), up to
+		However, even after applying our quality cuts, up to
 		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
-		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries.
+		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
 		The thick blue ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
@@
 		correction might be smaller.  We expect such a correction to be more important in
-		the sample of all Kepler stars than in the KOI host sample (see text).
+		the sample of all Kepler stars than in the KOI host sample.
 		\label{fig:hist_tgyro_corr}
@@
 {\it Binaries}:
-\replaced{Even intermediate-separation ($\sim$10\,AU) binaries
-are biased toward rapid rotation \citep[e.g.][and many studies
-thereafter]{Meibom_2007}.}{Stars in binaries are often biased
+Stars in binaries are often biased
 toward rapid rotation, even beyond the typical
-realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.}
+realm of tidal synchronization \citep[see][]{Meibom_2007,Gruner_2023}.
 \added{\citet{2019ApJ...871..174S} for instance found that 59\% of
@@
 of detectable ellipsoidal variations in \citet{2016AJ....151...68K}; the outer limit is set by the typical 
-orbital periods of astrometric binaries in Gaia DR3 \citep[see Figure~3 of][]{GaiaCollaboration2023}.}
-
-\added{ If we were not at all sensitive to
+orbital periods of astrometric binaries in Gaia DR3 }\citep[see Figure~3 of][]{GaiaCollaboration2023}.
+
+\added{If we were not at all sensitive to
 binaries with separations of 0.05-0.5\,AU, then based on
@@
 \ngyroappllttendays\ stars with $P_{\rm
-rot}$$<$10\,days.
-
-Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
+rot}$$<$10\,days. }
+
+\added{Figure~\ref{fig:hist_tgyro_corr} shows the rotation-based age distribution
 calculated by randomly dropping \fivepctnuniqstarsantosrotgyroappl\ of the
@@
 $\approx$4\,Gyr for Sun-like stars
-\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the observation of a ``long-period edge''
+\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the ``long-period edge''
 	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
 	may pile up over similar timescales \citep{David_2022}.  However, 
-	this pileup is thought to happen at $>$5\,Gyr for
-	late-G dwarfs and early K dwarfs \citep[see Figure~4
+	the stalling is thought to happen at $>$5\,Gyr for
+	late-G dwarfs and early K dwarfs \citep[e.g.~Figure~4
 	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
-	2.3\,Gyr peak that we see in the 4400-5400\,K age distribution in Figure~\ref{fig:hist_tgyro}. 
+	2.3\,Gyr peak in the 4400-5400\,K age distribution shown in Figure~\ref{fig:hist_tgyro}. 
 	Independently,
-	the analysis by \citet{2022ApJ...937...94M} fitted
+	the study by \citet{2022ApJ...937...94M} fitted
 	isochrones to a high-quality subset of Kepler data, and found that the
@@
 and $\approx$$\pm$60$^\circ$ \citep{Snodgrass1990}.
-This latitudinal surface differential rotation yields two separate limitations on our rotational ages.
+This latitudinal surface differential rotation yields two separate limitations on rotational ages.
 The first is
@@
 \deleted{Other age indicators may help in verifying the ages of the
-$\lesssim$3\,Gyr stars that are the \added{main }focus of this work.  Specifically,
+$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
 chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
@@
 provide age information that is truly independent from rotation.}
-
-%Incorporating
-%kinematics \citep{2021AJ....161..189L} seems likely to remain most
-%useful for stellar ensembles, due to the stochasticity of kinematic
-%heating.
 
@@
 minimal added information for most of the sample\added{ due to its astrophysical
-rate of depletion}, $t_{\rm Li}$ and
+depletion rate}, $t_{\rm Li}$ and
 $t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
@@
   \item Rather than being uniform, the age distributions of both the
-    Kepler target stars and the known Kepler planet-hosts show a
+    Kepler target stars and the known Kepler planet hosts show a
     demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
@@
     formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
-    than it was three billion years ago\added{.
-The statistical uncertainty on this figure is far smaller than the systematic uncertainty caused by the potential presence of unresolved photometric binaries (Section~\ref{subsec:caveats}); the star formation rate could have decreased by a factor of as much as 4.81$\pm$0.26}.
-\replaced{This}{Either} result \replaced{from
+    than it was three billion years ago\added{.  The statistical
+    uncertainty on this value is dwarfed by the systematic
+    uncertainty caused by the potential presence of unresolved
+    photometric binaries (Section~\ref{subsec:caveats}); the star
+    formation rate could have decreased by a factor of as much as
+    4.81$\pm$0.26 over this time interval}.  \replaced{This}{Either} result \replaced{from
     rotation-based ages broadly agrees}{is generally consistent} with recent reports of a
@@
   \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, the age distribution of stars in the Kepler field
+    billion years, this age distribution of stars in the Kepler field
     \replaced{implies}{suggests} that we should instead expect $\approx$250.
@@
 \acknowledgements
-\added{We thank the reviewer for their suggestions, which significantly improved the 
-quality of the manuscript.}
+We thank the reviewer for their suggestions, which significantly improved the 
+quality of the manuscript.
 This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
@@
   ``Confirmed'' planets appear in the machine-readable
-  version before``candidate'' planets.  
+  version before ``candidate'' planets.  
   Planetary sizes are mostly drawn from (in order of precedence)

### 2024-08-27 papers/paper/collect.sh

@@ -32,7 +32,7 @@ cp $r/li_vs_teff/li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf .
 
 # planets
 r="/Users/luke/Dropbox/proj/NEA_age_plots/results/rp_vs_period_scatter"
-cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung.pdf .
+cp $r/rp_vs_period_scatter_20240415_colorbyage_showaux-gyro_anyyoung_anyyoung_allowsgrazingandhighruwe.pdf .
 
 
 ########## 

### 2024-08-28 papers/paper/ms.tex
update ms

@@
   sub-gigayear systems can be attributed to the star formation rate in
-  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{a few}
+  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{3-5}
   over the past 3\,Gyr.  We observe this trend both for known planet

### 2024-08-29 papers/paper/ms.tex
update ms

@@
 % upon AAS submission
-\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
+%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
 %\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
-%\documentclass[11pt,twocolumn,tighten]{aastex63}
+\documentclass[11pt,twocolumn,tighten]{aastex63}
 \turnoffedit
@@
   This is about half the number expected under the classic assumption
-  of a uniform star formation history.  We find that the scarcity of
-  sub-gigayear systems can be attributed to the star formation rate in
-  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{3-5}
-  over the past 3\,Gyr.  We observe this trend both for known planet
-  hosts and for the parent stellar sample.
+  of a uniform star formation history.
   %
-  This ``demographic cliff'' in the Galaxy's star formation history
-  has been previously reported, and its confirmation helps clarify the
-  age distribution of the known transiting exoplanets.
+  \deleted{We find that the scarcity of sub-gigayear systems can be
+  attributed to the star formation rate in the Kepler field dropping
+  by a factor of \ratiosfr$\pm$\uncratiosfr over the past 3\,Gyr.  We
+  observe this trend both for known planet hosts and for the parent
+  stellar sample.  This ``demographic cliff'' in the Galaxy's star
+  formation history has been previously reported, and its confirmation
+  helps clarify the age distribution of the known transiting
+  exoplanets.}
   %
+  \added{The age distribution that we observe, rather than being uniform,
+  shows that the youngest stars in the Kepler field are 3-5 times
+  rarer than stars 3\,Gyr old.
+  This trend holds for both known planet hosts and for the parent
+  stellar sample.
+  We attribute the existence of this ``demographic cliff'' to a
+  combination of kinematic heating and a declining star formation rate
+  in the Galaxy's thin disk, and highlight its impact on the age
+  distribution of known transiting exoplanets.
+  }
 \end{abstract}
@@
 timeline for exoplanet evolution.  Informative individual exemplars
-include HIP~67522b, a Jupiter-sized planet with a sub-Neptunian mass
+include HIP~67522\,b, a Jupiter-sized planet with a sub-Neptunian mass
 (\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
@@
   %
-  \item We can identify binary stars \added{and high-metallicity outliers }with greater fidelity than
-    in the past, which can clarify otherwise problematic estimates of
+  \item We can identify \replaced{binary stars}{stars that are
+    binaries and that have high metallicity} with greater fidelity\deleted{ than
+    in the past}, which can clarify otherwise problematic estimates of
     rotation-based ages.
@@
 formation history at $\lesssim$3\,Gyr, at older ages incompleteness
-likely precludes any such interpretation.}
+precludes any such interpretation.}
 
@@
 
-Sub-gigayear planets can be particularly informative for studies of
-planet evolution.  Our catalog has \nplyounggyro\ confirmed and
-candidate planets with median ages below 1\,Gyr.  Requiring $t_{\rm
-gyro}$$<$1\,Gyr at 2$\sigma$ yields \nplyounggyrotwosigma\ planets
-orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim is most
-secure for the systems that are either in clusters (Section~\ref{subsec:clusterplanets}), or that have
-independent rotation and lithium-based ages (Section~\ref{subsec:rotnlithiumcomp}).
-These samples provide broader context for the planets around field
-stars that push the boundaries of the current young exoplanet census
-(Section~\ref{subsec:notables}).
+\deleted{Sub-gigayear planets
+can be particularly informative for studies of planet
+evolution.}
+\added{This analysis provides an important path toward expanding the population
+of young and evolving exoplanets.}
+Our catalog has \nplyounggyro\ confirmed
+and candidate planets with median ages below 1\,Gyr.  Requiring
+$t_{\rm gyro}$$<$1\,Gyr at 2$\sigma$ yields \nplyounggyrotwosigma\
+planets orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim
+is most secure for the systems that are either in clusters
+(Section~\ref{subsec:clusterplanets}), or that have independent
+rotation and lithium-based ages
+(Section~\ref{subsec:rotnlithiumcomp}).  These samples provide broader
+context for the planets around field stars that push the boundaries of
+the current young exoplanet census (Section~\ref{subsec:notables}).
 
@@
 Among these systems, four highlights are Kepler-1529, Kepler-1565,
-Kepler-1312, and Kepler-1629.  Kepler-1529b is a
+Kepler-1312, and Kepler-1629.  Kepler-1529\,b is a
 $\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation and
-lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
+lithium ages.  Kepler-1565\,b is an analogous $\approx$170-230\,Myr
 super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
@@
 mini-Neptune on a five-day orbit.  The Earth-sized planet,
-Kepler-1312~c, along with the near-identical Kepler-1561~b ($t_{\rm
+Kepler-1312\,c, along with the near-identical Kepler-1561\,b ($t_{\rm
 gyro}$=\kepfifteensixone\,Myr), rank among the youngest Earth-sized
-planets currently known, comparable to planets such as HD~63433d
+planets currently known, comparable to planets such as HD~63433\,d
 \citep[1.1\,$R_\oplus$, $414$$\pm$$23$\,Myr;][]{2024AJ....167...54C}
-and TOI-1807~b \citep[1.3\,$R_\oplus$,
-180$\pm$40\,Myr;][]{2021AJ....162...54H}.  Kepler-1629~b, with a size
+and TOI-1807\,b \citep[1.3\,$R_\oplus$,
+180$\pm$40\,Myr;][]{2021AJ....162...54H}.  Kepler-1629\,b, with a size
 just two-thirds that of Earth (0.67$\pm$0.05\,$R_\oplus$), is
@@
 
-\subsection{The Thin Disk's Star Formation History}
-
-Most \deleted{of the }stars observed by Kepler are in the thin disk.  This can
-be verified by following \citet{Gaia_2018}, and labeling stars with 2D
-tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
-and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
-\replaced{For all stars in the KIC, this yields }{These criteria yield }95{,}694 thin and 54{,}039 thick
-disk stars \replaced{respectively}{observed by Kepler}.  The thin disk stars have a threefold larger
-detected rotation period fraction: 35{,}675 rotators are in the thin
-disk, while 7{,}312 are in the thick disk.  Further imposing the
-\deleted{gyrochronology }quality flags discussed in Section~\ref{subsec:flags}
-yields 17{,}755 thin disk stars for which gyrochronology is
-applicable, and 2{,}462 thick disk stars.
-
-Classifying Kepler stars as thin vs.\ thick disk members helps connect
+\begin{figure*}[!t]
+	\begin{center}
+		\leavevmode
+		\subfloat{
+			\includegraphics[width=.99\textwidth]{threepanel_trilegal_comparison_twostepsfr.pdf}
+		}
+	\end{center}
+	\vspace{-0.5cm}
+	\caption{{\bf Impact of star formation rate (SFR) and kinematic heating on age distribution.}   
+		Most Kepler stars are a few hundred parsecs above the Galactic plane ($\hat{Z}$ points to the north galactic pole;  $\hat{Y}$ is the direction of
+		galactic rotation).  This is comparable to the $\approx$300\,pc mass-weighted scale height.
+		The dotted gray and solid orange lines show synthetic age distributions
+		from TRILEGAL assuming a constant SFR (middle) and a two-step
+		SFR (right), described in the text.
+		A combination of both kinematic heating and
+		a decreasing SFR qualitatively match the observed paucity of the youngest stars.
+		\label{fig:trilegal}
+	}
+\end{figure*}
+
+%\subsection{The Thin Disk's Star Formation History}
+\subsection{Kinematic Heating and Star Formation in the Thin Disk}
+\label{subsec:trilegal}
+
+\added{Two separate astrophysical effects might explain our observed 
+age distribution (Figure~\ref{fig:hist_tgyro}): a declining star formation
+rate (SFR), and kinematic heating.   Kinematic heating is associated with an
+increase in stellar velocity dispersions and the disk's vertical 
+scale height over time \citep[e.g.][]{Villumsen1983,Schmidt2024}.
+The Kepler field points toward $(l,b)$=(76.3$^\circ$,13.5$^\circ$).
+A typical Kepler star at 1\,kpc is therefore $\approx$230\,pc
+above the galactic plane, a significant fraction of the thin disk's 
+mass-weighted $\approx$300\,pc scale height \citep{Bland-Hawthorn2016}.
+}
+
+\added{We assessed the relative importance of these two effects
+	using the TRILEGAL Monte Carlo population synthesis tool }\citep[v1.6;][]{Girardi2005}.
+\added{TRILEGAL treats the Galaxy as a linear combination of a thin disk, thick disk,
+	halo, and bulge, and allows for two possible SFRs: constant, and a ``two-step''
+	SFR in which the thin disk SFR is 50\% larger from 1-4\,Gyr than at other ages.
+	This two-step SFR was introduced to improve the model's ability to
+	match star counts from Hipparcos and 2MASS }\citep[see Section~5.1 of][]{Girardi2005}.
+\added{For kinematic heating, TRILEGAL assumes that
+	the disk scale height grows as
+$h_{\rm z} = z_0 (1 + t/t_0)^\alpha$, for $\alpha$=5/3, $z_0$=94.7\,pc,
+	and $t_0$=5.55\,Gyr, similar to \citet{Rana1991}.
+	We queried the web interface toward the
+	Kepler field, and toward the galactic plane (at $l$=76.3$^\circ$) to
+	negate the effects of kinematic heating.
+	We limited the output to stars with apparent Kepler magnitudes brighter than 16,
+	and set the field of view to 10 ${\rm deg}^2$, the maximum
+	possible through the web interface.
+%	and set the Sun at $z_\odot$=20.8\,pc, $r$=8122\,pc from the Galactic center.
+}
+
+\added{Figure~\ref{fig:trilegal} compares the simulation results against
+	our observed age distribution.
+	The middle panel shows that assuming a constant SFR,
+	kinematic heating alone can yield a two-fold difference
+	in the abundance of the youngest stars relative to those $\approx$3\,Gyr old.
+	The right-most panel
+	shows that additionally including 
+	the time-variable SFR yields the best agreement with the slope of
+	our observed age distribution.
+%	An important note is that the previously noted parameters
+%	that were fine-tuned to fit different datasets, primarily photometric star counts
+%	along various sight-lines.  The exercise appears to have been predictive.
+}
+
+\deleted{Most stars observed by Kepler are in the thin disk.
+This can be verified by following \citet{Gaia_2018}, and labeling
+stars with 2D tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin
+disk members, and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick
+disk members.  These
+criteria yield 95{,}694 thin and 54{,}039 thick disk stars
+observed by Kepler.  The thin disk stars have
+a threefold larger detected rotation period fraction: 35{,}675
+rotators are in the thin disk, while 7{,}312 are in the thick disk.
+Further imposing the quality flags discussed
+in Section~\ref{subsec:flags} yields 17{,}755 thin disk stars for
+which gyrochronology is applicable, and 2{,}462 thick disk stars.}
+
+\deleted{Classifying Kepler stars as thin vs.\ thick disk members helps connect
 them to previous work on the Galaxy's star formation history.  On
@@
 since decreased by an order of magnitude
-\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
-galaxy, previous studies have measured star formation histories
+\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
+\added{Despite the expected differences between pencil-beam surveys
+and volume-complete samples, our observed age distribution qualitatively
+agrees with a number of previous studies.}
+\replaced{In our own
+galaxy, previous studies have}{Previous work has} measured star formation histories
 through CMD fitting of resolved stellar populations
-\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
+\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X},
 and by modeling the local white dwarf luminosity function
@@
 focused on stars in a wider 2\,kpc bubble, and additionally reported
-three local maxima in the galactic star formation rate, which they
+three local maxima in the SFR, which they
 associated with pericenter passages of the Sagittarius satellite
@@
 \citet{SilvaAguirre2018}.  The main novelty of our ages relative to
-\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.
+\replaced{those studies is their improved accuracy}{these previous studies is their accuracy} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.
 
@@
 influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
-metallicity, differential rotation, mergers, and tidal spin-up.  Observational
+metallicity, latitudinal differential rotation, mergers, and tidal spin-up.  Observational
 limitations include the difficulty of measuring small-amplitude,
@@
 evolutionary connection to their youth.  Other discoveries include
-Earth-sized planets \added{(e.g.~Kepler-1312~c and Kepler-1561~b) }with \deleted{new }ages of only a few hundred million years.
+Earth-sized planets \added{(e.g.~Kepler-1312\,c and Kepler-1561\,b) }with \deleted{new }ages of only a few hundred million years.
 
@@
 
-\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
+\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
   %
@@
     demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
-    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
+    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
     formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
-    than it was three billion years ago\added{.  The statistical
+    than it was three billion years ago.}
+\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
+    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 3\,Gyr old. The statistical
     uncertainty on this value is dwarfed by the systematic
-    uncertainty caused by the potential presence of unresolved
-    photometric binaries (Section~\ref{subsec:caveats}); the star
-    formation rate could have decreased by a factor of as much as
-    4.81$\pm$0.26 over this time interval}.  \replaced{This}{Either} result \replaced{from
-    rotation-based ages broadly agrees}{is generally consistent} with recent reports of a
-    declining star-formation rate from CMD fitting and white-dwarf
-    chronology, though with the \added{potential }advantage of better \replaced{sensitivity}{accuracy} for
+    uncertainty in the contamination rate of unresolved
+    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
+     \replaced{This}{Either} result \replaced{from
+    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
+    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} from CMD fitting and white-dwarf
+    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
     FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
@@
   \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, this age distribution of stars in the Kepler field
+    billion years, the age distribution of stars in the Kepler field
     \replaced{implies}{suggests} that we should instead expect $\approx$250.
@@
 \acknowledgements
-We thank the reviewer for their suggestions, which significantly improved the 
-quality of the manuscript.
+We thank Kevin Schlaufman and the reviewer for suggestions which significantly improved the 
+manuscript.
 This work was supported by the Heising-Simons 51~Pegasi~b Fellowship

### 2024-08-30 papers/paper/ms.tex
update ms

@@
 %\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
-%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
+\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
 % drafting / arxiv
-\documentclass[11pt,twocolumn,tighten]{aastex63}
+%\documentclass[11pt,twocolumn,tighten]{aastex63}
 \turnoffedit
@@
   %
-  We find that our rotational ages accurately recover the ages of
+  We find that our rotational ages \deleted{accurately }recover the ages of
   stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
@@
     These clusters offer stars that we can use to test
-    the reliability of the available single-star age-dating methods.
+    the reliability of \replaced{the available single-star}{our} age-dating methods.
 \end{enumerate}
@@
 stellar parameters other than ages in Section~\ref{sec:stellarprops}.
-We describe our age-dating methods in Sections~\ref{sec:rotage}
+We present our age-dating methods in Sections~\ref{sec:rotage}
 and~\ref{sec:liage}, and test them in
@@
   stars with temperatures of 4400--5400\,K, which \replaced{have more precise
-  ages due to their fast spin-down}{usually have more precise ages}.  The statistical uncertainty for a
+  ages due to their fast spin-down}{have more precise ages}.  The statistical uncertainty for a
   mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
@@
 \citep{Bouma_2022a,Bouma_2022b}.  Different sub-groups of the complex
-vary in age by $\approx$50\% (Kerr~et al.,~submitted).  The planets
+vary in age by $\approx$50\% \citep{Kerr2024}.  The planets
 themselves are all on close-in orbits (5-25\,days), with sizes from
@@
     and for the planets to similarly no have quality flags
-    raised.}{Transiting systems are selected to have $t$<1\,Gyr at $2$$\sigma$.
+    raised.}{Systems are selected to have $t$<1\,Gyr at $2$$\sigma$.
     Star symbols denote \nplyounggyrotwosigmanograzingnoruwe\ planets
@@
 sized planets below the radius valley.  While most of these planets
-are on orbital periods below 50\,days, \nlongperiodhighq\ are on more
+have orbital periods below 50\,days, \nlongperiodhighq\ are on more
 distant orbits.
@@
 	\caption{{\bf Impact of star formation rate (SFR) and kinematic heating on age distribution.}   
-		Most Kepler stars are a few hundred parsecs above the Galactic plane ($\hat{Z}$ points to the north galactic pole;  $\hat{Y}$ is the direction of
+		Most Kepler stars are a few hundred parsecs above the Galactic plane ($\hat{Z}$ points to the north galactic pole;  $\hat{Y}$ toward the direction of
 		galactic rotation).  This is comparable to the $\approx$300\,pc mass-weighted scale height.
@@
 		A combination of both kinematic heating and
-		a decreasing SFR qualitatively match the observed paucity of the youngest stars.
+		a decreasing SFR are needed to qualitatively match the observed paucity of the youngest stars.
 		\label{fig:trilegal}
@@
 rate (SFR), and kinematic heating.   Kinematic heating is associated with an
-increase in stellar velocity dispersions and the disk's vertical 
+increase in both stellar velocity dispersions and the disk's vertical 
 scale height over time \citep[e.g.][]{Villumsen1983,Schmidt2024}.
@@
 	We limited the output to stars with apparent Kepler magnitudes brighter than 16,
-	and set the field of view to 10 ${\rm deg}^2$, the maximum
-	possible through the web interface.
+	and set the field of view to 10\,${\rm deg}^2$, the maximum
+	available through the web interface.
 %	and set the Sun at $z_\odot$=20.8\,pc, $r$=8122\,pc from the Galactic center.
@@
 
-\added{Figure~\ref{fig:trilegal} compares the simulation results against
+\added{Figure~\ref{fig:trilegal} compares the results against
 	our observed age distribution.
@@
 mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
-since decreased by an order of magnitude
-\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
+since decreased by an order of magnitude. }
+%\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
 \added{Despite the expected differences between pencil-beam surveys
@@
 		However, even after applying our quality cuts, up to
-		\fivepctnuniqstarsantosrotgyroappl\ out of \ngyroappllttendays\ ``clean'' rotators with
+		\fivepctnuniqstarsantosrotgyroappl\ of \ngyroappllttendays\ ``clean'' rotators with
 		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
@@
 Gaia with rotation measurements from TESS. Given the $\sim$100\,Myr
-decoherence time of young clusters, the youngest ``field stars'' may
+decoherence time of young clusters, \replaced{the youngest ``field stars''}{these young stars} may
 have easily identifiable young neighbors.
@@
 depletion rate}, $t_{\rm Li}$ and
-$t_{\rm gyro}$ agreed in $\gtrsim$90\% of cases for which comparison
+$t_{\rm gyro}$ agreed in over 90\% of cases for which comparison
 was possible.  Our \replaced{resulting ages}{results} included two-sided $t_{\rm gyro}$
@@
 confirmed planets, and two-sided $t_{\rm gyro}$ with one-sided $t_{\rm
-Li}$ for \ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed
+Li}$ for \ltonegyrhighqconfirmedonesided\ \replaced{sub-gigayear confirmed}{such}
 planets.  Allowing for ``candidate'' planets, grazing transit
@@
     stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
-    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
+    formation rate today is \ratiosfr$\pm$\uncratiosfr\ times lower
     than it was three billion years ago.}
@@
     photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
-     \replaced{This}{Either} result \replaced{from
+     \replaced{This result}{This range of values} \replaced{from
     rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
@@
   %
-  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
+  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
     Kepler planets younger than 1\,Gyr at 2$\sigma$, and

### 2024-09-01 papers/paper/ms.tex
update ms

@@
 K2, TESS, and Gaia
-\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
+\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,Bouma_2020_toi837,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Zhou_2022,Zakhozhay_2022,Wood_2023}.
 The strategy pursued by most groups during the 2010s was to focus on
@@
 using a star's spin-down as a clock has a rich history
-\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015}.
+\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015,2023ApJ...954L..50E}.
 Empirical models now yield ages precise to $\lesssim$30\% for FGK
@@
 		\subfloat{
-			\includegraphics[width=0.43\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.45\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.43\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.45\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
+			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.png}
+			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
     candidate planets described in Section~\ref{subsec:plflags}.  The
-    number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
+    total number of stars $N_{\rm s}$ and planets $N_{\rm p}$ are noted in
     the relevant panels.  A simple estimate for the number of young
@@
 \citet{Santos_2019,Santos_2021}, hereafter \citetalias{Santos_2019}
-and \citetalias{Santos_2021}.  For a discussion of the work by
-\citet{Reinhold2023}, please see Appendix~\ref{app:reinhold23}.
+and \citetalias{Santos_2021}.  For discussion of the work by
+\citet{Reinhold2023}, see Appendix~\ref{app:reinhold23}.
 \citetalias{Santos_2019} and \citetalias{Santos_2021} combined a
@@
     Additional sources of uncertainty, including measurement
-    uncertainty, impose further limits on achievable precision.  These
+    uncertainty, impose further limits\deleted{ on achievable precision}.  These
     models are calibrated using data from open clusters.  The sizes of
@@
   ages due to their fast spin-down}{have more precise ages}.  The statistical uncertainty for a
-  mean star at 1, 2, and 3\,Gyr is shown, and is identical across each
+  single average star at 1, 2, and 3\,Gyr is shown, and is identical across each
   row.  
@@
 	\caption{{\bf Potential impact of unresolved binaries.}   This figure shows
-		the same raw (thin blue line) and quality-flag cleaned (``gyro applicable'') distributions
+		the same raw (thinnest line) and quality-flag cleaned (``gyro applicable'') distributions
 		as in Figure~\ref{fig:hist_tgyro}.
@@
 		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
-		The thick blue ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
+		The thick violet ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
 		what we estimate (estm.) to be the maximum possible number of such systems; the true
@@
 \item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
-    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 3\,Gyr old. The statistical
+    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
     uncertainty on this value is dwarfed by the systematic
@@
   Kepler \citep{Borucki10},
-  Keck:I (HIRES) \citep{Vogt94},
+  Keck:I (HIRES) \citep{vogt_hires_1994},
 	2MASS \citep{Skrutskie06},
@@
 \software{
-  astropy \citep{Astropy18},
+  astropy \citep{astropy:2013,astropy:2018,astropy:2022},
   claude \citep{claude2024anthropic},
@@
 		\subfloat{
-			\includegraphics[width=0.44\textwidth]{prot_teff_Santos19_Santos21_dquality.png}
+			\includegraphics[width=0.44\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.44\textwidth]{prot_teff_McQuillan2014only_dquality.png}
+			\includegraphics[width=0.44\textwidth]{prot_teff_McQuillan2014only_dquality.pdf}
 		}

### 2024-09-01 drivers/make_younghighlight_table.py
table1: comment update

@@ -287,6 +287,10 @@ def make_table(
         if k not in COMMENTDICT:
             COMMENTDICT[k] = '\checkmark \checkmark'
 
+    closeperiods = np.abs(pdf.Prot - pdf.adopted_period) < 0.1
+    for k in pdf[closeperiods].kepoi_name:
+        COMMENTDICT[k] = '$|P_\mathrm{rot}$-$P_\mathrm{orb}|$$<$0.1$\,\mathrm{days}$'
+
     _sel1 = (
         (pdf.min_age < 1000) &
         sel_gyro_quality &

### 2024-09-01 papers/paper/ms.tex
update ms

@@
 and by modeling the local white dwarf luminosity function
-\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
+\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} \replaced{and
 \citet{2019ApJ...878L..11I} for instance focused on stars (and white
-dwarfs) within a few hundred parsecs, and both found a peak in the
+dwarfs)}{and \citet{Mazzi2024} for instance focused on stars} within a few hundred parsecs, and both found a peak in the
 local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
@@
 than the $\approx$500 that would be expected under the assumption of a
-uniform star formation history for the $\approx$5{,}000 known planets?
+uniform star formation \replaced{history}{rate} for the $\approx$5{,}000 known planets?
 
@@
 
-Our main conclusions with respect to our original questions are as
+\added{Significant care is 
+needed when assessing not only the reliability of rotational ages, but also any implications of measured age distributions.
+Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
+and binarity, we expect that up to a third of the ``apparently clean'' 
+Kepler stars with $P_{\rm rot}$$<$10\,days
+could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
+A separate nuance is in understanding that the observed age distribution does not 
+directly map onto a star formation history: the kinematic heating of stars in the
+disk in particular produces a major bias in the age distribution of stars, and therefore of planets,
+observed by pencil-beam surveys such as Kepler.}
+
+Our \deleted{main }conclusions with respect to our original \added{two }questions\added{ -- what is the age distribution in the thin disk, and where are the missing young transiting exoplanets -- }are as
 follows.
@@
     rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
-    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} from CMD fitting and white-dwarf
+    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} CMD fitting and white-dwarf
     chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
@@
     billion years, the age distribution of stars in the Kepler field
-    \replaced{implies}{suggests} that we should instead expect $\approx$250.
+    \replaced{implies}{suggests} that we \replaced{should}{might} instead expect $\approx$250.
   %
@@
   following meaning.
-  Bit 0: Candidate not reliable?
+  Bit 0: Candidate disposition not reliable?
   Bit 1: MES$<$10?

### 2024-09-01 papers/paper/ms.tex
update ms

@@
 %%%%%%%%%%%%%%%%
-\newcommand{\caltech}{Department of Astronomy, MC 249-17, California Institute of Technology, Pasadena, CA 91125, USA}
+\newcommand{\caltech}{Department of Astronomy, California Institute of Technology, Pasadena, CA 91125, USA}
+\newcommand{\carnegie}{Observatories of the Carnegie Institution for Science, Pasadena, CA 91101, USA}
 \newcommand{\mitkavli}{MIT Kavli Institute and Department of Physics,
@@
 \received{2024 June 21}
-\revised{---}
+\revised{2024 September 5}
 \accepted{---}
@@
 \affiliation{\caltech}
+\affiliation{\carnegie}
 
@@
 		\subfloat{
-			\includegraphics[width=0.45\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
+			\includegraphics[width=0.42\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
 		}
@@
 		\subfloat{
-			\includegraphics[width=0.45\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
+			\includegraphics[width=0.42\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
 		}
@@
   based on factors including their non-binarity and proximity to the
-  main sequence (orange line in lower panel; see
+  main sequence (orange line in \replaced{lower}{upper} panel; see
   Section~\ref{subsec:flags}).  Some host ``confirmed'' or
@@
   and effective temperatures were derived photometrically by
-  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
+  \citet{Berger_2020a_catalog}.  Isochrones in the \replaced{lower}{upper} panel are
   from MIST \citep{Choi_2016}.
@@
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
+			\includegraphics[width=0.46\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
 		}
 		\subfloat{
-			\includegraphics[width=0.49\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
+			\includegraphics[width=0.46\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
 		}
@@
     planet hosts (right)}.  The gray lines are \replaced{``mean fits'' to the
-    rotation sequences of open clusters}{seventh-order polynomials
-    fitted to the slow rotation sequences of open clusters \citep[see Section~3.3 of][]{Bouma_2023}}.  The 2-D histogram of the
-    stellar sample (left) includes only apparently single stars near
+    rotation sequences of open clusters}{polynomials
+    fitted to the slow rotation sequences of open clusters}.  The 2-D histogram of the
+    stellar sample (left) shows only apparently single stars near
     the main sequence with $\log g$$>$4.2, RUWE$<$1.4, and
@@
 \citetalias{Santos_2021}, and our extended KOI list.\added{  The gray
-lines in this figure are the polynomials described by \citet{Bouma_2023}
+lines in this figure are the seventh-order polynomials described by \citet{Bouma_2023}
 that 
@@
 		\leavevmode
-		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
+		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
 	\end{center}
-	\vspace{-0.25cm}
+	\vspace{-0.5cm}
 	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
@@
 
-We began this work with two questions: how wrong is the assumption of
+We began \deleted{this work }with two questions: how wrong is the assumption of
 a uniform age distribution for stars in the galactic thin disk?  And
@@
 \added{Significant care is 
-needed when assessing not only the reliability of rotational ages, but also any implications of measured age distributions.
+needed not only when assessing the reliability of individual rotational ages, but also when interpreting distributions of rotational ages inferred from many stars.
 Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
-and binarity, we expect that up to a third of the ``apparently clean'' 
+and binarity, we expect that up to a third of our ``apparently clean'' 
 Kepler stars with $P_{\rm rot}$$<$10\,days
 could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
-A separate nuance is in understanding that the observed age distribution does not 
-directly map onto a star formation history: the kinematic heating of stars in the
-disk in particular produces a major bias in the age distribution of stars, and therefore of planets,
-observed by pencil-beam surveys such as Kepler.}
+An additional nuance is that even after correcting for effects
+that can bias rotational ages, our observed age distribution does not 
+directly yield a star formation history: 
+particularly for pencil-beam surveys such as Kepler,
+kinematic heating of stars in the
+thin disk significantly sculpts the stellar (and therefore planetary) age distribution.}
 

### 2024-09-02 papers/paper/ms.tex
update ms

@@
 \received{2024 June 21}
-\revised{2024 September 5}
+\revised{2024 September 4}
 \accepted{---}
@@
 \author[0000-0002-0514-5538]{Luke~G.~Bouma}
-\altaffiliation{51 Pegasi b Fellow}
+\altaffiliation{51 Pegasi b Fellow;  Carnegie Fellow}
 \affiliation{\caltech}
@@
 		However, even after applying our quality cuts, up to
-		\fivepctnuniqstarsantosrotgyroappl\ of \ngyroappllttendays\ ``clean'' rotators with
+		\fivepctnuniqstarsantosrotgyroappl\ of \ngyroappllttendays\ apparently clean rotators with
 		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
-		The thick violet ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
+		The black ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
 		what we estimate (estm.) to be the maximum possible number of such systems; the true
@@
 Kepler field, but also in the reliability of individual ages for \replaced{young
-stars known to host planets}{each star in our sample}.  We attempted to assess this reliability using
+stars known to host planets}{each star}.  We attempted to assess this reliability using
 quality bitmasks (Table~\ref{tab:planets}), a comparison against \deleted{open }cluster ages (Section~\ref{sec:litagecomparison}), and a comparison
 between rotation and lithium ages.  The \deleted{open }cluster comparison
-suggested that our ages were generally accurate at $\ll$0.5\,dex
+suggested that our ages were mostly accurate at $\ll$0.5\,dex
 across 0.04--2.5\,Gyr, similar to their quoted precision.
@@
 does show outliers, typically $\lesssim$10\% of the population
-in any cluster.  \added{While most of these outliers are known to have bad quality flags, some do not.} 
+in any cluster.  \added{While most of these outliers have known bad quality flags, some do not.} 
 These stars could be either field star contaminants,
@@
 
-\added{Other age indicators may help in verifying the ages of the
-$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
-chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
-can serve as an age tracer
-\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
-However, these age indicators are all in a sense ``rotation-powered''.
-The dynamo converts kinetic energy into magnetic energy, which is
-emitted through these chromospheric pathways.  We did not attempt to
-incorporate these indicators into this study due to concerns regarding
-sensitivity, homogeneity, and the question of whether they in fact
-provide age information that is truly independent from rotation.}
+%\added{Other age indicators may help in verifying the ages of the
+%$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
+%chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
+%can serve as an age tracer
+%\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
+%However, these age indicators are all in a sense ``rotation-powered''.
+%The dynamo converts kinetic energy into magnetic energy, which is
+%emitted through these chromospheric pathways.  We did not attempt to
+%incorporate these indicators into this study due to concerns regarding
+%sensitivity, homogeneity, and the question of whether they in fact
+%provide age information that is truly independent from rotation.}
 
@@
 \added{{\it Metallicity:}  
-Stars with higher metallicity are thought to have longer convective turnover
+Stars with higher metallicity are predicted to have longer convective turnover
 times, which may slow their rate of spin-down
 	\citep{Amard2020}.
-While observational studies of this effect have been suggestive, further work
-remains to definitively establish the importance of the effect \citep[e.g.][]{Amard2020b,2024arXiv240500779S}
-Nonetheless, based on the theoretical predictions, we flagged stars with known
-	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, based on either LAMOST or Keck/HIRES
+While recent observational work has empirically confirmed that such an effect is
+present in the Kepler data
+\citep[][]{2024arXiv240500779S},
+the anti-correlation between close,
+unresolved binaries and stellar metallicity \citep{Moe2019} could be a complicating factor.
+Nonetheless, based on these studies, we flagged stars with known
+	spectroscopic $|[{\rm Fe/H}]|$$>$0.3, using either LAMOST or Keck/HIRES
 	spectra \citep{Zong2018,Petigura_2022}.
@@
 $t_{\rm gyro}$$<$4\,Gyr had such spectra available.
-For the rest of the sample, primarily composed of fainter stars,
-this means that our quality flags were not sensitive to metallicity outliers.
-Based on the metallicity distribution of stars for which we do have spectra,
-this 
-implies that at most $\approx$5\% of stars in our sample might have $|[{\rm Fe/H}]|$$>$0.3,
+For the remainder of our sample,
+our quality flags were not sensitive to metallicity outliers.
+The metallicity distribution of stars for which we do have spectra suggests that
+at most $\approx$5\% of stars in our apparently clean sample might have $|[{\rm Fe/H}]|$$>$0.3 and not
+be flagged,
 with a slight bias toward lower metallicity stars over higher metallicity.
-Based on the results by \citet{Claytor2020},  this will be most important
+Based on the study by \citet{Claytor2020},  this will be most important
 for stars near the Kraft break, which could have ages over-estimated
@@
 The Sun's surface rotates once per $\approx$25-30\,days at latitudes between the equator
-and $\approx$$\pm$60$^\circ$ \citep{Snodgrass1990}.
+and $\approx$$\pm$60$^\circ$ \citep[e.g.][]{Snodgrass1990}.
 This latitudinal surface differential rotation yields two separate limitations on rotational ages.
@@
 This means that irrespective of the assumed $P_{\rm rot}$ measurement precision,
-our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter (perhaps differential-rotation induced)
+our model imposed a $\approx$5-10\% level of intrinsic $P_{\rm rot}$ scatter
 at old ages.
@@
 or slower than average stars of the same mass and age.  
-This effect can yield a $\approx$20\% age bias for individual Sun-like stars with ages
+This effect can yield a $\approx$20\% age error for individual Sun-like stars with ages
 of a few billion years; the statistical uncertainties in Table~\ref{tab:stars} should be
-treated with appropriate caution.
-}
+treated with appropriate caution.}
 
@@
 leverage either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, or else that combine both
+\citep{vanSaders_2016,2024ApJ...962..138S}, kinematics \citep{2021AJ....161..189L,2024arXiv240500779S},
+or else that combine both
 the evolution of stellar luminosity and stellar rotation
@@
 
-{\it \added{Planet} occurrence Rates}:
+{\it \added{Planet} occurrence rates}:
 The age-dependent trends predicted for exoplanet populations, such as
@@
 \citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
-Survey \citep{Wilson2023}.  Rotation-based age-dating from these surveys could
+Survey \citep{Wilson2023}.  Rotation-based ages from these surveys could
 yield new insight
-into whether the star formation history that we see in the Kepler field is
-universal across the thin disk.
+into \replaced{whether the star formation history that we see in the Kepler field is
+universal across the thin disk.}{the relative importance of star formation and
+kinematic heating across different components of the Galaxy.}
 
@@
 could potentially be influenced by unresolved binaries (Figure~\ref{fig:hist_tgyro_corr}).
-An additional nuance is that even after correcting for effects
-that can bias rotational ages, our observed age distribution does not 
-directly yield a star formation history: 
+An additional nuance is that even after correcting for such effects,
+the star formation history does not directly trace
+the observed age distribution: 
 particularly for pencil-beam surveys such as Kepler,
 kinematic heating of stars in the
-thin disk significantly sculpts the stellar (and therefore planetary) age distribution.}
+thin disk significantly sculpts the observed stellar (and therefore planetary) age distribution.}
 
@@
     than it was three billion years ago.}
-\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
+\item \added{The youngest Kepler stars (0-0.3\,Gyr)
     are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 2.7-3\,Gyr old. The statistical
@@
     chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
-    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
+    individual FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
   %
-  \item Rather than expecting $\approx$500 exoplanets younger than one
-    billion years, the age distribution of stars in the Kepler field
+  \item Rather than expecting $\approx$500 cumulative exoplanets younger than one
+    billion years, the age distribution of the Kepler field
     \replaced{implies}{suggests} that we \replaced{should}{might} instead expect $\approx$250.

### 2024-09-02 papers/paper/ms.tex
update ms

@@
 of young exoplanets relative to their older counterparts
-\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.
+\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,Vach2024}.
 
@@
 present in the Kepler data
-\citep[][]{2024arXiv240500779S},
+\citep[][]{See2024},
 the anti-correlation between close,
@@
 \deleted{Metallicity may also be relevant
-\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.  An additional caveat
+\citep{2020MNRAS.499.3481A,See2024}.  An additional caveat
 is that our rotation-based ages used photometric effective
@@
 leverage either asteroseismology
-\citep{vanSaders_2016,2024ApJ...962..138S}, kinematics \citep{2021AJ....161..189L,2024arXiv240500779S},
+\citep{vanSaders_2016,2024ApJ...962..138S}, kinematics \citep{2021AJ....161..189L,See2024},
 or else that combine both
@@
 \facilities{
-  Gaia \citep{Gaia_DR3_2022},
+  Gaia \citep{GaiaCollaboration2023},
   Kepler \citep{Borucki10},
@@
   bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
-  ten to be null yields \nuniqstarsantosrotgyroappl\ stars for
+	ten to be null (e.g.~using the suggested {\bf
+\href{https://gist.github.com/lgbouma/87ad8bde42625e766a5a8857cc5a183a}{python script}}) yields \nuniqstarsantosrotgyroappl\ stars for
   which gyrochronology is likely to be valid.  \label{tab:stars}}

### 2024-09-02 drivers/make_star_table.py

@@ -144,7 +144,7 @@ def make_star_table(
     # format ages
     sel = sdf.gyro_median.astype(float) > 4000
     _sdf.loc[sel, 't_gyro'] = '$> 4000$'
-    _sdf.sample(n=5).to_latex(
+    _sdf.sample(n=10).to_latex(
         latexpath, index=False, na_rep='--', escape=False, formatters=formatters
     )
 

### 2024-09-03 papers/paper/ms.tex
update ms

@@
 \added{{\it Weakened magnetic braking:} 
-If we lived in a
-universe in which stellar rotation rates and spot-induced photometric
-amplitudes remained {\it constant} after $\approx$3\,Gyr, then older
-stars could contaminate the peaks in
-Figure~\ref{fig:hist_tgyro}.  This
-scenario deserves consideration, given asteroseismic evidence for slowed spin-down rates after
-$\approx$4\,Gyr for Sun-like stars
-\citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the ``long-period edge''
-	in the $P_{\rm rot}$-$T_{\rm eff}$ distribution which
-	may pile up over similar timescales \citep{David_2022}.  However, 
-	the stalling is thought to happen at $>$5\,Gyr for
-	late-G dwarfs and early K dwarfs \citep[e.g.~Figure~4
-	by][]{2024ApJ...962..138S}.  This means that it cannot explain the
-	2.3\,Gyr peak in the 4400-5400\,K age distribution shown in Figure~\ref{fig:hist_tgyro}. 
-	Independently,
-	the study by \citet{2022ApJ...937...94M} fitted
-	isochrones to a high-quality subset of Kepler data, and found that the
-	\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly}, does not
-	include many stars older than $\approx$\,4\,Gyr.
-These arguments suggest that the peak
-in the age distribution is not caused by a pile-up of
-much older stars with lethargic spin-down rates, but that it could be
-	contaminated by them beyond $\approx$3\,Gyr, especially for stars hotter than the Sun.
-}
+If stellar rotation rates and spot-induced photometric amplitudes
+remained {\it constant} after $\approx$3\,Gyr, then older stars could
+contaminate the peaks in Figure~\ref{fig:hist_tgyro}.  This
+possibility warrants consideration, given asteroseismic evidence
+suggesting slowed spin-down rates after $\approx$4\,Gyr for Sun-like
+stars \citep[][]{vanSaders_2016,Hall2021,2024ApJ...962..138S}, and the
+``long-period edge'' in the $P_{\rm rot}$-$T_{\rm eff}$ distribution
+which may pile up over similar timescales \citep{David_2022}.
+However, this stalling is thought to happen at $>$5\,Gyr for late-G
+dwarfs and early K dwarfs \citep[e.g.~Figure~4
+by][]{2024ApJ...962..138S}, which means that it cannot account for the
+2.3\,Gyr peak in the 4400-5400\,K age distribution shown in
+Figure~\ref{fig:hist_tgyro}.}
+
+\added{Additionally, \citet{2022ApJ...937...94M} found that the
+\citet{McQuillan_2014} sample, discussed in Appendix~\ref{app:mcqonly},
+includes few stars older than $\approx$4\,Gyr, based on
+isochrone-fitting to a high-quality subset of Kepler planet hosts.
+Some solar twins have also been shown to have rotation periods of
+30-50 days
+\citep{LorenzoOliveira2019,LorenzoOliveira2020,doNascimento2020},
+indicating that a completely stalled spin-down may be unlikely. These
+arguments suggest that while the peak in the age distribution is not
+primarily caused by an accumulation of much older stars with slowed
+spin-down rates, contamination by such stars beyond $\approx$3\,Gyr
+remains possible, particularly for stars hotter than the Sun.}
 

### 2024-09-04 papers/paper/ms.tex
update ms

@@
 assumed planetary, ages are precise at the $\approx$10\% level.  
-An alternative strategy, facilitated by Gaia, is to select transiting
+An alternative strategy, facilitated by Gaia, is to select \deleted{transiting}
 planets based on empirical youth indicators, and to then search their local
-neighborhoods for comoving companions \citep[e.g.][]{Tofflemire_2021}.
+neighborhoods for comoving \added{stellar }companions \citep[e.g.][]{Tofflemire_2021}.
 
@@
 
-Rotation-based ages have been reported for various subsets of Kepler
+Rotational ages have been reported for various subsets of Kepler
 stars since the early data releases
@@
   \citet{Santos_2019,Santos_2021}.  About half of the stars with
-  rotation periods are suitable for gyrochronology (dark gray points),
-  based on factors including their non-binarity and proximity to the
+  rotation periods are suitable for gyrochronology (dark gray\deleted{ points}),
+  based on factors including their \added{apparent }non-binarity and proximity to the
   main sequence (orange line in \replaced{lower}{upper} panel; see
@@
 \citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
-and also for the broader set of all Kepler target stars
+and also for the broader \replaced{set of all}{sample of} Kepler target stars
 \citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021,Reinhold2023}.
@@
 are far from the main sequence.  Some of the challenges this introduces
-for rotation-based ages include unresolved binaries, metallicity,
+for rotational ages include unresolved binaries, metallicity,
 reddening, and drivers of rotation other than magnetic braking.
@@
 %main sequence.
-\added{{\it Metallicity (bit 10)}---Theoretical models
+\added{{\it Metallicity (bit 10)}---Models
 	suggest that stars with non-solar metallicities
@@
 	to verify due to a dearth of nearby non-solar metallicity open clusters.
-	However, if true, this effect could cause systematic
+	However, if true, this could cause systematic
 	biases for gyrochrone ages \citep[e.g.~Figure~9 of][]{Claytor2020}.
@@
 Visual inspection shows that although this flag selects many bona fide
-objects in these classes, it also selects most known young planets
+objects in these classes, it also selects known young planets
 with $P_{\rm rot}$$\lesssim$5\,day.  These objects are
@@
     others.  Horizontal scatter is added to visually clarify the
-    statistical age uncertainties.  At $<$1\,Gyr, our ages are
+    statistical age uncertainties.  Below 1\,Gyr, our ages are
     generally accurate, and their statistical uncertainties match
@@
 
-In Figure~\ref{fig:agescalecompone}, we compare our rotation-based
+In Figure~\ref{fig:agescalecompone}, we compare our rotational
 ages, and available ages from the literature, against the ages of
@@
 on parameters that do not appreciably change at $t$$\lesssim$1\,Gyr.
-For the rotation-based ages in \citet{2023ApJ...952..131M} and
+For the rotational ages in \citet{2023ApJ...952..131M} and
 \citet{2024AJ....167..159L}, the discrepancy is caused by details of
@@
   stars with temperatures of 4400--5400\,K, which \replaced{have more precise
-  ages due to their fast spin-down}{have more precise ages}.  The statistical uncertainty for a
-  single average star at 1, 2, and 3\,Gyr is shown, and is identical across each
+  ages due to their fast spin-down}{have more precise ages}.  The
+  statistical uncertainty for an average star at 1, 2, and 3\,Gyr is
+  shown, and is identical across each
   row.  
@@
 age, we similarly find that fewer systems at $t_{\rm
-gyro}$$<$0.4\,Gyr are detected with $P_{\rm orb}$$\gtrsim$30\,days
+gyro}$$<$0.4\,Gyr are detected with orbital periods beyond 30\,days
 than at $t_{\rm gyro}$$>$0.4\,Gyr.
@@
 this decrease continues linearly into the future, then a simple
-extrapolation would imply that in $\approx$1.4$\pm$0.3\,Gyr the rate
-of new stars forming in the Galaxy will reach zero.
+extrapolation would imply that in $\approx$1.4$\pm$0.3\,Gyr the
+abundance of newly-formed stars in this region of the Galaxy will
+reach zero.
 
@@
 
-Four stellar ensembles in the Kepler field have to date yielded a
-total of fourteen transiting planets.  Our analysis blindly recovers
-the youth of all of these planets.
+Four stellar ensembles in the Kepler field have to date yielded
+\deleted{a total of }fourteen transiting planets.  Our analysis
+blindly recovers the youth of all of these planets.
 
@@
 the rotation and cluster ages agree, and the lithium age is
-2.0$\sigma$ above the cluster age reported by \citet{Bouma_2022b} for
+2.0$\sigma$ above the cluster age \replaced{reported by}{from} \citet{Bouma_2022b} for
 RSG-5.
 
-{\it MELANGE-3}---\citet{Barber_2022} reported a $105\pm10$\,Myr
+{\it MELANGE-3}---\citet{Barber_2022} reported a $105$$\pm$$10$\,Myr
 association in the Kepler field containing two transiting
-mini-Neptunes: Kepler-970 and Kepler-1928.  We find rotation-based
+mini-Neptunes: Kepler-970 and Kepler-1928.  We find rotational
 ages of $t_{\rm gyro}=176_{-40}^{+120}$\,Myr and
 $144_{-88}^{+104}$\,Myr for Kepler-970 and Kepler-1928 respectively.
-The former rotation-based age being slightly older than the suggested
+The former \deleted{rotation-based }age being slightly older than the suggested
 Pleiades age for the association agrees with Figure~4 of
@@
 
-{\it Theia~520}---We find rotation-based ages of $\approx$350\,Myr
+{\it Theia~520}---We find rotational ages of $\approx$350\,Myr
 for Kepler-968 and Kepler-52.  Based on spatial and kinematic
@@
 as UBC-1 \citep{2018A&A...618A..59C}.  Based on isochrone and
-rotation-based age-dating, Theia~520 seems to be
+rotational age-dating, Theia~520 seems to be
 $\approx$230$\pm$70\,Myr old
@@
 Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
-We derive rotation-based ages for these systems of
+We derive rotational ages for these systems of
 \kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
@@
 Kepler planets, and \ncandidateplyounggyrotwosigmanograzingnoruwe\
-candidate planets, all of which are listed in Table~\ref{tab:planets}.
+candidate planets, all \deleted{of which are }listed in Table~\ref{tab:planets}.
 These include \nearthshighq\ objects Earth-sized or smaller,
@@
 Kepler-1312\,c, along with the near-identical Kepler-1561\,b ($t_{\rm
-gyro}$=\kepfifteensixone\,Myr), rank among the youngest Earth-sized
-planets currently known, comparable to planets such as HD~63433\,d
+gyro}$=\kepfifteensixone\,Myr), \replaced{rank among}{could be} the youngest Earth-sized
+planets currently known, \added{and are }comparable to planets such as HD~63433\,d
 \citep[1.1\,$R_\oplus$, $414$$\pm$$23$\,Myr;][]{2024AJ....167...54C}
@@
 		A combination of both kinematic heating and
-		a decreasing SFR are needed to qualitatively match the observed paucity of the youngest stars.
+		a decreasing SFR are needed to qualitatively match the observed
+    paucity of the youngest stars (blue line; as in
+    Figure~\ref{fig:hist_tgyro}.
 		\label{fig:trilegal}
@@
 \citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
-star sample (top row).  However, rotation-based ages are most precise
+star sample (top row).  However, rotational ages are most precise
 for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
@@
 		$P_{\rm rot}$$<$10\,days could be unresolved $\lesssim$0.5\,AU  binaries (see Section~\ref{subsec:caveats}).
-		The black ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
+		The dark blue ``5\% binary contamination'' distribution corrects for this possibility by randomly removing
 		what we estimate (estm.) to be the maximum possible number of such systems; the true
@@
 $P_{\rm rot}$$<$10\,day stars.
-Relative to the nominal ``gyro applicable'' age distribution, there are three important differences.
+Relative to the nominal age distribution, there are three important differences.
 There is a much larger drop in the star formation rate over the past three gigayears,
 by a factor of 4.81$\pm$0.26, rather than \ratiosfr$\pm$\uncratiosfr.
-The rate of decline in star formation
-since $\approx$2.8\,Gyr ago is also constant, rather than having a ``kink'' at 1.5\,Gyr.
+The rate of decline in the age distribution is also constant, rather than having a ``kink'' at 1.5\,Gyr.
 And finally, the corrected distributions show closer agreement with both the SFHs
@@
 with a slight bias toward lower metallicity stars over higher metallicity.
-Based on the study by \citet{Claytor2020},  this will be most important
+Based on work by \citet{Claytor2020},  this will be most important
 for stars near the Kraft break, which could have ages over-estimated
@@
 \citep{2022arXiv220606693G}, and the Roman Galactic Bulge Time Domain
-Survey \citep{Wilson2023}.  Rotation-based ages from these surveys could
+Survey \citep{Wilson2023}.  Rotational ages from these surveys could
 yield new insight
@@
 Despite a slew of stellar quality flags, including on metallicity, evolutionary state,
-and binarity, we expect that up to a third of our ``apparently clean'' 
+and binarity, up to a third of our ``apparently clean'' 
 Kepler stars with $P_{\rm rot}$$<$10\,days
@@
   %
-  \item We have \replaced{derived}{reported} rotation-based ages for \nplyounggyrotwosigma\
+  \item We have \replaced{derived}{reported} rotational ages for \nplyounggyrotwosigma\
     Kepler planets younger than 1\,Gyr at 2$\sigma$, and
@@
 
-\paragraph{Kepler-1644}---The rotation-based age of this system
+\paragraph{Kepler-1644}---The rotational age of this system
 (\kepsixteenfourfour\,Myr) is nominally much lower than the lithium
@@
 therefore point to this system being an unresolved binary, which calls
-the reliability of the rotation-based age into question.  The
+the reliability of the rotational age into question.  The
 non-detection of the companion from Robo-AO imaging at Palomar
@@
 detected in high-resolution imaging \citep{2017AJ....153...66Z}.  The
-rotation-based age is questionable, given that we do not know which
+rotational age is questionable, given that we do not know which
 source the rotation signal is from, or whether the stars have
@@
 Each bar denotes the $\pm$1$\sigma$ uncertainty for a star's
-rotation-based age:
+\deleted{rotation-based }age:
 this plot can be viewed as the transformation of

### 2024-09-04 papers/paper/ms.tex
update ms

@@
   stellar sample.
-  We attribute the existence of this ``demographic cliff'' to a
+  We attribute this ``demographic cliff'' to a
   combination of kinematic heating and a declining star formation rate
@@
 	\end{center}
-	\vspace{-0.5cm}
+	\vspace{-0.53cm}
 	\caption{{\bf Impact of star formation rate (SFR) and kinematic heating on age distribution.}   
@@
     paucity of the youngest stars (blue line; as in
-    Figure~\ref{fig:hist_tgyro}.
+    Figure~\ref{fig:hist_tgyro}).
 		\label{fig:trilegal}
@@
 \citet{SilvaAguirre2018}.  The main novelty of our ages relative to
-\replaced{those studies is their improved accuracy}{these previous studies is their accuracy} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.
+\replaced{those studies is their improved accuracy at $<$}{these previous studies is their accuracy below} 1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.
 

### 2024-09-04 drivers/plot_hall_age_comparison.py
tweak astseis comparison

@@ -8,7 +8,7 @@ from astropy.io import fits
 from astropy.table import Table
 from gyrojo.getters import select_by_quality_bits
 
-def plot_hall_age_comparison(smalllim=0):
+def get_merge():
 
     litdir = join(DATADIR, "literature")
     fitspath = join(litdir, "Hall_2021_asteroseismic.fits")
@@ -30,6 +30,13 @@ def plot_hall_age_comparison(smalllim=0):
     sel = ~pd.isnull(mdf.gyro_median)
     mdf = mdf[sel]
 
+    return mdf
+
+
+def plot_hall_age_comparison(smalllim=0):
+
+    mdf = get_merge()
+
     set_style('science')
     plt.close("all")
 
@@ -38,10 +45,13 @@ def plot_hall_age_comparison(smalllim=0):
     sdf = mdf
 
     mask = select_by_quality_bits(
-        sdf, [0, 1, 2, 3, 4, 5, 6, 7, 8, 9],
-            [0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
+        sdf, [0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10],
+             [0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0]
     )
     sdf['flag_is_gyro_applicable'] = mask
+
+    cols = 'KIC,adopted_Teff,gyro_median,gyro_-1sigma,gyro_+1sigma,Prot,Age,e_age_lc,e_Age,P,__Fe_H_'.split(",")
+    print(sdf[sdf['flag_is_gyro_applicable']][cols].sort_values(by='gyro_median'))
 
     yerr = 1e3*np.array(
         [sdf['e_age_lc'], sdf['e_Age']]

### 2024-09-04 papers/paper/ms.tex
update ms

@@
 by \citet{Barber_2022} and required ``offset'' tangential velocities
-below 2\,\kms.  For Cep-Her, we used candidate members from Kerr et
-al.~2024 (submitted) with $P_{\rm fin}$$>$0.8.  Even with
+below 2\,\kms.  For Cep-Her, we used candidate members from
+\citet{Kerr2024} with $P_{\rm fin}$$>$0.8.  Even with
 contaminants, we can compare the relative age distributions derived by
@@
 for instance has known ``stretches'' and ``compressions'' in its age
-scale relative to observed open clusters \citep[see][Sec.
-7.3]{2023ApJ...952..131M}.  The \citet{Mamajek_2008} calibration used
+scale relative to observed open clusters \citep[see][Sec.~7.3]{2023ApJ...952..131M}.  The \citet{Mamajek_2008} calibration used
 by \citet{Reinhold_2015} appears more accurate than the former two
@@
   row.  
-  Finally, the right panel compares the rotation-derived ages
-  against star formation histories derived using CMD fitting (dashed
+  Finally, the right panel compares the \replaced{rotation-derived}{rotational} ages
+  against star formation histories derived \replaced{using CMD
+  fitting}{from CMD fits to other regions of the Galaxy} (dashed
   and dotted lines).  Completeness in Kepler's $P$$_{\rm rot}$

### 2024-09-04 papers/paper/ms.tex
update ms

@@
 assumed planetary, ages are precise at the $\approx$10\% level.  
-An alternative strategy, facilitated by Gaia, is to select \deleted{transiting}
-planets based on empirical youth indicators, and to then search their local
+An alternative strategy, facilitated by Gaia, is to select \deleted{transiting }planets
+based on empirical youth indicators, and to then search their local
 neighborhoods for comoving \added{stellar }companions \citep[e.g.][]{Tofflemire_2021}.
@@
 \citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
-The implication for known planetary ages can appreciated by querying the NASA Exoplanet Archive
+The implication for known planetary ages can be appreciated by querying the NASA Exoplanet Archive
 \citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
 than 1\,Gyr.  Requiring $t$$<$1\,Gyr at 2$\sigma$ precision gives
-$\approx$50 such planets at the time of writing.  Most of these young
+$\approx$50 \deleted{such }planets at the time of writing.  Most of these young
 planets are in clusters, and were found by Kepler, K2, or TESS.
@@
     found in the Kepler field
-    \cite[e.g.][]{2019AJ....158..122K,Kerr_2021,Barber_2022}.
+    \cite[e.g.][]{2019AJ....158..122K,Barber_2022,Kerr2024}.
     These clusters offer stars that we can use to test

### 2024-10-07 papers/paper/ms.tex
aas journals accepted

@@
 \revised{2024 September 4}
-\accepted{---}
+\accepted{2024 October 7}
 \shorttitle{Kepler's Demographic Cliff}
@@
   exoplanets.}
-  %
   \added{The age distribution that we observe, rather than being uniform,
@@
   eagles \citep{Jeffries_2023},
-  gyro-interp \citep{Bouma_2023},
+  gyro-interp (\citealt{Bouma_2023}; \texttt{v0.6} available at doi:~\href{https://doi.org/10.5281/zenodo.13733242}{10.5281/zenodo.13733242}),
   matplotlib \citep{matplotlib},

### 2024-10-07 drivers/make_star_table.py

@@ -175,12 +175,24 @@ def make_star_table(
     print(f'Wrote {csvpath}')
 
     csvpath = join(PAPERDIR, 'table_star_gyro_agesformatted.csv')
-    # format ages
+    # Format ages
     sel = sdf.gyro_median.astype(float) > 4000
     sdf.loc[sel, 't_gyro'] = '$> 4000$'
     sdf.loc[sel, 'gyro_median'] = np.nan
     sdf.loc[sel, 'gyro_+1sigma'] = np.nan
     sdf.loc[sel, 'gyro_-1sigma'] = np.nan
+
+    # Remove "t_li" and "t_gyro" latex columns (which have the redunant
+    # values), and  instead report "t_li_lowerlimit" and "t_gyro_lowerlimit"
+    # columns
+    sdf['t_gyro_lowerlimit'] = ''
+    sel = sdf['t_gyro'].str.contains(">")
+    sdf.loc[sel, 't_gyro_lowerlimit'] = (
+        sdf.loc[sel, 't_gyro'].str.lstrip("$").str.rstrip("$")
+    )
+    sdf = sdf.drop(columns=['t_gyro'])
+
+    # Write
     sdf.to_csv(csvpath, index=False, na_rep='')
     print(f'Wrote {csvpath}')
 

### 2024-10-07 drivers/make_younghighlight_table.py

@@ -488,7 +488,7 @@ def make_table(
         outdf.to_csv(csvpath, index=False)
         print(f'Wrote {csvpath}')
 
-        # format ages
+        # Format ages
         sel = outdf['t_li'].str.contains(">")
         outdf.loc[sel, 'li_median'] = np.nan
         outdf.loc[sel, 'li_+1sigma'] = np.nan
@@ -499,6 +499,23 @@ def make_table(
         outdf.loc[sel, 'gyro_median'] = np.nan
         outdf.loc[sel, 'gyro_+1sigma'] = np.nan
         outdf.loc[sel, 'gyro_-1sigma'] = np.nan
+
+        # Remove "t_li" and "t_gyro" latex columns (which have the redunant
+        # values), and  instead report "t_li_lowerlimit" and "t_gyro_lowerlimit"
+        # columns
+        outdf['t_gyro_lowerlimit'] = ''
+        sel = outdf['t_gyro'].str.contains(">")
+        outdf.loc[sel, 't_gyro_lowerlimit'] = (
+            outdf.loc[sel, 't_gyro'].str.lstrip("$").str.rstrip("$")
+        )
+        outdf = outdf.drop(columns=['t_gyro'])
+
+        outdf['t_li_lowerlimit'] = ''
+        sel = outdf['t_li'].str.contains(">")
+        outdf.loc[sel, 't_li_lowerlimit'] = (
+            outdf.loc[sel, 't_li'].str.lstrip("$").str.rstrip("$")
+        )
+        outdf = outdf.drop(columns=['t_li'])
 
         csvpath = join(PAPERDIR, 'table_allageinfo_agesformatted.csv')
         outdf.to_csv(csvpath, index=False)

### 2024-10-08 papers/paper/ms.tex
arxiv upload

@@
 % Redefine footnote command to not have indent
-%\makeatletter
-%\long\def@makefntext#1{\noindent\footnotesize@makefnmark#1}
-%\makeatother
 \setlength{\footnotesep}{0pt}
@@
 
-%\title{Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field}
-%\title{A Cliff In The Age Distribution of Young Stars and Planets in the Kepler Field}
-%\title{A Paucity of Young Stars and Planets in the Kepler Field}
-%\title{A Non-Uniform Age Distribution of Stars and Planets in the Kepler Field}
-%\title{}
-% Rotational and Lithium-Based Ages of Stars and Planets in the Kepler Field
-% Stars And Planets Younger Than Four Billion Years In The Kepler Field
-%\title{Sub-Solar Aged Stars and Planets in the Kepler Field}
-%From Youth to Maturity: Ages of Stars and Planets in the Kepler Field
-%Ages of Stars and Planets in the Kepler Field Over The First Four Billion Years
-
-\title{Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years}
+\title{Ages of Stars and Planets in the Kepler Field Younger Than Four Billion Years}
 
 \correspondingauthor{Luke G. Bouma}
-\email{bouma.luke@gmail.com}
+\email{lbouma@carnegiescience.edu}
 
@@
 include HIP~67522\,b, a Jupiter-sized planet with a sub-Neptunian mass
-(\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
+(\citealt{Rizzuto_2020}; \citealt{Thao2024}), and V1298~Tau, a
 resonant chain of puffy planets that is a likely precusor to the
@@
 \citep[][]{2000MNRAS.318..658B,Nordstrom_2004}.  This approximation
-suggests that $\approx$1\% of nearby stars have $t$$<$100\,Myr, and
+suggests that $\approx$1\% of nearby stars have ages $t$$<$100\,Myr, and
 $\approx$10\% have $t$$<$1\,Gyr.  Studies of currently forming
@@
 
-%\subsection{Sample Summary}
-%\label{subsec:tally}
-%
-%The rotation periods for the star and planet samples are shown in
-%Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
-%observed by Kepler, requiring reported rotation periods and a reliable
-%DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
-%these stars to have temperatures of 3800--6200\,K yields
-%\nuniqstarsantosrotteffcut\ stars.  Further requiring the stars to be
-%apparently single, near the main sequence, with $\log g$$>$4.2,
-%$M_{\rm G}>3.9$, and not flagged as eclipsing binaries yielded
-%\nuniqstarsantosallbutruwe\ rotators.
-%Finally, imposing the full set of quality cuts,
-%i.e.~further requiring RUWE$<$1.4 and requiring no
-%resolved blends ($\Delta G$<2.5) within 4$''$ leaves
-%\nuniqstarsantosrotgyroappl\ stars for which we expect rotation-based
-%age dating to be secure.
-%
-%
-%In the planet sample, the cumulative KOI table included
-%\nnonfpcumkois\ confirmed and candidate planets, orbiting
-%\nnonfpcumkoihosts\ stars.  Assume that we apply all the same stellar
-%cuts, save for that on RUWE.  We can then require these confirmed and
-%candidate planets to have MES$>$10.  This combination of selection
-%criteria yield \nplwgyroagewithgrazingandhighruwe\ planets with
-%rotation-based ages orbiting
-%\nplhoststarwgyroagewithgrazingandhighruwe\ stars
-%(\nplhoststarwgyroagejusthighruwe\ of these stars with ``high'' RUWE).
-%If we additionally require the planets to be non-grazing ($b<0.8$), we
-%are left with \nplwgyroagenograzing\ planets orbiting
-%\nplhoststarwgyroagenograzing\ stars with rotatio-based ages.
-%This cut drops $\approx$41 more planets than one would expect if the initial sample
-%of \nplwgyroagewithgrazingandhighruwe\ planets had zero false positives.
-%The binomial probability that this could happen by coincidence is 0.1\%.
-
 \section{Cluster Age Comparison}
@@
     }
-%
-%	\vspace{-0.35cm}
-%	\subfloat{
-%		\includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240821_maxage4000_dropfracshortrot.pdf}
-%	}
   \end{center}
@@
 	degree by which the SFR in the Kepler field has decreased, by 7-14\%. }
-
-%\added{Other age indicators may help in verifying the ages of the
-%$\lesssim$3\,Gyr stars that are the main focus of this work.  Specifically,
-%chromospheric emission in the X-ray, \ion{Ca}{2} HK line, and the UV
-%can serve as an age tracer
-%\citep{Mamajek_2008,2014MNRAS.441.2361V,2024ApJ...960...62E}.
-%However, these age indicators are all in a sense ``rotation-powered''.
-%The dynamo converts kinetic energy into magnetic energy, which is
-%emitted through these chromospheric pathways.  We did not attempt to
-%incorporate these indicators into this study due to concerns regarding
-%sensitivity, homogeneity, and the question of whether they in fact
-%provide age information that is truly independent from rotation.}
 
@@
 This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
-(LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
+(LGB, EKP), the Carnegie Institution for Science's Carnegie Fellowship (LGB),
+and the Arthur R.~Adams SURF Fellowship (EKP).
 The HIRES data were obtained at the Keck Observatory.
